### 2011-07-27 Docs/Preproposal/preproposal.tex
Preproposal (1st draft) and template

\documentclass[11pt]{artikel3}
\usepackage{fullpage, setspace, graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{times}

\title{RIT Department of Computer Science\\MSc Project/Thesis Pre-Proposal:\\\emph{Implementation of a Push Based Signal Function FRP System and Evaluation Interface}}
\author{Edward Amsden}
\date{\today}

\begin{document}
\maketitle

The sections shown below are adapted from the topic analysis forms provided in ``Writing the Doctoral Dissertation" (2nd edition) by Davis and Parker (pages 82-88). Your final document should be 1-2 pages including references. {\bf The final pre-proposal may present the items below in any format, but using prose (not bulleted lists).}

\section{Problem Description}
Most of the work in Functional Reactive Programming centers around the construction and evaluation of FRP programs, and does not concern itself with the interface between FRP programs and the rest of the world. In fact, most literature on FRP ignores completely the problem  evaluation interface for FRP programs.

The signal function family of FRP semantics provides distinct advantages over the alternative signal/behavior family. These advantages include explicit inputs and the elimination of large classes of space and time leaks. However, there does not yet exist an implementation that provides "push-based" semantics (i.e. evaluation on event occurrences, rather than sampling for events) for signal functions.

Is it possible to create a push-based signal function FRP system with an intuitive and useful interface?

\section{Importance of Research}
FRP provides a declarative means of specifying reactive systems. This enables such systems to be created compositionally, and permits formal reasoning about such systems. However,
for FRP to be useful, there must be a useful way of evaluating programs with outside input and permitting the output of the FRP system to affect the outside world. This concern has yet to be explored
in the literature. An FRP system with a useful evaluation interface would provide a complete method of declaratively specifying reactive systems, specifying the inputs to a system, and specifying
what should be done with the output of the system.

\section{Related Work}

FRP is a well-explored field. There is a wealth of research on the best semantics and most efficient implementation of FRP systems. I previously surveyed
the literature of FRP for independent study credit~\cite{Amsden2011}. The proposed research would provide an efficient implementation of an elegant family of FRP semantics,
and a compositional and intuitive means of providing input to and using output from such systems.

\section{Methodology}
\label{sec:methodology}

I will utilize the well-explored signal function model for FRP, extended with the signal vector constructs described by Sculthorpe~\cite{Sculthorpe2011} to permit the system to distinguish events
and signals at the type level. I will explore the creation of a monadic interface which maintains the current FRP expression and other information such as event output actions as state, and provides
actions for pushing an event to an input of the FRP expression, incrementing the time, and sampling a signal output of the FRP expression at the current time.

I will make use of the HList type-level list construct to express signal vectors, and to provide "addressing" for inputs and outputs of the FRP expression.

I will consider making use of Template Haskell to smooth over the rather verbose type expressions necessitated by HLists, and to provide simple access to the type-level boilerplate necessary to permit addressing of the inputs and outputs of an FRP expression.

A successful FRP abstraction and evaluation interface will permit the user of the abstraction to control the steps of evaluation, permit a decoupling between supplying input and acting on signal output, permit the specification of actions to be taken whenever an event occurrence reaches an event output, not require access to the IO system of the host language (Haskell), provide a compositional interface to the IO system of the host language if the user of the abstraction wishes (for instance, a MonadIO instance), utilize a "push" (evaluate when input is supplied) rather than "pull" (continuously poll for input) strategy for evaluation, and possibly provide a low-overhead (in terms of code and computational cost) interface for selecting between multiple inputs to push events to and outputs to sample signal values from for the FRP expression being evaluated.

\section{Potential Outcomes}

\begin{itemize}
\item An interface meeting the requirements in Section \ref{sec:methodology} is created and demonstrated.
\item It is discovered that the type system of Haskell is not expressive enough to permit such an interface.
\item It is discovered that signal function semantics cannot support push-based evaluation
\item Unforeseen challenges in a lack of expressivity in the language or the signal function semantics prevent the creation of the interface. 
\end{itemize}

\end{document}

### 2011-07-27 Docs/Preproposal/preproposalTemplate.tex
Preproposal (1st draft) and template

\documentclass[11pt]{artikel3}
\usepackage{fullpage, setspace, graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{times}

\title{RIT Department of Computer Science\\MSc Project/Thesis Pre-Proposal:\\\emph{Proposed Project/Thesis Title}}
\author{FirstName LastName}
\date{\today}

\begin{document}
\maketitle

The sections shown below are adapted from the topic analysis forms provided in ``Writing the Doctoral Dissertation" (2nd edition) by Davis and Parker (pages 82-88). Your final document should be 1-2 pages including references. {\bf The final pre-proposal may present the items below in any format, but using prose (not bulleted lists).}

\section{Problem Description}
Identify what problem you are addressing, both in terms of the research area, and the \emph{specific} problem that you will be working on:
\begin{itemize}
	\item For a thesis, a hypothesis (`thesis statement') that you will test in your research.
	\item For a project, identify the work required (e.g. implementation and/or experiment) that needs to be completed. If you are completing a project, make sure to speak with your advisor about the expected deliverable; one deliverable will be a written project report.
\end{itemize}

\section{Importance of Research}
Motivate your problem. 
\begin{itemize}
	\item What is the significance of your problem? 
	\item What applications or new opportunities will solving your problem provide?
\end{itemize}

\section{Related Work}

Demonstrate the connection between your chosen problem and how it is related to existing work. 
\begin{itemize}
	\item What are the key theoretical models (e.g. process-based, formal language/complexity models, probability-based) and algorithms have been applied toward this problem previously? 
	\item What limitation and/or opportunity do you plan to address in your project/thesis?
	\item
In the related research literature, how is success measured (e.g. metrics and/or coverage of problem aspects)?
\end{itemize}

\section{Methodology}

What theory, model, or algorithm do you plan to modify or develop to address your research problem?
\begin{itemize}
\item What methods/techniques will you use to address your problem? 
\begin{itemize}
\item For theory-based projects and theses, what are the key theorems to be developed and/or proven? What proof techniques will be used?
\item For projects and theses involving experiments, what metrics
will you use to measure success? Commonly these include some subset of time, space, and accuracy (recognition rate, precision, recall, etc.).
\end{itemize}

\item How you will you measure success? Almost always, this should include reference to the evaluation methods described in the related work.

\item {\bf How will you know when you are done?}
\end{itemize}

\section{Potential Outcomes}

\begin{itemize}
\item Given your chosen methods, what are the possible outcomes of the work? 
\item What is the contribution/significance of the result for each outcome? 
\end{itemize}

\bibliographystyle{plain}
\bibliography{preproposal}

**(Omitted) As an exercise, modify this document to include the references in the {\tt plain.bib} file.

\end{document}

### 2011-08-25 Docs/Preproposal/preproposal.tex
Preproposal (first draft)

@@
 \usepackage{times}
+\usepackage{url}
 
@@
 \section{Problem Description}
-Most of the work in Functional Reactive Programming centers around the construction and evaluation of FRP programs, and does not concern itself with the interface between FRP programs and the rest of the world.
-In fact, most literature on FRP ignores completely the problem  evaluation interface for FRP programs.
-
 The signal function family of FRP semantics provides distinct advantages over the alternative signal/behavior family.
@@
 However, there does not yet exist an implementation that provides "push-based" semantics (i.e.
-evaluation on event occurrences, rather than sampling for events) for signal functions.
+evaluation on event occurrences, rather than polling for events) for signal functions.
 
-Is it possible to create a push-based signal function FRP system with an intuitive and useful interface?
+Most of the work in Functional Reactive Programming centers around the construction and evaluation of FRP programs, and does not concern itself with the interface between FRP programs and the rest of the world.
+In fact, most literature on FRP ignores completely the problem of evaluation interface for FRP programs.
+
+Do a monadic evaluation interface and signal vector typing enable a useful push-based FRP system?
 
 \section{Importance of Research}
-FRP provides a declarative means of specifying reactive systems. This enables such systems to be created compositionally, and permits formal reasoning about such systems. However,
-for FRP to be useful, there must be a useful way of evaluating programs with outside input and permitting the output of the FRP system to affect the outside world. This concern has yet to be explored
-in the literature. An FRP system with a useful evaluation interface would provide a complete method of declaratively specifying reactive systems, specifying the inputs to a system, and specifying
-what should be done with the output of the system.
+Functional reactive programming enables a denotative method of defining reactive systems. The efficient evaluation
+of such systems is still an open problem. Signal function style FRP semantics have several advantages over
+classic FRP semantics, but current implementations use ``pull'' based evaluation, which has the undesirable consequences of
+repeated unnecessary work in evaluating event non-occurrences, forcing input to be synchronous with output, and introducing sampling latency into event occurrences.
+
+A signal function FRP system which uses ``pull'' based evaluation would eliminate these consequences, permitting more
+efficient implementation with only the latency of processing, rather than the sampling interval. It would also permit
+the decoupling of the input and output of the signal function network.
 
@@
 
-FRP is a well-explored field. There is a wealth of research on the best semantics and most efficient implementation of FRP systems. I previously surveyed
-the literature of FRP for independent study credit~\cite{Amsden2011}. The proposed research would provide an efficient implementation of an elegant family of FRP semantics,
-and a compositional and intuitive means of providing input to and using output from such systems.
+FRP is a well-explored field.
+There is a wealth of research on the best semantics and most efficient implementation of FRP systems.
+I previously surveyed the literature regarding FRP for independent study credit~\cite{Amsden2011}.
+The proposed research would provide an efficient implementation of an elegant family of FRP semantics, and a compositional and intuitive means of providing input to and using output from such systems.
+
+Signal function FRP or Arrowized FRP (AFRP) is one of two main families of semantics for Functional Reactive 
+Programming~\cite{Nilsson2002}. It was recently extended with additional typechecked properties using the notion of signal 
+vectors.
+
+Push based FRP has been demonstrated for classic FRP semantics~\cite{Elliott2009}, but it has not yet been explored in the context of signal function semantics.
 
@@
 
-I will utilize the well-explored signal function model for FRP, extended with the signal vector constructs described by Sculthorpe~\cite{Sculthorpe2011} to permit the system to distinguish events
-and signals at the type level. I will explore the creation of a monadic interface which maintains the current FRP expression and other information such as event output actions as state, and provides
-actions for pushing an event to an input of the FRP expression, incrementing the time, and sampling a signal output of the FRP expression at the current time.
+I will explore an implementation which permits the separation of evaluation of continuous values and events.
+The implementation will hold the current signal function network as state in the monadic computation described using the evaluation monad.
+Event occurrences provided as input or produced within the network will mutate the network state.
 
-I will make use of the HList type-level list construct to express signal vectors, and to provide "addressing" for inputs and outputs of the FRP expression.
+The evaluation monad will provide monadic actions for ``pushing'' an event occurrence on an input to the signal function network, stepping time and evaluating continous values, and sampling continuous output values.
 
-I will consider making use of Template Haskell to smooth over the rather verbose type expressions necessitated by HLists, and to provide simple access to the type-level boilerplate necessary to permit addressing of the inputs and outputs of an FRP expression.
+Signal function networks will be described by a GADT using HLists~\cite{Kiselyov2004} and phantom types (Event $\alpha$ and Signal $\alpha$) to describe signal vectors.
 
-A successful FRP abstraction and evaluation interface will permit the user of the abstraction to control the steps of evaluation, permit a decoupling between supplying input and acting on signal output, permit the specification of actions to be taken whenever an event occurrence reaches an event output, not require access to the IO system of the host language (Haskell), provide a compositional interface to the IO system of the host language if the user of the abstraction wishes (for instance, a MonadIO instance), utilize a "push" (evaluate when input is supplied) rather than "pull" (continuously poll for input) strategy for evaluation, and possibly provide a low-overhead (in terms of code and computational cost) interface for selecting between multiple inputs to push events to and outputs to sample signal values from for the FRP expression being evaluated.
+Possibilities to be explored for implementation include ``callbacks'' to be activated on output event occurrences, or on time
+steps for continuous output, and a means of addressing multiple inputs or outputs to a signal function network.
 
@@
 
-\begin{itemize}
-\item An interface meeting the requirements in Section \ref{sec:methodology} is created and demonstrated.
-\item It is discovered that the type system of Haskell is not expressive enough to permit such an interface.
-\item It is discovered that signal function semantics cannot support push-based evaluation
-\item Unforeseen challenges in a lack of expressivity in the language or the signal function semantics prevent the creation of the interface. 
-\end{itemize}
+I expect to produce an implementation of signal function semantics that permits push-based evaluation.
+I expect that this implementation will include both constructs for producing a signal function network and a monadic interface for evaluating a signal function network.
+The constructs will include routing combinators, switching combinators (which will introduce information denoting where mutations to the network by events will take place), and stateful signal functions such as
+numeric integration.
+
+It is possible that this interface will also permit multiple inputs to and outputs from a signal function network.
+
+\bibliography{preproposal}{}
+\bibliographystyle{acm}
 

### 2011-08-30 Docs/Preproposal/preproposal.tex
Preproposal draft

@@
 
-\title{RIT Department of Computer Science\\MSc Project/Thesis Pre-Proposal:\\\emph{Implementation of a Push Based Signal Function FRP System and Evaluation Interface}}
+\title{RIT Department of Computer Science\\MSc Thesis Pre-Proposal:\\\emph{Implementation of a Push-Based Signal-Function FRP System and Evaluation Interface}}
 \author{Edward Amsden}
@@
 \maketitle
-
-The sections shown below are adapted from the topic analysis forms provided in ``Writing the Doctoral Dissertation" (2nd edition) by Davis and Parker (pages 82-88).
-Your final document should be 1-2 pages including references.
-{\bf The final pre-proposal may present the items below in any format, but using prose (not bulleted lists).}
 
@@
 
-Do a monadic evaluation interface and signal vector typing enable a useful push-based FRP system?
+I intend to show that a monadic evaluation interface and signal vector typing enable a useful push-based FRP system.
 
@@
 of such systems is still an open problem. Signal function style FRP semantics have several advantages over
-classic FRP semantics, but current implementations use ``pull'' based evaluation, which has the undesirable consequences of
-repeated unnecessary work in evaluating event non-occurrences, forcing input to be synchronous with output, and introducing sampling latency into event occurrences.
+classic FRP semantics, including the elimination of a large class of space and time leaks, as well as explicit input. However,
+current implementations use ``pull'' based evaluation, leading to undesirable consequences:
+repeating unnecessary work in evaluating event non-occurrences, forcing input to be synchronous with output, and introducing sampling latency into event occurrences.
 
-A signal function FRP system which uses ``pull'' based evaluation would eliminate these consequences, permitting more
+A signal function FRP system which uses ``push''-based evaluation would eliminate these negatives, permitting more
 efficient implementation with only the latency of processing, rather than the sampling interval. It would also permit
@@
 FRP is a well-explored field.
-There is a wealth of research on the best semantics and most efficient implementation of FRP systems.
+There is a wealth of research on the best semantics and most efficient implementation of FRP systems, though almost no discussion or exploration of the evaluation interface (the means of providing input and sampling output).
 I previously surveyed the literature regarding FRP for independent study credit~\cite{Amsden2011}.
@@
 Programming~\cite{Nilsson2002}. It was recently extended with additional typechecked properties using the notion of signal 
-vectors.
+vectors~\cite{Sculthorpe2011}.
 
@@
 
-I will explore an implementation which permits the separation of evaluation of continuous values and events.
+I will explore a semantic model of signal function FRP which considers the signal function network as state and events
+as mutators of the network. The network can of course evaluate events as well.
+
+The exploration of this model will lead to the design of an implementation which permits the separation of evaluation of
+continuous values and discrete events.
 The implementation will hold the current signal function network as state in the monadic computation described using the evaluation monad.
@@
 The evaluation monad will provide monadic actions for ``pushing'' an event occurrence on an input to the signal function network, stepping time and evaluating continous values, and sampling continuous output values.
+The evaluation monad is
+a key component of the design. It permits much finer control over FRP evaluation that in possible with existing models and
+implementations, and enables the push-based evaluation of the signal function network by threading state.
 
@@
 
-I expect to produce an implementation of signal function semantics that permits push-based evaluation.
+I expect to produce an implementation of signal function semantics that utilizes push-based evaluation.
 I expect that this implementation will include both constructs for producing a signal function network and a monadic interface for evaluating a signal function network.
@@
 
-It is possible that this interface will also permit multiple inputs to and outputs from a signal function network.
+I expect to explore type-safe means of permitting multiple inputs and outputs.
 

### 2011-08-30 Docs/Preproposal/preproposal.tex
proposed final draft of preproposal

@@
 \usepackage{fullpage, setspace, graphicx}
-\usepackage[margin=1in]{geometry}
+\usepackage[margin=.75in]{geometry}
 \usepackage{times}
@@
 
-\title{RIT Department of Computer Science\\MSc Thesis Pre-Proposal:\\\emph{Implementation of a Push-Based Signal-Function FRP System and Evaluation Interface}}
+\title{RIT Department of Computer Science\\MSc Thesis Pre-Proposal:\\\emph{Design and Implementation of a Push-Based Signal-Function FRP System}}
 \author{Edward Amsden}
@@
 
-I intend to show that a monadic evaluation interface and signal vector typing enable a useful push-based FRP system.
+I intend to show that a monadic evaluation interface and signal vector typing enable a useful and efficient push-based FRP system.
 
@@
 
+The implementation will be compared against existing pull-based signal function FRP systems.
+
 \section{Potential Outcomes}

### 2011-10-20 Code/TypeFamilies/ReactiveFunctions.hs
expressing routing with type families, fails on copy case (can't use type function from vector to route and instance for combining routes)

@@ -1,0 +1,49 @@
+{-# LANGUAGE GADTs, MultiParamTypeClasses, RankNTypes, FlexibleInstances, FunctionalDependencies #-}
+module ReactiveFunctions where
+
+import Routes
+import ReactiveVectors
+
+-- Event pushing
+class PushEvent rf v1 v2 
+  pushEvent :: rf v1 v2 -> EventRoute v1 -> (rf v1 v2, EventRoute v2)
+  
+-- Wrapper type for reactive functions
+data RF v1 v2 = forall c . PushEvent c v1 v2 => RF (c v1 v2)
+
+instance PushEvent RF v1 v2 where
+  pushEvent (RF f) r = let (f', r') = pushEvent f r 
+                       in (RF f', r')
+
+-- Reactive Function constructors
+data RFPureSignal v1 v2 where
+  RFPureSignal :: (a -> b) -> RFPureSignal (SVSignal a) (SVSignal b)
+  
+data RFPureEvent v1 v2 where
+  RFPureEvent :: (a -> b) -> RFPureEvent (SVEvent a) (SVEvent b)
+  
+data RFSequence v1 v2 where
+  RFSequence :: RF v1 v2 -> RF v2 v3 -> RFSequence v1 v3
+  
+data RFCopy v1 v2 where
+  RFCopy :: RFCopy v1 (SVAppend v1 v1)
+
+  
+-- Event pushing for reactive functions
+instance PushEvent RFPureSignal (SVSignal a) (SVSignal b) (EventRoute (SVSignal a)) (EventRoute (SVSignal b))where
+  pushEvent sf _ = (sf, EventRouteNowhere)
+  
+instance PushEvent RFPureEvent (SVEvent a) (SVEvent b) where
+  pushEvent ef EventRouteDoneNowhere = (ef, EventRouteDoneNowhere)
+  pushEvent ef@(RFPureEvent f) (EventRouteDone x) = (ef, EventRouteDone (f x))
+  
+instance PushEvent RFSequence v1 v2 where
+  pushEvent (RFSequence rf1 rf2) r = let (rf1', r') = pushEvent rf1 r
+                                         (rf2', r'') = pushEvent rf2 r'
+                                     in (RFSequence rf1' rf2', r'')
+{-                                        
+-- Fails with "no instance for EventRoutesCombine (EventRoute v1) (EventRoute v1)"
[…]

### 2011-10-20 Code/TypeFamilies/ReactiveVectors.hs

@@ -1,0 +1,15 @@
+{-# LANGUAGE EmptyDataDecls, TypeFamilies, UndecidableInstances #-}
+module ReactiveVectors where
+
+import Routes
+
+data SVEmpty
+data SVSignal a
+data SVEvent a
+data SVAppend v1 v2
+
+type family EventRoute v
+type instance EventRoute SVEmpty = EventRouteNowhere
+type instance EventRoute (SVSignal a) = EventRouteNowhere
+type instance EventRoute (SVEvent a) = EventRouteDone a
+type instance EventRoute (SVAppend v1 v2) = CombineEventRoutes (EventRoute v1) (EventRoute v2)

### 2011-10-20 Code/TypeFamilies/Routes.hs

@@ -1,0 +1,308 @@
+{-# LANGUAGE TypeFamilies, MultiParamTypeClasses #-}
+module Routes where
+
+-- Event Routes
+data EventRouteNowhere    = EventRouteNowhere
+
+data EventRouteDone a     = EventRouteDone a | 
+                            EventRouteDoneNowhere
+                            
+data EventRouteFirst r    = EventRouteFirst r |
+                            EventRouteFirstNowhere
+                            
+data EventRouteSecond r   = EventRouteSecond r | 
+                            EventRouteSecondNowhere
+                            
+data EventRouteBoth r1 r2 = EventRouteBoth r1 r2 |
+                            EventRouteOnlyFirst r1 |
+                            EventRouteOnlySecond r2 |
+                            EventRouteNeither
+                            
+-- Event Route Combination
+class EventRoutesCombine r1 r2 where
+  type CombineEventRoutes r1 r2
+  combineEventRoutes :: r1 -> r2 -> CombineEventRoutes r1 r2
+  eventRouteFirst :: CombineEventRoutes r1 r2 -> r1
+  eventRouteSecond :: CombineEventRoutes r1 r2 -> r2
+  
+instance EventRoutesCombine EventRouteNowhere EventRouteNowhere where
+  type CombineEventRoutes EventRouteNowhere EventRouteNowhere = EventRouteNowhere
+  combineEventRoutes _ _ = EventRouteNowhere
+  eventRouteFirst _ = EventRouteNowhere
+  eventRouteSecond _ = EventRouteNowhere
+  
+instance EventRoutesCombine EventRouteNowhere (EventRouteDone a) where
+  type CombineEventRoutes EventRouteNowhere (EventRouteDone a) = EventRouteSecond (EventRouteDone a)
+  combineEventRoutes _ EventRouteDoneNowhere = EventRouteSecondNowhere
+  combineEventRoutes _ r = EventRouteSecond r
+  eventRouteFirst _ = EventRouteNowhere
+  eventRouteSecond (EventRouteSecond r) = r
+  eventRouteSecond _ = EventRouteDoneNowhere
+  
[…]

### 2011-10-20 Code/old/SignalFunction.hs
pre-work

@@ -1,0 +1,100 @@
+{-# LANGUAGE EmptyDataDecls, GADTs, MultiParamTypeClasses, FlexibleInstances, UndecidableInstances, FunctionalDependencies, RankNTypes #-}
+module SignalFunction where
+
+-- | An empty type vector
+data EmptyVector
+-- | A single signal
+data Signal a
+-- | A single event
+data Event a
+-- | The concatenation of two vectors
+data AppendVector v1 v2
+
+{-
+instance ReactiveVector EmptyVector
+instance ReactiveVector (Signal a)
+instance ReactiveVector (Event a)
+instance (ReactiveVector v1, ReactiveVector v2) => ReactiveVector (AppendVector v1 v2)
+instance Singleton (Signal a)
+instance Singleton (Event a)
+instance SignalVector EmptyVector
+instance SignalVector (Signal a)
+instance (SignalVector v1, SignalVector v2) => SignalVector (AppendVector v1 v2)
+instance EventVector EmptyVector
+instance EventVector (Event a)
+instance (EventVector v1, EventVector v2) => EventVector (AppendVector v1 v2)
+
+-- | The head of a type level reactive vector
+class VectorHead v e | v -> e
+instance VectorHead (Signal a) (Signal a)
+instance VectorHead (Event a) (Event a)
+instance (VectorHead v e) => VectorHead (AppendVector v v') e 
+instance (VectorHead v e) => VectorHead (AppendVector EmptyVector v) e
+
+-- | The tail of a type level reactive vector
+class VectorTail v vt | v -> vt 
+instance VectorTail (Signal a) EmptyVector
+instance VectorTail (Event a) EmptyVector
+instance VectorTail (AppendVector (Signal a) v2) v2
+instance VectorTail (AppendVector (Event a) v2 ) v2
+instance (VectorTail v2 vt) => VectorTail (AppendVector EmptyVector v2) vt
+instance (VectorTail v1 vt) => VectorTail (AppendVector v1 v2) (AppendVector vt v2)
+
+-- | Equivalence test for two vectors
+class (ReactiveVector v1, ReactiveVector v2) => VectorEquivalent v1 v2
+instance VectorEquivalent (EmptyVector) (EmptyVector)
+instance VectorEquivalent (Signal a) (Signal a)
+instance VectorEquivalent (Event a) (Event a)
[…]

### 2011-10-20 Code/old/SignalVector.hs

@@ -1,0 +1,61 @@
+{-# LANGUAGE EmptyDataDecls, GADTs, MultiParamTypeClasses, RankNTypes #-}
+
+module Marrow where
+
+import Control.Monad
+import Control.Monad.State.Lazy
+import Data.HList
+
+data Event l a
+data Signal l a
+
+data NoLabel
+
+class Reactive a
+instance Reactive Event
+instance Reactive Signal
+
+-- We don't care about labels for equivalent signal vectors
+class SignalVectorEquivalent v1 v2
+instance SignalVectorEquivalent HNil HNil
+instance (SignalVectorEquivalent v1 v2) => SignalVectorEquivalent (HCons (Event l1 a) v1) (HCons (Event l2 a) v2)
+
+-- Reactive vectors: lists of reactive types
+class (HList a) => ReactiveVector a
+instance ReactiveVector HNil
+instance (Reactive r, ReactiveVector v) = ReactiveVector (HCons (r l a) v)
+
+-- Event vectors: lists of event-only types
+class EventVector a
+instance EventVector HNil
+instance (EventVector v) => EventVector (HCons (Event l a) v)
+
+-- Signal vectors
+class SignalVector a
+instance SignalVector HNil
+instance (SignalVector v) => SignalVector (HCons (Event l a) v)
+
+-- Only the event pieces of an HList
+class (HList l, HList l') => EventsOf l l' | l -> l'
+instance EventsOf HNil HNil
+instance (EventsOf l l') => EventsOf (HCons (Event l a) l) (HCons (Event l a) l')
+instance (EventsOf l l') => EventsOf (HCons (Signal l a) l) l'
+
+-- Only the signal pieces of an HList
+class (HList l, HList l') => SignalsOf l l' | l -> l'
+instance SignalsOf HNil HNil
+instance (SignalOf l l') => SignalsOf (HCons (Signal l a) l) (HCons (Signal l a) l')
+instance (SignalOf l l') => SignalsOf (HCons (Event l a) l) l'
+
+
+-- Reactive Function type
+data (ReactiveVector v1, ReactiveVector v2) => ReactiveFunction v1 v2 where
+  -- Pure functions
+  PureSignal :: (SignalVector v1, SignalVector v2, TypeVector v1 t1, TypeVector t2 v2) => (t1 -> t2) -> ReactiveFunction v1 v2
+
+  
[…]

### 2011-10-20 Code/old/TypedCombinators.hs

@@ -1,0 +1,142 @@
+{-# LANGUAGE FlexibleInstances, MultiParamTypeClasses, FunctionalDependencies, EmptyDataDecls, GADTs #-}
+module TypedCombinators where
+
+data Signal a
+data Event a
+data Empty
+data AppendV v1 v2
+
+-- Reifying dataflow in the type system
+data Nowhere = Nowhere
+data Done a = Done a | DoneNowhere
+data First r = First r | FirstNowhere
+data Second r = Second r | SecondNowhere
+data Both r1 r2 = Both r1 r2 | FirstOnly r1 | SecondOnly r2 | BothNowhere
+
+class Router r
+instance Router (Nowhere)
+instance Router (Done a)
+instance (Router r) => Router (First r)
+instance (Router r) => Router (Second r)
+instance (Router r1, Router r2) => Router (Both r1 r2)
+
+class SubRouter r1 r2 where
+  weakenRoute :: r1 -> r2
+instance SubRouter Nowhere (Done a) where
+  weakenRoute _ = DoneNowhere
+instance SubRouter Nowhere Nowhere where
+  weakenRoute _ = Nowhere
+instance SubRouter Nowhere (First r) where
+  weakenRoute _ = FirstNowhere
+instance SubRouter Nowhere (Second r) where
+  weakenRoute _ = SecondNowhere
+instance SubRouter Nowhere (Both r1 r2) where
+  weakenRoute _ = BothNowhere
+instance SubRouter (Done a) (Done a) where
+  weakenRoute = id
+instance (SubRouter r1 r2) => SubRouter (First r1) (First r2) where
+  weakenRoute (First r) = First (weakenRoute r)
+  weakenRoute FirstNowhere = FirstNowhere
+instance (SubRouter r1 r2) => SubRouter (Second r1) (Second r2) where
+  weakenRoute (Second r) = Second (weakenRoute r)
+  weakenRoute SecondNowhere = SecondNowhere
+instance (SubRouter r1 r3) => SubRouter (Second r1) (Both r2 r3) where
+  weakenRoute (Second r) = SecondOnly (weakenRoute r2)
+  weakenRoute SecondNowhere = BothNowhere
[…]

### 2011-10-20 Code/old/WrappedTypedCombinators.hs

@@ -1,0 +1,200 @@
+{-# LANGUAGE MultiParamTypeClasses, RankNQuantifiers, FlexibleInstances, FunctionalDependencies, GADTs #-}
+module WrappedTypedCombinators where
+
+-- Routing types
+data RouteNowhere    = RouteNowhere
+data RouteDone a     = RouteDone a | 
+                       RouteDoneNowhere
+data RouteFirst r    = RouteFirst r |
+                       RouteFirstNowhere
+data RouteSecond r   = RouteSecond r |
+                       RouteSecondNowhere
+data RouteBoth r1 r2 = RouteBoth r1 r2 |
+                       RouteOnlyFirst r1 |
+                       RouteOnlySecond r2 |
+                       RouteNeither
+                       
+                     
+-- Signal vector types
+data SVEmpty
+data SVEvent a
+data SVSignal a
+data SVAppend v1 v2
+
+-- Type families for routes
+type family CombineRoutes r1 r2 :: *
+type instance CombineRoutes RouteNowhere RouteNowhere = Nowhere
+type instance CombineRoutes (RouteDone a) Nowhere = RouteFirst (RouteDone a)
+type instance CombineRoutes (RouteFirst r) Nowhere = RouteFirst (RouteFirst r)
+type instance CombineRoutes (RouteSecond r) Nowhere = RouteFirst (RouteSecond r)
+type instance CombineRoutes (RouteBoth r1 r2) Nowhere = RouteFirst (RouteBoth r1 r2)
+type instance CombineRoutes Nowhere (RouteDone a) = RouteSecond (RouteDone a)
+type instance CombineRoutes Nowhere (RouteFirst r) = RouteSecond (RouteFirst r)
+type instance CombineRoutes Nowhere (RouteSecond r) = RouteSecond (RouteSecond r)
+type instance CombineRoutes Nowhere (RouteBoth r1 r2) = RouteSecond (RouteBoth r1 r2)
+type instance CombineRoutes (RouteDone a) (RouteDone b) = RouteBoth (RouteDone a) (RouteDone b)
+type instance CombineRoutes (RouteDone a) (RouteFirst r) = RouteBoth (RouteDone a) (RouteFirst r)
[…]

### 2011-10-20 Docs/Proposal/proposal.tex
Proposal initial

\documentclass[11pt]{artikel3}
\usepackage{fullpage, setspace, graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{times}

\title{RIT Department of Computer Science\\MSc Thesis Proposal:\\\emph{Implementation of a Push-Based Signal-Function FRP System and Evaluation Interface}}
\author{Edward Amsden}
\date{\today}

\begin{document}
\maketitle

\section{Background}
Functional Reactive Programming (FRP) is a family of languages for
declaratively specifying reactive systems. Functional reactive
programming enables declarative specification of reactive systems with
the advantages of functional languages, including ease of reasoning,
higher-order constructs, and composability. FRP includes the concepts
of {\em signals} (or {\em behaviors}) which are continuous, time
varying values, and {\em events} which are discrete time-value pairs.

FRP is split into two distinct families. Classic FRP (CFRP) presents
behaviors and events as first class values to be manipulated by
language constructs. While this approach enables simple reasoning
about programs, most implementations admit space and time leaks. In
addition, it is not clear how external inputs can be provided to
CFRP programs unless primitive behaviors or events for these
inputs are supplied by the library author.

In contrast, Signal Function FRP presents {\em signal functions} as
first class values. In Signal Function FRP (SF-FRP), a program is a
signal function with explicit input, which is supplied by an
evaluation function. SF-FRP programs enforce causality by not
permitting a signal function to evaluate another signal function
directly, thus, they eliminate a large class of time and space leaks.

The problem of efficient evaluation of FRP programs has been
extensively studied. {\em Push-based} evaluation has been demonstrated
for CFRP programs, enabling efficient evaluation of events. A system
enabling the type-safe addition of ad-hoc dynamic optimizations for
Yampa, an implementation of SF-FRP, has also been
demonstrated. However, all documented implementations of SF-FRP are
{\em pull-based}, meaning that they are re-evaluated whenever output
is requested and must synchronize inputs and outputs.

\section{Problem Statement}
The SF-FRP model seems inherently pull-based. However, a push-based
FRP model would eliminate sampling latency on events, as well as
wasted computation when non-occurrences of events are sampled. This
would result in much more responsive and less computationally taxing
SF-FRP systems. However, this requires a method of distinguishing
events from signals in the type systems while still permitting them to
influence each other (i.e. permitting signal functions to take
heterogenous collections of signals and events as both input and output).

The evaluation interface for SF-FRP systems is left unspecified in
literature. In practice, it has been implemented as an infinite loop
which runs a provided IO action to acquire input, and then provides
the signal function output to another IO action. This approach does
not lend itself to push-based evaluation, is not composable, is
awkward to supply input to, and permits the programmer no control
over the ``sampling rate'' of the FRP program. The structure of the
evaluation interface is essential both for the ease of efficiently
communicating input and output between the SF-FRP program and the
outside world, and for enabling push-based evaluation of the SF-FRP
program by permitting the supply of input to be decoupled from the
consuming of output.

\section{Proposed Research}

\section{Research Procedure}

\section{Summary}

\bibliography{proposal}{}
\bibliographystyle{acm}

\end{document}

### 2012-05-11 Code/TyConIndexedSignalVectors/SignalFunctions.hs
Initial signal vectors/functions implementation

@@ -1,0 +1,344 @@
+{-# LANGUAGE KindSignatures, EmptyDataDecls, GADTs #-}
+module SignalFunctions (
+  -- * Datatypes
+  SF(SF),
+  NonInitialized,
+  -- * Basic signal functions
+  identity,
+  -- * Lifting
+  pureSignalTransformer,
+  pureEventTransformer,
+  -- * Composition and Routing
+  (>>>),
+  first,
+  second,
+  swap,
+  copy,
+  ignore,
+  cancelLeft,
+  uncancelLeft,
+  cancelRight,
+  uncancelRight,
+  associate,
+  unassociate,
+  -- * Reactivity
+  switch,
+  -- * Feedback
+  loop,
+  -- * Time dependence
+  time,
+  delay,
+  TimeIntegrate,
+  integrate,
+  -- * Joining
+  union,
+  combineSignals
+  
+) where
+
+import SignalVectors
+
+-- | Signal function running or suspended
+data Initialized :: *
+
+-- | Signal function not yet running
+data NonInitialized :: *
+
+-- | Signal Function Type
+data SF :: * -> * -> * -> * where
+  SF     :: (SMemory Id svIn -> (SMemory Id svOut, SF Initialized svIn svOut)) 
+            -> SF NonInitialized svIn svOut
+  SFInit :: (Double -> ([SVIndex Id svOut], SF Initialized svIn svOut)) 
+            -> (SVIndex Id svIn -> ([SVIndex Id svOut], SF Initialized svIn svOut)) 
+            -> SF Initialized svIn svOut
+
+-- | Identity signal function
+identity :: SF NonInitialized sv sv
+identity = SF (\mem -> (mem, identityInit))
+
+identityInit :: SF Initialized sv sv
+identityInit = SFInit (\dt -> ([], identityInit)) (\idx -> ([idx], identityInit))
+
+
+-- | Lift a pure function to operate on signals
+pureSignalTransformer :: (a -> b) -> SF NonInitialized (SVSignal a) (SVSignal b)
+pureSignalTransformer f = SF (\mem -> (case mem of
+                                         (SMSignal (Id x)) -> (SMSignal (Id $ f x))
+                                         SMEmpty -> SMEmpty, pureSignalTransformerInit f))
+
+pureSignalTransformerInit :: (a -> b) -> SF Initialized (SVSignal a) (SVSignal b)
[…]

### 2012-05-11 Docs/Proposal/classic-frp-example-hs.tex
Proposal

\begin{Verbatim}[commandchars=\\\{\}]
\PY{c+c1}{-- Assumed primitives:}
\PY{c+c1}{-- mousePosition, mouseLeftClick, microphoneAudio}

\PY{c+c1}{-- onceE :: Event a -\PYZgt{} Event a, gives only the first occurrence}
\PY{c+c1}{-- restE :: Event a -\PYZgt{} Event a, gives all except the first occurrence}
\PY{c+c1}{-- switcher :: Behavior a -\PYZgt{} Event (Behavior a) -\PYZgt{} Behavior a,}
\PY{c+c1}{--             act like the given behavior, then each successive}
\PY{c+c1}{--             behavior in the event stream}

\PY{c+c1}{-- Assumed pre-existing functions:}
\PY{c+c1}{-- bandpassFilter :: Behavior Double -\PYZgt{} Behavior Double -\PYZgt{} Behavior Double -\PYZgt{} Behavior Double}
\PY{c+c1}{-- (Audio, low frequency, high frequency, audio output)}

\PY{k+kr}{import} \PY{n+nn}{Control.FRP} \PY{c+c1}{-- Our theoretical classic FRP module}
\PY{k+kr}{import} \PY{n+nn}{Control.Applicative} \PY{p}{(}\PY{o}{\PYZlt{}\PYZdl{}\PYZgt{}}\PY{p}{,} \PY{o}{\PYZlt{}*\PYZgt{}}\PY{p}{,} \PY{n+nf}{pure}\PY{p}{)} \PY{c+c1}{-- fmap and applicative functor application}

\PY{n+nf}{mouseX} \PY{o+ow}{::} \PY{k+kt}{Behavior} \PY{k+kt}{Double}
\PY{n+nf}{mouseX} \PY{o+ow}{=} \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{k+kr}{\PYZus{}}\PY{p}{)} \PY{o+ow}{-\PYZgt{}} \PY{n}{fromIntegral} \PY{n}{x} \PY{o}{/} \PY{n}{fromIntegral} \PY{n}{screenWidth}\PY{p}{)} \PY{o}{\PYZlt{}\PYZdl{}\PYZgt{}} \PY{n}{mousePosition}

\PY{n+nf}{mouseY} \PY{o+ow}{::} \PY{k+kt}{Behavior} \PY{k+kt}{Double}
\PY{n+nf}{mouseY} \PY{o+ow}{=} \PY{n}{fmap} \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{p}{(}\PY{k+kr}{\PYZus{}}\PY{p}{,} \PY{n}{y}\PY{p}{)} \PY{o+ow}{-\PYZgt{}} \PY{n}{fromIntegral} \PY{n}{y} \PY{o}{/} \PY{n}{fromIntegral} \PY{n}{screenHeight}\PY{p}{)} \PY{n}{mousePosition}

\PY{n+nf}{highFreq} \PY{o+ow}{::} \PY{k+kt}{Behavior} \PY{k+kt}{Double}
\PY{n+nf}{highFreq} \PY{o+ow}{=} \PY{n}{fmap} \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{n}{x} \PY{o+ow}{-\PYZgt{}} \PY{l+m+mi}{60} \PY{o}{*} \PY{n}{exp} \PY{p}{(}\PY{n}{x} \PY{o}{*} \PY{n}{ln} \PY{p}{(}\PY{l+m+mi}{20000} \PY{o}{/} \PY{l+m+mi}{60}\PY{p}{)}\PY{p}{)} \PY{n}{mouseX}

\PY{n+nf}{lowFreq} \PY{o+ow}{::} \PY{k+kt}{Behavior} \PY{k+kt}{Double}
\PY{n+nf}{lowFreq} \PY{o+ow}{=} \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{n}{y} \PY{n}{highF} \PY{o+ow}{-\PYZgt{}} \PY{l+m+mi}{60} \PY{o}{*} \PY{n}{exp} \PY{p}{(}\PY{n}{y} \PY{o}{*} \PY{n}{ln} \PY{p}{(}\PY{n}{highF} \PY{o}{/} \PY{l+m+mi}{60}\PY{p}{)}\PY{p}{)} \PY{o}{\PYZlt{}\PYZdl{}\PYZgt{}} \PY{n}{mouseY} \PY{o}{\PYZlt{}*\PYZgt{}} \PY{n}{highFreq}

\PY{n+nf}{audioOutput} \PY{o+ow}{::} \PY{k+kt}{Behavior} \PY{k+kt}{Double}
\PY{n+nf}{audioOutput} \PY{o+ow}{=} \PY{k+kr}{let} \PY{n}{filtered} \PY{o+ow}{=} \PY{n}{bandpass} \PY{n}{microphoneInput} \PY{n}{lowFreq} \PY{n}{highFreq}
                  \PY{n}{willStopFiltering} \PY{o+ow}{=} 
                     \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{n}{evt} \PY{o+ow}{-\PYZgt{}} \PY{n}{filtered} \PY{p}{`}\PY{n}{switcher}\PY{p}{`} 
                              \PY{n}{fmap} \PY{p}{(}\PY{n}{const} \PY{p}{(}\PY{n}{willStartFiltering} \PY{o}{\PYZdl{}} \PY{n}{restE} \PY{n}{evt}\PY{p}{)}\PY{p}{)} 
                              \PY{o}{\PYZdl{}} \PY{n}{onceE} \PY{n}{evt}\PY{p}{)}
                  \PY{n}{willStartFiltering} \PY{o+ow}{=} 
                     \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{n}{evt} \PY{o+ow}{-\PYZgt{}} \PY{n}{microphoneInput} \PY{p}{`}\PY{n}{switcher}\PY{p}{`} 
                              \PY{n}{fmap} \PY{p}{(}\PY{n}{const} \PY{p}{(}\PY{n}{willStopFiltering} \PY{o}{\PYZdl{}} \PY{n}{restE} \PY{n}{evt}\PY{p}{)}\PY{p}{)} 
                              \PY{o}{\PYZdl{}} \PY{n}{onceE} \PY{n}{evt}\PY{p}{)}
              \PY{k+kr}{in} \PY{n}{willStopFiltering} \PY{n}{mouseLeftClick}

\PY{n+nf}{main} \PY{o+ow}{::} \PY{k+kt}{IO} \PY{n}{a}
\PY{n+nf}{main} \PY{o+ow}{=} \PY{n}{playBehaviorAsSound} \PY{n}{audioOutput}
\end{Verbatim}

### 2012-05-11 Docs/Proposal/evalmonad-hs.tex
Proposal

\begin{Verbatim}[commandchars=\\\{\}]
\PY{c+c1}{-- SF Evaluation monad transformer}

\PY{k+kr}{data} \PY{k+kt}{SFEvalMonadT} \PY{o+ow}{::} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*}

\PY{k+kr}{instance} \PY{k+kt}{Monad} \PY{p}{(}\PY{k+kt}{SFEvalMonadT} \PY{n}{svIn} \PY{n}{svOut} \PY{n}{m}\PY{p}{)}

\PY{k+kr}{instance} \PY{k+kt}{SFMonadTransformer} \PY{p}{(}\PY{k+kt}{SFEvalMonadT} \PY{n}{svIn} \PY{n}{svOut}\PY{p}{)}

\PY{n+nf}{initSF} \PY{o+ow}{::} \PY{k+kt}{SF} \PY{n}{svIn} \PY{n}{svOut} \PY{o+ow}{-\PYZgt{}} \PY{k+kt}{SFState} \PY{n}{svIn} \PY{n}{svOut}

\PY{n+nf}{runSFEvalMonadT} \PY{o+ow}{::} \PY{p}{(}\PY{k+kt}{EventOnly} \PY{n}{svIn}\PY{p}{,} \PY{k+kt}{EventOnly} \PY{n}{svOut}\PY{p}{,} \PY{k+kt}{Monad} \PY{n}{m}\PY{p}{)} \PY{o+ow}{=\PYZgt{}} 
                   \PY{k+kt}{SFState} \PY{n}{svIn} \PY{n}{svOut} \PY{o+ow}{-\PYZgt{}}
                   \PY{p}{[}\PY{k+kt}{SVIndex} \PY{p}{(}\PY{o+ow}{-\PYZgt{}} \PY{n}{m} \PY{n+nb}{()}\PY{p}{)} \PY{n}{svOut}\PY{p}{]} \PY{o+ow}{-\PYZgt{}}
                   \PY{k+kt}{SFEvalMonadT} \PY{n}{svIn} \PY{n}{svOut} \PY{n}{m} \PY{n}{a} \PY{o+ow}{-\PYZgt{}}
                   \PY{n}{m} \PY{p}{(}\PY{n}{a}\PY{p}{,} \PY{k+kt}{SFState} \PY{n}{svIn} \PY{n}{svOut}\PY{p}{)}

\PY{n+nf}{push} \PY{o+ow}{::} \PY{p}{(}\PY{k+kt}{EventOnly} \PY{n}{svIn}\PY{p}{,} \PY{k+kt}{EventOnly} \PY{n}{svOut}\PY{p}{)} \PY{o+ow}{=\PYZgt{}}
		\PY{k+kt}{SVIndex} \PY{k+kt}{Id} \PY{n}{svIn} \PY{o+ow}{-\PYZgt{}}
		\PY{k+kt}{SFEvalMonadT} \PY{n}{svIn} \PY{n}{svOut} \PY{n}{m} \PY{n+nb}{()}
\end{Verbatim}

### 2012-05-11 Docs/Proposal/proposal.tex
Proposal

@@
-\documentclass[11pt]{artikel3}
-\usepackage{fullpage, setspace, graphicx}
-\usepackage[margin=1in]{geometry}
-\usepackage{times}
+\documentclass[letterpaper]{article}
+\usepackage[margin=1.2in]{geometry}
+\usepackage{ifthen}
+\usepackage{stmaryrd}
+\usepackage{mathpartir}
+\usepackage[usenames]{color}
+\usepackage{hyperref}
+\usepackage{multicol}
+\usepackage{fancyvrb}
 
-\title{RIT Department of Computer Science\\MSc Thesis Proposal:\\\emph{Implementation of a Push-Based Signal-Function FRP System and Evaluation Interface}}
+% For HsColour
+%\newsavebox{\spaceb}
+%\newsavebox{\tabb}
+%\savebox{\spaceb}[1ex]{~}
+%\savebox{\tabb}[4ex]{~}
+%\newcommand{\hsspace}{\usebox{\spaceb}}
+%\newcommand{\hstab}{\usebox{\tabb}}
+%\newcommand{\conceal}[1]{}
+
+\input{pygments.tex}
+\input{amsden-commands.tex}
+
+\title{Push-Based Signal-Function Functional Reactive Programming (Thesis Proposal)}
 \author{Edward Amsden}
@@
 \begin{document}
+
 \maketitle
+\input{abstract.tex}
+\input{introduction.tex}
+\input{background.tex}
+\input{proposed_work.tex}
 
-\section{Background}
-Functional Reactive Programming (FRP) is a family of languages for
-declaratively specifying reactive systems. Functional reactive
-programming enables declarative specification of reactive systems with
-the advantages of functional languages, including ease of reasoning,
-higher-order constructs, and composability. FRP includes the concepts
-of {\em signals} (or {\em behaviors}) which are continuous, time
-varying values, and {\em events} which are discrete time-value pairs.
-
-FRP is split into two distinct families. Classic FRP (CFRP) presents
-behaviors and events as first class values to be manipulated by
-language constructs. While this approach enables simple reasoning
-about programs, most implementations admit space and time leaks. In
-addition, it is not clear how external inputs can be provided to
-CFRP programs unless primitive behaviors or events for these
-inputs are supplied by the library author.
-
-In contrast, Signal Function FRP presents {\em signal functions} as
-first class values. In Signal Function FRP (SF-FRP), a program is a
-signal function with explicit input, which is supplied by an
-evaluation function. SF-FRP programs enforce causality by not
-permitting a signal function to evaluate another signal function
-directly, thus, they eliminate a large class of time and space leaks.
-
-The problem of efficient evaluation of FRP programs has been
-extensively studied. {\em Push-based} evaluation has been demonstrated
-for CFRP programs, enabling efficient evaluation of events. A system
-enabling the type-safe addition of ad-hoc dynamic optimizations for
-Yampa, an implementation of SF-FRP, has also been
-demonstrated. However, all documented implementations of SF-FRP are
-{\em pull-based}, meaning that they are re-evaluated whenever output
-is requested and must synchronize inputs and outputs.
-
-\section{Problem Statement}
-The SF-FRP model seems inherently pull-based. However, a push-based
-FRP model would eliminate sampling latency on events, as well as
-wasted computation when non-occurrences of events are sampled. This
-would result in much more responsive and less computationally taxing
-SF-FRP systems. However, this requires a method of distinguishing
-events from signals in the type systems while still permitting them to
-influence each other (i.e. permitting signal functions to take
-heterogenous collections of signals and events as both input and output).
-
-The evaluation interface for SF-FRP systems is left unspecified in
-literature. In practice, it has been implemented as an infinite loop
-which runs a provided IO action to acquire input, and then provides
-the signal function output to another IO action. This approach does
-not lend itself to push-based evaluation, is not composable, is
-awkward to supply input to, and permits the programmer no control
-over the ``sampling rate'' of the FRP program. The structure of the
-evaluation interface is essential both for the ease of efficiently
-communicating input and output between the SF-FRP program and the
-outside world, and for enabling push-based evaluation of the SF-FRP
-program by permitting the supply of input to be decoupled from the
-consuming of output.
-
-\section{Proposed Research}
-
-\section{Research Procedure}
-
-\section{Summary}
-
-\bibliography{proposal}{}
 \bibliographystyle{acm}
-
+\bibliography{proposal}
 \end{document}

### 2012-05-11 Docs/Proposal/proposed_work.tex
Proposal

\section{Proposed Work}
\label{section:ProposedWork}

Recent work on types for signal-function FRP systems has yielded N-ary FRP~\cite{Sculthorpe2011}, which types signal function inputs and outputs
with {\em signal vectors} rather than types of values. Signal vectors are phantom types: signal or event type constructor taking a value type,
or a product type of signal vectors. This permits individual components of a signal function's input or output to be designated as an event
or as a signal, and provides precisely the flexibility necessary to create a push-based implementation.

When signal functions are typed using value representations, where events are simply option values, the entire input and output must be constructed
for every step of a signal function. In particular, there is no obvious mechanism for producing signal values which contain only the updated components
of the signal, together with actually occurring events.

By making use of phantom types which designate components of a signal vector, together with Generalized Algebraic Datatypes (GADTs) which
permit the specification of the input and output types of a signal function and type-safe indexing of the input and output components,
individual components of the input of a signal function may be processed as they become available, and may produce individual components of
the output.

I will create a library of combinators for creating signal-function FRP systems which may be evaluated when input is provided, and which
is typed with phantom types describing signal vectors~(Figure~\ref{fig:proposed_work_example_signalvectors}). These combinators will produce continuations
which accept values together with signal vector indexes and produce replacement continuations together with collections of signal vector indexed values
~(Figure~\ref{fig:proposed_work_example_signalfunctions}). 
The indexes allow combinators to use only the continuations necessary to process a specific input, as opposed to requiring the entire input
to be constructed for a signal function. In particular, sequential composition will detect output changes or event
occurrences from the left-side signal function and only evaluate the right-side signal function as necessary, parallel composition ({\tt first} and {\tt second})
will only evaluate their left or right side signal function depending on the index, and routing combinators will produce outputs for a given index
in the output signal vector only when there is an input presented on the corresponding index in the input signal vector.

\begin{figure}
\input{signalvectors-hs.tex}
\hrule
\caption{Examples data types for push-based signal functions}
\label{fig:proposed_work_example_signalvectors}
\end{figure}

\begin{figure}
\input{signalfunctions-hs.tex}
\hrule
\caption{Example data type for push-based signal functions}
\label{fig:proposed_work_example_signalfunctions}
\end{figure}

Further, I will create an evaluation interface for signal functions created with this library, which will provide monadic actions to push
events into the signal function, activating predefined callbacks on output events (Figure \ref{fig:proposed_work_example_evalmonad}). Signals
will not be permitted on the inputs or outputs of evaluated signal functions, but only internally to the signal function.

\begin{figure}
\input{evalmonad-hs.tex}
\hrule
\caption{Example functions and monadic combinators for evaluation of push-based signal functions}
\label{fig:proposed_work_example_evalmonad}
\end{figure}

I will provide a formal semantic description for the combinator library and evaluation interface which will permit me to argue that the
new implementation solves a particularly thorny problem with ``merging'' events.

This implementation will be compared to the existing pull-based signal-function implementation, Yampa, by implementing a simple
2-dimensional game with both Yampa and the new implementation, combining both with OpenGL~\cite{Akeley2011} and GLUT~\cite{Kilgard96} for graphics. CPU usage, framerate,
and response latency will be measured for both systems and comparisons given. In  addition, I expect the proposed implementation to be usable with
GLUT and other event-loop based libraries, or even combinations of event-based libraries, simply and without an additional thread for the
FRP system, something not possible with the present evaluation function in Yampa or with current classic FRP systems.

\subsection{Timeline}
\label{subsection:ProposedWork:Timeline}

\begin{tabular}{rp{34em}}
 \withRightVertBar{April 2, 2012}  & Submit draft proposal with proposed work, timeline, and introduction for review by Professor Fluet. \\
                                   & \\
 \withRightVertBar{April 23, 2012} & Complete initial implementation of FRP library, begin debugging and testing. \\
                                   & \\
 \withRightVertBar{April 30, 2012} & Finish proposal including background. Submit final proposal to \newline
                                     Professor Fluet, Professor Nunes-Harwitt and Professor Butler to\newline
                                     convene committee. \\
                                   &\\
 \withRightVertBar{May 31, 2012}   & Submit ongoing work writeup to Haskell Symposium. Include semantics and an interactive demonstration.\\
                                   & \\
 \withRightVertBar{June 11, 2012}  & Complete demonstration application (2D graphical game) in new FRP library and in Yampa.\\
                                   & \\
 \withRightVertBar{June 18, 2012}  & Complete testing and comparisons of demonstration application.\\
                                   & \\
 \withRightVertBar{June 25, 2012}  & Thesis draft submission.\\
                                   & \\
 \withRightVertBar{July 8, 2012}   & Thesis defense.
\end{tabular}

### 2012-05-11 Docs/Proposal/sf-frp-example-hs.tex
Proposal

\begin{Verbatim}[commandchars=\\\{\}]
\PY{c+c1}{-- Assumed primitives:}
\PY{c+c1}{-- arr :: (a -\PYZgt{} b) -\PYZgt{} SF a b}
\PY{c+c1}{-- \PYZgt{}\PYZgt{}\PYZgt{} :: SF a b -\PYZgt{} SF b c -\PYZgt{} SF a c}
\PY{c+c1}{-- first :: SF a b -\PYZgt{} SF (a, c) (b, c)}
\PY{c+c1}{-- second :: SF a b -\PYZgt{} SF (c, a) (c, b)}
\PY{c+c1}{-- switch :: SF a (b, Event (SF a b)) -\PYZgt{} SF a b}
\PY{c+c1}{-- reactimateWithMouseAndSound :: SF (Event (), ((Int, Int), Double)) (Double) -\PYZgt{} IO ()}

\PY{c+c1}{-- Assumed given signal functions:}
\PY{c+c1}{-- bandpass :: SF (Double, (Double, Double)) Double}

\PY{k+kr}{import} \PY{n+nn}{Control.AFRP} \PY{c+c1}{-- Our theoretical AFRP}

\PY{n+nf}{mouseCoordsToDoubles} \PY{o+ow}{::} \PY{k+kt}{SF} \PY{p}{(}\PY{k+kt}{Int}\PY{p}{,} \PY{k+kt}{Int}\PY{p}{)} \PY{p}{(}\PY{k+kt}{Double}\PY{p}{,} \PY{k+kt}{Double}\PY{p}{)}
\PY{n+nf}{mouseCoordsToDoubles} \PY{o+ow}{=} 
  \PY{n}{arr} \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)} \PY{o+ow}{-\PYZgt{}} \PY{p}{(}\PY{n}{fromIntegral} \PY{n}{x} \PY{o}{/} \PY{n}{fromIntegral} \PY{n}{screenWidth}\PY{p}{,} 
                   \PY{n}{fromIntegral} \PY{n}{y} \PY{o}{/} \PY{n}{fromIntegral} \PY{n}{screenHeight}\PY{p}{)}\PY{p}{)}

\PY{n+nf}{highFreq} \PY{o+ow}{::} \PY{k+kt}{SF} \PY{k+kt}{Double} \PY{k+kt}{Double}
\PY{n+nf}{highFreq} \PY{o+ow}{=} \PY{n}{arr} \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{n}{x} \PY{o+ow}{-\PYZgt{}} \PY{l+m+mi}{60} \PY{o}{*} \PY{n}{exp}\PY{p}{(} \PY{n}{x} \PY{o}{*} \PY{n}{ln} \PY{p}{(}\PY{l+m+mi}{20000} \PY{o}{/} \PY{l+m+mi}{60}\PY{p}{)}\PY{p}{)}\PY{p}{)}

\PY{n+nf}{lowFreq} \PY{o+ow}{::} \PY{k+kt}{SF} \PY{p}{(}\PY{k+kt}{Double}\PY{p}{,} \PY{k+kt}{Double}\PY{p}{)} \PY{k+kt}{Double}
\PY{n+nf}{lowFreq} \PY{o+ow}{=} \PY{n}{arr} \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{p}{(}\PY{n}{high}\PY{p}{,} \PY{n}{y}\PY{p}{)} \PY{o+ow}{-\PYZgt{}} \PY{l+m+mi}{60} \PY{o}{*} \PY{n}{exp} \PY{p}{(}\PY{n}{y} \PY{o}{*} \PY{n}{ln} \PY{p}{(}\PY{n}{high} \PY{o}{/} \PY{l+m+mi}{60}\PY{p}{)}\PY{p}{)}\PY{p}{)}

\PY{n+nf}{lowHighFreq} \PY{o+ow}{::} \PY{k+kt}{SF} \PY{p}{(}\PY{k+kt}{Double}\PY{p}{,} \PY{k+kt}{Double}\PY{p}{)} \PY{p}{(}\PY{k+kt}{Double}\PY{p}{,} \PY{k+kt}{Double}\PY{p}{)}
\PY{n+nf}{lowHighFreq} \PY{o+ow}{=} \PY{n}{first} \PY{n}{highFreq} \PY{o}{\PYZgt{}\PYZgt{}\PYZgt{}} 
              \PY{n}{arr} \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{p}{(}\PY{n}{high}\PY{p}{,} \PY{n}{y}\PY{p}{)} \PY{o+ow}{-\PYZgt{}} \PY{p}{(}\PY{n}{high}\PY{p}{,} \PY{p}{(}\PY{n}{high}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{)}\PY{p}{)} \PY{o}{\PYZgt{}\PYZgt{}\PYZgt{}}
              \PY{n}{second} \PY{n}{lowFreq}

\PY{n+nf}{mouseCoordsToLowHighFreqs} \PY{o+ow}{::} \PY{k+kt}{SF} \PY{p}{(}\PY{k+kt}{Int}\PY{p}{,} \PY{k+kt}{Int}\PY{p}{)} \PY{p}{(}\PY{k+kt}{Double}\PY{p}{,} \PY{k+kt}{Double}\PY{p}{)} 
\PY{n+nf}{mouseCoordsToLowHighFreqs} \PY{o+ow}{=} \PY{n}{mouseCoordsToDoubles} \PY{o}{\PYZgt{}\PYZgt{}\PYZgt{}} \PY{n}{lowHighFreq}

\PY{n+nf}{toggleFilter} \PY{o+ow}{=} 
  \PY{k+kr}{let} \PY{n}{filterOn} \PY{o+ow}{=} 
        \PY{n}{switch} \PY{p}{(}\PY{n}{arr} \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{p}{(}\PY{n}{evt}\PY{p}{,} \PY{p}{(}\PY{n}{mouse}\PY{p}{,} \PY{n}{audio}\PY{p}{)}\PY{p}{)} \PY{o+ow}{-\PYZgt{}} 
                        \PY{p}{(}\PY{p}{(}\PY{n}{audio}\PY{p}{,} \PY{n}{mouse}\PY{p}{)}\PY{p}{,} \PY{n}{fmap} \PY{p}{(}\PY{n}{const} \PY{n}{filterOff}\PY{p}{)} \PY{n}{evt}\PY{p}{)}\PY{p}{)} \PY{o}{\PYZgt{}\PYZgt{}\PYZgt{}}
                \PY{n}{first} \PY{p}{(}\PY{n}{second} \PY{n}{mouseCoordsToLowHighFreqs} \PY{o}{\PYZgt{}\PYZgt{}\PYZgt{}} \PY{n}{bandpass}\PY{p}{)}\PY{p}{)}
      \PY{n}{filterOff} \PY{o+ow}{=} \PY{n}{switch} \PY{p}{(}\PY{n}{arr} \PY{p}{(}\PY{n+nf}{\PYZbs{}}\PY{p}{(}\PY{n}{evt}\PY{p}{,} \PY{p}{(}\PY{k+kr}{\PYZus{}}\PY{p}{,} \PY{n}{audio}\PY{p}{)}\PY{p}{)} \PY{o+ow}{-\PYZgt{}}
                                \PY{p}{(}\PY{n}{audio}\PY{p}{,} \PY{n}{fmap} \PY{p}{(}\PY{n}{const} \PY{n}{filterOn}\PY{p}{)} \PY{n}{evt}\PY{p}{)}\PY{p}{)}\PY{p}{)}
  \PY{k+kr}{in} \PY{n}{filterOn}

\PY{n+nf}{main} \PY{o+ow}{::} \PY{k+kt}{IO} \PY{n+nb}{()}
\PY{n+nf}{main} \PY{o+ow}{=} \PY{n}{reactimateWithMouseAndSound} \PY{n}{toggleFilter}
\end{Verbatim}

### 2012-05-11 Docs/Proposal/signalvectors-hs.tex
Proposal

\begin{Verbatim}[commandchars=\\\{\}]
\PY{c+c1}{-- Signal Vectors}
\PY{k+kr}{data} \PY{k+kt}{SVEmpty}
\PY{k+kr}{data} \PY{k+kt}{SVSignal} \PY{o+ow}{::} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*}
\PY{k+kr}{data} \PY{k+kt}{SVEvent}  \PY{o+ow}{::} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*}
\PY{k+kr}{data} \PY{k+kt}{SVAppend} \PY{o+ow}{::} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*}

\PY{c+c1}{-- Signal Vector indexes}
\PY{k+kr}{data} \PY{k+kt}{SVIndex} \PY{o+ow}{::} \PY{p}{(}\PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*}\PY{p}{)} \PY{o+ow}{-\PYZgt{}} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*} \PY{k+kr}{where}
  \PY{k+kt}{SVISignal} \PY{o+ow}{::} \PY{n}{a} \PY{o+ow}{-\PYZgt{}} \PY{k+kt}{SVIndex} \PY{n}{p} \PY{p}{(}\PY{k+kt}{SVSignal} \PY{n}{p} \PY{n}{a}\PY{p}{)}
  \PY{k+kt}{SVIEvent}  \PY{o+ow}{::} \PY{n}{a} \PY{o+ow}{-\PYZgt{}} \PY{k+kt}{SVIndex} \PY{n}{p} \PY{p}{(}\PY{k+kt}{SVEvent} \PY{n}{p} \PY{n}{a}\PY{p}{)}
  \PY{k+kt}{SVILeft}   \PY{o+ow}{::} \PY{k+kt}{SVIndex} \PY{n}{p} \PY{n}{svl} \PY{o+ow}{-\PYZgt{}} \PY{k+kt}{SVIndex} \PY{n}{p} \PY{p}{(}\PY{k+kt}{SVAppend} \PY{n}{svl} \PY{n}{svr}\PY{p}{)}
  \PY{k+kt}{SVIRight}  \PY{o+ow}{::} \PY{k+kt}{SVIndex} \PY{n}{p} \PY{n}{svr} \PY{o+ow}{-\PYZgt{}} \PY{k+kt}{SVIndex} \PY{n}{p} \PY{p}{(}\PY{k+kt}{SVAppend} \PY{n}{svl} \PY{n}{svr}\PY{p}{)}
\end{Verbatim}

### 2012-05-11 Docs/Proposal/abstract.tex

\section*{Abstract}
Functional Reactive Programming (FRP) is a promising class of abstractions for encoding interactive and time-dependent programs in
functional languages, but has proven difficult to implement efficiently. Signal-function FRP is a subclass of FRP which does not provide
direct access to the time-varying  values by the programmer, but rather provides {\em signal functions} as first class objects in the
program. All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation, producing output from
the FRP system whenever the consumer of the output is ready. This greatly simplifies the implementation of signal-function FRP systems, but
leads to inefficient and wasteful evaluation of the FRP system.

In contrast, an input-driven or ``push-based'' system evaluates the network whenever new input is available. This frees the system from
evaluating the network when nothing has changed, and then only components necessary to react to the input are reevaluated.
This form of evaluation has been partially applied to standard FRP systems~\cite{Elliott2009} but not to signal-function FRP systems.

I propose the implementation of a push-based signal-function FRP system. The semantics of the system will be discussed, and its performance
and expressiveness for practical examples of interactive programs will be compared to existing signal-function FRP systems.

### 2012-05-11 Docs/Proposal/background.tex

\section{Background}
\label{section:Background}

\begin{figure}
\input{classic-frp-example-hs.tex}
\hrule

\caption{The example application expressed using theoretical classic FRP primitives.}
\label{figure:classic-frp-example}
\end{figure}

\begin{figure}
\input{sf-frp-example-hs.tex}
\hrule

\caption{The example application expressed using theoretical signal-function FRP primitives.}
\label{figure:sf-frp-example}
\end{figure}

The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produces functions
as output and/or take functions as input. We say that functions are {\em first class values} in a language.

Most reactive programs are written in an imperative style, using low-level and non-composable abstractions including callbacks
or object based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models.

FRP implies that a model should keep the characteristics of functional programming (i.e. that the basic constructs of the language
should remain first-class) while incorporating reactivity into the language model. In particular, functions should be lifted to operate on reactive values,
and functions themselves ought to be reactive.

The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
abstractions of FRP are behaviors or signals, which are time-dependent values defined at every point in continuous time, and events, which are 
values defined at countably many points in time. An FRP system will provide functions to manipulate events and signals and to react
to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in functional reactive
models should of course be efficiently executable. This has proven to be the main challenge in implementing Functional Reactive Programming.

There are two general approaches to FRP are ``classic'' FRP, where behaviors and signals are first-class and reactive objects, and ``signal-function'' FRP,
where transformers of signals and events are first-class and reactive objects.

In order to illustrate the differences, it will be helpful to consider an example. In this case, we will consider
a toy bandpass audio filtering application which sets the low- and high-pass frequencies based on mouse position.

We would like the low-pass frequency to be exponentially related to the y-coordinate of the mouse, and similarly for the
high-pass frequency and x-coordinate. Further, left-clicking the mouse should toggle the effect.

\subsection{Classic FRP}
\label{subsection:classic_frp}

The first FRP system, Fran~\cite{Elliott1997}, was introduced as a system for compositionally describing interactive animations.
Though the term ``Functional Reactive Programming'' was not used in this work, it was the first to introduce the concept of first-class
behaviors and events as an abstract model for reactive programs. The key motivation was to separate the implementation of reactivity
from the modelling of reactivity, so that a programmer could be freed from details such as sampling time-varying values, time-slicing
to simulate updating values in parallel, and sampling of inputs.

The first implementation of FRP outside the Haskell language was Frapp\'{e}, an FRP implementation for the Java Beans framework. Frapp\'{e} built on
the notion of events and {\em bound properties} in the Beans framework, providing abstract interfaces for FRP events and behaviors, and combinators
as concrete classes implementing these interfaces. The evaluation of Frapp\'{e} used Bean components as sources and sinks, and the implementation of
Bean events and bound properties to propagate changes to the network~\cite{Courtney2001-2}.

The FrTime\footnote{FrTime is available in the current version of Dr Racket (5.2.1).}~\cite{Cooper2006} language extends the Scheme evaluator with a mutable
dependency graph, which is constructed by program evaluation. This graph is then updated by signal changes. FrTime does not provide a distinct concept of
events, and chooses branches of the dependency graph by conditional evaluation of signals, rather than by the substitution of signals used by FRP systems.

The Reactive~\cite{Elliott2009} system is a push-based FRP system with first-class behaviors and events. The primary insight of Reactive is
the separation of reactivity (or changes in response to events whose occurrence time could not be know beforehand) from time-dependence. This
gives rise to {\em reactive normal form}, which represents a behavior as a constant or simple time-varying value, with an event stream carrying values
which are also behaviors in reactive normal form. Push-based evaluation is achieved by forking a Haskell thread to evaluate the head behavior,
while waiting on the evaluation of the event stream. Upon an event occurrence, the current behavior thread is killed and a new thread
spawned to evaluate the new behavior. Unfortunately, The implmentation of Reactive uses a tenuous technique which depends on also forking threads to evaluate
two haskell values concurrently in order to implement event merging. This relies on the library author to ensure consistency when this technique is
used, and leads to thread leakage when one of the merged events is itself the merging of other events.

The reactive-banana~\cite{Apfelmus} library is a push-based FRP system designed for use with Haskell GUI frameworks. In particular, it features a monad for
the creation of behaviors and events which may then be composed and evaluated. This monad includes constructs for binding GUI library constructs to primitive events.
It must be ``compiled'' to a Haskell IO action for evaluation to take place. The implementation of reactive-banana is similar to FrTime, using a dependency graph to update the network on event occurences. Reactive-banana eschews generalized switching in favor of branching functions on behavior values, similarly to FrTime, but
maintains the distinction between behaviors and events. Rather than a generalized switching combinator which allows the replacement of arbitrary behaviors,
reactive-banana provides a step combinator which creates a stepwise behavior from the values of an event stream.

A recent thesis~\cite{Czaplicki2012} described Elm, a stand-alone language for reactivity. Elm provides combinators for manipulating discrete events, and
compiles to Javascript, making it useful for client-side web programming. However, Elm does not provide a notion of switching or of continuous time behaviors,
though an approximation is given using discrete-time events which are actuated at repeated intervals specified during the event definition. The thesis asserts
that Arrowized FRP (signal-function FRP, Section~\ref{subsection:signal_function_frp}) can be embedded in Elm, but provides no support for this assertion.

In classic FRP systems, it is common for the Behavior type to form an applicative functor, and for Events to form a functor. To construct our example application,
we lift the exponential function, with a tuning constant, to a behavior, which we then apply to the $x$ and $y$ values of the mouse. Finally, we assume
a function which takes a behavior for audio data, lowpass frequency, and highpass frequency, and produces a behavior for the filtered audio data. This
function would be written in terms of the integration primitive common to most FRP libraries. Figure~\ref{figure:classic-frp-example} provides an example.

\subsection{Signal Function FRP}
\label{subsection:signal_function_frp}

An alternative approach to FRP was first proposed in work on Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs. This library
utilized the abstraction of Arrows~\cite{Hughes2000} to structure {\em signal functions}. Arrows are abstract type constructors with input and output type
parameters, together with sequential ({\tt >>>}) and parallel ({\tt first} and {\tt second}) composition, branching ({\tt ***}), and lifting ({\tt arr}) functions
to the Arrow type. Signal functions are the first-class abstraction in this approach, they represent time-dependent and reactive transformers of events and signals, 
which are themselves not first class. This approach has two motivations: it increases modularity since both the input and output of signal functions may be transformed,
as opposed to signals or events which may only be transformed in their output, and it avoids a large class of time and space leaks which have emerged when
implementing FRP with first-class signals and events.

Similarly to FrTime, the netwire~\cite{Soylemez} library eschews dynamic switching, in this case in favor of {\em signal inhibition}. Netwire is written as an arrow
transformer, which, together with the Kliesli arrow instance for the IO monad\footnote{Any monad forms a Kliesli arrow.}, permits it to lift IO actions as sources and
sinks at arbitrary points in a signal function network. Signal inhibition is accomplished by making the output of signal functions a monoid, and then combining the
outputs of signal functions. An inhibited signal function will produce the monoid's zero as an output. Primitives have defined inhibition behavior, and composed signal
functions inhibit if their outputs combine to the monoid's zero.

Yampa~\cite{Nilsson2005} is an optimization of the Arrowized FRP system first utilized with Fruit (see above). The implementation of Yampa makes use of Generalized Algebraic Datatypes to permit a much larger class of type-safe datatypes for the signal function representation. This representation, together with ``smart''
constructors and combinators, enables the construction of a self-optimizing arrowized FRP system. Unfortunately, the primary inefficiency, that of unnecessary evaluation
steps due to pull-based evaluation, remains. Further, the optimization is ad-hoc and each new optimization requires the addition of new constructors, as well
as the updating of every primitive combinator to handle every combination of constructors. However, Yampa showed noticeable performance gains over previous Arrowized FRP implementations.

A recent PhD thesis~\cite{Sculthorpe2011} introduced N-Ary FRP, a technique for typing Arrowized FRP systems using dependent types. The bulk of the work consisted
in using the dependent type system to prove the correctness of the FRP system discussed. However, a key typing construct called signal vectors was presented,
which permits the distinction of signal and event types at the level of the FRP system, instead of making events merely a special type of signal. This distinction
is important to the approach I am proposing, as will be made clear in Section~\ref{section:ProposedWork}.

In the signal-function FRP example, we create a signal function with inputs for the audio data, mouse coordinates, and mouse clicks. These inputs are then routed
using the composition combinators to the switching combinator (for toggling the effect) or the input of the filter (Figure~\ref{figure:sf-frp-example}).

\subsection{Comparison of Examples}
\label{subsection:comparison_of_examples}

In the example for classic FRP (Figure~\ref{figure:classic-frp-example}), we note pure values which are not referentially transparent,
since they depend on real-world values. While notation in this manner is appealing, the behavior and event streams must be implemented either
by ``compiling'' to a dependency graph with continuations for sampling, as in reactive-banana, or by using backdoors which allow
IO actions to be treated as pure values, as in most other classic FRP systems. Further, note that event streams, in particular, must be 
explicitly split in order to avoid time leaks and violation of causality. Finally, our program is not modular, since it is not parameterized on
its inputs, and for the same reason, it cannot be suspended as a state value without reference to these input sources.

By contrast, the signal-function FRP program has somewhat less appealing notation (an issue addressed for Arrows~\cite{Paterson2001}), 
but requires no explicit mechanism to avoid time leaks. Further, the program is completely modular, each signal function may be further
composed sequentially either on its input, its output, or both; or in parallel. Finally, note that the evaluation function, not the signal
function, depends on the real-world inputs, thus, given a stepping rather than looping evaluation function, it would be entirely possible
to suspend a signal function as a state value.

\subsection{Outstanding Challenges}
\label{subsection:outstanding_challenges}

At present, there are two key issues apparent with FRP. First, while signal-function FRP is inherently safer and more modular than classic FRP, it has yet to be
efficiently implemented. Second, the interface between FRP programs and the many different sources of inputs and sinks for outputs available to a modern application
writer remains ad-hoc and is in most cases limited by the implementation. One key exception to this is the reactive-banana system, which provides a monad for
constructing primitive events and behaviors from which an FRP program may then be constructed. However, this approach is still inflexible as it requires library support
for the system which the FRP program will interact with. Further, classic FRP programs are vulnerable to time leaks and violations of causality due to the ability
to directly manipulate reactive values.

### 2012-05-11 Docs/Proposal/introduction.tex

\section{Introduction}
\label{section:Introduction}

Most of the useful programs which programmers are asked to write must react to inputs which are not available at the time the program starts,
and produce effects at many different times throughout the execution of the program. Examples of these programs include GUI applications,
web browsers, web applications, video games, multimedia authoring and playback tools, operating system shells and kernels, servers,
robotic control programs, and many others. This ability is called reactivity.

Functional reactive programming (FRP) is a promising class of abstractions for writing reactive programs in functional languages. FRP 
abstractions translate desirable properties of the underlying functional languages, such as higher-order functions and referential
transparency, to reactive constructs, generally without modifications to the underlying language. This permits the compositional
construction of reactive programs in purely functional languages.

Functional reactive programming was introduced, though not quite by name, with Fran~\cite{Elliott1997}, a compositional system for writing
reactive animations. From the start, two key challenges emerged: the enforcement of {\em causality}\footnote{Causality is the property 
that a value in an FRP program depends only on present or past values. Obviously, a program which violates this property is impossible to
evaluate, but in some FRP systems such programs are unfortunately quite easy to write down.} in FRP programs, and the efficient
evaluation of FRP programs.

The first challenge was addressed by representing FRP programs, not as compositions of signals and events, but as compositions of
{\em transformers} of behaviors and events called {\em signal functions}. By not permitting direct access to behaviors or events, but
representing them implicitly, signal functions prohibit accessing past or future time values, only values in the closure of the signal function
and current input values. Signal function FRP programs are written by composing signal functions, and are evaluated using a function which provides input to the
signal function and acts on the output values. This function is elided in existing literature on signal functions. 
A further advantage of signal function FRP programs is that they are more compositional, since additional signal functions may be composed
on the input or output side of a signal function, rather than simply the output side in classic (non-signal-function) FRP systems.

The second challenge was addressed for classic FRP by the creation of a hybrid push-pull FRP system~\cite{Elliott2009}. This system relied
on the distinction between reactivity and time dependence to decompose behaviors into phases of constant or simply time-dependent values,
with new phases initiated by event occurrences. However, this implementation did not address concerns of causality or compositionality.
Further, the existence of events as first-class values in classic FRP forced the use of an impure and resource-leaking technique to compare
events to determine which had the first occurrence after the current time. This technique made use of the Haskell {\tt unsafePerformIO} function
to fork two threads to compare the improving values of the next occurrence time for two events being merged. Further, since the classic FRP
interface permits access to only the output of a behavior or event, and is always bound to its implicit inputs, the best the push-based implementation
could do is to block when no computation needs to be performed. The computation cannot be suspended entirely as a value representation.

To address both challenges in a single FRP implementation, it seems desirable to combine the approaches, creating a push-based
signal-function FRP system. The current implementation approach for signal-function FRP is inherently pull-based~\cite{Nilsson2002}, but
recent work on N-ary FRP, a variant of signal-function FRP which enumerates inputs and distinguishes between events and behaviors\footnote{Behaviors are generally
referred to as {\em signals} in signal function literature. This is unfortunate, since a signal function may operate on events, signals, or some combination of the
two.}, provides a representation which suggests a method of push-based evaluation. In previous implementations of FRP, events were simply
option-valued signals. This approach has the advantage of being simple to implement in a pull-based setting, but does not have an obvious
semantics for event merging, and necessitates pull-based evaluation, since there is no way to separate the events which may be pushed from
the truly continuous values which must be repeatedly sampled.

N-ary FRP provides a model of a type system for a push-based signal-function FRP system. However, in order to represent signal functions
as continuations on collections of inputs rather than on single values, we represent the signal vectors as {\em phantom types}~\cite{Cheney03} over which
the signal function type is parameterized for both inputs and outputs. From these phantom types, we can type indices into the signal vector, which
contain the actual input or output values and form elements for a collection of inputs or outputs. Obviously the collection need not have all possible elements,
which forms the basis for evaluating only those parts of the network which depend on the updated values.

### 2012-05-11 Docs/Proposal/pygments.tex

\makeatletter
\def\PY@reset{\let\PY@it=\relax \let\PY@bf=\relax%
    \let\PY@ul=\relax \let\PY@tc=\relax%
    \let\PY@bc=\relax \let\PY@ff=\relax}
\def\PY@tok#1{\csname PY@tok@#1\endcsname}
\def\PY@toks#1+{\ifx\relax#1\empty\else%
    \PY@tok{#1}\expandafter\PY@toks\fi}
\def\PY@do#1{\PY@bc{\PY@tc{\PY@ul{%
    \PY@it{\PY@bf{\PY@ff{#1}}}}}}}
\def\PY#1#2{\PY@reset\PY@toks#1+\relax+\PY@do{#2}}

\expandafter\def\csname PY@tok@gd\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.63,0.00,0.00}{##1}}}
\expandafter\def\csname PY@tok@gu\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.50,0.00,0.50}{##1}}}
\expandafter\def\csname PY@tok@gt\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.00,0.25,0.82}{##1}}}
\expandafter\def\csname PY@tok@gs\endcsname{\let\PY@bf=\textbf}
\expandafter\def\csname PY@tok@gr\endcsname{\def\PY@tc##1{\textcolor[rgb]{1.00,0.00,0.00}{##1}}}
\expandafter\def\csname PY@tok@cm\endcsname{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.25,0.50,0.50}{##1}}}
\expandafter\def\csname PY@tok@vg\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.10,0.09,0.49}{##1}}}
\expandafter\def\csname PY@tok@m\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.40,0.40,0.40}{##1}}}
\expandafter\def\csname PY@tok@mh\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.40,0.40,0.40}{##1}}}
\expandafter\def\csname PY@tok@go\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.50,0.50,0.50}{##1}}}
\expandafter\def\csname PY@tok@ge\endcsname{\let\PY@it=\textit}
\expandafter\def\csname PY@tok@vc\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.10,0.09,0.49}{##1}}}
\expandafter\def\csname PY@tok@il\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.40,0.40,0.40}{##1}}}
\expandafter\def\csname PY@tok@cs\endcsname{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.25,0.50,0.50}{##1}}}
\expandafter\def\csname PY@tok@cp\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.74,0.48,0.00}{##1}}}
\expandafter\def\csname PY@tok@gi\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.00,0.63,0.00}{##1}}}
\expandafter\def\csname PY@tok@gh\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,0.50}{##1}}}
\expandafter\def\csname PY@tok@ni\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.60,0.60,0.60}{##1}}}
\expandafter\def\csname PY@tok@nl\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.63,0.63,0.00}{##1}}}
\expandafter\def\csname PY@tok@nn\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,1.00}{##1}}}
\expandafter\def\csname PY@tok@no\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.53,0.00,0.00}{##1}}}
\expandafter\def\csname PY@tok@na\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.49,0.56,0.16}{##1}}}
\expandafter\def\csname PY@tok@nb\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.00,0.50,0.00}{##1}}}
\expandafter\def\csname PY@tok@nc\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,1.00}{##1}}}
\expandafter\def\csname PY@tok@nd\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.67,0.13,1.00}{##1}}}
\expandafter\def\csname PY@tok@ne\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.82,0.25,0.23}{##1}}}
\expandafter\def\csname PY@tok@nf\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,1.00}{##1}}}
\expandafter\def\csname PY@tok@si\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.73,0.40,0.53}{##1}}}
\expandafter\def\csname PY@tok@s2\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.73,0.13,0.13}{##1}}}
\expandafter\def\csname PY@tok@vi\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.10,0.09,0.49}{##1}}}
\expandafter\def\csname PY@tok@nt\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.50,0.00}{##1}}}
\expandafter\def\csname PY@tok@nv\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.10,0.09,0.49}{##1}}}
\expandafter\def\csname PY@tok@s1\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.73,0.13,0.13}{##1}}}
\expandafter\def\csname PY@tok@sh\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.73,0.13,0.13}{##1}}}
\expandafter\def\csname PY@tok@sc\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.73,0.13,0.13}{##1}}}
\expandafter\def\csname PY@tok@sx\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.00,0.50,0.00}{##1}}}
\expandafter\def\csname PY@tok@bp\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.00,0.50,0.00}{##1}}}
\expandafter\def\csname PY@tok@c1\endcsname{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.25,0.50,0.50}{##1}}}
\expandafter\def\csname PY@tok@kc\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.50,0.00}{##1}}}
\expandafter\def\csname PY@tok@c\endcsname{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.25,0.50,0.50}{##1}}}
\expandafter\def\csname PY@tok@mf\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.40,0.40,0.40}{##1}}}
\expandafter\def\csname PY@tok@err\endcsname{\def\PY@bc##1{\setlength{\fboxsep}{0pt}\fcolorbox[rgb]{1.00,0.00,0.00}{1,1,1}{\strut ##1}}}
\expandafter\def\csname PY@tok@kd\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.50,0.00}{##1}}}
\expandafter\def\csname PY@tok@ss\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.10,0.09,0.49}{##1}}}
\expandafter\def\csname PY@tok@sr\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.73,0.40,0.53}{##1}}}
\expandafter\def\csname PY@tok@mo\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.40,0.40,0.40}{##1}}}
\expandafter\def\csname PY@tok@kn\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.50,0.00}{##1}}}
\expandafter\def\csname PY@tok@mi\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.40,0.40,0.40}{##1}}}
\expandafter\def\csname PY@tok@gp\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,0.50}{##1}}}
\expandafter\def\csname PY@tok@o\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.40,0.40,0.40}{##1}}}
\expandafter\def\csname PY@tok@kr\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.50,0.00}{##1}}}
\expandafter\def\csname PY@tok@s\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.73,0.13,0.13}{##1}}}
\expandafter\def\csname PY@tok@kp\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.00,0.50,0.00}{##1}}}
\expandafter\def\csname PY@tok@w\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.73,0.73,0.73}{##1}}}
\expandafter\def\csname PY@tok@kt\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.69,0.00,0.25}{##1}}}
\expandafter\def\csname PY@tok@ow\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.67,0.13,1.00}{##1}}}
\expandafter\def\csname PY@tok@sb\endcsname{\def\PY@tc##1{\textcolor[rgb]{0.73,0.13,0.13}{##1}}}
\expandafter\def\csname PY@tok@k\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.50,0.00}{##1}}}
\expandafter\def\csname PY@tok@se\endcsname{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.73,0.40,0.13}{##1}}}
\expandafter\def\csname PY@tok@sd\endcsname{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.73,0.13,0.13}{##1}}}

\def\PYZbs{\char`\\}
\def\PYZus{\char`\_}
\def\PYZob{\char`\{}
\def\PYZcb{\char`\}}
\def\PYZca{\char`\^}
\def\PYZam{\char`\&}
\def\PYZlt{\char`\<}
\def\PYZgt{\char`\>}
\def\PYZsh{\char`\#}
\def\PYZpc{\char`\%}
\def\PYZdl{\char`\$}
\def\PYZti{\char`\~}
% for compatibility with earlier versions
\def\PYZat{@}
\def\PYZlb{[}
\def\PYZrb{]}
\makeatother

### 2012-05-11 Docs/Proposal/signalfunctions-hs.tex

\begin{Verbatim}[commandchars=\\\{\}]
\PY{c+c1}{-- "Identity" type constructor}
\PY{k+kr}{newtype} \PY{k+kt}{Id} \PY{n}{a} \PY{o+ow}{=} \PY{k+kt}{Id} \PY{n}{a}

\PY{c+c1}{-- Signal Function (initialized) datatype}
\PY{k+kr}{data} \PY{k+kt}{SFInitialized} \PY{o+ow}{::} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*} \PY{o+ow}{-\PYZgt{}} \PY{o}{*} \PY{k+kr}{where}
  \PY{k+kt}{SFInitialized} \PY{o+ow}{::} \PY{p}{(}\PY{k+kt}{DTime} \PY{o+ow}{-\PYZgt{}} \PY{p}{(}\PY{p}{[}\PY{k+kt}{SVIndex} \PY{k+kt}{Id} \PY{n}{svOut}\PY{p}{]}\PY{p}{,} \PY{k+kt}{SFInitialized} \PY{n}{svIn} \PY{n}{svOut}\PY{p}{)}\PY{p}{)} \PY{o+ow}{-\PYZgt{}} 
                   \PY{p}{(}\PY{k+kt}{SVIndex} \PY{k+kt}{Id} \PY{n}{svIn} \PY{o+ow}{-\PYZgt{}} \PY{p}{(}\PY{p}{[}\PY{k+kt}{SVIndex} \PY{k+kt}{Id} \PY{n}{svOut}\PY{p}{]}\PY{p}{,} \PY{k+kt}{SFInitialized} \PY{n}{svIn} \PY{n}{svOut}\PY{p}{)}\PY{p}{)} \PY{o+ow}{-\PYZgt{}}
                   \PY{k+kt}{SFInitialized} \PY{n}{svIn} \PY{n}{svOut}
\end{Verbatim}

### 2012-05-11 Code/ReactiveVectors/ReactiveFunctions.hs
cleanup

@@ -1,285 +1,0 @@
-{-# LANGUAGE EmptyDataDecls, GADTs #-}
-module ReactiveFunctions (
-  -- * Reactive function type
-  ReactiveFunction, Initialized, NonInitialized
-  -- * Reactive function accessors (for use in eval implementation)
-  occurrenceContinuation, changeContinuation, timeContinuation, initFunction,
-  -- * Reactive function constructors
-  -- ** Routing functions
-  rfIdentity,
-  rfCompose,
-  rfFirst,     rfSecond,
-  rfCancelL,   rfCancelR,
-  rfUncancelL, rfUncancelR,
-  rfAssoc,     rfUnassoc
-  rfCopy,
-  rfDrop,
-  rfSwap,
-  rfLoop,
-  -- ** Signal functions
-  rfSigPure,
-  rfIntegral,
-  rfCombine,
-  rfHoldEvent,
-  -- ** Event functions
-  rfEventPure,
-  rfAfter,
-  rfCapture,
-  rfPredicate,
-  rfChange,
-  rfEventAccum,
-  -- ** Reactivity
-  rfSwitch
-
-) where
-
-import ReactiveVectors
-
-type DTime = Double
-
-data Initialized
-data NonInitialized
-
-data ReactiveFunction * -> * -> * -> * where
-  -- | Combinators and event functions for switches produce
-  --   non-initialized reactive functions. A non-initialized
-  --   reactive function may not depend on a signal being
-  --   remembered for any input in order to function properly.
-  ReactiveFunctionNonInitialized ::   (SignalMemory v1 -> (ReactiveFunction Initialized v1 v2, SignalMemory v2)) 
-                                      -> ReactiveFunction NonInitialized v1 v2
-  -- | Time dependent functions may produce new events or signals as a result of time advancing
-  --   Internally, signal changes and events should be propagated ahead of time advances.
-  ReactiveFunction  ::   (DTime -> ([Change v2], ReactiveFunction Initialized v2)) 
-                      -> (Change v1 -> ([Change v2], ReactiveFunction Initialized v2))
-                      -> ReactiveFunction Initialized v1 v2
-
-
--- | Selector for change continuations
-changeContinuation ::   ReactiveFunction Initialized v1 v2 
-                     -> Change v1 
-                     -> ([Change v2], ReactiveFunction Initialized v2)
-changeContinuation (ReactiveFunction _ cont) = cont
-
[…]

### 2012-05-11 Code/ReactiveVectors/ReactiveVectors.hs

@@ -1,104 +1,0 @@
-{-# LANGUAGE EmptyDataDecls, GADTs, KindSignatures #-}
-module ReactiveVectors (
-  -- * (Type level) Reactive vectors
-  -- $reactivevectors 
-  RVEmpty,
-  RVSignal,
-  RVEvent,
-  RVAppend
-
-  -- * Representations of reactive values
-  -- ** Signal vector representation
-  SignalMemory(..),
-  signalMemoryLeft,
-  signalMemoryRight,
-  signalMemoryCombine,
-  signalMemoryUpdate,
-
-  -- ** Signal change representation
-  SignalChange(..),
-  splitChanges,
-
-  -- ** Event occurrence representation
-  EventOccurrence(..),
-  splitOccurrences
-) where
-    
-
--- $reactivevectors
--- These are type level vectors of reactive elements, which describe
--- the inputs and outputs of reactive functions 
-
--- | Empty reactive vector
-data RVEmpty
-
--- | Singleton signal reactive vector
-data RVSignal a
-
--- | Singleton event reactive vector
-data RVEvent a
-
--- | Combine two reactive vectors
-data RVAppend v1 v2
-
--- | Signal vectors carry values for all signal elements in a corresponding type-level
--- reactive vector
-data SignalMemory :: * -> * where
-  SignalMemoryUnknown :: SignalMemory v1
-  SignalMemoryValue   :: a -> SignalMemory (RVSignal a)
-  SignalMemoryNode    :: SignalMemory v1 -> SignalMemory v2 -> SignalMemory v1 v2
-
-
-data Change :: * -> * where
-  ChangeSignal :: a -> ChangeValue (RVSignal a)
-  ChangeEvent  :: a -> ChangeEvent (RVEvent a)
-  ChangeLeft   :: Change v1 -> (ChangeLeft (RVAppend v1 v2))
-  ChangeRight  :: Change v2 -> (ChangeRight (RVAppend v1 v2))
-
-data SignalChange :: * -> * where
-  ChangeValue :: a -> ChangeValue (RVSignal a)
-  ChangeLeft  :: SignalChange v1 -> SignalChange (RVAppend v1 v2)
-  ChangeRight :: SignalChange v2 -> SignalChange (RVAppend v1 v2)
-
--- | Event occurrences carry typechecked information about which event
---  element in a type-level reactive vector the occurence corresponds to
-data EventOccurrence :: * -> * where
-  -- | An event occurrence
[…]

### 2012-05-11 Docs/Proposal/classic-frp-example.hs
Proposal: add source files for examples, remove source formatting test

@@ -1,0 +1,42 @@
+-- Assumed primitives:
+-- mousePosition, mouseLeftClick, microphoneAudio
+
+-- onceE :: Event a -> Event a, gives only the first occurrence
+-- restE :: Event a -> Event a, gives all except the first occurrence
+-- switcher :: Behavior a -> Event (Behavior a) -> Behavior a,
+--             act like the given behavior, then each successive
+--             behavior in the event stream
+
+-- Assumed pre-existing functions:
+-- bandpassFilter :: Behavior Double -> Behavior Double -> Behavior Double -> Behavior Double
+-- (Audio, low frequency, high frequency, audio output)
+
+import Control.FRP -- Our theoretical classic FRP module
+import Control.Applicative (<$>, <*>, pure) -- fmap and applicative functor application
+
+mouseX :: Behavior Double
+mouseX = (\(x, _) -> fromIntegral x / fromIntegral screenWidth) <$> mousePosition
+
+mouseY :: Behavior Double
+mouseY = fmap (\(_, y) -> fromIntegral y / fromIntegral screenHeight) mousePosition
+
+highFreq :: Behavior Double
+highFreq = fmap (\x -> 60 * exp (x * ln (20000 / 60)) mouseX
+
+lowFreq :: Behavior Double
+lowFreq = (\y highF -> 60 * exp (y * ln (highF / 60)) <$> mouseY <*> highFreq
+
+audioOutput :: Behavior Double
+audioOutput = let filtered = bandpass microphoneInput lowFreq highFreq
+                  willStopFiltering = 
+                     (\evt -> filtered `switcher` 
+                              fmap (const (willStartFiltering $ restE evt)) 
+                              $ onceE evt)
+                  willStartFiltering = 
+                     (\evt -> microphoneInput `switcher` 
+                              fmap (const (willStopFiltering $ restE evt)) 
+                              $ onceE evt)
+              in willStopFiltering mouseLeftClick
+
+main :: IO a
+main = playBehaviorAsSound audioOutput

### 2012-05-11 Docs/Proposal/evalmonad.hs

@@ -1,0 +1,20 @@
+-- SF Evaluation monad transformer
+
+data SFEvalMonadT :: * -> * -> * -> * -> *
+
+instance Monad (SFEvalMonadT svIn svOut m)
+
+instance SFMonadTransformer (SFEvalMonadT svIn svOut)
+
+initSF :: SF svIn svOut -> SFState svIn svOut
+
+runSFEvalMonadT :: (EventOnly svIn, EventOnly svOut, Monad m) => 
+                   SFState svIn svOut ->
+                   [SVIndex (-> m ()) svOut] ->
+                   SFEvalMonadT svIn svOut m a ->
+                   m (a, SFState svIn svOut)
+
+push :: (EventOnly svIn, EventOnly svOut) =>
+		SVIndex Id svIn ->
+		SFEvalMonadT svIn svOut m ()
+

### 2012-05-11 Docs/Proposal/sf-frp-example.hs

@@ -1,0 +1,43 @@
+-- Assumed primitives:
+-- arr :: (a -> b) -> SF a b
+-- >>> :: SF a b -> SF b c -> SF a c
+-- first :: SF a b -> SF (a, c) (b, c)
+-- second :: SF a b -> SF (c, a) (c, b)
+-- switch :: SF a (b, Event (SF a b)) -> SF a b
+-- reactimateWithMouseAndSound :: SF (Event (), ((Int, Int), Double)) (Double) -> IO ()
+
+-- Assumed given signal functions:
+-- bandpass :: SF (Double, (Double, Double)) Double
+
+import Control.AFRP -- Our theoretical AFRP
+
+mouseCoordsToDoubles :: SF (Int, Int) (Double, Double)
+mouseCoordsToDoubles = 
+  arr (\(x, y) -> (fromIntegral x / fromIntegral screenWidth, 
+                   fromIntegral y / fromIntegral screenHeight))
+
+highFreq :: SF Double Double
+highFreq = arr (\x -> 60 * exp( x * ln (20000 / 60)))
+
+lowFreq :: SF (Double, Double) Double
+lowFreq = arr (\(high, y) -> 60 * exp (y * ln (high / 60)))
+
+lowHighFreq :: SF (Double, Double) (Double, Double)
+lowHighFreq = first highFreq >>> 
+              arr (\(high, y) -> (high, (high, y))) >>>
+              second lowFreq
+
+mouseCoordsToLowHighFreqs :: SF (Int, Int) (Double, Double) 
+mouseCoordsToLowHighFreqs = mouseCoordsToDoubles >>> lowHighFreq
+
+toggleFilter = 
+  let filterOn = 
+        switch (arr (\(evt, (mouse, audio)) -> 
+                        ((audio, mouse), fmap (const filterOff) evt)) >>>
+                first (second mouseCoordsToLowHighFreqs >>> bandpass))
+      filterOff = switch (arr (\(evt, (_, audio)) ->
+                                (audio, fmap (const filterOn) evt)))
+  in filterOn
+
+main :: IO ()
+main = reactimateWithMouseAndSound toggleFilter

### 2012-05-11 Docs/Proposal/signalfunctions.hs

@@ -1,0 +1,9 @@
+-- "Identity" type constructor
+newtype Id a = Id a
+
+-- Signal Function (initialized) datatype
+data SFInitialized :: * -> * -> * where
+  SFInitialized :: (DTime -> ([SVIndex Id svOut], SFInitialized svIn svOut)) -> 
+                   (SVIndex Id svIn -> ([SVIndex Id svOut], SFInitialized svIn svOut)) ->
+                   SFInitialized svIn svOut
+

### 2012-05-11 Docs/Proposal/signalvectors.hs

@@ -1,0 +1,13 @@
+-- Signal Vectors
+data SVEmpty
+data SVSignal :: * -> *
+data SVEvent  :: * -> *
+data SVAppend :: * -> * -> *
+
+-- Signal Vector indexes
+data SVIndex :: (* -> *) -> * -> * where
+  SVISignal :: a -> SVIndex p (SVSignal p a)
+  SVIEvent  :: a -> SVIndex p (SVEvent p a)
+  SVILeft   :: SVIndex p svl -> SVIndex p (SVAppend svl svr)
+  SVIRight  :: SVIndex p svr -> SVIndex p (SVAppend svl svr)
+

### 2012-05-12 Docs/Proposal/background.tex
Sentence structure correction for proposal background

@@
 
-There are two general approaches to FRP are ``classic'' FRP, where behaviors and signals are first-class and reactive objects, and ``signal-function'' FRP,
+The two general approaches to FRP are ``classic'' FRP, where behaviors and signals are first-class and reactive objects, and ``signal-function'' FRP,
 where transformers of signals and events are first-class and reactive objects.

### 2012-05-12 Docs/HaskellSymposium2012/abstract.tex
Haskell2012 submission, initial commit

--- 2012-05-11 Docs/Proposal/abstract.tex
@@
-\section*{Abstract}
 Functional Reactive Programming (FRP) is a promising class of abstractions for encoding interactive and time-dependent programs in
 functional languages, but has proven difficult to implement efficiently. Signal-function FRP is a subclass of FRP which does not provide
-direct access to the time-varying  values by the programmer, but rather provides {\em signal functions} as first class objects in the
-program. All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation, producing output from
+direct access to the time-varying  values by the programmer, but rather provides {\em signal functions} as first class objects.
+All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation, producing output from
 the FRP system whenever the consumer of the output is ready. This greatly simplifies the implementation of signal-function FRP systems, but
@@
 
-I propose the implementation of a push-based signal-function FRP system. The semantics of the system will be discussed, and its performance
-and expressiveness for practical examples of interactive programs will be compared to existing signal-function FRP systems.
+I describe ongoing work toward the implementation of a push-based signal-function FRP system. Informal semantics
+and the current method of implementation are described, and challenges and possible solutions for a completed implementation
+are discussed.

### 2012-05-12 Docs/HaskellSymposium2012/frp.tex
Haskell2012 submission, initial commit

--- 2012-05-12 Docs/Proposal/background.tex
@@
-\section{Background}
-\label{section:Background}
-
-\begin{figure}
-\input{classic-frp-example-hs.tex}
-\hrule
-
-\caption{The example application expressed using theoretical classic FRP primitives.}
-\label{figure:classic-frp-example}
-\end{figure}
-
-\begin{figure}
-\input{sf-frp-example-hs.tex}
-\hrule
-
-\caption{The example application expressed using theoretical signal-function FRP primitives.}
-\label{figure:sf-frp-example}
-\end{figure}
-
-The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produces functions
-as output and/or take functions as input. We say that functions are {\em first class values} in a language.
+\section{Signal-Function FRP}
+\label{section:functional_reactive_programming}
 
@@
 
-In order to illustrate the differences, it will be helpful to consider an example. In this case, we will consider
-a toy bandpass audio filtering application which sets the low- and high-pass frequencies based on mouse position.
+In signal-function FRP, programs are written by composing signal functions, which take one or more signals and/or event streams as input
+and produce one or more signals and/or event-streams as output. Ideally, the programmer has only point-wise access to signals (values at the current
+point in time) and conditional access to event streams (access when there is an event occurrence).
 
-We would like the low-pass frequency to be exponentially related to the y-coordinate of the mouse, and similarly for the
-high-pass frequency and x-coordinate. Further, left-clicking the mouse should toggle the effect.
+In current signal-function FRP systems, such as Yampa~\cite{Nilsson2005}, a signal function transforms a single signal,
+and combinators combine signal functions by splitting input and/or combining output values. Event streams
+are represented as option-valued components of the signal. This leads to inefficient implementation, as event streams
+are sampled at every time step. Further, the latency between event occurrent and event response is, on average, one-half
+of the sampling rate~\cite{Elliott2009}. The pull-based evaluation is necessitated by the single signal value, since
+there is, in general, no way to separate components of a signal which have been updated or do carry event occurrences
+from those which have not.
 
-\subsection{Classic FRP}
-\label{subsection:classic_frp}
-
-The first FRP system, Fran~\cite{Elliott1997}, was introduced as a system for compositionally describing interactive animations.
-Though the term ``Functional Reactive Programming'' was not used in this work, it was the first to introduce the concept of first-class
-behaviors and events as an abstract model for reactive programs. The key motivation was to separate the implementation of reactivity
-from the modelling of reactivity, so that a programmer could be freed from details such as sampling time-varying values, time-slicing
-to simulate updating values in parallel, and sampling of inputs.
-
-The first implementation of FRP outside the Haskell language was Frapp\'{e}, an FRP implementation for the Java Beans framework. Frapp\'{e} built on
-the notion of events and {\em bound properties} in the Beans framework, providing abstract interfaces for FRP events and behaviors, and combinators
-as concrete classes implementing these interfaces. The evaluation of Frapp\'{e} used Bean components as sources and sinks, and the implementation of
-Bean events and bound properties to propagate changes to the network~\cite{Courtney2001-2}.
-
-The FrTime\footnote{FrTime is available in the current version of Dr Racket (5.2.1).}~\cite{Cooper2006} language extends the Scheme evaluator with a mutable
-dependency graph, which is constructed by program evaluation. This graph is then updated by signal changes. FrTime does not provide a distinct concept of
-events, and chooses branches of the dependency graph by conditional evaluation of signals, rather than by the substitution of signals used by FRP systems.
-
-The Reactive~\cite{Elliott2009} system is a push-based FRP system with first-class behaviors and events. The primary insight of Reactive is
-the separation of reactivity (or changes in response to events whose occurrence time could not be know beforehand) from time-dependence. This
-gives rise to {\em reactive normal form}, which represents a behavior as a constant or simple time-varying value, with an event stream carrying values
-which are also behaviors in reactive normal form. Push-based evaluation is achieved by forking a Haskell thread to evaluate the head behavior,
-while waiting on the evaluation of the event stream. Upon an event occurrence, the current behavior thread is killed and a new thread
-spawned to evaluate the new behavior. Unfortunately, The implmentation of Reactive uses a tenuous technique which depends on also forking threads to evaluate
-two haskell values concurrently in order to implement event merging. This relies on the library author to ensure consistency when this technique is
-used, and leads to thread leakage when one of the merged events is itself the merging of other events.
-
-The reactive-banana~\cite{Apfelmus} library is a push-based FRP system designed for use with Haskell GUI frameworks. In particular, it features a monad for
-the creation of behaviors and events which may then be composed and evaluated. This monad includes constructs for binding GUI library constructs to primitive events.
-It must be ``compiled'' to a Haskell IO action for evaluation to take place.
-The implementation of reactive-banana is similar to FrTime, using a dependency graph to update the network on event occurences.
-Reactive-banana eschews generalized switching in favor of branching functions on behavior values, similarly to FrTime, but
-maintains the distinction between behaviors and events. Rather than a generalized switching combinator which allows the replacement of arbitrary behaviors,
-reactive-banana provides a step combinator which creates a stepwise behavior from the values of an event stream.
-
-A recent thesis~\cite{Czaplicki2012} described Elm, a stand-alone language for reactivity. Elm provides combinators for manipulating discrete events, and
-compiles to Javascript, making it useful for client-side web programming. However, Elm does not provide a notion of switching or of continuous time behaviors,
-though an approximation is given using discrete-time events which are actuated at repeated intervals specified during the event definition. The thesis asserts
-that Arrowized FRP (signal-function FRP, Section~\ref{subsection:signal_function_frp}) can be embedded in Elm, but provides no support for this assertion.
-
-In classic FRP systems, it is common for the Behavior type to form an applicative functor, and for Events to form a functor. To construct our example application,
-we lift the exponential function, with a tuning constant, to a behavior, which we then apply to the $x$ and $y$ values of the mouse. Finally, we assume
-a function which takes a behavior for audio data, lowpass frequency, and highpass frequency, and produces a behavior for the filtered audio data. This
-function would be written in terms of the integration primitive common to most FRP libraries. Figure~\ref{figure:classic-frp-example} provides an example.
-
-\subsection{Signal Function FRP}
-\label{subsection:signal_function_frp}
-
-An alternative approach to FRP was first proposed in work on Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs. This library
-utilized the abstraction of Arrows~\cite{Hughes2000} to structure {\em signal functions}. Arrows are abstract type constructors with input and output type
-parameters, together with sequential ({\tt >>>}) and parallel ({\tt first} and {\tt second}) composition, branching ({\tt ***}), and lifting ({\tt arr}) functions
-to the Arrow type. Signal functions are the first-class abstraction in this approach, they represent time-dependent and reactive transformers of events and signals, 
-which are themselves not first class. This approach has two motivations: it increases modularity since both the input and output of signal functions may be transformed,
-as opposed to signals or events which may only be transformed in their output, and it avoids a large class of time and space leaks which have emerged when
-implementing FRP with first-class signals and events.
-
-Similarly to FrTime, the netwire~\cite{Soylemez} library eschews dynamic switching, in this case in favor of {\em signal inhibition}. Netwire is written as an arrow
-transformer, which, together with the Kliesli arrow instance for the IO monad\footnote{Any monad forms a Kliesli arrow.}, permits it to lift IO actions as sources and
-sinks at arbitrary points in a signal function network. Signal inhibition is accomplished by making the output of signal functions a monoid, and then combining the
-outputs of signal functions. An inhibited signal function will produce the monoid's zero as an output. Primitives have defined inhibition behavior, and composed signal
-functions inhibit if their outputs combine to the monoid's zero.
-
-Yampa~\cite{Nilsson2005} is an optimization of the Arrowized FRP system first utilized with Fruit (see above).
-The implementation of Yampa makes use of Generalized Algebraic Datatypes to permit a much larger class of type-safe datatypes for the signal function representation.
-This representation, together with ``smart''
-constructors and combinators, enables the construction of a self-optimizing arrowized FRP system. Unfortunately, the primary inefficiency, that of unnecessary evaluation
-steps due to pull-based evaluation, remains. Further, the optimization is ad-hoc and each new optimization requires the addition of new constructors, as well
-as the updating of every primitive combinator to handle every combination of constructors. However, Yampa showed noticeable performance gains over previous Arrowized FRP implementations.
-
-A recent PhD thesis~\cite{Sculthorpe2011} introduced N-Ary FRP, a technique for typing Arrowized FRP systems using dependent types. The bulk of the work consisted
-in using the dependent type system to prove the correctness of the FRP system discussed. However, a key typing construct called signal vectors was presented,
-which permits the distinction of signal and event types at the level of the FRP system, instead of making events merely a special type of signal. This distinction
-is important to the approach I am proposing, as will be made clear in Section~\ref{section:ProposedWork}.
-
-In the signal-function FRP example, we create a signal function with inputs for the audio data, mouse coordinates, and mouse clicks. These inputs are then routed
-using the composition combinators to the switching combinator (for toggling the effect) or the input of the filter (Figure~\ref{figure:sf-frp-example}).
-
-\subsection{Comparison of Examples}
-\label{subsection:comparison_of_examples}
-
-In the example for classic FRP (Figure~\ref{figure:classic-frp-example}), we note pure values which are not referentially transparent,
-since they depend on real-world values. While notation in this manner is appealing, the behavior and event streams must be implemented either
-by ``compiling'' to a dependency graph with continuations for sampling, as in reactive-banana, or by using backdoors which allow
-IO actions to be treated as pure values, as in most other classic FRP systems. Further, note that event streams, in particular, must be 
-explicitly split in order to avoid time leaks and violation of causality. Finally, our program is not modular, since it is not parameterized on
-its inputs, and for the same reason, it cannot be suspended as a state value without reference to these input sources.
-
-By contrast, the signal-function FRP program has somewhat less appealing notation (an issue addressed for Arrows~\cite{Paterson2001}), 
-but requires no explicit mechanism to avoid time leaks. Further, the program is completely modular, each signal function may be further
-composed sequentially either on its input, its output, or both; or in parallel. Finally, note that the evaluation function, not the signal
-function, depends on the real-world inputs, thus, given a stepping rather than looping evaluation function, it would be entirely possible
-to suspend a signal function as a state value.
-
-\subsection{Outstanding Challenges}
-\label{subsection:outstanding_challenges}
-
-At present, there are two key issues apparent with FRP. First, while signal-function FRP is inherently safer and more modular than classic FRP, it has yet to be
-efficiently implemented. Second, the interface between FRP programs and the many different sources of inputs and sinks for outputs available to a modern application
-writer remains ad-hoc and is in most cases limited by the implementation. One key exception to this is the reactive-banana system, which provides a monad for
-constructing primitive events and behaviors from which an FRP program may then be constructed. However, this approach is still inflexible as it requires library support
-for the system which the FRP program will interact with. Further, classic FRP programs are vulnerable to time leaks and violations of causality due to the ability
-to directly manipulate reactive values.
+In N-Ary FRP~\cite{Sculthorpe2011}, signal functions conceptually operate on {\em signal vectors}, which represent collections
+of signals and event streams at the type level. This opens the possibility of push-based evaluation, since there
+is now a clear distinction between individual signals and event streams. A push-based evaluation of an N-Ary FRP
+system can define reactive operations for signal functions in terms of individual inputs, with combinators
+maintaining the distinctions between elements of the signal vector and skipping evaluation of signal functions for
+which there is no input.

### 2012-05-12 Docs/HaskellSymposium2012/implementation.tex
Haskell2012 submission, initial commit

\section{Implementation}
\label{section:implementation}

Since the signal functions will operate on signal vectors, the first step in an implementation is to represent signal vectors.
We represent signal vectors at the type level as empty types:

\begin{verbatim}
-- An empty signal vector
data SVEmpty
-- A singleton vector of a signal
data SVSignal a
-- A singleton vector of an event stream
data SVEvent a
-- The concatenation of two signal vectors
data SVAppend svl svr 
\end{verbatim}

Once we have this type-level representation of signal vectors, we represent components of the signal vector as a GADT\footnote{Generalized Algebraic Datatype}
with phantom types:

\begin{verbatim}
data SVIndex ::   (* -> *) -> * -> * where
  SVISignal  ::   tc a -> SVIndex tc (SVSignal a)
  SVIEvent   ::   tc a -> SVIndex tc (SVEvent a)
  SVILeft    ::   SVIndex tc svl 
               -> SVIndex tc (SVAppend svl svr)
  SVIRight   ::   SVIndex tc svr 
               -> SVIndex tc (SVAppend svl svr)
\end{verbatim}

Note the addition of the higher-kinded type parameter, which allows us to represent components where the types of the data
in the component are dependent on but not equivalent to the the types of the signals and events in the vector. For instance,
we may instantiate $tc$ as $(\rightarrow IO ())$ to represent a new handler for the output of a signal function.

Another datatype provides a means of accumulating partial information about the state of a signal vector:
\begin{verbatim}
data SVMemory ::   (* -> *) -> * -> *
  SVMSignal   ::   tc a -> SVMemory tc (SVSignal a)
  SVMEvent    ::   tc a -> SVMemory tc (SVEvent a)
  SVMNothing  ::   SVMemory tc sv
  SVMBoth     ::   SVMemory tc svl
                -> SVMemory tc svr 
                -> SVMemory tc (SVAppend svl svr)
\end{verbatim}

Again notice the higher-kinded type parameter, which in this case might be used to produce a record of the current
handlers for the outputs of a signal function.

With these representations in hand we can describe an implementation type for signal functions:
\begin{verbatim}
newtype Id a = Id a

data NonInitialized
data Initialized

data SF  :: * -> * -> * -> * where
  SFInit ::  (SVIndex Id svIn 
              -> ([SVIndex Id svOut],
                  SF Initialized svIn svOut)) 
           -> SF Initialized svIn svOut
  SF     ::  (SVMemory svIn 
              -> (SVMemory Id svOut,
                  SF Initialized svInt svOut)) 
           -> SF NonInitialized svIn svOut
\end{verbatim}

Combinators produce the second constructor, which takes a partial memory of the input state of the signal function.
This enables preservation of the values of signals when a new signal function becomes a part of the network. 
Once a signal function has been supplied with the current state, it exists as a continuation (wrapped by the first constructor)
which takes a change (represented as an event occurrences or a new values for a signal in a signal vector index) to 
a list of output changes and a new signal function.

This implementation strategy is very similar to that used for other signal-function FRP systems, such as Yampa~\cite{Nilsson2005}.
The distinction is that, with the inputs separated, the continuation produced by a combinator may select the appropriate continuation
for a particular input, and components of the input may be responded to as they arrive, rather than all together at each time step.

The implementation strategy can be demonstrated with a few examples of combinators from the library:

The {\tt identity} signal function is perhaps the simplest. It is completely polymorphic in its input, and simply
copies inputs to its output:
\begin{verbatim}
identity :: SF NonInitialized sv sv
identity = SF (\mem -> (mem, identityInit))

identityInit :: SF Initialized sv sv
identityInit = 
  SFInit (\change -> ([change], identityInit))
\end{verbatim}

Note that every signal function must specify its replacement along with its output.
In this case, since identity does not react to but merely passes along inputs, its replacement is itself.

The sequential composition operator ({\tt >>>}) is perhaps the most crucial combinator.

The parallel composition functions {\tt first} and {\tt second} allow signal functions to
operate on one side of a signal vector, with the other side passed through unchanged.
{\tt first} is shown here.

The feedback combinator {\tt loop} allows a signal function to take its own input as output.

The remaining combinators are implemented similarly.

### 2012-05-12 Docs/HaskellSymposium2012/related.tex
Haskell2012 submission, initial commit

--- 2012-05-12 Docs/Proposal/background.tex
@@
-\section{Background}
-\label{section:Background}
-
-\begin{figure}
-\input{classic-frp-example-hs.tex}
-\hrule
-
-\caption{The example application expressed using theoretical classic FRP primitives.}
-\label{figure:classic-frp-example}
-\end{figure}
-
-\begin{figure}
-\input{sf-frp-example-hs.tex}
-\hrule
-
-\caption{The example application expressed using theoretical signal-function FRP primitives.}
-\label{figure:sf-frp-example}
-\end{figure}
-
-The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produces functions
-as output and/or take functions as input. We say that functions are {\em first class values} in a language.
-
-Most reactive programs are written in an imperative style, using low-level and non-composable abstractions including callbacks
-or object based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models.
-
-FRP implies that a model should keep the characteristics of functional programming (i.e. that the basic constructs of the language
-should remain first-class) while incorporating reactivity into the language model. In particular, functions should be lifted to operate on reactive values,
-and functions themselves ought to be reactive.
-
-The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
-abstractions of FRP are behaviors or signals, which are time-dependent values defined at every point in continuous time, and events, which are 
-values defined at countably many points in time. An FRP system will provide functions to manipulate events and signals and to react
-to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
-based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in functional reactive
-models should of course be efficiently executable. This has proven to be the main challenge in implementing Functional Reactive Programming.
-
-The two general approaches to FRP are ``classic'' FRP, where behaviors and signals are first-class and reactive objects, and ``signal-function'' FRP,
-where transformers of signals and events are first-class and reactive objects.
-
-In order to illustrate the differences, it will be helpful to consider an example. In this case, we will consider
-a toy bandpass audio filtering application which sets the low- and high-pass frequencies based on mouse position.
-
-We would like the low-pass frequency to be exponentially related to the y-coordinate of the mouse, and similarly for the
-high-pass frequency and x-coordinate. Further, left-clicking the mouse should toggle the effect.
+\section{Related Work}
+\label{section:related_work}
 
@@
 the creation of behaviors and events which may then be composed and evaluated. This monad includes constructs for binding GUI library constructs to primitive events.
-It must be ``compiled'' to a Haskell IO action for evaluation to take place.
-The implementation of reactive-banana is similar to FrTime, using a dependency graph to update the network on event occurences.
-Reactive-banana eschews generalized switching in favor of branching functions on behavior values, similarly to FrTime, but
+It must be ``compiled'' to a Haskell IO action for evaluation to take place. The implementation of reactive-banana is similar to FrTime, using a dependency graph to 
+update the network on event occurences. Reactive-banana eschews generalized switching in favor of branching functions on behavior values, similarly to FrTime, but
 maintains the distinction between behaviors and events. Rather than a generalized switching combinator which allows the replacement of arbitrary behaviors,
@@
 that Arrowized FRP (signal-function FRP, Section~\ref{subsection:signal_function_frp}) can be embedded in Elm, but provides no support for this assertion.
-
-In classic FRP systems, it is common for the Behavior type to form an applicative functor, and for Events to form a functor. To construct our example application,
-we lift the exponential function, with a tuning constant, to a behavior, which we then apply to the $x$ and $y$ values of the mouse. Finally, we assume
-a function which takes a behavior for audio data, lowpass frequency, and highpass frequency, and produces a behavior for the filtered audio data. This
-function would be written in terms of the integration primitive common to most FRP libraries. Figure~\ref{figure:classic-frp-example} provides an example.
 
@@
 
-Yampa~\cite{Nilsson2005} is an optimization of the Arrowized FRP system first utilized with Fruit (see above).
-The implementation of Yampa makes use of Generalized Algebraic Datatypes to permit a much larger class of type-safe datatypes for the signal function representation.
-This representation, together with ``smart''
+Yampa~\cite{Nilsson2005} is an optimization of the Arrowized FRP system first utilized with Fruit (see above). The implementation of Yampa makes use of Generalized 
+Algebraic Datatypes to permit a much larger class of type-safe datatypes for the signal function representation. This representation, together with ``smart''
 constructors and combinators, enables the construction of a self-optimizing arrowized FRP system. Unfortunately, the primary inefficiency, that of unnecessary evaluation
 steps due to pull-based evaluation, remains. Further, the optimization is ad-hoc and each new optimization requires the addition of new constructors, as well
-as the updating of every primitive combinator to handle every combination of constructors. However, Yampa showed noticeable performance gains over previous Arrowized FRP implementations.
+as the updating of every primitive combinator to handle every combination of constructors. However, Yampa showed noticeable performance gains over previous Arrowized FRP 
+implementations.
 
@@
 in using the dependent type system to prove the correctness of the FRP system discussed. However, a key typing construct called signal vectors was presented,
-which permits the distinction of signal and event types at the level of the FRP system, instead of making events merely a special type of signal. This distinction
-is important to the approach I am proposing, as will be made clear in Section~\ref{section:ProposedWork}.
-
-In the signal-function FRP example, we create a signal function with inputs for the audio data, mouse coordinates, and mouse clicks. These inputs are then routed
-using the composition combinators to the switching combinator (for toggling the effect) or the input of the filter (Figure~\ref{figure:sf-frp-example}).
-
-\subsection{Comparison of Examples}
-\label{subsection:comparison_of_examples}
-
-In the example for classic FRP (Figure~\ref{figure:classic-frp-example}), we note pure values which are not referentially transparent,
-since they depend on real-world values. While notation in this manner is appealing, the behavior and event streams must be implemented either
-by ``compiling'' to a dependency graph with continuations for sampling, as in reactive-banana, or by using backdoors which allow
-IO actions to be treated as pure values, as in most other classic FRP systems. Further, note that event streams, in particular, must be 
-explicitly split in order to avoid time leaks and violation of causality. Finally, our program is not modular, since it is not parameterized on
-its inputs, and for the same reason, it cannot be suspended as a state value without reference to these input sources.
-
-By contrast, the signal-function FRP program has somewhat less appealing notation (an issue addressed for Arrows~\cite{Paterson2001}), 
-but requires no explicit mechanism to avoid time leaks. Further, the program is completely modular, each signal function may be further
-composed sequentially either on its input, its output, or both; or in parallel. Finally, note that the evaluation function, not the signal
-function, depends on the real-world inputs, thus, given a stepping rather than looping evaluation function, it would be entirely possible
-to suspend a signal function as a state value.
-
-\subsection{Outstanding Challenges}
-\label{subsection:outstanding_challenges}
-
-At present, there are two key issues apparent with FRP. First, while signal-function FRP is inherently safer and more modular than classic FRP, it has yet to be
-efficiently implemented. Second, the interface between FRP programs and the many different sources of inputs and sinks for outputs available to a modern application
-writer remains ad-hoc and is in most cases limited by the implementation. One key exception to this is the reactive-banana system, which provides a monad for
-constructing primitive events and behaviors from which an FRP program may then be constructed. However, this approach is still inflexible as it requires library support
-for the system which the FRP program will interact with. Further, classic FRP programs are vulnerable to time leaks and violations of causality due to the ability
-to directly manipulate reactive values.
+which permits the distinction of signal and event types at the level of the FRP system, instead of making events merely a special type of signal. The implementation
+was pull-based in the style of Yampa.

### 2012-05-21 Docs/Proposal/abstract.tex
final proposal version

@@
 functional languages, but has proven difficult to implement efficiently. Signal-function FRP is a subclass of FRP which does not provide
-direct access to the time-varying  values by the programmer, but rather provides {\em signal functions} as first class objects in the
+direct access to the time-varying  values by the programmer, but rather provides {\em signal functions} as first-class objects in the
 program. All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation, producing output from

### 2012-05-21 Docs/Proposal/background.tex
final proposal version

@@
 The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produces functions
-as output and/or take functions as input. We say that functions are {\em first class values} in a language.
+as output and/or take functions as input. We say that functions are {\em first-class values} in a language.
 
 Most reactive programs are written in an imperative style, using low-level and non-composable abstractions including callbacks
-or object based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models.
+or object-based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models.
 
@@
 The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
-abstractions of FRP are behaviors or signals, which are time-dependent values defined at every point in continuous time, and events, which are 
+abstractions of FRP are behaviors or signals\footnote{Behaviors are generally referred to as {\em signals} in signal function literature. This is unfortunate, since a signal
+function may operate on events, signals, or some combination of the two.}, which are time-dependent values defined at every point in continuous time, and events, which are 
 values defined at countably many points in time. An FRP system will provide functions to manipulate events and signals and to react
@@
 while waiting on the evaluation of the event stream. Upon an event occurrence, the current behavior thread is killed and a new thread
-spawned to evaluate the new behavior. Unfortunately, The implmentation of Reactive uses a tenuous technique which depends on also forking threads to evaluate
+spawned to evaluate the new behavior. Unfortunately, the implmentation of Reactive uses a tenuous technique which depends on also forking threads to evaluate
 two haskell values concurrently in order to implement event merging. This relies on the library author to ensure consistency when this technique is
@@
 
-In classic FRP systems, it is common for the Behavior type to form an applicative functor, and for Events to form a functor. To construct our example application,
+In classic FRP systems, it is common for the Behavior type to form an applicative functor, and for the Event type to form a functor. To construct our example application,
 we lift the exponential function, with a tuning constant, to a behavior, which we then apply to the $x$ and $y$ values of the mouse. Finally, we assume
@@
 to the Arrow type. Signal functions are the first-class abstraction in this approach, they represent time-dependent and reactive transformers of events and signals, 
-which are themselves not first class. This approach has two motivations: it increases modularity since both the input and output of signal functions may be transformed,
+which are themselves not first class, since such values cannot be directly manipulated by the programmer.
+This approach has two motivations: it increases modularity since both the input and output of signal functions may be transformed,
 as opposed to signals or events which may only be transformed in their output, and it avoids a large class of time and space leaks which have emerged when
@@
 by ``compiling'' to a dependency graph with continuations for sampling, as in reactive-banana, or by using backdoors which allow
-IO actions to be treated as pure values, as in most other classic FRP systems. Further, note that event streams, in particular, must be 
-explicitly split in order to avoid time leaks and violation of causality. Finally, our program is not modular, since it is not parameterized on
-its inputs, and for the same reason, it cannot be suspended as a state value without reference to these input sources.
+IO actions to be treated as pure values, as in most other classic FRP systems.
+Further, note that event streams, in particular, must be explicitly split in order to avoid time leaks and violation of causality.
+Finally, our program is not modular, since it is not parameterized on its inputs, and for the same reason,it cannot be suspended as a state value without reference
+to these input sources. This requires evaluation to take place in a single event loop.
 

### 2012-05-21 Docs/Proposal/introduction.tex
final proposal version

@@
 {\em transformers} of behaviors and events called {\em signal functions}. By not permitting direct access to behaviors or events, but
-representing them implicitly, signal functions prohibit accessing past or future time values, only values in the closure of the signal function
+representing them implicitly, signal functions prohibit accessing past or future time values, only allowing access to values in the closure of the signal function
 and current input values. Signal function FRP programs are written by composing signal functions, and are evaluated using a function which provides input to the
-signal function and acts on the output values. This function is elided in existing literature on signal functions. 
+signal function and acts on the output values. This evaluation function is elided in existing literature on signal functions. 
 A further advantage of signal function FRP programs is that they are more compositional, since additional signal functions may be composed
@@
 
-To address both challenges in a single FRP implementation, it seems desirable to combine the approaches, creating a push-based
+To address both challenges in a FRP implementation, it seems desirable to combine the approaches, creating a push-based
 signal-function FRP system. The current implementation approach for signal-function FRP is inherently pull-based~\cite{Nilsson2002}, but
-recent work on N-ary FRP, a variant of signal-function FRP which enumerates inputs and distinguishes between events and behaviors\footnote{Behaviors are generally
-referred to as {\em signals} in signal function literature. This is unfortunate, since a signal function may operate on events, signals, or some combination of the
-two.}, provides a representation which suggests a method of push-based evaluation. In previous implementations of FRP, events were simply
+recent work on N-ary FRP, a variant of signal-function FRP which enumerates inputs and distinguishes between events and behaviors, provides a representation which suggests
+a method of push-based evaluation. In previous implementations of FRP, events were simply
 option-valued signals. This approach has the advantage of being simple to implement in a pull-based setting, but does not have an obvious
@@
 
-N-ary FRP provides a model of a type system for a push-based signal-function FRP system. However, in order to represent signal functions
-as continuations on collections of inputs rather than on single values, we represent the signal vectors as {\em phantom types}~\cite{Cheney03} over which
-the signal function type is parameterized for both inputs and outputs. From these phantom types, we can type indices into the signal vector, which
+N-ary FRP provides a model of a type system for a push-based signal-function FRP system. In order to represent signal functions
+as continuations on collections of inputs rather than on single values, we represent the signal vectors as types over which
+the signal function type is parameterized for both inputs and outputs. Indices into the signal vector are typed with the signal vector. These indices
 contain the actual input or output values and form elements for a collection of inputs or outputs. Obviously the collection need not have all possible elements,

### 2012-05-21 Docs/Proposal/proposed_work.tex
final proposal version

@@
 Recent work on types for signal-function FRP systems has yielded N-ary FRP~\cite{Sculthorpe2011}, which types signal function inputs and outputs
-with {\em signal vectors} rather than types of values. Signal vectors are phantom types: signal or event type constructor taking a value type,
-or a product type of signal vectors. This permits individual components of a signal function's input or output to be designated as an event
-or as a signal, and provides precisely the flexibility necessary to create a push-based implementation.
+with {\em signal vectors} rather than types of values. Signal vectors are types: a signal or event type constructor taking a value type,
+or a product type of signal vectors. A signal function then uses signal vectors for its input and output types. Since the actual values which are supplied to the
+representation of a signal function also parameterize over signal vectors, this representation permits us to restrict the type of values which may be supplied to the signal
+function, even though no values take the type of the signal vector as their whole type. This permits individual components of a signal function's input or output to be
+designated as an event or as a signal, and provides precisely the flexibility necessary to create a push-based implementation.
 

### 2012-06-06 Docs/Proposal/proposal.tex

@@
 \bibliographystyle{acm}
-\bibliography{proposal}
+%\bibliography{proposal}
+\begin{thebibliography}{10}
+
+\bibitem{Akeley2011}
+{\sc Akeley, K., and Segal, M.}
+\newblock {\em The OpenGL Graphics System: A Specification}, 2011.
+
+\bibitem{Apfelmus}
+{\sc Apfelmus, H.}
+\newblock reactive-banana library.
+\newblock \mbox{\url{http://hackage.haskell.org/package/reactive-banana}}.
+
+\bibitem{Cooper2006}
+{\sc Cooper, G.~H., and Krishnamurthi, S.}
+\newblock Embedding dynamic dataflow in a call-by-value language.
+\newblock In {\em European Symposium on Programming\/} (2006), pp.~294--308.
+
+\bibitem{Courtney2001-2}
+{\sc Courtney, A.}
+\newblock Frapp\'{e}: Functional reactive programming in {J}ava.
+\newblock In {\em Proceedings of the Third International Symposium on Practical
+  Aspects of Declarative Languages\/} (London, UK, UK, 2001), PADL '01,
+  Springer-Verlag, pp.~29--44.
+
+\bibitem{Courtney2001-1}
+{\sc Courtney, A., and Elliott, C.}
+\newblock Genuinely functional user interfaces.
+\newblock In {\em In Proceedings of the 2001 Haskell Workshop\/} (2001),
+  pp.~41--69.
+
+\bibitem{Czaplicki2012}
+{\sc Czaplicki, E.}
+\newblock Elm: Concurrent {FRP} for functional {GUI}s.
+\newblock
+  \url{http://www.testblogpleaseignore.com/wp-content/uploads/2012/03/thesis.pdf},
+  2012.
+
+\bibitem{Elliott2009}
+{\sc Elliott, C.}
+\newblock Push-pull functional reactive programming.
+\newblock In {\em Haskell Symposium\/} (2009).
+
+\bibitem{Elliott1997}
+{\sc Elliott, C., and Hudak, P.}
+\newblock Functional reactive animation.
+\newblock In {\em Proceedings of the second ACM SIGPLAN international
+  conference on Functional programming\/} (New York, NY, USA, 1997), ICFP '97,
+  ACM, pp.~263--273.
+
+\bibitem{Hughes2000}
+{\sc Hughes, J.}
+\newblock Generalising monads to arrows.
+\newblock {\em Science of Computer Programming 37}, 1–3 (2000), 67 -- 111.
+
+\bibitem{Kilgard96}
+{\sc Kilgard, M.~J.}
+\newblock {\em The OpenGL Utility Toolkit (GLUT) Programming Interface API
+  Version 3}, 1996.
+
+\bibitem{Nilsson2005}
+{\sc Nilsson, H.}
+\newblock Dynamic optimization for functional reactive programming using
+  generalized algebraic data types.
+\newblock In {\em Proceedings of the tenth ACM SIGPLAN international conference
+  on Functional programming\/} (New York, NY, USA, 2005), ICFP '05, ACM,
+  pp.~54--65.
+
+\bibitem{Nilsson2002}
+{\sc Nilsson, H., Courtney, A., and Peterson, J.}
+\newblock Functional reactive programming, continued.
+\newblock In {\em Proceedings of the 2002 ACM SIGPLAN workshop on Haskell\/}
+  (New York, NY, USA, 2002), Haskell '02, ACM, pp.~51--64.
+
+\bibitem{Paterson2001}
+{\sc Paterson, R.}
+\newblock A new notation for arrows.
+\newblock In {\em Proceedings of the sixth ACM SIGPLAN international conference
+  on Functional programming\/} (New York, NY, USA, 2001), ICFP '01, ACM,
+  pp.~229--240.
+
+\bibitem{Sculthorpe2011}
+{\sc Schulthorpe, N.}
+\newblock {\em Towards Safe and Efficient Functional Reactive Programming}.
+\newblock PhD thesis, University of Nottingham, UK, 2011.
+
+\bibitem{Soylemez}
+{\sc S\"{o}ylemez, E.}
+\newblock netwire library.
+\newblock \mbox{\url{http://hackage.haskell.org/package/netwire-3.1.0}}.
+
+\end{thebibliography}
+
 \end{document}

### 2012-06-08 Code/old/SignalFunction.hs
Remove old code

@@ -1,100 +1,0 @@
-{-# LANGUAGE EmptyDataDecls, GADTs, MultiParamTypeClasses, FlexibleInstances, UndecidableInstances, FunctionalDependencies, RankNTypes #-}
-module SignalFunction where
-
--- | An empty type vector
-data EmptyVector
--- | A single signal
-data Signal a
--- | A single event
-data Event a
--- | The concatenation of two vectors
-data AppendVector v1 v2
-
-{-
-instance ReactiveVector EmptyVector
-instance ReactiveVector (Signal a)
-instance ReactiveVector (Event a)
-instance (ReactiveVector v1, ReactiveVector v2) => ReactiveVector (AppendVector v1 v2)
-instance Singleton (Signal a)
-instance Singleton (Event a)
-instance SignalVector EmptyVector
-instance SignalVector (Signal a)
-instance (SignalVector v1, SignalVector v2) => SignalVector (AppendVector v1 v2)
-instance EventVector EmptyVector
-instance EventVector (Event a)
-instance (EventVector v1, EventVector v2) => EventVector (AppendVector v1 v2)
-
--- | The head of a type level reactive vector
-class VectorHead v e | v -> e
-instance VectorHead (Signal a) (Signal a)
-instance VectorHead (Event a) (Event a)
-instance (VectorHead v e) => VectorHead (AppendVector v v') e 
-instance (VectorHead v e) => VectorHead (AppendVector EmptyVector v) e
-
--- | The tail of a type level reactive vector
-class VectorTail v vt | v -> vt 
-instance VectorTail (Signal a) EmptyVector
-instance VectorTail (Event a) EmptyVector
-instance VectorTail (AppendVector (Signal a) v2) v2
-instance VectorTail (AppendVector (Event a) v2 ) v2
-instance (VectorTail v2 vt) => VectorTail (AppendVector EmptyVector v2) vt
-instance (VectorTail v1 vt) => VectorTail (AppendVector v1 v2) (AppendVector vt v2)
-
--- | Equivalence test for two vectors
-class (ReactiveVector v1, ReactiveVector v2) => VectorEquivalent v1 v2
-instance VectorEquivalent (EmptyVector) (EmptyVector)
-instance VectorEquivalent (Signal a) (Signal a)
-instance VectorEquivalent (Event a) (Event a)
[…]

### 2012-06-08 Code/old/SignalVector.hs

@@ -1,61 +1,0 @@
-{-# LANGUAGE EmptyDataDecls, GADTs, MultiParamTypeClasses, RankNTypes #-}
-
-module Marrow where
-
-import Control.Monad
-import Control.Monad.State.Lazy
-import Data.HList
-
-data Event l a
-data Signal l a
-
-data NoLabel
-
-class Reactive a
-instance Reactive Event
-instance Reactive Signal
-
--- We don't care about labels for equivalent signal vectors
-class SignalVectorEquivalent v1 v2
-instance SignalVectorEquivalent HNil HNil
-instance (SignalVectorEquivalent v1 v2) => SignalVectorEquivalent (HCons (Event l1 a) v1) (HCons (Event l2 a) v2)
-
--- Reactive vectors: lists of reactive types
-class (HList a) => ReactiveVector a
-instance ReactiveVector HNil
-instance (Reactive r, ReactiveVector v) = ReactiveVector (HCons (r l a) v)
-
--- Event vectors: lists of event-only types
-class EventVector a
-instance EventVector HNil
-instance (EventVector v) => EventVector (HCons (Event l a) v)
-
--- Signal vectors
-class SignalVector a
-instance SignalVector HNil
-instance (SignalVector v) => SignalVector (HCons (Event l a) v)
-
--- Only the event pieces of an HList
-class (HList l, HList l') => EventsOf l l' | l -> l'
-instance EventsOf HNil HNil
-instance (EventsOf l l') => EventsOf (HCons (Event l a) l) (HCons (Event l a) l')
-instance (EventsOf l l') => EventsOf (HCons (Signal l a) l) l'
-
--- Only the signal pieces of an HList
-class (HList l, HList l') => SignalsOf l l' | l -> l'
-instance SignalsOf HNil HNil
-instance (SignalOf l l') => SignalsOf (HCons (Signal l a) l) (HCons (Signal l a) l')
-instance (SignalOf l l') => SignalsOf (HCons (Event l a) l) l'
-
-
--- Reactive Function type
-data (ReactiveVector v1, ReactiveVector v2) => ReactiveFunction v1 v2 where
-  -- Pure functions
-  PureSignal :: (SignalVector v1, SignalVector v2, TypeVector v1 t1, TypeVector t2 v2) => (t1 -> t2) -> ReactiveFunction v1 v2
-
-  
-type Reactive v1 v2 a = ReactiveT v1 v2 Identity a
-
-type ReactiveIO v1 v2 a = ReactiveT v1 v2 IO a
-
-

### 2012-06-08 Code/old/TypedCombinators.hs

@@ -1,142 +1,0 @@
-{-# LANGUAGE FlexibleInstances, MultiParamTypeClasses, FunctionalDependencies, EmptyDataDecls, GADTs #-}
-module TypedCombinators where
-
-data Signal a
-data Event a
-data Empty
-data AppendV v1 v2
-
--- Reifying dataflow in the type system
-data Nowhere = Nowhere
-data Done a = Done a | DoneNowhere
-data First r = First r | FirstNowhere
-data Second r = Second r | SecondNowhere
-data Both r1 r2 = Both r1 r2 | FirstOnly r1 | SecondOnly r2 | BothNowhere
-
-class Router r
-instance Router (Nowhere)
-instance Router (Done a)
-instance (Router r) => Router (First r)
-instance (Router r) => Router (Second r)
-instance (Router r1, Router r2) => Router (Both r1 r2)
-
-class SubRouter r1 r2 where
-  weakenRoute :: r1 -> r2
-instance SubRouter Nowhere (Done a) where
-  weakenRoute _ = DoneNowhere
-instance SubRouter Nowhere Nowhere where
-  weakenRoute _ = Nowhere
-instance SubRouter Nowhere (First r) where
-  weakenRoute _ = FirstNowhere
-instance SubRouter Nowhere (Second r) where
-  weakenRoute _ = SecondNowhere
-instance SubRouter Nowhere (Both r1 r2) where
-  weakenRoute _ = BothNowhere
-instance SubRouter (Done a) (Done a) where
-  weakenRoute = id
-instance (SubRouter r1 r2) => SubRouter (First r1) (First r2) where
-  weakenRoute (First r) = First (weakenRoute r)
-  weakenRoute FirstNowhere = FirstNowhere
-instance (SubRouter r1 r2) => SubRouter (Second r1) (Second r2) where
-  weakenRoute (Second r) = Second (weakenRoute r)
-  weakenRoute SecondNowhere = SecondNowhere
-instance (SubRouter r1 r3) => SubRouter (Second r1) (Both r2 r3) where
-  weakenRoute (Second r) = SecondOnly (weakenRoute r2)
-  weakenRoute SecondNowhere = BothNowhere
-instance (SubRouter r1 r2) => SubRouter (First r1) (Both r2 r3) where
[…]

### 2012-06-08 Code/old/WrappedTypedCombinators.hs

@@ -1,200 +1,0 @@
-{-# LANGUAGE MultiParamTypeClasses, RankNQuantifiers, FlexibleInstances, FunctionalDependencies, GADTs #-}
-module WrappedTypedCombinators where
-
--- Routing types
-data RouteNowhere    = RouteNowhere
-data RouteDone a     = RouteDone a | 
-                       RouteDoneNowhere
-data RouteFirst r    = RouteFirst r |
-                       RouteFirstNowhere
-data RouteSecond r   = RouteSecond r |
-                       RouteSecondNowhere
-data RouteBoth r1 r2 = RouteBoth r1 r2 |
-                       RouteOnlyFirst r1 |
-                       RouteOnlySecond r2 |
-                       RouteNeither
-                       
-                     
--- Signal vector types
-data SVEmpty
-data SVEvent a
-data SVSignal a
-data SVAppend v1 v2
-
--- Type families for routes
-type family CombineRoutes r1 r2 :: *
-type instance CombineRoutes RouteNowhere RouteNowhere = Nowhere
-type instance CombineRoutes (RouteDone a) Nowhere = RouteFirst (RouteDone a)
-type instance CombineRoutes (RouteFirst r) Nowhere = RouteFirst (RouteFirst r)
-type instance CombineRoutes (RouteSecond r) Nowhere = RouteFirst (RouteSecond r)
-type instance CombineRoutes (RouteBoth r1 r2) Nowhere = RouteFirst (RouteBoth r1 r2)
-type instance CombineRoutes Nowhere (RouteDone a) = RouteSecond (RouteDone a)
-type instance CombineRoutes Nowhere (RouteFirst r) = RouteSecond (RouteFirst r)
-type instance CombineRoutes Nowhere (RouteSecond r) = RouteSecond (RouteSecond r)
-type instance CombineRoutes Nowhere (RouteBoth r1 r2) = RouteSecond (RouteBoth r1 r2)
-type instance CombineRoutes (RouteDone a) (RouteDone b) = RouteBoth (RouteDone a) (RouteDone b)
-type instance CombineRoutes (RouteDone a) (RouteFirst r) = RouteBoth (RouteDone a) (RouteFirst r)
[…]

### 2012-06-18 Code/SignalFunctions.hs
Added evaluation interface

@@ -1,4 +1,4 @@
-{-# LANGUAGE KindSignatures, EmptyDataDecls, GADTs, ScopedTypeVariables, NoMonomorphismRestriction #-}
+{-# LANGUAGE KindSignatures, EmptyDataDecls, GADTs, ScopedTypeVariables, NoMonomorphismRestriction, GeneralizedNewtypeDeriving #-}
 
 -- | Signal functions are stateful, reactive, and time-dependent entities
 -- which respond to inputs by producing outputs. Signal functions are typed
@@ -54,9 +54,37 @@ module SignalFunctions (
   integrate,
   -- * Joining
   union,
-  combineSignals
+  combineSignals,
+  -- * Evaluation
+  SignalDelta,
+  sd,
+  sdLeft,
+  sdRight,
+  sdBoth,
+  EventOccurrence,
+  eo,
+  eoLeft,
+  eoRight,
+  HandlerSet,
+  sdHS,
+  eoHS,
+  hsLeft,
+  hsRight,
+  hsBoth,
+  SFEvalT,
+  SFEvalIO,
+  SFEvalState,
+  runSFEvalT,
+  initSFEval,
+  push,
+  sample
   
 ) where
+
+import Control.Applicative
+import Control.Monad.IO.Class
+import Control.Monad.Trans.Class
+import Control.Monad.Trans.State
 
 import SignalVectors
 
@@ -464,3 +492,103 @@ combineSignalsInit f currentMem =
                   (\_ -> ([], sf))
   in sf
 
+-- | SFEvalState
+data SFEvalState m svIn svOut = SFEvalState { 
+                                esSF :: SF Initialized svIn svOut,
+                                esOutputHandlers :: SMemory (To (m ())) svOut
+                              }
+
+-- | A monad for evaluating signal functions
+newtype SFEvalT svIn svOut m a = SFEvalT (StateT (SFEvalState m svIn svOut) m a) deriving (Functor, Applicative, Monad, MonadIO)
+
+instance MonadTrans (SFEvalT svIn svOut) where
+  lift = SFEvalT . lift
+
+-- | The most likely use case:
+type SFEvalIO svIn svOut a = SFEvalT svIn svOut IO a
+
+-- | Descriptors of signal changes for a time step
+newtype SignalDelta sv = SignalDelta (SMemory Id sv)
+
+-- | Create a signal delta for a singleton input
+sd :: a -> SignalDelta (SVSignal a)
+sd = SignalDelta . SMSignal . Id
+
[…]

### 2012-06-18 Code/SignalVectors.hs

@@ -62,7 +62,9 @@ module SignalVectors
     updateSMemory,
     updateWithSMemory,
     combineSignalMemory,
-    toIndices
+    toIndices,
+    applySMTo,
+    applySMToSM
   )
   where
 
@@ -150,3 +152,20 @@ splitIndices ((SVILeft x):idxs) = let (xs, ys) = splitIndices idxs
                                 in (x:xs, ys)
 splitIndices ((SVIRight y):idxs) = let (xs, ys) = splitIndices idxs
                                  in (xs, y:ys)
+
+-- | Apply a signal memory holding functions to signal indices holding values
+applySMTo :: SMemory (To a) sv -> [SVIndex Id sv] -> [a]
+applySMTo (SMSignal (To f)) idxs = map (\(SVISignal (Id x)) -> f x) idxs
+applySMTo (SMEvent (To f)) idxs = map (\(SVIEvent (Id x)) -> f x) idxs
+applySMTo (SMBoth sml smr) idxs = let (leftidxs, rightidxs) = splitIndices idxs
+                                  in applySMTo sml leftidxs ++ applySMTo smr rightidxs
+applySMTo SMEmpty _ = []
+
+-- | Apply a signal memory holding functions to a signal memory holding values,
+-- producing a list of values.
+applySMToSM :: SMemory (To a) sv -> SMemory Id sv -> [a]
+applySMToSM _ SMEmpty = []
+applySMToSM SMEmpty _ = []
+applySMToSM (SMSignal (To f)) (SMSignal (Id x)) = [f x]
+applySMToSM (SMEvent (To f)) (SMEvent (Id x)) = [f x]
+applySMToSM (SMBoth sml smr) (SMBoth sml' smr') = applySMToSM sml sml' ++ applySMToSM smr smr'

### 2012-06-19 Code/time-flies/src/FRP/TimeFlies/SignalFunctions.hs
Cabalize

@@ -20,7 +20,7 @@
 -- while (in that context) /signal function/ will refer to the signal function
 -- produced by a combinator.
 
-module SignalFunctions (
+module FRP.TimeFlies.SignalFunctions (
   -- * Datatypes
   SF(),
   NonInitialized,
@@ -86,7 +86,7 @@ import Control.Monad.IO.Class
 import Control.Monad.Trans.Class
 import Control.Monad.Trans.State
 
-import SignalVectors
+import FRP.TimeFlies.SignalVectors
 
 -- | Signal function running or suspended
 data Initialized :: *

### 2012-06-19 Code/time-flies/src/FRP/TimeFlies/SignalVectors.hs

@@ -3,7 +3,7 @@
 -- of the input and output structure of signal functions, and value
 -- representations typed with signal vectors for use in implementing
 -- signal functions.
-module SignalVectors
+module FRP.TimeFlies.SignalVectors
   (
     -- * Type Representation of Signal Vectors
     -- | A signal vector is a type which describes

### 2012-06-19 Code/time-flies/src/FRP/TimeFlies/SignalFunctions.hs
Export signal vector types from SignalFunctions module

@@ -22,9 +22,13 @@
 
 module FRP.TimeFlies.SignalFunctions (
   -- * Datatypes
-  SF(),
+  SF,
   NonInitialized,
   Initialized,
+  SVEmpty,
+  SVSignal,
+  SVEvent,
+  SVAppend,
   -- * Basic signal functions
   identity,
   -- * Lifting

### 2012-06-19 Code/time-flies/src/FRP/TimeFlies/SignalVectors.hs
terminology update

@@ -68,13 +68,13 @@ module FRP.TimeFlies.SignalVectors
   )
   where
 
--- | Empty signal vector phantom type
+-- | Empty signal vector type
 data SVEmpty  :: *
--- | Singleton signal signal vector phantom type
+-- | Singleton signal signal vector type
 data SVSignal :: * -> *
--- | Singleton event signal vector phantom type
+-- | Singleton event signal vector type
 data SVEvent  :: * -> *
--- | Combine two signal vector phantom types
+-- | Combine two signal vector types
 data SVAppend :: * -> * -> *
 
 -- | \"Identity\" type constructor, for use as a functor on the element

### 2012-06-19 Code/mouse-color/src/Testing/MouseColor/Main.hs
Testing application skeleton

@@ -1,0 +1,20 @@
+module Main where
+
+import Graphics.Rendering.OpenGL
+import Graphics.UI.GLUT
+
+
+-- Skeleton from http://www.haskell.org/haskellwiki/OpenGLTutorial1
+myPoints :: [(GLfloat, GLfloat, GLfloat)]
+myPoints = map (\k -> (sin (2 * pi * k / 12) ,cos ( 2 * pi * k /12) ,0.0 ) ) [1..12]
+
+main :: IO ()
+main = do 
+  (progname, _) <- getArgsAndInitialize
+  createWindow "Hello World"
+  displayCallback $= display
+  mainLoop
+display = do 
+  clear [ColorBuffer]
+  renderPrimitive Points $ mapM_ (\(x, y, z) -> vertex $ Vertex3 x y z ) myPoints
+  flush

### 2012-06-20 Docs/Thesis/abstract.tex
Initial thesis layout

Functional Reactive Programming (FRP) is a promising family of abstractions for
encoding reactive and time-dependent programs in functional languages.
Signal-function FRP is a subfamily of FRP which provides {\em signal functions}
as first-class abstractions to the programmer, rather than signals and events as
is common in ``classic'' FRP systems.

### 2012-06-20 Docs/Thesis/background.tex
Initial thesis layout

\chapter{Background}
\label{chapter:Background}

The concept of FRP is traced back to the Fran system for declarative 
animation~\cite{Elliott1997}.

### 2012-06-20 Docs/Thesis/comparison.tex
Initial thesis layout

\chapter{Evaluation and Comparisons}
\label{chapter:Evaluation_and_Comparisons}

We have described the implementation of a push-based signal-function FRP system.
A push-based system should theoretically have advantages in event latency and
overall performance over a pull-based system. Further, we would like to compare
the evaluation interface in the current widely-used system with the one proposed
here, again in order to verify that the theoretical benefits do appear in
practice.

### 2012-06-20 Docs/Thesis/conclusions-and-further-work.tex
Initial thesis layout

\chapter{Conclusions and Further Work}
\label{chapter:Conclusions_and_Further_Work}
I have presented TimeFlies, a push-based signal-function FRP system. I have
demonstrated that TimeFlies does realize the theoretical benefits of a
push-based signal-function system.

### 2012-06-20 Docs/Thesis/implementation.tex
Initial thesis layout

\chapter{Implementation}
\label{chapter:Implemenation}

Having established our method of modeling reactivity, we proceed to a purely 
functional implementation of this method. For this implementation, we will make
significant use of certain advanced extensions to the Haskell type system.

### 2012-06-20 Docs/Thesis/introduction.tex
Initial thesis layout

\chapter{Introduction}
\label{chapter:Introduction}

It is common when considering computation to simplify programs to simple
functions from one input to one output. Alas, this simplication generally does
not hold for even the simplest applications of programming. Most applications
demand that a program respond to multiple inputs and produce multiple outputs
scattered through time.

The common model for such programs consists of an event loop which maintains a
queue of input events. These events are then fed to event handlers which perform
various observable actions and mutate the state of the application.

### 2012-06-20 Code/mouse-color/src/Testing/MouseColor/Main.hs
Testing a basic signal function

@@ -1,5 +1,12 @@
 module Main where
 
+import Control.Monad.IO.Class
+
+import Data.IORef
+import Data.Time.Clock
+
+import FRP.TimeFlies.SignalFunctions
+
 import Graphics.Rendering.OpenGL
 import Graphics.UI.GLUT
 
@@ -12,9 +19,50 @@ main :: IO ()
 main = do 
   (progname, _) <- getArgsAndInitialize
   createWindow "Hello World"
-  displayCallback $= display
+  displayRef <- newIORef $ return ()
+
+  time <- getCurrentTime
+  timeRef <- newIORef time
+  
+  let mouseApplicationState = initSFEval (sdHS (\act -> writeIORef displayRef act >> postRedisplay Nothing)) mouseApplicationSF
+  sfRef <- newIORef mouseApplicationState
+  
+  keyboardMouseCallback $= Just (\key keystate _ pos -> do sfSt <- readIORef sfRef
+                                                           ((), sfSt') <- runSFEvalT (do case keystate of
+                                                                                           Down -> push $ eoRight $ eo key
+                                                                                           Up -> return ()
+                                                                                         time <- liftIO getCurrentTime
+                                                                                         oldTime <- liftIO $ readIORef timeRef
+                                                                                         liftIO $ writeIORef timeRef time
+                                                                                         let dt = realToFrac $ time `diffUTCTime` oldTime
+                                                                                         sample dt $ sdLeft $ sd pos)
+                                                                                     sfSt
+                                                           writeIORef sfRef sfSt')
+
+  passiveMotionCallback $= Just (\pos -> do sfSt <- readIORef sfRef
+                                            ((), sfSt') <- runSFEvalT (do time <- liftIO getCurrentTime
+                                                                          oldTime <- liftIO $ readIORef timeRef
+                                                                          liftIO $ writeIORef timeRef time
+                                                                          let dt = realToFrac $ time `diffUTCTime` oldTime
+                                                                          sample dt $ sdLeft $ sd pos)
+                                                                      sfSt
+                                            writeIORef sfRef sfSt')  
+
+  initialDisplayMode $= [DoubleBuffered]
+  displayCallback $= do clear [ColorBuffer]
+                        act <- readIORef displayRef
+                        act
+                        swapBuffers
+                        flush
   mainLoop
[…]

### 2012-06-21 Code/time-flies/src/FRP/TimeFlies/SignalFunctions.hs
Update signal function documentation comments

@@ -237,7 +237,7 @@ copyInit :: SF Initialized sv (SVAppend sv sv)
 copyInit = SFInit (\_ mem -> (combineSignalMemory mem mem, [], copyInit)) 
                   (\change -> ([SVILeft change, SVIRight change], copyInit))
 
--- | Discard the input entirely and produce the left side of the output
+-- | Discard the input entirely and produce an empty output.
 ignore :: SF NonInitialized sv SVEmpty
 ignore = SF (\mem -> (SMEmpty, ignoreInit))
 

### 2012-06-21 Code/mouse-color/src/Testing/MouseColor/Main.hs
Add last sample time and intermediate signal deltas to evaluation interface, update test program

@@ -3,7 +3,7 @@ module Main where
 import Control.Monad.IO.Class
 
 import Data.IORef
-import Data.Time.Clock
+import Data.Time.Clock.POSIX
 
 import FRP.TimeFlies.SignalFunctions
 
@@ -21,32 +21,25 @@ main = do
   createWindow "Hello World"
   displayRef <- newIORef $ return ()
 
-  time <- getCurrentTime
-  timeRef <- newIORef time
+  time <- fmap realToFrac getPOSIXTime
   
-  let mouseApplicationState = initSFEval (sdHS (\act -> writeIORef displayRef act >> postRedisplay Nothing)) mouseApplicationSF
+  let mouseApplicationState = initSFEval (sdHS (\act -> writeIORef displayRef act >> postRedisplay Nothing)) time mouseApplicationSF
   sfRef <- newIORef mouseApplicationState
   
   keyboardMouseCallback $= Just (\key keystate _ pos -> do sfSt <- readIORef sfRef
                                                            ((), sfSt') <- runSFEvalT (do case keystate of
                                                                                            Down -> push $ eoRight $ eo key
                                                                                            Up -> return ()
-                                                                                         time <- liftIO getCurrentTime
-                                                                                         oldTime <- liftIO $ readIORef timeRef
-                                                                                         liftIO $ writeIORef timeRef time
-                                                                                         let dt = realToFrac $ time `diffUTCTime` oldTime
-                                                                                         sample dt $ sdLeft $ sd pos)
+                                                                                         update (sdLeft $ sd pos))
                                                                                      sfSt
                                                            writeIORef sfRef sfSt')
 
   passiveMotionCallback $= Just (\pos -> do sfSt <- readIORef sfRef
-                                            ((), sfSt') <- runSFEvalT (do time <- liftIO getCurrentTime
-                                                                          oldTime <- liftIO $ readIORef timeRef
-                                                                          liftIO $ writeIORef timeRef time
-                                                                          let dt = realToFrac $ time `diffUTCTime` oldTime
-                                                                          sample dt $ sdLeft $ sd pos)
+                                            ((), sfSt') <- runSFEvalT (update (sdLeft $ sd pos))
                                                                       sfSt
                                             writeIORef sfRef sfSt')  
+
+
 
   initialDisplayMode $= [DoubleBuffered]
   displayCallback $= do clear [ColorBuffer]
@@ -54,6 +47,15 @@ main = do
                         act
[…]

### 2012-06-21 Code/time-flies/src/FRP/TimeFlies/SignalFunctions.hs

@@ -81,6 +81,7 @@ module FRP.TimeFlies.SignalFunctions (
   runSFEvalT,
   initSFEval,
   push,
+  update,
   sample
   
 ) where
@@ -498,9 +499,11 @@ combineSignalsInit f currentMem =
 
 -- | SFEvalState
 data SFEvalState m svIn svOut = SFEvalState { 
-                                esSF :: SF Initialized svIn svOut,
-                                esOutputHandlers :: SMemory (To (m ())) svOut
-                              }
+                                  esSF :: SF Initialized svIn svOut,
+                                  esOutputHandlers :: SMemory (To (m ())) svOut,
+                                  esLastTime :: Double,
+                                  esDelta :: SMemory Id svIn
+                                }
 
 -- | A monad for evaluating signal functions
 newtype SFEvalT svIn svOut m a = SFEvalT (StateT (SFEvalState m svIn svOut) m a) deriving (Functor, Applicative, Monad, MonadIO)
@@ -573,8 +576,11 @@ hsBoth :: HandlerSet svl m -> HandlerSet svr m -> HandlerSet (SVAppend svl svr) 
 hsBoth (HandlerSet sml) (HandlerSet smr) = HandlerSet $ combineSignalMemory sml smr
 
 -- | Initialize the evaluation of a signal function, producing an SFEvalState.
-initSFEval :: HandlerSet svOut m -> SF NonInitialized svIn svOut -> SFEvalState m svIn svOut
-initSFEval (HandlerSet handlerMem) (SF memF) = let (_, sf) = memF SMEmpty in SFEvalState sf handlerMem
+initSFEval ::   HandlerSet svOut m -- ^ The set of handlers for the output of the signal function
+             -> Double  -- ^ The initial time of the signal function
+             -> SF NonInitialized svIn svOut -- ^ The signal function to evaluate
+             -> SFEvalState m svIn svOut
+initSFEval (HandlerSet handlerMem) initTime (SF memF) = let (_, sf) = memF SMEmpty in SFEvalState sf handlerMem initTime SMEmpty
 
 -- | Evaluate a signal function, whose current state is stored in an SFEvalState.
 runSFEvalT :: SFEvalT svIn svOut m a -> SFEvalState m svIn svOut -> m (a, SFEvalState m svIn svOut)
[…]

### 2012-06-23 Docs/Thesis/appendix-types.tex
Add typing terms glossary as appendix to thesis

\chapter{Glossary of Type System Terms}
\label{chapter:Glossary_of_Type_System_Terms}

One of the primary attractions of the Haskell language, and the reason for its
use throughout this work, is its advanced yet practical and usable type system.
This type system enables the use of compositional software design that would
be rendered infeasible without a type system to both inform and verify
composition and implementation. This glossary describes features of the type
system which may not be familiar to a general audience.

\begin{description}
\item[ADT] See {\em Algebraic Datatype}

\item[Algebraic Datatype] An Algebraic Datatype or ADT is a type whose terms
are {\em data constructors}. An ADT is defined by naming a
{\em type constructor} and its parameters (zero or more)
(as {\em type variables}), together with one or more data constructors and the
types of their members. Each data constructor takes a fixed number
(zero or more) data members, whose types are given following the constructor
name. These types are defined in terms of the type variables named as parameters
of the type constructor and any type constructors (including the type
constructor associated with the ADT) in scope in the module.

\item[Data Constructor] A Data Constructor is a component of an
Algebraic~Datatype which, when applied to values of the appropriate type, 
creates a value typed with the ADT. Data constructors are the primary element
which may be pattern matched in languages such as Haskell.

\item[Existential Type] An existential type is a means of ``hiding'' a
{\em type variable} in an {\em Algebraic Datatype} or
{\em Generalized Algebraic Datatype}. The Glasgow Haskell Compiler provides
existential types in two ways. The first is the {\tt Rank2Types} and
{\tt RankNTypes} extensions. With either of these extensions enabled,
type variables may be brought into scope for specific {\em data constructor} in
an Algebraic Datatype using the {\tt forall} keyword. Since this
type variable is not contained in the parameters of the type constructor, it
cannot be constrained by type inference. Thus, no external function can assume
anything about its type, except for equality between occurrences of the
variable, and {\em class constraints} given in the definition of the data
constructor. The other way is through Generalized Algebraic Datatypes. Any
variable occurring in the type of a data member of a data constructor of a
Generalized Algebraic Datatype, but not occurring in the type of that
constructor, is automatically existentially quantified.

\item[Type Constructor] A type constructor is a type level term which, when
applied to the proper number of types, produces a type. Type constructors,
together with {\em type variables}, form the basis of polymorphism in Haskell
and similar languages.

\item[Type Variable] A type variable is a type-level term which may be 
instantiated (by the typechecker via inference, or by the user via annotation)
with any type. Together with {\em type constructors}, type variables form the
basis of polymorphism in Haskell and similar languages.
\end{description}

### 2012-06-23 Docs/Thesis/abstract.tex
Intelligently import abstract, introduction, and background from proposal to thesis

@@
-Functional Reactive Programming (FRP) is a promising family of abstractions for
-encoding reactive and time-dependent programs in functional languages.
-Signal-function FRP is a subfamily of FRP which provides {\em signal functions}
-as first-class abstractions to the programmer, rather than signals and events as
-is common in ``classic'' FRP systems.
+Functional Reactive Programming (FRP) is a promising class of abstractions for encoding interactive and time-dependent programs in
+functional languages, but has proven difficult to implement efficiently. Signal-function FRP is a subclass of FRP which does not provide
+direct access to the time-varying  values by the programmer, but rather provides signal functions as first-class objects in the
+program. All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation, producing output from
+the FRP system whenever the consumer of the output is ready. This greatly simplifies the implementation of signal-function FRP systems, but
+leads to inefficient and wasteful evaluation of the FRP system.
+
+In contrast, an input-driven or ``push-based'' system evaluates the network whenever new input is available. This frees the system from
+evaluating the network when nothing has changed, and then only components necessary to react to the input are reevaluated.
+This form of evaluation has been partially applied to standard FRP systems~\cite{Elliott2009} but not to signal-function FRP systems.
+
+I describe the implementation of a push-based signal-function FRP system. The semantics of the system are discussed, and its performance
+and expressiveness for practical examples of interactive programs are be compared to existing signal-function FRP systems through the
+implementation of an example application.

### 2012-06-23 Docs/Thesis/background.tex
Intelligently import abstract, introduction, and background from proposal to thesis

@@
 
-The concept of FRP is traced back to the Fran system for declarative 
-animation~\cite{Elliott1997}.
+The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produces functions
+as output and/or take functions as input. We say that functions are {\em first-class values} in a language.
+
+Most reactive programs are written in an imperative style, using low-level and non-composable abstractions including callbacks
+or object-based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models.
+
+FRP implies that a model should keep the characteristics of functional programming (i.e. that the basic constructs of the language
+should remain first-class) while incorporating reactivity into the language model. In particular, functions should be lifted to operate on reactive values,
+and functions themselves ought to be reactive.
+
+The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
+abstractions of FRP are behaviors or signals\footnote{Behaviors are generally referred to as {\em signals} in signal function literature. This is unfortunate, since a signal
+function may operate on events, signals, or some combination of the two.}, which are time-dependent values defined at every point in continuous time, and events, which are 
+values defined at countably many points in time. An FRP system will provide functions to manipulate events and signals and to react
+to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
+based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in functional reactive
+models should of course be efficiently executable. This has proven to be the main challenge in implementing Functional Reactive Programming.
+
+The two general approaches to FRP are ``classic'' FRP, where behaviors and signals are first-class and reactive objects, and ``signal-function'' FRP,
+where transformers of signals and events are first-class and reactive objects.
+
+\section{Classic FRP}
+\label{section:Background-classic_frp}
+
+The first FRP system, Fran~\cite{Elliott1997}, was introduced as a system for compositionally describing interactive animations.
+Though the term ``Functional Reactive Programming'' was not used in this work, it was the first to introduce the concept of first-class
+behaviors and events as an abstract model for reactive programs. The key motivation was to separate the implementation of reactivity
+from the modelling of reactivity, so that a programmer could be freed from details such as sampling time-varying values, time-slicing
+to simulate updating values in parallel, and sampling of inputs.
+
+The first implementation of FRP outside the Haskell language was Frapp\'{e}, an FRP implementation for the Java Beans framework. Frapp\'{e} built on
+the notion of events and {\em bound properties} in the Beans framework, providing abstract interfaces for FRP events and behaviors, and combinators
+as concrete classes implementing these interfaces. The evaluation of Frapp\'{e} used Bean components as sources and sinks, and the implementation of
+Bean events and bound properties to propagate changes to the network~\cite{Courtney2001-2}.
+
+The FrTime\footnote{FrTime is available in the current version of Dr Racket (5.2.1).}~\cite{Cooper2006} language extends the Scheme evaluator with a mutable
+dependency graph, which is constructed by program evaluation. This graph is then updated by signal changes. FrTime does not provide a distinct concept of
+events, and chooses branches of the dependency graph by conditional evaluation of signals, rather than by the substitution of signals used by FRP systems.
+
+The Reactive~\cite{Elliott2009} system is a push-based FRP system with first-class behaviors and events. The primary insight of Reactive is
+the separation of reactivity (or changes in response to events whose occurrence time could not be know beforehand) from time-dependence. This
+gives rise to {\em reactive normal form}, which represents a behavior as a constant or simple time-varying value, with an event stream carrying values
+which are also behaviors in reactive normal form. Push-based evaluation is achieved by forking a Haskell thread to evaluate the head behavior,
+while waiting on the evaluation of the event stream. Upon an event occurrence, the current behavior thread is killed and a new thread
+spawned to evaluate the new behavior. Unfortunately, the implmentation of Reactive uses a tenuous technique which depends on also forking threads to evaluate
+two haskell values concurrently in order to implement event merging. This relies on the library author to ensure consistency when this technique is
+used, and leads to thread leakage when one of the merged events is itself the merging of other events.
+
+The reactive-banana~\cite{Apfelmus} library is a push-based FRP system designed for use with Haskell GUI frameworks. In particular, it features a monad for
+the creation of behaviors and events which may then be composed and evaluated. This monad includes constructs for binding GUI library constructs to primitive events.
+It must be ``compiled'' to a Haskell IO action for evaluation to take place.
+The implementation of reactive-banana is similar to FrTime, using a dependency graph to update the network on event occurences.
+Reactive-banana eschews generalized switching in favor of branching functions on behavior values, similarly to FrTime, but
+maintains the distinction between behaviors and events. Rather than a generalized switching combinator which allows the replacement of arbitrary behaviors,
+reactive-banana provides a step combinator which creates a stepwise behavior from the values of an event stream.
+
+A recent thesis~\cite{Czaplicki2012} described Elm, a stand-alone language for reactivity. Elm provides combinators for manipulating discrete events, and
+compiles to Javascript, making it useful for client-side web programming. However, Elm does not provide a notion of switching or of continuous time behaviors,
+though an approximation is given using discrete-time events which are actuated at repeated intervals specified during the event definition. The thesis asserts
+that Arrowized FRP (signal-function FRP, Section~\ref{section:Background-signal_function_frp}) can be embedded in Elm, but provides no support for this assertion.
+
+\section{Signal Function FRP}
+\label{section:Background-signal_function_frp}
+
+An alternative approach to FRP was first proposed in work on Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs. This library
+utilized the abstraction of Arrows~\cite{Hughes2000} to structure {\em signal functions}. Arrows are abstract type constructors with input and output type
+parameters, together with sequential ({\tt >>>}) and parallel ({\tt first} and {\tt second}) composition, branching ({\tt ***}), and lifting ({\tt arr}) functions
+to the Arrow type. Signal functions are the first-class abstraction in this approach, they represent time-dependent and reactive transformers of events and signals, 
+which are themselves not first class, since such values cannot be directly manipulated by the programmer.
+This approach has two motivations: it increases modularity since both the input and output of signal functions may be transformed,
+as opposed to signals or events which may only be transformed in their output, and it avoids a large class of time and space leaks which have emerged when
+implementing FRP with first-class signals and events.
+
+Similarly to FrTime, the netwire~\cite{Soylemez} library eschews dynamic switching, in this case in favor of {\em signal inhibition}. Netwire is written as an arrow
+transformer, which, together with the Kliesli arrow instance for the IO monad\footnote{Any monad forms a Kliesli arrow.}, permits it to lift IO actions as sources and
+sinks at arbitrary points in a signal function network. Signal inhibition is accomplished by making the output of signal functions a monoid, and then combining the
+outputs of signal functions. An inhibited signal function will produce the monoid's zero as an output. Primitives have defined inhibition behavior, and composed signal
+functions inhibit if their outputs combine to the monoid's zero.
+
+Yampa~\cite{Nilsson2005} is an optimization of the Arrowized FRP system first utilized with Fruit (see above).
+The implementation of Yampa makes use of Generalized Algebraic Datatypes to permit a much larger class of type-safe datatypes for the signal function representation.
+This representation, together with ``smart''
+constructors and combinators, enables the construction of a self-optimizing arrowized FRP system. Unfortunately, the primary inefficiency, that of unnecessary evaluation
+steps due to pull-based evaluation, remains. Further, the optimization is ad-hoc and each new optimization requires the addition of new constructors, as well
+as the updating of every primitive combinator to handle every combination of constructors. However, Yampa showed noticeable performance gains over previous Arrowized FRP implementations.
+
+A recent PhD thesis~\cite{Sculthorpe2011} introduced N-Ary FRP, a technique for typing Arrowized FRP systems using dependent types. The bulk of the work consisted
+in using the dependent type system to prove the correctness of the FRP system discussed. This work introduced the typing construct of signal vectors,
+which permit the distinction of signal and event types at the level of the FRP system, instead of making events merely a special type of signal.
+
+\section{Outstanding Challenges}
+\label{section:Background-outstanding_challenges}
+
+At present, there are two key issues apparent with FRP. First, while signal-function FRP is inherently safer and more modular than classic FRP, it has yet to be
+efficiently implemented. Second, the interface between FRP programs and the many different sources of inputs and sinks for outputs available to a modern application
+writer remains ad-hoc and is in most cases limited by the implementation. One key exception to this is the reactive-banana system, which provides a monad for
+constructing primitive events and behaviors from which an FRP program may then be constructed. However, this approach is still inflexible as it requires library support
+for the system which the FRP program will interact with. Further, classic FRP programs are vulnerable to time leaks and violations of causality due to the ability
+to directly manipulate reactive values.

### 2012-06-23 Docs/Thesis/introduction.tex
Intelligently import abstract, introduction, and background from proposal to thesis

@@
 
-It is common when considering computation to simplify programs to simple
-functions from one input to one output. Alas, this simplication generally does
-not hold for even the simplest applications of programming. Most applications
-demand that a program respond to multiple inputs and produce multiple outputs
-scattered through time.
+Most of the useful programs which programmers are asked to write must react to inputs which are not available at the time the program starts,
+and produce effects at many different times throughout the execution of the program. Examples of these programs include GUI applications,
+web browsers, web applications, video games, multimedia authoring and playback tools, operating system shells and kernels, servers,
+robotic control programs, and many others. This ability is called reactivity.
 
-The common model for such programs consists of an event loop which maintains a
-queue of input events. These events are then fed to event handlers which perform
-various observable actions and mutate the state of the application.
+Functional reactive programming (FRP) is a promising class of abstractions for writing reactive programs in functional languages. FRP 
+abstractions translate desirable properties of the underlying functional languages, such as higher-order functions and referential
+transparency, to reactive constructs, generally without modifications to the underlying language. This permits the compositional
+construction of reactive programs in purely functional languages.
+
+Functional reactive programming was introduced, though not quite by name, with Fran~\cite{Elliott1997}, a compositional system for writing
+reactive animations. From the start, two key challenges emerged: the enforcement of {\em causality}\footnote{Causality is the property 
+that a value in an FRP program depends only on present or past values. Obviously, a program which violates this property is impossible to
+evaluate, but in some FRP systems such programs are unfortunately quite easy to write down.} in FRP programs, and the efficient
+evaluation of FRP programs.
+
+The first challenge was addressed by representing FRP programs, not as compositions of signals and events, but as compositions of
+{\em transformers} of behaviors and events called {\em signal functions}. By not permitting direct access to behaviors or events, but
+representing them implicitly, signal functions prohibit accessing past or future time values, only allowing access to values in the closure of the signal function
+and current input values. Signal function FRP programs are written by composing signal functions, and are evaluated using a function which provides input to the
+signal function and acts on the output values. This evaluation function is elided in existing literature on signal functions. 
+A further advantage of signal function FRP programs is that they are more compositional, since additional signal functions may be composed
+on the input or output side of a signal function, rather than simply the output side in classic (non-signal-function) FRP systems.
+
+The second challenge was addressed for classic FRP by the creation of a hybrid push-pull FRP system~\cite{Elliott2009}. This system relied
+on the distinction between reactivity and time dependence to decompose behaviors into phases of constant or simply time-dependent values,
+with new phases initiated by event occurrences. However, this implementation did not address concerns of causality or compositionality.
+Further, the existence of events as first-class values in classic FRP forced the use of an impure and resource-leaking technique to compare
+events to determine which had the first occurrence after the current time. This technique made use of the Haskell {\tt unsafePerformIO} function
+to fork two threads to compare the improving values of the next occurrence time for two events being merged. Further, since the classic FRP
+interface permits access to only the output of a behavior or event, and is always bound to its implicit inputs, the best the push-based implementation
+could do is to block when no computation needs to be performed. The computation cannot be suspended entirely as a value representation.
+
+To address both challenges in a FRP implementation, it seems desirable to combine the approaches, creating a push-based
+signal-function FRP system. The current implementation approach for signal-function FRP is inherently pull-based~\cite{Nilsson2002}, but
+recent work on N-ary FRP, a variant of signal-function FRP which enumerates inputs and distinguishes between events and behaviors, provides a representation which suggests
+a method of push-based evaluation. In previous implementations of FRP, events were simply
+option-valued signals. This approach has the advantage of being simple to implement in a pull-based setting, but does not have an obvious
+semantics for event merging, and necessitates pull-based evaluation, since there is no way to separate the events which may be pushed from
+the truly continuous values which must be repeatedly sampled.
+
+N-ary FRP provides a model of a type system for a push-based signal-function FRP system. In order to represent signal functions
+as continuations on collections of inputs rather than on single values, we represent the signal vectors as types over which
+the signal function type is parameterized for both inputs and outputs. Indices into the signal vector are typed with the signal vector. These indices
+contain the actual input or output values and form elements for a collection of inputs or outputs. Obviously the collection need not have all possible elements,
+which forms the basis for evaluating only those parts of the network which depend on the updated values.

### 2012-06-23 Docs/Thesis/appendix-types.tex
Update type system glossary

@@
 
+\item[Generalized Algebraic Datatype] Similar to an {\em Algebraic Datatype},
+but {\em type variables} in the {\em type constructor} declaration serve merely
+to denote the number of type parameters (and thus may be replaced by a
+{\em kind signature}) and types are given for each {\em data constructor}. These
+types must have the type constructor as their top-level term, but may fill in
+the parameters of the type constructor with arbitrary types. Variables which
+appear in the data member types but on in the data constructor type are
+{\em existentially quantified}, and types appearing in the data constructor
+type but not the data member types may be instantiated arbitrarily.
+
+\item[Kind Signature] A means of specifying the number and kind of types which
+may instantiate type variables. Type variables in Haskell are not restricted to
+types, but may be instantiated by type constructors as well. The kind of a
+variable restricts what it may be instantiated with. A kind signature gives
+kinds to a type constructor, and thus to its parameters. Specifying the kind
+of a type constructor perfectly constrains the number of parameters.
+
 \item[Type Constructor] A type constructor is a type level term which, when

### 2012-06-23 Docs/Thesis/introduction.tex
Edits to introduction

@@
 signal-function FRP system. The current implementation approach for signal-function FRP is inherently pull-based~\cite{Nilsson2002}, but
-recent work on N-ary FRP, a variant of signal-function FRP which enumerates inputs and distinguishes between events and behaviors, provides a representation which suggests
+recent work on N-ary FRP~\cite{Sculthorpe2011}, a variant of signal-function FRP which enumerates inputs and distinguishes between events and behaviors, provides a representation which suggests
 a method of push-based evaluation. In previous implementations of FRP, events were simply

### 2012-06-23 Docs/Thesis/implementation-evalinterface.tex
Split implementation tex files

\section{Evaluation Interface}
\label{section:Implementation-Evaluation_Interface}
The evaluation interface provides the means of evaluating a signal function
with inputs and producing effects in response to the signal function's outputs.
We would like to produce a set of constructs that interacts well with Haskell's
system for external IO.

The evaluation interface translates between signal functions and the standard
Haskell construct for sequencing effects and external inputs, namely,
{\em monads}~\cite{PeytonJones1993}. The inspiration for monads is drawn from
the rather esoteric domain of category theory, but the concept as applied to
programming languages is actually rather simple.

In a programming language with type constructors, a monad is simply a 1-arity
type constructor together with two operations. We shall call this type
constructor $m$. The first operation takes a value of any type
(call the type $a$) and produces a value of type $m \: a$ ($m$ applied to $a$).
The second operation takes a value of type $m \: a$ and a function from $a$ to
$m \: b$, and produces a $m \: b$. This means that this second operation,
monadic application, applies the function to the value in the context of the
monad.

For our purposes, there are two interesting properties of monads. One is that
for a specific type constructor, opaque primitives may be defined which
can be used, along with return, to build functions for the second argument
of the monadic application operation. In our case, we would like operations
to push an event to a signal function, to update the signal components of
the signal function, and to increment the time and sample the signal function.

This property is what enables monads to be used for input and output. A Haskell
{\tt main} program is an IO action. An IO action is a value of the monadic type
IO. This action is comprised of a primitive action (either a {\tt return} or a 
true IO primitive), possibly together with a function from the output of that
action to another IO action. Note that values with the IO type may appear
anywhere in a Haskell program, but they are only execute when they are returned
as a part of the {\tt main} IO program.

The Haskell typeclass for monads is defined:

\begin{code}
class Monad m where
  return :: a -> m a
  -- | Infix operator for monadic application
  (>>=)  :: m a -> (a -> m b) -> m b 
\end{code}

The other interesting property of monads is the existence of a class of monads
called monad transformers. These are arity-2 type constructors which take as
their first parameter a monad type constructor. The partial application of
such a type constructor to any monad type constructor then results in a monad.
A construct is then provided to lift values of the underlying monad to values
in the transforming monad, which may then be sequenced as normal. This leads
to the powerful concept of monad transformer stacks, where monads with many
different capabilities are combined without any more plumbing than explicit
{\tt lift} operations.

This property will be useful because we wish our evaluation interface to be
part of a monad stack, the base of which is Haskell's IO monad. By formulating
the evaluation interface as a monad transformer, we need only define the
operations relevant to the evaluation of signal functions, and we can depend
on the constructs of the IO monad to interact with whatever inputs and outputs
are necessary. In some cases, we may not wish to use the IO monad at all (e.g
for testing or simulation). In this case, we can parameterize over another
monad, such as the Identity monad (which has no special operations and whose
context is just the value), or the State monad (which maintains an implicit 
state accessible by {\tt get} and {\tt put} operations).

The monad tranformer typeclass is defined as:

\begin{code}
class MonadTrans t where
  lift :: Monad m => m a -> t m a
\end{code}

Haskell also supports {\em do}-notation for monads. Do notation simply
makes monadic application implicit between lines and reverses the syntatic
order of lambda bindings, providing a way to use monads in a manner similar
to imperative programming. (See Figure~\ref{figure:monad-example} for an
example.)

\begin{figure}
\begin{code}
-- Do notation:
do putStr "Hello: "
   string <- getStr
   putStrLn $ "Nice to meet you " ++ string
   time <- getCurrentTime
   print time
   putStrLn "Look at the time! Bye."

-- Translates to:
putStr "Hello: " >> getStr >>= 
(\string -> (putStrLn $ "Nice to meet you " ++ string) >>
 getCurrentTime >>=
 (\time -> print time >> putStrLn "Look at the time! Bye."))
\end{code}
\hrule
\caption{An example of monadic $do$-notation and its equivalent monadic
expression.}
\label{figure:monad-example}
\end{figure}

\subsection{Constructs and Interface}
\label{section:Implementation-Evaluation_Interface-Constructs_and_Interface}

The evaluation interface is exported as follows:

\begin{code}
data SFEvalState svIn m
data SFEvalT svIn m
data HandlerSet sv m
data EventOccurrence sv
data SignalDelta sv

sd      :: a -> SignalDelta (SVSignal a)
sdLeft  :: SignalDelta svl -> SignalDelta (SVAppend svl svr)
sdRight :: SignalDelta svr -> SignalDelta (SVAppend svl svr)
sdBoth  :: SignalDelta svl -> SignalDelta svr 
        -> SignalDelta (SVAppend svl svr)

eo      :: a -> EventOccurrence (SVEvent a)
eoLeft  :: EventOccurrence svl -> EventOccurrence (SVAppend svl svr)
eoRight :: EventOccurrence svr -> EventOccurrence (SVAppend svl svr)

sdHS    :: (a -> m ()) -> HandlerSet m (SVSignal a)
eoHS    :: (a -> m ()) -> HandlerSet m (SVEvent a)
hsLeft  :: HandlerSet m svl -> HandlerSet m (SVAppend svl svr)
hsRight :: HandlerSet m svr -> HandlerSet m (SVAppend svl svr)
hsBoth  :: HandlerSet m svl -> HandlerSet m svr 
        -> HandlerSet m (SVAppend svl svr)

initSFEval :: HandlerSet svOut m 
           -> SF NonInitialized svIn svOut 
           -> SFEvalState svIn m

runSFEvalT :: SFEvalT svIn m a -> SFEvalState svIn m -> m a

push   :: EventOccurrence svIn -> SFEvalT svIn m ()
update :: SignalDelta svIn -> SFEvalT svIn m ()
sample :: Double -> SFEvalT svIn m ()
\end{code}

The {\tt SFEvalState} type constructor parameterizes over input types for signal
functions, and underlying monads for a monad transformer, but is not itself
a monad transformer. It describes the state of signal function evaluation.
It consists of a record with four members: the current signal function,
the set of handlers for outputs, the current input signal delta, and the last
sample time.

The {\tt SFEvalT} monad transformer is a newtype wrapper around the {\tt StateT}
monad available in the Haskell {\tt transformers} package. The StateT monad
wraps any other monad, providing a state which may be read using the {\tt get}
action and replaced with the {\tt put} action. An action in the underlying
monad is obtained by supplying a {\tt StateT} value and an initial state
value to the {\tt runStateT} function.

The {\tt SFEvalT} monad transformer does not make the {\tt get} and {\tt put}
actions available, but uses them in its implementation of the {\tt push},
{\tt update}, and {\tt sample} actions.

The {\tt push} action pushes an event onto the input of the signal function,
resulting in immediate evaluation of the relevant components signal function
and possible actuation of handlers specified for output events. It is
implemented by fetching the {\tt SFEvalState}, applying the event continuation
of the signal function contained by the {\tt SFEvalState} to the pushed event,
thus obtaining a new signal function and a list of events, applying the handlers
contained in the {\tt SFEvalState} to the output events, replacing the signal
function in the {\tt SFEvalState}, and replacing the {\tt SFEvalState}.

### 2012-06-23 Docs/Thesis/implementation-signalfunctions.tex
Split implementation tex files

\section{Signal Functions}
\label{section:Implementation-Signal_Functions}

To encode signal functions, we must create a compositional way to express
both the output and replacement responses to inputs. We must also have a means
of specifying the initial input and output signal of the signal function.

The interface types for signal functions are given in 
Figure~\ref{figure:signal-function-interface-types}.

\begin{figure}
\begin{minipage}[b]{0.5\linewidth}
\begin{code}
-- Signal memories
data SMemory p sv

-- Signal vector indices
data SVIndex p sv

-- Initialization types
data Initialized
data NonInitialized
\end{code}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[b]{0.5\linewidth}
\begin{code}
-- Signal vectors
data SVEmpty
data SVSignal a
data SVEvent a
data SVAppend svLeft svRight

-- Signal functions
data SF init svIn svOut
\end{code}
\end{minipage}
\vspace{\parskip}
\hrule
\caption{Interface types for signal functions}
\label{figure:signal-function-interface-types}
\end{figure}

Our implementation makes use of two value representations of signal vectors.
A signal memory is typed using a signal vector and carries values for a
(possibly empty) subset of the points in the signal vector. A signal vector
index is typed using a signal vector and carries a value for exactly one point
in the signal vector. In order to define memories and indices over signal
vectors, we make use of a type extension known as
Generalized~Algebraic~Datatypes~(GADTs)~\cite{Cheney2003,Xi2003}, which is
available in recent versions of the
Glasgow~Haskell~Compiler~(GHC)~\cite{PeytonJones2006}. The key distinction
of GADTs from standard Algebraic~Datatypes is that GADTs enable 
{\em type refinement}. A type constructor in a GADT specifies the specific type
with which each type parameter of the GADT's type constructor is instantiated.
This type is defined in terms of the types of the constructor's data members.
Any type variable which occurs in the type of a data member but does not occur
in the type of the constructor is existentially quantified\footnote{Existential
quantification, in the context of Algebraic Datatypes, is a means of ``hiding''
type variables in data member types. A function which pattern matches on a
constructor with existential quantification may not assume any instantiation of
the existentially quantified variable, but only typeclass constraints specified
in the constructor definition, and equality between identical type variables in
data members of the same constructor.}. GADTs generalize Algebraic Datatypes 
(ADTs) because an ADT may be translated to a GADT by simply using each unique
type variable in the types of the data members of a constructor to instantiate
the type parameter. As with ADTs, a type variable which appears in the type of a
data constructor but not in any data members may be arbitrarily instantiated.

The ability to specialize types is useful both for creating specialized
constructors for unique cases of the type parameters (e.g. for
optimization~\cite{Nilsson2005}) and for encoding properties at the type level
using empty types. We employ the second use here.% and the first later in Section~\ref{subsection:Implementation-Signal_Functions-Representation} when we describe the representation of signal functions.

Both signal memories and signal vector indices are GADTs:

\begin{code}
-- Signal memories
data SMemory :: (* -> *) -> * -> * where
  SMEmpty  :: SMemory p sv
  SMSignal :: p a -> SMemory p (SVSignal a)
  SMEvent  :: p a -> SMemory p (SVEvent a)
  SMBoth   :: SMemory p svLeft -> SMemory p svRight 
           -> SMemory p (SVAppend svLeft svRight)

-- Signal vector indices
data SVIndex :: (* -> *) -> * -> * where
  SVISignal :: p a -> SVIndex p (SVSignal a)
  SVIEvent  :: p a -> SVIndex p (SVEvent a)
  SVILeft   :: SVIndex p svLeft -> SVIndex p (SVAppend svLeft svRight)
  SVIRight  :: SVIndex p svRight -> SVIndex p (SVAppend svLeft svRight)

-- ``Identity'' newtype
newtype Id a = Id a
\end{code}

The higher-kind type parameter {\tt p} in both type constructors is a
convenience. It allows us to provide a type constructor which will be mapped
over the types of signals and events in the signal vector to produce the actual
type of values stored. This ability is used in
Section~\ref{section:Implementation-Evaluation_Interface} (Evaluation Interface),
permitting the memory type, in particular, to carry functions from the types of
points in the signal vector to the monadic action used to handle the
corresponding outputs.

Next we consider the type for signal functions. A signal function must be able
to have an initial signal input specified, and must be able to produce output
signals, output events, and replace itself, based on time steps and input
signals or event occurrences.

The definition of the {\tt SF} datatype is:

\begin{code}
data SF :: * -> * -> * -> * where
  SF     :: (SMemory Id svIn -> (SMemory Id svOut, SF Initialized svIn svOut))
         -> SF NonInitialized svIn svOut
  SFInit :: (Double -> SMemory Id svIn 
             -> (SMemory Id svOut, 
                 [SVIndex Id svOut],
                 SF Initialized svIn svOut))
         -> (SVIndex Id svIn 
             -> ([SVIndex Id svOu], SF Initialized svIn svOut))
         -> SF Initialized svIn svOut
\end{code}

The first constructor is used in the user-facing combinators to construct
non-initialized signal functions. It contains a single function from an intial
signal (which may contain empty values) to an initial output signal and
an initialized signal function with the same input and output types.

The second constructor is used in the implementation of both the non-initialized
and initialized signal functions to construct initialized signal functions. It
takes two functions. The first is a function from a time step and an updated
signal to an updated output signal, a set of output event occurrences, and a new
signal function. The second is a function from an event occurrence to a set
of event occurrences and a new signal function.

\subsection{Combinators}
\label{subsection:Implementation-Signal_Functions-Combinators}
From the perspective of the library user, signal functions are built from a set
of combinators which produce or modify signal functions. These combinators
include a set of primitive signal functions, a set of lifting functions to
produce signal functions from functions, a switching combinator,
time and integration combinators, and a set of routing combinators. A list
of the exported combinators is given in
Figure~\ref{figure:signal_function_combinators}.

\begin{figure}
\begin{code}
-- Simple signal functions
identity :: SignalFunction NonInitialized sv sv
constant :: a -> SignalFunction NonInitialized SVEmpty (SVSignal a)

-- Composition and routing
(>>>) :: SignalFunction NonInitialized svIn svBetween
      -> SignalFunction NonInitialized svBetween svOut
      -> SignalFunction NonInitialized svIn svOut

first :: SignalFunction NonInitialized svIn svOut
      -> SignalFunction NonInitialized (SVAppend svIn sv) (SVAppend svOut sv)

second :: SignalFunction NonInitialized svIn svOut
       -> SignalFunction NonInitialized (SVAppend sv svIn) (SVAppend sv svOut)

swap :: SignalFunction NonInitialized 
          (SVAppend svLeft svRight) (SVAppend svRight svLeft)
copy :: SignalFunction NonInitialized sv (SVAppend sv sv)
drop :: SignalFunction NonInitialized sv SVEmpty
\end{code}
\hrule
\caption{The combinators exported by the signal function interface}
\label{figure:signal_function_combinators}
\end{figure}

The simplest combinator is the {\tt identity} signal function. As its name would
suggest, the identity signal function simply replicates its input on its output.
It is implemented as:

\begin{code}
identity :: SF NonInitialized sv sv
identity = SF (\mem -> (identityInit, mem))

identityInit :: SF Initialized sv sv
identityInit = let sf = SFInit (\_ mem -> (mem, [], sf))
                               (\idx -> ([idx], sf))
               in sf
\end{code}

Only the first function is exported. In general, we use the convention that
the initialized version of a signal function is named with the signal function
name with {\tt Init} appended, and is not exported as part of the interface.

### 2012-06-23 Docs/Thesis/implementation.tex
Split implementation tex files

@@
 \chapter{Implementation}
-\label{chapter:Implemenation}
+\label{chapter:Implementation}
 
-Having established our method of modeling reactivity, we proceed to a purely 
-functional implementation of this method. For this implementation, we will make
-significant use of certain advanced extensions to the Haskell type system.
+Having established the semantics of the reactive system, I present a purely 
+functional implementation. For this implementation, I will make significant use
+of extensions to the Haskell type system. Properly used, these extensions allow
+greater expressive flexibility to define our types by permitting us to encode
+additional properties of values in their types.
+
+\input{implementation-signalfunctions}
+
+\input{implementation-evalinterface}

### 2012-06-28 Docs/Thesis/system_design_and_interface.tex
Change chapter semantics->system design and interface

\chapter{System Design and Interface}
\label{chapter:System_Design_and_Interface}

We have explored the current design space of FRP systems, and identified
strengths and weaknesses with the existing state-of-the-art systems. By
combining the best characteristics of both state-of-the-art systems, we hope
to produce an FRP system which combines the efficiency of push-based FRP system
with the modularity and causality safety of a signal-function FRP system.

The advantage of a push-based system is that it evaluates only those parts
of the network which are necessary to react to the input. In order to achieve
this behavior with signal functions, we must have a means of allowing the system
to distinguish between multiple inputs to a signal function. This is where the
concept of signal vectors from N-Ary FRP will be useful.

### 2012-06-28 Code/demo-app/src/Main.hs
Initial demo app (copied from mouse-color)

@@ -1,0 +1,70 @@
+module Main where
+
+import Control.Monad.IO.Class
+
+import Data.IORef
+import Data.Time.Clock.POSIX
+
+import FRP.TimeFlies.SignalFunctions
+
+import Graphics.Rendering.OpenGL
+import Graphics.UI.GLUT
+
+
+-- Skeleton from http://www.haskell.org/haskellwiki/OpenGLTutorial1
+myPoints :: [(GLfloat, GLfloat, GLfloat)]
+myPoints = map (\k -> (sin (2 * pi * k / 12) ,cos ( 2 * pi * k /12) ,0.0 ) ) [1..12]
+
+main :: IO ()
+main = do 
+  (progname, _) <- getArgsAndInitialize
+  createWindow "Hello World"
+  displayRef <- newIORef $ return ()
+
+  time <- fmap realToFrac getPOSIXTime
+  
+  let mouseApplicationState = initSFEval (sdHS (\act -> writeIORef displayRef act >> postRedisplay Nothing)) time mouseApplicationSF
+  sfRef <- newIORef mouseApplicationState
+  
+  keyboardMouseCallback $= Just (\key keystate _ pos -> do sfSt <- readIORef sfRef
+                                                           ((), sfSt') <- runSFEvalT (do case keystate of
+                                                                                           Down -> push $ eoRight $ eo key
+                                                                                           Up -> return ()
+                                                                                         update (sdLeft $ sd pos))
+                                                                                     sfSt
+                                                           writeIORef sfRef sfSt')
+
+  passiveMotionCallback $= Just (\pos -> do sfSt <- readIORef sfRef
+                                            ((), sfSt') <- runSFEvalT (update (sdLeft $ sd pos))
+                                                                      sfSt
+                                            writeIORef sfRef sfSt')  
+
+
+
+  initialDisplayMode $= [DoubleBuffered]
+  displayCallback $= do clear [ColorBuffer]
+                        act <- readIORef displayRef
+                        act
+                        swapBuffers
+                        flush
+
+  addTimerCallback 33 $ let tc = do addTimerCallback 33 tc
+                                    sfSt <- readIORef sfRef
+                                    ((), sfSt') <- runSFEvalT (do time <- fmap realToFrac $ liftIO getPOSIXTime
+                                                                  sample time)
+                                                   sfSt
[…]

### 2012-06-28 Code/demo-app/src/Main.hs
Update demo app to pay attention to window resize.

@@ -28,14 +28,14 @@ main = do
   
   keyboardMouseCallback $= Just (\key keystate _ pos -> do sfSt <- readIORef sfRef
                                                            ((), sfSt') <- runSFEvalT (do case keystate of
-                                                                                           Down -> push $ eoRight $ eo key
+                                                                                           Down -> push $ eoRight $ eoRight $ eo key
                                                                                            Up -> return ()
-                                                                                         update (sdLeft $ sd pos))
+                                                                                         update (sdRight $ sdLeft $ sd pos))
                                                                                      sfSt
                                                            writeIORef sfRef sfSt')
 
   passiveMotionCallback $= Just (\pos -> do sfSt <- readIORef sfRef
-                                            ((), sfSt') <- runSFEvalT (update (sdLeft $ sd pos))
+                                            ((), sfSt') <- runSFEvalT (update (sdRight $ sdLeft $ sd pos))
                                                                       sfSt
                                             writeIORef sfRef sfSt')  
 
@@ -48,6 +48,20 @@ main = do
                         swapBuffers
                         flush
 
+  reshapeCallback $= Just (\size -> do sfSt <- readIORef sfRef
+                                       ((), sfSt') <- runSFEvalT (update (sdLeft $ sd size)) sfSt
+                                       writeIORef sfRef sfSt'
+                                       viewport $= ((Position 0 0), size)
+                          )
+
+  sfSt <- readIORef sfRef
+  ((), sfSt') <- runSFEvalT (do size <- liftIO $ get windowSize
+                                update (sdLeft $ sd size))
+                            sfSt
+                 
+  writeIORef sfRef sfSt
+
+
   addTimerCallback 33 $ let tc = do addTimerCallback 33 tc
                                     sfSt <- readIORef sfRef
                                     ((), sfSt') <- runSFEvalT (do time <- fmap realToFrac $ liftIO getPOSIXTime
@@ -59,12 +73,14 @@ main = do
   mainLoop
 
 
-mouseApplicationSF :: SF NonInitialized (SVAppend (SVSignal Position) (SVEvent Key)) (SVSignal (IO ()))
-mouseApplicationSF = second ignore >>> cancelRight >>> pureSignalTransformer (\(Position x y) -> let r = fromIntegral (x `mod` 50) / 50 :: GLdouble
-                                                                                                     g = fromIntegral (y `mod` 50) / 50 :: GLdouble
-                                                                                                     b = 0.5 :: GLdouble
[…]

### 2012-06-29 Docs/Thesis/background.tex
Rewrite background

@@
 \section{Classic FRP}
-\label{section:Background-classic_frp}
+\label{section:Background-Classic_FRP}
 
-The first FRP system, Fran~\cite{Elliott1997}, was introduced as a system for compositionally describing interactive animations.
-Though the term ``Functional Reactive Programming'' was not used in this work, it was the first to introduce the concept of first-class
-behaviors and events as an abstract model for reactive programs. The key motivation was to separate the implementation of reactivity
-from the modelling of reactivity, so that a programmer could be freed from details such as sampling time-varying values, time-slicing
-to simulate updating values in parallel, and sampling of inputs.
+The earliest and still standard formulation of FRP is two primitive type
+constructors: {\tt Behavior} and {\tt Event}, together with combinators that
+produce values of this type. The easiest semantic definition for these types
+is given in Figure~\ref{figure:classic_frp_semantic_types}
 
-The first implementation of FRP outside the Haskell language was Frapp\'{e}, an FRP implementation for the Java Beans framework. Frapp\'{e} built on
-the notion of events and {\em bound properties} in the Beans framework, providing abstract interfaces for FRP events and behaviors, and combinators
-as concrete classes implementing these interfaces. The evaluation of Frapp\'{e} used Bean components as sources and sinks, and the implementation of
-Bean events and bound properties to propagate changes to the network~\cite{Courtney2001-2}.
+\begin{figure}
+\begin{code}
+type Event a = [(Time, a)]
+type Behavior a = Time -> a
+\end{code}
+\hrule
+\caption{Semantic types for Classic FRP}
+\label{figure:classic_frp_semantic_types}
+\end{figure}
 
-The FrTime\footnote{FrTime is available in the current version of Dr Racket (5.2.1).}~\cite{Cooper2006} language extends the Scheme evaluator with a mutable
-dependency graph, which is constructed by program evaluation. This graph is then updated by signal changes. FrTime does not provide a distinct concept of
-events, and chooses branches of the dependency graph by conditional evaluation of signals, rather than by the substitution of signals used by FRP systems.
+When these two type constructors are exposed directly, the system is known as a 
+{\em Classic FRP} system. If the type aliases are taken as given in the semantic
+definition, a simple, yet problematic, implementation is given in
+Figure~\ref{figure:classic_frp_simple_implementation}.
 
-The Reactive~\cite{Elliott2009} system is a push-based FRP system with first-class behaviors and events. The primary insight of Reactive is
-the separation of reactivity (or changes in response to events whose occurrence time could not be know beforehand) from time-dependence. This
-gives rise to {\em reactive normal form}, which represents a behavior as a constant or simple time-varying value, with an event stream carrying values
-which are also behaviors in reactive normal form. Push-based evaluation is achieved by forking a Haskell thread to evaluate the head behavior,
-while waiting on the evaluation of the event stream. Upon an event occurrence, the current behavior thread is killed and a new thread
-spawned to evaluate the new behavior. Unfortunately, the implmentation of Reactive uses a tenuous technique which depends on also forking threads to evaluate
-two haskell values concurrently in order to implement event merging. This relies on the library author to ensure consistency when this technique is
-used, and leads to thread leakage when one of the merged events is itself the merging of other events.
+\begin{figure}
+\begin{code}
+time :: Behavior Time
+time = id
 
-The reactive-banana~\cite{Apfelmus} library is a push-based FRP system designed for use with Haskell GUI frameworks. In particular, it features a monad for
-the creation of behaviors and events which may then be composed and evaluated. This monad includes constructs for binding GUI library constructs to primitive events.
-It must be ``compiled'' to a Haskell IO action for evaluation to take place.
-The implementation of reactive-banana is similar to FrTime, using a dependency graph to update the network on event occurences.
-Reactive-banana eschews generalized switching in favor of branching functions on behavior values, similarly to FrTime, but
-maintains the distinction between behaviors and events. Rather than a generalized switching combinator which allows the replacement of arbitrary behaviors,
-reactive-banana provides a step combinator which creates a stepwise behavior from the values of an event stream.
+constant :: a -> Behavior a
+constant = const
 
-A recent thesis~\cite{Czaplicki2012} described Elm, a stand-alone language for reactivity. Elm provides combinators for manipulating discrete events, and
-compiles to Javascript, making it useful for client-side web programming. However, Elm does not provide a notion of switching or of continuous time behaviors,
-though an approximation is given using discrete-time events which are actuated at repeated intervals specified during the event definition. The thesis asserts
-that Arrowized FRP (signal-function FRP, Section~\ref{section:Background-signal_function_frp}) can be embedded in Elm, but provides no support for this assertion.
+delayB :: a -> Time -> Behavior a -> Behavior a
+delayB a td b te = if te <= td
+                   then a
+                   else b (te - td)
+
+instance Functor Behavior where
+  fmap = (.)
+
+instance Applicative Behavior where
+  pure = constant
+  bf <*> ba = (\t -> bf t (ba t))
+
+never :: Event a
+never = []
+
+once :: a -> Event a
+once a = [(0, a)]
+
+delayE :: Time -> Event a -> Event a
+delayE td = map (\(t, a) -> (t + td, a))
+
+instance Functor Event where
+  fmap f = map (\(t, a) -> (t, f a))
+
+switcher :: Behavior a -> Event (Behavior a) -> Behavior a
+switcher b [] t = b t
+switcher b [(to, bo)] t = if t < to
+                          then b t
+                          else bo t
+\end{code}
+\hrule
+\caption{An obvious, yet inefficient and problematic, implementation of
+Classic FRP}
+\label{figure:classic_frp_simple_implementation}
+\end{figure}
+
+Obvious problems include the lack of ordering enforcement on event occurrences,
+the necessity of waiting for the first event occurrence in a switch, and the
+necessity of maintaining the full history of a behavior for evaluation.
+
+\subsection{History of Classic FRP}
+\label{subsection:Background-Classic_FRP-History_of_Classic_FRP}
+Classic FRP was originally described as the basis of Fran~\cite{Elliott1997}
+(Functional Reactive ANimation), a framework for declaratively specifying
+interactive animations. Fran represented behaviors as two sampling functions,
+one from a time to a value and a new behavior (so that history may be discared),
+and the other from a time interval (a lower and upper time bound) to a value
+interval and a new behavior. Events are represented as {\tt improving values},
+which, when sampled with a time, produce a lower bound on the time to the next
+occurrence, or the next occurrence if it has indeed occurred.
+
+The first implementation of FRP outside the Haskell language was Frapp\'{e}, an
+FRP implementation for the Java Beans framework. Frapp\'{e} built on the notion
+of events and {\em bound properties} in the Beans framework, providing abstract
+interfaces for FRP events and behaviors, and combinators as concrete classes
+implementing these interfaces. The evaluation of Frapp\'{e} used Bean components
+as sources and sinks, and the implementation of Bean events and bound properties
+to propagate changes to the network~\cite{Courtney2001-2}.
+
+\subsection{Current Classic FRP Systems}
+\label{subsection:Background-Classic_FRP-Current_Classic_FRP_Systems}
+
+The FrTime\footnote{FrTime is available in the current version of Dr Racket (5.2.1).}~\cite{Cooper2006}
+language extends the Scheme evaluator with a mutable dependency graph, which is
+constructed by program evaluation. This graph is then updated by signal changes.
+FrTime does not provide a distinct concept of events, and chooses branches of
+the dependency graph by conditional evaluation of signals, rather than by the
+substitution of signals used by FRP systems.
+
+The Reactive~\cite{Elliott2009} system is a push-based FRP system with
+first-class behaviors and events. The primary insight of Reactive is the
+separation of reactivity (or changes in response to events whose occurrence time
+could not be known beforehand) from time-dependence. This gives rise to {\em
+reactive normal form}, which represents a behavior as a constant or simple
+time-varying value, together with an event stream carrying values which are also
+behaviors in reactive normal form. Push-based evaluation is achieved by forking
+a Haskell thread to evaluate the head behavior, while waiting on the evaluation
+of the event stream. Upon an event occurrence, the current behavior thread is
+killed and a new thread spawned to evaluate the new behavior. Unfortunately, the
+implmentation of Reactive uses a tenuous technique which depends on also forking
+threads to evaluate two haskell values concurrently in order to implement event
+merging. This relies on the library author to ensure consistency when this
+technique is used, and leads to thread leakage when one of the merged events is
+itself the merging of other events.
+
+A recent thesis~\cite{Czaplicki2012} described Elm, a stand-alone language for
+reactivity. Elm provides combinators for manipulating discrete events, and
+compiles to Javascript, making it useful for client-side web programming.
+However, Elm does not provide a notion of switching or of continuous time
+behaviors, though an approximation is given using discrete-time events which are
+actuated at repeated intervals specified during the event definition. The thesis
+asserts that Arrowized FRP (signal-function FRP, Section~\ref{section:Background-signal_function_frp})
+can be embedded in Elm, but provides no support for this assertion.
+
+The reactive-banana~\cite{Apfelmus} library is a push-based FRP system designed
+for use with Haskell GUI frameworks. In particular, it features a monad for
+the creation of behaviors and events which may then be composed and evaluated.
+This monad includes constructs for binding GUI library constructs to primitive
+events. It must be ``compiled'' to a Haskell IO action for evaluation to take
+place. The implementation of reactive-banana is similar to FrTime, using a
+dependency graph to update the network on event occurences. Reactive-banana
+eschews generalized switching in favor of branching functions on behavior
+values, similarly to FrTime, but maintains the distinction between behaviors and
+events. Rather than a generalized switching combinator which allows the
+replacement of arbitrary behaviors, reactive-banana provides a step combinator
+which creates a stepwise behavior from the values of an event stream.
 
@@
 
-An alternative approach to FRP was first proposed in work on Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs. This library
-utilized the abstraction of Arrows~\cite{Hughes2000} to structure {\em signal functions}. Arrows are abstract type constructors with input and output type
-parameters, together with sequential ({\tt >>>}) and parallel ({\tt first} and {\tt second}) composition, branching ({\tt ***}), and lifting ({\tt arr}) functions
-to the Arrow type. Signal functions are the first-class abstraction in this approach, they represent time-dependent and reactive transformers of events and signals, 
-which are themselves not first class, since such values cannot be directly manipulated by the programmer.
-This approach has two motivations: it increases modularity since both the input and output of signal functions may be transformed,
-as opposed to signals or events which may only be transformed in their output, and it avoids a large class of time and space leaks which have emerged when
-implementing FRP with first-class signals and events.
+An alternative approach to FRP was first proposed in work on
+Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs.
+This library utilized the abstraction of Arrows~\cite{Hughes2000} to structure
+{\em signal functions}. Arrows are abstract type constructors with input and
+output type parameters, together with sequential ({\tt >>>}) and parallel
+({\tt first} and {\tt second}) composition, branching ({\tt ***}), and lifting
+({\tt arr}) functions to the Arrow type. Signal functions are the first-class
+abstraction in this approach, they represent time-dependent and reactive
+transformers of events and signals,  which are themselves not first class,
+since such values cannot be directly manipulated by the programmer. This
+approach has two motivations: it increases modularity since both the input and
+output of signal functions may be transformed, as opposed to signals or events
+which may only be transformed in their output, and it avoids a large class of
+time and space leaks which have emerged when implementing FRP with first-class
+signals and events.
 
-Similarly to FrTime, the netwire~\cite{Soylemez} library eschews dynamic switching, in this case in favor of {\em signal inhibition}. Netwire is written as an arrow
-transformer, which, together with the Kliesli arrow instance for the IO monad\footnote{Any monad forms a Kliesli arrow.}, permits it to lift IO actions as sources and
-sinks at arbitrary points in a signal function network. Signal inhibition is accomplished by making the output of signal functions a monoid, and then combining the
-outputs of signal functions. An inhibited signal function will produce the monoid's zero as an output. Primitives have defined inhibition behavior, and composed signal
-functions inhibit if their outputs combine to the monoid's zero.
+Similarly to FrTime, the netwire~\cite{Soylemez} library eschews dynamic
+switching, in this case in favor of {\em signal inhibition}. Netwire is written
+as an arrow transformer, which, together with the Kliesli arrow instance for the
+IO monad\footnote{Any monad forms a Kliesli arrow.}, permits it to lift IO
+actions as sources and sinks at arbitrary points in a signal function network.
+Signal inhibition is accomplished by making the output of signal functions a
+monoid, and then combining the outputs of signal functions. An inhibited signal
+function will produce the monoid's zero as an output. Primitives have defined
+inhibition behavior, and composed signal functions inhibit if their outputs
+combine to the monoid's zero.
 
-Yampa~\cite{Nilsson2005} is an optimization of the Arrowized FRP system first utilized with Fruit (see above).
-The implementation of Yampa makes use of Generalized Algebraic Datatypes to permit a much larger class of type-safe datatypes for the signal function representation.
-This representation, together with ``smart''
-constructors and combinators, enables the construction of a self-optimizing arrowized FRP system. Unfortunately, the primary inefficiency, that of unnecessary evaluation
-steps due to pull-based evaluation, remains. Further, the optimization is ad-hoc and each new optimization requires the addition of new constructors, as well
-as the updating of every primitive combinator to handle every combination of constructors. However, Yampa showed noticeable performance gains over previous Arrowized FRP implementations.
+Yampa~\cite{Nilsson2005} is an optimization of the Arrowized FRP system first
+utilized with Fruit (see above). The implementation of Yampa makes use of
+Generalized Algebraic Datatypes to permit a much larger class of type-safe
+datatypes for the signal function representation. This representation,
+together with ``smart'' constructors and combinators, enables the construction
+of a self-optimizing arrowized FRP system. Unfortunately, the primary
+inefficiency, that of unnecessary evaluation steps due to pull-based evaluation,
+remains. Further, the optimization is ad-hoc and each new optimization requires
+the addition of new constructors, as well as the updating of every primitive
+combinator to handle every combination of constructors. However, Yampa showed
+noticeable performance gains over previous Arrowized FRP implementations.
 
-A recent PhD thesis~\cite{Sculthorpe2011} introduced N-Ary FRP, a technique for typing Arrowized FRP systems using dependent types. The bulk of the work consisted
-in using the dependent type system to prove the correctness of the FRP system discussed. This work introduced the typing construct of signal vectors,
-which permit the distinction of signal and event types at the level of the FRP system, instead of making events merely a special type of signal.
+A recent PhD thesis~\cite{Sculthorpe2011} introduced N-Ary FRP, a technique for
+typing Arrowized FRP systems using dependent types. The bulk of the work
+consisted in using the dependent type system to prove the correctness of the FRP
+system discussed. This work introduced the typing construct of signal vectors,
+which permit the distinction of signal and event types at the level of the FRP
+system, instead of making events merely a special type of signal.
 
@@
 
-At present, there are two key issues apparent with FRP. First, while signal-function FRP is inherently safer and more modular than classic FRP, it has yet to be
-efficiently implemented. Second, the interface between FRP programs and the many different sources of inputs and sinks for outputs available to a modern application
-writer remains ad-hoc and is in most cases limited by the implementation. One key exception to this is the reactive-banana system, which provides a monad for
-constructing primitive events and behaviors from which an FRP program may then be constructed. However, this approach is still inflexible as it requires library support
-for the system which the FRP program will interact with. Further, classic FRP programs are vulnerable to time leaks and violations of causality due to the ability
-to directly manipulate reactive values.
+At present, there are two key issues apparent with FRP. First, while
+signal-function FRP is inherently safer and more modular than classic FRP, it
+has yet to be efficiently implemented. Classic FRP programs are vulnerable to
+time leaks and violations of causality due to the ability to directly manipulate
+reactive values. Second, the interface between FRP programs and the many
+different sources of inputs and sinks for outputs available to a modern
+application writer remains ad-hoc and is in most cases limited by the
+implementation.
+
+One key exception to this is the reactive-banana system, which provides a monad
+for constructing primitive events and behaviors from which an FRP program may
+then be constructed. However, this approach is still inflexible as it requires
+library support for the system which the FRP program will interact with.
+Further, being a Classic FRP system, reactive-banana lacks the ability to
+transform the inputs of behaviors and events, since all inputs are implicit.

### 2012-07-12 Docs/Thesis/system_design_and_interface.tex
Partial draft of system design chapter

@@
 
-We have explored the current design space of FRP systems, and identified
-strengths and weaknesses with the existing state-of-the-art systems. By
-combining the best characteristics of both state-of-the-art systems, we hope
-to produce an FRP system which combines the efficiency of push-based FRP system
-with the modularity and causality safety of a signal-function FRP system.
+\section{Goals}
+\label{section:System_Design_and_Interface-Goals}
 
-The advantage of a push-based system is that it evaluates only those parts
-of the network which are necessary to react to the input. In order to achieve
-this behavior with signal functions, we must have a means of allowing the system
-to distinguish between multiple inputs to a signal function. This is where the
-concept of signal vectors from N-Ary FRP will be useful.
+The (not yet fully realized) goal of FRP is to provide an efficient, declarative
+abstraction for creating reactive programs. Towards this overall goal, there are
+three goals which this system is intended to meet.
+
+\subsection{Efficient and Push-Based Evaluation}
+\label{subsection:System_Design_and_Interface-Goals-Efficient_and_Push_based_Evaluation}
+
+Efficient evaluation is the for push-based evaluation. Since FRP programs are
+expected to  interact with an outside world in real time, efficiency cannot be
+measured by runtime. Thus, when speaking of efficiency, we are expressing a
+desire that the system utilize as few system resources as possible for the task
+at hand, while responding as quickly as possible to external inputs and
+producing output at a consistently high sample rate.
+
+\subsection{Composability}
+\label{subsection:System_Design_and_Interface-Goals-Composability}
+
+A composable abstraction is one in which values in that abstraction may be
+combined in such a way that reasoning about their actions together involves
+little more than reasoning about their actions separately. In a signal function
+system, the only interaction between composed signal functions ought to be that
+the output of one is the input of another. Composability permits a particularly
+attractive form of software engineering in which successively larger systems are
+created from by combining smaller systems, without having to reason about the 
+components of the systems being combined.
+
+\subsection{Ease of Integration}
+\label{subsection:System_Design_and_Interface-Goals-Ease_of_Integration}
+
+It is fine for a system to be composable with regards to itself, but an FRP
+system must interact with the outside world. Since we cannot anticipate every
+possible form of input and output that the system will be asked to interact
+with, we must interface with Haskell's IO system. In particular, most libraries
+for user interaction (e.g. GUI and graphics libraries such as GTK+ and GLUT) and
+most libraries for time-dependent IO (e.g. audio and video systems) make use of
+the event loop abstraction. In this abstraction, event handlers are registered
+with the system, and then a command is issued to run a loop which detects events
+and runs the handlers, and uses the results of the handlers to render the
+appropriate output.
+
+We would like for the FRP system to be easily to integrate with such IO systems,
+while being flexible enough to enable its use with other forms of IO systems,
+such as simple imperative systems, threaded systems, or network servers.
+
+\section{Types}
+\label{section:System_Design_and_Interface-Types}
+
+In a strongly and statically typed functional language, types are a key part of
+an interface. Types provide a mechanism for describing and ensuring properties
+of the interface's components and about the systems created with these
+components.
+
+\subsection{Signal Vectors}
+\label{subsection:System_Design_and_Interface-Types-Signal_Vectors}
+
+In order to type signal functions, we must be able to describe their input and
+output. In most signal function systems, a signal function takes exactly one
+input and produces exactly one output. Multiple inputs or outputs are handled
+by making the output a tuple, and combinators which combine or split the inputs
+or outputs of a signal assume this. Events are represented at the type level
+as a particular type of signal, and at the value level as an option, either an
+event occurrence or not.
+
+This method of typing excludes push-based evaluation at the outset.
+It is not possible to construct a "partial tuple" nor in general is it possible
+to construct only part of any type of value. Push-based evaluation depends on
+evaluating only that part of the system which is updated, which means evaluating
+only that part of the input which is updated.
+
+In order to permit the construction of partial inputs and outputs, we make use
+of signal vectors. Signal vectors are uninhabited types which describe the input
+and output of a signal function. Singleton vectors are parameterized over the
+type carried by the signal or by event occurrences. The definition of the signal
+vector type is shown in Figure~\ref{figure:signal_vector_types}.
+
+Having an uninhabited signal vector type allows us to construct representations
+of inputs and outputs which are hidden from the user of the system, and are
+designed for partial representations.
+
+\begin{figure}
+\begin{code}
+data SVEmpty    -- An empty signal vector component,
+                -- neither event nor signal
+data SVSignal a -- A signal, carrying values of type a
+data SVEvent a  -- An event, whose occurrences carry values of type a
+data SVAppend svLeft svRight -- The combination of the signal vectors
+                             -- svLeft and svRight
+\end{code}
+\hrule
+\caption{Signal vector types}
+\label{figure:signal_vector_types}
+\end{figure}
+
+\subsection{Signal Functions}
+\label{subsection:System_Design_and_Interface-Types-Signal_Functions}
+
+The type constructor for signal functions is shown in
+Figure~\ref{figure:signal_function_types}. For the {\tt init} parameter, only
+one possible instantiation is shown. The usefulness of this type parameter,
+along with another instantation which is hidden from users of the library,
+is discussed in the section on implementation of signal functions
+(Section~\ref{section:Implementation-Signal_Functions}).
+
+The representation of signal functions is also discussed in
+Section~\ref{section:Implementation-Signal_Functions}. Here it suffices to say
+that the use of GADTs permits the construction of values which make use of
+uninhabited types as instantiations of type parameters.
+
+The type synonyms {\tt :~>} and {\tt :\textasciicircum:} are included for readability and are
+not crucial to the FRP system.
+
+\begin{figure}
+\begin{code}
+-- Signal functions
+-- init: The initialization type for 
+-- the signal function, always NonInitialized
+-- for exported signal functions
+-- svIn: The input signal vector
+-- svOut: The output signal vector
+data SF init svIn svOut
+
+data NonInitialized
+
+type svIn :~> svOut = SF NonInitialized svIn svOut
+type svLeft :^: svRight = SVAppend svLeft svRight
+\end{code}
+\hrule
+\caption{Signal function types.}
+\label{figure:signal_function_types}
+\end{figure}
+
+\subsection{Evaluation Monad}
+\label{section:System_Design_and_Interface-Types-Evaluation_Monad}
+
+A {\em monad} is a standard, composable abstraction for writing functions with
+a context, used in Haskell for IO~\cite{PeytonJones1993,PeytonJones2001} among
+other tasks. A monad is simply a 1-arity type constructor together with two
+functions. The first function, {\tt return}, takes a value of type {\tt a} and
+produces a value of type {\tt m a}, where {\tt m} is the type constructor. The
+second, called {\tt bind} and stylized in the Haskell standard library as the
+infix operator {\tt (>>=)}, takes a value of type {\tt m a} and a function
+from {\tt a} to {\tt m b} and produces a value of type {\tt m b}. This allows
+a value to be operated on out of the context and a new context to be assigned.
+
+A monad can have other primitives which manipulate the context in some way. For 
+instance, the primtives in Haskell's {\tt IO} monad produce actions which, when
+interpreted as part of the {\tt main} action, produce some side-effect. The
+{\tt State} monad provides {\tt get} and {\tt put} operations to work with a 
+state value stored in the context.
+
+Monad transformers~\cite{Jones1995} provide a means to combine the functionality
+of multiple monads. A monad transformer is a monad with an extra type parameter.
+This type parameter is instantiated with the type constructor of the underlying
+monad, and an extra operation {\tt lift} is provided which converts values in
+the underlying monad to values in the monad transformer.
+
+The evaluation monad is a monad transformer. This permits it to be used in
+conjunction with the {\tt IO} monad (or any other monad) to describe how input
+is to be obtained for the signal function being evaluated, and how outputs are
+to be handled.
+
+The evaluation monad, in addition to the standard monad operators, provides a
+means of {\em initializing} a signal function, and a means of translating the
+monadic value describing evaluation to a value in the underlying monad. This
+means, for instance, that we can obtain an action in the {\tt IO} monad to
+evaluate a signal function.
+
+The type of the evaluation monad must track the input type of the signal
+function. The monad's context stores a mapping from outputs to handling actions.
+An existential type can thus be used to ``hide'' the output type of the signal
+function. However, inputs must come from external values, so the input type
+cannot be hidden. There are thus three type parameters to the monad's type
+constructor: the input signal vector, the type of the underlying monad, and the
+monadic type parameter. The type is shown in Figure~\ref{figure:evaluation_monad_types}.
+
+\begin{figure}
+\begin{code}
+-- A signal function's evaluations state
+data SFEvalState svIn m
+-- The evaluation monad
+data SFEvalT svIn m a
+\end{code}
+\hrule
+\caption{Evaluation monad types.}
+\label{figure:evaluation_monad_types}
+\end{figure}
+
+\section{Combinators}
+\label{section:System_Design_and_Interface-Combinators}
+
+Signal functions are constructed from combinators, which are primitive signal
+functions and operations to combine these primitives. These combinators are
+grouped as basic signal functions, lifting operations for pure functions,
+routing, reactivity, feedback, and time dependence.
+
+\subsection{Basic Signal Functions}
+\label{subsection:System_Design_and_Interface-Combinators-Basic_Signal_Functions}
+
+The basic signal functions (Figure~\ref{figure:basic_signal_functions})
+provide very simple operations. The {\tt identity} signal function, as expected,
+simply copies its input to its output. The {\tt constant} signal function
+produces the provided value as a signal at all times. The {\tt never} signal
+function has an event output which never produces occurrences. The {\tt asap}
+function produces an event occurrence with the given value at the first time
+step after it is switched into the network. The {\tt after} function waits for
+the specified amount of time before producing the event occurrence.
+
+With the exception of {\tt identity}, all of the basic signal functions have
+empty inputs. This allows these signal functions to be used to insert values
+into the network which are known when the signal function is created, without
+having to route those values from an input.
+
+\begin{figure}
+\begin{code}
+identity :: SF NonInitialized sv sv
+constant :: a -> SF NonInitialized SVEmpty (SVSignal a)
+never    :: SF NonInitialized SVEmpty (SVEvent a)
+asap     :: a -> SF NonInitialized SVEmpty (SVEvent a)
+after    :: Double -> a -> SF NonInitialized SVEmpty (SVEvent a)
+\end{code}
+\hrule
+\caption{Basic signal functions.}
+\label{figure:basic_signal_functions}
+\end{figure}
+
+\subsection{Lifting Pure Functions}
+\label{subsection:System_Design_and_Interface-Combinators-Lifting_Pure_Functions}
+
+Two combinators are provided to lift pure functions to signal functions (Figure~\ref{figure:lifting_pure_functions}).
+The {\tt pureSignal} combinator applies the pure function to a signal at every
+sample point. The {\tt pureEvent} combinator applies the function to each
+occurrence of an input event.
+
+\begin{figure}
+\begin{code}
+pureSignal ::    (a -> b) 
+              -> SF NonInitialized (SVSignal a) (SVSignal b)
+pureEvent  ::    (a -> b)
+              -> SF NonInitialized (SVSignal a) (SVSignal b)
+\end{code}
+\hrule
+\caption{Lifting pure functions.}
+\label{figure:lifting_pure_functions}
+\end{figure}
+
+\subsection{Routing}
+\label{subsection:System_Design_and_Interface-Combinators-Routing}
+
+\begin{figure}
+\begin{code}
+(>>>)       ::    SF NonInitialized svIn svBetween
+               -> SF NonInitialized svBetween svOut
+               -> SF NonInitialized svIn svOut
+first       ::    SF NonInitialized svIn svOut
+               -> SF NonInitialized (SVAppend svIn sv)
+                                    (SVAppend svOut sv)
+second      ::    SF NonInitialized svIn svOut
+               -> SF NonInitialized (SVAppend sv svIn)
+                                    (SVAppend sv svOut)
+swap        ::    SF NonInitialized (SVAppend svLeft svRight) 
+                                    (SVAppend svRight svLeft)
+\end{code}
+\hrule
+\caption{Routing combinators.}
+\label{figure:routing_combinators}
+\end{figure}
+
+\section{Evaluation}
+\label{section:System_Design_and_Interface-Evaluator}
+
+\subsection{Initialization and Running}
+
+\subsection{Input and Output}
+
+\subsection{Monad Lifting}

### 2012-07-31 Docs/IFL2012/paper.tex
Oops: additional source files for IFL2012 paper

\documentclass[draft]{llncs}

\usepackage{fancyvrb}                                       % For code environment
\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small} %

\title{Push-Pull Signal-Function Functional Reactive Programming}
\titlerunning{Push-Pull Signal-Function FRP}

\author{Edward Amsden}
\institute{Rochester Institute of Technology\\\email{eca7215@cs.rit.edu}}

\begin{document}
\maketitle

\begin{abstract}
Functional Reactive Programming is a promising class of systems for writing
interactive and time-dependent programs. Signal-function FRP is a subclass of
these systems which provides advantages of modularity and correctness, but
has proven difficult to efficiently implement.

The abstraction of signal vectors provides the necessary type apparatus to
distinguish components of the input and output of signal functions which benefit
from a push-based implementation from those which benefit from a pull-based
implementation, and to combine both implementation strategies in a single system.

We describe a signal-function FRP system which provides push-based evaluation
for events, pull-based evaluation for signals, and a simple monadic evaluation
interface which permits the system to be easily integrated with one or more
IO systems.
\end{abstract}

\section{Introduction}
\label{section:Introduction}

Functional Reactive Programming (FRP) is a class of systems for describing
reactive programs. Reactive programs are programs which, rather than taking
a single input and producing a single output, must accept multiple inputs and 
alter temporal behavior, including the production of multiple outputs, based
on these inputs.

An FRP system will provide a means of manipulating {\em behaviors} and
{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
but the definition is the same. A behavior or signal is, semantically, a
function from time to a value. An event is a discrete, possibly infinite, and 
time-ordered sequence of occurrences, which are times paired with values.

FRP systems can generally be categorized as ``classic FRP,'' 
which corresponds to the originally described FRP system in that behaviors
and events are manipulated directly and are first-class values in the FRP
system, or ``signal-function FRP,'' in which behaviors (generally termed
signals in this approach) and events are not first-class values, but signal
functions are first class values. Signal functions are time-dependent and
reactive transformers of signals, events, or combinations of signals and events.

FRP combines behaviors and events through the use of {\em switching}, in which
a behavior (in classic FRP) or a signal function (in signal-function FRP) is
replaced by a new behavior or signal function carried by an event occurrence.

Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
Recent work on classic FRP has focused on efficient implementation. Reactive is
a system for push-pull FRP~\cite{Elliott2009}. Push-based evaluation evaluates a
system only when input is available, and is thus suitable for discrete inputs
such as events. Pull-based evaluation evaluates the system as quickly as
possible, polling for input, and is preferable for behaviors and signals.
The initial implementations of FRP made use of pull-based evaluation for both
behaviors and events. Reactive, as well as more recent systems such as
``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
events and pull-based evaluation for behaviors.

All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
have used pull-based evaluation. This is due to the ease of implementation of
pull-based evaluation, and the types used for signal functions which do not
permit distinguishing signals and events, or constructing only part of the
input (for instance, one event occurrence.)

A recent extension of signal-function FRP called N-Ary FRP~\cite{Sculthorpe2011}
describes a method of typing signal functions which, as we will show, enables
the push-based evaluation of events in a signal-function FRP system. The notion
of signal vectors allows the representation of signal function inputs and
outputs as combinations of signals and events, rather than a single signal which
may contain multiple values, including option values for events. Signal vectors
are uninhabited types, which can be used to type partial or full representations
of the signal function inputs and outputs.

We present TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
favorite quotation of one of the author's philosophy instructors, used to
demonstrate the ambiguity of language. The origin of the quotation is unknown.}
a push-pull signal-function FRP system. We hope to demonstrate the feasibility
of such an approach to FRP, and provide a basis for further research into
efficient implementation of signal-function FRP. We also describe a powerful
evaluation interface for TimeFlies, which permits us to use TimeFlies to
describe applications which make use of multiple and differing IO libraries.

Section~\ref{section:System_Design} describes design choices for the system,
and provides an overview of the interface. Section~\ref{section:Implementation}
describes how the system is implemented, and how the separation of evaluation
between events and signals is achieved. Section~\ref{section:Discussion} is a
discussion of the usefulness of our implementation. 
Section~\ref{section:Ongoing_and_Further_Work} describes the current and future
work on this system. Section~\ref{section:Related_Work} gives an overview of
related efforts. Section~\ref{section:Conclusion} concludes.

\section{System Design}
\label{section:System_Design}

Our goal is to produce a composable and efficient FRP system. Signal function
FRP has an advantage in terms of composability, because it permits the
construction of self-contained objects through both input and output composition,
rather than purely through output composition as in classic FRP. Signal-function
FRP also avoids problematic properties common to classic FRP systems such as a
large class of time and space leaks~\cite{Liu2007}.

Efficient implementations of signal-function FRP have been approached through
runtime optimization~\cite{Nilsson2005}, but all implementations have been
pull-based for both signals and events. A truly efficient implementation will
likely combine run-time optimization with push-based evaluation for events.

The concept of N-Ary FRP was to encode additional safety properties into the
types of a signal-function FRP system~\cite{Sculthorpe2011}. This system
introduced the concept of {\em signal vectors} as input and output types for
signal functions. Signal vectors are combinations of signals and events. In
N-Ary FRP, signal functions are represented at the type level, and type-level
functions (type families in Haskell) are used to specify the representations
of signal functions.

In our system, we construct representations of signal vectors using
Generalized Algebraic Datatypes~\cite{Cheney2003,Xi2003}. The use of GADTs
enables us to use signal vectors to instantiate type parameters in the types
of signal vector representations, as well as in the types of signal functions.
GADTs are available as an extension in the Glasgow Haskell Compiler~\cite{PeytonJones2006}.

This approach permits us to construct partial inputs and outputs for signal
functions. For instance, we can construct a representation for a single event
in a signal vector, and another representation of updated values for a subset of
the signals in the signal vector, and yet another carrying values for all
signals in a signal vector. The signal vector types are shown in
Figure~\ref{figure:signal_vectors}.

\begin{figure}[t]
\begin{code}
data SVEmpty
data SVSignal a
data SVEvent a
data SVAppend svLeft svRight
\end{code}
\hrule
\caption{Signal vectors.}
\label{figure:signal_vectors}
\end{figure}

With the ability to construct partial representations of signal vectors, we can
represent signal functions with a datatype carrying multiple functions, one for
each type of input. This allows us to separate the pull-based processing of
signals from the push-based reaction to events.

The exposed interface is a set of combinators for constructing signal functions,
as well as combinators for describing the evaluation of a signal function. The
interface is shown in Fig.~\ref{figure:signal_function_interface}.

\begin{figure}
\begin{code}

-- Signal Functions
type :~> svIn svOut

-- Infix type alias for SVAppend
type :^: svLeft svRight = SVAppend svLeft svRight

-- Basic signal functions
identity :: sv :~> sv
constant :: a -> SVEmpty :~> SVSignal a
never    :: SVEmpty :~> SVEvent a
asap     :: a -> SVEmpty :~> SVEvent a
after    :: Double -> a -> SVEmpty :~> SVEvent a

-- Lifting pure functions
pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal a
pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent a

-- Composition and routing
(>>>)           :: (svIn :~> svMiddle) -> (svMiddle :~> svOut) 
                   -> (svIn :~> svOut)
first           :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
second          :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut) 
swap            :: (svLeft :^: svRight) :~> (svRight :^: svLeft)
copy            :: sv :~> (sv :^: sv)
ignore          :: sv :~> SVEmpty
cancelLeft      :: (SVEmpty :^: sv) :~> sv
cancelRight     :: (sv :^: SVEmpty) :~> sv
uncancelLeft    :: sv :~> (SVEmpty :^: sv)
uncancelRight   :: sv :~> (sv :^: SVEmpty)
associate       :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3)) 
unassociate     :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)

-- Reactivity
switch :: (svIn :~> (svOut :^: SVEvent (svIn :~> svOut))) 
          -> svIn :~> svOut

-- Feedback
loop :: ((svIn :^: svLoop) :~> (svOut :^: svLoop)) -> svIn :~> svOut

-- Time dependence
time  :: SVEmpty :~> Double
delay :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
class TimeIntegrate

-- Joining
union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
combineSignals :: (a -> b -> c) 
                  -> (SVSignal a :^: SVSignal b) :~> SVSignal c
capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent a

-- Events
filter         :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
\end{code}
\hrule
\caption{Signal function interface.}
\label{figure:signal_function_interface}
\end{figure}

Signal functions are produced by combining primitive signal functions using
the {\tt >>>} (sequential composition), {\tt first}, and {\tt second} routing
combinators. Other routing signal functions are provided, but they are intended
to be combined with, rather than to modify, other signal functions.

\subsection{Examples of Signal Function Primitives}
\label{subsection:Examples_of_Signal_Function_Primitives}
The most basic signal is the {\tt identity} signal function, which, as its
name suggests, simply passes its input to its output.

The {\tt constant} signal function has an empty input, and produces a constant
signal as its output.

The {\tt asap} signal function produces an event at the first time interval
after it begins, and never again.

The {\tt filter} signal function applies a predicate to the values of event
occurrences and passes them along only if the predicate is true of the value.
The predicate outputs {\tt Maybe} values to allow the user to avoid partial
functions on types such as {\tt Maybe} or {\tt Either}.

The {\tt associate} signal function is a routing function, which takes as input
a combination of three signal vectors to which the signal vector append
constructor is combined left-associatively, and produce the same signal vector
as output, but with the signal vector append constructor combined
right-associatively.

The {\tt switch} function is the essential signal function for reactivity. It
acts as the supplied signal function until that signal function produces an
event on its right output. This event's value is a new signal function, which
replaces the switch signal function.

The {\tt loop} function permits a signal function to receive its own output
as an input.

Implementations for these examples will be discussed in Sec.~\ref{subsection:Signal_Function_Implementations}.

\subsection{Evaluation Interface}
The evaluation interface provides a monad~\cite{PeytonJones1993,PeytonJones2001}
for specifying input to signal functions and handling their output. This
interface also allows input actions to be separated from output actions, and
from each other, so that a signal function may receive input from multiple IO
systems and have its output handled by multiple output systems.

The evaluation interfaceis a monad transformer~\cite{Liang1995}. This allows
a signal function evaluation to be constructed from primitive evaluation actions
(event pushing, input signal updating, and sampling) as well as actions from the
underlying monad. These actions can then be run in the underlying monad
to actuate the signal function. Because the actuation takes a signal function's
state as input and produces a new state as output, in addition to the monadic
side effects of handling the signal function's output, different evaluation
actions may be taken on the same signal function from distinct locations within
an application's code. This allows the evaluation interface to be easily
integrated with event-loop style systems as well as traditional imperative
IO systems.

The evaluation interface consists of the monad transformer type, functions for
constructing event inputs, initial input signal samples, and input signal
updates, functions for constructing the vector of output handlers for a signal
function, and evaluation actions. The full interface is shown in Fig.~\ref{figure:evaluation_interface}.

\begin{figure}
\begin{code}
-- Input helpers
class SVRoutable -- Instances: SVSignalUpdate,
                 -- SVEventOccurrence
svLeft :: (SVRoutable r) => 
          r svLeft -> r (SVAppend svLeft svRight)
svRight :: (SVRoutable r) =>
           r svRight -> r (SVAppend svLeft svRight)

-- Signal inputs
data SVSignalUpdate
data SVSample
sample :: a -> SVSample (SVSignal a)
sampleEvt :: SVSample (SVEvent a)
sampleNothing :: SVSample SVEmpty
combineSamples :: SVSample svLeft -> svSample svRight 
                  -> svSample (svLeft :^: svRight)
svSig :: a -> SVSignalUpdate (SVSignal a)

-- Event inputs
data SVEventInput
svOcc :: a -> SVEventInput (SVEvent a)
-- Actuation
update :: (Monad m) => SVSignalUpdate svIn 
           -> SFEvalT svIn svOut m ()
push   :: (Monad m) => SVEventInput svIn
          -> SFEvalT svIn svOut m ()
sample :: SVEvalT svIn svOut m ()
-- Running
data SVHandler
emptyHandler :: SVHandler m SVEmpty
eventHandler :: (a -> m ()) 
                -> SVHandler m (SVEvent a)
signalHandler :: (a -> m ()) 
                 -> SVHandler m (SVSignal a)
combineHandlers :: SVHandler m svLeft
                   -> SVHandler m svRight
                   -> SVHandler m (svLeft :^: svRight)
initSFEvalT :: SVHandler m svOut -> SVSample svIn 
               -> Double ->(svIn :~> svOut)
               -> SFEvalState m svIn svOut
runSFEvalT :: SFEvalT svIn svOut m a
              -> SFEvalState m svIn svOut
              -> m a
\end{code}
\hrule
\caption{Evaluation interface.}
\label{figure:evaluation_interface}
\end{figure}

\section{Implementation}
\label{section:Implementation}

We now turn our attention to the implementation of the signal function
system. We will discuss representations of inputs, outputs, and signal functions,
as well as the implementations of specific signal function combinators.

\subsection{Input and Output Representations}
\label{subsection:Input_and_Output_Representations}
As discussed in Section~\ref{section:System_Design}, our implementation requires
representations of the inputs and outputs of signal functions. Since signal
functions are typed using signal vectors, the representation types must be
parameterized over signal vectors. But signal vectors are uninhabited types, and
in ordinary Haskell we cannot declare a data constructor which uses a component
of a type parameter. Further, we cannot restrict which types may instantiate a
type parameter for a particular data constructor.

GADTs lift these restrictions, permitting type constraints to be applied
to individual data constructors. In the declaration of a GADT, a type signature
is provided for each data constructor, including the types of the constructor's
parameters. When a GADT is pattern-matched, the types of the parameters are
inferred using the constraints given in this signature. This is called type
refinement.

Using GADTs, we can construct representations of signal vectors to use in our
implementation. The first such representation carries a value for each signal
component of a signal vector, and is shown in Figure~\ref{figure:signal_sample_representation}.

\begin{figure}[t]
\begin{code}
data SVSample sv where
  SVSample    :: a -> SVSample (SVSignal a)
  SVSampleEvt :: SVSample (SVEvent a)
  SVNothing   :: SVSample SVEmpty
  SVBoth      :: SVSample svLeft -> SVSample svRight
                 -> SVSample (svLeft :^: svRight)

sample         :: a -> SVSample a
sampleEvt      :: SVSample (SVEvent a)
sampleNothing  :: SVSample SVEmpty
combineSamples :: SVSample svLeft -> SVSample svRight
                  -> SVSample (svLeft :^: svRight)
splitSample    :: SVSample (svLeft :^: svRight)
                  -> (SVSample svLeft, svSample svRight)
sampleValue    :: SVSample (SVSignal a) -> a
\end{code}
\hrule
\caption{Signal sample representation.}
\label{figure:signal_sample_representation}
\end{figure}

A value of this type represents the entire vector, though it has nullary
constructors at event leaves and empty leaves. For event occurrences, we want
to represent only a single point in the vector, as shown in Figure~\ref{figure:event_occurrence_representation}.
This representation uses two constructors, each of which leaves one or the
other of the type variables in the signal vector {\tt SVAppend} constructor
unconstrained.

\begin{figure}[t]
\begin{code}
data SVOccurrence sv where
  SVOccurrence :: a -> SVOccurrence (SVEvent a)
  SVOccLeft    :: SVOccurrence svLeft
                  -> SVOccurrence (svLeft :^: svRight)
  SVOccRight   :: SVOccurrence svRight
                  -> SVOccurrence (svLeft :^: svRight)

occurrence        :: a -> SVOccurrence (SVEvent a)
occLeft           :: SVOccurrence svLeft
                     -> SVOccurrence (svLeft :^: svRight)
occRight          :: SVOccurrence svRight
                     -> SVOccurrence (svLeft :^: svRight)
chooseOccurrences :: SVOccurrence (svLeft :^: svRight)
                     -> Either (SVOccurrence svLeft) (SVOccurrence svRight)
fromOccurrence    :: SVOccurrence (SVEvent a) -> a
\end{code}
\hrule
\caption{Event occurrence representation.}
\label{figure:event_occurrence_representation}
\end{figure}

The final representation ``signal deltas,'' carries replacement values for some
points in a signal vector. Here, we combine the unconstrained type variable
from the occurrence representation with the ``Both'' constructor from the sample
representation. The unconstrained type variable will be placed in a separate
constructor, which allows the construction of empty signal deltas, as is shown
in Figure~\ref{figure:signal_delta_representation}.

\begin{figure}[t]
\begin{code}
data SVDelta sv where
  SVDelta        :: a -> SVDelta (SVSignal a)
  SVDeltaNothing :: SVDelta sv
  SVDeltaBoth    :: SVDelta svLeft -> SVDelta svRight
                    -> SVDelta (svLeft :^: svRight)
delta         :: a -> SVDelta (SVSignal a)
deltaNothing  :: SVDelta sv
combineDeltas :: SVDelta svLeft -> SVDelta svRight
                 -> SVDelta (svLeft :^: svRight)
updateSample  :: SVDelta sv -> SVSample sv -> SVSample sv
\end{code}
\hrule
\caption{Signal delta representation.}
\label{figure:signal_delta_representation}
\end{figure}

\subsection{Signal Function Representations}
\label{subsection:Signal_Function_Representations}
With representations for signal function inputs and outputs in place, we turn
to the representation of signal functions. A signal function must be able to
respond to time increments and new signal samples, as well as event occurrences.
Since we wish to be able to ``push'' event occurrences independently of
sampling, events are handled using the last-updated time and signal sample.

When a signal function is asked to respond to either of these types of inputs,
it must produce an output consisting of zero or more event occurrences, a new
signal function with the same type (to enable reactivity and state), and, if
it is responding to a time and sample update, a delta which represents updates
to the sample of its output signals.

Finally, there is a special case for a signal function at time 0. It must be
provided with an initial input sample before it can respond to incremental time
and sample updates or to event occurrences.

Signal functions are represented as a datatype with two constructors. The first
constructor wraps a function from an input sample to an output sample and an
initialized signal function. The second wraps two functions. The first accepts
a time delta and a signal delta, and produces a signal delta, a list of event
occurrences, and a new signal function as the output. The second accepts an
event occurrence and produces a list of event occurrences and a new signal
function as the output. The signal function representation is shown in Figure~\ref{figure:signal_function_representation}.

\begin{figure}[t]
\begin{code}
data Initialized
data NonInitialized

type :~> svIn svOut = SF NonInitialized svIn svOut

data SF init svIn svOut where
  SF :: (SVSample svIn -> (SVSample svOut, SF Initialized svIn svOut))
        -> SF NonInitialized svIn svOut
  SFInit :: (Double -> SVDelta svIn 
             -> (SVDelta svOut, [SVOccurrence svOut],
                 SF Initialized svIn svOut))
            -> (SVOccurrence svIn -> ([SVOccurrence svOut],
                                      SF Initialized svIn svOut))
            -> SF Initialized svIn svOut
\end{code}
\hrule
\caption{Signal function representation}
\label{figure:signal_function_representation}
\end{figure}

This representation allows a signal function to respond to events and time
updates separately, and does not enforce the representation of events during
time updates, as previous signal function systems do.

\subsection{Signal Function Implementations}
\label{subsection:Signal_Function_Implementations}

Now that we have established representations for signal functions and their
inputs and outputs, we show several examples of how signal function combinators
are implemented.

The {\tt identity} signal function's initialization function takes a sample,
and returns that sample as its output sample. The initialized function returned
is the {\tt identityInit} function, which is not exported by the module.
The time and sample function ignores the provided time delta, produces the
input signal delta as the output signal delta, and returns {\tt identityInit} as
the new signal function. The event function takes an event occurrence and
returns a singleton list containing that occurrence, along with the
{\tt identityInit} function as the replacement signal function. The
implementation is shown in Figure~\ref{figure:identity_signal_function_implementation}.

\begin{figure}[t]
\begin{code}
identity :: sv :~> sv
identity = SF (\sample -> (sample, identityInit))

identityInit :: SF Initialized sv sv
identityInit = SFInit (\_ delta -> (delta, [], identityInit))
                      (\occ -> ([occ], identityInit))
\end{code}
\hrule
\caption{{\tt identity} signal function implementation.}
\label{figure:identity_signal_function_implementation}
\end{figure}

The {\tt constant} signal function is initialized with a value, which it then
produces as its output forever. Thus, the initialized version of {\tt constant}
never produces any output other than itself as a replacement
signal function, and an empty signal delta. The implementation of {\tt constant}
is shown in Fig.~\ref{figure:constant_signal_function_implementation}.

\begin{figure}
\begin{code}
constant :: a -> SVEmpty :~> SVSignal a
constant x = SF (\_ -> (constantInit, sample x))

constantInit :: SF Initialized SVEmpty (SVSignal a)
constantInit = SFInit (\_ _ -> (deltaNothing, [], constantInit))
                      (\_ -> ([], constantInit))
\end{code}
\hrule
\caption{{\tt constant} signal function implementation.}
\label{figure:constant_signal_function_implementation}
\end{figure}

The {\tt asap} signal function implementation requires a parameter for the
initialized version of the signal function, to specify the value of the event
occurrence. It replaces itself with the initialized version of the {\tt never}
signal function after the first time step. The implementation is shown in
Fig.~\ref{figure:asap_signal_function_implementation}.

\begin{figure}
\begin{code}
asap :: a -> SVEmpty :~> SVEvent a
asap x = SF (\_ -> (sampleEvt, asapInit x))

asapInit :: a -> SF Initialized SVEmpty (SVEvent a)
asapInit x = SFInit (\_ _ -> (deltaNothing, [occurrence x], neverInit))
                    (\_ -> (never))
\end{code}
\hrule
\caption{{\tt asap} signal function implementation.}
\label{figure:asap_signal_function_implementation}
\end{figure}

The {\tt filter} function accepts events, applies a {\tt Maybe} predicate to
their values, and produces the value produced by the predicate as an event
occurrence, or no occurrence if the value is {\tt Nothing}. It is implemented
by having the event occurrence response function apply the {\tt maybe} function
from the Haskell prelude to the value returned by the predicate, as shown in 
Fig.~\ref{figure:filter_signal_function_implementation}

\begin{figure}[t]
\begin{code}
filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
filter p = SF (\_ -> (sampleEvt, filterInit p))

filterInit :: (a -> Maybe b) -> SF Initialized (SVEvent a) (SVEvent b)
filterInit p = SFInit (\_ _ -> (deltaNothing, [], filterInit p))
                      (\evtOcc -> (maybe [] ((:[]) . occurrence) $
                                   fromOccurrence evtOcc,
                                   filterInit p))
\end{code}
\hrule
\caption{{\tt filter} signal function implementation.}
\label{figure:filter_signal_function_implementation}
\end{figure}

The {\tt associate} signal function is a routing signal function. It transforms
a signal vector which is left-associated at the top level to one that is
right-associated at the top level. Its implementation is representative of all
of the routing functions, and is shown in Fig.~\ref{figure:associate_signal_function_implementation}.

\begin{figure}[t]
\begin{code}
associate :: SVAppend (SVAppend sv1 sv2) sv3
             :~> SVAppend sv1 (SVAppend sv2 sv3)
associate = 
  SF (\sigSample -> let (sigSampleLeft, sigSampleRight) = 
                          splitSample sigSample
                        (sigSampleLeftLeft, sigSampleLeftRight) =
                          splitSample sigSampleLeft
                    in (combineSamples sigSampleLeftLeft 
                        (combineSamples sigSampleLeftRight sigSampleRight),
                        associateInit))

associateInit :: SF Initialized (SVAppend (SVAppend sv1 sv2) sv3)
                                (SVAppend sv1 (SVAppend sv2 sv3))
associateInit = SFInit (\_ sigDelta -> let (sigDeltaLeft, sigDeltaRight) =
                                             splitDelta sigDelta
                                           (sigDeltaLeftLeft,
                                            sigDeltaLeftRight) =
                                             splitDelta sigDeltaLeft
                                       in (combineDeltas sigDeltaLeftLeft
                                           (combineDeltas sigDeltaLeftRight
                                                          sigDeltaRight),
                                           [], associateInit)) 
                       (\evtOcc -> (case chooseOccurrence evtOcc of
                                      Left leftOcc -> 
                                        case chooseOccurrence leftOcc of
                                          Left leftLeftOcc ->
                                            [occLeft leftLeftOcc]
                                          Right leftRightOcc ->
                                            [occRight $ occLeft leftRightOcc]
                                      Right rightOcc ->
                                        [occRight $ occRight $ rightOcc],
                                    associateInit))
\end{code}
\hrule
\caption{{\tt associate} signal function implementation.}
\label{figure:associate_signal_function_implementation}
\end{figure}

The {\tt switch} function is the basic combinator used to introduce reactivity.
It is given a signal function whose output is the append of two signal vectors.
The left side of the signal vector is passed on as the output of the reactive
signal function, and the right side is an event carrying signal functions. When
an event occurrence is present on the right-side event output, the signal
function carried by this occurrence replaces the signal function constructed
by switch. Due to space concerns, the implementation is elided, but a brief
description will suffice. The signal function stores the input sample
provided during initialization, and updates it with deltas. When an event
occurrence carrying a signal function is produced by either the event or 
time function of the wrapped signal function, the stored signal sample is used
to initialize the new signal function. If this occurs while handling an event
input, the sample output by the new signal function's initialization is stored
and is combined with its first output delta to produce the output delta at the
next time step.

The {\tt loop} function allows a signal function to see a component of its own
output as input. This is primarily useful when a signal function has components
which mutually depend on each others outputs, such as in physics simulations
or games. Care must be taken that the feedback output is not immediately
dependent on the feedback input, or sampling the signal function will not
terminate. Loop is implemented by generating a recursive list (which will
be infinite if the feedback is not decoupled) of event inputs, and by splitting
the output signal delta and using the right side output delta as the right
side input delta within a recursive let-binding. (The let-binding in Haskell
always admits recursive bindings.) If the system is not completely decoupled,
this will result in non-termination during an evaluation step. Decoupling can
be achieved using {\tt delay} primitive.

\subsection{Evaluation Interface Implementation}
\label{subsection:Evaluation_Interface_Implementation}

The evaluation interface must maintain state including the current signal
function, the current time (to produce time deltas), updates to the input
sample which have yet to be sampled, and the output handlers.

The output handlers are stored in a structure similar to that for a signal
sample. The difference is that both signal and event leaves contain values, and
these values are functions from the leaf type to another type. The handler
datatype is shown in Fig.~\ref{figure:handler_datatype}.

\begin{figure}[t]
\begin{code}
data SVHandler out sv where
  SVHandlerEmpty  :: SVHandler out SVEmpty
  SVHandlerSignal :: (a -> out) -> SVHandler out (SVSignal a)
  SVHandlerEvent  :: (a -> out) -> SVHandler out (SVEvent a)
  SVHandlerBoth   :: SVHandler out svLeft -> SVHandler out svRight 
                     -> SVHandler out (SVAppend svLeft svRight)
\end{code}
\hrule
\caption{Handler datatype.}
\label{figure:handler_datatype}
\end{figure}

We now need a datatype to hold the various components of the evaluation state.
This type is shown in Fig~\ref{figure:evaluation_state_datatype}.

\begin{figure}[t]
\begin{code}
data SFEvalState m svIn svOut 
  = SFEvalState { 
                  esSF :: SF Initialized svIn svOut,
                  esOutputHandlers :: SVHandler (m ()) svOut,
                  esLastTime :: Double,
                  esDelta :: SVDelta svIn
                }

\end{code}
\hrule
\caption{Evaluation state datatype.}
\label{figure:evaluation_state_datatype}
\end{figure}

The evaluation interface itself is a monad transformer, which we implement
as a Haskell newtype wrapping the {\tt StateT} monad transformer. We do not
export the raw {\tt put} and {\tt get} actions of the state monad, but instead
implement the push, update, and sampling operations for signal function
evaluation using {\tt put} and {\tt get}. The evaluation monad transformer is
shown in Fig.~\ref{figure:signal_function_evaluation_monad_transformer}.
Instances of typeclasses including {\tt Monad} and {\tt MonadTrans} are derived
using the ``GeneralizedNewtypeDeriving'' extension to the Glasgow Haskell
Compiler.

\begin{figure}[t]
\begin{code}
newtype SFEvalT svIn svOut m a = StateT (SFEvalState m svIn svOut) m a
\end{code}
\hrule
\caption{Signal function evaluation monad transformer.}
\label{figure:signal_function_evaluation_monad_transformer}
\end{figure}

\section{Discussion}
\label{section:Discussion}

The system presented here, TimeFlies, demonstrates how using signal vectors
to type inputs and outputs enables push-based evaluation of events in a
signal-function system. We take advantage of this representation in several ways.

First, by separating components of inputs and outputs in the types, we are free
to create distinct, and often partial, representations of the input or output
of a signal function. This enables us to represent only the event occurrence
being pushed at that time.

Second, this separation also permits us to separate the process of gathering
the input to a signal function, and the process of handling its output, into
different points in a program. Using the evaluation interface described, an
event occurrence may be pushed onto one input of a signal function from one
point in a program (e.g. a mouse click handler), an input signal
may be updated in another (e.g. a mouse movement handler), and finally the
system may be sampled in a third place (e.g. an animation or audio timed
callback).

Finally, this approach enables further work on the implementation of the signal
function system to be separated from changes in the interface. By enabling
differing representations of the inputs and outputs of signal functions, we are
free to change these representations without the need to further constrain the
input and output types.

\section{Ongoing and Further Work}
\label{section:Ongoing_and_Further_Work}

TimeFlies, the system described here, has been implemented, but not
extensively tested. The immediate goal is to create a real-time application
which will permit a performance and implementation comparison of TimeFlies with
Yampa, the current state-of-the-art pull-based signal-function system.

In the future, we hope to apply run-time optimizations, using the technique used
for Yampa, to create a push-pull self-optimizing signal-function system. Further,
we hope to use this system as a basis for exploring signal-function FRP as a
basis for general-purpose application frameworks.

\section{Related Work}
\label{section:Related_Work}

Signal Function FRP was introduced as a model for Graphical User Interfaces~\cite{Courtney2001-1}.
The system was originally termed ``AFRP'' (Arrowized FRP). Yampa is a rewrite of
AFRP where signal functions apply a number of ad-hoc optimizatons to themselves
as they evolve. Yampa demonstrated a modest performance improvement
over AFRP~\cite{Nilsson2005}.

Reactive is a classic FRP system which implements push-based evaluation for events
by transforming behaviors to ``reactive~normal~form,'' where a behavior
is a non-reactive behavior running inside a switch, whose event stream carries
behaviors in reactive normal form. The system is evaluated by forking a Haskell
thread to repeatedly sample the non-reactive behavior, and then blocking on the
evaluation of the first occurrence in the event stream. When this occurrence
is yielded, the evaluation thread for the behavior is killed and a new
thread forked to evaluate the new behavior~\cite{Elliott2009}.

\section{Conclusion}
\label{section:Conclusion}

We have presented TimeFlies, a system for push-pull signal-function Functional
Reactive Programming, and have shown how the use of a signal vectors as input
and output types for signal functions, together with GADT-based representations
of the inputs and outputs, permits the implementation of a push-pull system.

We have also described a general and flexible monadic evaluation interface for
TimeFlies, which permits us to interface the TimeFlies system with different
styles of IO systems, including multiple IO systems in the same application.

This opens up the exciting possibility that a signal-function FRP could become
an efficient and general framework for writing interactive applications.

\bibliographystyle{splncs}
\bibliography{thesis}

\end{document}

### 2012-08-01 Code/time-flies/src/FRP/TimeFlies/SignalFunctions.hs
Filter SF

@@ -62,6 +62,8 @@ module FRP.TimeFlies.SignalFunctions (
   delay,
   TimeIntegrate,
   integrate,
+  -- * Events
+  filter,
   -- * Joining
   union,
   combineSignals,
@@ -93,6 +95,8 @@ import Data.Maybe
 import qualified Data.Tuple as T
 
 import FRP.TimeFlies.SignalVectors
+
+import Prelude hiding (filter)
 
 -- | Signal function running or suspended
 data Initialized
@@ -342,6 +346,14 @@ unassociateInit = SFInit (\_ sigDelta -> let (sigDeltaLeft, sigDeltaRight) = spl
                                                             Left rightLeftOcc -> [occLeft $ occRight rightLeftOcc]
                                                             Right rightRightOcc -> [occRight rightRightOcc],
                                       unassociateInit))
+
+-- | Filter event occurrences
+filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
+filter f = SF (\_ -> (sampleEvt, filterInit f))
+
+filterInit :: (a -> Maybe b) -> SF Initialized (SVEvent a) (SVEvent b)
+filterInit f = SFInit (\_ _ -> (deltaNothing, [], filterInit f))
+                      (\evtOcc -> (maybe [] ((:[]) . occurrence) $ f $ fromOccurrence evtOcc, filterInit f))
 
 -- | Provide the input as input to the given signal function, producing
 -- the left side of the given signal function's output as output. Upon

### 2012-08-01 Code/time-flies/src/FRP/TimeFlies/SignalFunctions.hs
Use infix type alias for SF NonInitialized

@@ -131,14 +131,14 @@ applySF sf indices = foldr (\evtOcc (changes, SFInit _ changeCont) -> let (newCh
                      indices
 
 -- | Identity signal function: reproduce the input exactly as the output.
-identity :: SF NonInitialized sv sv
+identity :: sv :~> sv
 identity = SF (\initSample -> (initSample, identityInit))
 
 identityInit :: SF Initialized sv sv
 identityInit = SFInit (\dt sigDelta -> (sigDelta, [], identityInit)) (\evtOcc -> ([evtOcc], identityInit))
 
 -- | Constant signal function: Produce the signal
-constant :: a -> SF NonInitialized SVEmpty (SVSignal a)
+constant :: a -> SVEmpty :~> SVSignal a
 constant x = SF (\_ -> (sample x, constantInit))
 
 constantInit :: SF Initialized SVEmpty (SVSignal a)
@@ -146,7 +146,7 @@ constantInit = SFInit (\_ _ -> (deltaNothing, [], constantInit))
                       (\_ -> ([], constantInit))
 
 -- | Produce an event at the first time step after being switched in
-asap :: a -> SF NonInitialized SVEmpty (SVEvent a)
+asap :: a -> SVEmpty :~> SVEvent a
 asap x = SF (\_ -> (sampleEvt, asapInit x))
 
 asapInit :: a -> SF Initialized SVEmpty (SVEvent a)
@@ -154,7 +154,7 @@ asapInit x = SFInit (\_ _ -> (deltaNothing, [occurrence x], neverInit))
                     (\_ -> ([], asapInit x))
 
 -- | Never produce an event
-never :: SF NonInitialized SVEmpty (SVEvent a)
+never :: SVEmpty :~> SVEvent a
 never = SF (\_ -> (sampleEvt, neverInit))
 
 neverInit :: SF Initialized SVEmpty (SVEvent a)
@@ -162,7 +162,7 @@ neverInit = SFInit (\_ _ -> (deltaNothing, [], neverInit))
                    (\_ -> ([], neverInit))
 
 -- | Produce an event after some time
-after :: Double -> a -> SF NonInitialized SVEmpty (SVEvent a)
+after :: Double -> a -> SVEmpty :~> SVEvent a
 after dt x = SF (\_ -> (sampleEvt, afterInit dt x))
 
 afterInit :: Double -> a -> SF Initialized SVEmpty (SVEvent a)
[…]

### 2012-08-27 Docs/IFL2012/presentation/presentation.tex
Preliminary presentation source

\documentclass{beamer}
\usetheme{JuanLesPins}
\usecolortheme{seahorse}

\usepackage{fancyvrb}
\usepackage{graphicx}
\usepackage[style=authortitle-icomp]{biblatex}

%gets rid of bottom navigation bars
\setbeamertemplate{footline}[]{}

%gets rid of navigation symbols
\setbeamertemplate{navigation symbols}{}

\bibliography{thesis}

\title{Push-Pull Signal-Function Functional Reactive Programming}
\author{Edward Amsden}
\institute{Rochester Institute of Technology}

\begin{document}

\AtBeginSection[]
{
\begin{frame}
\frametitle{\insertsection}
\end{frame}
}

\begin{frame}[plain]
\titlepage
\includegraphics[width=9em]{RIT_GCCIS1}
\end{frame}

\section{Introduction}

\begin{frame}
\frametitle{Functional Reactive Programming (FRP)}
    \begin{itemize}
        \item {\em Functional} First-class and higher-order functions.
        \item {\em Reactive} Behavior changes in response to temporal inputs.
        \item Basic abstractions
        \begin{itemize}
            \item Signals - Functions of time\footnote{Often also called behaviors.}.
            \item Events - Sequences of temporally ordered and labeled discrete values.
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Signal-Function FRP}
    \begin{itemize}
        \item Signal functions are transformers of events and signals.
        \item Signal functions are first class in Signal-Function FRP.
        \item Signals and events are not first-class in Signal Function FRP.
        \item This approach is more composable than first-class signals and events (since input may be transformed)\footcite{Courtney2001-1}.
    \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Push vs. Pull Evaluation}
    \begin{itemize}
        \item When to evaluate what?
        \item Pull evaluation: (``demand-driven'') Evaluate when output is needed.
        \item Push evaluation: (``data-driven'') Evaluate when input is available.
        \item Ideally, FRP uses both: pull for signals and push for events.
    \end{itemize}
\end{frame}

\section{Problem}

\begin{frame}[fragile]
\frametitle{Separating Events and Signals}
    \begin{itemize}
        \item[] 
\begin{Verbatim}
class (Category a) => Arrow a where
  arr :: (b -> c) -> a b c
  ...
\end{Verbatim}
        \item Traditional Signal-Function FRP encodes signal functions as a Haskell Arrow.
\begin{Verbatim}
data SF a b = ...

instance Arrow SF where
  ...
\end{Verbatim}
    \end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Separating Events and Signals}
    \begin{itemize}
        \item Necessary to lift any function to a corresponding signal function, without input/output type annotation
        \item Events encoded as optional signals:
\begin{Verbatim}
type Event a = Maybe a
\end{Verbatim}
        \item Combined signals and events encoded as one signal with pair type:
    \end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Separating Events and Signals}
    \begin{itemize}
        \item[]
\begin{Verbatim}
class (Category a) => Arrow a where
  ...
  first :: a b c -> a (b, d) (c, d)
  ...

capture :: SF (a, Event b) (Event a)

ghci> :t first . arr
first . arr
    :: (Arrow a) => (b -> c) -> a (b, d) (c, d)
\end{Verbatim}
    \end{itemize}
\end{frame}

\section{Approach}

\begin{frame}[fragile]
\frametitle{Type Signal Functions with Signal Vectors}

    \begin{itemize}
        \item Described by Sculthorpe\footcite{Sculthorpe2011}
        \item Describe separation of individual signals and events
        \item In Haskell (with {\tt -XEmptyDataDecls}):
\begin{Verbatim}
data SVEmpty
data SVSignal a
data SVEvent a
data SVAppend svl svr
type :^: svl svr = SVAppend svl svr
\end{Verbatim}
        \item Would like to use {\tt -XDataKinds} but it's not working well yet.
    \end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Type Signal Functions with Signal Vectors}

    \begin{itemize}
        \item Two combinator examples:
        \item Lifting a pure function to transform a signal:
\begin{Verbatim}
pureSignal ::     (a -> b)
               -> SF (SVSignal a) (SVSignal b)
\end{Verbatim}
        \item Passing through input on the right:
\begin{Verbatim}
first ::    SF svIn svOut
         -> SF (svIn :^: svRight) (svOut :^: svRight)
\end{Verbatim}
        \item Composing these leads to:
\begin{Verbatim}
ghci> :t first . pureSignal
first . pureSignal
    :: (a -> b) -> SF (SVSignal a :^: sv)
                      (SVSignal b :^: sv)
\end{Verbatim}
    \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Partial Representations of Signal Vectors}
    \begin{itemize}
        \item To evaluate signals and events differently we must represent them separately.
        \item Signal vectors enable this by distinguishing them in the types.
        \item We can construct several representations of a signal vector.
        \begin{itemize}
          \item Represent signal leaves, event leaves, or both.
          \item Represent one leaf, a subset of leaves, or all applicable leaves.
          \item Transform the type at the leaf, or don't.
        \end{itemize}
    \end{itemize}
\end{frame}

\section{Implementation}

\begin{frame}[fragile]
\frametitle{Signal Representation}
    \begin{itemize} 
        \item Represent an entire signal sample\\ (for initializing a signal function at time zero):
\begin{Verbatim}
data SVSample where
    SVSample      ::    a -> SVSample (SVSignal a)
    SVSampleEvent ::    SVSample (SVEvent a)
    SVSampleEmpty ::    SVSample SVEmpty
    SVSampleBoth  ::    SVSample svLeft 
                     -> SVSample svRight
                     -> SVSample (svLeft :^: svRight) 
\end{Verbatim}
    \end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Signal Representation}
    \begin{itemize} 
        \item Represent a signal delta\\ (replacement values for a subset of signals):
\begin{Verbatim}
data SVDelta where
     SVDelta        ::   a -> SVDelta (SVSignal a)
     SVDeltaNothing ::   SVDelta sv
     SVDeltaBoth    ::   SVDelta svLeft 
                      -> SVDelta svRight
                      -> SVDelta (svLeft :^: svRight) 
\end{Verbatim}
    \end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Event Representation}
    \begin{itemize} 
        \item Represent an event occurrence:
\begin{Verbatim}
data SVOccurrence where
     SVOccurrence ::   a -> SVOccurrence (SVEvent a)
     SVOccLeft    ::   SVOccurrence svLeft
                    -> SVOccurrence
                         (svLeft :^: svRight)
     SVOccRight   ::   SVOccurrence svRight
                    -> SVOccurrence
                         (svLeft :^: svRight)
\end{Verbatim}
    \end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Signal Function Representation}
    \begin{itemize}
        \item Separate continuations for time advancement and event occurrences:
\begin{Verbatim}
data Initialized
data NonInitialized

data SF init svIn svOut where
  SF ::   (SVSample svIn
           -> (SVSample svOut,
               SF Initialized svIn svOut))
       -> SF NonInitialized svIn svOut
...
\end{Verbatim}
    \end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Signal Function Representation}
    \begin{itemize}
    \item Separate continuations for time advancement and event occurrences.
\begin{Verbatim}
...
  SFInit ::   (Double 
               -> SVDelta svIn
               -> (SVDelta svOut,
                   [SVOccurrence svOut],
                   SF Initialized svIn svOut))
           -> (SVOccurrence svIn
               -> ([SVOccurrence svOut],
                   SF Initialized svIn svOut)
           -> SF Initialized svIn svOut
\end{Verbatim}
    \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Evaluation}
    \begin{itemize}
        \item Yampa/AFRP: Supply SF and input/output actions to an evaluation loop ({\tt reactimate}).
        \item Here: initialize an evaluation state with:
        \begin{itemize}
            \item A signal function.
            \item Initial values for all input signals.
            \item Handlers for all outputs.
        \end{itemize}
        \item Then, the evaluation monad carries this state and provides the actions:
            \begin{itemize}
                \item {\tt push} Push an event (which will be immediately reacted to).
                \item {\tt update} Update the value of an input signal (with no immediate effect).
                \item {\tt step} Update the time and evaluate new values of signals.
            \end{itemize}
    \end{itemize}
\end{frame}

\section{Conclusion}

% Add conclusion

\begin{frame}
\frametitle{Further Work}
\begin{itemize}
  \item Dynamic collections: Dynamically switch between collections of signal functions.
  \item Semantics/correctness proof (especially for event merging).
  \item Time-independence optimization.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Questions?}
\end{frame}

\section{In Detail}

\begin{frame}
\frametitle{Combinator Implementations: Serial Composition ({\tt >>>})}
    \begin{itemize}
        \item Initialization:
        \item Running:
    \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Combinator Implementations: Parallel Composition ({\tt first})}
    \begin{itemize}
        \item Initialization:
        \item Running:
    \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Combinator Implementations: Parallel Composition ({\tt second})}
    \begin{itemize}
        \item Initialization:
        \item Running:
    \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Time-independence optimization}
\end{frame}

\begin{frame}
\frametitle{Event Merging}
\end{frame}

\end{document}

### 2012-08-29 Docs/IFL2012/presentation/presentation.tex
Presentation source update

@@
 \institute{Rochester Institute of Technology}
+\date{IFL 2012\\ \today}
 
@@
 \begin{frame}
-\frametitle{\insertsection}
 \end{frame}
@@
         \item {\em Reactive} Behavior changes in response to temporal inputs.
-        \item Basic abstractions
+        \item Basic abstractions:
         \begin{itemize}
@@
         \item When to evaluate what?
-        \item Pull evaluation: (``demand-driven'') Evaluate when output is needed.
-        \item Push evaluation: (``data-driven'') Evaluate when input is available.
-        \item Ideally, FRP uses both: pull for signals and push for events.
+        \item Pull evaluation (``demand-driven''):\\ Evaluate when output is needed.
+        \item Push evaluation (``data-driven''):\\ Evaluate when input is available.
+        \item Ideally, an FRP system uses push for events and pull for signals\footcite{Elliott2009}.
     \end{itemize}
@@
 \begin{frame}
-\frametitle{Combinator Implementations: Serial Composition ({\tt >>>})}
+\frametitle{Time-independence optimization}
     \begin{itemize}
-        \item Initialization:
-        \item Running:
+        \item Inefficiency: running every time continuation every sample step.
+        \item This happens even if the signal delta is empty (no change)
+        \item Currently this is necessary because the {\em time} delta might
+              cause some output.
+        \item Optimization:
+        \begin{itemize}
+            \item Mark time-independent SFs using a separate constructor.
+            \item "Smart" composition-routing to make a composite SF time-independent
+                  if all of its components are.
+            \item Time-independent time continuation has no time input, only delta.
+            \item Only called if non-empty delta. 
+        \end{itemize}
     \end{itemize}
@@
 \begin{frame}
-\frametitle{Combinator Implementations: Parallel Composition ({\tt first})}
+\frametitle{Event Merging (Semantics Idea)}
     \begin{itemize}
-        \item Initialization:
-        \item Running:
+        \item Push-based events are not tied to samples.
+        \item We thus know only what the last sample time was for an event,
+              not the actual time.
+        \item Can we hide event merging in this loss of precision?
     \end{itemize}
@@
 
-\begin{frame}
-\frametitle{Combinator Implementations: Parallel Composition ({\tt second})}
-    \begin{itemize}
-        \item Initialization:
-        \item Running:
-    \end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Time-independence optimization}
-\end{frame}
-
-\begin{frame}
-\frametitle{Event Merging}
-\end{frame}
-
 \end{document}

### 2012-10-03 Code/time-flies/src/FRP/TimeFlies/SignalFunctions.hs
Add filterList and event accumulation (accumulate, accumulateList) functions

@@ -64,6 +64,9 @@ module FRP.TimeFlies.SignalFunctions (
   integrate,
   -- * Events
   filter,
+  filterList,
+  accumulate,
+  accumulateList,
   -- * Joining
   union,
   combineSignals,
@@ -354,6 +357,30 @@ filter f = SF (\_ -> (sampleEvt, filterInit f))
 filterInit :: (a -> Maybe b) -> SF Initialized (SVEvent a) (SVEvent b)
 filterInit f = SFInit (\_ _ -> (deltaNothing, [], filterInit f))
                       (\evtOcc -> (maybe [] ((:[]) . occurrence) $ f $ fromOccurrence evtOcc, filterInit f))
+                      
+-- | Filter event occurrences with a list-producing function, producing each element of the list as an event.
+filterList :: (a -> [b]) -> SVEvent a :~> SVEvent b
+filterList f = SF (\_ -> (sampleEvt, filterListInit f))
+
+filterListInit f = let fli = SFInit (\_ _ -> (deltaNothing, [], fli))
+                                    (\evtOcc -> (map occurrence $ f $ fromOccurrence evtOcc, fli))
+                   in fli
+                                    
+-- | Accumulate over event occurrences
+accumulate :: (a -> b -> (Maybe c, a)) -> a -> SVEvent b :~> SVEvent c
+accumulate f a = acc a
+  where acc a = switch (pureEventTransformer (f a) >>>
+                        copy >>>
+                        first (pureEventTransformer fst >>> filter id) >>>
+                        second (pureEventTransformer (acc . snd)))
+
+-- | Accumulate over event occurrences, with lists of event outputs
+accumulateList :: (a -> b -> ([c], a)) -> a -> SVEvent b :~> SVEvent c
+accumulateList f a = acc a
+  where acc a = switch (pureEventTransformer (f a) >>>
+                        copy >>>
+                        first (pureEventTransformer fst >>> filterList id) >>>
+                        second (pureEventTransformer (acc . snd)))
 
 -- | Provide the input as input to the given signal function, producing
 -- the left side of the given signal function's output as output. Upon

### 2012-10-03 Docs/IFL2012/presentation/presentation.tex

@@
     \begin{itemize}
-        \item {\em Functional} First-class and higher-order functions.
-        \item {\em Reactive} Behavior changes in response to temporal inputs.
+        \item {\em Functional}: First-class and higher-order functions.
+        \item {\em Reactive}: Behavior changes in response to temporal inputs.
         \item Basic abstractions:
@@
             \item Signals - Functions of time\footnote{Often also called behaviors.}.
-            \item Events - Sequences of temporally ordered and labeled discrete values.
+            \item Events - Sequences of temporally ordered and labeled\\
+                  discrete values.
         \end{itemize}
+    \item ``Classic'' FRP - Signals and events are first-class.
     \end{itemize}
@@
 \begin{frame}
-\frametitle{Signal-Function FRP}
+\frametitle{Signal-Function FRP (SF-FRP)}
     \begin{itemize}
         \item Signal functions are transformers of events and signals.
-        \item Signal functions are first class in Signal-Function FRP.
-        \item Signals and events are not first-class in Signal Function FRP.
-        \item This approach is more composable than first-class signals and events (since input may be transformed)\footcite{Courtney2001-1}.
+        \item Signal functions are first-class in Signal-Function FRP.
+        \item Signals and events are not first-class in Signal-Function FRP.
+        \item This approach is more composable than first-class\\
+              signals and events (since input may be transformed)\footcite{Courtney2001-1}.
     \end{itemize}
@@
         \item Push evaluation (``data-driven''):\\ Evaluate when input is available.
-        \item Ideally, an FRP system uses push for events and pull for signals\footcite{Elliott2009}.
+        \item Ideally, an FRP system uses push for events and\\
+              pull for signals\footcite{Elliott2009}.
+        \item This has been achieved for ``classic'' FRP
+              (first class signals and events) but not SF-FRP\footcite{Elliott2009}.
     \end{itemize}
@@
     \begin{itemize}
-        \item Necessary to lift any function to a corresponding signal function, without input/output type annotation
+        \item The abstraction must lift any function to a corresponding signal function, without input/output type annotation.
         \item Events encoded as optional signals:
@@
 \end{Verbatim}
-        \item Combined signals and events encoded as one signal with pair type:
+        \item Combined signals and events encoded as one signal with\\
+              a pair type:
     \end{itemize}
@@
     \begin{itemize}
-        \item Described by Sculthorpe\footcite{Sculthorpe2011}
-        \item Describe separation of individual signals and events
+        \item Described by Sculthorpe\footcite{Sculthorpe2011}.
+        \item Describe separation of individual signals and events.
         \item In Haskell (with {\tt -XEmptyDataDecls}):
@@
 
-% Add conclusion
+\begin{frame}
+\frametitle{Conclusion}
+    \begin{itemize}
+        \item Signal vectors fix an abstraction leak in Signal-Function FRP.
+        \item Representing events distinctly from signals in the type permits
+              heterogeneous (push-pull) evaluation of signal functions.
+    \end{itemize}
+\end{frame}
 
@@
         \item Dynamic collections: Dynamically switch between collections of signal functions.
+        \item Demonstrations and performance comparisons.
         \item Semantics/correctness proof (especially for event merging).

### 2012-10-13 Docs/Thesis/system_design_and_interface.tex
Thesis: evaluation interface description

@@
 
-\subsection{Efficient and Push-Based Evaluation}
+\subsection{Efficient Evaluation}
 \label{subsection:System_Design_and_Interface-Goals-Efficient_and_Push_based_Evaluation}
 
-Efficient evaluation is the for push-based evaluation. Since FRP programs are
-expected to  interact with an outside world in real time, efficiency cannot be
-measured by runtime. Thus, when speaking of efficiency, we are expressing a
-desire that the system utilize as few system resources as possible for the task
-at hand, while responding as quickly as possible to external inputs and
+Efficient evaluation is the motivation for push-based evaluation of events.
+Since FRP programs are expected to  interact with an outside world in real time,
+efficiency cannot be measured by the runtime of a program. Thus, when speaking of efficiency,
+we are expressing a desire that the system utilize as few system resources as possible
+for the task at hand, while responding as quickly as possible to external inputs and
 producing output at a consistently high sample rate.
@@
 
-We would like for the FRP system to be easily to integrate with such IO systems,
+We would like for the FRP system to be easy to integrate with such IO systems,
 while being flexible enough to enable its use with other forms of IO systems,
@@
 
+The evaluation interface provides a modified state monad which holds a signal
+function, together with some additional information, as its state. Rather than
+monadic instructions to put and get the state, the monad provides instructions
+to trigger an input event, update an input signal, and trigger sampling of
+signals in the signal function. Additional state includes the current set of
+modifications to the input signals (since the last sample) and a set of
+handlers which actuate effects based on output events or changes to the output
+signal. The interface combinators are shown in Figure \ref{figure:evaluation_combinators}.
+
+\begin{figure}
+\begin{code}
+data SVSample sv
+
+data SVEventInput sv
+instance SVRoutable SVEventInput sv
+
+data SVSignalUpdate sv
+instance SVRoutable SVSignalUpdate sv
+
+data SVHandler out sv
+
+svLeft          :: SVRoutable r =>
+                   r svLeft -> r (svLeft :^: svRight)
+svRight         :: SVRoutable r =>
+                   r svRight -> r (svLeft :^: svRight)
+
+svOcc           :: a -> SVEventInput (SVEvent a)
+svSig           :: a -> SVSignalUpdate (SVSignal a)
+
+emptyHandler    :: SVHandler out SVEmpty
+signalHandler   :: (a -> out) -> SVHandler out (SVSignal a)
+eventHandler    :: (a -> out) -> SVHandler out (SVEvent a)
+combineHandlers ::    SVHandler out svLeft
+                   -> SVHandler out svRight
+                   -> SVHandler out (svLeft :^: svRight)
+
+sample          :: a -> SVSample (SVSignal a)
+sampleEvt       :: SVSample (SVEvent a)
+sampleNothing   :: SVSample SVEmpty
+combineSamples  ::    SVSample svLeft
+                   -> SVSample svRight
+                   -> SVSample (svLeft :^: svRight)
+
+data SFEvalState m svIn svOut
+initSFEval ::    SVHandler (m ()) svOut
+              -> SVSample svIn
+              -> Double
+              -> (svIn :~> svOut)
+              -> SFEvalState m svIn svOut
+
+data SFEvalT svIn svOut m a
+instance MonadTrans (SFEvalT svIn svOut)
+instance (Monad m) => Monad (SFEvalT svIn svOut m)
+instance (Functor m) => Functor (SFEvalT svIn svOut m)
+instance (Monad m, Functor m) => Applicative (SFEvalT svIn svOut m)
+instance (MonadIO m) => MonadIO (SVEvalT svIn svOut m)
+
+runSFEvalT ::    SFEvalT svIn svOut m a
+              -> SFEvalState m svIn svOut
+              -> m (a, SFEvalState m svIn svOut)
+
+push :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
+update :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
+step :: (Monad m) => Double -> SFEvalT svIn svOut m ()
+\end{code}
+\hrule
+\caption{Evaluation combinators.}
+\label{figure:evaluation_combinators}
+\end{figure}
+
 \subsection{Initialization and Running}
+To evaluate a signal function, one first constructs an initial state
+with the {\tt initSFEval} function. This function takes a set of
+handlers, a set of initial samples for the signal inputs, the original
+time (as a {\tt Double}) and the signal function to evaluate.
 
-\subsection{Input and Output}
+The signal function replaces itself after each event occurrence or time
+step. The current signal function is maintained in the state. Calling
+{\tt push} causes the signal function to react to the supplied event 
+occurrence, and events which subsequently occur on the signal function's
+output are passed to the supplied handlers. Calling {\tt update} updates
+a set of new signal inputs. This set is nullified after it is passed in
+along with a new time by calling {\tt step}. This allows various
+components of a system to update individual signal inputs in a manner which is
+decoupled from the sample timer.
+
+When a monadic evaluation action is passed to the {\tt runSFEvalT} function,
+along with a state, the result is an action in the underlying monad. This
+action, when run in the underlying monad ({\tt IO}, for example) produces
+a pair of an output value and a new state. This new state is what should be 
+supplied to the next invocation of {\tt runSFEvalT}.
 
 \subsection{Monad Lifting}
+
+{\tt SFEvalT} is not only a monad, but a monad transformer. This means that
+it is parameterized by another monad, and that "running" an {\tt SFEvalT}
+action produces an action in this underlying monad. It means also that actions
+from this underlying monad may be ``lifted'' into actions of {\tt SFEvalT}.
+So if the {\tt IO} monad is the underlying monad, {\tt IO} actions can be 
+used to obtain input and arrange for timing. The underlying monad is also the
+monad whose actions the output handlers produce, which provides the means of
+actuating the output of the signal function.

### 2012-10-13 Docs/Thesis/abstract.tex
Thesis: edited abstract

@@
 Functional Reactive Programming (FRP) is a promising class of abstractions for encoding interactive and time-dependent programs in
-functional languages, but has proven difficult to implement efficiently. Signal-function FRP is a subclass of FRP which does not provide
+functional languages which has proven difficult to implement efficiently. Signal-function FRP is a subclass of FRP which does not provide
 direct access to the time-varying values by the programmer, but rather provides signal functions as first-class objects in the
-program. All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation, producing output from
-the FRP system whenever the consumer of the output is ready. This greatly simplifies the implementation of signal-function FRP systems, but
-leads to inefficient and wasteful evaluation of the FRP system.
+program. FRP systems provide values defined at all points in time ({\em behaviors} or {\em signals}) and values defined
+at countably many points in time ({\em events}).
+
+All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation for both events and signals,
+producing output from the FRP system whenever the consumer of the output is ready. This greatly simplifies the implementation of signal-function
+FRP systems, but leads to inefficient and wasteful evaluation of the FRP system.
 
@@
 evaluating the network when nothing has changed, and then only components necessary to react to the input are reevaluated.
-This form of evaluation has been partially applied to standard FRP systems~\cite{Elliott2009} but not to signal-function FRP systems.
+This form of evaluation has been applied to events in standard FRP systems~\cite{Elliott2009} but not in signal-function FRP systems.
 
 I describe the implementation of a push-based signal-function FRP system. The semantics of the system are discussed, and its performance
-and expressiveness for practical examples of interactive programs are be compared to existing signal-function FRP systems through the
-implementation of an example application.
+and expressiveness for practical examples of interactive programs are compared to existing signal-function FRP systems through the
+implementation of a networking application.

### 2012-10-13 Docs/Thesis/introduction.tex
Thesis: rewrote introduction ending

@@
 The first challenge was addressed by representing FRP programs, not as compositions of signals and events, but as compositions of
-{\em transformers} of behaviors and events called {\em signal functions}. By not permitting direct access to behaviors or events, but
+transformers of behaviors and events called {\em signal functions}. By not permitting direct access to behaviors or events, but
 representing them implicitly, signal functions prohibit accessing past or future time values, only allowing access to values in the closure of the signal function
@@
 
-N-ary FRP provides a model of a type system for a push-based signal-function FRP system. In order to represent signal functions
-as continuations on collections of inputs rather than on single values, we represent the signal vectors as types over which
-the signal function type is parameterized for both inputs and outputs. Indices into the signal vector are typed with the signal vector. These indices
-contain the actual input or output values and form elements for a collection of inputs or outputs. Obviously the collection need not have all possible elements,
-which forms the basis for evaluating only those parts of the network which depend on the updated values.
+This thesis describes the design and implementation of Time-Flies, a push-pull signal-function FRP system. This system is presented as a library
+of combinators for constructing and evaluating FRP programs.

### 2012-10-13 Docs/Thesis/background.tex
Thesis: Background edits

@@
 
-The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produces functions
+The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produce functions
 as output and/or take functions as input. We say that functions are {\em first-class values} in a language.
 
-Most reactive programs are written in an imperative style, using low-level and non-composable abstractions including callbacks
+Even in functional languages, reactive programs are generally written in an imperative style, using low-level and non-composable abstractions including callbacks
 or object-based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models.
 
-FRP implies that a model should keep the characteristics of functional programming (i.e. that the basic constructs of the language
+{\em Functional} Reactive Programming implies that a model should keep the characteristics of functional programming (i.e. that the basic constructs of the language
 should remain first-class) while incorporating reactivity into the language model. In particular, functions should be lifted to operate on reactive values,
@@
 to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
-based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in functional reactive
-models should of course be efficiently executable. This has proven to be the main challenge in implementing Functional Reactive Programming.
+based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in FRP languages
+should of course be efficiently executable. This has proven to be the main challenge in implementing FRP.
 
@@
 
-The earliest and still standard formulation of FRP is two primitive type
+The earliest and still standard formulation of FRP provides two primitive type
 constructors: {\tt Behavior} and {\tt Event}, together with combinators that
-produce values of this type. The easiest semantic definition for these types
-is given in Figure~\ref{figure:classic_frp_semantic_types}
+produce values of these types. The easiest semantic definition for these types
+is given in Figure~\ref{figure:classic_frp_semantic_types}\footnote{The list constructor {\tt []} should be considered to contain infinite as well as finite lists.}.
 
@@
 and the other from a time interval (a lower and upper time bound) to a value
-interval and a new behavior. Events are represented as {\tt improving values},
+interval and a new behavior. Events are represented as ``improving values'',
 which, when sampled with a time, produce a lower bound on the time to the next
@@
 FRP implementation for the Java Beans framework. Frapp\'{e} built on the notion
-of events and {\em bound properties} in the Beans framework, providing abstract
+of events and ``bound properties'' in the Beans framework, providing abstract
 interfaces for FRP events and behaviors, and combinators as concrete classes
@@
 
-The Reactive~\cite{Elliott2009} system is a push-based FRP system with
+The Reactive~\cite{Elliott2009} system is a push-pull FRP system with
 first-class behaviors and events. The primary insight of Reactive is the
@@
 {\em signal functions}. Arrows are abstract type constructors with input and
-output type parameters, together with sequential ({\tt >>>}) and parallel
-({\tt first} and {\tt second}) composition, branching ({\tt ***}), and lifting
-({\tt arr}) functions to the Arrow type. Signal functions are the first-class
+output type parameters, together with a set of routing combinators (see Figure~\ref{figure:arrow_combinators}).
+Signal functions are the first-class
 abstraction in this approach, they represent time-dependent and reactive
@@
 signals and events.
+
+\begin{figure}
+\begin{code}
+(>>>)  :: (Arrow a) => a b c -> a c d -> a b d
+arr    :: (Arrow a) => (b -> c) -> a b c
+first  :: (Arrow a) => a b c -> a (b, d) (c, d)
+second :: (Arrow a) => a b c -> a (d, b) (d, c)
+(***)  :: (Arrow a) => a b c -> a b d -> a b (c, d)
+loop   :: (Arrow a) => a (b, d) (c, d) => a b c
+\end{code}
+\hrule
+\caption{Arrow combinators.}
+\label{figure:arrow_combinators}
+\end{figure}
 

### 2012-10-13 Code/openflow-switch/src/openflow-switch.hs
Non-FRP openflow switch for comparison

@@ -1,0 +1,171 @@
+module Main where
+
+import           Control.Monad (mapM_)                -- Monad utilities
+import           Data.HList                           -- Parsed frames are HLists
+import           Data.IORef                           -- Storing the table
+import qualified Data.Map                        as M -- Table data structure
+import           Data.Word                            -- Representation-specific numbers
+import           GHC.Event                            -- GHC event loop
+import           Nettle.OpenFlow                      -- OpenFlow API
+import           Nettle.Servers.EventServer           -- OpenFlow event-based server
+import           System.Clock (getTime,               -- A monotonic clock
+                               Clock(Monotonic),
+                               TimeSpec,
+                               sec,
+                               nsec)
+
+port :: Word16
+port = 6633
+
+tableTime :: Double
+tableTime = 5.0
+
+cleanInterval :: Int
+cleanInterval = 10000
+
+main :: IO ()
+main = do tableRef <- newIORef M.empty
+          ofEventManager <- 
+            openFlowEventManager Nothing port handleSwitchConnect
+                                 (handleSCMessage tableRef)
+          let evtMgr = getEventManager ofEventManager
+          cleanTable cleanInterval tableTime tableRef evtMgr
+          loop $ evtMgr
+          
+
+-- | Get the monotonic time as a Double number of seconds
+getMonoTimeAsDouble :: IO Double
+getMonoTimeAsDouble =
+  fmap (\ts -> fromIntegral (sec ts) + 1.0e-9 * fromIntegral (nsec ts)) $
+    getTime Monotonic
+
+-- | An entry in the switch table
+data SwitchEntry =
+  SwitchEntry { 
+    swEntryHandle         :: SwitchHandle EthernetFrame,
+    swEntryAddress        :: EthernetAddress,
+    swEntryPort           :: PortID,
+    swEntryCurrentMatches :: [Match],
+    swEntryTime           :: Double
+  }
+  
+
+type SwitchTableKey = (SwitchID, EthernetAddress)
+type SwitchTable = M.Map SwitchTableKey SwitchEntry
+  
+addMatchToSwitchEntry :: Match -> SwitchEntry -> SwitchEntry
+addMatchToSwitchEntry m swe =
+  swe { swEntryCurrentMatches = m : (swEntryCurrentMatches swe) }
+
+addMatchToTable :: Match -> SwitchTableKey -> SwitchTable -> SwitchTable
+addMatchToTable = M.adjust . addMatchToSwitchEntry
+
[…]

### 2012-10-23 Code/time-flies/src/FRP/TimeFlies/SignalFunctions.hs
Make capture keep original event data

@@ -530,12 +530,15 @@ combineSignalsInit f currentSample = SFInit (\dt sigDelta -> let newSample = upd
 -- | Combine a signal and an event by producing an output event occurrence
 -- for each input event occurrence, but with the value of the signal
 -- at that time interval
-capture :: (SVSignal a :^: SVEvent b) :~> SVEvent a
+capture :: (SVSignal a :^: SVEvent b) :~> SVEvent (b, a)
 capture = SF ((sampleEvt,) . captureInit . sampleValue . fst . splitSample)
 
-captureInit :: a -> SF Initialized (SVAppend (SVSignal a) (SVEvent b)) (SVEvent a)
-captureInit x = SFInit (flip (const . (deltaNothing, [],) . maybe (captureInit x) captureInit . deltaValue . fst . splitDelta))
-                       ((, captureInit x) . const [occurrence x])
+captureInit :: a -> SF Initialized (SVAppend (SVSignal a) (SVEvent b)) (SVEvent (b, a))
+captureInit x = let cap = SFInit (flip (const . (deltaNothing, [],) . maybe (captureInit x) captureInit . deltaValue . fst . splitDelta))
+                                 (\evtOcc -> ([case chooseOccurrence evtOcc of
+                                                 Right evtOccRight -> occurrence (fromOccurrence evtOccRight, x)
+                                              ], cap))
+                in cap
 
 -- | SFEvalState
 data SFEvalState m svIn svOut = SFEvalState { 

### 2012-10-23 Code/openflow-switch-timeflies/src/openflow-switch-timeflies.hs
TimeFlies based openflow learning switch

@@ -1,0 +1,228 @@
+{-# LANGUAGE TypeOperators #-}
+module Main where
+
+import           Control.Monad (mapM_)                -- Monad utilities
+import           Data.HList                           -- Parsed frames are HLists
+import           Data.IORef                           -- Storing the table
+import qualified Data.Map                        as M -- Table data structure
+import           Data.Word                            -- Representation-specific
+                                                      -- numbers
+import           FRP.TimeFlies.SignalFunctions        -- FRP
+                   hiding (loop)
+import           GHC.Event hiding (step)              -- GHC event loop
+import           Nettle.OpenFlow                      -- OpenFlow API
+import           Nettle.Servers.EventServer           -- OpenFlow event-based server
+import           System.Clock (getTime,               -- A monotonic clock
+                               Clock(Monotonic),
+                               TimeSpec,
+                               sec,
+                               nsec)
+import           System.Environment (getArgs)                               
+
+import qualified Data.List as L (filter)
+import Prelude hiding (filter)
+
+-- | The port for the server to listen on.
+port :: Word16
+port = 6633
+
+-- | The amount of time a rule should persist in the table
+tableTime :: Double
+tableTime = 5.0
+
+-- | Entry point
+main :: IO ()
+main = do (argS:_) <- getArgs
+          let sampleTime = read argS
+          time <- getMonoTimeAsDouble
+          reactiveRef <- newIORef $ initSFEval (eventHandler id) (combineSamples sampleEvt sampleEvt) time learningSwitch
+          ofEventManager <- 
+            openFlowEventManager Nothing port (\handle -> do rState <- readIORef reactiveRef
+                                                             ((), rState') <- runSFEvalT (push $ svLeft $ svOcc handle) rState
+                                                             writeIORef reactiveRef rState')
+                                 (\((tid, msg), handle) -> do rState <- readIORef reactiveRef
+                                                              ((), rState') <- runSFEvalT (push $ svRight $ svOcc (handle, tid, msg)) rState
+                                                              writeIORef reactiveRef rState')
+          let evtMgr = getEventManager ofEventManager
+              sample = void $ registerTimeout evtMgr sampleTime $ do rState <- readIORef reactiveRef
+                                                                     time <- getMonoTimeAsDouble
+                                                                     ((), rState') <- runSFEvalT (step time) rState
[…]

### 2012-10-23 Code/openflow-switch-yampa/src/openflow-switch-yampa.hs
Yampa-based OpenFlow learning switch

@@ -1,0 +1,247 @@
+{-# LANGUAGE TypeOperators #-}
+module Main where
+
+import           Control.Concurrent
+import           Control.Concurrent.MVar
+import           Control.Monad (mapM_)                -- Monad utilities
+import           Data.HList                           -- Parsed frames are HLists
+import           Data.IORef                           -- Storing the table
+import qualified Data.Map                        as M -- Table data structure
+import           Data.Word                            -- Representation-specific
+                                                      -- numbers
+import           FRP.Yampa                            -- FRP
+                   hiding (loop)
+import           Nettle.Servers.Server                -- OpenFlow server
+import           Nettle.OpenFlow                      -- OpenFlow API
+import           System.Clock (getTime,               -- A monotonic clock
+                               Clock(Monotonic),
+                               TimeSpec,
+                               sec,
+                               nsec)
+import           System.Environment (getArgs)
+                               
+import qualified Data.List as L (filter)
+import Prelude hiding (filter)
+
+-- | The port for the server to listen on.
+port :: Word16
+port = 6633
+
+-- | The amount of time a rule should persist in the table
+tableTime :: Double
+tableTime = 5.0
+
+
+-- | Entry point
+main :: IO ()
+main = do (argS:_) <- getArgs
+          let sampleTime = read argS
+          inputVar <- newEmptyMVar
+          server <- startOpenFlowServer Nothing port
+          forkIO $ let acceptLoop = do handle <- acceptSwitch server
+                                       handshake handle
+                                       forkIO $ let readLoop = do mMsg <- receiveFromSwitch handle
+                                                                  case mMsg of
+                                                                    Just (tid, msg) -> putMVar inputVar (handle, tid, msg)
+                                                                    _ -> return ()
+                                                                  readLoop
+                                                in readLoop
+                                       acceptLoop
+                   in acceptLoop
+          do timeRef <- newIORef 0
+             readRef <- newIORef True
+             reactimate (do time <- getMonoTimeAsDouble
+                            writeIORef timeRef time
+                            return NoEvent
+                        )
+                        (const $ do read <- readIORef readRef
+                                    time <- getMonoTimeAsDouble
+                                    oldTime <- readIORef timeRef
+                                    let dt = time - oldTime
+                                    if read
[…]

### 2012-10-24 Code/openflow-switch-timeflies/src/openflow-switch-timeflies.hs
Resolve scoping problem

@@ -10,6 +10,7 @@ import           Data.Word                            -- Representation-specific
 import           FRP.TimeFlies.SignalFunctions        -- FRP
                    hiding (loop)
 import           GHC.Event hiding (step)              -- GHC event loop
+import           Nettle.Servers.Server (handshake)
 import           Nettle.OpenFlow                      -- OpenFlow API
 import           Nettle.Servers.EventServer           -- OpenFlow event-based server
 import           System.Clock (getTime,               -- A monotonic clock

### 2012-10-24 Docs/Thesis/comparison.tex
First draft of comparison chapter

@@
 practice.
+
+One way to evaluate this is to write an application in two systems, and benchmark
+these systems for comparison. In order to compare TimeFlies with Yampa, the most
+recently released pull-based signal-function FRP system, we decided to implement
+a suitable application in both.
+
+The application of choice is a learning switch controller for OpenFlow switches.
+this application must handle input events triggered by switches signalling that
+they have received a packet for which they have no handling rules, and produce
+output events which install handling rules for dealing with these packets.
+
+The reactive algorithm is simple. A table from machine addresses to switch ports
+is maintained. This table is updated in two ways. When a packet is sent to the
+controller, meta-data for the packet indicates which port it arrived on. The
+source address of the packet can then be mapped to that port in the table.
+Once this occurs, rules are installed for each packet with a destination address
+matching the source, along with reciprocal rules.
+
+Each mapping in the table has a timestamp. After a certain amount of time, the
+table entry is deleted. The table entry is also deleted if a packet is received
+with the same source address and a different port. When an entry is deleted,
+the switch is signaled to remove all corresponding rules.
+
+In both Yampa and TimeFlies, this algorithm is implemented by using an accumulator,
+which applies functions carried by event occurrences to a state value. It stores
+the new state and outputs new event occurrences based on the output of the function.
+
+The state stored is the table (implemented using the {\tt Data.Map} module in the
+Haskell {\tt containers} library) and the output event occurrences carry messages
+to be sent to the switch. The events are sourced internally to the learning
+switch signal function (using the {\tt after} combinator in TimeFlies and the {\tt repeatedly}
+function in Yampa) to delete expired rules, and externally, by transforming messages
+from the switch to the controller into accumulator functions.
+
+The controllers were benchmarked using the {\tt cbench} program~\cite{cbench}. The
+amount to time to pause between time and signal updates was passed as a parameter to
+both programs. The {\tt cbench} benchmark operates by sending a simulated packet in
+event, waiting for a rule response, and then repeating. The output of the benchmark
+is the number of times this occurs in a second. This is thus a very good test of
+event latency in a reactive system.
+
+Figure \ref{fig:timeflies-yampa-comparison} shows the performance comparison.
+Toward the left of the graph is lower pause times (more frequent sampling). The
+y-axis is a higher rate of responses (higher numbers are preferred).
+
+\begin{figure}
+\epsfig{file=graph.eps,angle=-90,width=\linewidth}
+\hrule
+\caption{Comparison of timeflies vs. Yampa implementing an OpenFlow learning switch controller}
+\label{fig:timeflies-yampa-comparison}
+\end{figure}
+
+There is some unexpected correlation between the sampling rate and the response rate
+for TimeFlies, which is likely due to the more frequent size reductions in the table.
+Nevertheless, TimeFlies outperforms Yampa in event response by several orders of magnitude,
+even at a very high sampling rate. This is due to two factors. First, TimeFlies can react
+to individual events without evaluating the whole signal function. Second, the
+interface of time-flies permits us to connect its evaluation directly to GHC's event
+manager, while the design of Yampa requires us to use threads and synchronized communication\footnote{Yampa does include a step-wise evaluation function, but it still couples event
+evaluation firmly to time steps, and is not well documented.}.

### 2012-10-24 Docs/Thesis/conclusions-and-further-work.tex
First draft of comclusions chapter

@@
 push-based signal-function system.
+
+The TimeFlies system is a full-fledged and documented FRP library which may
+be extended with utility functions and further optimized. Its performance in
+responding to events is demonstrated to be superior to that of the predominant
+pull-based signal-function FRP library.
+
+Further, the model of events used by TimeFlies subverts problematic semantic
+questions about the evaluation of events in a system. By using the N-Ary FRP
+type model and separating the evaluation of events from the time steps used
+for signals, TimeFlies fully supports non-deterministic merging of events,
+and provides a semantic guarantee that events are not ''lost'' during evaluation.
+
+The TimeFlies system would benefit from attentive microbenchmarking and performance
+tuning, as well as optimizations to avoid evaluating irrelevant parts of the network
+during the the evaluation of time steps. A formal semantic justification for the
+formulation of event evaluation (which would require a full formal semantics of FRP)
+would enable a far more robust correctness argument, as well as providing a basis for
+semantic extensions to signal-function FRP.
+
+FRP is not yet mature, and has not been the subject of focused application development.
+Thus, there is a dearth of design patterns for FRP applications. Such design patterns
+would yield necessary feedback as to which generalizations and restrictions of FRP would
+be appropriate and useful, and clarify the necessity of various utility combinators to
+be included in the standard libraries of FRP systems.
+
+In order to improve the performance of FRP yet further, it may be productive to attempt
+to introduce parallel evaluation into FRP, taking advantage of the functional purity in
+the implementation of the signal function combinators. This may involve, for instance,
+evaluating several time-steps at once in a data-parallel manner,task parallelism
+between different branches of a signal function, or speculative evaluation of switch
+combinators.
+
+Many classes of reactive application would benefit from a ``dynamic collections'' combinator
+similar to {\tt pSwitch} in Yampa. Such a combinator allows a collection of signal functions
+to be evaluated as one signal function, with addition or removal of signal functions instead
+of the total replacement given by the {\tt switch} combinator. This is useful, for instance,
+when simulating objects for games or computer models, as the behavior of each object can be
+modeled as a signal function, and these signal functions can be added to and removed from the
+collection.
+
+The TimeFlies system provides a principled and performant system for future experimentation
+on FRP, as well as implementation of FRP applications.

### 2012-10-24 Docs/Thesis/system_design_and_interface.tex
Add semantics section to system design chapter

@@
 such as simple imperative systems, threaded systems, or network servers.
+
+\section{Semantics}
+\label{section:System_Design_and_Interface-Semantics}
+
+A rigorous and formal elucidation of the semantics of signal-function FRP remains
+unattempted, but there is a sufficient change to the practical semantics of
+signal-function FRP between previous signal-function systems and TimeFlies to warrant
+some description.
+
+In previous systems such as Yampa, events were understood (and typed) as option-valued
+signals. This approach is undesirable for several reasons. The most pressing reason is
+that it prohibits push-based evaluation of events, because events are embedded in the
+samples of a signal and must be searched for.
+
+Another concern is that this approach limits the event rate to the sampling rate.
+The rate of sampling should, at some level, not matter to the FRP system. Events which
+occur between sampling intervals are never observed by the system.
+
+This concern drives the next concern. Events are not instantaneous in this formulation.
+If a signal is option valued, the sampling instant must fall within the interval where
+there is an event occurrence present for that event to be observed. If events are
+instantaneous, the probability of observing an event occurrence is zero.
+
+Therefore, TimeFlies employs the N-Ary FRP type formulation to represent signals and
+events as distinct entities in the inputs and outputs of signal functions. This means
+we are now free to choose our representation of events, and to separate it from the
+representation and evaluation of signals.
+
+This freedom yields the additional ability to make events independent of the sampling
+interval altogether. The semantics of event handling in TimeFlies is that an event occurrence
+is responded to immediately, and does not wait for the next sampling instant. This allows events
+to be instantaneous, and further, allows multiple events to occur within a single sampling interval.
+
+There are two tradeoffs intrinsic to this approach. The first is that events are only partially ordered
+temporally. There is no way to guarantee the order of observation of event occurrences occurring in the
+same sampling interval. Further, the precise time of an event occurrence cannot be observed, only the 
+time of the last sample prior to the occurrence.
+
+In return for giving up total ordering and precise observation of the timing of events, we obtain the
+ability to employ push-based evaluation for event occurrences, and the ability to non-deterministically
+merge event occurrences. When events being input to a non-deterministic merge have simultaneous occurrences,
+we simply select one arbitrarly to occur first. This does not violate any guarantee about time values, since
+they will both have the same time value in either case, and does not violate any guarantee about ordering,
+since no guarantee of their order is given.
+
+A formal semantic description of signal function FRP would clarify the consequences of this decision somewhat,
+but is outside the scope of this thesis.
 

### 2012-10-25 Docs/Thesis/system_design_and_interface.tex
Updates to system design chapter, still need more combinator descriptions

@@
 
-The representation of signal functions is also discussed in
+Signal functions with signal vectors as input and output types form a
+Haskell {\tt GArrow}~\cite{Megacz2011}. Specifically, the signal function
+type constructor (with the initialization parameter fixed) forms the arrow
+type, the {\tt SVAppend} type constructor forms the product type, and the
+{\tt SVEmpty} type constructor forms the unit type.
+
+The representation of signal functions is discussed in
 Section~\ref{section:Implementation-Signal_Functions}. Here it suffices to say
@@
 
-The type synonyms {\tt :~>} and {\tt :\textasciicircum:} are included for readability and are
+The type synonyms {\tt :\textasciitilde>} and {\tt :\textasciicircum:} are included for readability and are
 not crucial to the FRP system.
@@
 \begin{code}
-identity :: SF NonInitialized sv sv
-constant :: a -> SF NonInitialized SVEmpty (SVSignal a)
-never    :: SF NonInitialized SVEmpty (SVEvent a)
-asap     :: a -> SF NonInitialized SVEmpty (SVEvent a)
-after    :: Double -> a -> SF NonInitialized SVEmpty (SVEvent a)
+-- Pass the input unmodified to the output
+identity :: sv :~> sv
+
+-- Produce a signal which is at all times the supplied value
+constant :: a -> SVEmpty :~> SVSignal a
+
+-- An event with no occurrences
+never    :: SVEmpty :~> SVEvent a
+
+-- An event with one occurrence, as soon as possible after
+-- the signal function is initialized
+asap     :: a -> SVEmpty :~> SVEvent a
+
+-- An event after the specified amount of time has elapsed.
+after    :: Double -> a -> SVEmpty :~> SVEvent a
 \end{code}
@@
 \begin{code}
-pureSignal ::    (a -> b) 
-              -> SF NonInitialized (SVSignal a) (SVSignal b)
-pureEvent  ::    (a -> b)
-              -> SF NonInitialized (SVSignal a) (SVSignal b)
+-- Apply the given function to a signal at all points in time
+pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal b
+
+-- Apply the given function to each event occurrence
+pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent b
 \end{code}
@@
 
+The routing combinators are used to combine signal functions, and
+to re-arrange signal vectors in order to connect signal functions.
+The routing combinators are shown in Figure~\ref{figure:routing_combinators}.
+
+Only those combinators which modify or combine signal functions
+({\tt (>>>)}, {\tt first}, {\tt second}) are reactive, and then
+only if they inherit their reactivity from the signal function(s)
+they modify. The rest do not react to or modify the input in any
+way, except to re-arrange it, copy it, or discard it altogether.
+
 \begin{figure}
 \begin{code}
-(>>>)       ::    SF NonInitialized svIn svBetween
-               -> SF NonInitialized svBetween svOut
-               -> SF NonInitialized svIn svOut
-first       ::    SF NonInitialized svIn svOut
-               -> SF NonInitialized (SVAppend svIn sv)
-                                    (SVAppend svOut sv)
-second      ::    SF NonInitialized svIn svOut
-               -> SF NonInitialized (SVAppend sv svIn)
-                                    (SVAppend sv svOut)
-swap        ::    SF NonInitialized (SVAppend svLeft svRight) 
-                                    (SVAppend svRight svLeft)
+-- Use the output of one signal function as the input for another
+(>>>) :: (svIn :~> svBetween) -> (svBetween :~> svOut) -> svIn :~> svOut
+
+-- Pass through the right side of the input unchanged
+first :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
+
+-- Pass through the left side of the input unchanged
+second :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut)
+
+-- Swap the left and right sides
+swap :: (svLeft :^: svRight) :~> (svRight :^: svLeft)
+
+-- Duplicate the input
+copy :: sv :~> (sv :^: sv)
+
+-- Ignore the input
+ignore :: sv :~> svEmpty
+
+-- Remove an empty vector on the left
+cancelLeft :: (SVEmpty :^: sv) :~> sv
+
+-- Add an empty vector on the left
+uncancelLeft :: sv :~> (SVEmpty :^: sv)
+
+-- Remove an empty vector on the right
+cancelRight :: (sv :^: SVEmpty) :~> sv
+
+-- Add an empty vector on the right
+uncancelRight :: sv :~> (sv :^: SVEmpty)
+
+-- Make right-associative
+associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))
+
+-- Make left-associative
+unassociate :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
 \end{code}
@@
 
+\subsection{Reactivity}
+\label{subsection:System_Design_and_Interface-Combinators-Reactivity}
+
+Reactivity is introduced by means of the {\tt switch} combinator
+(Figure~\ref{figure:switch_combinator}). The design of this combinator
+allows modular packaging of reactivity. A signal function can determine
+autonomously when to replace itself, based only on its input and state,
+by emitting an event occurrence carrying its replacement. The combinator
+consumes and hides the event carrying the replacement signal function, 
+so the reactivity is not exposed by the resulting reactive signal function.
+
+\begin{figure}
+\begin{code}
+switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
+          -> svIn :~> svOut
+\end{code}
+\hrule
+\caption{Combinator for reactivity.}
+\label{figure:switch_combinator}
+\end{figure}
+
+There are other formulations of a reactive combinator which may be implemented
+using the one supplied. These are shown in Figure~\ref{figure:alternate_switching_combinators}
+and may be provided in a future version of the TimeFlies library.
+
+\begin{figure}
+\begin{code}
+-- Alternate version of switch,
+-- implemented in terms of supplied version
+switch_gen ::    (svIn :~> (svOut :^: SVEvent a))
+              -> (a -> svIn :~> svOut)
+              -> svIn :~> svOut
+switch_gen sf f =
+  switch (sf >>> second (pureEventTransformer f))
+
+-- Supplied version in terms of alternate version
+switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
+          -> svIn :~> svOut
+switch sf = switch_gen sf id
+
+-- Repeated switch, which takes replacement signal functions
+-- externally.
+rswitch ::    (svIn :~> svOut)
+           -> (svIn :^: SVEvent (svIn :~> svOut)) :~> svOut
+rswitch sf =
+  switch (first sf >>> second (pureEventTransformer rswitch))
+\end{code}
+\hrule
+\caption{Alternate combinators for reactivity.}
+\label{figure:alternate_switching_combinators}
+\end{figure}
+
 \section{Evaluation}
@@
 The evaluation interface provides a modified state monad which holds a signal
-function, together with some additional information, as its state. Rather than
-monadic instructions to put and get the state, the monad provides instructions
+function, together with some additional information, as its state (shown in Figure~\ref{figure:evaluation_state}.
+Rather than monadic instructions to put and get the state, the monad provides instructions
 to trigger an input event, update an input signal, and trigger sampling of
@@
 handlers which actuate effects based on output events or changes to the output
-signal. The interface combinators are shown in Figure \ref{figure:evaluation_combinators}.
+signal.
 
@@
 \begin{code}
-data SVSample sv
-
-data SVEventInput sv
-instance SVRoutable SVEventInput sv
-
-data SVSignalUpdate sv
-instance SVRoutable SVSignalUpdate sv
-
+-- A vector of handlers for outputs
 data SVHandler out sv
 
-svLeft          :: SVRoutable r =>
-                   r svLeft -> r (svLeft :^: svRight)
-svRight         :: SVRoutable r =>
-                   r svRight -> r (svLeft :^: svRight)
+-- A dummy handler for an empty output
+emptyHandler    :: SVHandler out SVEmpty
 
-svOcc           :: a -> SVEventInput (SVEvent a)
-svSig           :: a -> SVSignalUpdate (SVSignal a)
+-- A handler for an updated signal sample
+signalHandler   :: (a -> out) -> SVHandler out (SVSignal a)
 
-emptyHandler    :: SVHandler out SVEmpty
-signalHandler   :: (a -> out) -> SVHandler out (SVSignal a)
+-- A handler for an event occurrence
 eventHandler    :: (a -> out) -> SVHandler out (SVEvent a)
+
+-- Combine handlers for a vector
 combineHandlers ::    SVHandler out svLeft
@@
 
-sample          :: a -> SVSample (SVSignal a)
-sampleEvt       :: SVSample (SVEvent a)
-sampleNothing   :: SVSample SVEmpty
-combineSamples  ::    SVSample svLeft
-                   -> SVSample svRight
-                   -> SVSample (svLeft :^: svRight)
+-- The state maintained when evaluating a signal function
+data SFEvalState m svIn svOut
 
-data SFEvalState m svIn svOut
+-- Create the initial state for evaluating a signal function
 initSFEval ::    SVHandler (m ()) svOut
@@
               -> SFEvalState m svIn svOut
+\end{code}
+\hrule
+\caption{State maintained when evaluating a signal function}
+\label{figure:evaluation_state}
+\end{figure}
 
+In order to initialize the state, the user must supply a set of handlers, the
+signal function to evaluate, and initial values for all of the signal inputs
+(Figure~\ref{figure:initial_input}).
+
+\begin{figure}
+\begin{code}
+-- A sample for all leaves of a signal vector
+data SVSample sv
+
+-- Create a sample for a signal leaf
+sample          :: a -> SVSample (SVSignal a)
+
+-- A dummy sample for an event leaf
+sampleEvt       :: SVSample (SVEvent a)
+
+-- A dummy sample for an empty leaf
+sampleNothing   :: SVSample SVEmpty
+
+-- Combine two samples
+combineSamples  ::    SVSample svLeft
+                   -> SVSample svRight
+                   -> SVSample (svLeft :^: svRight)
+\end{code}
+\hrule
+\caption{Data type for initial input}
+\label{figure:initial_input}
+\end{figure}
+
+This state can then be passed to a monadic action which will supply input to
+the signal function. Inputs are constructed using a simple interface with
+functions to construct sample updates and event occurrences, and to specify
+their place in the vector (Figure~\ref{figure:ongoing_input}).
+
+\begin{figure}
+\begin{code}
+-- Class to overload left and right functions
+class SVRoutable r where
+  svLeft          :: r svLeft -> r (svLeft :^: svRight)
+  svRight         :: r svRight -> r (svLeft :^: svRight)
+
+-- An input event occurrence
+data SVEventInput sv
+instance SVRoutable SVEventInput sv
+
+-- An updated sample for a signal
+data SVSignalUpdate sv
+instance SVRoutable SVSignalUpdate sv
+
+-- Create an event occurrence
+svOcc           :: a -> SVEventInput (SVEvent a)
+
+-- Create an updated sample
+svSig           :: a -> SVSignalUpdate (SVSignal a)
+\end{code}
+\hrule
+\caption{Data types for ongoing input.}
+\label{figure:ongoing_input}
+\end{figure}
+
+The {\tt SFEvalT} monad is actually a monad transformer, that is, it is
+parameterized over an underlying monad whose actions may be lifted to
+{\tt SFEvalT}. In the usual case, this will be the {\tt IO} monad.
+
+{\tt SFEvalT} actions are constructed using combinators to push events,
+update inputs, and step time, as well as actions lifted from the underlying
+monad (used to obtain these inputs). An action in the underlying monad
+which produces the result and a new state is obtained with the {\tt runSFEvalT}
+function. These combinators are shown in Figure~\ref{figure:evaluation_combinators}.
+
+\begin{figure}
+\begin{code}
+-- The evaluation monad
 data SFEvalT svIn svOut m a
@@
 
+-- Obtain an action in the underlying monad
+-- from an SFEvalT and a new state.
 runSFEvalT ::    SFEvalT svIn svOut m a
@@
 
+-- Push an event occurrence.
 push :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
+
+-- Update the value of an input signal sample
+-- (not immediately observed)
 update :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
+
+-- Step forward in time, observing the updated signal values
 step :: (Monad m) => Double -> SFEvalT svIn svOut m ()
@@
 \hrule
-\caption{Evaluation combinators.}
+\caption{Evaluation combinators}
 \label{figure:evaluation_combinators}
 \end{figure}
-
-\subsection{Initialization and Running}
-To evaluate a signal function, one first constructs an initial state
-with the {\tt initSFEval} function. This function takes a set of
-handlers, a set of initial samples for the signal inputs, the original
-time (as a {\tt Double}) and the signal function to evaluate.
-
-The signal function replaces itself after each event occurrence or time
-step. The current signal function is maintained in the state. Calling
-{\tt push} causes the signal function to react to the supplied event 
-occurrence, and events which subsequently occur on the signal function's
-output are passed to the supplied handlers. Calling {\tt update} updates
-a set of new signal inputs. This set is nullified after it is passed in
-along with a new time by calling {\tt step}. This allows various
-components of a system to update individual signal inputs in a manner which is
-decoupled from the sample timer.
-
-When a monadic evaluation action is passed to the {\tt runSFEvalT} function,
-along with a state, the result is an action in the underlying monad. This
-action, when run in the underlying monad ({\tt IO}, for example) produces
-a pair of an output value and a new state. This new state is what should be 
-supplied to the next invocation of {\tt runSFEvalT}.
-
-\subsection{Monad Lifting}
-
-{\tt SFEvalT} is not only a monad, but a monad transformer. This means that
-it is parameterized by another monad, and that "running" an {\tt SFEvalT}
-action produces an action in this underlying monad. It means also that actions
-from this underlying monad may be ``lifted'' into actions of {\tt SFEvalT}.
-So if the {\tt IO} monad is the underlying monad, {\tt IO} actions can be 
-used to obtain input and arrange for timing. The underlying monad is also the
-monad whose actions the output handlers produce, which provides the means of
-actuating the output of the signal function.

### 2012-10-25 Docs/Thesis/system_design_and_interface.tex
Full initial draft of system design chapter

@@
 
+\subsection{Feedback}
+\label{subsection:System_Design_and_Interface-Combinators-Feedback}
+
+It is often useful for a signal function to receive a portion of its
+own output as input. This is especially useful when we have two
+signal functions which we would like to mutually interact. We cannot
+just serially compose them, we must also bring the output of the second
+back around to the first. Many signal-processing algorithms also depend
+on feedback. The combinator which provides this ability is shown
+in Figure~\ref{figure:feedback_combinator}.
+
+\begin{figure}
+\begin{code}
+loop ::    ((svIn :^: svLoop) :~> (svOut :^: svLoop))
+        -> svIn :~> svOut
+\end{code}
+\hrule
+\caption{Feedback combinator.}
+\label{figure:feedback_combinator}
+\end{figure}
+
+This combinator provides decoupling for signals
+(the input signal is the output signal at the previous time-step)
+but not events (event occurrences are supplied to the combinator immediately).
+This means that the programmer has the responsibility to ensure that feedback
+does not generate an infinite sequence of events in a single time-step.
+
+\subsection{Event-Specific Combinators}
+\label{subsection:System_Design_and_Interface-Combinators-Event_specific_combinators}
+
+Several combinators are provided for manipulating, suppressing, and generating events.
+Output events, in these combinators, are triggered by input events.
+
+Each of the combinators has an option variant and a list variant. The option variant
+produces an output event occurrence whenever the application of the supplied function
+to the input event produces a value. The list version produces an event occurrence for
+each of the elements of the output list. The combinators are shown in Figure~\ref{figure:event_specific_combinators}.
+
+\begin{figure}
+\begin{code}
+-- Apply the function to each input occurrence,
+-- and produce an occurrence for each Just.
+filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
+
+-- Apply the function to each input occurrence,
+-- and produce an occurrence for each list element
+filterList :: (a -> [b]) -> SVEvent a :~> SVEvent b
+
+-- Apply the function to the stored accumulator
+-- and the event occurrence, replacing the accumulator
+-- and possibly outputting an occurrence
+accumulate ::    (a -> b -> (Maybe c, a))
+              -> a
+              -> SVEvent b :~> SVEvent c
+
+-- Apply the function to the stored accumulator
+-- and the event occurrence, replacing the
+-- accumulator and outputting an event occurrence
+-- for each element of the list
+accumulateList ::    (a -> b -> ([c]))
+                  -> a
+                  -> SVEvent b :~> SVEvent c
+\end{code}
+\hrule
+\caption{Event-specific combinators.}
+\label{figure:event_specific_combinators}
+\end{figure}
+
+The filter combinators are stateless, and thus apply the function to only the new
+input value. They are useful for suppressing events, as well as for extracting one
+of multiple cases of a datatype. For instance, a splitter for events carrying
+{\tt Either}-valued occurrences could be written as:
+
+\begin{code}
+getLeft :: Either a b -> Maybe a
+getLeft (Left x) = Just x
+getLeft _ = Nothing
+
+getRight :: Either a b -> Maybe b
+getRight (Right x) = Just x
+getRight _ = Nothing
+
+split :: SVEvent (Either a b) :~> (SVEvent a :^: SVEvent b)
+split = copy >>> first (filter getLeft) >>> second (filter getRight)
+\end{code}
+
+The accumulate combinators are stateful, applying the supplied function
+to both the input value and an accumulator. This function has two results:
+the option or list of output event occurrence values, and the new value
+for the accumulator.
+
+The accumulator is useful when responses to multiple event occurrences
+(from one or more sources) must be coordinated. For instance, in the
+benchmark application (see Chapter~\ref{chapter:Evaluation_and_Comparisons},
+a table is maintained that allows knowledge from previous event occurrences
+(packets from a network switch) to be used in deciding where the present
+packet ought to go.
+
+\subsection{Joining}
+\label{subsection:System_Design_and_Interface-Combinators-Joining}
+
+The joining combinators provide the ability to combine two event
+streams, two signals, or a signal and an event stream. These
+combinators are shown in Figure~\ref{figure:joining_combinators}
+
+The {\tt union} combinator is a non-deterministic merge of event
+streams. Any event which occurs on either input will occur
+on the output. For simultaneous event occurrences, the order of occurrence
+is not guaranteed, but the occurrence itself is. This construct
+is also guaranteed to respect the relationship of event occurrences to sampling
+intervals.
+
+The {\tt combineSignals} combinator applies a binary function pointwise to two
+signals, and produces the result of this application as a third signal.
+
+The {\tt capture} combinator adds the last-sampled value of a signal at the time
+of an event occurrence to that event occurrence.
+
+These three combinators together provide the ability to combine elements of
+signal vectors. By combining these combinators, arbitrary signal vectors can
+be reduced.
+
+\begin{figure}
+\begin{code}
+union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
+combineSignals :: (a -> b -> c) -> (SVEvent a :^: SVEvent b) :~> SVEvent c
+capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent (b, a)
+\end{code}
+\hrule
+\caption{Joining combinators.}
+\label{figure:joining_combinators}
+\end{figure}
+
 \section{Evaluation}

### 2012-10-29 Docs/Thesis/implementation-signalfunctions.tex
Completed first re-draft of signal function implementation section

@@
 
-To encode signal functions, we must create a compositional way to express
-both the output and replacement responses to inputs. We must also have a means
-of specifying the initial input and output signal of the signal function.
+The design of signal functions specifies a family of types for the inputs and
+outputs of signal functions. Signal functions are not functions in the purest
+sense, however. They are not mappings from a single instance of their input
+type to a single instance of their output type. They must be implemented with
+respect to the temporal semantics of their inputs and outputs.
 
-The interface types for signal functions are given in 
-Figure~\ref{figure:signal-function-interface-types}.
+We therefore start by creating a set of concrete datatypes for the inputs and
+outputs of signal functions. These datatypes will be parameterized by the input
+and output types of the signal function, and will not be exposed to the user of
+the library. Rather, they will specify how data is represented during the
+temporal evaluation of signal functions.
 
-\begin{figure}
-\begin{minipage}[b]{0.5\linewidth}
-\begin{code}
--- Signal memories
-data SMemory p sv
+In order to create these concrete representations, we employ the Haskell type
+system extension known as Generalized Algebraic Datatypes, which we divert
+briefly to explain before exploring its use in the implementation.
 
--- Signal vector indices
-data SVIndex p sv
+We then describe how signal functions are implemented using these concrete
+types, along with higher-order functions and continuations.
 
--- Initialization types
-data Initialized
-data NonInitialized
-\end{code}
-\end{minipage}
-\hspace{0.5cm}
-\begin{minipage}[b]{0.5\linewidth}
-\begin{code}
--- Signal vectors
-data SVEmpty
-data SVSignal a
-data SVEvent a
-data SVAppend svLeft svRight
+\subsection{Generalized Algebraic Datatypes}
+\label{subsection:Implementation-Signal_Functions-Generalized_Algebraic_Datatypes}
+The first consideration is how to make use of the information encoded in the
+signal vectors used as the input and output types of signal functions. In order
+to parameterize over these types, we turn to an extension of the Haskell type
+system known as Generalized Algebraic Datatypes (GADTs).
 
--- Signal functions
-data SF init svIn svOut
-\end{code}
-\end{minipage}
-\vspace{\parskip}
-\hrule
-\caption{Interface types for signal functions}
-\label{figure:signal-function-interface-types}
-\end{figure}
-
-Our implementation makes use of two value representations of signal vectors.
-A signal memory is typed using a signal vector and carries values for a
-(possibly empty) subset of the points in the signal vector. A signal vector
-index is typed using a signal vector and carries a value for exactly one point
-in the signal vector. In order to define memories and indices over signal
-vectors, we make use of a type extension known as
-Generalized~Algebraic~Datatypes~(GADTs)~\cite{Cheney2003,Xi2003}, which is
-available in recent versions of the
-Glasgow~Haskell~Compiler~(GHC)~\cite{PeytonJones2006}. The key distinction
-of GADTs from standard Algebraic~Datatypes is that GADTs enable 
-{\em type refinement}. A type constructor in a GADT specifies the specific type
-with which each type parameter of the GADT's type constructor is instantiated.
-This type is defined in terms of the types of the constructor's data members.
-Any type variable which occurs in the type of a data member but does not occur
-in the type of the constructor is existentially quantified\footnote{Existential
-quantification, in the context of Algebraic Datatypes, is a means of ``hiding''
-type variables in data member types. A function which pattern matches on a
-constructor with existential quantification may not assume any instantiation of
-the existentially quantified variable, but only typeclass constraints specified
-in the constructor definition, and equality between identical type variables in
-data members of the same constructor.}. GADTs generalize Algebraic Datatypes 
-(ADTs) because an ADT may be translated to a GADT by simply using each unique
-type variable in the types of the data members of a constructor to instantiate
-the type parameter. As with ADTs, a type variable which appears in the type of a
-data constructor but not in any data members may be arbitrarily instantiated.
-
-The ability to specialize types is useful both for creating specialized
-constructors for unique cases of the type parameters (e.g. for
-optimization~\cite{Nilsson2005}) and for encoding properties at the type level
-using empty types.
-We employ the second use here.% and the first later in Section~\ref{subsection:Implementation-Signal_Functions-Representation} when we describe the representation of signal functions.
-
-Both signal memories and signal vector indices are GADTs:
+Algebraic Datatypes (ADTs) are the means by which new primitive types are
+introduced into Haskell programs. For instance, the following datatypes are
+declared in the Haskell Prelude (the Haskell module which is in scope for every
+Haskell program):
 
 \begin{code}
--- Signal memories
-data SMemory :: (* -> *) -> * -> * where
-  SMEmpty  :: SMemory p sv
-  SMSignal :: p a -> SMemory p (SVSignal a)
-  SMEvent  :: p a -> SMemory p (SVEvent a)
-  SMBoth   :: SMemory p svLeft -> SMemory p svRight 
-           -> SMemory p (SVAppend svLeft svRight)
-
--- Signal vector indices
-data SVIndex :: (* -> *) -> * -> * where
-  SVISignal :: p a -> SVIndex p (SVSignal a)
-  SVIEvent  :: p a -> SVIndex p (SVEvent a)
-  SVILeft   :: SVIndex p svLeft -> SVIndex p (SVAppend svLeft svRight)
-  SVIRight  :: SVIndex p svRight -> SVIndex p (SVAppend svLeft svRight)
-
--- ``Identity'' newtype
-newtype Id a = Id a
+data Bool    = True   | False
+data Maybe a = Just a | Nothing
 \end{code}
 
-The higher-kind type parameter {\tt p} in both type constructors is a
-convenience. It allows us to provide a type constructor which will be mapped
-over the types of signals and events in the signal vector to produce the actual
-type of values stored. This ability is used in
-Section~\ref{section:Implementation-Evaluation_Interface} (Evaluation Interface),
-permitting the memory type, in particular, to carry functions from the types of
-points in the signal vector to the monadic action used to handle the
-corresponding outputs.
-
-Next we consider the type for signal functions. A signal function must be able
-to have an initial signal input specified, and must be able to produce output
-signals, output events, and replace itself, based on time steps and input
-signals or event occurrences.
-
-The definition of the {\tt SF} datatype is:
+The first example is simply a set of alternatives, representing the values in
+Boolean algebras. These alternatives may be pattern-matched in a Haskell
+function either at the top level:
 
 \begin{code}
-data SF :: * -> * -> * -> * where
-  SF     :: (SMemory Id svIn -> (SMemory Id svOut, SF Initialized svIn svOut))
-         -> SF NonInitialized svIn svOut
-  SFInit :: (Double -> SMemory Id svIn 
-             -> (SMemory Id svOut, 
-                 [SVIndex Id svOut],
-                 SF Initialized svIn svOut))
-         -> (SVIndex Id svIn 
-             -> ([SVIndex Id svOu], SF Initialized svIn svOut))
-         -> SF Initialized svIn svOut
+boolToString :: Bool -> String
+boolToString True  = "Boolean True"
+boolToString False = "Boolean False"
 \end{code}
 
-The first constructor is used in the user-facing combinators to construct
-non-initialized signal functions. It contains a single function from an intial
-signal (which may contain empty values) to an initial output signal and
-an initialized signal function with the same input and output types.
+Or in a {\tt case} expression:
 
-The second constructor is used in the implementation of both the non-initialized
-and initialized signal functions to construct initialized signal functions. It
-takes two functions. The first is a function from a time step and an updated
-signal to an updated output signal, a set of output event occurrences, and a new
-signal function. The second is a function from an event occurrence to a set
-of event occurrences and a new signal function.
+\begin{code}
+boolToString :: Bool -> String
+boolToString b = "Boolean " ++ case b of
+                                 True  -> "True"
+                                 False -> "False"
+\end{code}
 
-\subsection{Combinators}
-\label{subsection:Implementation-Signal_Functions-Combinators}
-From the perspective of the library user, signal functions are built from a set
-of combinators which produce or modify signal functions. These combinators
-include a set of primitive signal functions, a set of lifting functions to
-produce signal functions from functions, a switching combinator,
-time and integration combinators, and a set of routing combinators. A list
-of the exported combinators is given in
-Figure~\ref{figure:signal_function_combinators}.
+The second example is a parameterized ADT, or type constructor. The actual type
+is constructed by filling in a concrete type for the variable {\tt a}.
+Polymorphic functions can fill in quantified variables for this type or a
+component of it. However, there are no restrictions on what types may be used.
+In the {\tt Just} data constructor, the use of the type variable indicates a
+parameter to the data constructor which must take the type filled in for {\tt a}.
+
+This allows the {\tt Maybe} type to generalize the pattern of an optional value.
+The type {\tt Maybe Int} represents 0 or 1 integer values, and the type
+{\tt Maybe String} represents 0 or 1 strings. In order to consume values of these
+optional types, we can pattern match\footnote{The underscore character {\tt\_} 
+represents an input to the pattern which we disregard. It is called a wildcard.}:
+
+\begin{code}
+stringMaybe :: String -> Maybe String -> String
+stringMaybe s Nothing = s
+stringMaybe _ (Just s)  = s
+\end{code}
+
+But since we can fill in a type variable for the parameter of the Maybe type
+constructor, we can generalize this function:
+
+\begin{code}
+maybe :: a -> Maybe a -> a
+maybe x Nothing  = x
+maybe _ (Just x) = x
+\end{code}
+
+GADTs permit us to specify what types fill in the type parameters for specific
+constructors. For instance, if we wish to build a tiny expression language,
+we can write:
+
+\begin{code}
+data Exp a = Const a | Plus (Exp a) (Exp a)
+\end{code}
+
+Let us assume for the moment that the Haskell addition function is typed:
+
+\begin{code}
+(+) :: Int -> Int -> Int
+\end{code}
+
+The type actually involves a typeclass constraint, but the point is that the
+function's type is not parametric.
+
+An attempt to write an evaluation function for our expression type is:
+
+\begin{code}
+eval :: Exp a -> a
+eval (Const x) = x
+eval (Plus x y) = eval x + eval y
+\end{code}
+
+But this will not typecheck, since we cannot assume that {\tt a} is {\tt Int},
+so the output type of {\tt (+)} does not match the output type of our function,
+and the input types to {\tt (+)} do not match the input types to our function.
+
+Let's try a slightly modified ADT:
+
+\begin{code}
+data Exp a = Const a | Plus (Exp Int) (Exp Int)
+\end{code}
+
+The code for our our evaluator is the same. Now the input types match, but
+the output type still does not.
+
+Here is our expression type as a GADT:
+\begin{code}
+data Exp a where
+  Const  ::  a -> Exp a
+  Plus   :: Exp Int -> Exp Int -> Exp Int
+\end{code}
+
+Now our evaluation function will typecheck. But why? The type of the {\tt Plus}
+constructor has had its {\em output} type restricted to {\tt Int}, and this
+allows us to assume, in the case where {\tt Plus} is matched, that the type
+{\tt a} has been restricted to {\tt Int}. Put another way, if a value of an
+{\tt Exp} type is not an {\tt Exp Int}, then the {\tt Plus} case will not occur.
+
+This capacity to constrain the output types of data constructors, and thus, to
+constrain the types of expressions in the scope of pattern matches of these data
+constructors, is called {\em type refinement}. We will make use of this ability
+to parameterize concrete datatypes over abstract type structures, rather than to
+permit typechecking in specific cases, but the principle is the same.
+
+\subsection{Concrete Representations of Signal Vectors}
+\label{subsection:Concrete_Representations_of_Signal_Vectors}
+
+In Chapter~\ref{chapter:System_Design_and_Interface} we present signal vectors
+as a set of types. In order to be completely principled, we should isolate these
+types into their own {\tt kind} (a sort of type of types), however, the Haskell
+extension for this was far from stable at the time this system was created.
+
+The types are therefore expressed in the system exactly as they were described
+in Chapter~\ref{chapter:System_Design_and_Interface}. (To refresh, see
+Figure~\ref{figure:signal_vector_types}.) The striking observation about these
+types is that they have {\em no data constructors}. There are no values which
+take these types.
+
+Instead, we will create concrete representations which are parameterized over
+these types. These concrete representations will be expressed as GADTs, allowing
+each data constructor of the representation to fill in a specific signal vector
+type for the parameter of the representation.
+
+To get our feet wet, we create a representation which carries a value for
+every {\tt SVSignal} leaf of a signal vector. In order to do this, we restrict
+each of our constructors to a single signal vector type. This ensures that the
+only way to represent a sample leaf is with the {\tt SVSample} constructor,
+which carries a value of the appropriate type. The datatype is shown in
+Figure~\ref{figure:signal_sample_datatype}.
 
@@
 \begin{code}
--- Simple signal functions
-identity :: SignalFunction NonInitialized sv sv
-constant :: a -> SignalFunction NonInitialized SVEmpty (SVSignal a)
-
--- Composition and routing
-(>>>) :: SignalFunction NonInitialized svIn svBetween
-      -> SignalFunction NonInitialized svBetween svOut
-      -> SignalFunction NonInitialized svIn svOut
-
-first :: SignalFunction NonInitialized svIn svOut
-      -> SignalFunction NonInitialized (SVAppend svIn sv) (SVAppend svOut sv)
-
-second :: SignalFunction NonInitialized svIn svOut
-       -> SignalFunction NonInitialized (SVAppend sv svIn) (SVAppend sv svOut)
-
-swap :: SignalFunction NonInitialized 
-          (SVAppend svLeft svRight) (SVAppend svRight svLeft)
-copy :: SignalFunction NonInitialized sv (SVAppend sv sv)
-drop :: SignalFunction NonInitialized sv SVEmpty
+data SVSample sv where
+  SVSample      ::    a
+                   -> SVSample (SVSignal a)
+  SVSampleEvent ::    SVSample (SVEvent a)
+  SVSampleEmpty ::    SVSample SVEmpty
+  SVSampleBoth  ::    SVSample svLeft
+                   -> SVSample svRight
+                   -> SVSample (SVAppend svLeft svRight)
 \end{code}
 \hrule
-\caption{The combinators exported by the signal function interface}
-\label{figure:signal_function_combinators}
+\caption{Datatype for signal samples.}
+\label{figure:signal_sample_datatype}
 \end{figure}
 
-The simplest combinator is the {\tt identity} signal function. As its name would
-suggest, the identity signal function simply replicates its input on its output.
-It is implemented as:
+So we now have a representation for the signal components of a signal vector.
+What about the event components? We want to represent event occurrences, each
+of which will correspond to at most one event in the vector. So a different
+representation is called for. In this case, there will be only three
+constructors. One constructor will represent an event leaf, and the other will
+represent a single value on the left or right side of the node ({\tt SVAppend}),
+ignoring all of the type structure on the other side. This representation
+describes a path from the root of the signal vector, terminating at an event
+leaf with a value.
 
+By pattern matching on the path constructors, we can determine which subvector
+of a signal vector an event occurrence belongs to, repeatedly refining it until
+we determine which event in the vector the occurrence corresponds to. The
+datatype for occurrences is shown in Figure~\ref{figure:event_occurrence_datatype}.
+
+\begin{figure}
 \begin{code}
-identity :: SF NonInitialized sv sv
-identity = SF (\mem -> (identityInit, mem))
+data SVOccurrence sv where
+  SVOccurrence ::    a
+                  -> SVOccurrence (SVEvent a)
+  SVOccLeft    ::    SVOccurrence svLeft
+                  -> SVOccurrence (SVAppend svLeft svRight)
+  SVOccRight   ::    SVOccurrence svRight 
+                  -> SVOccurrence (SVAppend svLeft svRight)
+\end{code}
+\hrule
+\caption{Datatype for event occurrences.}
+\label{figure:event_occurrence_datatype}
+\end{figure}
+
+We add one more representation for signals, in order to avoid uneccessary
+representations of the values of all signals when not all signals have changed
+their values. This representation allows us to represent the values of zero or
+more of the signals in a signal vector. To accomplish this, we replace the
+individual constructors for the {\tt SVEmpty} and {\tt SVEvent} leaves with %there is a better word, something about "useless but not really", for this
+a single, unconstrained constructor. This constructor can represent an arbitrary
+signal vector. We can use the constructor for signal vector nodes and the 
+constructor for sample leaves to represent the updated values, while filling
+in the unchanged portions of the signal vector with this general constructor.
+This datatype is shown in Figure~\ref{figure:signal_update_datatype}.
+
+\begin{figure}
+\begin{code}
+data SVDelta sv where
+  SVDeltaSignal  ::    a
+                    -> SVDelta (SVSignal a)
+  SVDeltaNothing ::    SVDelta sv
+  SVDeltaBoth    ::    SVDelta svLeft
+                    -> SVDelta svRight
+                    -> SVDelta (SVAppend svLeft svRight)
+\end{code}
+\hrule
+\caption{Datatype for signal updates.}
+\label{figure:signal_update_datatype}
+\end{figure}
+
+\subsection{Signal Function Implementation Structure}
+\label{subsection:Implementation-Signal_Functions-Signal_Function_Implementation_Structure}
+
+We now have concrete datatypes for an implementation to operate on. Our next
+task is to represent transformers of temporal data, which themselves may change
+with time. The common approach to this task is sampling, in which a program
+repeatedly checks for updated information, evaluates it, updates some state,
+and produces an output. This is the essence of pull-based evaluation.
+
+Another strategy is notification, in which the program exposes an interface
+which the source of updated information may invoke. This is a repeated entry
+point to the program, which causes the program to perform the same tasks
+listed above, namely, evaluate the updated information, update state, and
+produce output. The strategy of notification as opposed to repeated checking is
+the essence of push-based evaluation.
+
+Signal functions are declarative objects, and not running processes. They have
+no way to invoke sampling themselves. They can, however, expose separate
+interfaces for when sampling is invoked, and when they are notified of an event
+occurrence. This creates two control paths through a signal function. One of
+these control paths is intended to be invoked regularly and frequently with
+updates to the time and sample values, and the other is intended to be invoked
+only when an event occurs. The benefit of separating these control paths is that
+events are no longer defined in terms of sampling intervals, and need not even
+be considered in sampling, except when they are generated by a condition on a
+sample. On the other hand, events can be responded to even if the time has not
+yet come for another sample, and multiple events can be responded to in a single
+sampling interval.
+
+We represent signal functions as a datatype with two constructors.
+The signal functions that a user will compose are non-initialized signal
+functions. They must be provided with an initial set of input signals
+(corresponding to time zero). When provided with this input, they produce their
+time-zero output, and an initialized signal function. The datatype is shown in
+Figure~\ref{figure:signal_function_datatype}
+
+\begin{figure}
+\begin{code}
+data Initialized
+
+data NonInitialized
+
+data SF init svIn svOut where
+  SF     ::    (SVSample svIn 
+                  -> (SVSample svOut,
+                      SF Initialized svIn svOut)) 
+            -> SF NonInitialized svIn svOut
+  SFInit ::    (Double 
+                  -> SVDelta svIn
+                  -> (SVDelta svOut,
+                      [SVOccurrence svOut],
+                      SF Initialized svIn svOut)) 
+            -> (SVOccurrence svIn
+                  -> ([SVOccurrence svOut],
+                      SF Initialized svIn svOut))
+            -> SF Initialized svIn svOut
+\end{code}
+\hrule
+\caption{Datatype and empty types for signal functions.}
+\label{figure:signal_function_datatype}
+\end{figure}
+
+Initialized signal functions carry two continuations. The first continuation
+takes a time differential and a set of signal updates, and returns a set of
+signal updates, a collection of event occurrences, and a new initialized signal
+function of the same type. This is the continuation called when sampling.
+
+The second continuation takes an event occurrence, and returns a collection of
+event occurrences and a new signal function of the same type. This continuation
+is only called when there is an event occurrence to be input to the signal
+function.
+
+Note that each of these continuations uses one or more of the concrete
+representations of signal vectors, and applies the type constructor for the
+representation to the input or output signal vector for the signal function.
+
+\subsection{Implementation of Signal Function Combinators}
+\label{subsection:Implementation-Signal_Functions-Implementation_of_Signal_Function_Combinators}
+
+Having specified a datatype for signal functions, we must now provide
+combinators which produce signal functions of this type. Each combinator's
+implementation must specify how it is initialized, how it samples its input, and
+how it responds to event occurrences.
+
+We will not detail every combinator here, but we will discuss each of the
+implementation challenges encountered.
+
+The simplest combinator is the {\tt identity} signal function. This signal
+function simply passes all of its inputs along as outputs. The initialization
+function simply passes along the received sample and outputs the initialized
+version of the signal function. The initialized version of the input is similar,
+but is self-referential. It replaces itself with itself after every signal
+function or event. The implementation is shown in Figure~\ref{figure:identity_implementation}.
+
+\begin{figure}
+\begin{code}
+identity :: sv :~> sv
+identity =
+  SF (\initSample -> (initSample, identityInit))
 
 identityInit :: SF Initialized sv sv
-identityInit = let sf = SFInit (\_ mem -> (mem, [], sf))
-                               (\idx -> ([idx], sf))
-               in sf
+identityInit =
+  SFInit (\dt sigDelta -> (sigDelta, [], identityInit))
+         (\evtOcc -> ([evtOcc], identityInit))
 \end{code}
+\hrule
+\caption{Implementation of the {\tt identity} combinator.}
+\label{figure:identity_implementation}
+\end{figure}
 
-Only the first function is exported. In general, we use the convention that
-the initialized version of a signal function is named with the signal function
-name with {\tt Init} appended, and is not exported as part of the interface.
+In order for our primitive signal functions to be useful, we need a means of
+composing them. Serial composition creates one signal function from two, by
+using the output of one as the input of the other. The serial composition
+combinator is styled {\tt (>>>)}. The implementation of this operator is one
+place where the advantage of responding to events independently from signal
+samples becomes clear. The implementation is shown in Figure~\ref{figure:serial_composition_implementation}.
+
+\begin{figure}
+\begin{code}
+(>>>) ::    (svIn :~> svBetween) 
+         -> (svBetween :~> svOut)
+         -> (svIn :~> svOut)
+(SF sigSampleF1) >>> (SF sigSampleF2) =
+  SF (\sigSample -> let (sigSample', sfInit1) = sigSampleF1 sigSample
+                        (sigSample'', sfInit2) = sigSampleF2 sigSample'
+                    in (sigSample'', composeInit sfInit1 sfInit2))
+
+composeInit ::     SF Initialized svIn svBetween
+                -> SF Initialized svBetween svOut
+                -> SF Initialized svIn svOut
+composeInit (SFInit dtCont1 inputCont1) sf2@(SFInit dtCont2 inputCont2) =
+  SFInit
+    (\dt sigDelta -> 
+       let (sf1MemOutput, sf1EvtOutputs, sf1New) = dtCont1 dt sigDelta
+           (sf2MemOutput, sf2EvtOutputs, sf2New) = dtCont2 dt sf1MemOutput
+           (sf2EvtEvtOutputs, sf2Newest) = applySF sf2New sf1EvtOutputs
+       in (sf2MemOutput,
+           sf2EvtOutputs ++ sf2EvtEvtOutputs,
+           composeInit sf1New sf2Newest)
+    )
+    (\evtOcc -> 
+      let (sf1Outputs, newSf1) = inputCont1 evtOcc
+          (sf2FoldOutputs, newSf2) = applySF sf2 sf1Outputs
+      in (sf2FoldOutputs, composeInit newSf1 newSf2)   
+    )
+
+applySF ::    SF Initialized svIn svOut
+           -> [SVOccurrence svIn]
+           -> ([SVOccurrence svOut],
+               SF Initialized svIn svOut)
+applySF sf indices =
+  foldr (\evtOcc (changes, SFInit _ changeCont) ->
+           let (newChanges, nextSF) = changeCont evtOcc
+               in (newChanges ++ changes, nextSF))
+        ([], sf)
+        indices
+\end{code}
+\hrule
+\caption{Implementation of serial composition.}
+\label{figure:serial_composition_implementation}
+\end{figure}
+
+The implementation of the non-initialized version is simple. It takes two
+non-initialized signal functions, and returns the sample output of applying the
+second to the sample output of the first, along with an initialized signal
+function constructed by applying the initialized version of the combinator to
+the initialized signal functions given by the inputs.
+
+The initialized version constructs two continuations. The first receives a time
+and signal update input, and applies the signal continuation of the first
+signal function to it. The output is fed, along with the time update, into the
+second signal function, first the signal update, and then events. The output
+of the second, along with the accumulated updated signal functions with the
+initialized serial composition recursively applied, is returned as output.
+
+The event continuation is similar but simpler, as there is no need to deal
+with signal updates.
+
+The switch combinator is the means of introducing reactivity into a signal
+function. This combinator allows a signal function to replace itself by
+producing an event occurrence. The switch combinator stores the full sample
+for its input vector (which is identical to the input vector of the supplied
+signal function) to initialize the new signal function. This also demands that
+it continue wrapping the signal function until the next sampling interval, so
+that it can actuate the output of initialization as a replacement signal. This
+has some performance implications, which are discussed in
+Chapter~\ref{chapter:Evaluation_and_Comparisons}. The implementation is shown in
+Figure~\ref{figure:switch_implementation}.
+
+\begin{figure}
+\begin{code}
+switch :: (svIn :~> (svOut :^: SVEvent (svIn :~> svOut))) -> svIn :~> svOut
+switch (SF sigSampleF) =
+  SF (\sigSample -> let (sigSampleSF, sf) = sigSampleF sigSample
+                        (sigSampleOut, _) = splitSample sigSampleSF
+                    in (sigSampleOut, switchInit sigSample sf))
+
+-- Pre-occurrence
+switchInit ::    SVSample svIn
+              -> SF Initialized svIn 
+                    (SVAppend svOut 
+                     (SVEvent (SF NonInitialized svIn svOut)))
+              -> SF Initialized svIn svOut
+switchInit inputSample sf@(SFInit timeCont changeCont) =
+  SFInit (\dt sigDelta ->
+            let (sigDeltaSF, occsSF, newSF) = timeCont dt sigDelta
+                newInputSample = updateSample inputSample sigDelta
+                (sigDeltaOut, _) = splitDelta sigDeltaSF
+                (occsOut, occsSwitch) = splitOccurrences occsSF
+                (outputDelta, nextSF) =
+                  maybe (sigDeltaOut, switchInit newInputSample newSF)
+                    (\(SF sigSampleF) ->
+                       let (outputSample, switchSF) =
+                             sigSampleF newInputSample
+                           outputDelta = sampleDelta outputSample
+                       in (outputDelta, switchSF))
+                  $ occurrenceListToMaybe occsSwitch
+            in (outputDelta, occsOut, nextSF))
+         (\evtOcc ->
+            let (occsSF, newSF) = changeCont evtOcc
+                (occsOut, occsSwitch) = splitOccurrences occsSF
+                nextSF = maybe (switchInit inputSample newSF) 
+                               (\(SF sigSampleF) ->
+                                  let (sampleOut, switchSF) =
+                                        sigSampleF inputSample
+                                  in switchWait sampleOut switchSF)
+                         $ occurrenceListToMaybe occsSwitch
+            in (occsOut, nextSF))
+
+switchWait ::    SVSample svOut
+              -> SF Initialized svIn svOut
+              -> SF Initialized svIn svOut
+switchWait outputSample sf@(SFInit timeCont changeCont) =
+  SFInit (\dt sigDelta -> let (sigDeltaOut, outOccs, newSF) =
+                                timeCont dt sigDelta
+                          in (sampleDelta $
+                                updateSample outputSample sigDeltaOut,
+                              outOccs,
+                              newSF))
+         \evtOcc -> let (outOccs, newSF) = changeCont evtOcc
+                    in (outOccs, switchWait outputSample newSF))
+\end{code} 
+\hrule
+\caption{Implementation of reactivity.}
+\label{figure:switch_implementation}
+\end{figure}
+
+Most of the routing combinators are simple to implement. The only task is to add
+remove, or replace routing constructors on signal updates and event occurrences.
+Since these signal functions are stateless and primitive, they can simply
+return themselves as their replacements. The {\tt swap} combinator is shown as
+an example in Figure~\ref{figure:swap_implementation}
+
+\begin{figure}
+\begin{code}
+-- T.swap is imported from Data.Tuple
+swap :: (sv1 :^: sv2) :~> (sv2 :^: sv1)
+swap =
+  SF ((, swapInit) . 
+      uncurry combineSamples .
+      T.swap . splitSample)
+
+swapInit :: SF Initialized (SVAppend sv1 sv2) (SVAppend sv2 sv1)
+swapInit =
+  SFInit (flip (const .
+                (, [], swapInit) .
+                uncurry combineDeltas .
+                T.swap . splitDelta))
+          (\evtOcc ->
+             (case chooseOccurrence evtOcc of
+                Left lOcc  -> [occRight lOcc]
+                Right rOcc -> [occLeft rOcc], swapInit))
+
+\end{code}
+\hrule
+\caption{Swap routing combinator implementation.}
+\label{figure:swap_implementation}
+\end{figure}
+
+The {\tt first} and {\tt second} combinators are not quite as simple, as they
+must transform a provided signal function. For signal changes, they must split
+the set of changes into those which will be passed to the signal function and
+those which will be simply passed along to the output, and then recombine them
+on the other side. For event occurrences, the occurrence must be pattern-matched
+to determine whether to call the event continuation from the provided signal
+function or passed through, and output event occurrences must have the suitable
+routing constructor re-applied. In any case, when a continuation has been
+applied, the combinator must be recursively applied to the new signal function.
+The implementation of {\tt first} is shown in Figure~\ref{figure:first_implementation}.
+The implementation of {\tt second} is the obvious reversal of this
+implementation.
+
+\begin{figure}
+\begin{code}
+first :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
+first (SF sigSampleF) =
+  SF (\sigSample -> let (leftSample, rightSample) = splitSample sigSample
+                        (leftSampleOut, sf) = sigSampleF leftSample
+                    in (combineSamples leftSampleOut rightSample,
+                        firstInit sf))
+
+firstInit ::    SF Initialized svIn svOut
+             -> SF Initialized (SVAppend svIn sv) (SVAppend svOut sv)
+firstInit (SFInit timeCont inputCont) = 
+  let firstInitSF =
+    SFInit (\dt sigDelta ->
+               let (input, rightOutput) = splitDelta sigDelta
+                   (sigDeltaOutput, evtOutput, sf1New) = timeCont dt input
+               in (combineDeltas sigDeltaOutput rightOutput,
+                   map occLeft evtOutput,
+                   firstInit sf1New))
+           (\evtOcc ->
+              case chooseOccurrence evtOcc of
+                Left  lChange -> let (changes, sf) = inputCont lChange
+                                 in (map occLeft changes, firstInit sf)
+                Right rChange -> ([occRight rChange], firstInitSF))
+  in firstInitSF
+\end{code}
+\hrule
+\caption{Implemenation of {\tt first} combinator.}
+\label{figure:first_implementation}
+\end{figure}
+
+The looping feedback combinator is particularly tricky. As it is currently
+implemented, the initial sample for the right side of the input signal vector to
+the supplied function is the right side of the output sample. This is acceptable,
+given Haskell's non-strict evaluation strategy, but it is necessary that the
+right side of the signal function's output not be immediately dependent on its
+input. The feedback combinator makes use of Haskell's lazy evaluation to
+feed events back into the combinator, and stores signal updates until the next
+sample. Signal samples are thus automatically decoupled after initialization.
+The implementation, which makes extensive use of the recursive nature of
+Haskell's {\tt let} construct, is shown in Figure~\ref{figure:loop_implementation}.
+
+\begin{figure}
+\begin{code}
+loop :: (svIn :^: svLoop) :~> (svOut :^: svLoop) -> svIn :~> svOut
+loop (SF sigSampleF) =
+  SF (\sigSample -> 
+        let (sigSampleOut, sfInit) =
+              sigSampleF (combineSamples sigSample sigSampleOutRight)
+            (sigSampleOutLeft, sigSampleOutRight) =
+              splitSample sigSampleOut
+        in (sigSampleOutLeft, loopInit deltaNothing sfInit))
+
+loopInit ::    SVDelta svLoop
+            -> SF Initialized (SVAppend svIn svLoop) (SVAppend svOut svLoop)
+            -> SF Initialized svIn svOut
+loopInit loopDelta (SFInit timeCont changeCont) =
+  SFInit (\dt sigDelta ->
+            let (sigDeltaOut, occsOut, newSF) =
+                  timeCont dt (combineDeltas sigDelta loopDelta)
+                (occsOutput, loopOccs) = splitOccurrences occsOut
+                (occsOutput', loopOccs') = splitOccurrences foldOccs
+                (foldOccs, newSF') =
+                  applySF newSF (map occRight (loopOccs ++ loopOccs'))
+                (outputDelta, newLoopDelta) = splitDelta sigDeltaOut
+            in (outputDelta,
+                occsOutput ++ occsOutput',
+                loopInit newLoopDelta newSF'))
+         (\change -> let (occsOut, newSF) = changeCont $ occLeft change
+                         (occsOutput, loopOccs) =
+                           splitOccurrences occsOut
+                         (occsOutput', loopOccs') =
+                           splitOccurrences foldOccs
+                         (foldOccs, newSF') =
+                           applySF newSF
+                             (map occRight (loopOccs ++ loopOccs'))
+                     in (occsOutput ++ occsOutput',
+                         loopInit loopDelta newSF'))
+\end{code}
+\hrule
+\caption{Implementation of feedback.}
+\label{figure:loop_implementation}
+\end{figure}
+
+The {\tt filter} combinators are simple to implement. Their sampling
+continuation is superfluous, and the event continuation merely applies the
+supplied function, and constructs an output list based on the result.
+
+The {\tt accumulate} combinators are implemented in terms of the {\tt filter}
+and {\tt switch} combinators, as shown in
+Figure~\ref{figure:accumulate_implementation}.
+
+\begin{figure}
+\begin{code}
+-- | Accumulate over event occurrences
+accumulate :: (a -> b -> (Maybe c, a)) -> a -> SVEvent b :~> SVEvent c
+accumulate f a = acc a
+  where acc a = switch (pureEventTransformer (f a) >>>
+                        copy >>>
+                        first (pureEventTransformer fst >>> filter id) >>>
+                        second (pureEventTransformer (acc . snd)))
+
+-- | Accumulate over event occurrences, with lists of event outputs
+accumulateList :: (a -> b -> ([c], a)) -> a -> SVEvent b :~> SVEvent c
+accumulateList f a = acc a
+  where acc a = switch (pureEventTransformer (f a) >>>
+                        copy >>>
+                        first (pureEventTransformer fst >>> filterList id) >>>
+                        second (pureEventTransformer (acc . snd)))
+\end{code}
+\hrule
+\caption{Implementation of event accumulators.}
+\label{figure:accumulate_implementation}
+\end{figure}
+
+The implementation of the joining combinators is simple. The {\tt union}
+combinator simply passes along every event occurrence it receives on either
+input, stripping off the left and right combinator. This is acceptable since we
+do not insist on a total ordering of events, or an event time resolution greater
+than the sampling rate. The {\tt combineSignals} combinator maintains the values
+of both signals, and applies the combination function whenever one is updated.
+The {\tt capture} combinator maintains the input signal value, and adds it to
+each event occurrence.
+
+Time dependence is introduced by the {\tt time}, {\tt delay}, and {\tt integrate}
+combinators. The time combinator simply sums the time updates and provides the
+sum as a signal output. The {\tt delay} combinator keeps a table of events
+which have come in, along with their schedule occurrence time, and produces
+them as output when time advances far enough. The integrate combinator performs
+rectangle-rule integration on signal samples with respect to time.
+
+The implementation strategy leaves room for optimizations. In particular, an
+additional constructor for time-independent signal functions would allow
+portions of a signal function to forgo evaluation during time steps unless they
+had signal updates. Optimizations in the style of Yampa, observed by keeping
+an updated for the signal function and pattern-matching on it when switching,
+might further improve performance. Nevertheless, this implementation performs
+quite well as it currently exists, as we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparison}.

### 2012-10-29 Docs/Thesis/implementation.tex
Completed first re-draft of signal function implementation section

@@
 
-Having established the semantics of the reactive system, I present a purely 
+Having established the design and semantics of the reactive system, I present a purely 
 functional implementation. For this implementation, I will make significant use
-of extensions to the Haskell type system. Properly used, these extensions allow
-greater expressive flexibility to define our types by permitting us to encode
-additional properties of values in their types.
+of extensions to the Haskell type system. These extensions provide mechanisms for expressing
+abstractions not easily expressible in a standard Hindley-Milner type system, while maintaining
+the type soundness of Haskell.
 

### 2012-10-29 Docs/Thesis/abstract.tex
Edits

@@
-Functional Reactive Programming (FRP) is a promising class of abstractions for encoding interactive and time-dependent programs in
-functional languages which has proven difficult to implement efficiently. Signal-function FRP is a subclass of FRP which does not provide
-direct access to the time-varying values by the programmer, but rather provides signal functions as first-class objects in the
-program. FRP systems provide values defined at all points in time ({\em behaviors} or {\em signals}) and values defined
-at countably many points in time ({\em events}).
+Functional Reactive Programming (FRP)is a promising class of abstractions for
+interactive programs. FRP systems provide values defined at all points in time
+({\em behaviors} or {\em signals}) and values defined at countably many points
+in time ({\em events}) as abstractions.
 
-All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation for both events and signals,
-producing output from the FRP system whenever the consumer of the output is ready. This greatly simplifies the implementation of signal-function
-FRP systems, but leads to inefficient and wasteful evaluation of the FRP system.
+Signal-function FRP is a subclass of FRP which does not provide direct access to
+the time-varying values to the programmer, but rather provides signal functions,
+which are reactive transformers of signals and events, as first-class objects in the program.
 
-In contrast, an input-driven or ``push-based'' system evaluates the network whenever new input is available. This frees the system from
-evaluating the network when nothing has changed, and then only components necessary to react to the input are reevaluated.
-This form of evaluation has been applied to events in standard FRP systems~\cite{Elliott2009} but not in signal-function FRP systems.
+All signal-function implementations of FRP to date have utilized demand-driven
+or ``pull-based''  evaluation for both events and signals, producing output from
+the FRP system whenever the consumer of the output is ready. This greatly
+simplifies the implementation of signal-function FRP systems, but leads to
+inefficient and wasteful evaluation of the FRP system when this strategy is
+employed to evaluate events.
 
-I describe the implementation of a push-based signal-function FRP system. The semantics of the system are discussed, and its performance
-and expressiveness for practical examples of interactive programs are compared to existing signal-function FRP systems through the
-implementation of a networking application.
+In contrast, an input-driven or ``push-based'' system evaluates the network
+whenever new input is available. This frees the system from evaluating the
+network when nothing has changed, and then only components necessary to react to
+the input are reevaluated. This form of evaluation has been applied to events in
+standard FRP systems~\cite{Elliott2009} but not in signal-function FRP systems.
+
+I describe the implementation of a signal-function FRP system which applies
+pull-based evaluation to signals and push-based evaluation to events (a
+"push-pull" system). The semantics of the system are discussed, and its
+performance and expressiveness for practical examples of interactive programs
+are compared to existing signal-function FRP systems through the implementation
+of a networking application.

### 2012-10-29 Docs/Thesis/background.tex
Fix figure captions
Edits

@@
 as output and/or take functions as input. We say that functions are {\em first-class values} in a language.
+
+In contrast, the more popular imperative model of programming takes statements,
+or actions which modify the state of the world, as the primary building blocks
+of programs. Such programs have inherently sequential control flow, and require
+reasoning about side-effects. They are thus inherently resistant to
+compositional reasoning.
 
@@
 function may operate on events, signals, or some combination of the two.}, which are time-dependent values defined at every point in continuous time, and events, which are 
-values defined at countably many points in time. An FRP system will provide functions to manipulate events and signals and to react
+values defined at countably many points in time. An FRP system will provide combinators to manipulate events and signals and to react
 to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
 based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in FRP languages
-should of course be efficiently executable. This has proven to be the main challenge in implementing FRP.
+should of course be efficiently executable, but this has proven to be the main challenge in implementing FRP.
 
@@
 produce values of these types. The easiest semantic definition for these types
-is given in Figure~\ref{figure:classic_frp_semantic_types}\footnote{The list constructor {\tt []} should be considered to contain infinite as well as finite lists.}.
+is given in Figure~\ref{figure:classic_frp_semantic_types}\footnote{The list type constructor {\tt []} should be considered to contain infinite as well as finite lists.}.
 
@@
 \hrule
-\caption{Semantic types for Classic FRP}
+\caption{Semantic types for Classic FRP.}
 \label{figure:classic_frp_semantic_types}
@@
 \caption{An obvious, yet inefficient and problematic, implementation of
-Classic FRP}
+Classic FRP.}
 \label{figure:classic_frp_simple_implementation}
@@
 
-Obvious problems include the lack of ordering enforcement on event occurrences,
-the necessity of waiting for the first event occurrence in a switch, and the
-necessity of maintaining the full history of a behavior for evaluation.
+There are several obvious problems with this implementation of course, but it
+suffices to show the intuition behind the classic FRP model. Problems in this
+implementation which are addressed by real implementations include the necessity
+of waiting for event occurrences, the necessity of maintaining the full history
+of a behavior in memory, and the lack of an enforced order for event
+occurrences.
 
@@
 FRP implementation for the Java Beans framework. Frapp\'{e} built on the notion
-of events and ``bound properties'' in the Beans framework, providing abstract
-interfaces for FRP events and behaviors, and combinators as concrete classes
-implementing these interfaces. The evaluation of Frapp\'{e} used Bean components
-as sources and sinks, and the implementation of Bean events and bound properties
-to propagate changes to the network~\cite{Courtney2001-2}.
+of events\footnote{These are semantic objects to which callbacks may be added,
+and not events in the FRP sense, though they are related.} and ``bound
+properties'' in the Beans framework, providing abstract interfaces for FRP
+events and behaviors, and combinators as concrete classes implementing these
+interfaces. The evaluation of Frapp\'{e} used Bean components as sources and
+sinks, and the implementation of Bean events and bound properties to propagate
+changes to the network~\cite{Courtney2001-2}.
 

### 2012-10-29 Docs/Thesis/introduction.tex
Edits

@@
 web browsers, web applications, video games, multimedia authoring and playback tools, operating system shells and kernels, servers,
-robotic control programs, and many others. This ability is called reactivity.
+robotic control programs, and many others. This attribute of a program is called
+reactivity.
 
@@
 
-The first challenge was addressed by representing FRP programs, not as compositions of signals and events, but as compositions of
-transformers of behaviors and events called {\em signal functions}. By not permitting direct access to behaviors or events, but
-representing them implicitly, signal functions prohibit accessing past or future time values, only allowing access to values in the closure of the signal function
-and current input values. Signal function FRP programs are written by composing signal functions, and are evaluated using a function which provides input to the
-signal function and acts on the output values. This evaluation function is elided in existing literature on signal functions. 
-A further advantage of signal function FRP programs is that they are more compositional, since additional signal functions may be composed
-on the input or output side of a signal function, rather than simply the output side in classic (non-signal-function) FRP systems.
+The first challenge was addressed by representing FRP programs, not as
+compositions of signals and events, but as compositions of transformers of
+behaviors and events called {\em signal functions}. By not permitting direct
+access to behaviors or events, but representing them implicitly, signal
+functions prohibit accessing past or future time values, only allowing access to
+values in the closure of the signal function and current input values. Signal
+function FRP programs are written by composing signal functions, and are
+evaluated using a function which provides input to the signal function and acts
+on the output values. This evaluation function is elided in existing literature
+on signal functions, but we provide a documented and improved interface for
+evaluating signal functions. A further advantage of signal function FRP programs
+is that they are more composable, since additional signal functions may be
+composed on the input or output side of a signal function, rather than simply
+the output side in classic (non-signal-function) FRP systems.
 
-The second challenge was addressed for classic FRP by the creation of a hybrid push-pull FRP system~\cite{Elliott2009}. This system relied
-on the distinction between reactivity and time dependence to decompose behaviors into phases of constant or simply time-dependent values,
-with new phases initiated by event occurrences. However, this implementation did not address concerns of causality or compositionality.
-Further, the existence of events as first-class values in classic FRP forced the use of an impure and resource-leaking technique to compare
-events to determine which had the first occurrence after the current time. This technique made use of the Haskell {\tt unsafePerformIO} function
-to fork two threads to compare the improving values of the next occurrence time for two events being merged. Further, since the classic FRP
-interface permits access to only the output of a behavior or event, and is always bound to its implicit inputs, the best the push-based implementation
-could do is to block when no computation needs to be performed. The computation cannot be suspended entirely as a value representation.
+The second challenge was addressed for classic FRP by the creation of a hybrid
+push-pull FRP system~\cite{Elliott2009}. This system relied on the distinction
+between reactivity and time dependence to decompose behaviors into phases of
+constant or simply time-dependent values, with new phases initiated by event
+occurrences. However, this implementation did not address concerns of causality
+or compositionality. Further, the existence of events as first-class values in
+classic FRP forced the use of an impure and resource-leaking technique to
+compare events when merging, to determine which had the first occurrence after
+the current time. This technique made use of the Haskell {\tt unsafePerformIO}
+function to fork two threads to compare the improving values of the next
+occurrence time for two events being merged. Further, since the classic FRP
+interface permits access to only the output of a behavior or event, and is
+always bound to its implicit inputs, the best the push-based implementation
+could do is to block when no computation needs to be performed. The computation
+cannot be suspended entirely as a value representation.
 
-To address both challenges in a FRP implementation, it seems desirable to combine the approaches, creating a push-based
-signal-function FRP system. The current implementation approach for signal-function FRP is inherently pull-based~\cite{Nilsson2002}, but
-recent work on N-ary FRP~\cite{Sculthorpe2011}, a variant of signal-function FRP which enumerates inputs and distinguishes between events and behaviors, provides a representation which suggests
-a method of push-based evaluation. In previous implementations of FRP, events were simply
-option-valued signals. This approach has the advantage of being simple to implement in a pull-based setting, but does not have an obvious
-semantics for event merging, and necessitates pull-based evaluation, since there is no way to separate the events which may be pushed from
-the truly continuous values which must be repeatedly sampled.
+To address both challenges in a FRP implementation, we to combine the approaches
+and demonstrate a push-based signal-function FRP system. The current
+implementation approach for signal-function FRP is inherently
+pull-based~\cite{Nilsson2002}, but recent work on N-ary
+FRP~\cite{Sculthorpe2011}, a variant of signal-function FRP which enumerates
+inputs and distinguishes between events and behaviors, provides a representation
+which suggests a method of push-based evaluation. In previous implementations of
+FRP, events were simply option-valued signals. This approach has the advantage
+of being simple to implement in a pull-based setting. However, it does not have
+an obvious semantics for event merging. Further, it necessitates pull-based
+evaluation, since there is no way to separate the events which may be pushed
+from the truly continuous values which must be repeatedly sampled.
 
-This thesis describes the design and implementation of Time-Flies, a push-pull signal-function FRP system. This system is presented as a library
+This thesis describes the design and implementation of TimeFlies, a push-pull signal-function FRP system. This system is presented as a library
 of combinators for constructing and evaluating FRP programs.

### 2012-10-30 Docs/Thesis/abstract.tex
Thesis draft edits

@@
-Functional Reactive Programming (FRP)is a promising class of abstractions for
+Functional Reactive Programming (FRP) is a promising class of abstractions for
 interactive programs. FRP systems provide values defined at all points in time
@@
 Signal-function FRP is a subclass of FRP which does not provide direct access to
-the time-varying values to the programmer, but rather provides signal functions,
-which are reactive transformers of signals and events, as first-class objects in the program.
+time-varying values to the programmer, but instead provides signal functions,
+which are reactive transformers of signals and events, as first-class objects in
+the program.
 
@@
 network when nothing has changed, and then only components necessary to react to
-the input are reevaluated. This form of evaluation has been applied to events in
-standard FRP systems~\cite{Elliott2009} but not in signal-function FRP systems.
+the input are re-evaluated. This form of evaluation has been applied to events
+in standard FRP systems~\cite{Elliott2009} but not in signal-function FRP
+systems.
 
-I describe the implementation of a signal-function FRP system which applies
-pull-based evaluation to signals and push-based evaluation to events (a
+I describe the design and implementation of a signal-function FRP system which
+applies pull-based evaluation to signals and push-based evaluation to events (a
 "push-pull" system). The semantics of the system are discussed, and its

### 2012-10-30 Docs/Thesis/background.tex
Thesis draft edits

@@
 or actions which modify the state of the world, as the primary building blocks
-of programs. Such programs have inherently sequential control flow, and require
-reasoning about side-effects. They are thus inherently resistant to
-compositional reasoning.
+of programs. Such programs have sequential control flow, and require reasoning
+about side-effects. They are thus inherently resistant to compositional
+reasoning.
 
@@
 interactive animations. Fran represented behaviors as two sampling functions,
-one from a time to a value and a new behavior (so that history may be discared),
+one from a time to a value and a new behavior (so that history may be discarded),
 and the other from a time interval (a lower and upper time bound) to a value
@@
 
-The first implementation of FRP outside the Haskell language was Frapp\'{e}, an
-FRP implementation for the Java Beans framework. Frapp\'{e} built on the notion
+The first implementation of FRP outside the Haskell language was Frapp\'{e}, 
+implemented in the Java Beans framework. Frapp\'{e} built on the notion
 of events\footnote{These are semantic objects to which callbacks may be added,
@@
 could not be known beforehand) from time-dependence. This gives rise to {\em
-reactive normal form}, which represents a behavior as a constant or simple
-time-varying value, together with an event stream carrying values which are also
+reactive normal form}, which represents a behavior as a constant or simply
+time-dependent value, together with an event stream carrying values which are also
 behaviors in reactive normal form. Push-based evaluation is achieved by forking
@@
 killed and a new thread spawned to evaluate the new behavior. Unfortunately, the
-implmentation of Reactive uses a tenuous technique which depends on also forking
-threads to evaluate two haskell values concurrently in order to implement event
+implementation of Reactive uses a tenuous technique which depends on forking
+threads to evaluate two Haskell values concurrently, in order to implement event
 merging. This relies on the library author to ensure consistency when this
@@
 compiles to Javascript, making it useful for client-side web programming.
-However, Elm does not provide a notion of switching or of continuous time
+However, Elm does not provide a notion of switching or continuous time
 behaviors, though an approximation is given using discrete-time events which are
@@
 asserts that Arrowized FRP (signal-function FRP, Section~\ref{section:Background-signal_function_frp})
-can be embedded in Elm, but provides no support for this assertion.
+can be embedded in Elm, but provides little support for this assertion\footnote{
+A form of Arrowized FRP employing applicative functors is presented, and
+justified by the assertion that applicative functors are just arrows with the
+input type parameter fixed. While this is true, it ignores the compositional
+benefits of the general Arrow framework, and it is not clear that it provides
+the same enforcement of causality.}
 
@@
 output type parameters, together with a set of routing combinators (see Figure~\ref{figure:arrow_combinators}).
-Signal functions are the first-class
-abstraction in this approach, they represent time-dependent and reactive
-transformers of events and signals,  which are themselves not first class,
-since such values cannot be directly manipulated by the programmer. This
-approach has two motivations: it increases modularity since both the input and
-output of signal functions may be transformed, as opposed to signals or events
-which may only be transformed in their output, and it avoids a large class of
-time and space leaks which have emerged when implementing FRP with first-class
-signals and events.
+Signal functions are the first-class abstraction in this approach. They
+represent time-dependent and reactive transformers of events and signals, which
+are themselves not first class, since such values cannot be directly manipulated
+by the programmer. This approach has two motivations: it increases modularity
+since both the input and output of signal functions may be transformed (as
+opposed to signals or events which may only be transformed in their output) and
+it avoids a large class of time and space leaks which have emerged when
+implementing FRP with first-class signals and events.
 

### 2012-10-30 Docs/Thesis/comparison.tex
Fix figure captions
Thesis draft edits

@@
 
-We have described the implementation of a push-based signal-function FRP system.
-A push-based system should theoretically have advantages in event latency and
-overall performance over a pull-based system. Further, we would like to compare
+We have described the implementation of a push-pull signal-function FRP system.
+Push-based event evaluation should have advantages in event latency and overall
+performance over pull-based event evaluation. We would also like to compare
 the evaluation interface in the current widely-used system with the one proposed
@@
 
-The application of choice is a learning switch controller for OpenFlow switches.
-this application must handle input events triggered by switches signalling that
-they have received a packet for which they have no handling rules, and produce
-output events which install handling rules for dealing with these packets.
+\section{Evaluation Application}
+\label{section:Evaluation_and_Comparisons-Evaluation_Application}
+
+The application of choice is a learning switch controller for OpenFlow~\cite{OpenflowSpec}
+switches. this application must handle input events triggered by switches
+signalling that they have received a packet for which they have no handling
+rules, and produce output events which install handling rules for dealing with
+these packets.
 
@@
 
+\section{Methodology and Results}
+\label{section:Evaluation_and_Comparisons-Methodology_and_Results}
+
 The controllers were benchmarked using the {\tt cbench} program~\cite{cbench}. The
@@
 \hrule
-\caption{Comparison of timeflies vs. Yampa implementing an OpenFlow learning switch controller}
+\caption{Comparison of timeflies vs. Yampa implementing an OpenFlow learning switch controller.}
 \label{fig:timeflies-yampa-comparison}
@@
 
-There is some unexpected correlation between the sampling rate and the response rate
-for TimeFlies, which is likely due to the more frequent size reductions in the table.
+There is some correlation between the sampling rate and the response rate
+for TimeFlies, which is due to the continued wrapping of a replacement signal
+function by {\tt switch} until the next time step (see Section~\ref{subsection:Implementation-Signal_Functions-Implementation_of_Signal_Function_Combinators}).
 Nevertheless, TimeFlies outperforms Yampa in event response by several orders of magnitude,
@@
 interface of time-flies permits us to connect its evaluation directly to GHC's event
-manager, while the design of Yampa requires us to use threads and synchronized communication\footnote{Yampa does include a step-wise evaluation function, but it still couples event
-evaluation firmly to time steps, and is not well documented.}.
+manager, while the design of Yampa requires us to use threads and synchronized
+communication\footnote{Yampa does include a step-wise evaluation function, but
+it still couples event evaluation firmly to time steps, and is not well
+documented.}.

### 2012-10-30 Docs/Thesis/conclusions-and-further-work.tex
Thesis draft edits

@@
 
-The TimeFlies system is a full-fledged and documented FRP library which may
-be extended with utility functions and further optimized. Its performance in
+\section{Conclusions}
+\label{section:Conclusions_and_Further_Work-Conclusions}
+The TimeFlies system is a fully-implemented and -documented FRP library which
+may be extended with utility functions and further optimized. Its performance in
 responding to events is demonstrated to be superior to that of the predominant
@@
 Further, the model of events used by TimeFlies subverts problematic semantic
-questions about the evaluation of events in a system. By using the N-Ary FRP
-type model and separating the evaluation of events from the time steps used
+questions about the evaluation of events in an FRP system. By using the N-Ary
+FRP type model and separating the evaluation of events from the time steps used
 for signals, TimeFlies fully supports non-deterministic merging of events,
@@
 
+TimeFlies includes a fully-documented evaluation interface which clarifies and
+simplifies the task of integrating the TimeFlies system with the many IO systems
+available to Haskell programmers.
+
+\section{Further Work}
+\label{section:Conclusions_and_Further_Work-Further_Work}
 The TimeFlies system would benefit from attentive microbenchmarking and performance

### 2012-10-30 Docs/Thesis/implementation-signalfunctions.tex
Thesis draft edits

@@
 constructors. For instance, if we wish to build a tiny expression language,
-we can write:
+we can use a standard ADT:
 
@@
 
-The code for our our evaluator is the same. Now the input types match, but
-the output type still does not.
+The code for our our evaluator is the same. Now the input types to {\tt (+)}
+match, but the output type still does not.
 
@@
 
-We represent signal functions as a datatype with two constructors.
-The signal functions that a user will compose are non-initialized signal
-functions. They must be provided with an initial set of input signals
-(corresponding to time zero). When provided with this input, they produce their
-time-zero output, and an initialized signal function. The datatype is shown in
-Figure~\ref{figure:signal_function_datatype}
+We represent signal functions as a GADT with three type parameters and two 
+constructors. The first type parameter represents the initialization state,
+and is specialized to {\tt Initialized} or {\tt NonInitialized} depending on the
+constructor. The other two type parameters are the input and output signal
+vectors, respectively. The signal functions that a user will compose are\
+non-initialized signal functions. They must be provided with an initial set of
+input signals (corresponding to time zero). When provided with this input, they
+produce their time-zero output, and an initialized signal function. The datatype
+is shown in Figure~\ref{figure:signal_function_datatype}.
 
@@
 might further improve performance. Nevertheless, this implementation performs
-quite well as it currently exists, as we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparison}.
+quite well as it currently exists, as we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparisons}.

### 2012-10-30 Docs/Thesis/implementation.tex
Thesis draft edits

@@
 functional implementation. For this implementation, I will make significant use
-of extensions to the Haskell type system. These extensions provide mechanisms for expressing
-abstractions not easily expressible in a standard Hindley-Milner type system, while maintaining
-the type soundness of Haskell.
+of an extension to the Haskell type system. This extension, called
+GADTs~\cite{Cheney2003,Xi2003}, provides a mechanism for expressing abstractions
+which are not easily expressible in a standard Hindley-Milner type system, while
+maintaining the type soundness of Haskell. In particular, GADTs provide the
+power we require to take advantage of the signal vector representation of signal
+function inputs and outputs.
 

### 2012-10-30 Docs/Thesis/introduction.tex
Thesis draft edits

@@
 occurrences. However, this implementation did not address concerns of causality
-or compositionality. Further, the existence of events as first-class values in
+or composability. Further, the existence of events as first-class values in
 classic FRP forced the use of an impure and resource-leaking technique to
-compare events when merging, to determine which had the first occurrence after
-the current time. This technique made use of the Haskell {\tt unsafePerformIO}
-function to fork two threads to compare the improving values of the next
-occurrence time for two events being merged. Further, since the classic FRP
-interface permits access to only the output of a behavior or event, and is
-always bound to its implicit inputs, the best the push-based implementation
-could do is to block when no computation needs to be performed. The computation
-cannot be suspended entirely as a value representation.
+compare events when merging, in order to determine which event had the first
+occurrence after the current time. This technique made use of the Haskell
+{\tt unsafePerformIO} function to fork two threads to compare the improving
+values of the next occurrence time for two events being merged. Further, since
+the classic FRP interface permits access to only the output of a behavior or
+event, and is always bound to its implicit inputs, the best the push-based
+implementation could do is to block when no computation needs to be performed.
+The computation cannot be suspended entirely as a value representation.
 
-To address both challenges in a FRP implementation, we to combine the approaches
+To address both challenges in a FRP implementation, we combine the approaches
 and demonstrate a push-based signal-function FRP system. The current
@@
 FRP, events were simply option-valued signals. This approach has the advantage
-of being simple to implement in a pull-based setting. However, it does not have
-an obvious semantics for event merging. Further, it necessitates pull-based
-evaluation, since there is no way to separate the events which may be pushed
-from the truly continuous values which must be repeatedly sampled.
+of being simple to implement. However, it does not have obvious semantics for
+event merging. Further, it necessitates pull-based evaluation, since there is no
+way to separate the events which may be pushed from the truly continuous values
+which must be repeatedly sampled.
 

### 2012-10-30 Docs/Thesis/system_design_and_interface.tex
Fix figure captions
Thesis draft edits

@@
 
-The (not yet fully realized) goal of FRP is to provide an efficient, declarative
-abstraction for creating reactive programs. Towards this overall goal, there are
-three goals which this system is intended to meet.
+The goal of FRP is to provide an efficient, declarative abstraction for creating
+reactive programs. Towards this overall goal, there are three goals which this
+system is intended to meet.
 
@@
 created from by combining smaller systems, without having to reason about the 
-components of the systems being combined.
+implementation of the components of the systems being combined.
 
@@
 that it prohibits push-based evaluation of events, because events are embedded in the
-samples of a signal and must be searched for.
+samples of a signal and must be searched for when sampling.
 
@@
 merge event occurrences. When events being input to a non-deterministic merge have simultaneous occurrences,
-we simply select one arbitrarly to occur first. This does not violate any guarantee about time values, since
+we arbitrarily select one to occur first. This does not violate any guarantee about time values, since
 they will both have the same time value in either case, and does not violate any guarantee about ordering,
@@
 \hrule
-\caption{Signal vector types}
+\caption{Signal vector types.}
 \label{figure:signal_vector_types}
@@
 The representation of signal functions is discussed in
-Section~\ref{section:Implementation-Signal_Functions}. Here it suffices to say
-that the use of GADTs permits the construction of values which make use of
-uninhabited types as instantiations of type parameters.
-
-The type synonyms {\tt :\textasciitilde>} and {\tt :\textasciicircum:} are included for readability and are
-not crucial to the FRP system.
+Section~\ref{section:Implementation-Signal_Functions}. The type synonyms
+{\tt :\textasciitilde>} and {\tt :\textasciicircum:} are included for
+readability and are not crucial to the FRP system.
 
@@
 
-The type of the evaluation monad must track the input type of the signal
-function. The monad's context stores a mapping from outputs to handling actions.
-An existential type can thus be used to ``hide'' the output type of the signal
-function. However, inputs must come from external values, so the input type
-cannot be hidden. There are thus three type parameters to the monad's type
-constructor: the input signal vector, the type of the underlying monad, and the
-monadic type parameter. The type is shown in Figure~\ref{figure:evaluation_monad_types}.
+The type of the evaluation monad must track the input and output type of the
+signal function. The monad's context stores a mapping from outputs to handling
+actions. There are thus four type parameters to the monad's type
+constructor: the input signal vector, the output signal vector, the type of the
+underlying monad, and the monadic type parameter. The type is shown in
+Figure~\ref{figure:evaluation_monad_types}.
 
@@
 grouped as basic signal functions, lifting operations for pure functions,
-routing, reactivity, feedback, and time dependence.
+routing, reactivity, feedback, event processing, joining, and time dependence.
 
@@
 \hrule
-\caption{Basic signal functions.}
+\caption{Basic signal function combinators.}
 \label{figure:basic_signal_functions}
@@
 \hrule
-\caption{Combinator for reactivity.}
+\caption{Reactivity combinator.}
 \label{figure:switch_combinator}
@@
 \hrule
-\caption{Alternate combinators for reactivity.}
+\caption{Alternate reactivity combinators.}
 \label{figure:alternate_switching_combinators}
@@
 Several combinators are provided for manipulating, suppressing, and generating events.
-Output events, in these combinators, are triggered by input events.
-
 Each of the combinators has an option variant and a list variant. The option variant
@@
 (from one or more sources) must be coordinated. For instance, in the
-benchmark application (see Chapter~\ref{chapter:Evaluation_and_Comparisons},
+benchmark application (see Chapter~\ref{chapter:Evaluation_and_Comparisons}),
 a table is maintained that allows knowledge from previous event occurrences
@@
 These three combinators together provide the ability to combine elements of
-signal vectors. By combining these combinators, arbitrary signal vectors can
-be reduced.
+signal vectors. By combining these combinators, along with the {\tt cancelLeft}
+and {\tt cancelRight} routing combinators, arbitrary signal vectors can be
+reduced.
 
@@
 \end{figure}
+
+\subsection{Time Dependence}
+\label{subsection:System_Design_and_Interface-Signal_Functions-Time_Dependence}
+
+A set of combinators are provided for making use of time-dependence in a signal
+function. These combinators allow the modification of signals and events with
+respect to time, and the observation of the current time value. The combinators
+are shown in Figure~\ref{figure:time_combinators}
+
+\begin{figure}
+\begin{code}
+time      :: SVEmpty :~> SVSignal Double
+delay     :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
+integrate :: TimeIntegrate i => SVSignal i :~> SVSignal i
+\end{code}
+\hrule
+\caption{Time-dependent combinators.}
+\label{figure:time_combinators}
+\end{figure}
+
+The simplest time-dependent combinator is {\tt time}, which simply outputs
+the time since it began to be evaluated. This does not necessarily correspond to
+the time since the global signal function began to be evaluated, since the
+{\tt time} combinator may have been introduced through {\tt switch}.
+
+The {\tt delay} signal function allows events to be delayed. An initial delay
+time is given, and event occurrences on the right input can carry a new delay
+time. Event occurrences on the left input are stored and output when their delay
+time has passed. Changing the delay time does not affect occurrences already
+waiting.
+
+The {\tt integrate} combinator outputs the rectangle-rule approximation of the
+integral of its input signal with respect to time.
 
@@
 \hrule
-\caption{State maintained when evaluating a signal function}
+\caption{State maintained when evaluating a signal function.}
 \label{figure:evaluation_state}

### 2012-11-05 Docs/Thesis/abstract.tex
Thesis edits

@@
 inefficient and wasteful evaluation of the FRP system when this strategy is
-employed to evaluate events.
+employed to evaluate events, because the components of the signal function which
+process events must be computed whether or not there is an event occurrence.
 
@@
 whenever new input is available. This frees the system from evaluating the
-network when nothing has changed, and then only components necessary to react to
-the input are re-evaluated. This form of evaluation has been applied to events
-in standard FRP systems~\cite{Elliott2009} but not in signal-function FRP
+network when nothing has changed, and then only the components necessary to
+react to the input are re-evaluated. This form of evaluation has been applied to
+events in standard FRP systems~\cite{Elliott2009} but not in signal-function FRP
 systems.
@@
 applies pull-based evaluation to signals and push-based evaluation to events (a
-"push-pull" system). The semantics of the system are discussed, and its
+``push-pull'' system). The semantics of the system are discussed, and its
 performance and expressiveness for practical examples of interactive programs

### 2012-11-05 Docs/Thesis/appendix-types.tex
Thesis edits

@@
 \begin{description}
-\item[ADT] See {\em Algebraic Datatype}
+\item[ADT] See {\em Algebraic Datatype}.
 
@@
 
-\item[Existential Type] An existential type is a means of ``hiding'' a
-{\em type variable} in an {\em Algebraic Datatype} or
-{\em Generalized Algebraic Datatype}. The Glasgow Haskell Compiler provides
-existential types in two ways. The first is the {\tt Rank2Types} and
-{\tt RankNTypes} extensions. With either of these extensions enabled,
-type variables may be brought into scope for specific {\em data constructor} in
-an Algebraic Datatype using the {\tt forall} keyword. Since this
-type variable is not contained in the parameters of the type constructor, it
-cannot be constrained by type inference. Thus, no external function can assume
-anything about its type, except for equality between occurrences of the
-variable, and {\em class constraints} given in the definition of the data
-constructor. The other way is through Generalized Algebraic Datatypes. Any
-variable occurring in the type of a data member of a data constructor of a
-Generalized Algebraic Datatype, but not occurring in the type of that
-constructor, is automatically existentially quantified.
+\item[GADT] See {\em Generalized Algebraic Datatype}.
 
@@
 type but not the data member types may be instantiated arbitrarily.
+
+\item[Kind] A ``type of types.'' Kinds are used to verify that types are
+consistent during typechecking. The kind of types which contain values is
+{\tt *}, and the kind of single-parameter type constructors which take a
+type is {\tt * -> *}. Other kinds may also be introduced. For instance,
+signal vectors should be their own separate kind, but the Haskell type mechanism
+was not mature enough to support this at the time of this writing..
 
@@
 instantiated (by the typechecker via inference, or by the user via annotation)
-with any type. Together with {\em type constructors}, type variables form the
-basis of polymorphism in Haskell and similar languages.
+with any type at the point where the value so typed is used. 
+Together with {\em type constructors}, type variables form the basis of
+polymorphism in Haskell and similar languages.
 \end{description}

### 2012-11-05 Docs/Thesis/background.tex
Thesis edits

@@
 
-The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produce functions
-as output and/or take functions as input. We say that functions are {\em first-class values} in a language.
+The key abstraction of functional programming is the function, which takes an
+input and produces an output. Functions may produce functions as output and/or
+take functions as input. We say that functions are {\em first-class values} in a
+functional programming language.
 
@@
 
-The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
-abstractions of FRP are behaviors or signals\footnote{Behaviors are generally referred to as {\em signals} in signal function literature. This is unfortunate, since a signal
-function may operate on events, signals, or some combination of the two.}, which are time-dependent values defined at every point in continuous time, and events, which are 
-values defined at countably many points in time. An FRP system will provide combinators to manipulate events and signals and to react
-to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
-based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in FRP languages
-should of course be efficiently executable, but this has proven to be the main challenge in implementing FRP.
+The goal of FRP is to provide compositional and high-level abstractions for
+creating reactive programs. The key abstractions of FRP are behaviors or
+signals\footnote{Behaviors are generally referred to as {\em signals} in
+signal-function literature. This is unfortunate, since a {\em signal function}
+may operate on events, signals, or some combination of the two.}, which are
+time-dependent values defined at every point in continuous time, and events,
+which are values defined at countably many points in time. An FRP system will
+provide combinators to manipulate events and signals and to react to events by
+replacing a portion of the running program in response to an event. Behaviors
+and events, or some abstraction based on them, will be first class, in keeping
+with the spirit of functional programming. Programs implemented in FRP languages
+should of course be efficiently executable, but this has proven to be the main
+challenge in implementing FRP.
 
@@
 reactivity. Elm provides combinators for manipulating discrete events, and
-compiles to Javascript, making it useful for client-side web programming.
+compiles to JavaScript, making it useful for client-side web programming.
 However, Elm does not provide a notion of switching or continuous time
@@
 output type parameters, together with a set of routing combinators (see Figure~\ref{figure:arrow_combinators}).
-Signal functions are the first-class abstraction in this approach. They
-represent time-dependent and reactive transformers of events and signals, which
-are themselves not first class, since such values cannot be directly manipulated
-by the programmer. This approach has two motivations: it increases modularity
-since both the input and output of signal functions may be transformed (as
-opposed to signals or events which may only be transformed in their output) and
-it avoids a large class of time and space leaks which have emerged when
-implementing FRP with first-class signals and events.
+Signal functions form an arrow. Signal functions are the first-class abstraction
+in this approach. They represent time-dependent and reactive transformers of
+events and signals, which are themselves not first class, since such values
+cannot be directly manipulated by the programmer. This approach has two
+motivations: it increases modularity since both the input and output of signal
+functions may be transformed (as opposed to signals or events which may only be
+transformed in their output) and it avoids a large class of time and space leaks
+which have emerged when implementing FRP with first-class signals and events.
 
@@
 switching, in this case in favor of {\em signal inhibition}. Netwire is written
-as an arrow transformer, which, together with the Kliesli arrow instance for the
-IO monad\footnote{Any monad forms a Kliesli arrow.}, permits it to lift IO
-actions as sources and sinks at arbitrary points in a signal function network.
-Signal inhibition is accomplished by making the output of signal functions a
-monoid, and then combining the outputs of signal functions. An inhibited signal
-function will produce the monoid's zero as an output. Primitives have defined
-inhibition behavior, and composed signal functions inhibit if their outputs
-combine to the monoid's zero.
+as an arrow transformer, and permits the lifting of IO actions as sources and
+sinks at arbitrary points in a signal function network. Signal inhibition is
+accomplished by making the output of signal functions a monoid, and then
+combining the outputs of signal functions. An inhibited signal function will
+produce the monoid's zero as an output. Primitives have defined inhibition
+behavior, and composed signal functions inhibit if their outputs combine to the
+monoid's zero.
 
@@
 consisted in using the dependent type system to prove the correctness of the FRP
-system discussed. This work introduced the typing construct of signal vectors,
-which permit the distinction of signal and event types at the level of the FRP
+system discussed. This work introduced a signal vectors, a typing construct
+which permits the distinction of signal and event types at the level of the FRP
 system, instead of making events merely a special type of signal.

### 2012-11-05 Docs/Thesis/comparison.tex
Thesis edits

@@
 The application of choice is a learning switch controller for OpenFlow~\cite{OpenflowSpec}
-switches. this application must handle input events triggered by switches
+switches. This application must handle input events triggered by switches
 signalling that they have received a packet for which they have no handling
@@
 The controllers were benchmarked using the {\tt cbench} program~\cite{cbench}. The
-amount to time to pause between time and signal updates was passed as a parameter to
-both programs. The {\tt cbench} benchmark operates by sending a simulated packet in
+amount to time to pause between time and signal updates (the inverse of the sampling rate for signals)
+was passed as a parameter to both programs. The {\tt cbench} benchmark operates by sending a simulated packet in
 event, waiting for a rule response, and then repeating. The output of the benchmark

### 2012-11-05 Docs/Thesis/conclusions-and-further-work.tex
Thesis edits

@@
 for signals, TimeFlies fully supports non-deterministic merging of events,
-and provides a semantic guarantee that events are not ''lost'' during evaluation.
+and provides a semantic guarantee that events are not ``lost'' during evaluation.
 
@@
 the implementation of the signal function combinators. This may involve, for instance,
-evaluating several time-steps at once in a data-parallel manner,task parallelism
+evaluating several time-steps at once in a data-parallel manner, task parallelism
 between different branches of a signal function, or speculative evaluation of switch

### 2012-11-05 Docs/Thesis/implementation-evalinterface.tex
Thesis edits

@@
 
-This property will be useful because we wish our evaluation interface to be
+This property will be useful because we want our evaluation interface to be
 part of a monad stack, the base of which is Haskell's IO monad. By formulating
@@
 
-The evaluation interface is exported as follows:
-
-\begin{code}
-data SFEvalState svIn m
-data SFEvalT svIn m
-data HandlerSet sv m
-data EventOccurrence sv
-data SignalDelta sv
-
-sd      :: a -> SignalDelta (SVSignal a)
-sdLeft  :: SignalDelta svl -> SignalDelta (SVAppend svl svr)
-sdRight :: SignalDelta svr -> SignalDelta (SVAppend svl svr)
-sdBoth  :: SignalDelta svl -> SignalDelta svr 
-        -> SignalDelta (SVAppend svl svr)
-
-eo      :: a -> EventOccurrence (SVEvent a)
-eoLeft  :: EventOccurrence svl -> EventOccurrence (SVAppend svl svr)
-eoRight :: EventOccurrence svr -> EventOccurrence (SVAppend svl svr)
-
-sdHS    :: (a -> m ()) -> HandlerSet m (SVSignal a)
-eoHS    :: (a -> m ()) -> HandlerSet m (SVEvent a)
-hsLeft  :: HandlerSet m svl -> HandlerSet m (SVAppend svl svr)
-hsRight :: HandlerSet m svr -> HandlerSet m (SVAppend svl svr)
-hsBoth  :: HandlerSet m svl -> HandlerSet m svr 
-        -> HandlerSet m (SVAppend svl svr)
-
-initSFEval :: HandlerSet svOut m 
-           -> SF NonInitialized svIn svOut 
-           -> SFEvalState svIn m
-
-runSFEvalT :: SFEvalT svIn m a -> SFEvalState svIn m -> m a
-
-push   :: EventOccurrence svIn -> SFEvalT svIn m ()
-update :: SignalDelta svIn -> SFEvalT svIn m ()
-sample :: Double -> SFEvalT svIn m ()
-\end{code}
+The evaluation interface is exported as shown in
+Section~\ref{section:System_Design_and_Interface-Evaluator}.
 
@@
 function in the {\tt SFEvalState}, and replacing the {\tt SFEvalState}.
+
+The {\tt update} action updates the current input signal delta, which will be
+sampled at the next {\tt step} action. It has no immediately observable effect.
+It simply updates the value of one signal in the input signal vector, a delta
+of which is stored in the {\tt SFEvalState} record.
+
+The {\tt step} action takes a time delta, and calls the signal continuation of
+the stored signal function. It actuates the outputs using the stored handlers,
+replaces the stored delta with an empty delta, and stores the resulting new
+signal function in the state.

### 2012-11-05 Docs/Thesis/implementation-signalfunctions.tex
Thesis edits

@@
 temporal evaluation of signal functions.
-
-In order to create these concrete representations, we employ the Haskell type
-system extension known as Generalized Algebraic Datatypes, which we divert
-briefly to explain before exploring its use in the implementation.
 
@@
 
-In Chapter~\ref{chapter:System_Design_and_Interface} we present signal vectors
+In Chapter~\ref{chapter:System_Design_and_Interface} we presented signal vectors
 as a set of types. In order to be completely principled, we should isolate these
-types into their own {\tt kind} (a sort of type of types), however, the Haskell
+types into their own {\em kind} (a sort of type of types); however, the Haskell
 extension for this was far from stable at the time this system was created.
@@
 
-To get our feet wet, we create a representation which carries a value for
-every {\tt SVSignal} leaf of a signal vector. In order to do this, we restrict
-each of our constructors to a single signal vector type. This ensures that the
+The first thing to represent is {\em samples}, which are sets of values for
+the signal components of a signal vector. Therefore, we create a representation
+which carries a value for every {\tt SVSignal} leaf of a signal vector. In order
+to do this, we restrict each of our constructors to a single signal vector type.
+So there are three leaf constructors: {\tt SVSample}, which carries a value; and
+{\tt SVSampleEvent} and {\tt SVSampleEmpty}, which do not. This ensures that the
 only way to represent a sample leaf is with the {\tt SVSample} constructor,
@@
 
-So we now have a representation for the signal components of a signal vector.
-What about the event components? We want to represent event occurrences, each
-of which will correspond to at most one event in the vector. So a different
+What about the event components? We want to represent event occurrences,
+each of which will correspond to at most one event in the vector. So a different
 representation is called for. In this case, there will be only three
@@
 
-Another strategy is notification, in which the program exposes an interface
+Another approach is notification, in which the program exposes an interface
 which the source of updated information may invoke. This is a repeated entry
@@
 
-The simplest combinator is the {\tt identity} signal function. This signal
-function simply passes all of its inputs along as outputs. The initialization
-function simply passes along the received sample and outputs the initialized
-version of the signal function. The initialized version of the input is similar,
-but is self-referential. It replaces itself with itself after every signal
-function or event. The implementation is shown in Figure~\ref{figure:identity_implementation}.
+As an example of the implementation of combinators, we show the implementation
+of the {\tt identity} signal function in Figure~\ref{figure:identity_implementation}.
+This signal function simply passes all of its inputs along as outputs. The
+initialization function simply passes along the received sample and outputs the
+initialized version of the signal function. The initialized version of the input
+is similar, but is self-referential. It outputs itself as its replacement. This
+is standard for simple and routing combinators which are not reactive, and
+simply move samples and event occurrences around.
 
@@
 place where the advantage of responding to events independently from signal
-samples becomes clear. The implementation is shown in Figure~\ref{figure:serial_composition_implementation}.
+samples becomes clear.
+
+This is the only primitive combinator which takes two signal functions, and
+thus, it is the only way to combine signal functions. Parallel, branching, and
+joining composition can be achieved by modifying signal functions with the
+{\tt first} and {\tt second} combinators and composing them with the
+routing and joining combinators.
+
+Combinators which take one or more signal functions as input must recursively
+apply themselves, as is shown in the implementation of serial composition
+(Figure~\ref{figure:serial_composition_implementation}). They must also
+handle initialization, retaining the initialized signal functions and passing
+them to the initialized version of the combinator.
 
@@
 
-The implementation of the non-initialized version is simple. It takes two
-non-initialized signal functions, and returns the sample output of applying the
-second to the sample output of the first, along with an initialized signal
-function constructed by applying the initialized version of the combinator to
-the initialized signal functions given by the inputs.
-
-The initialized version constructs two continuations. The first receives a time
-and signal update input, and applies the signal continuation of the first
-signal function to it. The output is fed, along with the time update, into the
-second signal function, first the signal update, and then events. The output
-of the second, along with the accumulated updated signal functions with the
-initialized serial composition recursively applied, is returned as output.
-
-The event continuation is similar but simpler, as there is no need to deal
-with signal updates.
-
 The switch combinator is the means of introducing reactivity into a signal
 function. This combinator allows a signal function to replace itself by
-producing an event occurrence. The switch combinator stores the full sample
+producing an event occurrence. The combinator wraps a signal function, and 
+observes an event on the right side of the output signal vector. At the first
+occurrence of the event, the signal function carried by the occurrence replaces
+the signal function.
+
+The switch combinator stores the input sample provided during initialization,
+and updates it with the input signal updates. When the wrapped signal function
+produces an occurrence carrying a new signal function, that signal function is
+initialized with the stored input sample. It is then "wrapped" by another
+function which closes over its output sample, and outputs the sample as a signal
+update as the next time step. After this, it acts as the new signal function.
+This wrapping has some performance implications, which are discussed
+
+This combinator checks the outputs of the wrapped
+signal function for an event occurrence from which an uninitialized signal
+function is extracted. The switch combinator stores the full sample
 for its input vector (which is identical to the input vector of the supplied
 signal function) to initialize the new signal function. This also demands that
-it continue wrapping the signal function until the next sampling interval, so
-that it can actuate the output of initialization as a replacement signal. This
-has some performance implications, which are discussed in
-Chapter~\ref{chapter:Evaluation_and_Comparisons}. The implementation is shown in
-Figure~\ref{figure:switch_implementation}.
-
-\begin{figure}
-\begin{code}
-switch :: (svIn :~> (svOut :^: SVEvent (svIn :~> svOut))) -> svIn :~> svOut
-switch (SF sigSampleF) =
-  SF (\sigSample -> let (sigSampleSF, sf) = sigSampleF sigSample
-                        (sigSampleOut, _) = splitSample sigSampleSF
-                    in (sigSampleOut, switchInit sigSample sf))
-
--- Pre-occurrence
-switchInit ::    SVSample svIn
-              -> SF Initialized svIn 
-                    (SVAppend svOut 
-                     (SVEvent (SF NonInitialized svIn svOut)))
-              -> SF Initialized svIn svOut
-switchInit inputSample sf@(SFInit timeCont changeCont) =
-  SFInit (\dt sigDelta ->
-            let (sigDeltaSF, occsSF, newSF) = timeCont dt sigDelta
-                newInputSample = updateSample inputSample sigDelta
-                (sigDeltaOut, _) = splitDelta sigDeltaSF
-                (occsOut, occsSwitch) = splitOccurrences occsSF
-                (outputDelta, nextSF) =
-                  maybe (sigDeltaOut, switchInit newInputSample newSF)
-                    (\(SF sigSampleF) ->
-                       let (outputSample, switchSF) =
-                             sigSampleF newInputSample
-                           outputDelta = sampleDelta outputSample
-                       in (outputDelta, switchSF))
-                  $ occurrenceListToMaybe occsSwitch
-            in (outputDelta, occsOut, nextSF))
-         (\evtOcc ->
-            let (occsSF, newSF) = changeCont evtOcc
-                (occsOut, occsSwitch) = splitOccurrences occsSF
-                nextSF = maybe (switchInit inputSample newSF) 
-                               (\(SF sigSampleF) ->
-                                  let (sampleOut, switchSF) =
-                                        sigSampleF inputSample
-                                  in switchWait sampleOut switchSF)
-                         $ occurrenceListToMaybe occsSwitch
-            in (occsOut, nextSF))
-
-switchWait ::    SVSample svOut
-              -> SF Initialized svIn svOut
-              -> SF Initialized svIn svOut
-switchWait outputSample sf@(SFInit timeCont changeCont) =
-  SFInit (\dt sigDelta -> let (sigDeltaOut, outOccs, newSF) =
-                                timeCont dt sigDelta
-                          in (sampleDelta $
-                                updateSample outputSample sigDeltaOut,
-                              outOccs,
-                              newSF))
-         \evtOcc -> let (outOccs, newSF) = changeCont evtOcc
-                    in (outOccs, switchWait outputSample newSF))
-\end{code} 
-\hrule
-\caption{Implementation of reactivity.}
-\label{figure:switch_implementation}
-\end{figure}
+it add a wrapper to the new signal function which waits for the next sampling
+interval and actuates the sample output at initialization as an output set
+of changes to the signal. This has some performance implications, which are
+discussed in Chapter~\ref{chapter:Evaluation_and_Comparisons}.
 
@@
 \hrule
-\caption{Swap routing combinator implementation.}
+\caption{Implementation of the {\tt swap} routing combinator.}
 \label{figure:swap_implementation}
@@
 
-The {\tt first} and {\tt second} combinators are not quite as simple, as they
-must transform a provided signal function. For signal changes, they must split
-the set of changes into those which will be passed to the signal function and
-those which will be simply passed along to the output, and then recombine them
-on the other side. For event occurrences, the occurrence must be pattern-matched
-to determine whether to call the event continuation from the provided signal
-function or passed through, and output event occurrences must have the suitable
-routing constructor re-applied. In any case, when a continuation has been
-applied, the combinator must be recursively applied to the new signal function.
-The implementation of {\tt first} is shown in Figure~\ref{figure:first_implementation}.
-The implementation of {\tt second} is the obvious reversal of this
-implementation.
-
-\begin{figure}
-\begin{code}
-first :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
-first (SF sigSampleF) =
-  SF (\sigSample -> let (leftSample, rightSample) = splitSample sigSample
-                        (leftSampleOut, sf) = sigSampleF leftSample
-                    in (combineSamples leftSampleOut rightSample,
-                        firstInit sf))
-
-firstInit ::    SF Initialized svIn svOut
-             -> SF Initialized (SVAppend svIn sv) (SVAppend svOut sv)
-firstInit (SFInit timeCont inputCont) = 
-  let firstInitSF =
-    SFInit (\dt sigDelta ->
-               let (input, rightOutput) = splitDelta sigDelta
-                   (sigDeltaOutput, evtOutput, sf1New) = timeCont dt input
-               in (combineDeltas sigDeltaOutput rightOutput,
-                   map occLeft evtOutput,
-                   firstInit sf1New))
-           (\evtOcc ->
-              case chooseOccurrence evtOcc of
-                Left  lChange -> let (changes, sf) = inputCont lChange
-                                 in (map occLeft changes, firstInit sf)
-                Right rChange -> ([occRight rChange], firstInitSF))
-  in firstInitSF
-\end{code}
-\hrule
-\caption{Implemenation of {\tt first} combinator.}
-\label{figure:first_implementation}
-\end{figure}
+The {\tt first} and {\tt second} combinators are similar to serial composition,
+but they transform only one signal function. For signal changes, they must split
+the set of input changes into those which will be passed to the signal function
+and those which will be simply passed along to the output, and then recombine
+them on the other side. For event occurrences, the occurrence must be
+pattern-matched to determine whether to call the event continuation from the
+provided signal function or passed through, and output event occurrences must
+have the suitable routing constructor re-applied. In any case, when a
+continuation has been applied, the combinator must be recursively applied to the
+new signal function.
 
@@
 sample. Signal samples are thus automatically decoupled after initialization.
-The implementation, which makes extensive use of the recursive nature of
-Haskell's {\tt let} construct, is shown in Figure~\ref{figure:loop_implementation}.
-
-\begin{figure}
-\begin{code}
-loop :: (svIn :^: svLoop) :~> (svOut :^: svLoop) -> svIn :~> svOut
-loop (SF sigSampleF) =
-  SF (\sigSample -> 
-        let (sigSampleOut, sfInit) =
-              sigSampleF (combineSamples sigSample sigSampleOutRight)
-            (sigSampleOutLeft, sigSampleOutRight) =
-              splitSample sigSampleOut
-        in (sigSampleOutLeft, loopInit deltaNothing sfInit))
-
-loopInit ::    SVDelta svLoop
-            -> SF Initialized (SVAppend svIn svLoop) (SVAppend svOut svLoop)
-            -> SF Initialized svIn svOut
-loopInit loopDelta (SFInit timeCont changeCont) =
-  SFInit (\dt sigDelta ->
-            let (sigDeltaOut, occsOut, newSF) =
-                  timeCont dt (combineDeltas sigDelta loopDelta)
-                (occsOutput, loopOccs) = splitOccurrences occsOut
-                (occsOutput', loopOccs') = splitOccurrences foldOccs
-                (foldOccs, newSF') =
-                  applySF newSF (map occRight (loopOccs ++ loopOccs'))
-                (outputDelta, newLoopDelta) = splitDelta sigDeltaOut
-            in (outputDelta,
-                occsOutput ++ occsOutput',
-                loopInit newLoopDelta newSF'))
-         (\change -> let (occsOut, newSF) = changeCont $ occLeft change
-                         (occsOutput, loopOccs) =
-                           splitOccurrences occsOut
-                         (occsOutput', loopOccs') =
-                           splitOccurrences foldOccs
-                         (foldOccs, newSF') =
-                           applySF newSF
-                             (map occRight (loopOccs ++ loopOccs'))
-                     in (occsOutput ++ occsOutput',
-                         loopInit loopDelta newSF'))
-\end{code}
-\hrule
-\caption{Implementation of feedback.}
-\label{figure:loop_implementation}
-\end{figure}
+The implementation makes use of the recursive nature of the {\tt let} construct
+in Haskell, and the non-strict evaluation of Haskell, to implement feedback.
 
@@
 had signal updates. Optimizations in the style of Yampa, observed by keeping
-an updated for the signal function and pattern-matching on it when switching,
-might further improve performance. Nevertheless, this implementation performs
-quite well as it currently exists, as we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparisons}.
+an updated AST for the signal function and pattern-matching on it when switching,
+might further improve performance. In particular, collapsing nested or
+serially-composed versions of the {\tt switchWait} step when switching would
+remove at least some of the observed dependence of performance on sampling rate.
+Nevertheless, this implementation performs quite well as it currently exists, as
+we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparisons}.

### 2012-11-05 Docs/Thesis/implementation.tex
Thesis edits

@@
 of an extension to the Haskell type system. This extension, called
-GADTs~\cite{Cheney2003,Xi2003}, provides a mechanism for expressing abstractions
-which are not easily expressible in a standard Hindley-Milner type system, while
-maintaining the type soundness of Haskell. In particular, GADTs provide the
-power we require to take advantage of the signal vector representation of signal
-function inputs and outputs.
+Generalized Algebraic Datatypes (GADTs)~\cite{Cheney2003,Xi2003}, provides a mechanism
+for expressing abstractions which are not easily expressible in a standard
+Hindley-Milner type system, while maintaining the type soundness of Haskell.
+In particular, GADTs provide the power we require to take advantage of the
+signal vector representation of signal function inputs and outputs.
 

### 2012-11-05 Docs/Thesis/introduction.tex
Thesis edits

@@
 web browsers, web applications, video games, multimedia authoring and playback tools, operating system shells and kernels, servers,
-robotic control programs, and many others. This attribute of a program is called
-reactivity.
+robotic control programs, and many others. This attribute of a program is called reactivity.
 
-Functional reactive programming (FRP) is a promising class of abstractions for writing reactive programs in functional languages. FRP 
-abstractions translate desirable properties of the underlying functional languages, such as higher-order functions and referential
-transparency, to reactive constructs, generally without modifications to the underlying language. This permits the compositional
+Functional reactive programming (FRP) is a promising class of abstractions for
+writing reactive programs in functional languages. The FRP abstractions are
+behaviors or signals, which are functions of continuous time, and events, which
+are sequences of time-value pairs. These abstractions translate desirable
+properties of the underlying functional languages, such as higher-order
+functions and referential transparency, to reactive constructs, generally
+without modifications to the underlying language. This permits the compositional
 construction of reactive programs in purely functional languages.
@@
 evaluated using a function which provides input to the signal function and acts
-on the output values. This evaluation function is elided in existing literature
-on signal functions, but we provide a documented and improved interface for
-evaluating signal functions. A further advantage of signal function FRP programs
-is that they are more composable, since additional signal functions may be
-composed on the input or output side of a signal function, rather than simply
-the output side in classic (non-signal-function) FRP systems.
+on the output values. This evaluation function is generally elided in existing
+literature on signal functions, but we provide a documented and improved
+interface for evaluating signal functions. A further advantage of signal
+function FRP programs is that they are more readily composed, since additional
+signal functions may be composed on the input or output side of a signal
+function, rather than simply the output side as in classic (non-signal-function)
+FRP systems.
 
@@
 compare events when merging, in order to determine which event had the first
-occurrence after the current time. This technique made use of the Haskell
-{\tt unsafePerformIO} function to fork two threads to compare the improving
-values of the next occurrence time for two events being merged. Further, since
-the classic FRP interface permits access to only the output of a behavior or
-event, and is always bound to its implicit inputs, the best the push-based
-implementation could do is to block when no computation needs to be performed.
-The computation cannot be suspended entirely as a value representation.
+occurrence after the current time. Further, since the classic FRP interface
+permits access to only the output of a behavior or event, and is always bound to
+its implicit inputs, the best the push-based implementation can do is to block
+when no computation needs to be performed. The computation cannot be suspended
+entirely as a value representation.
 
@@
 FRP, events were simply option-valued signals. This approach has the advantage
-of being simple to implement. However, it does not have obvious semantics for
+of being simple to implement. However, it does not have an obvious semantics for
 event merging. Further, it necessitates pull-based evaluation, since there is no

### 2012-11-05 Docs/Thesis/system_design_and_interface.tex
Thesis edits

@@
 Since FRP programs are expected to  interact with an outside world in real time,
-efficiency cannot be measured by the runtime of a program. Thus, when speaking of efficiency,
+efficiency cannot simply be measured by the runtime of a program. Thus, when speaking of efficiency,
 we are expressing a desire that the system utilize as few system resources as possible
@@
 A composable abstraction is one in which values in that abstraction may be
-combined in such a way that reasoning about their actions together involves
-little more than reasoning about their actions separately. In a signal function
+combined in such a way that reasoning about their combined actions involves
+little more than reasoning about their individual actions. In a signal function
 system, the only interaction between composed signal functions ought to be that
@@
 
-\subsection{Ease of Integration}
+\subsection{Simple Integration}
 \label{subsection:System_Design_and_Interface-Goals-Ease_of_Integration}
@@
 
-In previous systems such as Yampa, events were understood (and typed) as option-valued
+In previous systems such as Yampa, events were understood (and implemented) as option-valued
 signals. This approach is undesirable for several reasons. The most pressing reason is
@@
 Another concern is that this approach limits the event rate to the sampling rate.
-The rate of sampling should, at some level, not matter to the FRP system. Events which
-occur between sampling intervals are never observed by the system.
+The rate of sampling should, at some level, not matter to the FRP system. Events
+which occur between sampling intervals are never observed by the system. Since
+the signal function is never evaluated except at the sampling intervals, no
+events can otherwise be observed.
 
@@
 input and produces exactly one output. Multiple inputs or outputs are handled
-by making the output a tuple, and combinators which combine or split the inputs
-or outputs of a signal assume this. Events are represented at the type level
-as a particular type of signal, and at the value level as an option, either an
-event occurrence or not.
+by making the input or output a tuple, and combinators which combine or split
+the inputs or outputs of a signal assume this. Events are represented at the
+type level as a particular type of signal, and at the value level as an option,
+either an event occurrence or not.
 
 This method of typing excludes push-based evaluation at the outset.
-It is not possible to construct a "partial tuple" nor in general is it possible
-to construct only part of any type of value. Push-based evaluation depends on
-evaluating only that part of the system which is updated, which means evaluating
-only that part of the input which is updated.
+It is not possible to construct a ``partial tuple'' nor in general is it
+possible to construct only part of any type of value. Push-based evaluation
+depends on evaluating only that part of the system which is updated, which means
+evaluating only that part of the input which is updated.
 
@@
 A {\em monad} is a standard, composable abstraction for writing functions with
-a context, used in Haskell for IO~\cite{PeytonJones1993,PeytonJones2001} among
+a context, used in Haskell for IO~\cite{PeytonJones1993,PeytonJones2001} and
 other tasks. A monad is simply a 1-arity type constructor together with two
 functions. The first function, {\tt return}, takes a value of type {\tt a} and
-produces a value of type {\tt m a}, where {\tt m} is the type constructor. The
-second, called {\tt bind} and stylized in the Haskell standard library as the
-infix operator {\tt (>>=)}, takes a value of type {\tt m a} and a function
-from {\tt a} to {\tt m b} and produces a value of type {\tt m b}. This allows
-a value to be operated on out of the context and a new context to be assigned.
+produces a value of type {\tt m a}. The second, called {\tt bind} and stylized
+in the Haskell standard library as the infix operator {\tt (>>=)}, takes a value
+of type {\tt m a} and a function of type {\tt a -> m b} and produces a value of
+type {\tt m b}. This allows a value to be operated on out of the context and a
+new context to be assigned.
 
@@
 -- A signal function's evaluations state
-data SFEvalState svIn m
+data SFEvalState svIn svOut m
 -- The evaluation monad
-data SFEvalT svIn m a
+data SFEvalT svIn svOut m a
+
+instance Monad m => Monad (SFEvalT svIn svOut m)
 \end{code}
@@
 step after it is switched into the network. The {\tt after} function waits for
-the specified amount of time before producing the event occurrence.
+the specified amount of time before producing an event occurrence.
 
@@
 
--- An event after the specified amount of time has elapsed.
+-- An event with one occurrence
+-- after the specified amount of time has elapsed.
 after    :: Double -> a -> SVEmpty :~> SVEvent a
@@
 
-Two combinators are provided to lift pure functions to signal functions (Figure~\ref{figure:lifting_pure_functions}).
-The {\tt pureSignal} combinator applies the pure function to a signal at every
-sample point. The {\tt pureEvent} combinator applies the function to each
-occurrence of an input event.
+Two combinators are provided to lift pure functions to signal functions
+(Figure~\ref{figure:lifting_pure_functions}). The {\tt pureSignalTransformer}
+combinator applies the pure function to a signal at every sample point. The
+{\tt pureEventTransformer} combinator applies the function to each occurrence of
+an input event.
 
@@
 Only those combinators which modify or combine signal functions
-({\tt (>>>)}, {\tt first}, {\tt second}) are reactive, and then
-only if they inherit their reactivity from the signal function(s)
-they modify. The rest do not react to or modify the input in any
-way, except to re-arrange it, copy it, or discard it altogether.
+({\tt (>>>)}, {\tt first}, {\tt second}) are reactive (may replace themselves in
+response to events, and then only if they inherit their reactivity from the
+signal function(s) which are inputs to combinator. The rest do not
+react to or modify the input in any way, except to re-arrange it, copy it, or
+discard it altogether.
 
@@
 
-There are other formulations of a reactive combinator which may be implemented
-using the one supplied. These are shown in Figure~\ref{figure:alternate_switching_combinators}
-and may be provided in a future version of the TimeFlies library.
+There are other formulations of a reactive combinator found in FRP literature
+and libraries, which may be implemented using the one supplied. These are shown
+in Figure~\ref{figure:alternate_switching_combinators} and may be provided in a
+future version of the TimeFlies library.
 
@@
 -- for each element of the list
-accumulateList ::    (a -> b -> ([c]))
+accumulateList ::    (a -> b -> ([c], a))
                   -> a
@@
 The {\tt combineSignals} combinator applies a binary function pointwise to two
-signals, and produces the result of this application as a third signal.
+signals, and produces the result of this application as a third signal. The
+combining function is necessary because we will have two input samples at each
+time step, and must produce exactly one output sample.
 
@@
 union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
-combineSignals :: (a -> b -> c) -> (SVEvent a :^: SVEvent b) :~> SVEvent c
+combineSignals :: (a -> b -> c) -> (SVSignal a :^: SVSignal b) :~> SVSignal c
 capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent (b, a)
@@
 the time since the global signal function began to be evaluated, since the
-{\tt time} combinator may have been introduced through {\tt switch}.
+signal function in which the {\tt time} combinator is used may have been
+introduced through {\tt switch}.
 
@@
 -- Create the initial state for evaluating a signal function
-initSFEval ::    SVHandler (m ()) svOut
-              -> SVSample svIn
-              -> Double
-              -> (svIn :~> svOut)
+initSFEval ::    SVHandler (m ()) svOut   -- Output handlers
+
+              -> SVSample svIn            -- Initial input samples
+
+              -> Double                   -- Initial external time,
+                                          -- corresponding to time 0 for
+                                          -- the signal function
+
+              -> (svIn :~> svOut)         -- Signal function to evaluate
               -> SFEvalState m svIn svOut
@@
 \end{figure}
+
+Sets which correspond to signal vectors are built with ``smart'' constructors.
+For instance, to construct a set of handlers, individual handling functions are
+lifted to handler sets with the {\tt signalHandler} and {\tt eventHandler}
+functions, and then combined with each other and {\tt emptyHandler} leaves
+using the {\tt combineHandlers} function.
+
+Building the initial input sample is similar, but {\tt sampleEvt} leaves do
+not carry a value.
 

### 2012-11-24 Docs/IFL2012/paper.tex
Beginning edits to IFL paper

@@
 Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
-Recent work on classic FRP has focused on efficient implementation. Reactive is
-a system for push-pull FRP~\cite{Elliott2009}. Push-based evaluation evaluates a
+Recent work on classic FRP has focused on efficient implementation. One approach to
+efficiency is to separate the evaluation of behaviors and events, since suitable 
+strategies give best performance in each case. Push-based evaluation evaluates a
 system only when input is available, and is thus suitable for discrete inputs
@@
 The initial implementations of FRP made use of pull-based evaluation for both
-behaviors and events. Reactive, as well as more recent systems such as
+behaviors and events. Reactive~\cite{Elliott2009}, as well as more recent systems such as
 ``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
-events and pull-based evaluation for behaviors.
+events and pull-based evaluation for behaviors. This is known as ``push-pull''
+evaluation.
 
 All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
-have used pull-based evaluation. This is due to the ease of implementation of
-pull-based evaluation, and the types used for signal functions which do not
-permit distinguishing signals and events, or constructing only part of the
-input (for instance, one event occurrence.)
+have used pull-based evaluation for both signals and events. This is due to
+the ease of implementation of pull-based evaluation, and the types used for
+signal functions which do not permit distinguishing signals and events, or
+constructing only part of the input (for instance, one event occurrence.)
 
@@
 
-\subsection{Examples of Signal Function Primitives}
-\label{subsection:Examples_of_Signal_Function_Primitives}
-The most basic signal is the {\tt identity} signal function, which, as its
-name suggests, simply passes its input to its output.
-
-The {\tt constant} signal function has an empty input, and produces a constant
-signal as its output.
-
-The {\tt asap} signal function produces an event at the first time interval
-after it begins, and never again.
-
-The {\tt filter} signal function applies a predicate to the values of event
-occurrences and passes them along only if the predicate is true of the value.
-The predicate outputs {\tt Maybe} values to allow the user to avoid partial
-functions on types such as {\tt Maybe} or {\tt Either}.
-
-The {\tt associate} signal function is a routing function, which takes as input
-a combination of three signal vectors to which the signal vector append
-constructor is combined left-associatively, and produce the same signal vector
-as output, but with the signal vector append constructor combined
-right-associatively.
-
-The {\tt switch} function is the essential signal function for reactivity. It
-acts as the supplied signal function until that signal function produces an
-event on its right output. This event's value is a new signal function, which
-replaces the switch signal function.
-
-The {\tt loop} function permits a signal function to receive its own output
-as an input.
-
-Implementations for these examples will be discussed in Sec.~\ref{subsection:Signal_Function_Implementations}.
-
 \subsection{Evaluation Interface}
@@
 
-The evaluation interfaceis a monad transformer~\cite{Liang1995}. This allows
+The evaluation interface is a monad transformer~\cite{Liang1995}. This allows
 a signal function evaluation to be constructed from primitive evaluation actions
@@
 Using GADTs, we can construct representations of signal vectors to use in our
-implementation. The first such representation carries a value for each signal
-component of a signal vector, and is shown in Figure~\ref{figure:signal_sample_representation}.
+implementation.
 
-\begin{figure}[t]
+When constructing a representation, we consider how to represent each
+component of a signal vector, whether it is a signal element, an empty
+element, or an append node. For instance, in order to represent a signal
+sample for all signals in an element, we construct the GADT shown in Figure~\ref{figure:signal_sample_representation}.
+
+\begin{figure}
 \begin{code}
@@
   SVSampleEvt     ::    SVSample (SVEvent a)
-  SVNothing   :: SVSample SVEmpty
-  SVBoth      :: SVSample svLeft -> SVSample svRight
+  SVSampleNothing ::    SVSample SVEmpty
+  SVSampleBoth    ::    SVSample svLeft
+                     -> SVSample svRight
                      -> SVSample (svLeft :^: svRight)
-
-sample         :: a -> SVSample a
-sampleEvt      :: SVSample (SVEvent a)
-sampleNothing  :: SVSample SVEmpty
-combineSamples :: SVSample svLeft -> SVSample svRight
-                  -> SVSample (svLeft :^: svRight)
-splitSample    :: SVSample (svLeft :^: svRight)
-                  -> (SVSample svLeft, svSample svRight)
-sampleValue    :: SVSample (SVSignal a) -> a
 \end{code}
 \hrule
-\caption{Signal sample representation.}
+\caption{A representation which carries a value for each signal in a signal vector.}
 \label{figure:signal_sample_representation}
-\end{figure}
-
-A value of this type represents the entire vector, though it has nullary
-constructors at event leaves and empty leaves. For event occurrences, we want
-to represent only a single point in the vector, as shown in Figure~\ref{figure:event_occurrence_representation}.
-This representation uses two constructors, each of which leaves one or the
-other of the type variables in the signal vector {\tt SVAppend} constructor
-unconstrained.
-
-\begin{figure}[t]
-\begin{code}
-data SVOccurrence sv where
-  SVOccurrence :: a -> SVOccurrence (SVEvent a)
-  SVOccLeft    :: SVOccurrence svLeft
-                  -> SVOccurrence (svLeft :^: svRight)
-  SVOccRight   :: SVOccurrence svRight
-                  -> SVOccurrence (svLeft :^: svRight)
-
-occurrence        :: a -> SVOccurrence (SVEvent a)
-occLeft           :: SVOccurrence svLeft
-                     -> SVOccurrence (svLeft :^: svRight)
-occRight          :: SVOccurrence svRight
-                     -> SVOccurrence (svLeft :^: svRight)
-chooseOccurrences :: SVOccurrence (svLeft :^: svRight)
-                     -> Either (SVOccurrence svLeft) (SVOccurrence svRight)
-fromOccurrence    :: SVOccurrence (SVEvent a) -> a
-\end{code}
-\hrule
-\caption{Event occurrence representation.}
-\label{figure:event_occurrence_representation}
-\end{figure}
-
-The final representation ``signal deltas,'' carries replacement values for some
-points in a signal vector. Here, we combine the unconstrained type variable
-from the occurrence representation with the ``Both'' constructor from the sample
-representation. The unconstrained type variable will be placed in a separate
-constructor, which allows the construction of empty signal deltas, as is shown
-in Figure~\ref{figure:signal_delta_representation}.
-
-\begin{figure}[t]
-\begin{code}
-data SVDelta sv where
-  SVDelta        :: a -> SVDelta (SVSignal a)
-  SVDeltaNothing :: SVDelta sv
-  SVDeltaBoth    :: SVDelta svLeft -> SVDelta svRight
-                    -> SVDelta (svLeft :^: svRight)
-delta         :: a -> SVDelta (SVSignal a)
-deltaNothing  :: SVDelta sv
-combineDeltas :: SVDelta svLeft -> SVDelta svRight
-                 -> SVDelta (svLeft :^: svRight)
-updateSample  :: SVDelta sv -> SVSample sv -> SVSample sv
-\end{code}
-\hrule
-\caption{Signal delta representation.}
-\label{figure:signal_delta_representation}
 \end{figure}

### 2012-11-26 Docs/IFL2012/paper.tex
Copypasta thesis chapters into IFL paper. To edit later.

@@
 
-\section{System Design}
-\label{section:System_Design}
-
-Our goal is to produce a composable and efficient FRP system. Signal function
-FRP has an advantage in terms of composability, because it permits the
-construction of self-contained objects through both input and output composition,
-rather than purely through output composition as in classic FRP. Signal-function
-FRP also avoids problematic properties common to classic FRP systems such as a
-large class of time and space leaks~\cite{Liu2007}.
-
-Efficient implementations of signal-function FRP have been approached through
-runtime optimization~\cite{Nilsson2005}, but all implementations have been
-pull-based for both signals and events. A truly efficient implementation will
-likely combine run-time optimization with push-based evaluation for events.
-
-The concept of N-Ary FRP was to encode additional safety properties into the
-types of a signal-function FRP system~\cite{Sculthorpe2011}. This system
-introduced the concept of {\em signal vectors} as input and output types for
-signal functions. Signal vectors are combinations of signals and events. In
-N-Ary FRP, signal functions are represented at the type level, and type-level
-functions (type families in Haskell) are used to specify the representations
-of signal functions.
-
-In our system, we construct representations of signal vectors using
-Generalized Algebraic Datatypes~\cite{Cheney2003,Xi2003}. The use of GADTs
-enables us to use signal vectors to instantiate type parameters in the types
-of signal vector representations, as well as in the types of signal functions.
-GADTs are available as an extension in the Glasgow Haskell Compiler~\cite{PeytonJones2006}.
-
-This approach permits us to construct partial inputs and outputs for signal
-functions. For instance, we can construct a representation for a single event
-in a signal vector, and another representation of updated values for a subset of
-the signals in the signal vector, and yet another carrying values for all
-signals in a signal vector. The signal vector types are shown in
-Figure~\ref{figure:signal_vectors}.
-
-\begin{figure}[t]
-\begin{code}
-data SVEmpty
-data SVSignal a
-data SVEvent a
-data SVAppend svLeft svRight
-\end{code}
-\hrule
-\caption{Signal vectors.}
-\label{figure:signal_vectors}
-\end{figure}
-
-With the ability to construct partial representations of signal vectors, we can
-represent signal functions with a datatype carrying multiple functions, one for
-each type of input. This allows us to separate the pull-based processing of
-signals from the push-based reaction to events.
-
-The exposed interface is a set of combinators for constructing signal functions,
-as well as combinators for describing the evaluation of a signal function. The
-interface is shown in Fig.~\ref{figure:signal_function_interface}.
-
-\begin{figure}
-\begin{code}
-
--- Signal Functions
-type :~> svIn svOut
-
--- Infix type alias for SVAppend
-type :^: svLeft svRight = SVAppend svLeft svRight
-
--- Basic signal functions
+\section{System Design and Interface}
+\label{section:System_Design_and_Interface}
+
+\subsection{Goals}
+\label{section:System_Design_and_Interface-Goals}
+
+The goal of FRP is to provide an efficient, declarative abstraction for creating
+reactive programs. Towards this overall goal, there are three goals which this
+system is intended to meet.
+
+Efficient evaluation is the motivation for push-based evaluation of events.
+Since FRP programs are expected to  interact with an outside world in real time,
+efficiency cannot simply be measured by the runtime of a program. Thus, when speaking of efficiency,
+we are expressing a desire that the system utilize as few system resources as possible
+for the task at hand, while responding as quickly as possible to external inputs and
+producing output at a consistently high sample rate.
+
+A composable abstraction is one in which values in that abstraction may be
+combined in such a way that reasoning about their combined actions involves
+little more than reasoning about their individual actions. In a signal function
+system, the only interaction between composed signal functions ought to be that
+the output of one is the input of another. Composability permits a particularly
+attractive form of software engineering in which successively larger systems are
+created from by combining smaller systems, without having to reason about the 
+implementation of the components of the systems being combined.
+
+It is fine for a system to be composable with regards to itself, but an FRP
+system must interact with the outside world. Since we cannot anticipate every
+possible form of input and output that the system will be asked to interact
+with, we must interface with Haskell's IO system. In particular, most libraries
+for user interaction (e.g. GUI and graphics libraries such as GTK+ and GLUT) and
+most libraries for time-dependent IO (e.g. audio and video systems) make use of
+the event loop abstraction. In this abstraction, event handlers are registered
+with the system, and then a command is issued to run a loop which detects events
+and runs the handlers, and uses the results of the handlers to render the
+appropriate output.
+
+We would like for the FRP system to be easy to integrate with such IO systems,
+while being flexible enough to enable its use with other forms of IO systems,
+such as simple imperative systems, threaded systems, or network servers.
+
+\subsection{Semantics}
+\label{subsection:System_Design_and_Interface-Semantics}
+
+A rigorous and formal elucidation of the semantics of signal-function FRP remains
+unattempted, but there is a sufficient change to the practical semantics of
+signal-function FRP between previous signal-function systems and TimeFlies to warrant
+some description.
+
+In previous systems such as Yampa, events were understood (and implemented) as option-valued
+signals. This approach is undesirable for several reasons. The most pressing reason is
+that it prohibits push-based evaluation of events, because events are embedded in the
+samples of a signal and must be searched for when sampling.
+
+Another concern is that this approach limits the event rate to the sampling rate.
+The rate of sampling should, at some level, not matter to the FRP system. Events
+which occur between sampling intervals are never observed by the system. Since
+the signal function is never evaluated except at the sampling intervals, no
+events can otherwise be observed.
+
+This concern drives the next concern. Events are not instantaneous in this formulation.
+If a signal is option valued, the sampling instant must fall within the interval where
+there is an event occurrence present for that event to be observed. If events are
+instantaneous, the probability of observing an event occurrence is zero.
+
+Therefore, TimeFlies employs the N-Ary FRP type formulation to represent signals and
+events as distinct entities in the inputs and outputs of signal functions. This means
+we are now free to choose our representation of events, and to separate it from the
+representation and evaluation of signals.
+
+This freedom yields the additional ability to make events independent of the sampling
+interval altogether. The semantics of event handling in TimeFlies is that an event occurrence
+is responded to immediately, and does not wait for the next sampling instant. This allows events
+to be instantaneous, and further, allows multiple events to occur within a single sampling interval.
+
+There are two tradeoffs intrinsic to this approach. The first is that events are only partially ordered
+temporally. There is no way to guarantee the order of observation of event occurrences occurring in the
+same sampling interval. Further, the precise time of an event occurrence cannot be observed, only the 
+time of the last sample prior to the occurrence.
+
+In return for giving up total ordering and precise observation of the timing of events, we obtain the
+ability to employ push-based evaluation for event occurrences, and the ability to non-deterministically
+merge event occurrences. When events being input to a non-deterministic merge have simultaneous occurrences,
+we arbitrarily select one to occur first. This does not violate any guarantee about time values, since
+they will both have the same time value in either case, and does not violate any guarantee about ordering,
+since no guarantee of their order is given.
+
+A formal semantic description of signal function FRP would clarify the consequences of this decision somewhat,
+but is outside the scope of this thesis.
+
+\subsection{Types}
+\label{subsection:System_Design_and_Interface-Types}
+
+In a strongly and statically typed functional language, types are a key part of
+an interface. Types provide a mechanism for describing and ensuring properties
+of the interface's components and about the systems created with these
+components.
+
+In order to type signal functions, we must be able to describe their input and
+output. In most signal function systems, a signal function takes exactly one
+input and produces exactly one output. Multiple inputs or outputs are handled
+by making the input or output a tuple, and combinators which combine or split
+the inputs or outputs of a signal assume this. Events are represented at the
+type level as a particular type of signal, and at the value level as an option,
+either an event occurrence or not.
+
+This method of typing excludes push-based evaluation at the outset.
+It is not possible to construct a ``partial tuple'' nor in general is it
+possible to construct only part of any type of value. Push-based evaluation
+depends on evaluating only that part of the system which is updated, which means
+evaluating only that part of the input which is updated.
+
+In order to permit the construction of partial inputs and outputs, we make use
+of signal vectors. Signal vectors are uninhabited types which describe the input
+and output of a signal function. Singleton vectors are parameterized over the
+type carried by the signal or by event occurrences. The definition of the signal
+vector type is shown in Figure~\ref{figure:signal_vector_types}.
+
+Having an uninhabited signal vector type allows us to construct representations
+of inputs and outputs which are hidden from the user of the system, and are
+designed for partial representations.
+
+\begin{figure}
+\begin{code}
+data SVEmpty    -- An empty signal vector component,
+                -- neither event nor signal
+data SVSignal a -- A signal, carrying values of type a
+data SVEvent a  -- An event, whose occurrences carry values of type a
+data SVAppend svLeft svRight -- The combination of the signal vectors
+                             -- svLeft and svRight
+\end{code}
+\hrule
+\caption{Signal vector types.}
+\label{figure:signal_vector_types}
+\end{figure}
+
+The type constructor for signal functions is shown in
+Figure~\ref{figure:signal_function_types}. For the {\tt init} parameter, only
+one possible instantiation is shown. The usefulness of this type parameter,
+along with another instantation which is hidden from users of the library,
+is discussed in the section on implementation of signal functions
+(Section~\ref{section:Implementation}).
+
+Signal functions with signal vectors as input and output types form a
+Haskell {\tt GArrow}~\cite{Megacz2011}. Specifically, the signal function
+type constructor (with the initialization parameter fixed) forms the arrow
+type, the {\tt SVAppend} type constructor forms the product type, and the
+{\tt SVEmpty} type constructor forms the unit type.
+
+The representation of signal functions is discussed in
+Section~\ref{section:Implementation-Signal_Functions}. The type synonyms
+{\tt :\textasciitilde>} and {\tt :\textasciicircum:} are included for
+readability and are not crucial to the FRP system.
+
+\begin{figure}
+\begin{code}
+-- Signal functions
+-- init: The initialization type for 
+--   the signal function, always NonInitialized
+--   for exported signal functions
+-- svIn: The input signal vector
+-- svOut: The output signal vector
+data SF init svIn svOut
+
+data NonInitialized
+
+type svIn :~> svOut = SF NonInitialized svIn svOut
+type svLeft :^: svRight = SVAppend svLeft svRight
+\end{code}
+\hrule
+\caption{Signal function types.}
+\label{figure:signal_function_types}
+\end{figure}
+
+A {\em monad} is a standard, composable abstraction for writing functions with
+a context, used in Haskell for IO~\cite{PeytonJones1993,PeytonJones2001} and
+other tasks. A monad is simply a 1-arity type constructor together with two
+functions. The first function, {\tt return}, takes a value of type {\tt a} and
+produces a value of type {\tt m a}. The second, called {\tt bind} and stylized
+in the Haskell standard library as the infix operator {\tt (>>=)}, takes a value
+of type {\tt m a} and a function of type {\tt a -> m b} and produces a value of
+type {\tt m b}. This allows a value to be operated on out of the context and a
+new context to be assigned.
+
+A monad can have other primitives which manipulate the context in some way. For 
+instance, the primtives in Haskell's {\tt IO} monad produce actions which, when
+interpreted as part of the {\tt main} action, produce some side-effect. The
+{\tt State} monad provides {\tt get} and {\tt put} operations to work with a 
+state value stored in the context.
+
+Monad transformers~\cite{Jones1995} provide a means to combine the functionality
+of multiple monads. A monad transformer is a monad with an extra type parameter.
+This type parameter is instantiated with the type constructor of the underlying
+monad, and an extra operation {\tt lift} is provided which converts values in
+the underlying monad to values in the monad transformer.
+
+The evaluation monad is a monad transformer. This permits it to be used in
+conjunction with the {\tt IO} monad (or any other monad) to describe how input
+is to be obtained for the signal function being evaluated, and how outputs are
+to be handled.
+
+The evaluation monad, in addition to the standard monad operators, provides a
+means of {\em initializing} a signal function, and a means of translating the
+monadic value describing evaluation to a value in the underlying monad. This
+means, for instance, that we can obtain an action in the {\tt IO} monad to
+evaluate a signal function.
+
+The type of the evaluation monad must track the input and output type of the
+signal function. The monad's context stores a mapping from outputs to handling
+actions. There are thus four type parameters to the monad's type
+constructor: the input signal vector, the output signal vector, the type of the
+underlying monad, and the monadic type parameter. The type is shown in
+Figure~\ref{figure:evaluation_monad_types}.
+
+\begin{figure}
+\begin{code}
+-- A signal function's evaluations state
+data SFEvalState svIn svOut m
+-- The evaluation monad
+data SFEvalT svIn svOut m a
+
+instance Monad m => Monad (SFEvalT svIn svOut m)
+\end{code}
+\hrule
+\caption{Evaluation monad types.}
+\label{figure:evaluation_monad_types}
+\end{figure}
+
+\subsection{Combinators}
+\label{section:System_Design_and_Interface-Combinators}
+
+Signal functions are constructed from combinators, which are primitive signal
+functions and operations to combine these primitives. These combinators are
+grouped as basic signal functions, lifting operations for pure functions,
+routing, reactivity, feedback, event processing, joining, and time dependence.
+
+The basic signal functions (Figure~\ref{figure:basic_signal_functions})
+provide very simple operations. The {\tt identity} signal function, as expected,
+simply copies its input to its output. The {\tt constant} signal function
+produces the provided value as a signal at all times. The {\tt never} signal
+function has an event output which never produces occurrences. The {\tt asap}
+function produces an event occurrence with the given value at the first time
+step after it is switched into the network. The {\tt after} function waits for
+the specified amount of time before producing an event occurrence.
+
+With the exception of {\tt identity}, all of the basic signal functions have
+empty inputs. This allows these signal functions to be used to insert values
+into the network which are known when the signal function is created, without
+having to route those values from an input.
+
+\begin{figure}
+\begin{code}
+-- Pass the input unmodified to the output
 identity :: sv :~> sv
+
+-- Produce a signal which is at all times the supplied value
 constant :: a -> SVEmpty :~> SVSignal a
+
+-- An event with no occurrences
 never    :: SVEmpty :~> SVEvent a
+
+-- An event with one occurrence, as soon as possible after
+-- the signal function is initialized
 asap     :: a -> SVEmpty :~> SVEvent a
+
+-- An event with one occurrence
+-- after the specified amount of time has elapsed.
 after    :: Double -> a -> SVEmpty :~> SVEvent a
-
--- Lifting pure functions
-pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal a
-pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent a
-
--- Composition and routing
-(>>>)           :: (svIn :~> svMiddle) -> (svMiddle :~> svOut) 
-                   -> (svIn :~> svOut)
+\end{code}
+\hrule
+\caption{Basic signal function combinators.}
+\label{figure:basic_signal_functions}
+\end{figure}
+
+Two combinators are provided to lift pure functions to signal functions
+(Figure~\ref{figure:lifting_pure_functions}). The {\tt pureSignalTransformer}
+combinator applies the pure function to a signal at every sample point. The
+{\tt pureEventTransformer} combinator applies the function to each occurrence of
+an input event.
+
+\begin{figure}
+\begin{code}
+-- Apply the given function to a signal at all points in time
+pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal b
+
+-- Apply the given function to each event occurrence
+pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent b
+\end{code}
+\hrule
+\caption{Lifting pure functions.}
+\label{figure:lifting_pure_functions}
+\end{figure}
+
+The routing combinators are used to combine signal functions, and
+to re-arrange signal vectors in order to connect signal functions.
+The routing combinators are shown in Figure~\ref{figure:routing_combinators}.
+
+Only those combinators which modify or combine signal functions
+({\tt (>>>)}, {\tt first}, {\tt second}) are reactive (may replace themselves in
+response to events, and then only if they inherit their reactivity from the
+signal function(s) which are inputs to combinator. The rest do not
+react to or modify the input in any way, except to re-arrange it, copy it, or
+discard it altogether.
+
+\begin{figure}
+\begin{code}
+-- Use the output of one signal function as the input for another
+(>>>) :: (svIn :~> svBetween) -> (svBetween :~> svOut) -> svIn :~> svOut
+
+-- Pass through the right side of the input unchanged
 first :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
+
+-- Pass through the left side of the input unchanged
 second :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut)
+
+-- Swap the left and right sides
 swap :: (svLeft :^: svRight) :~> (svRight :^: svLeft)
+
+-- Duplicate the input
 copy :: sv :~> (sv :^: sv)
-ignore          :: sv :~> SVEmpty
+
+-- Ignore the input
+ignore :: sv :~> svEmpty
+
+-- Remove an empty vector on the left
 cancelLeft :: (SVEmpty :^: sv) :~> sv
+
+-- Add an empty vector on the left
+uncancelLeft :: sv :~> (SVEmpty :^: sv)
+
+-- Remove an empty vector on the right
 cancelRight :: (sv :^: SVEmpty) :~> sv
-uncancelLeft    :: sv :~> (SVEmpty :^: sv)
+
+-- Add an empty vector on the right
 uncancelRight :: sv :~> (sv :^: SVEmpty)
+
+-- Make right-associative
 associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))
+
+-- Make left-associative
 unassociate :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
-
--- Reactivity
+\end{code}
+\hrule
+\caption{Routing combinators.}
+\label{figure:routing_combinators}
+\end{figure}
+
+Reactivity is introduced by means of the {\tt switch} combinator
+(Figure~\ref{figure:switch_combinator}). The design of this combinator
+allows modular packaging of reactivity. A signal function can determine
+autonomously when to replace itself, based only on its input and state,
+by emitting an event occurrence carrying its replacement. The combinator
+consumes and hides the event carrying the replacement signal function, 
+so the reactivity is not exposed by the resulting reactive signal function.
+
+\begin{figure}
+\begin{code}
 switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
           -> svIn :~> svOut
-
--- Feedback
-loop :: ((svIn :^: svLoop) :~> (svOut :^: svLoop)) -> svIn :~> svOut
-
--- Time dependence
-time  :: SVEmpty :~> Double
+\end{code}
+\hrule
+\caption{Reactivity combinator.}
+\label{figure:switch_combinator}
+\end{figure}
+
+There are other formulations of a reactive combinator found in FRP literature
+and libraries, which may be implemented using the one supplied. These are shown
+in Figure~\ref{figure:alternate_switching_combinators} and may be provided in a
+future version of the TimeFlies library.
+
+\begin{figure}
+\begin{code}
+-- Alternate version of switch,
+-- implemented in terms of supplied version
+switch_gen ::    (svIn :~> (svOut :^: SVEvent a))
+              -> (a -> svIn :~> svOut)
+              -> svIn :~> svOut
+switch_gen sf f =
+  switch (sf >>> second (pureEventTransformer f))
+
+-- Supplied version in terms of alternate version
+switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
+          -> svIn :~> svOut
+switch sf = switch_gen sf id
+
+-- Repeated switch, which takes replacement signal functions
+-- externally.
+rswitch ::    (svIn :~> svOut)
+           -> (svIn :^: SVEvent (svIn :~> svOut)) :~> svOut
+rswitch sf =
+  switch (first sf >>> second (pureEventTransformer rswitch))
+\end{code}
+\hrule
+\caption{Alternate reactivity combinators.}
+\label{figure:alternate_switching_combinators}
+\end{figure}
+
+It is often useful for a signal function to receive a portion of its
+own output as input. This is especially useful when we have two
+signal functions which we would like to mutually interact. We cannot
+just serially compose them, we must also bring the output of the second
+back around to the first. Many signal-processing algorithms also depend
+on feedback. The combinator which provides this ability is shown
+in Figure~\ref{figure:feedback_combinator}.
+
+\begin{figure}
+\begin{code}
+loop ::    ((svIn :^: svLoop) :~> (svOut :^: svLoop))
+        -> svIn :~> svOut
+\end{code}
+\hrule
+\caption{Feedback combinator.}
+\label{figure:feedback_combinator}
+\end{figure}
+
+This combinator provides decoupling for signals
+(the input signal is the output signal at the previous time-step)
+but not events (event occurrences are supplied to the combinator immediately).
+This means that the programmer has the responsibility to ensure that feedback
+does not generate an infinite sequence of events in a single time-step.
+
+Several combinators are provided for manipulating, suppressing, and generating events.
+Each of the combinators has an option variant and a list variant. The option variant
+produces an output event occurrence whenever the application of the supplied function
+to the input event produces a value. The list version produces an event occurrence for
+each of the elements of the output list. The combinators are shown in Figure~\ref{figure:event_specific_combinators}.
+
+\begin{figure}
+\begin{code}
+-- Apply the function to each input occurrence,
+-- and produce an occurrence for each Just.
+filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
+
+-- Apply the function to each input occurrence,
+-- and produce an occurrence for each list element
+filterList :: (a -> [b]) -> SVEvent a :~> SVEvent b
+
+-- Apply the function to the stored accumulator
+-- and the event occurrence, replacing the accumulator
+-- and possibly outputting an occurrence
+accumulate ::    (a -> b -> (Maybe c, a))
+              -> a
+              -> SVEvent b :~> SVEvent c
+
+-- Apply the function to the stored accumulator
+-- and the event occurrence, replacing the
+-- accumulator and outputting an event occurrence
+-- for each element of the list
+accumulateList ::    (a -> b -> ([c], a))
+                  -> a
+                  -> SVEvent b :~> SVEvent c
+\end{code}
+\hrule
+\caption{Event-specific combinators.}
+\label{figure:event_specific_combinators}
+\end{figure}
+
+The filter combinators are stateless, and thus apply the function to only the new
+input value. They are useful for suppressing events, as well as for extracting one
+of multiple cases of a datatype. For instance, a splitter for events carrying
+{\tt Either}-valued occurrences could be written as:
+
+\begin{code}
+getLeft :: Either a b -> Maybe a
+getLeft (Left x) = Just x
+getLeft _ = Nothing
+
+getRight :: Either a b -> Maybe b
+getRight (Right x) = Just x
+getRight _ = Nothing
+
+split :: SVEvent (Either a b) :~> (SVEvent a :^: SVEvent b)
+split = copy >>> first (filter getLeft) >>> second (filter getRight)
+\end{code}
+
+The accumulate combinators are stateful, applying the supplied function
+to both the input value and an accumulator. This function has two results:
+the option or list of output event occurrence values, and the new value
+for the accumulator.
+
+The accumulator is useful when responses to multiple event occurrences
+(from one or more sources) must be coordinated. For instance, in the
+benchmark application (see Chapter~\ref{chapter:Evaluation_and_Comparisons}),
+a table is maintained that allows knowledge from previous event occurrences
+(packets from a network switch) to be used in deciding where the present
+packet ought to go.
+
+The joining combinators provide the ability to combine two event
+streams, two signals, or a signal and an event stream. These
+combinators are shown in Figure~\ref{figure:joining_combinators}
+
+The {\tt union} combinator is a non-deterministic merge of event
+streams. Any event which occurs on either input will occur
+on the output. For simultaneous event occurrences, the order of occurrence
+is not guaranteed, but the occurrence itself is. This construct
+is also guaranteed to respect the relationship of event occurrences to sampling
+intervals.
+
+The {\tt combineSignals} combinator applies a binary function pointwise to two
+signals, and produces the result of this application as a third signal. The
+combining function is necessary because we will have two input samples at each
+time step, and must produce exactly one output sample.
+
+The {\tt capture} combinator adds the last-sampled value of a signal at the time
+of an event occurrence to that event occurrence.
+
+These three combinators together provide the ability to combine elements of
+signal vectors. By combining these combinators, along with the {\tt cancelLeft}
+and {\tt cancelRight} routing combinators, arbitrary signal vectors can be
+reduced.
+
+\begin{figure}
+\begin{code}
+union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
+combineSignals :: (a -> b -> c) -> (SVSignal a :^: SVSignal b) :~> SVSignal c
+capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent (b, a)
+\end{code}
+\hrule
+\caption{Joining combinators.}
+\label{figure:joining_combinators}
+\end{figure}
+
+A set of combinators are provided for making use of time-dependence in a signal
+function. These combinators allow the modification of signals and events with
+respect to time, and the observation of the current time value. The combinators
+are shown in Figure~\ref{figure:time_combinators}
+
+\begin{figure}
+\begin{code}
+time      :: SVEmpty :~> SVSignal Double
 delay     :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
-class TimeIntegrate
-
--- Joining
-union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
-combineSignals :: (a -> b -> c) 
-                  -> (SVSignal a :^: SVSignal b) :~> SVSignal c
-capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent a
-
--- Events
-filter         :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-\end{code}
-\hrule
-\caption{Signal function interface.}
-\label{figure:signal_function_interface}
-\end{figure}
-
-Signal functions are produced by combining primitive signal functions using
-the {\tt >>>} (sequential composition), {\tt first}, and {\tt second} routing
-combinators. Other routing signal functions are provided, but they are intended
-to be combined with, rather than to modify, other signal functions.
+integrate :: TimeIntegrate i => SVSignal i :~> SVSignal i
+\end{code}
+\hrule
+\caption{Time-dependent combinators.}
+\label{figure:time_combinators}
+\end{figure}
+
+The simplest time-dependent combinator is {\tt time}, which simply outputs
+the time since it began to be evaluated. This does not necessarily correspond to
+the time since the global signal function began to be evaluated, since the
+signal function in which the {\tt time} combinator is used may have been
+introduced through {\tt switch}.
+
+The {\tt delay} signal function allows events to be delayed. An initial delay
+time is given, and event occurrences on the right input can carry a new delay
+time. Event occurrences on the left input are stored and output when their delay
+time has passed. Changing the delay time does not affect occurrences already
+waiting.
+
+The {\tt integrate} combinator outputs the rectangle-rule approximation of the
+integral of its input signal with respect to time.
 
 \subsection{Evaluation Interface}
-The evaluation interface provides a monad~\cite{PeytonJones1993,PeytonJones2001}
-for specifying input to signal functions and handling their output. This
-interface also allows input actions to be separated from output actions, and
-from each other, so that a signal function may receive input from multiple IO
-systems and have its output handled by multiple output systems.
-
-The evaluation interface is a monad transformer~\cite{Liang1995}. This allows
-a signal function evaluation to be constructed from primitive evaluation actions
-(event pushing, input signal updating, and sampling) as well as actions from the
-underlying monad. These actions can then be run in the underlying monad
-to actuate the signal function. Because the actuation takes a signal function's
-state as input and produces a new state as output, in addition to the monadic
-side effects of handling the signal function's output, different evaluation
-actions may be taken on the same signal function from distinct locations within
-an application's code. This allows the evaluation interface to be easily
-integrated with event-loop style systems as well as traditional imperative
-IO systems.
-
-The evaluation interface consists of the monad transformer type, functions for
-constructing event inputs, initial input signal samples, and input signal
-updates, functions for constructing the vector of output handlers for a signal
-function, and evaluation actions. The full interface is shown in Fig.~\ref{figure:evaluation_interface}.
-
-\begin{figure}
-\begin{code}
--- Input helpers
-class SVRoutable -- Instances: SVSignalUpdate,
-                 -- SVEventOccurrence
-svLeft :: (SVRoutable r) => 
-          r svLeft -> r (SVAppend svLeft svRight)
-svRight :: (SVRoutable r) =>
-           r svRight -> r (SVAppend svLeft svRight)
-
--- Signal inputs
-data SVSignalUpdate
-data SVSample
+\label{subsection:System_Design_and_Interface-Evaluation_Interface}
+
+The evaluation interface provides a modified state monad which holds a signal
+function, together with some additional information, as its state (shown in Figure~\ref{figure:evaluation_state}.
+Rather than monadic instructions to put and get the state, the monad provides instructions
+to trigger an input event, update an input signal, and trigger sampling of
+signals in the signal function. Additional state includes the current set of
+modifications to the input signals (since the last sample) and a set of
+handlers which actuate effects based on output events or changes to the output
+signal.
+
+\begin{figure}
+\begin{code}
+-- A vector of handlers for outputs
+data SVHandler out sv
+
+-- A dummy handler for an empty output
+emptyHandler    :: SVHandler out SVEmpty
+
+-- A handler for an updated signal sample
+signalHandler   :: (a -> out) -> SVHandler out (SVSignal a)
+
+-- A handler for an event occurrence
+eventHandler    :: (a -> out) -> SVHandler out (SVEvent a)
+
+-- Combine handlers for a vector
+combineHandlers ::    SVHandler out svLeft
+                   -> SVHandler out svRight
+                   -> SVHandler out (svLeft :^: svRight)
+
+-- The state maintained when evaluating a signal function
+data SFEvalState m svIn svOut
+
+-- Create the initial state for evaluating a signal function
+initSFEval ::    SVHandler (m ()) svOut   -- Output handlers
+
+              -> SVSample svIn            -- Initial input samples
+
+              -> Double                   -- Initial external time,
+                                          -- corresponding to time 0 for
+                                          -- the signal function
+
+              -> (svIn :~> svOut)         -- Signal function to evaluate
+              -> SFEvalState m svIn svOut
+\end{code}
+\hrule
+\caption{State maintained when evaluating a signal function.}
+\label{figure:evaluation_state}
+\end{figure}
+
+Sets which correspond to signal vectors are built with ``smart'' constructors.
+For instance, to construct a set of handlers, individual handling functions are
+lifted to handler sets with the {\tt signalHandler} and {\tt eventHandler}
+functions, and then combined with each other and {\tt emptyHandler} leaves
+using the {\tt combineHandlers} function.
+
+Building the initial input sample is similar, but {\tt sampleEvt} leaves do
+not carry a value.
+
+In order to initialize the state, the user must supply a set of handlers, the
+signal function to evaluate, and initial values for all of the signal inputs
+(Figure~\ref{figure:initial_input}).
+
+\begin{figure}
+\begin{code}
+-- A sample for all leaves of a signal vector
+data SVSample sv
+
+-- Create a sample for a signal leaf
 sample          :: a -> SVSample (SVSignal a)
+
+-- A dummy sample for an event leaf
 sampleEvt       :: SVSample (SVEvent a)
+
+-- A dummy sample for an empty leaf
 sampleNothing   :: SVSample SVEmpty
-combineSamples :: SVSample svLeft -> svSample svRight 
-                  -> svSample (svLeft :^: svRight)
+
+-- Combine two samples
+combineSamples  ::    SVSample svLeft
+                   -> SVSample svRight
+                   -> SVSample (svLeft :^: svRight)
+\end{code}
+\hrule
+\caption{Data type for initial input}
+\label{figure:initial_input}
+\end{figure}
+
+This state can then be passed to a monadic action which will supply input to
+the signal function. Inputs are constructed using a simple interface with
+functions to construct sample updates and event occurrences, and to specify
+their place in the vector (Figure~\ref{figure:ongoing_input}).
+
+\begin{figure}
+\begin{code}
+-- Class to overload left and right functions
+class SVRoutable r where
+  svLeft          :: r svLeft -> r (svLeft :^: svRight)
+  svRight         :: r svRight -> r (svLeft :^: svRight)
+
+-- An input event occurrence
+data SVEventInput sv
+instance SVRoutable SVEventInput sv
+
+-- An updated sample for a signal
+data SVSignalUpdate sv
+instance SVRoutable SVSignalUpdate sv
+
+-- Create an event occurrence
+svOcc           :: a -> SVEventInput (SVEvent a)
+
+-- Create an updated sample
 svSig           :: a -> SVSignalUpdate (SVSignal a)
-
--- Event inputs
-data SVEventInput
-svOcc :: a -> SVEventInput (SVEvent a)
--- Actuation
-update :: (Monad m) => SVSignalUpdate svIn 
-           -> SFEvalT svIn svOut m ()
-push   :: (Monad m) => SVEventInput svIn
-          -> SFEvalT svIn svOut m ()
-sample :: SVEvalT svIn svOut m ()
--- Running
-data SVHandler
-emptyHandler :: SVHandler m SVEmpty
-eventHandler :: (a -> m ()) 
-                -> SVHandler m (SVEvent a)
-signalHandler :: (a -> m ()) 
-                 -> SVHandler m (SVSignal a)
-combineHandlers :: SVHandler m svLeft
-                   -> SVHandler m svRight
-                   -> SVHandler m (svLeft :^: svRight)
-initSFEvalT :: SVHandler m svOut -> SVSample svIn 
-               -> Double ->(svIn :~> svOut)
-               -> SFEvalState m svIn svOut
+\end{code}
+\hrule
+\caption{Data types for ongoing input.}
+\label{figure:ongoing_input}
+\end{figure}
+
+The {\tt SFEvalT} monad is actually a monad transformer, that is, it is
+parameterized over an underlying monad whose actions may be lifted to
+{\tt SFEvalT}. In the usual case, this will be the {\tt IO} monad.
+
+{\tt SFEvalT} actions are constructed using combinators to push events,
+update inputs, and step time, as well as actions lifted from the underlying
+monad (used to obtain these inputs). An action in the underlying monad
+which produces the result and a new state is obtained with the {\tt runSFEvalT}
+function. These combinators are shown in Figure~\ref{figure:evaluation_combinators}.
+
+\begin{figure}
+\begin{code}
+-- The evaluation monad
+data SFEvalT svIn svOut m a
+instance MonadTrans (SFEvalT svIn svOut)
+instance (Monad m) => Monad (SFEvalT svIn svOut m)
+instance (Functor m) => Functor (SFEvalT svIn svOut m)
+instance (Monad m, Functor m) => Applicative (SFEvalT svIn svOut m)
+instance (MonadIO m) => MonadIO (SVEvalT svIn svOut m)
+
+-- Obtain an action in the underlying monad
+-- from an SFEvalT and a new state.
 runSFEvalT ::    SFEvalT svIn svOut m a
               -> SFEvalState m svIn svOut
-              -> m a
-\end{code}
-\hrule
-\caption{Evaluation interface.}
-\label{figure:evaluation_interface}
+              -> m (a, SFEvalState m svIn svOut)
+
+-- Push an event occurrence.
+push :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
+
+-- Update the value of an input signal sample
+-- (not immediately observed)
+update :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
+
+-- Step forward in time, observing the updated signal values
+step :: (Monad m) => Double -> SFEvalT svIn svOut m ()
+\end{code}
+\hrule
+\caption{Evaluation combinators}
+\label{figure:evaluation_combinators}
 \end{figure}
@@
 
-\subsection{Input and Output Representations}
-\label{subsection:Input_and_Output_Representations}
-As discussed in Section~\ref{section:System_Design}, our implementation requires
-representations of the inputs and outputs of signal functions. Since signal
-functions are typed using signal vectors, the representation types must be
-parameterized over signal vectors. But signal vectors are uninhabited types, and
-in ordinary Haskell we cannot declare a data constructor which uses a component
-of a type parameter. Further, we cannot restrict which types may instantiate a
-type parameter for a particular data constructor.
-
-GADTs lift these restrictions, permitting type constraints to be applied
-to individual data constructors. In the declaration of a GADT, a type signature
-is provided for each data constructor, including the types of the constructor's
-parameters. When a GADT is pattern-matched, the types of the parameters are
-inferred using the constraints given in this signature. This is called type
-refinement.
-
-Using GADTs, we can construct representations of signal vectors to use in our
-implementation.
-
-When constructing a representation, we consider how to represent each
-component of a signal vector, whether it is a signal element, an empty
-element, or an append node. For instance, in order to represent a signal
-sample for all signals in an element, we construct the GADT shown in Figure~\ref{figure:signal_sample_representation}.
+\subsection{Signal Functions}
+\label{subsection:Implementation-Signal_Functions}
+
+The design of signal functions specifies a family of types for the inputs and
+outputs of signal functions. Signal functions are not functions in the purest
+sense, however. They are not mappings from a single instance of their input
+type to a single instance of their output type. They must be implemented with
+respect to the temporal semantics of their inputs and outputs.
+
+We therefore start by creating a set of concrete datatypes for the inputs and
+outputs of signal functions. These datatypes will be parameterized by the input
+and output types of the signal function, and will not be exposed to the user of
+the library. Rather, they will specify how data is represented during the
+temporal evaluation of signal functions.
+
+We then describe how signal functions are implemented using these concrete
+types, along with higher-order functions and continuations.
+
+In Section~\ref{section:System_Design_and_Interface} we presented signal vectors
+as a set of types. In order to be completely principled, we should isolate these
+types into their own {\em kind} (a sort of type of types); however, the Haskell
+extension for this was far from stable at the time this system was created.
+
+The types are therefore expressed in the system exactly as they were described
+in Section~\ref{section:System_Design_and_Interface}. (To refresh, see
+Figure~\ref{figure:signal_vector_types}.) The striking observation about these
+types is that they have {\em no data constructors}. There are no values which
+take these types.
+
+Instead, we will create concrete representations which are parameterized over
+these types. These concrete representations will be expressed as GADTs, allowing
+each data constructor of the representation to fill in a specific signal vector
+type for the parameter of the representation.
+
+The first thing to represent is {\em samples}, which are sets of values for
+the signal components of a signal vector. Therefore, we create a representation
+which carries a value for every {\tt SVSignal} leaf of a signal vector. In order
+to do this, we restrict each of our constructors to a single signal vector type.
+So there are three leaf constructors: {\tt SVSample}, which carries a value; and
+{\tt SVSampleEvent} and {\tt SVSampleEmpty}, which do not. This ensures that the
+only way to represent a sample leaf is with the {\tt SVSample} constructor,
+which carries a value of the appropriate type. The datatype is shown in
+Figure~\ref{figure:signal_sample_datatype}.
 
@@
 data SVSample sv where
-  SVSample        ::    a -> SVSample (SVSignal a)
-  SVSampleEvt     ::    SVSample (SVEvent a)
-  SVSampleNothing ::    SVSample SVEmpty
+  SVSample      ::    a
+                   -> SVSample (SVSignal a)
+  SVSampleEvent ::    SVSample (SVEvent a)
+  SVSampleEmpty ::    SVSample SVEmpty
   SVSampleBoth  ::    SVSample svLeft
                    -> SVSample svRight
-                     -> SVSample (svLeft :^: svRight)
-\end{code}
-\hrule
-\caption{A representation which carries a value for each signal in a signal vector.}
-\label{figure:signal_sample_representation}
-\end{figure}
-
-\subsection{Signal Function Representations}
-\label{subsection:Signal_Function_Representations}
-With representations for signal function inputs and outputs in place, we turn
-to the representation of signal functions. A signal function must be able to
-respond to time increments and new signal samples, as well as event occurrences.
-Since we wish to be able to ``push'' event occurrences independently of
-sampling, events are handled using the last-updated time and signal sample.
-
-When a signal function is asked to respond to either of these types of inputs,
-it must produce an output consisting of zero or more event occurrences, a new
-signal function with the same type (to enable reactivity and state), and, if
-it is responding to a time and sample update, a delta which represents updates
-to the sample of its output signals.
-
-Finally, there is a special case for a signal function at time 0. It must be
-provided with an initial input sample before it can respond to incremental time
-and sample updates or to event occurrences.
-
-Signal functions are represented as a datatype with two constructors. The first
-constructor wraps a function from an input sample to an output sample and an
-initialized signal function. The second wraps two functions. The first accepts
-a time delta and a signal delta, and produces a signal delta, a list of event
-occurrences, and a new signal function as the output. The second accepts an
-event occurrence and produces a list of event occurrences and a new signal
-function as the output. The signal function representation is shown in Figure~\ref{figure:signal_function_representation}.
-
-\begin{figure}[t]
+                   -> SVSample (SVAppend svLeft svRight)
+\end{code}
+\hrule
+\caption{Datatype for signal samples.}
+\label{figure:signal_sample_datatype}
+\end{figure}
+
+What about the event components? We want to represent event occurrences,
+each of which will correspond to at most one event in the vector. So a different
+representation is called for. In this case, there will be only three
+constructors. One constructor will represent an event leaf, and the other will
+represent a single value on the left or right side of the node ({\tt SVAppend}),
+ignoring all of the type structure on the other side. This representation
+describes a path from the root of the signal vector, terminating at an event
+leaf with a value.
+
+By pattern matching on the path constructors, we can determine which subvector
+of a signal vector an event occurrence belongs to, repeatedly refining it until
+we determine which event in the vector the occurrence corresponds to. The
+datatype for occurrences is shown in Figure~\ref{figure:event_occurrence_datatype}.
+
+\begin{figure}
+\begin{code}
+data SVOccurrence sv where
+  SVOccurrence ::    a
+                  -> SVOccurrence (SVEvent a)
+  SVOccLeft    ::    SVOccurrence svLeft
+                  -> SVOccurrence (SVAppend svLeft svRight)
+  SVOccRight   ::    SVOccurrence svRight 
+                  -> SVOccurrence (SVAppend svLeft svRight)
+\end{code}
+\hrule
+\caption{Datatype for event occurrences.}
+\label{figure:event_occurrence_datatype}
+\end{figure}
+
+We add one more representation for signals, in order to avoid uneccessary
+representations of the values of all signals when not all signals have changed
+their values. This representation allows us to represent the values of zero or
+more of the signals in a signal vector. To accomplish this, we replace the
+individual constructors for the {\tt SVEmpty} and {\tt SVEvent} leaves with %there is a better word, something about "useless but not really", for this
+a single, unconstrained constructor. This constructor can represent an arbitrary
+signal vector. We can use the constructor for signal vector nodes and the 
+constructor for sample leaves to represent the updated values, while filling
+in the unchanged portions of the signal vector with this general constructor.
+This datatype is shown in Figure~\ref{figure:signal_update_datatype}.
+
+\begin{figure}
+\begin{code}
+data SVDelta sv where
+  SVDeltaSignal  ::    a
+                    -> SVDelta (SVSignal a)
+  SVDeltaNothing ::    SVDelta sv
+  SVDeltaBoth    ::    SVDelta svLeft
+                    -> SVDelta svRight
+                    -> SVDelta (SVAppend svLeft svRight)
+\end{code}
+\hrule
+\caption{Datatype for signal updates.}
+\label{figure:signal_update_datatype}
+\end{figure}
+
+We now have concrete datatypes for an implementation to operate on. Our next
+task is to represent transformers of temporal data, which themselves may change
+with time. The common approach to this task is sampling, in which a program
+repeatedly checks for updated information, evaluates it, updates some state,
+and produces an output. This is the essence of pull-based evaluation.
+
+Another approach is notification, in which the program exposes an interface
+which the source of updated information may invoke. This is a repeated entry
+point to the program, which causes the program to perform the same tasks
+listed above, namely, evaluate the updated information, update state, and
+produce output. The strategy of notification as opposed to repeated checking is
+the essence of push-based evaluation.
+
+Signal functions are declarative objects, and not running processes. They have
+no way to invoke sampling themselves. They can, however, expose separate
+interfaces for when sampling is invoked, and when they are notified of an event
+occurrence. This creates two control paths through a signal function. One of
+these control paths is intended to be invoked regularly and frequently with
+updates to the time and sample values, and the other is intended to be invoked
+only when an event occurs. The benefit of separating these control paths is that
+events are no longer defined in terms of sampling intervals, and need not even
+be considered in sampling, except when they are generated by a condition on a
+sample. On the other hand, events can be responded to even if the time has not
+yet come for another sample, and multiple events can be responded to in a single
+sampling interval.
+
+We represent signal functions as a GADT with three type parameters and two 
+constructors. The first type parameter represents the initialization state,
+and is specialized to {\tt Initialized} or {\tt NonInitialized} depending on the
+constructor. The other two type parameters are the input and output signal
+vectors, respectively. The signal functions that a user will compose are\
+non-initialized signal functions. They must be provided with an initial set of
+input signals (corresponding to time zero). When provided with this input, they
+produce their time-zero output, and an initialized signal function. The datatype
+is shown in Figure~\ref{figure:signal_function_datatype}.
+
+\begin{figure}
 \begin{code}
 data Initialized
+
 data NonInitialized
 
-type :~> svIn svOut = SF NonInitialized svIn svOut
-
 data SF init svIn svOut where
-  SF :: (SVSample svIn -> (SVSample svOut, SF Initialized svIn svOut))
+  SF     ::    (SVSample svIn 
+                  -> (SVSample svOut,
+                      SF Initialized svIn svOut)) 
             -> SF NonInitialized svIn svOut
-  SFInit :: (Double -> SVDelta svIn 
-             -> (SVDelta svOut, [SVOccurrence svOut],
+  SFInit ::    (Double 
+                  -> SVDelta svIn
+                  -> (SVDelta svOut,
+                      [SVOccurrence svOut],
                       SF Initialized svIn svOut)) 
-            -> (SVOccurrence svIn -> ([SVOccurrence svOut],
+            -> (SVOccurrence svIn
+                  -> ([SVOccurrence svOut],
                       SF Initialized svIn svOut))
@@
 \hrule
-\caption{Signal function representation}
-\label{figure:signal_function_representation}
-\end{figure}
-
-This representation allows a signal function to respond to events and time
-updates separately, and does not enforce the representation of events during
-time updates, as previous signal function systems do.
-
-\subsection{Signal Function Implementations}
-\label{subsection:Signal_Function_Implementations}
-
-Now that we have established representations for signal functions and their
-inputs and outputs, we show several examples of how signal function combinators
-are implemented.
-
-The {\tt identity} signal function's initialization function takes a sample,
-and returns that sample as its output sample. The initialized function returned
-is the {\tt identityInit} function, which is not exported by the module.
-The time and sample function ignores the provided time delta, produces the
-input signal delta as the output signal delta, and returns {\tt identityInit} as
-the new signal function. The event function takes an event occurrence and
-returns a singleton list containing that occurrence, along with the
-{\tt identityInit} function as the replacement signal function. The
-implementation is shown in Figure~\ref{figure:identity_signal_function_implementation}.
-
-\begin{figure}[t]
+\caption{Datatype and empty types for signal functions.}
+\label{figure:signal_function_datatype}
+\end{figure}
+
+Initialized signal functions carry two continuations. The first continuation
+takes a time differential and a set of signal updates, and returns a set of
+signal updates, a collection of event occurrences, and a new initialized signal
+function of the same type. This is the continuation called when sampling.
+
+The second continuation takes an event occurrence, and returns a collection of
+event occurrences and a new signal function of the same type. This continuation
+is only called when there is an event occurrence to be input to the signal
+function.
+
+Note that each of these continuations uses one or more of the concrete
+representations of signal vectors, and applies the type constructor for the
+representation to the input or output signal vector for the signal function.
+
+Having specified a datatype for signal functions, we must now provide
+combinators which produce signal functions of this type. Each combinator's
+implementation must specify how it is initialized, how it samples its input, and
+how it responds to event occurrences.
+
+We will not detail every combinator here, but we will discuss each of the
+implementation challenges encountered.
+
+As an example of the implementation of combinators, we show the implementation
+of the {\tt identity} signal function in Figure~\ref{figure:identity_implementation}.
+This signal function simply passes all of its inputs along as outputs. The
+initialization function simply passes along the received sample and outputs the
+initialized version of the signal function. The initialized version of the input
+is similar, but is self-referential. It outputs itself as its replacement. This
+is standard for simple and routing combinators which are not reactive, and
+simply move samples and event occurrences around.
+
+\begin{figure}
 \begin{code}
 identity :: sv :~> sv
-identity = SF (\sample -> (sample, identityInit))
+identity =
+  SF (\initSample -> (initSample, identityInit))
 
 identityInit :: SF Initialized sv sv
-identityInit = SFInit (\_ delta -> (delta, [], identityInit))
-                      (\occ -> ([occ], identityInit))
-\end{code}
-\hrule
-\caption{{\tt identity} signal function implementation.}
-\label{figure:identity_signal_function_implementation}
-\end{figure}
-
-The {\tt constant} signal function is initialized with a value, which it then
-produces as its output forever. Thus, the initialized version of {\tt constant}
-never produces any output other than itself as a replacement
-signal function, and an empty signal delta. The implementation of {\tt constant}
-is shown in Fig.~\ref{figure:constant_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-constant :: a -> SVEmpty :~> SVSignal a
-constant x = SF (\_ -> (constantInit, sample x))
-
-constantInit :: SF Initialized SVEmpty (SVSignal a)
-constantInit = SFInit (\_ _ -> (deltaNothing, [], constantInit))
-                      (\_ -> ([], constantInit))
-\end{code}
-\hrule
-\caption{{\tt constant} signal function implementation.}
-\label{figure:constant_signal_function_implementation}
-\end{figure}
-
-The {\tt asap} signal function implementation requires a parameter for the
-initialized version of the signal function, to specify the value of the event
-occurrence. It replaces itself with the initialized version of the {\tt never}
-signal function after the first time step. The implementation is shown in
-Fig.~\ref{figure:asap_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-asap :: a -> SVEmpty :~> SVEvent a
-asap x = SF (\_ -> (sampleEvt, asapInit x))
-
-asapInit :: a -> SF Initialized SVEmpty (SVEvent a)
-asapInit x = SFInit (\_ _ -> (deltaNothing, [occurrence x], neverInit))
-                    (\_ -> (never))
-\end{code}
-\hrule
-\caption{{\tt asap} signal function implementation.}
-\label{figure:asap_signal_function_implementation}
-\end{figure}
-
-The {\tt filter} function accepts events, applies a {\tt Maybe} predicate to
-their values, and produces the value produced by the predicate as an event
-occurrence, or no occurrence if the value is {\tt Nothing}. It is implemented
-by having the event occurrence response function apply the {\tt maybe} function
-from the Haskell prelude to the value returned by the predicate, as shown in 
-Fig.~\ref{figure:filter_signal_function_implementation}
-
-\begin{figure}[t]
-\begin{code}
-filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-filter p = SF (\_ -> (sampleEvt, filterInit p))
-
-filterInit :: (a -> Maybe b) -> SF Initialized (SVEvent a) (SVEvent b)
-filterInit p = SFInit (\_ _ -> (deltaNothing, [], filterInit p))
-                      (\evtOcc -> (maybe [] ((:[]) . occurrence) $
-                                   fromOccurrence evtOcc,
-                                   filterInit p))
-\end{code}
-\hrule
-\caption{{\tt filter} signal function implementation.}
-\label{figure:filter_signal_function_implementation}
-\end{figure}
-
-The {\tt associate} signal function is a routing signal function. It transforms
-a signal vector which is left-associated at the top level to one that is
-right-associated at the top level. Its implementation is representative of all
-of the routing functions, and is shown in Fig.~\ref{figure:associate_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-associate :: SVAppend (SVAppend sv1 sv2) sv3
-             :~> SVAppend sv1 (SVAppend sv2 sv3)
-associate = 
-  SF (\sigSample -> let (sigSampleLeft, sigSampleRight) = 
-                          splitSample sigSample
-                        (sigSampleLeftLeft, sigSampleLeftRight) =
-                          splitSample sigSampleLeft
-                    in (combineSamples sigSampleLeftLeft 
-                        (combineSamples sigSampleLeftRight sigSampleRight),
-                        associateInit))
-
-associateInit :: SF Initialized (SVAppend (SVAppend sv1 sv2) sv3)
-                                (SVAppend sv1 (SVAppend sv2 sv3))
-associateInit = SFInit (\_ sigDelta -> let (sigDeltaLeft, sigDeltaRight) =
-                                             splitDelta sigDelta
-                                           (sigDeltaLeftLeft,
-                                            sigDeltaLeftRight) =
-                                             splitDelta sigDeltaLeft
-                                       in (combineDeltas sigDeltaLeftLeft
-                                           (combineDeltas sigDeltaLeftRight
-                                                          sigDeltaRight),
-                                           [], associateInit)) 
-                       (\evtOcc -> (case chooseOccurrence evtOcc of
-                                      Left leftOcc -> 
-                                        case chooseOccurrence leftOcc of
-                                          Left leftLeftOcc ->
-                                            [occLeft leftLeftOcc]
-                                          Right leftRightOcc ->
-                                            [occRight $ occLeft leftRightOcc]
-                                      Right rightOcc ->
-                                        [occRight $ occRight $ rightOcc],
-                                    associateInit))
-\end{code}
-\hrule
-\caption{{\tt associate} signal function implementation.}
-\label{figure:associate_signal_function_implementation}
-\end{figure}
-
-The {\tt switch} function is the basic combinator used to introduce reactivity.
-It is given a signal function whose output is the append of two signal vectors.
-The left side of the signal vector is passed on as the output of the reactive
-signal function, and the right side is an event carrying signal functions. When
-an event occurrence is present on the right-side event output, the signal
-function carried by this occurrence replaces the signal function constructed
-by switch. Due to space concerns, the implementation is elided, but a brief
-description will suffice. The signal function stores the input sample
-provided during initialization, and updates it with deltas. When an event
-occurrence carrying a signal function is produced by either the event or 
-time function of the wrapped signal function, the stored signal sample is used
-to initialize the new signal function. If this occurs while handling an event
-input, the sample output by the new signal function's initialization is stored
-and is combined with its first output delta to produce the output delta at the
-next time step.
-
-The {\tt loop} function allows a signal function to see a component of its own
-output as input. This is primarily useful when a signal function has components
-which mutually depend on each others outputs, such as in physics simulations
-or games. Care must be taken that the feedback output is not immediately
-dependent on the feedback input, or sampling the signal function will not
-terminate. Loop is implemented by generating a recursive list (which will
-be infinite if the feedback is not decoupled) of event inputs, and by splitting
-the output signal delta and using the right side output delta as the right
-side input delta within a recursive let-binding. (The let-binding in Haskell
-always admits recursive bindings.) If the system is not completely decoupled,
-this will result in non-termination during an evaluation step. Decoupling can
-be achieved using {\tt delay} primitive.
-
-\subsection{Evaluation Interface Implementation}
-\label{subsection:Evaluation_Interface_Implementation}
-
-The evaluation interface must maintain state including the current signal
-function, the current time (to produce time deltas), updates to the input
-sample which have yet to be sampled, and the output handlers.
-
-The output handlers are stored in a structure similar to that for a signal
-sample. The difference is that both signal and event leaves contain values, and
-these values are functions from the leaf type to another type. The handler
-datatype is shown in Fig.~\ref{figure:handler_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SVHandler out sv where
-  SVHandlerEmpty  :: SVHandler out SVEmpty
-  SVHandlerSignal :: (a -> out) -> SVHandler out (SVSignal a)
-  SVHandlerEvent  :: (a -> out) -> SVHandler out (SVEvent a)
-  SVHandlerBoth   :: SVHandler out svLeft -> SVHandler out svRight 
-                     -> SVHandler out (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Handler datatype.}
-\label{figure:handler_datatype}
-\end{figure}
-
-We now need a datatype to hold the various components of the evaluation state.
-This type is shown in Fig~\ref{figure:evaluation_state_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SFEvalState m svIn svOut 
-  = SFEvalState { 
-                  esSF :: SF Initialized svIn svOut,
-                  esOutputHandlers :: SVHandler (m ()) svOut,
-                  esLastTime :: Double,
-                  esDelta :: SVDelta svIn
-                }
-
-\end{code}
-\hrule
-\caption{Evaluation state datatype.}
-\label{figure:evaluation_state_datatype}
-\end{figure}
-
-The evaluation interface itself is a monad transformer, which we implement
-as a Haskell newtype wrapping the {\tt StateT} monad transformer. We do not
-export the raw {\tt put} and {\tt get} actions of the state monad, but instead
-implement the push, update, and sampling operations for signal function
-evaluation using {\tt put} and {\tt get}. The evaluation monad transformer is
-shown in Fig.~\ref{figure:signal_function_evaluation_monad_transformer}.
-Instances of typeclasses including {\tt Monad} and {\tt MonadTrans} are derived
-using the ``GeneralizedNewtypeDeriving'' extension to the Glasgow Haskell
-Compiler.
-
-\begin{figure}[t]
-\begin{code}
-newtype SFEvalT svIn svOut m a = StateT (SFEvalState m svIn svOut) m a
-\end{code}
-\hrule
-\caption{Signal function evaluation monad transformer.}
-\label{figure:signal_function_evaluation_monad_transformer}
-\end{figure}
+identityInit =
+  SFInit (\dt sigDelta -> (sigDelta, [], identityInit))
+         (\evtOcc -> ([evtOcc], identityInit))
+\end{code}
+\hrule
+\caption{Implementation of the {\tt identity} combinator.}
+\label{figure:identity_implementation}
+\end{figure}
+
+In order for our primitive signal functions to be useful, we need a means of
+composing them. Serial composition creates one signal function from two, by
+using the output of one as the input of the other. The serial composition
+combinator is styled {\tt (>>>)}. The implementation of this operator is one
+place where the advantage of responding to events independently from signal
+samples becomes clear.
+
+This is the only primitive combinator which takes two signal functions, and
+thus, it is the only way to combine signal functions. Parallel, branching, and
+joining composition can be achieved by modifying signal functions with the
+{\tt first} and {\tt second} combinators and composing them with the
+routing and joining combinators.
+
+Combinators which take one or more signal functions as input must recursively
+apply themselves, as is shown in the implementation of serial composition
+(Figure~\ref{figure:serial_composition_implementation}). They must also
+handle initialization, retaining the initialized signal functions and passing
+them to the initialized version of the combinator.
+
+\begin{figure}
+\begin{code}
+(>>>) ::    (svIn :~> svBetween) 
+         -> (svBetween :~> svOut)
+         -> (svIn :~> svOut)
+(SF sigSampleF1) >>> (SF sigSampleF2) =
+  SF (\sigSample -> let (sigSample', sfInit1) = sigSampleF1 sigSample
+                        (sigSample'', sfInit2) = sigSampleF2 sigSample'
+                    in (sigSample'', composeInit sfInit1 sfInit2))
+
+composeInit ::     SF Initialized svIn svBetween
+                -> SF Initialized svBetween svOut
+                -> SF Initialized svIn svOut
+composeInit (SFInit dtCont1 inputCont1) sf2@(SFInit dtCont2 inputCont2) =
+  SFInit
+    (\dt sigDelta -> 
+       let (sf1MemOutput, sf1EvtOutputs, sf1New) = dtCont1 dt sigDelta
+           (sf2MemOutput, sf2EvtOutputs, sf2New) = dtCont2 dt sf1MemOutput
+           (sf2EvtEvtOutputs, sf2Newest) = applySF sf2New sf1EvtOutputs
+       in (sf2MemOutput,
+           sf2EvtOutputs ++ sf2EvtEvtOutputs,
+           composeInit sf1New sf2Newest)
+    )
+    (\evtOcc -> 
+      let (sf1Outputs, newSf1) = inputCont1 evtOcc
+          (sf2FoldOutputs, newSf2) = applySF sf2 sf1Outputs
+      in (sf2FoldOutputs, composeInit newSf1 newSf2)   
+    )
+
+applySF ::    SF Initialized svIn svOut
+           -> [SVOccurrence svIn]
+           -> ([SVOccurrence svOut],
+               SF Initialized svIn svOut)
+applySF sf indices =
+  foldr (\evtOcc (changes, SFInit _ changeCont) ->
+           let (newChanges, nextSF) = changeCont evtOcc
+               in (newChanges ++ changes, nextSF))
+        ([], sf)
+        indices
+\end{code}
+\hrule
+\caption{Implementation of serial composition.}
+\label{figure:serial_composition_implementation}
+\end{figure}
+
+The switch combinator is the means of introducing reactivity into a signal
+function. This combinator allows a signal function to replace itself by
+producing an event occurrence. The combinator wraps a signal function, and 
+observes an event on the right side of the output signal vector. At the first
+occurrence of the event, the signal function carried by the occurrence replaces
+the signal function.
+
+The switch combinator stores the input sample provided during initialization,
+and updates it with the input signal updates. When the wrapped signal function
+produces an occurrence carrying a new signal function, that signal function is
+initialized with the stored input sample. It is then "wrapped" by another
+function which closes over its output sample, and outputs the sample as a signal
+update as the next time step. After this, it acts as the new signal function.
+This wrapping has some performance implications, which are discussed
+
+This combinator checks the outputs of the wrapped
+signal function for an event occurrence from which an uninitialized signal
+function is extracted. The switch combinator stores the full sample
+for its input vector (which is identical to the input vector of the supplied
+signal function) to initialize the new signal function. This also demands that
+it add a wrapper to the new signal function which waits for the next sampling
+interval and actuates the sample output at initialization as an output set
+of changes to the signal. This has some performance implications, which are
+discussed in Chapter~\ref{chapter:Evaluation_and_Comparisons}.
+
+Most of the routing combinators are simple to implement. The only task is to add
+remove, or replace routing constructors on signal updates and event occurrences.
+Since these signal functions are stateless and primitive, they can simply
+return themselves as their replacements. The {\tt swap} combinator is shown as
+an example in Figure~\ref{figure:swap_implementation}
+
+\begin{figure}
+\begin{code}
+-- T.swap is imported from Data.Tuple
+swap :: (sv1 :^: sv2) :~> (sv2 :^: sv1)
+swap =
+  SF ((, swapInit) . 
+      uncurry combineSamples .
+      T.swap . splitSample)
+
+swapInit :: SF Initialized (SVAppend sv1 sv2) (SVAppend sv2 sv1)
+swapInit =
+  SFInit (flip (const .
+                (, [], swapInit) .
+                uncurry combineDeltas .
+                T.swap . splitDelta))
+          (\evtOcc ->
+             (case chooseOccurrence evtOcc of
+                Left lOcc  -> [occRight lOcc]
+                Right rOcc -> [occLeft rOcc], swapInit))
+
+\end{code}
+\hrule
+\caption{Implementation of the {\tt swap} routing combinator.}
+\label{figure:swap_implementation}
+\end{figure}
+
+The {\tt first} and {\tt second} combinators are similar to serial composition,
+but they transform only one signal function. For signal changes, they must split
+the set of input changes into those which will be passed to the signal function
+and those which will be simply passed along to the output, and then recombine
+them on the other side. For event occurrences, the occurrence must be
+pattern-matched to determine whether to call the event continuation from the
+provided signal function or passed through, and output event occurrences must
+have the suitable routing constructor re-applied. In any case, when a
+continuation has been applied, the combinator must be recursively applied to the
+new signal function.
+
+The looping feedback combinator is particularly tricky. As it is currently
+implemented, the initial sample for the right side of the input signal vector to
+the supplied function is the right side of the output sample. This is acceptable,
+given Haskell's non-strict evaluation strategy, but it is necessary that the
+right side of the signal function's output not be immediately dependent on its
+input. The feedback combinator makes use of Haskell's lazy evaluation to
+feed events back into the combinator, and stores signal updates until the next
+sample. Signal samples are thus automatically decoupled after initialization.
+The implementation makes use of the recursive nature of the {\tt let} construct
+in Haskell, and the non-strict evaluation of Haskell, to implement feedback.
+
+The {\tt filter} combinators are simple to implement. Their sampling
+continuation is superfluous, and the event continuation merely applies the
+supplied function, and constructs an output list based on the result.
+
+The {\tt accumulate} combinators are implemented in terms of the {\tt filter}
+and {\tt switch} combinators, as shown in
+Figure~\ref{figure:accumulate_implementation}.
+
+\begin{figure}
+\begin{code}
+-- | Accumulate over event occurrences
+accumulate :: (a -> b -> (Maybe c, a)) -> a -> SVEvent b :~> SVEvent c
+accumulate f a = acc a
+  where acc a = switch (pureEventTransformer (f a) >>>
+                        copy >>>
+                        first (pureEventTransformer fst >>> filter id) >>>
+                        second (pureEventTransformer (acc . snd)))
+
+-- | Accumulate over event occurrences, with lists of event outputs
+accumulateList :: (a -> b -> ([c], a)) -> a -> SVEvent b :~> SVEvent c
+accumulateList f a = acc a
+  where acc a = switch (pureEventTransformer (f a) >>>
+                        copy >>>
+                        first (pureEventTransformer fst >>> filterList id) >>>
+                        second (pureEventTransformer (acc . snd)))
+\end{code}
+\hrule
+\caption{Implementation of event accumulators.}
+\label{figure:accumulate_implementation}
+\end{figure}
+
+The implementation of the joining combinators is simple. The {\tt union}
+combinator simply passes along every event occurrence it receives on either
+input, stripping off the left and right combinator. This is acceptable since we
+do not insist on a total ordering of events, or an event time resolution greater
+than the sampling rate. The {\tt combineSignals} combinator maintains the values
+of both signals, and applies the combination function whenever one is updated.
+The {\tt capture} combinator maintains the input signal value, and adds it to
+each event occurrence.
+
+Time dependence is introduced by the {\tt time}, {\tt delay}, and {\tt integrate}
+combinators. The time combinator simply sums the time updates and provides the
+sum as a signal output. The {\tt delay} combinator keeps a table of events
+which have come in, along with their schedule occurrence time, and produces
+them as output when time advances far enough. The integrate combinator performs
+rectangle-rule integration on signal samples with respect to time.
+
+The implementation strategy leaves room for optimizations. In particular, an
+additional constructor for time-independent signal functions would allow
+portions of a signal function to forgo evaluation during time steps unless they
+had signal updates. Optimizations in the style of Yampa, observed by keeping
+an updated AST for the signal function and pattern-matching on it when switching,
+might further improve performance. In particular, collapsing nested or
+serially-composed versions of the {\tt switchWait} step when switching would
+remove at least some of the observed dependence of performance on sampling rate.
+Nevertheless, this implementation performs quite well as it currently exists, as
+we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparisons}.
+
+\subsection{Evaluation Interface}
+\label{subsection:Implementation-Evaluation_Interface}
+The evaluation interface provides the means of evaluating a signal function
+with inputs and producing effects in response to the signal function's outputs.
+We would like to produce a set of constructs that interacts well with Haskell's
+system for external IO.
+
+The evaluation interface translates between signal functions and the standard
+Haskell construct for sequencing effects and external inputs, namely,
+{\em monads}~\cite{PeytonJones1993}. The inspiration for monads is drawn from
+the rather esoteric domain of category theory, but the concept as applied to
+programming languages is actually rather simple.
+
+In a programming language with type constructors, a monad is simply a 1-arity
+type constructor together with two operations. We shall call this type
+constructor $m$. The first operation takes a value of any type
+(call the type $a$) and produces a value of type $m \: a$ ($m$ applied to $a$).
+The second operation takes a value of type $m \: a$ and a function from $a$ to
+$m \: b$, and produces a $m \: b$. This means that this second operation,
+monadic application, applies the function to the value in the context of the
+monad.
+
+For our purposes, there are two interesting properties of monads. One is that
+for a specific type constructor, opaque primitives may be defined which
+can be used, along with return, to build functions for the second argument
+of the monadic application operation. In our case, we would like operations
+to push an event to a signal function, to update the signal components of
+the signal function, and to increment the time and sample the signal function.
+
+This property is what enables monads to be used for input and output. A Haskell
+{\tt main} program is an IO action. An IO action is a value of the monadic type
+IO. This action is comprised of a primitive action (either a {\tt return} or a 
+true IO primitive), possibly together with a function from the output of that
+action to another IO action. Note that values with the IO type may appear
+anywhere in a Haskell program, but they are only execute when they are returned
+as a part of the {\tt main} IO program.
+
+The Haskell typeclass for monads is defined:
+
+\begin{code}
+class Monad m where
+  return :: a -> m a
+  -- | Infix operator for monadic application
+  (>>=)  :: m a -> (a -> m b) -> m b 
+\end{code}
+
+The other interesting property of monads is the existence of a class of monads
+called monad transformers. These are arity-2 type constructors which take as
+their first parameter a monad type constructor. The partial application of
+such a type constructor to any monad type constructor then results in a monad.
+A construct is then provided to lift values of the underlying monad to values
+in the transforming monad, which may then be sequenced as normal. This leads
+to the powerful concept of monad transformer stacks, where monads with many
+different capabilities are combined without any more plumbing than explicit
+{\tt lift} operations.
+
+This property will be useful because we want our evaluation interface to be
+part of a monad stack, the base of which is Haskell's IO monad. By formulating
+the evaluation interface as a monad transformer, we need only define the
+operations relevant to the evaluation of signal functions, and we can depend
+on the constructs of the IO monad to interact with whatever inputs and outputs
+are necessary. In some cases, we may not wish to use the IO monad at all (e.g
+for testing or simulation). In this case, we can parameterize over another
+monad, such as the Identity monad (which has no special operations and whose
+context is just the value), or the State monad (which maintains an implicit 
+state accessible by {\tt get} and {\tt put} operations).
+
+The monad tranformer typeclass is defined as:
+
+\begin{code}
+class MonadTrans t where
+  lift :: Monad m => m a -> t m a
+\end{code}
+
+Haskell also supports {\em do}-notation for monads. Do notation simply
+makes monadic application implicit between lines and reverses the syntatic
+order of lambda bindings, providing a way to use monads in a manner similar
+to imperative programming. (See Figure~\ref{figure:monad-example} for an
+example.)
+
+\begin{figure}
+\begin{code}
+-- Do notation:
+do putStr "Hello: "
+   string <- getStr
+   putStrLn $ "Nice to meet you " ++ string
+   time <- getCurrentTime
+   print time
+   putStrLn "Look at the time! Bye."
+
+-- Translates to:
+putStr "Hello: " >> getStr >>= 
+(\string -> (putStrLn $ "Nice to meet you " ++ string) >>
+ getCurrentTime >>=
+ (\time -> print time >> putStrLn "Look at the time! Bye."))
+\end{code}
+\hrule
+\caption{An example of monadic $do$-notation and its equivalent monadic
+expression.}
+\label{figure:monad-example}
+\end{figure}
+
+The evaluation interface is exported as shown in
+Section~\ref{section:System_Design_and_Interface-Evaluator}.
+
+The {\tt SFEvalState} type constructor parameterizes over input types for signal
+functions, and underlying monads for a monad transformer, but is not itself
+a monad transformer. It describes the state of signal function evaluation.
+It consists of a record with four members: the current signal function,
+the set of handlers for outputs, the current input signal delta, and the last
+sample time.
+
+The {\tt SFEvalT} monad transformer is a newtype wrapper around the {\tt StateT}
+monad available in the Haskell {\tt transformers} package. The StateT monad
+wraps any other monad, providing a state which may be read using the {\tt get}
+action and replaced with the {\tt put} action. An action in the underlying
+monad is obtained by supplying a {\tt StateT} value and an initial state
+value to the {\tt runStateT} function.
+
+The {\tt SFEvalT} monad transformer does not make the {\tt get} and {\tt put}
+actions available, but uses them in its implementation of the {\tt push},
+{\tt update}, and {\tt sample} actions.
+
+The {\tt push} action pushes an event onto the input of the signal function,
+resulting in immediate evaluation of the relevant components signal function
+and possible actuation of handlers specified for output events. It is
+implemented by fetching the {\tt SFEvalState}, applying the event continuation
+of the signal function contained by the {\tt SFEvalState} to the pushed event,
+thus obtaining a new signal function and a list of events, applying the handlers
+contained in the {\tt SFEvalState} to the output events, replacing the signal
+function in the {\tt SFEvalState}, and replacing the {\tt SFEvalState}.
+
+The {\tt update} action updates the current input signal delta, which will be
+sampled at the next {\tt step} action. It has no immediately observable effect.
+It simply updates the value of one signal in the input signal vector, a delta
+of which is stored in the {\tt SFEvalState} record.
+
+The {\tt step} action takes a time delta, and calls the signal continuation of
+the stored signal function. It actuates the outputs using the stored handlers,
+replaces the stored delta with an empty delta, and stores the resulting new
+signal function in the state.
 
@@
 
+An alternate push-based system, based on {\em signal segments}, was presented in
+work on the Curry-Howard correspondence between a form of temporal logic (LTL)
+and FRP~\cite{Jeffrey2012}. However, this system continues to represent events
+as option-valued signals. It therefore cannot be push-based in its response to
+individual events, and cannot be push-pull, evaluating signals by pull-based
+evaluation and events by push-based evaluation. It is push-based not in the
+sense of evaluating events only when they occur, but rather in that it provides
+a means to push signal segments (samples of a signal over intervals of time) to
+the system and wait on its response.
+
 \section{Conclusion}

### 2012-11-27 Docs/IFL2012/paper.tex
Draft update for IFL paper

@@
 
-\subsection{Goals}
-\label{section:System_Design_and_Interface-Goals}
-
 The goal of FRP is to provide an efficient, declarative abstraction for creating
@@
 
-A {\em monad} is a standard, composable abstraction for writing functions with
-a context, used in Haskell for IO~\cite{PeytonJones1993,PeytonJones2001} and
-other tasks. A monad is simply a 1-arity type constructor together with two
-functions. The first function, {\tt return}, takes a value of type {\tt a} and
-produces a value of type {\tt m a}. The second, called {\tt bind} and stylized
-in the Haskell standard library as the infix operator {\tt (>>=)}, takes a value
-of type {\tt m a} and a function of type {\tt a -> m b} and produces a value of
-type {\tt m b}. This allows a value to be operated on out of the context and a
-new context to be assigned.
-
-A monad can have other primitives which manipulate the context in some way. For 
-instance, the primtives in Haskell's {\tt IO} monad produce actions which, when
-interpreted as part of the {\tt main} action, produce some side-effect. The
-{\tt State} monad provides {\tt get} and {\tt put} operations to work with a 
-state value stored in the context.
-
 Monad transformers~\cite{Jones1995} provide a means to combine the functionality
@@
 The evaluation monad is a monad transformer. This permits it to be used in
-conjunction with the {\tt IO} monad (or any other monad) to describe how input
-is to be obtained for the signal function being evaluated, and how outputs are
-to be handled.
+conjunction with Haskell's {\tt IO} monad (or any other monad) to describe how
+input is to be obtained for the signal function being evaluated, and how outputs
+are to be handled.
 
@@
 
--- Pass through the left side of the input unchanged
-second :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut)
-
 -- Swap the left and right sides
@@
 
--- Remove an empty vector on the right
-cancelRight :: (sv :^: SVEmpty) :~> sv
-
--- Add an empty vector on the right
-uncancelRight :: sv :~> (sv :^: SVEmpty)
-
 -- Make right-associative
 associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))
-
--- Make left-associative
-unassociate :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
 \end{code}
@@
 
-There are other formulations of a reactive combinator found in FRP literature
-and libraries, which may be implemented using the one supplied. These are shown
-in Figure~\ref{figure:alternate_switching_combinators} and may be provided in a
-future version of the TimeFlies library.
-
-\begin{figure}
-\begin{code}
--- Alternate version of switch,
--- implemented in terms of supplied version
-switch_gen ::    (svIn :~> (svOut :^: SVEvent a))
-              -> (a -> svIn :~> svOut)
-              -> svIn :~> svOut
-switch_gen sf f =
-  switch (sf >>> second (pureEventTransformer f))
-
--- Supplied version in terms of alternate version
-switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
-          -> svIn :~> svOut
-switch sf = switch_gen sf id
-
--- Repeated switch, which takes replacement signal functions
--- externally.
-rswitch ::    (svIn :~> svOut)
-           -> (svIn :^: SVEvent (svIn :~> svOut)) :~> svOut
-rswitch sf =
-  switch (first sf >>> second (pureEventTransformer rswitch))
-\end{code}
-\hrule
-\caption{Alternate reactivity combinators.}
-\label{figure:alternate_switching_combinators}
-\end{figure}
-
 It is often useful for a signal function to receive a portion of its
@@
 
--- Apply the function to each input occurrence,
--- and produce an occurrence for each list element
-filterList :: (a -> [b]) -> SVEvent a :~> SVEvent b
-
 -- Apply the function to the stored accumulator
@@
 accumulate ::    (a -> b -> (Maybe c, a))
-              -> a
-              -> SVEvent b :~> SVEvent c
-
--- Apply the function to the stored accumulator
--- and the event occurrence, replacing the
--- accumulator and outputting an event occurrence
--- for each element of the list
-accumulateList ::    (a -> b -> ([c], a))
               -> a
@@
 
--- A dummy handler for an empty output
-emptyHandler    :: SVHandler out SVEmpty
-
--- A handler for an updated signal sample
-signalHandler   :: (a -> out) -> SVHandler out (SVSignal a)
-
--- A handler for an event occurrence
-eventHandler    :: (a -> out) -> SVHandler out (SVEvent a)
-
--- Combine handlers for a vector
-combineHandlers ::    SVHandler out svLeft
-                   -> SVHandler out svRight
-                   -> SVHandler out (svLeft :^: svRight)
-
 -- The state maintained when evaluating a signal function
@@
 
-\begin{figure}
-\begin{code}
--- A sample for all leaves of a signal vector
-data SVSample sv
-
--- Create a sample for a signal leaf
-sample          :: a -> SVSample (SVSignal a)
-
--- A dummy sample for an event leaf
-sampleEvt       :: SVSample (SVEvent a)
-
--- A dummy sample for an empty leaf
-sampleNothing   :: SVSample SVEmpty
-
--- Combine two samples
-combineSamples  ::    SVSample svLeft
-                   -> SVSample svRight
-                   -> SVSample (svLeft :^: svRight)
-\end{code}
-\hrule
-\caption{Data type for initial input}
-\label{figure:initial_input}
-\end{figure}
-
 This state can then be passed to a monadic action which will supply input to
@@
 data SFEvalT svIn svOut m a
-instance MonadTrans (SFEvalT svIn svOut)
-instance (Monad m) => Monad (SFEvalT svIn svOut m)
-instance (Functor m) => Functor (SFEvalT svIn svOut m)
-instance (Monad m, Functor m) => Applicative (SFEvalT svIn svOut m)
-instance (MonadIO m) => MonadIO (SVEvalT svIn svOut m)
+-- Appropriate instances for Monad, MonadTrans, MonadIO,
+-- Functor, Applicative, etc
 
@@
 the library. Rather, they will specify how data is represented during the
-temporal evaluation of signal functions.
-
-We then describe how signal functions are implemented using these concrete
-types, along with higher-order functions and continuations.
+temporal evaluation of signal functions. We then implement signal functions
+as records of functions from these concrete types to these concrete types paired
+with new signal functions. The evaluation system for signal functions then
+maintains this record for the signal function it is evaluating, calls the
+correct function for the current input, actuates the outputs (using supplied
+functions from the outputs to monadic actions) and replaces the record with the
+newly supplied signal function record.
 
@@
 
-The types are therefore expressed in the system exactly as they were described
-in Section~\ref{section:System_Design_and_Interface}. (To refresh, see
-Figure~\ref{figure:signal_vector_types}.) The striking observation about these
-types is that they have {\em no data constructors}. There are no values which
-take these types.
-
-Instead, we will create concrete representations which are parameterized over
+We create concrete representations which are parameterized over
 these types. These concrete representations will be expressed as GADTs, allowing
@@
 
-Having specified a datatype for signal functions, we must now provide
-combinators which produce signal functions of this type. Each combinator's
-implementation must specify how it is initialized, how it samples its input, and
-how it responds to event occurrences.
-
-We will not detail every combinator here, but we will discuss each of the
-implementation challenges encountered.
-
-As an example of the implementation of combinators, we show the implementation
-of the {\tt identity} signal function in Figure~\ref{figure:identity_implementation}.
-This signal function simply passes all of its inputs along as outputs. The
-initialization function simply passes along the received sample and outputs the
-initialized version of the signal function. The initialized version of the input
-is similar, but is self-referential. It outputs itself as its replacement. This
-is standard for simple and routing combinators which are not reactive, and
-simply move samples and event occurrences around.
-
-\begin{figure}
-\begin{code}
-identity :: sv :~> sv
-identity =
-  SF (\initSample -> (initSample, identityInit))
-
-identityInit :: SF Initialized sv sv
-identityInit =
-  SFInit (\dt sigDelta -> (sigDelta, [], identityInit))
-         (\evtOcc -> ([evtOcc], identityInit))
-\end{code}
-\hrule
-\caption{Implementation of the {\tt identity} combinator.}
-\label{figure:identity_implementation}
-\end{figure}
+Having specified a type constructor for signal functions, we must now provide
+combinators which produce signal functions typed with this type constructor.
+Space precludes a complete description of the combinators in the TimeFlies
+system, but we will present
+
+The simplest example of the implementation of is the {\tt identity} combinator,
+shown in Figure~\ref{figure:identity_implementation}. This signal function
+simply passes all of its inputs along as outputs. The initialization function
+simply passes along the received sample and outputs the initialized version of
+the signal function. The initialized version of the input is similar, but is
+self-referential. It outputs itself as its replacement. This is standard for
+simple and routing combinators which are not reactive, and simply re-arrange,
+discard, or combine signals and events.
 
@@
 them to the initialized version of the combinator.
-
-\begin{figure}
-\begin{code}
-(>>>) ::    (svIn :~> svBetween) 
-         -> (svBetween :~> svOut)
-         -> (svIn :~> svOut)
-(SF sigSampleF1) >>> (SF sigSampleF2) =
-  SF (\sigSample -> let (sigSample', sfInit1) = sigSampleF1 sigSample
-                        (sigSample'', sfInit2) = sigSampleF2 sigSample'
-                    in (sigSample'', composeInit sfInit1 sfInit2))
-
-composeInit ::     SF Initialized svIn svBetween
-                -> SF Initialized svBetween svOut
-                -> SF Initialized svIn svOut
-composeInit (SFInit dtCont1 inputCont1) sf2@(SFInit dtCont2 inputCont2) =
-  SFInit
-    (\dt sigDelta -> 
-       let (sf1MemOutput, sf1EvtOutputs, sf1New) = dtCont1 dt sigDelta
-           (sf2MemOutput, sf2EvtOutputs, sf2New) = dtCont2 dt sf1MemOutput
-           (sf2EvtEvtOutputs, sf2Newest) = applySF sf2New sf1EvtOutputs
-       in (sf2MemOutput,
-           sf2EvtOutputs ++ sf2EvtEvtOutputs,
-           composeInit sf1New sf2Newest)
-    )
-    (\evtOcc -> 
-      let (sf1Outputs, newSf1) = inputCont1 evtOcc
-          (sf2FoldOutputs, newSf2) = applySF sf2 sf1Outputs
-      in (sf2FoldOutputs, composeInit newSf1 newSf2)   
-    )
-
-applySF ::    SF Initialized svIn svOut
-           -> [SVOccurrence svIn]
-           -> ([SVOccurrence svOut],
-               SF Initialized svIn svOut)
-applySF sf indices =
-  foldr (\evtOcc (changes, SFInit _ changeCont) ->
-           let (newChanges, nextSF) = changeCont evtOcc
-               in (newChanges ++ changes, nextSF))
-        ([], sf)
-        indices
-\end{code}
-\hrule
-\caption{Implementation of serial composition.}
-\label{figure:serial_composition_implementation}
-\end{figure}
 
@@
 an example in Figure~\ref{figure:swap_implementation}
-
-\begin{figure}
-\begin{code}
--- T.swap is imported from Data.Tuple
-swap :: (sv1 :^: sv2) :~> (sv2 :^: sv1)
-swap =
-  SF ((, swapInit) . 
-      uncurry combineSamples .
-      T.swap . splitSample)
-
-swapInit :: SF Initialized (SVAppend sv1 sv2) (SVAppend sv2 sv1)
-swapInit =
-  SFInit (flip (const .
-                (, [], swapInit) .
-                uncurry combineDeltas .
-                T.swap . splitDelta))
-          (\evtOcc ->
-             (case chooseOccurrence evtOcc of
-                Left lOcc  -> [occRight lOcc]
-                Right rOcc -> [occLeft rOcc], swapInit))
-
-\end{code}
-\hrule
-\caption{Implementation of the {\tt swap} routing combinator.}
-\label{figure:swap_implementation}
-\end{figure}
 
@@
 
-\begin{figure}
-\begin{code}
--- | Accumulate over event occurrences
-accumulate :: (a -> b -> (Maybe c, a)) -> a -> SVEvent b :~> SVEvent c
-accumulate f a = acc a
-  where acc a = switch (pureEventTransformer (f a) >>>
-                        copy >>>
-                        first (pureEventTransformer fst >>> filter id) >>>
-                        second (pureEventTransformer (acc . snd)))
-
--- | Accumulate over event occurrences, with lists of event outputs
-accumulateList :: (a -> b -> ([c], a)) -> a -> SVEvent b :~> SVEvent c
-accumulateList f a = acc a
-  where acc a = switch (pureEventTransformer (f a) >>>
-                        copy >>>
-                        first (pureEventTransformer fst >>> filterList id) >>>
-                        second (pureEventTransformer (acc . snd)))
-\end{code}
-\hrule
-\caption{Implementation of event accumulators.}
-\label{figure:accumulate_implementation}
-\end{figure}
-
 The implementation of the joining combinators is simple. The {\tt union}
@@
 
-In a programming language with type constructors, a monad is simply a 1-arity
-type constructor together with two operations. We shall call this type
-constructor $m$. The first operation takes a value of any type
-(call the type $a$) and produces a value of type $m \: a$ ($m$ applied to $a$).
-The second operation takes a value of type $m \: a$ and a function from $a$ to
-$m \: b$, and produces a $m \: b$. This means that this second operation,
-monadic application, applies the function to the value in the context of the
-monad.
-
-For our purposes, there are two interesting properties of monads. One is that
-for a specific type constructor, opaque primitives may be defined which
-can be used, along with return, to build functions for the second argument
-of the monadic application operation. In our case, we would like operations
-to push an event to a signal function, to update the signal components of
-the signal function, and to increment the time and sample the signal function.
-
-This property is what enables monads to be used for input and output. A Haskell
-{\tt main} program is an IO action. An IO action is a value of the monadic type
-IO. This action is comprised of a primitive action (either a {\tt return} or a 
-true IO primitive), possibly together with a function from the output of that
-action to another IO action. Note that values with the IO type may appear
-anywhere in a Haskell program, but they are only execute when they are returned
-as a part of the {\tt main} IO program.
-
-The Haskell typeclass for monads is defined:
-
-\begin{code}
-class Monad m where
-  return :: a -> m a
-  -- | Infix operator for monadic application
-  (>>=)  :: m a -> (a -> m b) -> m b 
-\end{code}
-
-The other interesting property of monads is the existence of a class of monads
-called monad transformers. These are arity-2 type constructors which take as
-their first parameter a monad type constructor. The partial application of
-such a type constructor to any monad type constructor then results in a monad.
-A construct is then provided to lift values of the underlying monad to values
-in the transforming monad, which may then be sequenced as normal. This leads
-to the powerful concept of monad transformer stacks, where monads with many
-different capabilities are combined without any more plumbing than explicit
-{\tt lift} operations.
-
-This property will be useful because we want our evaluation interface to be
-part of a monad stack, the base of which is Haskell's IO monad. By formulating
+By formulating
 the evaluation interface as a monad transformer, we need only define the
@@
 state accessible by {\tt get} and {\tt put} operations).
-
-The monad tranformer typeclass is defined as:
-
-\begin{code}
-class MonadTrans t where
-  lift :: Monad m => m a -> t m a
-\end{code}
-
-Haskell also supports {\em do}-notation for monads. Do notation simply
-makes monadic application implicit between lines and reverses the syntatic
-order of lambda bindings, providing a way to use monads in a manner similar
-to imperative programming. (See Figure~\ref{figure:monad-example} for an
-example.)
-
-\begin{figure}
-\begin{code}
--- Do notation:
-do putStr "Hello: "
-   string <- getStr
-   putStrLn $ "Nice to meet you " ++ string
-   time <- getCurrentTime
-   print time
-   putStrLn "Look at the time! Bye."
-
--- Translates to:
-putStr "Hello: " >> getStr >>= 
-(\string -> (putStrLn $ "Nice to meet you " ++ string) >>
- getCurrentTime >>=
- (\time -> print time >> putStrLn "Look at the time! Bye."))
-\end{code}
-\hrule
-\caption{An example of monadic $do$-notation and its equivalent monadic
-expression.}
-\label{figure:monad-example}
-\end{figure}
 

### 2012-11-27 Docs/IFL2012/abstract.tex
Break IFL paper into multiple files.

--- 2012-11-24 Docs/IFL2012/paper.tex
@@
-\documentclass[draft]{llncs}
-
-\usepackage{fancyvrb}                                       % For code environment
-\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small} %
-
-\title{Push-Pull Signal-Function Functional Reactive Programming}
-\titlerunning{Push-Pull Signal-Function FRP}
-
-\author{Edward Amsden}
-\institute{Rochester Institute of Technology\\\email{eca7215@cs.rit.edu}}
-
-\begin{document}
-\maketitle
-
 \begin{abstract}
@@
 \end{abstract}
-
-\section{Introduction}
-\label{section:Introduction}
-
-Functional Reactive Programming (FRP) is a class of systems for describing
-reactive programs. Reactive programs are programs which, rather than taking
-a single input and producing a single output, must accept multiple inputs and 
-alter temporal behavior, including the production of multiple outputs, based
-on these inputs.
-
-An FRP system will provide a means of manipulating {\em behaviors} and
-{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
-but the definition is the same. A behavior or signal is, semantically, a
-function from time to a value. An event is a discrete, possibly infinite, and 
-time-ordered sequence of occurrences, which are times paired with values.
-
-FRP systems can generally be categorized as ``classic FRP,'' 
-which corresponds to the originally described FRP system in that behaviors
-and events are manipulated directly and are first-class values in the FRP
-system, or ``signal-function FRP,'' in which behaviors (generally termed
-signals in this approach) and events are not first-class values, but signal
-functions are first class values. Signal functions are time-dependent and
-reactive transformers of signals, events, or combinations of signals and events.
-
-FRP combines behaviors and events through the use of {\em switching}, in which
-a behavior (in classic FRP) or a signal function (in signal-function FRP) is
-replaced by a new behavior or signal function carried by an event occurrence.
-
-Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
-Recent work on classic FRP has focused on efficient implementation. One approach to
-efficiency is to separate the evaluation of behaviors and events, since suitable 
-strategies give best performance in each case. Push-based evaluation evaluates a
-system only when input is available, and is thus suitable for discrete inputs
-such as events. Pull-based evaluation evaluates the system as quickly as
-possible, polling for input, and is preferable for behaviors and signals.
-The initial implementations of FRP made use of pull-based evaluation for both
-behaviors and events. Reactive~\cite{Elliott2009}, as well as more recent systems such as
-``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
-events and pull-based evaluation for behaviors. This is known as ``push-pull''
-evaluation.
-
-All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
-have used pull-based evaluation for both signals and events. This is due to
-the ease of implementation of pull-based evaluation, and the types used for
-signal functions which do not permit distinguishing signals and events, or
-constructing only part of the input (for instance, one event occurrence.)
-
-A recent extension of signal-function FRP called N-Ary FRP~\cite{Sculthorpe2011}
-describes a method of typing signal functions which, as we will show, enables
-the push-based evaluation of events in a signal-function FRP system. The notion
-of signal vectors allows the representation of signal function inputs and
-outputs as combinations of signals and events, rather than a single signal which
-may contain multiple values, including option values for events. Signal vectors
-are uninhabited types, which can be used to type partial or full representations
-of the signal function inputs and outputs.
-
-We present TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
-favorite quotation of one of the author's philosophy instructors, used to
-demonstrate the ambiguity of language. The origin of the quotation is unknown.}
-a push-pull signal-function FRP system. We hope to demonstrate the feasibility
-of such an approach to FRP, and provide a basis for further research into
-efficient implementation of signal-function FRP. We also describe a powerful
-evaluation interface for TimeFlies, which permits us to use TimeFlies to
-describe applications which make use of multiple and differing IO libraries.
-
-Section~\ref{section:System_Design} describes design choices for the system,
-and provides an overview of the interface. Section~\ref{section:Implementation}
-describes how the system is implemented, and how the separation of evaluation
-between events and signals is achieved. Section~\ref{section:Discussion} is a
-discussion of the usefulness of our implementation. 
-Section~\ref{section:Ongoing_and_Further_Work} describes the current and future
-work on this system. Section~\ref{section:Related_Work} gives an overview of
-related efforts. Section~\ref{section:Conclusion} concludes.
-
-\section{System Design}
-\label{section:System_Design}
-
-Our goal is to produce a composable and efficient FRP system. Signal function
-FRP has an advantage in terms of composability, because it permits the
-construction of self-contained objects through both input and output composition,
-rather than purely through output composition as in classic FRP. Signal-function
-FRP also avoids problematic properties common to classic FRP systems such as a
-large class of time and space leaks~\cite{Liu2007}.
-
-Efficient implementations of signal-function FRP have been approached through
-runtime optimization~\cite{Nilsson2005}, but all implementations have been
-pull-based for both signals and events. A truly efficient implementation will
-likely combine run-time optimization with push-based evaluation for events.
-
-The concept of N-Ary FRP was to encode additional safety properties into the
-types of a signal-function FRP system~\cite{Sculthorpe2011}. This system
-introduced the concept of {\em signal vectors} as input and output types for
-signal functions. Signal vectors are combinations of signals and events. In
-N-Ary FRP, signal functions are represented at the type level, and type-level
-functions (type families in Haskell) are used to specify the representations
-of signal functions.
-
-In our system, we construct representations of signal vectors using
-Generalized Algebraic Datatypes~\cite{Cheney2003,Xi2003}. The use of GADTs
-enables us to use signal vectors to instantiate type parameters in the types
-of signal vector representations, as well as in the types of signal functions.
-GADTs are available as an extension in the Glasgow Haskell Compiler~\cite{PeytonJones2006}.
-
-This approach permits us to construct partial inputs and outputs for signal
-functions. For instance, we can construct a representation for a single event
-in a signal vector, and another representation of updated values for a subset of
-the signals in the signal vector, and yet another carrying values for all
-signals in a signal vector. The signal vector types are shown in
-Figure~\ref{figure:signal_vectors}.
-
-\begin{figure}[t]
-\begin{code}
-data SVEmpty
-data SVSignal a
-data SVEvent a
-data SVAppend svLeft svRight
-\end{code}
-\hrule
-\caption{Signal vectors.}
-\label{figure:signal_vectors}
-\end{figure}
-
-With the ability to construct partial representations of signal vectors, we can
-represent signal functions with a datatype carrying multiple functions, one for
-each type of input. This allows us to separate the pull-based processing of
-signals from the push-based reaction to events.
-
-The exposed interface is a set of combinators for constructing signal functions,
-as well as combinators for describing the evaluation of a signal function. The
-interface is shown in Fig.~\ref{figure:signal_function_interface}.
-
-\begin{figure}
-\begin{code}
-
--- Signal Functions
-type :~> svIn svOut
-
--- Infix type alias for SVAppend
-type :^: svLeft svRight = SVAppend svLeft svRight
-
--- Basic signal functions
-identity :: sv :~> sv
-constant :: a -> SVEmpty :~> SVSignal a
-never    :: SVEmpty :~> SVEvent a
-asap     :: a -> SVEmpty :~> SVEvent a
-after    :: Double -> a -> SVEmpty :~> SVEvent a
-
--- Lifting pure functions
-pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal a
-pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent a
-
--- Composition and routing
-(>>>)           :: (svIn :~> svMiddle) -> (svMiddle :~> svOut) 
-                   -> (svIn :~> svOut)
-first           :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
-second          :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut) 
-swap            :: (svLeft :^: svRight) :~> (svRight :^: svLeft)
-copy            :: sv :~> (sv :^: sv)
-ignore          :: sv :~> SVEmpty
-cancelLeft      :: (SVEmpty :^: sv) :~> sv
-cancelRight     :: (sv :^: SVEmpty) :~> sv
-uncancelLeft    :: sv :~> (SVEmpty :^: sv)
-uncancelRight   :: sv :~> (sv :^: SVEmpty)
-associate       :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3)) 
-unassociate     :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
-
--- Reactivity
-switch :: (svIn :~> (svOut :^: SVEvent (svIn :~> svOut))) 
-          -> svIn :~> svOut
-
--- Feedback
-loop :: ((svIn :^: svLoop) :~> (svOut :^: svLoop)) -> svIn :~> svOut
-
--- Time dependence
-time  :: SVEmpty :~> Double
-delay :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
-class TimeIntegrate
-
--- Joining
-union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
-combineSignals :: (a -> b -> c) 
-                  -> (SVSignal a :^: SVSignal b) :~> SVSignal c
-capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent a
-
--- Events
-filter         :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-\end{code}
-\hrule
-\caption{Signal function interface.}
-\label{figure:signal_function_interface}
-\end{figure}
-
-Signal functions are produced by combining primitive signal functions using
-the {\tt >>>} (sequential composition), {\tt first}, and {\tt second} routing
-combinators. Other routing signal functions are provided, but they are intended
-to be combined with, rather than to modify, other signal functions.
-
-\subsection{Evaluation Interface}
-The evaluation interface provides a monad~\cite{PeytonJones1993,PeytonJones2001}
-for specifying input to signal functions and handling their output. This
-interface also allows input actions to be separated from output actions, and
-from each other, so that a signal function may receive input from multiple IO
-systems and have its output handled by multiple output systems.
-
-The evaluation interface is a monad transformer~\cite{Liang1995}. This allows
-a signal function evaluation to be constructed from primitive evaluation actions
-(event pushing, input signal updating, and sampling) as well as actions from the
-underlying monad. These actions can then be run in the underlying monad
-to actuate the signal function. Because the actuation takes a signal function's
-state as input and produces a new state as output, in addition to the monadic
-side effects of handling the signal function's output, different evaluation
-actions may be taken on the same signal function from distinct locations within
-an application's code. This allows the evaluation interface to be easily
-integrated with event-loop style systems as well as traditional imperative
-IO systems.
-
-The evaluation interface consists of the monad transformer type, functions for
-constructing event inputs, initial input signal samples, and input signal
-updates, functions for constructing the vector of output handlers for a signal
-function, and evaluation actions. The full interface is shown in Fig.~\ref{figure:evaluation_interface}.
-
-\begin{figure}
-\begin{code}
--- Input helpers
-class SVRoutable -- Instances: SVSignalUpdate,
-                 -- SVEventOccurrence
-svLeft :: (SVRoutable r) => 
-          r svLeft -> r (SVAppend svLeft svRight)
-svRight :: (SVRoutable r) =>
-           r svRight -> r (SVAppend svLeft svRight)
-
--- Signal inputs
-data SVSignalUpdate
-data SVSample
-sample :: a -> SVSample (SVSignal a)
-sampleEvt :: SVSample (SVEvent a)
-sampleNothing :: SVSample SVEmpty
-combineSamples :: SVSample svLeft -> svSample svRight 
-                  -> svSample (svLeft :^: svRight)
-svSig :: a -> SVSignalUpdate (SVSignal a)
-
--- Event inputs
-data SVEventInput
-svOcc :: a -> SVEventInput (SVEvent a)
--- Actuation
-update :: (Monad m) => SVSignalUpdate svIn 
-           -> SFEvalT svIn svOut m ()
-push   :: (Monad m) => SVEventInput svIn
-          -> SFEvalT svIn svOut m ()
-sample :: SVEvalT svIn svOut m ()
--- Running
-data SVHandler
-emptyHandler :: SVHandler m SVEmpty
-eventHandler :: (a -> m ()) 
-                -> SVHandler m (SVEvent a)
-signalHandler :: (a -> m ()) 
-                 -> SVHandler m (SVSignal a)
-combineHandlers :: SVHandler m svLeft
-                   -> SVHandler m svRight
-                   -> SVHandler m (svLeft :^: svRight)
-initSFEvalT :: SVHandler m svOut -> SVSample svIn 
-               -> Double ->(svIn :~> svOut)
-               -> SFEvalState m svIn svOut
-runSFEvalT :: SFEvalT svIn svOut m a
-              -> SFEvalState m svIn svOut
-              -> m a
-\end{code}
-\hrule
-\caption{Evaluation interface.}
-\label{figure:evaluation_interface}
-\end{figure}
-
-\section{Implementation}
-\label{section:Implementation}
-
-We now turn our attention to the implementation of the signal function
-system. We will discuss representations of inputs, outputs, and signal functions,
-as well as the implementations of specific signal function combinators.
-
-\subsection{Input and Output Representations}
-\label{subsection:Input_and_Output_Representations}
-As discussed in Section~\ref{section:System_Design}, our implementation requires
-representations of the inputs and outputs of signal functions. Since signal
-functions are typed using signal vectors, the representation types must be
-parameterized over signal vectors. But signal vectors are uninhabited types, and
-in ordinary Haskell we cannot declare a data constructor which uses a component
-of a type parameter. Further, we cannot restrict which types may instantiate a
-type parameter for a particular data constructor.
-
-GADTs lift these restrictions, permitting type constraints to be applied
-to individual data constructors. In the declaration of a GADT, a type signature
-is provided for each data constructor, including the types of the constructor's
-parameters. When a GADT is pattern-matched, the types of the parameters are
-inferred using the constraints given in this signature. This is called type
-refinement.
-
-Using GADTs, we can construct representations of signal vectors to use in our
-implementation.
-
-When constructing a representation, we consider how to represent each
-component of a signal vector, whether it is a signal element, an empty
-element, or an append node. For instance, in order to represent a signal
-sample for all signals in an element, we construct the GADT shown in Figure~\ref{figure:signal_sample_representation}.
-
-\begin{figure}
-\begin{code}
-data SVSample sv where
-  SVSample        ::    a -> SVSample (SVSignal a)
-  SVSampleEvt     ::    SVSample (SVEvent a)
-  SVSampleNothing ::    SVSample SVEmpty
-  SVSampleBoth    ::    SVSample svLeft
-                     -> SVSample svRight
-                     -> SVSample (svLeft :^: svRight)
-\end{code}
-\hrule
-\caption{A representation which carries a value for each signal in a signal vector.}
-\label{figure:signal_sample_representation}
-\end{figure}
-
-\subsection{Signal Function Representations}
-\label{subsection:Signal_Function_Representations}
-With representations for signal function inputs and outputs in place, we turn
-to the representation of signal functions. A signal function must be able to
-respond to time increments and new signal samples, as well as event occurrences.
-Since we wish to be able to ``push'' event occurrences independently of
-sampling, events are handled using the last-updated time and signal sample.
-
-When a signal function is asked to respond to either of these types of inputs,
-it must produce an output consisting of zero or more event occurrences, a new
-signal function with the same type (to enable reactivity and state), and, if
-it is responding to a time and sample update, a delta which represents updates
-to the sample of its output signals.
-
-Finally, there is a special case for a signal function at time 0. It must be
-provided with an initial input sample before it can respond to incremental time
-and sample updates or to event occurrences.
-
-Signal functions are represented as a datatype with two constructors. The first
-constructor wraps a function from an input sample to an output sample and an
-initialized signal function. The second wraps two functions. The first accepts
-a time delta and a signal delta, and produces a signal delta, a list of event
-occurrences, and a new signal function as the output. The second accepts an
-event occurrence and produces a list of event occurrences and a new signal
-function as the output. The signal function representation is shown in Figure~\ref{figure:signal_function_representation}.
-
-\begin{figure}[t]
-\begin{code}
-data Initialized
-data NonInitialized
-
-type :~> svIn svOut = SF NonInitialized svIn svOut
-
-data SF init svIn svOut where
-  SF :: (SVSample svIn -> (SVSample svOut, SF Initialized svIn svOut))
-        -> SF NonInitialized svIn svOut
-  SFInit :: (Double -> SVDelta svIn 
-             -> (SVDelta svOut, [SVOccurrence svOut],
-                 SF Initialized svIn svOut))
-            -> (SVOccurrence svIn -> ([SVOccurrence svOut],
-                                      SF Initialized svIn svOut))
-            -> SF Initialized svIn svOut
-\end{code}
-\hrule
-\caption{Signal function representation}
-\label{figure:signal_function_representation}
-\end{figure}
-
-This representation allows a signal function to respond to events and time
-updates separately, and does not enforce the representation of events during
-time updates, as previous signal function systems do.
-
-\subsection{Signal Function Implementations}
-\label{subsection:Signal_Function_Implementations}
-
-Now that we have established representations for signal functions and their
-inputs and outputs, we show several examples of how signal function combinators
-are implemented.
-
-The {\tt identity} signal function's initialization function takes a sample,
-and returns that sample as its output sample. The initialized function returned
-is the {\tt identityInit} function, which is not exported by the module.
-The time and sample function ignores the provided time delta, produces the
-input signal delta as the output signal delta, and returns {\tt identityInit} as
-the new signal function. The event function takes an event occurrence and
-returns a singleton list containing that occurrence, along with the
-{\tt identityInit} function as the replacement signal function. The
-implementation is shown in Figure~\ref{figure:identity_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-identity :: sv :~> sv
-identity = SF (\sample -> (sample, identityInit))
-
-identityInit :: SF Initialized sv sv
-identityInit = SFInit (\_ delta -> (delta, [], identityInit))
-                      (\occ -> ([occ], identityInit))
-\end{code}
-\hrule
-\caption{{\tt identity} signal function implementation.}
-\label{figure:identity_signal_function_implementation}
-\end{figure}
-
-The {\tt constant} signal function is initialized with a value, which it then
-produces as its output forever. Thus, the initialized version of {\tt constant}
-never produces any output other than itself as a replacement
-signal function, and an empty signal delta. The implementation of {\tt constant}
-is shown in Fig.~\ref{figure:constant_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-constant :: a -> SVEmpty :~> SVSignal a
-constant x = SF (\_ -> (constantInit, sample x))
-
-constantInit :: SF Initialized SVEmpty (SVSignal a)
-constantInit = SFInit (\_ _ -> (deltaNothing, [], constantInit))
-                      (\_ -> ([], constantInit))
-\end{code}
-\hrule
-\caption{{\tt constant} signal function implementation.}
-\label{figure:constant_signal_function_implementation}
-\end{figure}
-
-The {\tt asap} signal function implementation requires a parameter for the
-initialized version of the signal function, to specify the value of the event
-occurrence. It replaces itself with the initialized version of the {\tt never}
-signal function after the first time step. The implementation is shown in
-Fig.~\ref{figure:asap_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-asap :: a -> SVEmpty :~> SVEvent a
-asap x = SF (\_ -> (sampleEvt, asapInit x))
-
-asapInit :: a -> SF Initialized SVEmpty (SVEvent a)
-asapInit x = SFInit (\_ _ -> (deltaNothing, [occurrence x], neverInit))
-                    (\_ -> (never))
-\end{code}
-\hrule
-\caption{{\tt asap} signal function implementation.}
-\label{figure:asap_signal_function_implementation}
-\end{figure}
-
-The {\tt filter} function accepts events, applies a {\tt Maybe} predicate to
-their values, and produces the value produced by the predicate as an event
-occurrence, or no occurrence if the value is {\tt Nothing}. It is implemented
-by having the event occurrence response function apply the {\tt maybe} function
-from the Haskell prelude to the value returned by the predicate, as shown in 
-Fig.~\ref{figure:filter_signal_function_implementation}
-
-\begin{figure}[t]
-\begin{code}
-filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-filter p = SF (\_ -> (sampleEvt, filterInit p))
-
-filterInit :: (a -> Maybe b) -> SF Initialized (SVEvent a) (SVEvent b)
-filterInit p = SFInit (\_ _ -> (deltaNothing, [], filterInit p))
-                      (\evtOcc -> (maybe [] ((:[]) . occurrence) $
-                                   fromOccurrence evtOcc,
-                                   filterInit p))
-\end{code}
-\hrule
-\caption{{\tt filter} signal function implementation.}
-\label{figure:filter_signal_function_implementation}
-\end{figure}
-
-The {\tt associate} signal function is a routing signal function. It transforms
-a signal vector which is left-associated at the top level to one that is
-right-associated at the top level. Its implementation is representative of all
-of the routing functions, and is shown in Fig.~\ref{figure:associate_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-associate :: SVAppend (SVAppend sv1 sv2) sv3
-             :~> SVAppend sv1 (SVAppend sv2 sv3)
-associate = 
-  SF (\sigSample -> let (sigSampleLeft, sigSampleRight) = 
-                          splitSample sigSample
-                        (sigSampleLeftLeft, sigSampleLeftRight) =
-                          splitSample sigSampleLeft
-                    in (combineSamples sigSampleLeftLeft 
-                        (combineSamples sigSampleLeftRight sigSampleRight),
-                        associateInit))
-
-associateInit :: SF Initialized (SVAppend (SVAppend sv1 sv2) sv3)
-                                (SVAppend sv1 (SVAppend sv2 sv3))
-associateInit = SFInit (\_ sigDelta -> let (sigDeltaLeft, sigDeltaRight) =
-                                             splitDelta sigDelta
-                                           (sigDeltaLeftLeft,
-                                            sigDeltaLeftRight) =
-                                             splitDelta sigDeltaLeft
-                                       in (combineDeltas sigDeltaLeftLeft
-                                           (combineDeltas sigDeltaLeftRight
-                                                          sigDeltaRight),
-                                           [], associateInit)) 
-                       (\evtOcc -> (case chooseOccurrence evtOcc of
-                                      Left leftOcc -> 
-                                        case chooseOccurrence leftOcc of
-                                          Left leftLeftOcc ->
-                                            [occLeft leftLeftOcc]
-                                          Right leftRightOcc ->
-                                            [occRight $ occLeft leftRightOcc]
-                                      Right rightOcc ->
-                                        [occRight $ occRight $ rightOcc],
-                                    associateInit))
-\end{code}
-\hrule
-\caption{{\tt associate} signal function implementation.}
-\label{figure:associate_signal_function_implementation}
-\end{figure}
-
-The {\tt switch} function is the basic combinator used to introduce reactivity.
-It is given a signal function whose output is the append of two signal vectors.
-The left side of the signal vector is passed on as the output of the reactive
-signal function, and the right side is an event carrying signal functions. When
-an event occurrence is present on the right-side event output, the signal
-function carried by this occurrence replaces the signal function constructed
-by switch. Due to space concerns, the implementation is elided, but a brief
-description will suffice. The signal function stores the input sample
-provided during initialization, and updates it with deltas. When an event
-occurrence carrying a signal function is produced by either the event or 
-time function of the wrapped signal function, the stored signal sample is used
-to initialize the new signal function. If this occurs while handling an event
-input, the sample output by the new signal function's initialization is stored
-and is combined with its first output delta to produce the output delta at the
-next time step.
-
-The {\tt loop} function allows a signal function to see a component of its own
-output as input. This is primarily useful when a signal function has components
-which mutually depend on each others outputs, such as in physics simulations
-or games. Care must be taken that the feedback output is not immediately
-dependent on the feedback input, or sampling the signal function will not
-terminate. Loop is implemented by generating a recursive list (which will
-be infinite if the feedback is not decoupled) of event inputs, and by splitting
-the output signal delta and using the right side output delta as the right
-side input delta within a recursive let-binding. (The let-binding in Haskell
-always admits recursive bindings.) If the system is not completely decoupled,
-this will result in non-termination during an evaluation step. Decoupling can
-be achieved using {\tt delay} primitive.
-
-\subsection{Evaluation Interface Implementation}
-\label{subsection:Evaluation_Interface_Implementation}
-
-The evaluation interface must maintain state including the current signal
-function, the current time (to produce time deltas), updates to the input
-sample which have yet to be sampled, and the output handlers.
-
-The output handlers are stored in a structure similar to that for a signal
-sample. The difference is that both signal and event leaves contain values, and
-these values are functions from the leaf type to another type. The handler
-datatype is shown in Fig.~\ref{figure:handler_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SVHandler out sv where
-  SVHandlerEmpty  :: SVHandler out SVEmpty
-  SVHandlerSignal :: (a -> out) -> SVHandler out (SVSignal a)
-  SVHandlerEvent  :: (a -> out) -> SVHandler out (SVEvent a)
-  SVHandlerBoth   :: SVHandler out svLeft -> SVHandler out svRight 
-                     -> SVHandler out (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Handler datatype.}
-\label{figure:handler_datatype}
-\end{figure}
-
-We now need a datatype to hold the various components of the evaluation state.
-This type is shown in Fig~\ref{figure:evaluation_state_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SFEvalState m svIn svOut 
-  = SFEvalState { 
-                  esSF :: SF Initialized svIn svOut,
-                  esOutputHandlers :: SVHandler (m ()) svOut,
-                  esLastTime :: Double,
-                  esDelta :: SVDelta svIn
-                }
-
-\end{code}
-\hrule
-\caption{Evaluation state datatype.}
-\label{figure:evaluation_state_datatype}
-\end{figure}
-
-The evaluation interface itself is a monad transformer, which we implement
-as a Haskell newtype wrapping the {\tt StateT} monad transformer. We do not
-export the raw {\tt put} and {\tt get} actions of the state monad, but instead
-implement the push, update, and sampling operations for signal function
-evaluation using {\tt put} and {\tt get}. The evaluation monad transformer is
-shown in Fig.~\ref{figure:signal_function_evaluation_monad_transformer}.
-Instances of typeclasses including {\tt Monad} and {\tt MonadTrans} are derived
-using the ``GeneralizedNewtypeDeriving'' extension to the Glasgow Haskell
-Compiler.
-
-\begin{figure}[t]
-\begin{code}
-newtype SFEvalT svIn svOut m a = StateT (SFEvalState m svIn svOut) m a
-\end{code}
-\hrule
-\caption{Signal function evaluation monad transformer.}
-\label{figure:signal_function_evaluation_monad_transformer}
-\end{figure}
-
-\section{Discussion}
-\label{section:Discussion}
-
-The system presented here, TimeFlies, demonstrates how using signal vectors
-to type inputs and outputs enables push-based evaluation of events in a
-signal-function system. We take advantage of this representation in several ways.
-
-First, by separating components of inputs and outputs in the types, we are free
-to create distinct, and often partial, representations of the input or output
-of a signal function. This enables us to represent only the event occurrence
-being pushed at that time.
-
-Second, this separation also permits us to separate the process of gathering
-the input to a signal function, and the process of handling its output, into
-different points in a program. Using the evaluation interface described, an
-event occurrence may be pushed onto one input of a signal function from one
-point in a program (e.g. a mouse click handler), an input signal
-may be updated in another (e.g. a mouse movement handler), and finally the
-system may be sampled in a third place (e.g. an animation or audio timed
-callback).
-
-Finally, this approach enables further work on the implementation of the signal
-function system to be separated from changes in the interface. By enabling
-differing representations of the inputs and outputs of signal functions, we are
-free to change these representations without the need to further constrain the
-input and output types.
-
-\section{Ongoing and Further Work}
-\label{section:Ongoing_and_Further_Work}
-
-TimeFlies, the system described here, has been implemented, but not
-extensively tested. The immediate goal is to create a real-time application
-which will permit a performance and implementation comparison of TimeFlies with
-Yampa, the current state-of-the-art pull-based signal-function system.
-
-In the future, we hope to apply run-time optimizations, using the technique used
-for Yampa, to create a push-pull self-optimizing signal-function system. Further,
-we hope to use this system as a basis for exploring signal-function FRP as a
-basis for general-purpose application frameworks.
-
-\section{Related Work}
-\label{section:Related_Work}
-
-Signal Function FRP was introduced as a model for Graphical User Interfaces~\cite{Courtney2001-1}.
-The system was originally termed ``AFRP'' (Arrowized FRP). Yampa is a rewrite of
-AFRP where signal functions apply a number of ad-hoc optimizatons to themselves
-as they evolve. Yampa demonstrated a modest performance improvement
-over AFRP~\cite{Nilsson2005}.
-
-Reactive is a classic FRP system which implements push-based evaluation for events
-by transforming behaviors to ``reactive~normal~form,'' where a behavior
-is a non-reactive behavior running inside a switch, whose event stream carries
-behaviors in reactive normal form. The system is evaluated by forking a Haskell
-thread to repeatedly sample the non-reactive behavior, and then blocking on the
-evaluation of the first occurrence in the event stream. When this occurrence
-is yielded, the evaluation thread for the behavior is killed and a new
-thread forked to evaluate the new behavior~\cite{Elliott2009}.
-
-\section{Conclusion}
-\label{section:Conclusion}
-
-We have presented TimeFlies, a system for push-pull signal-function Functional
-Reactive Programming, and have shown how the use of a signal vectors as input
-and output types for signal functions, together with GADT-based representations
-of the inputs and outputs, permits the implementation of a push-pull system.
-
-We have also described a general and flexible monadic evaluation interface for
-TimeFlies, which permits us to interface the TimeFlies system with different
-styles of IO systems, including multiple IO systems in the same application.
-
-This opens up the exciting possibility that a signal-function FRP could become
-an efficient and general framework for writing interactive applications.
-
-\bibliographystyle{splncs}
-\bibliography{thesis}
-
-\end{document}

### 2012-11-27 Docs/IFL2012/conclusion.tex
Break IFL paper into multiple files.

--- 2012-11-24 Docs/IFL2012/paper.tex
@@
-\documentclass[draft]{llncs}
-
-\usepackage{fancyvrb}                                       % For code environment
-\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small} %
-
-\title{Push-Pull Signal-Function Functional Reactive Programming}
-\titlerunning{Push-Pull Signal-Function FRP}
-
-\author{Edward Amsden}
-\institute{Rochester Institute of Technology\\\email{eca7215@cs.rit.edu}}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-Functional Reactive Programming is a promising class of systems for writing
-interactive and time-dependent programs. Signal-function FRP is a subclass of
-these systems which provides advantages of modularity and correctness, but
-has proven difficult to efficiently implement.
-
-The abstraction of signal vectors provides the necessary type apparatus to
-distinguish components of the input and output of signal functions which benefit
-from a push-based implementation from those which benefit from a pull-based
-implementation, and to combine both implementation strategies in a single system.
-
-We describe a signal-function FRP system which provides push-based evaluation
-for events, pull-based evaluation for signals, and a simple monadic evaluation
-interface which permits the system to be easily integrated with one or more
-IO systems.
-\end{abstract}
-
-\section{Introduction}
-\label{section:Introduction}
-
-Functional Reactive Programming (FRP) is a class of systems for describing
-reactive programs. Reactive programs are programs which, rather than taking
-a single input and producing a single output, must accept multiple inputs and 
-alter temporal behavior, including the production of multiple outputs, based
-on these inputs.
-
-An FRP system will provide a means of manipulating {\em behaviors} and
-{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
-but the definition is the same. A behavior or signal is, semantically, a
-function from time to a value. An event is a discrete, possibly infinite, and 
-time-ordered sequence of occurrences, which are times paired with values.
-
-FRP systems can generally be categorized as ``classic FRP,'' 
-which corresponds to the originally described FRP system in that behaviors
-and events are manipulated directly and are first-class values in the FRP
-system, or ``signal-function FRP,'' in which behaviors (generally termed
-signals in this approach) and events are not first-class values, but signal
-functions are first class values. Signal functions are time-dependent and
-reactive transformers of signals, events, or combinations of signals and events.
-
-FRP combines behaviors and events through the use of {\em switching}, in which
-a behavior (in classic FRP) or a signal function (in signal-function FRP) is
-replaced by a new behavior or signal function carried by an event occurrence.
-
-Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
-Recent work on classic FRP has focused on efficient implementation. One approach to
-efficiency is to separate the evaluation of behaviors and events, since suitable 
-strategies give best performance in each case. Push-based evaluation evaluates a
-system only when input is available, and is thus suitable for discrete inputs
-such as events. Pull-based evaluation evaluates the system as quickly as
-possible, polling for input, and is preferable for behaviors and signals.
-The initial implementations of FRP made use of pull-based evaluation for both
-behaviors and events. Reactive~\cite{Elliott2009}, as well as more recent systems such as
-``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
-events and pull-based evaluation for behaviors. This is known as ``push-pull''
-evaluation.
-
-All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
-have used pull-based evaluation for both signals and events. This is due to
-the ease of implementation of pull-based evaluation, and the types used for
-signal functions which do not permit distinguishing signals and events, or
-constructing only part of the input (for instance, one event occurrence.)
-
-A recent extension of signal-function FRP called N-Ary FRP~\cite{Sculthorpe2011}
-describes a method of typing signal functions which, as we will show, enables
-the push-based evaluation of events in a signal-function FRP system. The notion
-of signal vectors allows the representation of signal function inputs and
-outputs as combinations of signals and events, rather than a single signal which
-may contain multiple values, including option values for events. Signal vectors
-are uninhabited types, which can be used to type partial or full representations
-of the signal function inputs and outputs.
-
-We present TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
-favorite quotation of one of the author's philosophy instructors, used to
-demonstrate the ambiguity of language. The origin of the quotation is unknown.}
-a push-pull signal-function FRP system. We hope to demonstrate the feasibility
-of such an approach to FRP, and provide a basis for further research into
-efficient implementation of signal-function FRP. We also describe a powerful
-evaluation interface for TimeFlies, which permits us to use TimeFlies to
-describe applications which make use of multiple and differing IO libraries.
-
-Section~\ref{section:System_Design} describes design choices for the system,
-and provides an overview of the interface. Section~\ref{section:Implementation}
-describes how the system is implemented, and how the separation of evaluation
-between events and signals is achieved. Section~\ref{section:Discussion} is a
-discussion of the usefulness of our implementation. 
-Section~\ref{section:Ongoing_and_Further_Work} describes the current and future
-work on this system. Section~\ref{section:Related_Work} gives an overview of
-related efforts. Section~\ref{section:Conclusion} concludes.
-
-\section{System Design}
-\label{section:System_Design}
-
-Our goal is to produce a composable and efficient FRP system. Signal function
-FRP has an advantage in terms of composability, because it permits the
-construction of self-contained objects through both input and output composition,
-rather than purely through output composition as in classic FRP. Signal-function
-FRP also avoids problematic properties common to classic FRP systems such as a
-large class of time and space leaks~\cite{Liu2007}.
-
-Efficient implementations of signal-function FRP have been approached through
-runtime optimization~\cite{Nilsson2005}, but all implementations have been
-pull-based for both signals and events. A truly efficient implementation will
-likely combine run-time optimization with push-based evaluation for events.
-
-The concept of N-Ary FRP was to encode additional safety properties into the
-types of a signal-function FRP system~\cite{Sculthorpe2011}. This system
-introduced the concept of {\em signal vectors} as input and output types for
-signal functions. Signal vectors are combinations of signals and events. In
-N-Ary FRP, signal functions are represented at the type level, and type-level
-functions (type families in Haskell) are used to specify the representations
-of signal functions.
-
-In our system, we construct representations of signal vectors using
-Generalized Algebraic Datatypes~\cite{Cheney2003,Xi2003}. The use of GADTs
-enables us to use signal vectors to instantiate type parameters in the types
-of signal vector representations, as well as in the types of signal functions.
-GADTs are available as an extension in the Glasgow Haskell Compiler~\cite{PeytonJones2006}.
-
-This approach permits us to construct partial inputs and outputs for signal
-functions. For instance, we can construct a representation for a single event
-in a signal vector, and another representation of updated values for a subset of
-the signals in the signal vector, and yet another carrying values for all
-signals in a signal vector. The signal vector types are shown in
-Figure~\ref{figure:signal_vectors}.
-
-\begin{figure}[t]
-\begin{code}
-data SVEmpty
-data SVSignal a
-data SVEvent a
-data SVAppend svLeft svRight
-\end{code}
-\hrule
-\caption{Signal vectors.}
-\label{figure:signal_vectors}
-\end{figure}
-
-With the ability to construct partial representations of signal vectors, we can
-represent signal functions with a datatype carrying multiple functions, one for
-each type of input. This allows us to separate the pull-based processing of
-signals from the push-based reaction to events.
-
-The exposed interface is a set of combinators for constructing signal functions,
-as well as combinators for describing the evaluation of a signal function. The
-interface is shown in Fig.~\ref{figure:signal_function_interface}.
-
-\begin{figure}
-\begin{code}
-
--- Signal Functions
-type :~> svIn svOut
-
--- Infix type alias for SVAppend
-type :^: svLeft svRight = SVAppend svLeft svRight
-
--- Basic signal functions
-identity :: sv :~> sv
-constant :: a -> SVEmpty :~> SVSignal a
-never    :: SVEmpty :~> SVEvent a
-asap     :: a -> SVEmpty :~> SVEvent a
-after    :: Double -> a -> SVEmpty :~> SVEvent a
-
--- Lifting pure functions
-pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal a
-pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent a
-
--- Composition and routing
-(>>>)           :: (svIn :~> svMiddle) -> (svMiddle :~> svOut) 
-                   -> (svIn :~> svOut)
-first           :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
-second          :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut) 
-swap            :: (svLeft :^: svRight) :~> (svRight :^: svLeft)
-copy            :: sv :~> (sv :^: sv)
-ignore          :: sv :~> SVEmpty
-cancelLeft      :: (SVEmpty :^: sv) :~> sv
-cancelRight     :: (sv :^: SVEmpty) :~> sv
-uncancelLeft    :: sv :~> (SVEmpty :^: sv)
-uncancelRight   :: sv :~> (sv :^: SVEmpty)
-associate       :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3)) 
-unassociate     :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
-
--- Reactivity
-switch :: (svIn :~> (svOut :^: SVEvent (svIn :~> svOut))) 
-          -> svIn :~> svOut
-
--- Feedback
-loop :: ((svIn :^: svLoop) :~> (svOut :^: svLoop)) -> svIn :~> svOut
-
--- Time dependence
-time  :: SVEmpty :~> Double
-delay :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
-class TimeIntegrate
-
--- Joining
-union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
-combineSignals :: (a -> b -> c) 
-                  -> (SVSignal a :^: SVSignal b) :~> SVSignal c
-capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent a
-
--- Events
-filter         :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-\end{code}
-\hrule
-\caption{Signal function interface.}
-\label{figure:signal_function_interface}
-\end{figure}
-
-Signal functions are produced by combining primitive signal functions using
-the {\tt >>>} (sequential composition), {\tt first}, and {\tt second} routing
-combinators. Other routing signal functions are provided, but they are intended
-to be combined with, rather than to modify, other signal functions.
-
-\subsection{Evaluation Interface}
-The evaluation interface provides a monad~\cite{PeytonJones1993,PeytonJones2001}
-for specifying input to signal functions and handling their output. This
-interface also allows input actions to be separated from output actions, and
-from each other, so that a signal function may receive input from multiple IO
-systems and have its output handled by multiple output systems.
-
-The evaluation interface is a monad transformer~\cite{Liang1995}. This allows
-a signal function evaluation to be constructed from primitive evaluation actions
-(event pushing, input signal updating, and sampling) as well as actions from the
-underlying monad. These actions can then be run in the underlying monad
-to actuate the signal function. Because the actuation takes a signal function's
-state as input and produces a new state as output, in addition to the monadic
-side effects of handling the signal function's output, different evaluation
-actions may be taken on the same signal function from distinct locations within
-an application's code. This allows the evaluation interface to be easily
-integrated with event-loop style systems as well as traditional imperative
-IO systems.
-
-The evaluation interface consists of the monad transformer type, functions for
-constructing event inputs, initial input signal samples, and input signal
-updates, functions for constructing the vector of output handlers for a signal
-function, and evaluation actions. The full interface is shown in Fig.~\ref{figure:evaluation_interface}.
-
-\begin{figure}
-\begin{code}
--- Input helpers
-class SVRoutable -- Instances: SVSignalUpdate,
-                 -- SVEventOccurrence
-svLeft :: (SVRoutable r) => 
-          r svLeft -> r (SVAppend svLeft svRight)
-svRight :: (SVRoutable r) =>
-           r svRight -> r (SVAppend svLeft svRight)
-
--- Signal inputs
-data SVSignalUpdate
-data SVSample
-sample :: a -> SVSample (SVSignal a)
-sampleEvt :: SVSample (SVEvent a)
-sampleNothing :: SVSample SVEmpty
-combineSamples :: SVSample svLeft -> svSample svRight 
-                  -> svSample (svLeft :^: svRight)
-svSig :: a -> SVSignalUpdate (SVSignal a)
-
--- Event inputs
-data SVEventInput
-svOcc :: a -> SVEventInput (SVEvent a)
--- Actuation
-update :: (Monad m) => SVSignalUpdate svIn 
-           -> SFEvalT svIn svOut m ()
-push   :: (Monad m) => SVEventInput svIn
-          -> SFEvalT svIn svOut m ()
-sample :: SVEvalT svIn svOut m ()
--- Running
-data SVHandler
-emptyHandler :: SVHandler m SVEmpty
-eventHandler :: (a -> m ()) 
-                -> SVHandler m (SVEvent a)
-signalHandler :: (a -> m ()) 
-                 -> SVHandler m (SVSignal a)
-combineHandlers :: SVHandler m svLeft
-                   -> SVHandler m svRight
-                   -> SVHandler m (svLeft :^: svRight)
-initSFEvalT :: SVHandler m svOut -> SVSample svIn 
-               -> Double ->(svIn :~> svOut)
-               -> SFEvalState m svIn svOut
-runSFEvalT :: SFEvalT svIn svOut m a
-              -> SFEvalState m svIn svOut
-              -> m a
-\end{code}
-\hrule
-\caption{Evaluation interface.}
-\label{figure:evaluation_interface}
-\end{figure}
-
-\section{Implementation}
-\label{section:Implementation}
-
-We now turn our attention to the implementation of the signal function
-system. We will discuss representations of inputs, outputs, and signal functions,
-as well as the implementations of specific signal function combinators.
-
-\subsection{Input and Output Representations}
-\label{subsection:Input_and_Output_Representations}
-As discussed in Section~\ref{section:System_Design}, our implementation requires
-representations of the inputs and outputs of signal functions. Since signal
-functions are typed using signal vectors, the representation types must be
-parameterized over signal vectors. But signal vectors are uninhabited types, and
-in ordinary Haskell we cannot declare a data constructor which uses a component
-of a type parameter. Further, we cannot restrict which types may instantiate a
-type parameter for a particular data constructor.
-
-GADTs lift these restrictions, permitting type constraints to be applied
-to individual data constructors. In the declaration of a GADT, a type signature
-is provided for each data constructor, including the types of the constructor's
-parameters. When a GADT is pattern-matched, the types of the parameters are
-inferred using the constraints given in this signature. This is called type
-refinement.
-
-Using GADTs, we can construct representations of signal vectors to use in our
-implementation.
-
-When constructing a representation, we consider how to represent each
-component of a signal vector, whether it is a signal element, an empty
-element, or an append node. For instance, in order to represent a signal
-sample for all signals in an element, we construct the GADT shown in Figure~\ref{figure:signal_sample_representation}.
-
-\begin{figure}
-\begin{code}
-data SVSample sv where
-  SVSample        ::    a -> SVSample (SVSignal a)
-  SVSampleEvt     ::    SVSample (SVEvent a)
-  SVSampleNothing ::    SVSample SVEmpty
-  SVSampleBoth    ::    SVSample svLeft
-                     -> SVSample svRight
-                     -> SVSample (svLeft :^: svRight)
-\end{code}
-\hrule
-\caption{A representation which carries a value for each signal in a signal vector.}
-\label{figure:signal_sample_representation}
-\end{figure}
-
-\subsection{Signal Function Representations}
-\label{subsection:Signal_Function_Representations}
-With representations for signal function inputs and outputs in place, we turn
-to the representation of signal functions. A signal function must be able to
-respond to time increments and new signal samples, as well as event occurrences.
-Since we wish to be able to ``push'' event occurrences independently of
-sampling, events are handled using the last-updated time and signal sample.
-
-When a signal function is asked to respond to either of these types of inputs,
-it must produce an output consisting of zero or more event occurrences, a new
-signal function with the same type (to enable reactivity and state), and, if
-it is responding to a time and sample update, a delta which represents updates
-to the sample of its output signals.
-
-Finally, there is a special case for a signal function at time 0. It must be
-provided with an initial input sample before it can respond to incremental time
-and sample updates or to event occurrences.
-
-Signal functions are represented as a datatype with two constructors. The first
-constructor wraps a function from an input sample to an output sample and an
-initialized signal function. The second wraps two functions. The first accepts
-a time delta and a signal delta, and produces a signal delta, a list of event
-occurrences, and a new signal function as the output. The second accepts an
-event occurrence and produces a list of event occurrences and a new signal
-function as the output. The signal function representation is shown in Figure~\ref{figure:signal_function_representation}.
-
-\begin{figure}[t]
-\begin{code}
-data Initialized
-data NonInitialized
-
-type :~> svIn svOut = SF NonInitialized svIn svOut
-
-data SF init svIn svOut where
-  SF :: (SVSample svIn -> (SVSample svOut, SF Initialized svIn svOut))
-        -> SF NonInitialized svIn svOut
-  SFInit :: (Double -> SVDelta svIn 
-             -> (SVDelta svOut, [SVOccurrence svOut],
-                 SF Initialized svIn svOut))
-            -> (SVOccurrence svIn -> ([SVOccurrence svOut],
-                                      SF Initialized svIn svOut))
-            -> SF Initialized svIn svOut
-\end{code}
-\hrule
-\caption{Signal function representation}
-\label{figure:signal_function_representation}
-\end{figure}
-
-This representation allows a signal function to respond to events and time
-updates separately, and does not enforce the representation of events during
-time updates, as previous signal function systems do.
-
-\subsection{Signal Function Implementations}
-\label{subsection:Signal_Function_Implementations}
-
-Now that we have established representations for signal functions and their
-inputs and outputs, we show several examples of how signal function combinators
-are implemented.
-
-The {\tt identity} signal function's initialization function takes a sample,
-and returns that sample as its output sample. The initialized function returned
-is the {\tt identityInit} function, which is not exported by the module.
-The time and sample function ignores the provided time delta, produces the
-input signal delta as the output signal delta, and returns {\tt identityInit} as
-the new signal function. The event function takes an event occurrence and
-returns a singleton list containing that occurrence, along with the
-{\tt identityInit} function as the replacement signal function. The
-implementation is shown in Figure~\ref{figure:identity_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-identity :: sv :~> sv
-identity = SF (\sample -> (sample, identityInit))
-
-identityInit :: SF Initialized sv sv
-identityInit = SFInit (\_ delta -> (delta, [], identityInit))
-                      (\occ -> ([occ], identityInit))
-\end{code}
-\hrule
-\caption{{\tt identity} signal function implementation.}
-\label{figure:identity_signal_function_implementation}
-\end{figure}
-
-The {\tt constant} signal function is initialized with a value, which it then
-produces as its output forever. Thus, the initialized version of {\tt constant}
-never produces any output other than itself as a replacement
-signal function, and an empty signal delta. The implementation of {\tt constant}
-is shown in Fig.~\ref{figure:constant_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-constant :: a -> SVEmpty :~> SVSignal a
-constant x = SF (\_ -> (constantInit, sample x))
-
-constantInit :: SF Initialized SVEmpty (SVSignal a)
-constantInit = SFInit (\_ _ -> (deltaNothing, [], constantInit))
-                      (\_ -> ([], constantInit))
-\end{code}
-\hrule
-\caption{{\tt constant} signal function implementation.}
-\label{figure:constant_signal_function_implementation}
-\end{figure}
-
-The {\tt asap} signal function implementation requires a parameter for the
-initialized version of the signal function, to specify the value of the event
-occurrence. It replaces itself with the initialized version of the {\tt never}
-signal function after the first time step. The implementation is shown in
-Fig.~\ref{figure:asap_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-asap :: a -> SVEmpty :~> SVEvent a
-asap x = SF (\_ -> (sampleEvt, asapInit x))
-
-asapInit :: a -> SF Initialized SVEmpty (SVEvent a)
-asapInit x = SFInit (\_ _ -> (deltaNothing, [occurrence x], neverInit))
-                    (\_ -> (never))
-\end{code}
-\hrule
-\caption{{\tt asap} signal function implementation.}
-\label{figure:asap_signal_function_implementation}
-\end{figure}
-
-The {\tt filter} function accepts events, applies a {\tt Maybe} predicate to
-their values, and produces the value produced by the predicate as an event
-occurrence, or no occurrence if the value is {\tt Nothing}. It is implemented
-by having the event occurrence response function apply the {\tt maybe} function
-from the Haskell prelude to the value returned by the predicate, as shown in 
-Fig.~\ref{figure:filter_signal_function_implementation}
-
-\begin{figure}[t]
-\begin{code}
-filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-filter p = SF (\_ -> (sampleEvt, filterInit p))
-
-filterInit :: (a -> Maybe b) -> SF Initialized (SVEvent a) (SVEvent b)
-filterInit p = SFInit (\_ _ -> (deltaNothing, [], filterInit p))
-                      (\evtOcc -> (maybe [] ((:[]) . occurrence) $
-                                   fromOccurrence evtOcc,
-                                   filterInit p))
-\end{code}
-\hrule
-\caption{{\tt filter} signal function implementation.}
-\label{figure:filter_signal_function_implementation}
-\end{figure}
-
-The {\tt associate} signal function is a routing signal function. It transforms
-a signal vector which is left-associated at the top level to one that is
-right-associated at the top level. Its implementation is representative of all
-of the routing functions, and is shown in Fig.~\ref{figure:associate_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-associate :: SVAppend (SVAppend sv1 sv2) sv3
-             :~> SVAppend sv1 (SVAppend sv2 sv3)
-associate = 
-  SF (\sigSample -> let (sigSampleLeft, sigSampleRight) = 
-                          splitSample sigSample
-                        (sigSampleLeftLeft, sigSampleLeftRight) =
-                          splitSample sigSampleLeft
-                    in (combineSamples sigSampleLeftLeft 
-                        (combineSamples sigSampleLeftRight sigSampleRight),
-                        associateInit))
-
-associateInit :: SF Initialized (SVAppend (SVAppend sv1 sv2) sv3)
-                                (SVAppend sv1 (SVAppend sv2 sv3))
-associateInit = SFInit (\_ sigDelta -> let (sigDeltaLeft, sigDeltaRight) =
-                                             splitDelta sigDelta
-                                           (sigDeltaLeftLeft,
-                                            sigDeltaLeftRight) =
-                                             splitDelta sigDeltaLeft
-                                       in (combineDeltas sigDeltaLeftLeft
-                                           (combineDeltas sigDeltaLeftRight
-                                                          sigDeltaRight),
-                                           [], associateInit)) 
-                       (\evtOcc -> (case chooseOccurrence evtOcc of
-                                      Left leftOcc -> 
-                                        case chooseOccurrence leftOcc of
-                                          Left leftLeftOcc ->
-                                            [occLeft leftLeftOcc]
-                                          Right leftRightOcc ->
-                                            [occRight $ occLeft leftRightOcc]
-                                      Right rightOcc ->
-                                        [occRight $ occRight $ rightOcc],
-                                    associateInit))
-\end{code}
-\hrule
-\caption{{\tt associate} signal function implementation.}
-\label{figure:associate_signal_function_implementation}
-\end{figure}
-
-The {\tt switch} function is the basic combinator used to introduce reactivity.
-It is given a signal function whose output is the append of two signal vectors.
-The left side of the signal vector is passed on as the output of the reactive
-signal function, and the right side is an event carrying signal functions. When
-an event occurrence is present on the right-side event output, the signal
-function carried by this occurrence replaces the signal function constructed
-by switch. Due to space concerns, the implementation is elided, but a brief
-description will suffice. The signal function stores the input sample
-provided during initialization, and updates it with deltas. When an event
-occurrence carrying a signal function is produced by either the event or 
-time function of the wrapped signal function, the stored signal sample is used
-to initialize the new signal function. If this occurs while handling an event
-input, the sample output by the new signal function's initialization is stored
-and is combined with its first output delta to produce the output delta at the
-next time step.
-
-The {\tt loop} function allows a signal function to see a component of its own
-output as input. This is primarily useful when a signal function has components
-which mutually depend on each others outputs, such as in physics simulations
-or games. Care must be taken that the feedback output is not immediately
-dependent on the feedback input, or sampling the signal function will not
-terminate. Loop is implemented by generating a recursive list (which will
-be infinite if the feedback is not decoupled) of event inputs, and by splitting
-the output signal delta and using the right side output delta as the right
-side input delta within a recursive let-binding. (The let-binding in Haskell
-always admits recursive bindings.) If the system is not completely decoupled,
-this will result in non-termination during an evaluation step. Decoupling can
-be achieved using {\tt delay} primitive.
-
-\subsection{Evaluation Interface Implementation}
-\label{subsection:Evaluation_Interface_Implementation}
-
-The evaluation interface must maintain state including the current signal
-function, the current time (to produce time deltas), updates to the input
-sample which have yet to be sampled, and the output handlers.
-
-The output handlers are stored in a structure similar to that for a signal
-sample. The difference is that both signal and event leaves contain values, and
-these values are functions from the leaf type to another type. The handler
-datatype is shown in Fig.~\ref{figure:handler_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SVHandler out sv where
-  SVHandlerEmpty  :: SVHandler out SVEmpty
-  SVHandlerSignal :: (a -> out) -> SVHandler out (SVSignal a)
-  SVHandlerEvent  :: (a -> out) -> SVHandler out (SVEvent a)
-  SVHandlerBoth   :: SVHandler out svLeft -> SVHandler out svRight 
-                     -> SVHandler out (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Handler datatype.}
-\label{figure:handler_datatype}
-\end{figure}
-
-We now need a datatype to hold the various components of the evaluation state.
-This type is shown in Fig~\ref{figure:evaluation_state_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SFEvalState m svIn svOut 
-  = SFEvalState { 
-                  esSF :: SF Initialized svIn svOut,
-                  esOutputHandlers :: SVHandler (m ()) svOut,
-                  esLastTime :: Double,
-                  esDelta :: SVDelta svIn
-                }
-
-\end{code}
-\hrule
-\caption{Evaluation state datatype.}
-\label{figure:evaluation_state_datatype}
-\end{figure}
-
-The evaluation interface itself is a monad transformer, which we implement
-as a Haskell newtype wrapping the {\tt StateT} monad transformer. We do not
-export the raw {\tt put} and {\tt get} actions of the state monad, but instead
-implement the push, update, and sampling operations for signal function
-evaluation using {\tt put} and {\tt get}. The evaluation monad transformer is
-shown in Fig.~\ref{figure:signal_function_evaluation_monad_transformer}.
-Instances of typeclasses including {\tt Monad} and {\tt MonadTrans} are derived
-using the ``GeneralizedNewtypeDeriving'' extension to the Glasgow Haskell
-Compiler.
-
-\begin{figure}[t]
-\begin{code}
-newtype SFEvalT svIn svOut m a = StateT (SFEvalState m svIn svOut) m a
-\end{code}
-\hrule
-\caption{Signal function evaluation monad transformer.}
-\label{figure:signal_function_evaluation_monad_transformer}
-\end{figure}
-
-\section{Discussion}
-\label{section:Discussion}
-
-The system presented here, TimeFlies, demonstrates how using signal vectors
-to type inputs and outputs enables push-based evaluation of events in a
-signal-function system. We take advantage of this representation in several ways.
-
-First, by separating components of inputs and outputs in the types, we are free
-to create distinct, and often partial, representations of the input or output
-of a signal function. This enables us to represent only the event occurrence
-being pushed at that time.
-
-Second, this separation also permits us to separate the process of gathering
-the input to a signal function, and the process of handling its output, into
-different points in a program. Using the evaluation interface described, an
-event occurrence may be pushed onto one input of a signal function from one
-point in a program (e.g. a mouse click handler), an input signal
-may be updated in another (e.g. a mouse movement handler), and finally the
-system may be sampled in a third place (e.g. an animation or audio timed
-callback).
-
-Finally, this approach enables further work on the implementation of the signal
-function system to be separated from changes in the interface. By enabling
-differing representations of the inputs and outputs of signal functions, we are
-free to change these representations without the need to further constrain the
-input and output types.
-
-\section{Ongoing and Further Work}
-\label{section:Ongoing_and_Further_Work}
-
-TimeFlies, the system described here, has been implemented, but not
-extensively tested. The immediate goal is to create a real-time application
-which will permit a performance and implementation comparison of TimeFlies with
-Yampa, the current state-of-the-art pull-based signal-function system.
-
-In the future, we hope to apply run-time optimizations, using the technique used
-for Yampa, to create a push-pull self-optimizing signal-function system. Further,
-we hope to use this system as a basis for exploring signal-function FRP as a
-basis for general-purpose application frameworks.
-
-\section{Related Work}
-\label{section:Related_Work}
-
-Signal Function FRP was introduced as a model for Graphical User Interfaces~\cite{Courtney2001-1}.
-The system was originally termed ``AFRP'' (Arrowized FRP). Yampa is a rewrite of
-AFRP where signal functions apply a number of ad-hoc optimizatons to themselves
-as they evolve. Yampa demonstrated a modest performance improvement
-over AFRP~\cite{Nilsson2005}.
-
-Reactive is a classic FRP system which implements push-based evaluation for events
-by transforming behaviors to ``reactive~normal~form,'' where a behavior
-is a non-reactive behavior running inside a switch, whose event stream carries
-behaviors in reactive normal form. The system is evaluated by forking a Haskell
-thread to repeatedly sample the non-reactive behavior, and then blocking on the
-evaluation of the first occurrence in the event stream. When this occurrence
-is yielded, the evaluation thread for the behavior is killed and a new
-thread forked to evaluate the new behavior~\cite{Elliott2009}.
-
 \section{Conclusion}
@@
 an efficient and general framework for writing interactive applications.
-
-\bibliographystyle{splncs}
-\bibliography{thesis}
-
-\end{document}

### 2012-11-27 Docs/IFL2012/design.tex
Break IFL paper into multiple files.

--- 2012-11-27 Docs/IFL2012/paper.tex
@@
-\documentclass[draft]{llncs}
-
-\usepackage{fancyvrb}                                       % For code environment
-\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small} %
-
-\title{Push-Pull Signal-Function Functional Reactive Programming}
-\titlerunning{Push-Pull Signal-Function FRP}
-
-\author{Edward Amsden}
-\institute{Rochester Institute of Technology\\\email{eca7215@cs.rit.edu}}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-Functional Reactive Programming is a promising class of systems for writing
-interactive and time-dependent programs. Signal-function FRP is a subclass of
-these systems which provides advantages of modularity and correctness, but
-has proven difficult to efficiently implement.
-
-The abstraction of signal vectors provides the necessary type apparatus to
-distinguish components of the input and output of signal functions which benefit
-from a push-based implementation from those which benefit from a pull-based
-implementation, and to combine both implementation strategies in a single system.
-
-We describe a signal-function FRP system which provides push-based evaluation
-for events, pull-based evaluation for signals, and a simple monadic evaluation
-interface which permits the system to be easily integrated with one or more
-IO systems.
-\end{abstract}
-
-\section{Introduction}
-\label{section:Introduction}
-
-Functional Reactive Programming (FRP) is a class of systems for describing
-reactive programs. Reactive programs are programs which, rather than taking
-a single input and producing a single output, must accept multiple inputs and 
-alter temporal behavior, including the production of multiple outputs, based
-on these inputs.
-
-An FRP system will provide a means of manipulating {\em behaviors} and
-{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
-but the definition is the same. A behavior or signal is, semantically, a
-function from time to a value. An event is a discrete, possibly infinite, and 
-time-ordered sequence of occurrences, which are times paired with values.
-
-FRP systems can generally be categorized as ``classic FRP,'' 
-which corresponds to the originally described FRP system in that behaviors
-and events are manipulated directly and are first-class values in the FRP
-system, or ``signal-function FRP,'' in which behaviors (generally termed
-signals in this approach) and events are not first-class values, but signal
-functions are first class values. Signal functions are time-dependent and
-reactive transformers of signals, events, or combinations of signals and events.
-
-FRP combines behaviors and events through the use of {\em switching}, in which
-a behavior (in classic FRP) or a signal function (in signal-function FRP) is
-replaced by a new behavior or signal function carried by an event occurrence.
-
-Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
-Recent work on classic FRP has focused on efficient implementation. One approach to
-efficiency is to separate the evaluation of behaviors and events, since suitable 
-strategies give best performance in each case. Push-based evaluation evaluates a
-system only when input is available, and is thus suitable for discrete inputs
-such as events. Pull-based evaluation evaluates the system as quickly as
-possible, polling for input, and is preferable for behaviors and signals.
-The initial implementations of FRP made use of pull-based evaluation for both
-behaviors and events. Reactive~\cite{Elliott2009}, as well as more recent systems such as
-``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
-events and pull-based evaluation for behaviors. This is known as ``push-pull''
-evaluation.
-
-All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
-have used pull-based evaluation for both signals and events. This is due to
-the ease of implementation of pull-based evaluation, and the types used for
-signal functions which do not permit distinguishing signals and events, or
-constructing only part of the input (for instance, one event occurrence.)
-
-A recent extension of signal-function FRP called N-Ary FRP~\cite{Sculthorpe2011}
-describes a method of typing signal functions which, as we will show, enables
-the push-based evaluation of events in a signal-function FRP system. The notion
-of signal vectors allows the representation of signal function inputs and
-outputs as combinations of signals and events, rather than a single signal which
-may contain multiple values, including option values for events. Signal vectors
-are uninhabited types, which can be used to type partial or full representations
-of the signal function inputs and outputs.
-
-We present TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
-favorite quotation of one of the author's philosophy instructors, used to
-demonstrate the ambiguity of language. The origin of the quotation is unknown.}
-a push-pull signal-function FRP system. We hope to demonstrate the feasibility
-of such an approach to FRP, and provide a basis for further research into
-efficient implementation of signal-function FRP. We also describe a powerful
-evaluation interface for TimeFlies, which permits us to use TimeFlies to
-describe applications which make use of multiple and differing IO libraries.
-
-Section~\ref{section:System_Design} describes design choices for the system,
-and provides an overview of the interface. Section~\ref{section:Implementation}
-describes how the system is implemented, and how the separation of evaluation
-between events and signals is achieved. Section~\ref{section:Discussion} is a
-discussion of the usefulness of our implementation. 
-Section~\ref{section:Ongoing_and_Further_Work} describes the current and future
-work on this system. Section~\ref{section:Related_Work} gives an overview of
-related efforts. Section~\ref{section:Conclusion} concludes.
-
 \section{System Design and Interface}
@@
 \end{figure}
-
-\section{Implementation}
-\label{section:Implementation}
-
-We now turn our attention to the implementation of the signal function
-system. We will discuss representations of inputs, outputs, and signal functions,
-as well as the implementations of specific signal function combinators.
-
-\subsection{Signal Functions}
-\label{subsection:Implementation-Signal_Functions}
-
-The design of signal functions specifies a family of types for the inputs and
-outputs of signal functions. Signal functions are not functions in the purest
-sense, however. They are not mappings from a single instance of their input
-type to a single instance of their output type. They must be implemented with
-respect to the temporal semantics of their inputs and outputs.
-
-We therefore start by creating a set of concrete datatypes for the inputs and
-outputs of signal functions. These datatypes will be parameterized by the input
-and output types of the signal function, and will not be exposed to the user of
-the library. Rather, they will specify how data is represented during the
-temporal evaluation of signal functions. We then implement signal functions
-as records of functions from these concrete types to these concrete types paired
-with new signal functions. The evaluation system for signal functions then
-maintains this record for the signal function it is evaluating, calls the
-correct function for the current input, actuates the outputs (using supplied
-functions from the outputs to monadic actions) and replaces the record with the
-newly supplied signal function record.
-
-In Section~\ref{section:System_Design_and_Interface} we presented signal vectors
-as a set of types. In order to be completely principled, we should isolate these
-types into their own {\em kind} (a sort of type of types); however, the Haskell
-extension for this was far from stable at the time this system was created.
-
-We create concrete representations which are parameterized over
-these types. These concrete representations will be expressed as GADTs, allowing
-each data constructor of the representation to fill in a specific signal vector
-type for the parameter of the representation.
-
-The first thing to represent is {\em samples}, which are sets of values for
-the signal components of a signal vector. Therefore, we create a representation
-which carries a value for every {\tt SVSignal} leaf of a signal vector. In order
-to do this, we restrict each of our constructors to a single signal vector type.
-So there are three leaf constructors: {\tt SVSample}, which carries a value; and
-{\tt SVSampleEvent} and {\tt SVSampleEmpty}, which do not. This ensures that the
-only way to represent a sample leaf is with the {\tt SVSample} constructor,
-which carries a value of the appropriate type. The datatype is shown in
-Figure~\ref{figure:signal_sample_datatype}.
-
-\begin{figure}
-\begin{code}
-data SVSample sv where
-  SVSample      ::    a
-                   -> SVSample (SVSignal a)
-  SVSampleEvent ::    SVSample (SVEvent a)
-  SVSampleEmpty ::    SVSample SVEmpty
-  SVSampleBoth  ::    SVSample svLeft
-                   -> SVSample svRight
-                   -> SVSample (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Datatype for signal samples.}
-\label{figure:signal_sample_datatype}
-\end{figure}
-
-What about the event components? We want to represent event occurrences,
-each of which will correspond to at most one event in the vector. So a different
-representation is called for. In this case, there will be only three
-constructors. One constructor will represent an event leaf, and the other will
-represent a single value on the left or right side of the node ({\tt SVAppend}),
-ignoring all of the type structure on the other side. This representation
-describes a path from the root of the signal vector, terminating at an event
-leaf with a value.
-
-By pattern matching on the path constructors, we can determine which subvector
-of a signal vector an event occurrence belongs to, repeatedly refining it until
-we determine which event in the vector the occurrence corresponds to. The
-datatype for occurrences is shown in Figure~\ref{figure:event_occurrence_datatype}.
-
-\begin{figure}
-\begin{code}
-data SVOccurrence sv where
-  SVOccurrence ::    a
-                  -> SVOccurrence (SVEvent a)
-  SVOccLeft    ::    SVOccurrence svLeft
-                  -> SVOccurrence (SVAppend svLeft svRight)
-  SVOccRight   ::    SVOccurrence svRight 
-                  -> SVOccurrence (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Datatype for event occurrences.}
-\label{figure:event_occurrence_datatype}
-\end{figure}
-
-We add one more representation for signals, in order to avoid uneccessary
-representations of the values of all signals when not all signals have changed
-their values. This representation allows us to represent the values of zero or
-more of the signals in a signal vector. To accomplish this, we replace the
-individual constructors for the {\tt SVEmpty} and {\tt SVEvent} leaves with %there is a better word, something about "useless but not really", for this
-a single, unconstrained constructor. This constructor can represent an arbitrary
-signal vector. We can use the constructor for signal vector nodes and the 
-constructor for sample leaves to represent the updated values, while filling
-in the unchanged portions of the signal vector with this general constructor.
-This datatype is shown in Figure~\ref{figure:signal_update_datatype}.
-
-\begin{figure}
-\begin{code}
-data SVDelta sv where
-  SVDeltaSignal  ::    a
-                    -> SVDelta (SVSignal a)
-  SVDeltaNothing ::    SVDelta sv
-  SVDeltaBoth    ::    SVDelta svLeft
-                    -> SVDelta svRight
-                    -> SVDelta (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Datatype for signal updates.}
-\label{figure:signal_update_datatype}
-\end{figure}
-
-We now have concrete datatypes for an implementation to operate on. Our next
-task is to represent transformers of temporal data, which themselves may change
-with time. The common approach to this task is sampling, in which a program
-repeatedly checks for updated information, evaluates it, updates some state,
-and produces an output. This is the essence of pull-based evaluation.
-
-Another approach is notification, in which the program exposes an interface
-which the source of updated information may invoke. This is a repeated entry
-point to the program, which causes the program to perform the same tasks
-listed above, namely, evaluate the updated information, update state, and
-produce output. The strategy of notification as opposed to repeated checking is
-the essence of push-based evaluation.
-
-Signal functions are declarative objects, and not running processes. They have
-no way to invoke sampling themselves. They can, however, expose separate
-interfaces for when sampling is invoked, and when they are notified of an event
-occurrence. This creates two control paths through a signal function. One of
-these control paths is intended to be invoked regularly and frequently with
-updates to the time and sample values, and the other is intended to be invoked
-only when an event occurs. The benefit of separating these control paths is that
-events are no longer defined in terms of sampling intervals, and need not even
-be considered in sampling, except when they are generated by a condition on a
-sample. On the other hand, events can be responded to even if the time has not
-yet come for another sample, and multiple events can be responded to in a single
-sampling interval.
-
-We represent signal functions as a GADT with three type parameters and two 
-constructors. The first type parameter represents the initialization state,
-and is specialized to {\tt Initialized} or {\tt NonInitialized} depending on the
-constructor. The other two type parameters are the input and output signal
-vectors, respectively. The signal functions that a user will compose are\
-non-initialized signal functions. They must be provided with an initial set of
-input signals (corresponding to time zero). When provided with this input, they
-produce their time-zero output, and an initialized signal function. The datatype
-is shown in Figure~\ref{figure:signal_function_datatype}.
-
-\begin{figure}
-\begin{code}
-data Initialized
-
-data NonInitialized
-
-data SF init svIn svOut where
-  SF     ::    (SVSample svIn 
-                  -> (SVSample svOut,
-                      SF Initialized svIn svOut)) 
-            -> SF NonInitialized svIn svOut
-  SFInit ::    (Double 
-                  -> SVDelta svIn
-                  -> (SVDelta svOut,
-                      [SVOccurrence svOut],
-                      SF Initialized svIn svOut)) 
-            -> (SVOccurrence svIn
-                  -> ([SVOccurrence svOut],
-                      SF Initialized svIn svOut))
-            -> SF Initialized svIn svOut
-\end{code}
-\hrule
-\caption{Datatype and empty types for signal functions.}
-\label{figure:signal_function_datatype}
-\end{figure}
-
-Initialized signal functions carry two continuations. The first continuation
-takes a time differential and a set of signal updates, and returns a set of
-signal updates, a collection of event occurrences, and a new initialized signal
-function of the same type. This is the continuation called when sampling.
-
-The second continuation takes an event occurrence, and returns a collection of
-event occurrences and a new signal function of the same type. This continuation
-is only called when there is an event occurrence to be input to the signal
-function.
-
-Note that each of these continuations uses one or more of the concrete
-representations of signal vectors, and applies the type constructor for the
-representation to the input or output signal vector for the signal function.
-
-Having specified a type constructor for signal functions, we must now provide
-combinators which produce signal functions typed with this type constructor.
-Space precludes a complete description of the combinators in the TimeFlies
-system, but we will present
-
-The simplest example of the implementation of is the {\tt identity} combinator,
-shown in Figure~\ref{figure:identity_implementation}. This signal function
-simply passes all of its inputs along as outputs. The initialization function
-simply passes along the received sample and outputs the initialized version of
-the signal function. The initialized version of the input is similar, but is
-self-referential. It outputs itself as its replacement. This is standard for
-simple and routing combinators which are not reactive, and simply re-arrange,
-discard, or combine signals and events.
-
-In order for our primitive signal functions to be useful, we need a means of
-composing them. Serial composition creates one signal function from two, by
-using the output of one as the input of the other. The serial composition
-combinator is styled {\tt (>>>)}. The implementation of this operator is one
-place where the advantage of responding to events independently from signal
-samples becomes clear.
-
-This is the only primitive combinator which takes two signal functions, and
-thus, it is the only way to combine signal functions. Parallel, branching, and
-joining composition can be achieved by modifying signal functions with the
-{\tt first} and {\tt second} combinators and composing them with the
-routing and joining combinators.
-
-Combinators which take one or more signal functions as input must recursively
-apply themselves, as is shown in the implementation of serial composition
-(Figure~\ref{figure:serial_composition_implementation}). They must also
-handle initialization, retaining the initialized signal functions and passing
-them to the initialized version of the combinator.
-
-The switch combinator is the means of introducing reactivity into a signal
-function. This combinator allows a signal function to replace itself by
-producing an event occurrence. The combinator wraps a signal function, and 
-observes an event on the right side of the output signal vector. At the first
-occurrence of the event, the signal function carried by the occurrence replaces
-the signal function.
-
-The switch combinator stores the input sample provided during initialization,
-and updates it with the input signal updates. When the wrapped signal function
-produces an occurrence carrying a new signal function, that signal function is
-initialized with the stored input sample. It is then "wrapped" by another
-function which closes over its output sample, and outputs the sample as a signal
-update as the next time step. After this, it acts as the new signal function.
-This wrapping has some performance implications, which are discussed
-
-This combinator checks the outputs of the wrapped
-signal function for an event occurrence from which an uninitialized signal
-function is extracted. The switch combinator stores the full sample
-for its input vector (which is identical to the input vector of the supplied
-signal function) to initialize the new signal function. This also demands that
-it add a wrapper to the new signal function which waits for the next sampling
-interval and actuates the sample output at initialization as an output set
-of changes to the signal. This has some performance implications, which are
-discussed in Chapter~\ref{chapter:Evaluation_and_Comparisons}.
-
-Most of the routing combinators are simple to implement. The only task is to add
-remove, or replace routing constructors on signal updates and event occurrences.
-Since these signal functions are stateless and primitive, they can simply
-return themselves as their replacements. The {\tt swap} combinator is shown as
-an example in Figure~\ref{figure:swap_implementation}
-
-The {\tt first} and {\tt second} combinators are similar to serial composition,
-but they transform only one signal function. For signal changes, they must split
-the set of input changes into those which will be passed to the signal function
-and those which will be simply passed along to the output, and then recombine
-them on the other side. For event occurrences, the occurrence must be
-pattern-matched to determine whether to call the event continuation from the
-provided signal function or passed through, and output event occurrences must
-have the suitable routing constructor re-applied. In any case, when a
-continuation has been applied, the combinator must be recursively applied to the
-new signal function.
-
-The looping feedback combinator is particularly tricky. As it is currently
-implemented, the initial sample for the right side of the input signal vector to
-the supplied function is the right side of the output sample. This is acceptable,
-given Haskell's non-strict evaluation strategy, but it is necessary that the
-right side of the signal function's output not be immediately dependent on its
-input. The feedback combinator makes use of Haskell's lazy evaluation to
-feed events back into the combinator, and stores signal updates until the next
-sample. Signal samples are thus automatically decoupled after initialization.
-The implementation makes use of the recursive nature of the {\tt let} construct
-in Haskell, and the non-strict evaluation of Haskell, to implement feedback.
-
-The {\tt filter} combinators are simple to implement. Their sampling
-continuation is superfluous, and the event continuation merely applies the
-supplied function, and constructs an output list based on the result.
-
-The {\tt accumulate} combinators are implemented in terms of the {\tt filter}
-and {\tt switch} combinators, as shown in
-Figure~\ref{figure:accumulate_implementation}.
-
-The implementation of the joining combinators is simple. The {\tt union}
-combinator simply passes along every event occurrence it receives on either
-input, stripping off the left and right combinator. This is acceptable since we
-do not insist on a total ordering of events, or an event time resolution greater
-than the sampling rate. The {\tt combineSignals} combinator maintains the values
-of both signals, and applies the combination function whenever one is updated.
-The {\tt capture} combinator maintains the input signal value, and adds it to
-each event occurrence.
-
-Time dependence is introduced by the {\tt time}, {\tt delay}, and {\tt integrate}
-combinators. The time combinator simply sums the time updates and provides the
-sum as a signal output. The {\tt delay} combinator keeps a table of events
-which have come in, along with their schedule occurrence time, and produces
-them as output when time advances far enough. The integrate combinator performs
-rectangle-rule integration on signal samples with respect to time.
-
-The implementation strategy leaves room for optimizations. In particular, an
-additional constructor for time-independent signal functions would allow
-portions of a signal function to forgo evaluation during time steps unless they
-had signal updates. Optimizations in the style of Yampa, observed by keeping
-an updated AST for the signal function and pattern-matching on it when switching,
-might further improve performance. In particular, collapsing nested or
-serially-composed versions of the {\tt switchWait} step when switching would
-remove at least some of the observed dependence of performance on sampling rate.
-Nevertheless, this implementation performs quite well as it currently exists, as
-we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparisons}.
-
-\subsection{Evaluation Interface}
-\label{subsection:Implementation-Evaluation_Interface}
-The evaluation interface provides the means of evaluating a signal function
-with inputs and producing effects in response to the signal function's outputs.
-We would like to produce a set of constructs that interacts well with Haskell's
-system for external IO.
-
-The evaluation interface translates between signal functions and the standard
-Haskell construct for sequencing effects and external inputs, namely,
-{\em monads}~\cite{PeytonJones1993}. The inspiration for monads is drawn from
-the rather esoteric domain of category theory, but the concept as applied to
-programming languages is actually rather simple.
-
-By formulating
-the evaluation interface as a monad transformer, we need only define the
-operations relevant to the evaluation of signal functions, and we can depend
-on the constructs of the IO monad to interact with whatever inputs and outputs
-are necessary. In some cases, we may not wish to use the IO monad at all (e.g
-for testing or simulation). In this case, we can parameterize over another
-monad, such as the Identity monad (which has no special operations and whose
-context is just the value), or the State monad (which maintains an implicit 
-state accessible by {\tt get} and {\tt put} operations).
-
-The evaluation interface is exported as shown in
-Section~\ref{section:System_Design_and_Interface-Evaluator}.
-
-The {\tt SFEvalState} type constructor parameterizes over input types for signal
-functions, and underlying monads for a monad transformer, but is not itself
-a monad transformer. It describes the state of signal function evaluation.
-It consists of a record with four members: the current signal function,
-the set of handlers for outputs, the current input signal delta, and the last
-sample time.
-
-The {\tt SFEvalT} monad transformer is a newtype wrapper around the {\tt StateT}
-monad available in the Haskell {\tt transformers} package. The StateT monad
-wraps any other monad, providing a state which may be read using the {\tt get}
-action and replaced with the {\tt put} action. An action in the underlying
-monad is obtained by supplying a {\tt StateT} value and an initial state
-value to the {\tt runStateT} function.
-
-The {\tt SFEvalT} monad transformer does not make the {\tt get} and {\tt put}
-actions available, but uses them in its implementation of the {\tt push},
-{\tt update}, and {\tt sample} actions.
-
-The {\tt push} action pushes an event onto the input of the signal function,
-resulting in immediate evaluation of the relevant components signal function
-and possible actuation of handlers specified for output events. It is
-implemented by fetching the {\tt SFEvalState}, applying the event continuation
-of the signal function contained by the {\tt SFEvalState} to the pushed event,
-thus obtaining a new signal function and a list of events, applying the handlers
-contained in the {\tt SFEvalState} to the output events, replacing the signal
-function in the {\tt SFEvalState}, and replacing the {\tt SFEvalState}.
-
-The {\tt update} action updates the current input signal delta, which will be
-sampled at the next {\tt step} action. It has no immediately observable effect.
-It simply updates the value of one signal in the input signal vector, a delta
-of which is stored in the {\tt SFEvalState} record.
-
-The {\tt step} action takes a time delta, and calls the signal continuation of
-the stored signal function. It actuates the outputs using the stored handlers,
-replaces the stored delta with an empty delta, and stores the resulting new
-signal function in the state.
-
-\section{Discussion}
-\label{section:Discussion}
-
-The system presented here, TimeFlies, demonstrates how using signal vectors
-to type inputs and outputs enables push-based evaluation of events in a
-signal-function system. We take advantage of this representation in several ways.
-
-First, by separating components of inputs and outputs in the types, we are free
-to create distinct, and often partial, representations of the input or output
-of a signal function. This enables us to represent only the event occurrence
-being pushed at that time.
-
-Second, this separation also permits us to separate the process of gathering
-the input to a signal function, and the process of handling its output, into
-different points in a program. Using the evaluation interface described, an
-event occurrence may be pushed onto one input of a signal function from one
-point in a program (e.g. a mouse click handler), an input signal
-may be updated in another (e.g. a mouse movement handler), and finally the
-system may be sampled in a third place (e.g. an animation or audio timed
-callback).
-
-Finally, this approach enables further work on the implementation of the signal
-function system to be separated from changes in the interface. By enabling
-differing representations of the inputs and outputs of signal functions, we are
-free to change these representations without the need to further constrain the
-input and output types.
-
-\section{Ongoing and Further Work}
-\label{section:Ongoing_and_Further_Work}
-
-TimeFlies, the system described here, has been implemented, but not
-extensively tested. The immediate goal is to create a real-time application
-which will permit a performance and implementation comparison of TimeFlies with
-Yampa, the current state-of-the-art pull-based signal-function system.
-
-In the future, we hope to apply run-time optimizations, using the technique used
-for Yampa, to create a push-pull self-optimizing signal-function system. Further,
-we hope to use this system as a basis for exploring signal-function FRP as a
-basis for general-purpose application frameworks.
-
-\section{Related Work}
-\label{section:Related_Work}
-
-Signal Function FRP was introduced as a model for Graphical User Interfaces~\cite{Courtney2001-1}.
-The system was originally termed ``AFRP'' (Arrowized FRP). Yampa is a rewrite of
-AFRP where signal functions apply a number of ad-hoc optimizatons to themselves
-as they evolve. Yampa demonstrated a modest performance improvement
-over AFRP~\cite{Nilsson2005}.
-
-Reactive is a classic FRP system which implements push-based evaluation for events
-by transforming behaviors to ``reactive~normal~form,'' where a behavior
-is a non-reactive behavior running inside a switch, whose event stream carries
-behaviors in reactive normal form. The system is evaluated by forking a Haskell
-thread to repeatedly sample the non-reactive behavior, and then blocking on the
-evaluation of the first occurrence in the event stream. When this occurrence
-is yielded, the evaluation thread for the behavior is killed and a new
-thread forked to evaluate the new behavior~\cite{Elliott2009}.
-
-An alternate push-based system, based on {\em signal segments}, was presented in
-work on the Curry-Howard correspondence between a form of temporal logic (LTL)
-and FRP~\cite{Jeffrey2012}. However, this system continues to represent events
-as option-valued signals. It therefore cannot be push-based in its response to
-individual events, and cannot be push-pull, evaluating signals by pull-based
-evaluation and events by push-based evaluation. It is push-based not in the
-sense of evaluating events only when they occur, but rather in that it provides
-a means to push signal segments (samples of a signal over intervals of time) to
-the system and wait on its response.
-
-\section{Conclusion}
-\label{section:Conclusion}
-
-We have presented TimeFlies, a system for push-pull signal-function Functional
-Reactive Programming, and have shown how the use of a signal vectors as input
-and output types for signal functions, together with GADT-based representations
-of the inputs and outputs, permits the implementation of a push-pull system.
-
-We have also described a general and flexible monadic evaluation interface for
-TimeFlies, which permits us to interface the TimeFlies system with different
-styles of IO systems, including multiple IO systems in the same application.
-
-This opens up the exciting possibility that a signal-function FRP could become
-an efficient and general framework for writing interactive applications.
-
-\bibliographystyle{splncs}
-\bibliography{thesis}
-
-\end{document}

### 2012-11-27 Docs/IFL2012/discussion.tex
Break IFL paper into multiple files.

--- 2012-11-24 Docs/IFL2012/paper.tex
@@
-\documentclass[draft]{llncs}
-
-\usepackage{fancyvrb}                                       % For code environment
-\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small} %
-
-\title{Push-Pull Signal-Function Functional Reactive Programming}
-\titlerunning{Push-Pull Signal-Function FRP}
-
-\author{Edward Amsden}
-\institute{Rochester Institute of Technology\\\email{eca7215@cs.rit.edu}}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-Functional Reactive Programming is a promising class of systems for writing
-interactive and time-dependent programs. Signal-function FRP is a subclass of
-these systems which provides advantages of modularity and correctness, but
-has proven difficult to efficiently implement.
-
-The abstraction of signal vectors provides the necessary type apparatus to
-distinguish components of the input and output of signal functions which benefit
-from a push-based implementation from those which benefit from a pull-based
-implementation, and to combine both implementation strategies in a single system.
-
-We describe a signal-function FRP system which provides push-based evaluation
-for events, pull-based evaluation for signals, and a simple monadic evaluation
-interface which permits the system to be easily integrated with one or more
-IO systems.
-\end{abstract}
-
-\section{Introduction}
-\label{section:Introduction}
-
-Functional Reactive Programming (FRP) is a class of systems for describing
-reactive programs. Reactive programs are programs which, rather than taking
-a single input and producing a single output, must accept multiple inputs and 
-alter temporal behavior, including the production of multiple outputs, based
-on these inputs.
-
-An FRP system will provide a means of manipulating {\em behaviors} and
-{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
-but the definition is the same. A behavior or signal is, semantically, a
-function from time to a value. An event is a discrete, possibly infinite, and 
-time-ordered sequence of occurrences, which are times paired with values.
-
-FRP systems can generally be categorized as ``classic FRP,'' 
-which corresponds to the originally described FRP system in that behaviors
-and events are manipulated directly and are first-class values in the FRP
-system, or ``signal-function FRP,'' in which behaviors (generally termed
-signals in this approach) and events are not first-class values, but signal
-functions are first class values. Signal functions are time-dependent and
-reactive transformers of signals, events, or combinations of signals and events.
-
-FRP combines behaviors and events through the use of {\em switching}, in which
-a behavior (in classic FRP) or a signal function (in signal-function FRP) is
-replaced by a new behavior or signal function carried by an event occurrence.
-
-Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
-Recent work on classic FRP has focused on efficient implementation. One approach to
-efficiency is to separate the evaluation of behaviors and events, since suitable 
-strategies give best performance in each case. Push-based evaluation evaluates a
-system only when input is available, and is thus suitable for discrete inputs
-such as events. Pull-based evaluation evaluates the system as quickly as
-possible, polling for input, and is preferable for behaviors and signals.
-The initial implementations of FRP made use of pull-based evaluation for both
-behaviors and events. Reactive~\cite{Elliott2009}, as well as more recent systems such as
-``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
-events and pull-based evaluation for behaviors. This is known as ``push-pull''
-evaluation.
-
-All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
-have used pull-based evaluation for both signals and events. This is due to
-the ease of implementation of pull-based evaluation, and the types used for
-signal functions which do not permit distinguishing signals and events, or
-constructing only part of the input (for instance, one event occurrence.)
-
-A recent extension of signal-function FRP called N-Ary FRP~\cite{Sculthorpe2011}
-describes a method of typing signal functions which, as we will show, enables
-the push-based evaluation of events in a signal-function FRP system. The notion
-of signal vectors allows the representation of signal function inputs and
-outputs as combinations of signals and events, rather than a single signal which
-may contain multiple values, including option values for events. Signal vectors
-are uninhabited types, which can be used to type partial or full representations
-of the signal function inputs and outputs.
-
-We present TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
-favorite quotation of one of the author's philosophy instructors, used to
-demonstrate the ambiguity of language. The origin of the quotation is unknown.}
-a push-pull signal-function FRP system. We hope to demonstrate the feasibility
-of such an approach to FRP, and provide a basis for further research into
-efficient implementation of signal-function FRP. We also describe a powerful
-evaluation interface for TimeFlies, which permits us to use TimeFlies to
-describe applications which make use of multiple and differing IO libraries.
-
-Section~\ref{section:System_Design} describes design choices for the system,
-and provides an overview of the interface. Section~\ref{section:Implementation}
-describes how the system is implemented, and how the separation of evaluation
-between events and signals is achieved. Section~\ref{section:Discussion} is a
-discussion of the usefulness of our implementation. 
-Section~\ref{section:Ongoing_and_Further_Work} describes the current and future
-work on this system. Section~\ref{section:Related_Work} gives an overview of
-related efforts. Section~\ref{section:Conclusion} concludes.
-
-\section{System Design}
-\label{section:System_Design}
-
-Our goal is to produce a composable and efficient FRP system. Signal function
-FRP has an advantage in terms of composability, because it permits the
-construction of self-contained objects through both input and output composition,
-rather than purely through output composition as in classic FRP. Signal-function
-FRP also avoids problematic properties common to classic FRP systems such as a
-large class of time and space leaks~\cite{Liu2007}.
-
-Efficient implementations of signal-function FRP have been approached through
-runtime optimization~\cite{Nilsson2005}, but all implementations have been
-pull-based for both signals and events. A truly efficient implementation will
-likely combine run-time optimization with push-based evaluation for events.
-
-The concept of N-Ary FRP was to encode additional safety properties into the
-types of a signal-function FRP system~\cite{Sculthorpe2011}. This system
-introduced the concept of {\em signal vectors} as input and output types for
-signal functions. Signal vectors are combinations of signals and events. In
-N-Ary FRP, signal functions are represented at the type level, and type-level
-functions (type families in Haskell) are used to specify the representations
-of signal functions.
-
-In our system, we construct representations of signal vectors using
-Generalized Algebraic Datatypes~\cite{Cheney2003,Xi2003}. The use of GADTs
-enables us to use signal vectors to instantiate type parameters in the types
-of signal vector representations, as well as in the types of signal functions.
-GADTs are available as an extension in the Glasgow Haskell Compiler~\cite{PeytonJones2006}.
-
-This approach permits us to construct partial inputs and outputs for signal
-functions. For instance, we can construct a representation for a single event
-in a signal vector, and another representation of updated values for a subset of
-the signals in the signal vector, and yet another carrying values for all
-signals in a signal vector. The signal vector types are shown in
-Figure~\ref{figure:signal_vectors}.
-
-\begin{figure}[t]
-\begin{code}
-data SVEmpty
-data SVSignal a
-data SVEvent a
-data SVAppend svLeft svRight
-\end{code}
-\hrule
-\caption{Signal vectors.}
-\label{figure:signal_vectors}
-\end{figure}
-
-With the ability to construct partial representations of signal vectors, we can
-represent signal functions with a datatype carrying multiple functions, one for
-each type of input. This allows us to separate the pull-based processing of
-signals from the push-based reaction to events.
-
-The exposed interface is a set of combinators for constructing signal functions,
-as well as combinators for describing the evaluation of a signal function. The
-interface is shown in Fig.~\ref{figure:signal_function_interface}.
-
-\begin{figure}
-\begin{code}
-
--- Signal Functions
-type :~> svIn svOut
-
--- Infix type alias for SVAppend
-type :^: svLeft svRight = SVAppend svLeft svRight
-
--- Basic signal functions
-identity :: sv :~> sv
-constant :: a -> SVEmpty :~> SVSignal a
-never    :: SVEmpty :~> SVEvent a
-asap     :: a -> SVEmpty :~> SVEvent a
-after    :: Double -> a -> SVEmpty :~> SVEvent a
-
--- Lifting pure functions
-pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal a
-pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent a
-
--- Composition and routing
-(>>>)           :: (svIn :~> svMiddle) -> (svMiddle :~> svOut) 
-                   -> (svIn :~> svOut)
-first           :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
-second          :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut) 
-swap            :: (svLeft :^: svRight) :~> (svRight :^: svLeft)
-copy            :: sv :~> (sv :^: sv)
-ignore          :: sv :~> SVEmpty
-cancelLeft      :: (SVEmpty :^: sv) :~> sv
-cancelRight     :: (sv :^: SVEmpty) :~> sv
-uncancelLeft    :: sv :~> (SVEmpty :^: sv)
-uncancelRight   :: sv :~> (sv :^: SVEmpty)
-associate       :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3)) 
-unassociate     :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
-
--- Reactivity
-switch :: (svIn :~> (svOut :^: SVEvent (svIn :~> svOut))) 
-          -> svIn :~> svOut
-
--- Feedback
-loop :: ((svIn :^: svLoop) :~> (svOut :^: svLoop)) -> svIn :~> svOut
-
--- Time dependence
-time  :: SVEmpty :~> Double
-delay :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
-class TimeIntegrate
-
--- Joining
-union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
-combineSignals :: (a -> b -> c) 
-                  -> (SVSignal a :^: SVSignal b) :~> SVSignal c
-capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent a
-
--- Events
-filter         :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-\end{code}
-\hrule
-\caption{Signal function interface.}
-\label{figure:signal_function_interface}
-\end{figure}
-
-Signal functions are produced by combining primitive signal functions using
-the {\tt >>>} (sequential composition), {\tt first}, and {\tt second} routing
-combinators. Other routing signal functions are provided, but they are intended
-to be combined with, rather than to modify, other signal functions.
-
-\subsection{Evaluation Interface}
-The evaluation interface provides a monad~\cite{PeytonJones1993,PeytonJones2001}
-for specifying input to signal functions and handling their output. This
-interface also allows input actions to be separated from output actions, and
-from each other, so that a signal function may receive input from multiple IO
-systems and have its output handled by multiple output systems.
-
-The evaluation interface is a monad transformer~\cite{Liang1995}. This allows
-a signal function evaluation to be constructed from primitive evaluation actions
-(event pushing, input signal updating, and sampling) as well as actions from the
-underlying monad. These actions can then be run in the underlying monad
-to actuate the signal function. Because the actuation takes a signal function's
-state as input and produces a new state as output, in addition to the monadic
-side effects of handling the signal function's output, different evaluation
-actions may be taken on the same signal function from distinct locations within
-an application's code. This allows the evaluation interface to be easily
-integrated with event-loop style systems as well as traditional imperative
-IO systems.
-
-The evaluation interface consists of the monad transformer type, functions for
-constructing event inputs, initial input signal samples, and input signal
-updates, functions for constructing the vector of output handlers for a signal
-function, and evaluation actions. The full interface is shown in Fig.~\ref{figure:evaluation_interface}.
-
-\begin{figure}
-\begin{code}
--- Input helpers
-class SVRoutable -- Instances: SVSignalUpdate,
-                 -- SVEventOccurrence
-svLeft :: (SVRoutable r) => 
-          r svLeft -> r (SVAppend svLeft svRight)
-svRight :: (SVRoutable r) =>
-           r svRight -> r (SVAppend svLeft svRight)
-
--- Signal inputs
-data SVSignalUpdate
-data SVSample
-sample :: a -> SVSample (SVSignal a)
-sampleEvt :: SVSample (SVEvent a)
-sampleNothing :: SVSample SVEmpty
-combineSamples :: SVSample svLeft -> svSample svRight 
-                  -> svSample (svLeft :^: svRight)
-svSig :: a -> SVSignalUpdate (SVSignal a)
-
--- Event inputs
-data SVEventInput
-svOcc :: a -> SVEventInput (SVEvent a)
--- Actuation
-update :: (Monad m) => SVSignalUpdate svIn 
-           -> SFEvalT svIn svOut m ()
-push   :: (Monad m) => SVEventInput svIn
-          -> SFEvalT svIn svOut m ()
-sample :: SVEvalT svIn svOut m ()
--- Running
-data SVHandler
-emptyHandler :: SVHandler m SVEmpty
-eventHandler :: (a -> m ()) 
-                -> SVHandler m (SVEvent a)
-signalHandler :: (a -> m ()) 
-                 -> SVHandler m (SVSignal a)
-combineHandlers :: SVHandler m svLeft
-                   -> SVHandler m svRight
-                   -> SVHandler m (svLeft :^: svRight)
-initSFEvalT :: SVHandler m svOut -> SVSample svIn 
-               -> Double ->(svIn :~> svOut)
-               -> SFEvalState m svIn svOut
-runSFEvalT :: SFEvalT svIn svOut m a
-              -> SFEvalState m svIn svOut
-              -> m a
-\end{code}
-\hrule
-\caption{Evaluation interface.}
-\label{figure:evaluation_interface}
-\end{figure}
-
-\section{Implementation}
-\label{section:Implementation}
-
-We now turn our attention to the implementation of the signal function
-system. We will discuss representations of inputs, outputs, and signal functions,
-as well as the implementations of specific signal function combinators.
-
-\subsection{Input and Output Representations}
-\label{subsection:Input_and_Output_Representations}
-As discussed in Section~\ref{section:System_Design}, our implementation requires
-representations of the inputs and outputs of signal functions. Since signal
-functions are typed using signal vectors, the representation types must be
-parameterized over signal vectors. But signal vectors are uninhabited types, and
-in ordinary Haskell we cannot declare a data constructor which uses a component
-of a type parameter. Further, we cannot restrict which types may instantiate a
-type parameter for a particular data constructor.
-
-GADTs lift these restrictions, permitting type constraints to be applied
-to individual data constructors. In the declaration of a GADT, a type signature
-is provided for each data constructor, including the types of the constructor's
-parameters. When a GADT is pattern-matched, the types of the parameters are
-inferred using the constraints given in this signature. This is called type
-refinement.
-
-Using GADTs, we can construct representations of signal vectors to use in our
-implementation.
-
-When constructing a representation, we consider how to represent each
-component of a signal vector, whether it is a signal element, an empty
-element, or an append node. For instance, in order to represent a signal
-sample for all signals in an element, we construct the GADT shown in Figure~\ref{figure:signal_sample_representation}.
-
-\begin{figure}
-\begin{code}
-data SVSample sv where
-  SVSample        ::    a -> SVSample (SVSignal a)
-  SVSampleEvt     ::    SVSample (SVEvent a)
-  SVSampleNothing ::    SVSample SVEmpty
-  SVSampleBoth    ::    SVSample svLeft
-                     -> SVSample svRight
-                     -> SVSample (svLeft :^: svRight)
-\end{code}
-\hrule
-\caption{A representation which carries a value for each signal in a signal vector.}
-\label{figure:signal_sample_representation}
-\end{figure}
-
-\subsection{Signal Function Representations}
-\label{subsection:Signal_Function_Representations}
-With representations for signal function inputs and outputs in place, we turn
-to the representation of signal functions. A signal function must be able to
-respond to time increments and new signal samples, as well as event occurrences.
-Since we wish to be able to ``push'' event occurrences independently of
-sampling, events are handled using the last-updated time and signal sample.
-
-When a signal function is asked to respond to either of these types of inputs,
-it must produce an output consisting of zero or more event occurrences, a new
-signal function with the same type (to enable reactivity and state), and, if
-it is responding to a time and sample update, a delta which represents updates
-to the sample of its output signals.
-
-Finally, there is a special case for a signal function at time 0. It must be
-provided with an initial input sample before it can respond to incremental time
-and sample updates or to event occurrences.
-
-Signal functions are represented as a datatype with two constructors. The first
-constructor wraps a function from an input sample to an output sample and an
-initialized signal function. The second wraps two functions. The first accepts
-a time delta and a signal delta, and produces a signal delta, a list of event
-occurrences, and a new signal function as the output. The second accepts an
-event occurrence and produces a list of event occurrences and a new signal
-function as the output. The signal function representation is shown in Figure~\ref{figure:signal_function_representation}.
-
-\begin{figure}[t]
-\begin{code}
-data Initialized
-data NonInitialized
-
-type :~> svIn svOut = SF NonInitialized svIn svOut
-
-data SF init svIn svOut where
-  SF :: (SVSample svIn -> (SVSample svOut, SF Initialized svIn svOut))
-        -> SF NonInitialized svIn svOut
-  SFInit :: (Double -> SVDelta svIn 
-             -> (SVDelta svOut, [SVOccurrence svOut],
-                 SF Initialized svIn svOut))
-            -> (SVOccurrence svIn -> ([SVOccurrence svOut],
-                                      SF Initialized svIn svOut))
-            -> SF Initialized svIn svOut
-\end{code}
-\hrule
-\caption{Signal function representation}
-\label{figure:signal_function_representation}
-\end{figure}
-
-This representation allows a signal function to respond to events and time
-updates separately, and does not enforce the representation of events during
-time updates, as previous signal function systems do.
-
-\subsection{Signal Function Implementations}
-\label{subsection:Signal_Function_Implementations}
-
-Now that we have established representations for signal functions and their
-inputs and outputs, we show several examples of how signal function combinators
-are implemented.
-
-The {\tt identity} signal function's initialization function takes a sample,
-and returns that sample as its output sample. The initialized function returned
-is the {\tt identityInit} function, which is not exported by the module.
-The time and sample function ignores the provided time delta, produces the
-input signal delta as the output signal delta, and returns {\tt identityInit} as
-the new signal function. The event function takes an event occurrence and
-returns a singleton list containing that occurrence, along with the
-{\tt identityInit} function as the replacement signal function. The
-implementation is shown in Figure~\ref{figure:identity_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-identity :: sv :~> sv
-identity = SF (\sample -> (sample, identityInit))
-
-identityInit :: SF Initialized sv sv
-identityInit = SFInit (\_ delta -> (delta, [], identityInit))
-                      (\occ -> ([occ], identityInit))
-\end{code}
-\hrule
-\caption{{\tt identity} signal function implementation.}
-\label{figure:identity_signal_function_implementation}
-\end{figure}
-
-The {\tt constant} signal function is initialized with a value, which it then
-produces as its output forever. Thus, the initialized version of {\tt constant}
-never produces any output other than itself as a replacement
-signal function, and an empty signal delta. The implementation of {\tt constant}
-is shown in Fig.~\ref{figure:constant_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-constant :: a -> SVEmpty :~> SVSignal a
-constant x = SF (\_ -> (constantInit, sample x))
-
-constantInit :: SF Initialized SVEmpty (SVSignal a)
-constantInit = SFInit (\_ _ -> (deltaNothing, [], constantInit))
-                      (\_ -> ([], constantInit))
-\end{code}
-\hrule
-\caption{{\tt constant} signal function implementation.}
-\label{figure:constant_signal_function_implementation}
-\end{figure}
-
-The {\tt asap} signal function implementation requires a parameter for the
-initialized version of the signal function, to specify the value of the event
-occurrence. It replaces itself with the initialized version of the {\tt never}
-signal function after the first time step. The implementation is shown in
-Fig.~\ref{figure:asap_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-asap :: a -> SVEmpty :~> SVEvent a
-asap x = SF (\_ -> (sampleEvt, asapInit x))
-
-asapInit :: a -> SF Initialized SVEmpty (SVEvent a)
-asapInit x = SFInit (\_ _ -> (deltaNothing, [occurrence x], neverInit))
-                    (\_ -> (never))
-\end{code}
-\hrule
-\caption{{\tt asap} signal function implementation.}
-\label{figure:asap_signal_function_implementation}
-\end{figure}
-
-The {\tt filter} function accepts events, applies a {\tt Maybe} predicate to
-their values, and produces the value produced by the predicate as an event
-occurrence, or no occurrence if the value is {\tt Nothing}. It is implemented
-by having the event occurrence response function apply the {\tt maybe} function
-from the Haskell prelude to the value returned by the predicate, as shown in 
-Fig.~\ref{figure:filter_signal_function_implementation}
-
-\begin{figure}[t]
-\begin{code}
-filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-filter p = SF (\_ -> (sampleEvt, filterInit p))
-
-filterInit :: (a -> Maybe b) -> SF Initialized (SVEvent a) (SVEvent b)
-filterInit p = SFInit (\_ _ -> (deltaNothing, [], filterInit p))
-                      (\evtOcc -> (maybe [] ((:[]) . occurrence) $
-                                   fromOccurrence evtOcc,
-                                   filterInit p))
-\end{code}
-\hrule
-\caption{{\tt filter} signal function implementation.}
-\label{figure:filter_signal_function_implementation}
-\end{figure}
-
-The {\tt associate} signal function is a routing signal function. It transforms
-a signal vector which is left-associated at the top level to one that is
-right-associated at the top level. Its implementation is representative of all
-of the routing functions, and is shown in Fig.~\ref{figure:associate_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-associate :: SVAppend (SVAppend sv1 sv2) sv3
-             :~> SVAppend sv1 (SVAppend sv2 sv3)
-associate = 
-  SF (\sigSample -> let (sigSampleLeft, sigSampleRight) = 
-                          splitSample sigSample
-                        (sigSampleLeftLeft, sigSampleLeftRight) =
-                          splitSample sigSampleLeft
-                    in (combineSamples sigSampleLeftLeft 
-                        (combineSamples sigSampleLeftRight sigSampleRight),
-                        associateInit))
-
-associateInit :: SF Initialized (SVAppend (SVAppend sv1 sv2) sv3)
-                                (SVAppend sv1 (SVAppend sv2 sv3))
-associateInit = SFInit (\_ sigDelta -> let (sigDeltaLeft, sigDeltaRight) =
-                                             splitDelta sigDelta
-                                           (sigDeltaLeftLeft,
-                                            sigDeltaLeftRight) =
-                                             splitDelta sigDeltaLeft
-                                       in (combineDeltas sigDeltaLeftLeft
-                                           (combineDeltas sigDeltaLeftRight
-                                                          sigDeltaRight),
-                                           [], associateInit)) 
-                       (\evtOcc -> (case chooseOccurrence evtOcc of
-                                      Left leftOcc -> 
-                                        case chooseOccurrence leftOcc of
-                                          Left leftLeftOcc ->
-                                            [occLeft leftLeftOcc]
-                                          Right leftRightOcc ->
-                                            [occRight $ occLeft leftRightOcc]
-                                      Right rightOcc ->
-                                        [occRight $ occRight $ rightOcc],
-                                    associateInit))
-\end{code}
-\hrule
-\caption{{\tt associate} signal function implementation.}
-\label{figure:associate_signal_function_implementation}
-\end{figure}
-
-The {\tt switch} function is the basic combinator used to introduce reactivity.
-It is given a signal function whose output is the append of two signal vectors.
-The left side of the signal vector is passed on as the output of the reactive
-signal function, and the right side is an event carrying signal functions. When
-an event occurrence is present on the right-side event output, the signal
-function carried by this occurrence replaces the signal function constructed
-by switch. Due to space concerns, the implementation is elided, but a brief
-description will suffice. The signal function stores the input sample
-provided during initialization, and updates it with deltas. When an event
-occurrence carrying a signal function is produced by either the event or 
-time function of the wrapped signal function, the stored signal sample is used
-to initialize the new signal function. If this occurs while handling an event
-input, the sample output by the new signal function's initialization is stored
-and is combined with its first output delta to produce the output delta at the
-next time step.
-
-The {\tt loop} function allows a signal function to see a component of its own
-output as input. This is primarily useful when a signal function has components
-which mutually depend on each others outputs, such as in physics simulations
-or games. Care must be taken that the feedback output is not immediately
-dependent on the feedback input, or sampling the signal function will not
-terminate. Loop is implemented by generating a recursive list (which will
-be infinite if the feedback is not decoupled) of event inputs, and by splitting
-the output signal delta and using the right side output delta as the right
-side input delta within a recursive let-binding. (The let-binding in Haskell
-always admits recursive bindings.) If the system is not completely decoupled,
-this will result in non-termination during an evaluation step. Decoupling can
-be achieved using {\tt delay} primitive.
-
-\subsection{Evaluation Interface Implementation}
-\label{subsection:Evaluation_Interface_Implementation}
-
-The evaluation interface must maintain state including the current signal
-function, the current time (to produce time deltas), updates to the input
-sample which have yet to be sampled, and the output handlers.
-
-The output handlers are stored in a structure similar to that for a signal
-sample. The difference is that both signal and event leaves contain values, and
-these values are functions from the leaf type to another type. The handler
-datatype is shown in Fig.~\ref{figure:handler_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SVHandler out sv where
-  SVHandlerEmpty  :: SVHandler out SVEmpty
-  SVHandlerSignal :: (a -> out) -> SVHandler out (SVSignal a)
-  SVHandlerEvent  :: (a -> out) -> SVHandler out (SVEvent a)
-  SVHandlerBoth   :: SVHandler out svLeft -> SVHandler out svRight 
-                     -> SVHandler out (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Handler datatype.}
-\label{figure:handler_datatype}
-\end{figure}
-
-We now need a datatype to hold the various components of the evaluation state.
-This type is shown in Fig~\ref{figure:evaluation_state_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SFEvalState m svIn svOut 
-  = SFEvalState { 
-                  esSF :: SF Initialized svIn svOut,
-                  esOutputHandlers :: SVHandler (m ()) svOut,
-                  esLastTime :: Double,
-                  esDelta :: SVDelta svIn
-                }
-
-\end{code}
-\hrule
-\caption{Evaluation state datatype.}
-\label{figure:evaluation_state_datatype}
-\end{figure}
-
-The evaluation interface itself is a monad transformer, which we implement
-as a Haskell newtype wrapping the {\tt StateT} monad transformer. We do not
-export the raw {\tt put} and {\tt get} actions of the state monad, but instead
-implement the push, update, and sampling operations for signal function
-evaluation using {\tt put} and {\tt get}. The evaluation monad transformer is
-shown in Fig.~\ref{figure:signal_function_evaluation_monad_transformer}.
-Instances of typeclasses including {\tt Monad} and {\tt MonadTrans} are derived
-using the ``GeneralizedNewtypeDeriving'' extension to the Glasgow Haskell
-Compiler.
-
-\begin{figure}[t]
-\begin{code}
-newtype SFEvalT svIn svOut m a = StateT (SFEvalState m svIn svOut) m a
-\end{code}
-\hrule
-\caption{Signal function evaluation monad transformer.}
-\label{figure:signal_function_evaluation_monad_transformer}
-\end{figure}
-
 \section{Discussion}
@@
 basis for general-purpose application frameworks.
-
-\section{Related Work}
-\label{section:Related_Work}
-
-Signal Function FRP was introduced as a model for Graphical User Interfaces~\cite{Courtney2001-1}.
-The system was originally termed ``AFRP'' (Arrowized FRP). Yampa is a rewrite of
-AFRP where signal functions apply a number of ad-hoc optimizatons to themselves
-as they evolve. Yampa demonstrated a modest performance improvement
-over AFRP~\cite{Nilsson2005}.
-
-Reactive is a classic FRP system which implements push-based evaluation for events
-by transforming behaviors to ``reactive~normal~form,'' where a behavior
-is a non-reactive behavior running inside a switch, whose event stream carries
-behaviors in reactive normal form. The system is evaluated by forking a Haskell
-thread to repeatedly sample the non-reactive behavior, and then blocking on the
-evaluation of the first occurrence in the event stream. When this occurrence
-is yielded, the evaluation thread for the behavior is killed and a new
-thread forked to evaluate the new behavior~\cite{Elliott2009}.
-
-\section{Conclusion}
-\label{section:Conclusion}
-
-We have presented TimeFlies, a system for push-pull signal-function Functional
-Reactive Programming, and have shown how the use of a signal vectors as input
-and output types for signal functions, together with GADT-based representations
-of the inputs and outputs, permits the implementation of a push-pull system.
-
-We have also described a general and flexible monadic evaluation interface for
-TimeFlies, which permits us to interface the TimeFlies system with different
-styles of IO systems, including multiple IO systems in the same application.
-
-This opens up the exciting possibility that a signal-function FRP could become
-an efficient and general framework for writing interactive applications.
-
-\bibliographystyle{splncs}
-\bibliography{thesis}
-
-\end{document}

### 2012-11-27 Docs/IFL2012/implementation.tex
Break IFL paper into multiple files.

--- 2012-11-27 Docs/IFL2012/paper.tex
@@
-\documentclass[draft]{llncs}
-
-\usepackage{fancyvrb}                                       % For code environment
-\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small} %
-
-\title{Push-Pull Signal-Function Functional Reactive Programming}
-\titlerunning{Push-Pull Signal-Function FRP}
-
-\author{Edward Amsden}
-\institute{Rochester Institute of Technology\\\email{eca7215@cs.rit.edu}}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-Functional Reactive Programming is a promising class of systems for writing
-interactive and time-dependent programs. Signal-function FRP is a subclass of
-these systems which provides advantages of modularity and correctness, but
-has proven difficult to efficiently implement.
-
-The abstraction of signal vectors provides the necessary type apparatus to
-distinguish components of the input and output of signal functions which benefit
-from a push-based implementation from those which benefit from a pull-based
-implementation, and to combine both implementation strategies in a single system.
-
-We describe a signal-function FRP system which provides push-based evaluation
-for events, pull-based evaluation for signals, and a simple monadic evaluation
-interface which permits the system to be easily integrated with one or more
-IO systems.
-\end{abstract}
-
-\section{Introduction}
-\label{section:Introduction}
-
-Functional Reactive Programming (FRP) is a class of systems for describing
-reactive programs. Reactive programs are programs which, rather than taking
-a single input and producing a single output, must accept multiple inputs and 
-alter temporal behavior, including the production of multiple outputs, based
-on these inputs.
-
-An FRP system will provide a means of manipulating {\em behaviors} and
-{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
-but the definition is the same. A behavior or signal is, semantically, a
-function from time to a value. An event is a discrete, possibly infinite, and 
-time-ordered sequence of occurrences, which are times paired with values.
-
-FRP systems can generally be categorized as ``classic FRP,'' 
-which corresponds to the originally described FRP system in that behaviors
-and events are manipulated directly and are first-class values in the FRP
-system, or ``signal-function FRP,'' in which behaviors (generally termed
-signals in this approach) and events are not first-class values, but signal
-functions are first class values. Signal functions are time-dependent and
-reactive transformers of signals, events, or combinations of signals and events.
-
-FRP combines behaviors and events through the use of {\em switching}, in which
-a behavior (in classic FRP) or a signal function (in signal-function FRP) is
-replaced by a new behavior or signal function carried by an event occurrence.
-
-Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
-Recent work on classic FRP has focused on efficient implementation. One approach to
-efficiency is to separate the evaluation of behaviors and events, since suitable 
-strategies give best performance in each case. Push-based evaluation evaluates a
-system only when input is available, and is thus suitable for discrete inputs
-such as events. Pull-based evaluation evaluates the system as quickly as
-possible, polling for input, and is preferable for behaviors and signals.
-The initial implementations of FRP made use of pull-based evaluation for both
-behaviors and events. Reactive~\cite{Elliott2009}, as well as more recent systems such as
-``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
-events and pull-based evaluation for behaviors. This is known as ``push-pull''
-evaluation.
-
-All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
-have used pull-based evaluation for both signals and events. This is due to
-the ease of implementation of pull-based evaluation, and the types used for
-signal functions which do not permit distinguishing signals and events, or
-constructing only part of the input (for instance, one event occurrence.)
-
-A recent extension of signal-function FRP called N-Ary FRP~\cite{Sculthorpe2011}
-describes a method of typing signal functions which, as we will show, enables
-the push-based evaluation of events in a signal-function FRP system. The notion
-of signal vectors allows the representation of signal function inputs and
-outputs as combinations of signals and events, rather than a single signal which
-may contain multiple values, including option values for events. Signal vectors
-are uninhabited types, which can be used to type partial or full representations
-of the signal function inputs and outputs.
-
-We present TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
-favorite quotation of one of the author's philosophy instructors, used to
-demonstrate the ambiguity of language. The origin of the quotation is unknown.}
-a push-pull signal-function FRP system. We hope to demonstrate the feasibility
-of such an approach to FRP, and provide a basis for further research into
-efficient implementation of signal-function FRP. We also describe a powerful
-evaluation interface for TimeFlies, which permits us to use TimeFlies to
-describe applications which make use of multiple and differing IO libraries.
-
-Section~\ref{section:System_Design} describes design choices for the system,
-and provides an overview of the interface. Section~\ref{section:Implementation}
-describes how the system is implemented, and how the separation of evaluation
-between events and signals is achieved. Section~\ref{section:Discussion} is a
-discussion of the usefulness of our implementation. 
-Section~\ref{section:Ongoing_and_Further_Work} describes the current and future
-work on this system. Section~\ref{section:Related_Work} gives an overview of
-related efforts. Section~\ref{section:Conclusion} concludes.
-
-\section{System Design and Interface}
-\label{section:System_Design_and_Interface}
-
-The goal of FRP is to provide an efficient, declarative abstraction for creating
-reactive programs. Towards this overall goal, there are three goals which this
-system is intended to meet.
-
-Efficient evaluation is the motivation for push-based evaluation of events.
-Since FRP programs are expected to  interact with an outside world in real time,
-efficiency cannot simply be measured by the runtime of a program. Thus, when speaking of efficiency,
-we are expressing a desire that the system utilize as few system resources as possible
-for the task at hand, while responding as quickly as possible to external inputs and
-producing output at a consistently high sample rate.
-
-A composable abstraction is one in which values in that abstraction may be
-combined in such a way that reasoning about their combined actions involves
-little more than reasoning about their individual actions. In a signal function
-system, the only interaction between composed signal functions ought to be that
-the output of one is the input of another. Composability permits a particularly
-attractive form of software engineering in which successively larger systems are
-created from by combining smaller systems, without having to reason about the 
-implementation of the components of the systems being combined.
-
-It is fine for a system to be composable with regards to itself, but an FRP
-system must interact with the outside world. Since we cannot anticipate every
-possible form of input and output that the system will be asked to interact
-with, we must interface with Haskell's IO system. In particular, most libraries
-for user interaction (e.g. GUI and graphics libraries such as GTK+ and GLUT) and
-most libraries for time-dependent IO (e.g. audio and video systems) make use of
-the event loop abstraction. In this abstraction, event handlers are registered
-with the system, and then a command is issued to run a loop which detects events
-and runs the handlers, and uses the results of the handlers to render the
-appropriate output.
-
-We would like for the FRP system to be easy to integrate with such IO systems,
-while being flexible enough to enable its use with other forms of IO systems,
-such as simple imperative systems, threaded systems, or network servers.
-
-\subsection{Semantics}
-\label{subsection:System_Design_and_Interface-Semantics}
-
-A rigorous and formal elucidation of the semantics of signal-function FRP remains
-unattempted, but there is a sufficient change to the practical semantics of
-signal-function FRP between previous signal-function systems and TimeFlies to warrant
-some description.
-
-In previous systems such as Yampa, events were understood (and implemented) as option-valued
-signals. This approach is undesirable for several reasons. The most pressing reason is
-that it prohibits push-based evaluation of events, because events are embedded in the
-samples of a signal and must be searched for when sampling.
-
-Another concern is that this approach limits the event rate to the sampling rate.
-The rate of sampling should, at some level, not matter to the FRP system. Events
-which occur between sampling intervals are never observed by the system. Since
-the signal function is never evaluated except at the sampling intervals, no
-events can otherwise be observed.
-
-This concern drives the next concern. Events are not instantaneous in this formulation.
-If a signal is option valued, the sampling instant must fall within the interval where
-there is an event occurrence present for that event to be observed. If events are
-instantaneous, the probability of observing an event occurrence is zero.
-
-Therefore, TimeFlies employs the N-Ary FRP type formulation to represent signals and
-events as distinct entities in the inputs and outputs of signal functions. This means
-we are now free to choose our representation of events, and to separate it from the
-representation and evaluation of signals.
-
-This freedom yields the additional ability to make events independent of the sampling
-interval altogether. The semantics of event handling in TimeFlies is that an event occurrence
-is responded to immediately, and does not wait for the next sampling instant. This allows events
-to be instantaneous, and further, allows multiple events to occur within a single sampling interval.
-
-There are two tradeoffs intrinsic to this approach. The first is that events are only partially ordered
-temporally. There is no way to guarantee the order of observation of event occurrences occurring in the
-same sampling interval. Further, the precise time of an event occurrence cannot be observed, only the 
-time of the last sample prior to the occurrence.
-
-In return for giving up total ordering and precise observation of the timing of events, we obtain the
-ability to employ push-based evaluation for event occurrences, and the ability to non-deterministically
-merge event occurrences. When events being input to a non-deterministic merge have simultaneous occurrences,
-we arbitrarily select one to occur first. This does not violate any guarantee about time values, since
-they will both have the same time value in either case, and does not violate any guarantee about ordering,
-since no guarantee of their order is given.
-
-A formal semantic description of signal function FRP would clarify the consequences of this decision somewhat,
-but is outside the scope of this thesis.
-
-\subsection{Types}
-\label{subsection:System_Design_and_Interface-Types}
-
-In a strongly and statically typed functional language, types are a key part of
-an interface. Types provide a mechanism for describing and ensuring properties
-of the interface's components and about the systems created with these
-components.
-
-In order to type signal functions, we must be able to describe their input and
-output. In most signal function systems, a signal function takes exactly one
-input and produces exactly one output. Multiple inputs or outputs are handled
-by making the input or output a tuple, and combinators which combine or split
-the inputs or outputs of a signal assume this. Events are represented at the
-type level as a particular type of signal, and at the value level as an option,
-either an event occurrence or not.
-
-This method of typing excludes push-based evaluation at the outset.
-It is not possible to construct a ``partial tuple'' nor in general is it
-possible to construct only part of any type of value. Push-based evaluation
-depends on evaluating only that part of the system which is updated, which means
-evaluating only that part of the input which is updated.
-
-In order to permit the construction of partial inputs and outputs, we make use
-of signal vectors. Signal vectors are uninhabited types which describe the input
-and output of a signal function. Singleton vectors are parameterized over the
-type carried by the signal or by event occurrences. The definition of the signal
-vector type is shown in Figure~\ref{figure:signal_vector_types}.
-
-Having an uninhabited signal vector type allows us to construct representations
-of inputs and outputs which are hidden from the user of the system, and are
-designed for partial representations.
-
-\begin{figure}
-\begin{code}
-data SVEmpty    -- An empty signal vector component,
-                -- neither event nor signal
-data SVSignal a -- A signal, carrying values of type a
-data SVEvent a  -- An event, whose occurrences carry values of type a
-data SVAppend svLeft svRight -- The combination of the signal vectors
-                             -- svLeft and svRight
-\end{code}
-\hrule
-\caption{Signal vector types.}
-\label{figure:signal_vector_types}
-\end{figure}
-
-The type constructor for signal functions is shown in
-Figure~\ref{figure:signal_function_types}. For the {\tt init} parameter, only
-one possible instantiation is shown. The usefulness of this type parameter,
-along with another instantation which is hidden from users of the library,
-is discussed in the section on implementation of signal functions
-(Section~\ref{section:Implementation}).
-
-Signal functions with signal vectors as input and output types form a
-Haskell {\tt GArrow}~\cite{Megacz2011}. Specifically, the signal function
-type constructor (with the initialization parameter fixed) forms the arrow
-type, the {\tt SVAppend} type constructor forms the product type, and the
-{\tt SVEmpty} type constructor forms the unit type.
-
-The representation of signal functions is discussed in
-Section~\ref{section:Implementation-Signal_Functions}. The type synonyms
-{\tt :\textasciitilde>} and {\tt :\textasciicircum:} are included for
-readability and are not crucial to the FRP system.
-
-\begin{figure}
-\begin{code}
--- Signal functions
--- init: The initialization type for 
---   the signal function, always NonInitialized
---   for exported signal functions
--- svIn: The input signal vector
--- svOut: The output signal vector
-data SF init svIn svOut
-
-data NonInitialized
-
-type svIn :~> svOut = SF NonInitialized svIn svOut
-type svLeft :^: svRight = SVAppend svLeft svRight
-\end{code}
-\hrule
-\caption{Signal function types.}
-\label{figure:signal_function_types}
-\end{figure}
-
-Monad transformers~\cite{Jones1995} provide a means to combine the functionality
-of multiple monads. A monad transformer is a monad with an extra type parameter.
-This type parameter is instantiated with the type constructor of the underlying
-monad, and an extra operation {\tt lift} is provided which converts values in
-the underlying monad to values in the monad transformer.
-
-The evaluation monad is a monad transformer. This permits it to be used in
-conjunction with Haskell's {\tt IO} monad (or any other monad) to describe how
-input is to be obtained for the signal function being evaluated, and how outputs
-are to be handled.
-
-The evaluation monad, in addition to the standard monad operators, provides a
-means of {\em initializing} a signal function, and a means of translating the
-monadic value describing evaluation to a value in the underlying monad. This
-means, for instance, that we can obtain an action in the {\tt IO} monad to
-evaluate a signal function.
-
-The type of the evaluation monad must track the input and output type of the
-signal function. The monad's context stores a mapping from outputs to handling
-actions. There are thus four type parameters to the monad's type
-constructor: the input signal vector, the output signal vector, the type of the
-underlying monad, and the monadic type parameter. The type is shown in
-Figure~\ref{figure:evaluation_monad_types}.
-
-\begin{figure}
-\begin{code}
--- A signal function's evaluations state
-data SFEvalState svIn svOut m
--- The evaluation monad
-data SFEvalT svIn svOut m a
-
-instance Monad m => Monad (SFEvalT svIn svOut m)
-\end{code}
-\hrule
-\caption{Evaluation monad types.}
-\label{figure:evaluation_monad_types}
-\end{figure}
-
-\subsection{Combinators}
-\label{section:System_Design_and_Interface-Combinators}
-
-Signal functions are constructed from combinators, which are primitive signal
-functions and operations to combine these primitives. These combinators are
-grouped as basic signal functions, lifting operations for pure functions,
-routing, reactivity, feedback, event processing, joining, and time dependence.
-
-The basic signal functions (Figure~\ref{figure:basic_signal_functions})
-provide very simple operations. The {\tt identity} signal function, as expected,
-simply copies its input to its output. The {\tt constant} signal function
-produces the provided value as a signal at all times. The {\tt never} signal
-function has an event output which never produces occurrences. The {\tt asap}
-function produces an event occurrence with the given value at the first time
-step after it is switched into the network. The {\tt after} function waits for
-the specified amount of time before producing an event occurrence.
-
-With the exception of {\tt identity}, all of the basic signal functions have
-empty inputs. This allows these signal functions to be used to insert values
-into the network which are known when the signal function is created, without
-having to route those values from an input.
-
-\begin{figure}
-\begin{code}
--- Pass the input unmodified to the output
-identity :: sv :~> sv
-
--- Produce a signal which is at all times the supplied value
-constant :: a -> SVEmpty :~> SVSignal a
-
--- An event with no occurrences
-never    :: SVEmpty :~> SVEvent a
-
--- An event with one occurrence, as soon as possible after
--- the signal function is initialized
-asap     :: a -> SVEmpty :~> SVEvent a
-
--- An event with one occurrence
--- after the specified amount of time has elapsed.
-after    :: Double -> a -> SVEmpty :~> SVEvent a
-\end{code}
-\hrule
-\caption{Basic signal function combinators.}
-\label{figure:basic_signal_functions}
-\end{figure}
-
-Two combinators are provided to lift pure functions to signal functions
-(Figure~\ref{figure:lifting_pure_functions}). The {\tt pureSignalTransformer}
-combinator applies the pure function to a signal at every sample point. The
-{\tt pureEventTransformer} combinator applies the function to each occurrence of
-an input event.
-
-\begin{figure}
-\begin{code}
--- Apply the given function to a signal at all points in time
-pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal b
-
--- Apply the given function to each event occurrence
-pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent b
-\end{code}
-\hrule
-\caption{Lifting pure functions.}
-\label{figure:lifting_pure_functions}
-\end{figure}
-
-The routing combinators are used to combine signal functions, and
-to re-arrange signal vectors in order to connect signal functions.
-The routing combinators are shown in Figure~\ref{figure:routing_combinators}.
-
-Only those combinators which modify or combine signal functions
-({\tt (>>>)}, {\tt first}, {\tt second}) are reactive (may replace themselves in
-response to events, and then only if they inherit their reactivity from the
-signal function(s) which are inputs to combinator. The rest do not
-react to or modify the input in any way, except to re-arrange it, copy it, or
-discard it altogether.
-
-\begin{figure}
-\begin{code}
--- Use the output of one signal function as the input for another
-(>>>) :: (svIn :~> svBetween) -> (svBetween :~> svOut) -> svIn :~> svOut
-
--- Pass through the right side of the input unchanged
-first :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
-
--- Swap the left and right sides
-swap :: (svLeft :^: svRight) :~> (svRight :^: svLeft)
-
--- Duplicate the input
-copy :: sv :~> (sv :^: sv)
-
--- Ignore the input
-ignore :: sv :~> svEmpty
-
--- Remove an empty vector on the left
-cancelLeft :: (SVEmpty :^: sv) :~> sv
-
--- Add an empty vector on the left
-uncancelLeft :: sv :~> (SVEmpty :^: sv)
-
--- Make right-associative
-associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))
-\end{code}
-\hrule
-\caption{Routing combinators.}
-\label{figure:routing_combinators}
-\end{figure}
-
-Reactivity is introduced by means of the {\tt switch} combinator
-(Figure~\ref{figure:switch_combinator}). The design of this combinator
-allows modular packaging of reactivity. A signal function can determine
-autonomously when to replace itself, based only on its input and state,
-by emitting an event occurrence carrying its replacement. The combinator
-consumes and hides the event carrying the replacement signal function, 
-so the reactivity is not exposed by the resulting reactive signal function.
-
-\begin{figure}
-\begin{code}
-switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
-          -> svIn :~> svOut
-\end{code}
-\hrule
-\caption{Reactivity combinator.}
-\label{figure:switch_combinator}
-\end{figure}
-
-It is often useful for a signal function to receive a portion of its
-own output as input. This is especially useful when we have two
-signal functions which we would like to mutually interact. We cannot
-just serially compose them, we must also bring the output of the second
-back around to the first. Many signal-processing algorithms also depend
-on feedback. The combinator which provides this ability is shown
-in Figure~\ref{figure:feedback_combinator}.
-
-\begin{figure}
-\begin{code}
-loop ::    ((svIn :^: svLoop) :~> (svOut :^: svLoop))
-        -> svIn :~> svOut
-\end{code}
-\hrule
-\caption{Feedback combinator.}
-\label{figure:feedback_combinator}
-\end{figure}
-
-This combinator provides decoupling for signals
-(the input signal is the output signal at the previous time-step)
-but not events (event occurrences are supplied to the combinator immediately).
-This means that the programmer has the responsibility to ensure that feedback
-does not generate an infinite sequence of events in a single time-step.
-
-Several combinators are provided for manipulating, suppressing, and generating events.
-Each of the combinators has an option variant and a list variant. The option variant
-produces an output event occurrence whenever the application of the supplied function
-to the input event produces a value. The list version produces an event occurrence for
-each of the elements of the output list. The combinators are shown in Figure~\ref{figure:event_specific_combinators}.
-
-\begin{figure}
-\begin{code}
--- Apply the function to each input occurrence,
--- and produce an occurrence for each Just.
-filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-
--- Apply the function to the stored accumulator
--- and the event occurrence, replacing the accumulator
--- and possibly outputting an occurrence
-accumulate ::    (a -> b -> (Maybe c, a))
-              -> a
-              -> SVEvent b :~> SVEvent c
-\end{code}
-\hrule
-\caption{Event-specific combinators.}
-\label{figure:event_specific_combinators}
-\end{figure}
-
-The filter combinators are stateless, and thus apply the function to only the new
-input value. They are useful for suppressing events, as well as for extracting one
-of multiple cases of a datatype. For instance, a splitter for events carrying
-{\tt Either}-valued occurrences could be written as:
-
-\begin{code}
-getLeft :: Either a b -> Maybe a
-getLeft (Left x) = Just x
-getLeft _ = Nothing
-
-getRight :: Either a b -> Maybe b
-getRight (Right x) = Just x
-getRight _ = Nothing
-
-split :: SVEvent (Either a b) :~> (SVEvent a :^: SVEvent b)
-split = copy >>> first (filter getLeft) >>> second (filter getRight)
-\end{code}
-
-The accumulate combinators are stateful, applying the supplied function
-to both the input value and an accumulator. This function has two results:
-the option or list of output event occurrence values, and the new value
-for the accumulator.
-
-The accumulator is useful when responses to multiple event occurrences
-(from one or more sources) must be coordinated. For instance, in the
-benchmark application (see Chapter~\ref{chapter:Evaluation_and_Comparisons}),
-a table is maintained that allows knowledge from previous event occurrences
-(packets from a network switch) to be used in deciding where the present
-packet ought to go.
-
-The joining combinators provide the ability to combine two event
-streams, two signals, or a signal and an event stream. These
-combinators are shown in Figure~\ref{figure:joining_combinators}
-
-The {\tt union} combinator is a non-deterministic merge of event
-streams. Any event which occurs on either input will occur
-on the output. For simultaneous event occurrences, the order of occurrence
-is not guaranteed, but the occurrence itself is. This construct
-is also guaranteed to respect the relationship of event occurrences to sampling
-intervals.
-
-The {\tt combineSignals} combinator applies a binary function pointwise to two
-signals, and produces the result of this application as a third signal. The
-combining function is necessary because we will have two input samples at each
-time step, and must produce exactly one output sample.
-
-The {\tt capture} combinator adds the last-sampled value of a signal at the time
-of an event occurrence to that event occurrence.
-
-These three combinators together provide the ability to combine elements of
-signal vectors. By combining these combinators, along with the {\tt cancelLeft}
-and {\tt cancelRight} routing combinators, arbitrary signal vectors can be
-reduced.
-
-\begin{figure}
-\begin{code}
-union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
-combineSignals :: (a -> b -> c) -> (SVSignal a :^: SVSignal b) :~> SVSignal c
-capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent (b, a)
-\end{code}
-\hrule
-\caption{Joining combinators.}
-\label{figure:joining_combinators}
-\end{figure}
-
-A set of combinators are provided for making use of time-dependence in a signal
-function. These combinators allow the modification of signals and events with
-respect to time, and the observation of the current time value. The combinators
-are shown in Figure~\ref{figure:time_combinators}
-
-\begin{figure}
-\begin{code}
-time      :: SVEmpty :~> SVSignal Double
-delay     :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
-integrate :: TimeIntegrate i => SVSignal i :~> SVSignal i
-\end{code}
-\hrule
-\caption{Time-dependent combinators.}
-\label{figure:time_combinators}
-\end{figure}
-
-The simplest time-dependent combinator is {\tt time}, which simply outputs
-the time since it began to be evaluated. This does not necessarily correspond to
-the time since the global signal function began to be evaluated, since the
-signal function in which the {\tt time} combinator is used may have been
-introduced through {\tt switch}.
-
-The {\tt delay} signal function allows events to be delayed. An initial delay
-time is given, and event occurrences on the right input can carry a new delay
-time. Event occurrences on the left input are stored and output when their delay
-time has passed. Changing the delay time does not affect occurrences already
-waiting.
-
-The {\tt integrate} combinator outputs the rectangle-rule approximation of the
-integral of its input signal with respect to time.
-
-\subsection{Evaluation Interface}
-\label{subsection:System_Design_and_Interface-Evaluation_Interface}
-
-The evaluation interface provides a modified state monad which holds a signal
-function, together with some additional information, as its state (shown in Figure~\ref{figure:evaluation_state}.
-Rather than monadic instructions to put and get the state, the monad provides instructions
-to trigger an input event, update an input signal, and trigger sampling of
-signals in the signal function. Additional state includes the current set of
-modifications to the input signals (since the last sample) and a set of
-handlers which actuate effects based on output events or changes to the output
-signal.
-
-\begin{figure}
-\begin{code}
--- A vector of handlers for outputs
-data SVHandler out sv
-
--- The state maintained when evaluating a signal function
-data SFEvalState m svIn svOut
-
--- Create the initial state for evaluating a signal function
-initSFEval ::    SVHandler (m ()) svOut   -- Output handlers
-
-              -> SVSample svIn            -- Initial input samples
-
-              -> Double                   -- Initial external time,
-                                          -- corresponding to time 0 for
-                                          -- the signal function
-
-              -> (svIn :~> svOut)         -- Signal function to evaluate
-              -> SFEvalState m svIn svOut
-\end{code}
-\hrule
-\caption{State maintained when evaluating a signal function.}
-\label{figure:evaluation_state}
-\end{figure}
-
-Sets which correspond to signal vectors are built with ``smart'' constructors.
-For instance, to construct a set of handlers, individual handling functions are
-lifted to handler sets with the {\tt signalHandler} and {\tt eventHandler}
-functions, and then combined with each other and {\tt emptyHandler} leaves
-using the {\tt combineHandlers} function.
-
-Building the initial input sample is similar, but {\tt sampleEvt} leaves do
-not carry a value.
-
-In order to initialize the state, the user must supply a set of handlers, the
-signal function to evaluate, and initial values for all of the signal inputs
-(Figure~\ref{figure:initial_input}).
-
-This state can then be passed to a monadic action which will supply input to
-the signal function. Inputs are constructed using a simple interface with
-functions to construct sample updates and event occurrences, and to specify
-their place in the vector (Figure~\ref{figure:ongoing_input}).
-
-\begin{figure}
-\begin{code}
--- Class to overload left and right functions
-class SVRoutable r where
-  svLeft          :: r svLeft -> r (svLeft :^: svRight)
-  svRight         :: r svRight -> r (svLeft :^: svRight)
-
--- An input event occurrence
-data SVEventInput sv
-instance SVRoutable SVEventInput sv
-
--- An updated sample for a signal
-data SVSignalUpdate sv
-instance SVRoutable SVSignalUpdate sv
-
--- Create an event occurrence
-svOcc           :: a -> SVEventInput (SVEvent a)
-
--- Create an updated sample
-svSig           :: a -> SVSignalUpdate (SVSignal a)
-\end{code}
-\hrule
-\caption{Data types for ongoing input.}
-\label{figure:ongoing_input}
-\end{figure}
-
-The {\tt SFEvalT} monad is actually a monad transformer, that is, it is
-parameterized over an underlying monad whose actions may be lifted to
-{\tt SFEvalT}. In the usual case, this will be the {\tt IO} monad.
-
-{\tt SFEvalT} actions are constructed using combinators to push events,
-update inputs, and step time, as well as actions lifted from the underlying
-monad (used to obtain these inputs). An action in the underlying monad
-which produces the result and a new state is obtained with the {\tt runSFEvalT}
-function. These combinators are shown in Figure~\ref{figure:evaluation_combinators}.
-
-\begin{figure}
-\begin{code}
--- The evaluation monad
-data SFEvalT svIn svOut m a
--- Appropriate instances for Monad, MonadTrans, MonadIO,
--- Functor, Applicative, etc
-
--- Obtain an action in the underlying monad
--- from an SFEvalT and a new state.
-runSFEvalT ::    SFEvalT svIn svOut m a
-              -> SFEvalState m svIn svOut
-              -> m (a, SFEvalState m svIn svOut)
-
--- Push an event occurrence.
-push :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
-
--- Update the value of an input signal sample
--- (not immediately observed)
-update :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
-
--- Step forward in time, observing the updated signal values
-step :: (Monad m) => Double -> SFEvalT svIn svOut m ()
-\end{code}
-\hrule
-\caption{Evaluation combinators}
-\label{figure:evaluation_combinators}
-\end{figure}
-
 \section{Implementation}
@@
 signal function in the state.
-
-\section{Discussion}
-\label{section:Discussion}
-
-The system presented here, TimeFlies, demonstrates how using signal vectors
-to type inputs and outputs enables push-based evaluation of events in a
-signal-function system. We take advantage of this representation in several ways.
-
-First, by separating components of inputs and outputs in the types, we are free
-to create distinct, and often partial, representations of the input or output
-of a signal function. This enables us to represent only the event occurrence
-being pushed at that time.
-
-Second, this separation also permits us to separate the process of gathering
-the input to a signal function, and the process of handling its output, into
-different points in a program. Using the evaluation interface described, an
-event occurrence may be pushed onto one input of a signal function from one
-point in a program (e.g. a mouse click handler), an input signal
-may be updated in another (e.g. a mouse movement handler), and finally the
-system may be sampled in a third place (e.g. an animation or audio timed
-callback).
-
-Finally, this approach enables further work on the implementation of the signal
-function system to be separated from changes in the interface. By enabling
-differing representations of the inputs and outputs of signal functions, we are
-free to change these representations without the need to further constrain the
-input and output types.
-
-\section{Ongoing and Further Work}
-\label{section:Ongoing_and_Further_Work}
-
-TimeFlies, the system described here, has been implemented, but not
-extensively tested. The immediate goal is to create a real-time application
-which will permit a performance and implementation comparison of TimeFlies with
-Yampa, the current state-of-the-art pull-based signal-function system.
-
-In the future, we hope to apply run-time optimizations, using the technique used
-for Yampa, to create a push-pull self-optimizing signal-function system. Further,
-we hope to use this system as a basis for exploring signal-function FRP as a
-basis for general-purpose application frameworks.
-
-\section{Related Work}
-\label{section:Related_Work}
-
-Signal Function FRP was introduced as a model for Graphical User Interfaces~\cite{Courtney2001-1}.
-The system was originally termed ``AFRP'' (Arrowized FRP). Yampa is a rewrite of
-AFRP where signal functions apply a number of ad-hoc optimizatons to themselves
-as they evolve. Yampa demonstrated a modest performance improvement
-over AFRP~\cite{Nilsson2005}.
-
-Reactive is a classic FRP system which implements push-based evaluation for events
-by transforming behaviors to ``reactive~normal~form,'' where a behavior
-is a non-reactive behavior running inside a switch, whose event stream carries
-behaviors in reactive normal form. The system is evaluated by forking a Haskell
-thread to repeatedly sample the non-reactive behavior, and then blocking on the
-evaluation of the first occurrence in the event stream. When this occurrence
-is yielded, the evaluation thread for the behavior is killed and a new
-thread forked to evaluate the new behavior~\cite{Elliott2009}.
-
-An alternate push-based system, based on {\em signal segments}, was presented in
-work on the Curry-Howard correspondence between a form of temporal logic (LTL)
-and FRP~\cite{Jeffrey2012}. However, this system continues to represent events
-as option-valued signals. It therefore cannot be push-based in its response to
-individual events, and cannot be push-pull, evaluating signals by pull-based
-evaluation and events by push-based evaluation. It is push-based not in the
-sense of evaluating events only when they occur, but rather in that it provides
-a means to push signal segments (samples of a signal over intervals of time) to
-the system and wait on its response.
-
-\section{Conclusion}
-\label{section:Conclusion}
-
-We have presented TimeFlies, a system for push-pull signal-function Functional
-Reactive Programming, and have shown how the use of a signal vectors as input
-and output types for signal functions, together with GADT-based representations
-of the inputs and outputs, permits the implementation of a push-pull system.
-
-We have also described a general and flexible monadic evaluation interface for
-TimeFlies, which permits us to interface the TimeFlies system with different
-styles of IO systems, including multiple IO systems in the same application.
-
-This opens up the exciting possibility that a signal-function FRP could become
-an efficient and general framework for writing interactive applications.
-
-\bibliographystyle{splncs}
-\bibliography{thesis}
-
-\end{document}

### 2012-11-27 Docs/IFL2012/introduction.tex
Break IFL paper into multiple files.

--- 2012-11-24 Docs/IFL2012/paper.tex
@@
-\documentclass[draft]{llncs}
-
-\usepackage{fancyvrb}                                       % For code environment
-\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small} %
-
-\title{Push-Pull Signal-Function Functional Reactive Programming}
-\titlerunning{Push-Pull Signal-Function FRP}
-
-\author{Edward Amsden}
-\institute{Rochester Institute of Technology\\\email{eca7215@cs.rit.edu}}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-Functional Reactive Programming is a promising class of systems for writing
-interactive and time-dependent programs. Signal-function FRP is a subclass of
-these systems which provides advantages of modularity and correctness, but
-has proven difficult to efficiently implement.
-
-The abstraction of signal vectors provides the necessary type apparatus to
-distinguish components of the input and output of signal functions which benefit
-from a push-based implementation from those which benefit from a pull-based
-implementation, and to combine both implementation strategies in a single system.
-
-We describe a signal-function FRP system which provides push-based evaluation
-for events, pull-based evaluation for signals, and a simple monadic evaluation
-interface which permits the system to be easily integrated with one or more
-IO systems.
-\end{abstract}
-
 \section{Introduction}
@@
 
-Section~\ref{section:System_Design} describes design choices for the system,
+Section~\ref{section:System_Design_and_Interface} describes design choices for the system,
 and provides an overview of the interface. Section~\ref{section:Implementation}
@@
 related efforts. Section~\ref{section:Conclusion} concludes.
-
-\section{System Design}
-\label{section:System_Design}
-
-Our goal is to produce a composable and efficient FRP system. Signal function
-FRP has an advantage in terms of composability, because it permits the
-construction of self-contained objects through both input and output composition,
-rather than purely through output composition as in classic FRP. Signal-function
-FRP also avoids problematic properties common to classic FRP systems such as a
-large class of time and space leaks~\cite{Liu2007}.
-
-Efficient implementations of signal-function FRP have been approached through
-runtime optimization~\cite{Nilsson2005}, but all implementations have been
-pull-based for both signals and events. A truly efficient implementation will
-likely combine run-time optimization with push-based evaluation for events.
-
-The concept of N-Ary FRP was to encode additional safety properties into the
-types of a signal-function FRP system~\cite{Sculthorpe2011}. This system
-introduced the concept of {\em signal vectors} as input and output types for
-signal functions. Signal vectors are combinations of signals and events. In
-N-Ary FRP, signal functions are represented at the type level, and type-level
-functions (type families in Haskell) are used to specify the representations
-of signal functions.
-
-In our system, we construct representations of signal vectors using
-Generalized Algebraic Datatypes~\cite{Cheney2003,Xi2003}. The use of GADTs
-enables us to use signal vectors to instantiate type parameters in the types
-of signal vector representations, as well as in the types of signal functions.
-GADTs are available as an extension in the Glasgow Haskell Compiler~\cite{PeytonJones2006}.
-
-This approach permits us to construct partial inputs and outputs for signal
-functions. For instance, we can construct a representation for a single event
-in a signal vector, and another representation of updated values for a subset of
-the signals in the signal vector, and yet another carrying values for all
-signals in a signal vector. The signal vector types are shown in
-Figure~\ref{figure:signal_vectors}.
-
-\begin{figure}[t]
-\begin{code}
-data SVEmpty
-data SVSignal a
-data SVEvent a
-data SVAppend svLeft svRight
-\end{code}
-\hrule
-\caption{Signal vectors.}
-\label{figure:signal_vectors}
-\end{figure}
-
-With the ability to construct partial representations of signal vectors, we can
-represent signal functions with a datatype carrying multiple functions, one for
-each type of input. This allows us to separate the pull-based processing of
-signals from the push-based reaction to events.
-
-The exposed interface is a set of combinators for constructing signal functions,
-as well as combinators for describing the evaluation of a signal function. The
-interface is shown in Fig.~\ref{figure:signal_function_interface}.
-
-\begin{figure}
-\begin{code}
-
--- Signal Functions
-type :~> svIn svOut
-
--- Infix type alias for SVAppend
-type :^: svLeft svRight = SVAppend svLeft svRight
-
--- Basic signal functions
-identity :: sv :~> sv
-constant :: a -> SVEmpty :~> SVSignal a
-never    :: SVEmpty :~> SVEvent a
-asap     :: a -> SVEmpty :~> SVEvent a
-after    :: Double -> a -> SVEmpty :~> SVEvent a
-
--- Lifting pure functions
-pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal a
-pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent a
-
--- Composition and routing
-(>>>)           :: (svIn :~> svMiddle) -> (svMiddle :~> svOut) 
-                   -> (svIn :~> svOut)
-first           :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
-second          :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut) 
-swap            :: (svLeft :^: svRight) :~> (svRight :^: svLeft)
-copy            :: sv :~> (sv :^: sv)
-ignore          :: sv :~> SVEmpty
-cancelLeft      :: (SVEmpty :^: sv) :~> sv
-cancelRight     :: (sv :^: SVEmpty) :~> sv
-uncancelLeft    :: sv :~> (SVEmpty :^: sv)
-uncancelRight   :: sv :~> (sv :^: SVEmpty)
-associate       :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3)) 
-unassociate     :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
-
--- Reactivity
-switch :: (svIn :~> (svOut :^: SVEvent (svIn :~> svOut))) 
-          -> svIn :~> svOut
-
--- Feedback
-loop :: ((svIn :^: svLoop) :~> (svOut :^: svLoop)) -> svIn :~> svOut
-
--- Time dependence
-time  :: SVEmpty :~> Double
-delay :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
-class TimeIntegrate
-
--- Joining
-union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
-combineSignals :: (a -> b -> c) 
-                  -> (SVSignal a :^: SVSignal b) :~> SVSignal c
-capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent a
-
--- Events
-filter         :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-\end{code}
-\hrule
-\caption{Signal function interface.}
-\label{figure:signal_function_interface}
-\end{figure}
-
-Signal functions are produced by combining primitive signal functions using
-the {\tt >>>} (sequential composition), {\tt first}, and {\tt second} routing
-combinators. Other routing signal functions are provided, but they are intended
-to be combined with, rather than to modify, other signal functions.
-
-\subsection{Evaluation Interface}
-The evaluation interface provides a monad~\cite{PeytonJones1993,PeytonJones2001}
-for specifying input to signal functions and handling their output. This
-interface also allows input actions to be separated from output actions, and
-from each other, so that a signal function may receive input from multiple IO
-systems and have its output handled by multiple output systems.
-
-The evaluation interface is a monad transformer~\cite{Liang1995}. This allows
-a signal function evaluation to be constructed from primitive evaluation actions
-(event pushing, input signal updating, and sampling) as well as actions from the
-underlying monad. These actions can then be run in the underlying monad
-to actuate the signal function. Because the actuation takes a signal function's
-state as input and produces a new state as output, in addition to the monadic
-side effects of handling the signal function's output, different evaluation
-actions may be taken on the same signal function from distinct locations within
-an application's code. This allows the evaluation interface to be easily
-integrated with event-loop style systems as well as traditional imperative
-IO systems.
-
-The evaluation interface consists of the monad transformer type, functions for
-constructing event inputs, initial input signal samples, and input signal
-updates, functions for constructing the vector of output handlers for a signal
-function, and evaluation actions. The full interface is shown in Fig.~\ref{figure:evaluation_interface}.
-
-\begin{figure}
-\begin{code}
--- Input helpers
-class SVRoutable -- Instances: SVSignalUpdate,
-                 -- SVEventOccurrence
-svLeft :: (SVRoutable r) => 
-          r svLeft -> r (SVAppend svLeft svRight)
-svRight :: (SVRoutable r) =>
-           r svRight -> r (SVAppend svLeft svRight)
-
--- Signal inputs
-data SVSignalUpdate
-data SVSample
-sample :: a -> SVSample (SVSignal a)
-sampleEvt :: SVSample (SVEvent a)
-sampleNothing :: SVSample SVEmpty
-combineSamples :: SVSample svLeft -> svSample svRight 
-                  -> svSample (svLeft :^: svRight)
-svSig :: a -> SVSignalUpdate (SVSignal a)
-
--- Event inputs
-data SVEventInput
-svOcc :: a -> SVEventInput (SVEvent a)
--- Actuation
-update :: (Monad m) => SVSignalUpdate svIn 
-           -> SFEvalT svIn svOut m ()
-push   :: (Monad m) => SVEventInput svIn
-          -> SFEvalT svIn svOut m ()
-sample :: SVEvalT svIn svOut m ()
--- Running
-data SVHandler
-emptyHandler :: SVHandler m SVEmpty
-eventHandler :: (a -> m ()) 
-                -> SVHandler m (SVEvent a)
-signalHandler :: (a -> m ()) 
-                 -> SVHandler m (SVSignal a)
-combineHandlers :: SVHandler m svLeft
-                   -> SVHandler m svRight
-                   -> SVHandler m (svLeft :^: svRight)
-initSFEvalT :: SVHandler m svOut -> SVSample svIn 
-               -> Double ->(svIn :~> svOut)
-               -> SFEvalState m svIn svOut
-runSFEvalT :: SFEvalT svIn svOut m a
-              -> SFEvalState m svIn svOut
-              -> m a
-\end{code}
-\hrule
-\caption{Evaluation interface.}
-\label{figure:evaluation_interface}
-\end{figure}
-
-\section{Implementation}
-\label{section:Implementation}
-
-We now turn our attention to the implementation of the signal function
-system. We will discuss representations of inputs, outputs, and signal functions,
-as well as the implementations of specific signal function combinators.
-
-\subsection{Input and Output Representations}
-\label{subsection:Input_and_Output_Representations}
-As discussed in Section~\ref{section:System_Design}, our implementation requires
-representations of the inputs and outputs of signal functions. Since signal
-functions are typed using signal vectors, the representation types must be
-parameterized over signal vectors. But signal vectors are uninhabited types, and
-in ordinary Haskell we cannot declare a data constructor which uses a component
-of a type parameter. Further, we cannot restrict which types may instantiate a
-type parameter for a particular data constructor.
-
-GADTs lift these restrictions, permitting type constraints to be applied
-to individual data constructors. In the declaration of a GADT, a type signature
-is provided for each data constructor, including the types of the constructor's
-parameters. When a GADT is pattern-matched, the types of the parameters are
-inferred using the constraints given in this signature. This is called type
-refinement.
-
-Using GADTs, we can construct representations of signal vectors to use in our
-implementation.
-
-When constructing a representation, we consider how to represent each
-component of a signal vector, whether it is a signal element, an empty
-element, or an append node. For instance, in order to represent a signal
-sample for all signals in an element, we construct the GADT shown in Figure~\ref{figure:signal_sample_representation}.
-
-\begin{figure}
-\begin{code}
-data SVSample sv where
-  SVSample        ::    a -> SVSample (SVSignal a)
-  SVSampleEvt     ::    SVSample (SVEvent a)
-  SVSampleNothing ::    SVSample SVEmpty
-  SVSampleBoth    ::    SVSample svLeft
-                     -> SVSample svRight
-                     -> SVSample (svLeft :^: svRight)
-\end{code}
-\hrule
-\caption{A representation which carries a value for each signal in a signal vector.}
-\label{figure:signal_sample_representation}
-\end{figure}
-
-\subsection{Signal Function Representations}
-\label{subsection:Signal_Function_Representations}
-With representations for signal function inputs and outputs in place, we turn
-to the representation of signal functions. A signal function must be able to
-respond to time increments and new signal samples, as well as event occurrences.
-Since we wish to be able to ``push'' event occurrences independently of
-sampling, events are handled using the last-updated time and signal sample.
-
-When a signal function is asked to respond to either of these types of inputs,
-it must produce an output consisting of zero or more event occurrences, a new
-signal function with the same type (to enable reactivity and state), and, if
-it is responding to a time and sample update, a delta which represents updates
-to the sample of its output signals.
-
-Finally, there is a special case for a signal function at time 0. It must be
-provided with an initial input sample before it can respond to incremental time
-and sample updates or to event occurrences.
-
-Signal functions are represented as a datatype with two constructors. The first
-constructor wraps a function from an input sample to an output sample and an
-initialized signal function. The second wraps two functions. The first accepts
-a time delta and a signal delta, and produces a signal delta, a list of event
-occurrences, and a new signal function as the output. The second accepts an
-event occurrence and produces a list of event occurrences and a new signal
-function as the output. The signal function representation is shown in Figure~\ref{figure:signal_function_representation}.
-
-\begin{figure}[t]
-\begin{code}
-data Initialized
-data NonInitialized
-
-type :~> svIn svOut = SF NonInitialized svIn svOut
-
-data SF init svIn svOut where
-  SF :: (SVSample svIn -> (SVSample svOut, SF Initialized svIn svOut))
-        -> SF NonInitialized svIn svOut
-  SFInit :: (Double -> SVDelta svIn 
-             -> (SVDelta svOut, [SVOccurrence svOut],
-                 SF Initialized svIn svOut))
-            -> (SVOccurrence svIn -> ([SVOccurrence svOut],
-                                      SF Initialized svIn svOut))
-            -> SF Initialized svIn svOut
-\end{code}
-\hrule
-\caption{Signal function representation}
-\label{figure:signal_function_representation}
-\end{figure}
-
-This representation allows a signal function to respond to events and time
-updates separately, and does not enforce the representation of events during
-time updates, as previous signal function systems do.
-
-\subsection{Signal Function Implementations}
-\label{subsection:Signal_Function_Implementations}
-
-Now that we have established representations for signal functions and their
-inputs and outputs, we show several examples of how signal function combinators
-are implemented.
-
-The {\tt identity} signal function's initialization function takes a sample,
-and returns that sample as its output sample. The initialized function returned
-is the {\tt identityInit} function, which is not exported by the module.
-The time and sample function ignores the provided time delta, produces the
-input signal delta as the output signal delta, and returns {\tt identityInit} as
-the new signal function. The event function takes an event occurrence and
-returns a singleton list containing that occurrence, along with the
-{\tt identityInit} function as the replacement signal function. The
-implementation is shown in Figure~\ref{figure:identity_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-identity :: sv :~> sv
-identity = SF (\sample -> (sample, identityInit))
-
-identityInit :: SF Initialized sv sv
-identityInit = SFInit (\_ delta -> (delta, [], identityInit))
-                      (\occ -> ([occ], identityInit))
-\end{code}
-\hrule
-\caption{{\tt identity} signal function implementation.}
-\label{figure:identity_signal_function_implementation}
-\end{figure}
-
-The {\tt constant} signal function is initialized with a value, which it then
-produces as its output forever. Thus, the initialized version of {\tt constant}
-never produces any output other than itself as a replacement
-signal function, and an empty signal delta. The implementation of {\tt constant}
-is shown in Fig.~\ref{figure:constant_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-constant :: a -> SVEmpty :~> SVSignal a
-constant x = SF (\_ -> (constantInit, sample x))
-
-constantInit :: SF Initialized SVEmpty (SVSignal a)
-constantInit = SFInit (\_ _ -> (deltaNothing, [], constantInit))
-                      (\_ -> ([], constantInit))
-\end{code}
-\hrule
-\caption{{\tt constant} signal function implementation.}
-\label{figure:constant_signal_function_implementation}
-\end{figure}
-
-The {\tt asap} signal function implementation requires a parameter for the
-initialized version of the signal function, to specify the value of the event
-occurrence. It replaces itself with the initialized version of the {\tt never}
-signal function after the first time step. The implementation is shown in
-Fig.~\ref{figure:asap_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-asap :: a -> SVEmpty :~> SVEvent a
-asap x = SF (\_ -> (sampleEvt, asapInit x))
-
-asapInit :: a -> SF Initialized SVEmpty (SVEvent a)
-asapInit x = SFInit (\_ _ -> (deltaNothing, [occurrence x], neverInit))
-                    (\_ -> (never))
-\end{code}
-\hrule
-\caption{{\tt asap} signal function implementation.}
-\label{figure:asap_signal_function_implementation}
-\end{figure}
-
-The {\tt filter} function accepts events, applies a {\tt Maybe} predicate to
-their values, and produces the value produced by the predicate as an event
-occurrence, or no occurrence if the value is {\tt Nothing}. It is implemented
-by having the event occurrence response function apply the {\tt maybe} function
-from the Haskell prelude to the value returned by the predicate, as shown in 
-Fig.~\ref{figure:filter_signal_function_implementation}
-
-\begin{figure}[t]
-\begin{code}
-filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-filter p = SF (\_ -> (sampleEvt, filterInit p))
-
-filterInit :: (a -> Maybe b) -> SF Initialized (SVEvent a) (SVEvent b)
-filterInit p = SFInit (\_ _ -> (deltaNothing, [], filterInit p))
-                      (\evtOcc -> (maybe [] ((:[]) . occurrence) $
-                                   fromOccurrence evtOcc,
-                                   filterInit p))
-\end{code}
-\hrule
-\caption{{\tt filter} signal function implementation.}
-\label{figure:filter_signal_function_implementation}
-\end{figure}
-
-The {\tt associate} signal function is a routing signal function. It transforms
-a signal vector which is left-associated at the top level to one that is
-right-associated at the top level. Its implementation is representative of all
-of the routing functions, and is shown in Fig.~\ref{figure:associate_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-associate :: SVAppend (SVAppend sv1 sv2) sv3
-             :~> SVAppend sv1 (SVAppend sv2 sv3)
-associate = 
-  SF (\sigSample -> let (sigSampleLeft, sigSampleRight) = 
-                          splitSample sigSample
-                        (sigSampleLeftLeft, sigSampleLeftRight) =
-                          splitSample sigSampleLeft
-                    in (combineSamples sigSampleLeftLeft 
-                        (combineSamples sigSampleLeftRight sigSampleRight),
-                        associateInit))
-
-associateInit :: SF Initialized (SVAppend (SVAppend sv1 sv2) sv3)
-                                (SVAppend sv1 (SVAppend sv2 sv3))
-associateInit = SFInit (\_ sigDelta -> let (sigDeltaLeft, sigDeltaRight) =
-                                             splitDelta sigDelta
-                                           (sigDeltaLeftLeft,
-                                            sigDeltaLeftRight) =
-                                             splitDelta sigDeltaLeft
-                                       in (combineDeltas sigDeltaLeftLeft
-                                           (combineDeltas sigDeltaLeftRight
-                                                          sigDeltaRight),
-                                           [], associateInit)) 
-                       (\evtOcc -> (case chooseOccurrence evtOcc of
-                                      Left leftOcc -> 
-                                        case chooseOccurrence leftOcc of
-                                          Left leftLeftOcc ->
-                                            [occLeft leftLeftOcc]
-                                          Right leftRightOcc ->
-                                            [occRight $ occLeft leftRightOcc]
-                                      Right rightOcc ->
-                                        [occRight $ occRight $ rightOcc],
-                                    associateInit))
-\end{code}
-\hrule
-\caption{{\tt associate} signal function implementation.}
-\label{figure:associate_signal_function_implementation}
-\end{figure}
-
-The {\tt switch} function is the basic combinator used to introduce reactivity.
-It is given a signal function whose output is the append of two signal vectors.
-The left side of the signal vector is passed on as the output of the reactive
-signal function, and the right side is an event carrying signal functions. When
-an event occurrence is present on the right-side event output, the signal
-function carried by this occurrence replaces the signal function constructed
-by switch. Due to space concerns, the implementation is elided, but a brief
-description will suffice. The signal function stores the input sample
-provided during initialization, and updates it with deltas. When an event
-occurrence carrying a signal function is produced by either the event or 
-time function of the wrapped signal function, the stored signal sample is used
-to initialize the new signal function. If this occurs while handling an event
-input, the sample output by the new signal function's initialization is stored
-and is combined with its first output delta to produce the output delta at the
-next time step.
-
-The {\tt loop} function allows a signal function to see a component of its own
-output as input. This is primarily useful when a signal function has components
-which mutually depend on each others outputs, such as in physics simulations
-or games. Care must be taken that the feedback output is not immediately
-dependent on the feedback input, or sampling the signal function will not
-terminate. Loop is implemented by generating a recursive list (which will
-be infinite if the feedback is not decoupled) of event inputs, and by splitting
-the output signal delta and using the right side output delta as the right
-side input delta within a recursive let-binding. (The let-binding in Haskell
-always admits recursive bindings.) If the system is not completely decoupled,
-this will result in non-termination during an evaluation step. Decoupling can
-be achieved using {\tt delay} primitive.
-
-\subsection{Evaluation Interface Implementation}
-\label{subsection:Evaluation_Interface_Implementation}
-
-The evaluation interface must maintain state including the current signal
-function, the current time (to produce time deltas), updates to the input
-sample which have yet to be sampled, and the output handlers.
-
-The output handlers are stored in a structure similar to that for a signal
-sample. The difference is that both signal and event leaves contain values, and
-these values are functions from the leaf type to another type. The handler
-datatype is shown in Fig.~\ref{figure:handler_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SVHandler out sv where
-  SVHandlerEmpty  :: SVHandler out SVEmpty
-  SVHandlerSignal :: (a -> out) -> SVHandler out (SVSignal a)
-  SVHandlerEvent  :: (a -> out) -> SVHandler out (SVEvent a)
-  SVHandlerBoth   :: SVHandler out svLeft -> SVHandler out svRight 
-                     -> SVHandler out (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Handler datatype.}
-\label{figure:handler_datatype}
-\end{figure}
-
-We now need a datatype to hold the various components of the evaluation state.
-This type is shown in Fig~\ref{figure:evaluation_state_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SFEvalState m svIn svOut 
-  = SFEvalState { 
-                  esSF :: SF Initialized svIn svOut,
-                  esOutputHandlers :: SVHandler (m ()) svOut,
-                  esLastTime :: Double,
-                  esDelta :: SVDelta svIn
-                }
-
-\end{code}
-\hrule
-\caption{Evaluation state datatype.}
-\label{figure:evaluation_state_datatype}
-\end{figure}
-
-The evaluation interface itself is a monad transformer, which we implement
-as a Haskell newtype wrapping the {\tt StateT} monad transformer. We do not
-export the raw {\tt put} and {\tt get} actions of the state monad, but instead
-implement the push, update, and sampling operations for signal function
-evaluation using {\tt put} and {\tt get}. The evaluation monad transformer is
-shown in Fig.~\ref{figure:signal_function_evaluation_monad_transformer}.
-Instances of typeclasses including {\tt Monad} and {\tt MonadTrans} are derived
-using the ``GeneralizedNewtypeDeriving'' extension to the Glasgow Haskell
-Compiler.
-
-\begin{figure}[t]
-\begin{code}
-newtype SFEvalT svIn svOut m a = StateT (SFEvalState m svIn svOut) m a
-\end{code}
-\hrule
-\caption{Signal function evaluation monad transformer.}
-\label{figure:signal_function_evaluation_monad_transformer}
-\end{figure}
-
-\section{Discussion}
-\label{section:Discussion}
-
-The system presented here, TimeFlies, demonstrates how using signal vectors
-to type inputs and outputs enables push-based evaluation of events in a
-signal-function system. We take advantage of this representation in several ways.
-
-First, by separating components of inputs and outputs in the types, we are free
-to create distinct, and often partial, representations of the input or output
-of a signal function. This enables us to represent only the event occurrence
-being pushed at that time.
-
-Second, this separation also permits us to separate the process of gathering
-the input to a signal function, and the process of handling its output, into
-different points in a program. Using the evaluation interface described, an
-event occurrence may be pushed onto one input of a signal function from one
-point in a program (e.g. a mouse click handler), an input signal
-may be updated in another (e.g. a mouse movement handler), and finally the
-system may be sampled in a third place (e.g. an animation or audio timed
-callback).
-
-Finally, this approach enables further work on the implementation of the signal
-function system to be separated from changes in the interface. By enabling
-differing representations of the inputs and outputs of signal functions, we are
-free to change these representations without the need to further constrain the
-input and output types.
-
-\section{Ongoing and Further Work}
-\label{section:Ongoing_and_Further_Work}
-
-TimeFlies, the system described here, has been implemented, but not
-extensively tested. The immediate goal is to create a real-time application
-which will permit a performance and implementation comparison of TimeFlies with
-Yampa, the current state-of-the-art pull-based signal-function system.
-
-In the future, we hope to apply run-time optimizations, using the technique used
-for Yampa, to create a push-pull self-optimizing signal-function system. Further,
-we hope to use this system as a basis for exploring signal-function FRP as a
-basis for general-purpose application frameworks.
-
-\section{Related Work}
-\label{section:Related_Work}
-
-Signal Function FRP was introduced as a model for Graphical User Interfaces~\cite{Courtney2001-1}.
-The system was originally termed ``AFRP'' (Arrowized FRP). Yampa is a rewrite of
-AFRP where signal functions apply a number of ad-hoc optimizatons to themselves
-as they evolve. Yampa demonstrated a modest performance improvement
-over AFRP~\cite{Nilsson2005}.
-
-Reactive is a classic FRP system which implements push-based evaluation for events
-by transforming behaviors to ``reactive~normal~form,'' where a behavior
-is a non-reactive behavior running inside a switch, whose event stream carries
-behaviors in reactive normal form. The system is evaluated by forking a Haskell
-thread to repeatedly sample the non-reactive behavior, and then blocking on the
-evaluation of the first occurrence in the event stream. When this occurrence
-is yielded, the evaluation thread for the behavior is killed and a new
-thread forked to evaluate the new behavior~\cite{Elliott2009}.
-
-\section{Conclusion}
-\label{section:Conclusion}
-
-We have presented TimeFlies, a system for push-pull signal-function Functional
-Reactive Programming, and have shown how the use of a signal vectors as input
-and output types for signal functions, together with GADT-based representations
-of the inputs and outputs, permits the implementation of a push-pull system.
-
-We have also described a general and flexible monadic evaluation interface for
-TimeFlies, which permits us to interface the TimeFlies system with different
-styles of IO systems, including multiple IO systems in the same application.
-
-This opens up the exciting possibility that a signal-function FRP could become
-an efficient and general framework for writing interactive applications.
-
-\bibliographystyle{splncs}
-\bibliography{thesis}
-
-\end{document}

### 2012-11-27 Docs/IFL2012/paper.tex
Break IFL paper into multiple files.

--- 2012-11-24 Docs/IFL2012/paper.tex
@@
 
-\begin{abstract}
-Functional Reactive Programming is a promising class of systems for writing
-interactive and time-dependent programs. Signal-function FRP is a subclass of
-these systems which provides advantages of modularity and correctness, but
-has proven difficult to efficiently implement.
-
-The abstraction of signal vectors provides the necessary type apparatus to
-distinguish components of the input and output of signal functions which benefit
-from a push-based implementation from those which benefit from a pull-based
-implementation, and to combine both implementation strategies in a single system.
-
-We describe a signal-function FRP system which provides push-based evaluation
-for events, pull-based evaluation for signals, and a simple monadic evaluation
-interface which permits the system to be easily integrated with one or more
-IO systems.
-\end{abstract}
-
-\section{Introduction}
-\label{section:Introduction}
-
-Functional Reactive Programming (FRP) is a class of systems for describing
-reactive programs. Reactive programs are programs which, rather than taking
-a single input and producing a single output, must accept multiple inputs and 
-alter temporal behavior, including the production of multiple outputs, based
-on these inputs.
-
-An FRP system will provide a means of manipulating {\em behaviors} and
-{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
-but the definition is the same. A behavior or signal is, semantically, a
-function from time to a value. An event is a discrete, possibly infinite, and 
-time-ordered sequence of occurrences, which are times paired with values.
-
-FRP systems can generally be categorized as ``classic FRP,'' 
-which corresponds to the originally described FRP system in that behaviors
-and events are manipulated directly and are first-class values in the FRP
-system, or ``signal-function FRP,'' in which behaviors (generally termed
-signals in this approach) and events are not first-class values, but signal
-functions are first class values. Signal functions are time-dependent and
-reactive transformers of signals, events, or combinations of signals and events.
-
-FRP combines behaviors and events through the use of {\em switching}, in which
-a behavior (in classic FRP) or a signal function (in signal-function FRP) is
-replaced by a new behavior or signal function carried by an event occurrence.
-
-Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
-Recent work on classic FRP has focused on efficient implementation. One approach to
-efficiency is to separate the evaluation of behaviors and events, since suitable 
-strategies give best performance in each case. Push-based evaluation evaluates a
-system only when input is available, and is thus suitable for discrete inputs
-such as events. Pull-based evaluation evaluates the system as quickly as
-possible, polling for input, and is preferable for behaviors and signals.
-The initial implementations of FRP made use of pull-based evaluation for both
-behaviors and events. Reactive~\cite{Elliott2009}, as well as more recent systems such as
-``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
-events and pull-based evaluation for behaviors. This is known as ``push-pull''
-evaluation.
-
-All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
-have used pull-based evaluation for both signals and events. This is due to
-the ease of implementation of pull-based evaluation, and the types used for
-signal functions which do not permit distinguishing signals and events, or
-constructing only part of the input (for instance, one event occurrence.)
-
-A recent extension of signal-function FRP called N-Ary FRP~\cite{Sculthorpe2011}
-describes a method of typing signal functions which, as we will show, enables
-the push-based evaluation of events in a signal-function FRP system. The notion
-of signal vectors allows the representation of signal function inputs and
-outputs as combinations of signals and events, rather than a single signal which
-may contain multiple values, including option values for events. Signal vectors
-are uninhabited types, which can be used to type partial or full representations
-of the signal function inputs and outputs.
-
-We present TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
-favorite quotation of one of the author's philosophy instructors, used to
-demonstrate the ambiguity of language. The origin of the quotation is unknown.}
-a push-pull signal-function FRP system. We hope to demonstrate the feasibility
-of such an approach to FRP, and provide a basis for further research into
-efficient implementation of signal-function FRP. We also describe a powerful
-evaluation interface for TimeFlies, which permits us to use TimeFlies to
-describe applications which make use of multiple and differing IO libraries.
-
-Section~\ref{section:System_Design} describes design choices for the system,
-and provides an overview of the interface. Section~\ref{section:Implementation}
-describes how the system is implemented, and how the separation of evaluation
-between events and signals is achieved. Section~\ref{section:Discussion} is a
-discussion of the usefulness of our implementation. 
-Section~\ref{section:Ongoing_and_Further_Work} describes the current and future
-work on this system. Section~\ref{section:Related_Work} gives an overview of
-related efforts. Section~\ref{section:Conclusion} concludes.
-
-\section{System Design}
-\label{section:System_Design}
-
-Our goal is to produce a composable and efficient FRP system. Signal function
-FRP has an advantage in terms of composability, because it permits the
-construction of self-contained objects through both input and output composition,
-rather than purely through output composition as in classic FRP. Signal-function
-FRP also avoids problematic properties common to classic FRP systems such as a
-large class of time and space leaks~\cite{Liu2007}.
-
-Efficient implementations of signal-function FRP have been approached through
-runtime optimization~\cite{Nilsson2005}, but all implementations have been
-pull-based for both signals and events. A truly efficient implementation will
-likely combine run-time optimization with push-based evaluation for events.
-
-The concept of N-Ary FRP was to encode additional safety properties into the
-types of a signal-function FRP system~\cite{Sculthorpe2011}. This system
-introduced the concept of {\em signal vectors} as input and output types for
-signal functions. Signal vectors are combinations of signals and events. In
-N-Ary FRP, signal functions are represented at the type level, and type-level
-functions (type families in Haskell) are used to specify the representations
-of signal functions.
-
-In our system, we construct representations of signal vectors using
-Generalized Algebraic Datatypes~\cite{Cheney2003,Xi2003}. The use of GADTs
-enables us to use signal vectors to instantiate type parameters in the types
-of signal vector representations, as well as in the types of signal functions.
-GADTs are available as an extension in the Glasgow Haskell Compiler~\cite{PeytonJones2006}.
-
-This approach permits us to construct partial inputs and outputs for signal
-functions. For instance, we can construct a representation for a single event
-in a signal vector, and another representation of updated values for a subset of
-the signals in the signal vector, and yet another carrying values for all
-signals in a signal vector. The signal vector types are shown in
-Figure~\ref{figure:signal_vectors}.
-
-\begin{figure}[t]
-\begin{code}
-data SVEmpty
-data SVSignal a
-data SVEvent a
-data SVAppend svLeft svRight
-\end{code}
-\hrule
-\caption{Signal vectors.}
-\label{figure:signal_vectors}
-\end{figure}
-
-With the ability to construct partial representations of signal vectors, we can
-represent signal functions with a datatype carrying multiple functions, one for
-each type of input. This allows us to separate the pull-based processing of
-signals from the push-based reaction to events.
-
-The exposed interface is a set of combinators for constructing signal functions,
-as well as combinators for describing the evaluation of a signal function. The
-interface is shown in Fig.~\ref{figure:signal_function_interface}.
-
-\begin{figure}
-\begin{code}
-
--- Signal Functions
-type :~> svIn svOut
-
--- Infix type alias for SVAppend
-type :^: svLeft svRight = SVAppend svLeft svRight
-
--- Basic signal functions
-identity :: sv :~> sv
-constant :: a -> SVEmpty :~> SVSignal a
-never    :: SVEmpty :~> SVEvent a
-asap     :: a -> SVEmpty :~> SVEvent a
-after    :: Double -> a -> SVEmpty :~> SVEvent a
-
--- Lifting pure functions
-pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal a
-pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent a
-
--- Composition and routing
-(>>>)           :: (svIn :~> svMiddle) -> (svMiddle :~> svOut) 
-                   -> (svIn :~> svOut)
-first           :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
-second          :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut) 
-swap            :: (svLeft :^: svRight) :~> (svRight :^: svLeft)
-copy            :: sv :~> (sv :^: sv)
-ignore          :: sv :~> SVEmpty
-cancelLeft      :: (SVEmpty :^: sv) :~> sv
-cancelRight     :: (sv :^: SVEmpty) :~> sv
-uncancelLeft    :: sv :~> (SVEmpty :^: sv)
-uncancelRight   :: sv :~> (sv :^: SVEmpty)
-associate       :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3)) 
-unassociate     :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
-
--- Reactivity
-switch :: (svIn :~> (svOut :^: SVEvent (svIn :~> svOut))) 
-          -> svIn :~> svOut
-
--- Feedback
-loop :: ((svIn :^: svLoop) :~> (svOut :^: svLoop)) -> svIn :~> svOut
-
--- Time dependence
-time  :: SVEmpty :~> Double
-delay :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
-class TimeIntegrate
-
--- Joining
-union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
-combineSignals :: (a -> b -> c) 
-                  -> (SVSignal a :^: SVSignal b) :~> SVSignal c
-capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent a
-
--- Events
-filter         :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-\end{code}
-\hrule
-\caption{Signal function interface.}
-\label{figure:signal_function_interface}
-\end{figure}
-
-Signal functions are produced by combining primitive signal functions using
-the {\tt >>>} (sequential composition), {\tt first}, and {\tt second} routing
-combinators. Other routing signal functions are provided, but they are intended
-to be combined with, rather than to modify, other signal functions.
-
-\subsection{Evaluation Interface}
-The evaluation interface provides a monad~\cite{PeytonJones1993,PeytonJones2001}
-for specifying input to signal functions and handling their output. This
-interface also allows input actions to be separated from output actions, and
-from each other, so that a signal function may receive input from multiple IO
-systems and have its output handled by multiple output systems.
-
-The evaluation interface is a monad transformer~\cite{Liang1995}. This allows
-a signal function evaluation to be constructed from primitive evaluation actions
-(event pushing, input signal updating, and sampling) as well as actions from the
-underlying monad. These actions can then be run in the underlying monad
-to actuate the signal function. Because the actuation takes a signal function's
-state as input and produces a new state as output, in addition to the monadic
-side effects of handling the signal function's output, different evaluation
-actions may be taken on the same signal function from distinct locations within
-an application's code. This allows the evaluation interface to be easily
-integrated with event-loop style systems as well as traditional imperative
-IO systems.
-
-The evaluation interface consists of the monad transformer type, functions for
-constructing event inputs, initial input signal samples, and input signal
-updates, functions for constructing the vector of output handlers for a signal
-function, and evaluation actions. The full interface is shown in Fig.~\ref{figure:evaluation_interface}.
-
-\begin{figure}
-\begin{code}
--- Input helpers
-class SVRoutable -- Instances: SVSignalUpdate,
-                 -- SVEventOccurrence
-svLeft :: (SVRoutable r) => 
-          r svLeft -> r (SVAppend svLeft svRight)
-svRight :: (SVRoutable r) =>
-           r svRight -> r (SVAppend svLeft svRight)
-
--- Signal inputs
-data SVSignalUpdate
-data SVSample
-sample :: a -> SVSample (SVSignal a)
-sampleEvt :: SVSample (SVEvent a)
-sampleNothing :: SVSample SVEmpty
-combineSamples :: SVSample svLeft -> svSample svRight 
-                  -> svSample (svLeft :^: svRight)
-svSig :: a -> SVSignalUpdate (SVSignal a)
-
--- Event inputs
-data SVEventInput
-svOcc :: a -> SVEventInput (SVEvent a)
--- Actuation
-update :: (Monad m) => SVSignalUpdate svIn 
-           -> SFEvalT svIn svOut m ()
-push   :: (Monad m) => SVEventInput svIn
-          -> SFEvalT svIn svOut m ()
-sample :: SVEvalT svIn svOut m ()
--- Running
-data SVHandler
-emptyHandler :: SVHandler m SVEmpty
-eventHandler :: (a -> m ()) 
-                -> SVHandler m (SVEvent a)
-signalHandler :: (a -> m ()) 
-                 -> SVHandler m (SVSignal a)
-combineHandlers :: SVHandler m svLeft
-                   -> SVHandler m svRight
-                   -> SVHandler m (svLeft :^: svRight)
-initSFEvalT :: SVHandler m svOut -> SVSample svIn 
-               -> Double ->(svIn :~> svOut)
-               -> SFEvalState m svIn svOut
-runSFEvalT :: SFEvalT svIn svOut m a
-              -> SFEvalState m svIn svOut
-              -> m a
-\end{code}
-\hrule
-\caption{Evaluation interface.}
-\label{figure:evaluation_interface}
-\end{figure}
-
-\section{Implementation}
-\label{section:Implementation}
-
-We now turn our attention to the implementation of the signal function
-system. We will discuss representations of inputs, outputs, and signal functions,
-as well as the implementations of specific signal function combinators.
-
-\subsection{Input and Output Representations}
-\label{subsection:Input_and_Output_Representations}
-As discussed in Section~\ref{section:System_Design}, our implementation requires
-representations of the inputs and outputs of signal functions. Since signal
-functions are typed using signal vectors, the representation types must be
-parameterized over signal vectors. But signal vectors are uninhabited types, and
-in ordinary Haskell we cannot declare a data constructor which uses a component
-of a type parameter. Further, we cannot restrict which types may instantiate a
-type parameter for a particular data constructor.
-
-GADTs lift these restrictions, permitting type constraints to be applied
-to individual data constructors. In the declaration of a GADT, a type signature
-is provided for each data constructor, including the types of the constructor's
-parameters. When a GADT is pattern-matched, the types of the parameters are
-inferred using the constraints given in this signature. This is called type
-refinement.
-
-Using GADTs, we can construct representations of signal vectors to use in our
-implementation.
-
-When constructing a representation, we consider how to represent each
-component of a signal vector, whether it is a signal element, an empty
-element, or an append node. For instance, in order to represent a signal
-sample for all signals in an element, we construct the GADT shown in Figure~\ref{figure:signal_sample_representation}.
-
-\begin{figure}
-\begin{code}
-data SVSample sv where
-  SVSample        ::    a -> SVSample (SVSignal a)
-  SVSampleEvt     ::    SVSample (SVEvent a)
-  SVSampleNothing ::    SVSample SVEmpty
-  SVSampleBoth    ::    SVSample svLeft
-                     -> SVSample svRight
-                     -> SVSample (svLeft :^: svRight)
-\end{code}
-\hrule
-\caption{A representation which carries a value for each signal in a signal vector.}
-\label{figure:signal_sample_representation}
-\end{figure}
-
-\subsection{Signal Function Representations}
-\label{subsection:Signal_Function_Representations}
-With representations for signal function inputs and outputs in place, we turn
-to the representation of signal functions. A signal function must be able to
-respond to time increments and new signal samples, as well as event occurrences.
-Since we wish to be able to ``push'' event occurrences independently of
-sampling, events are handled using the last-updated time and signal sample.
-
-When a signal function is asked to respond to either of these types of inputs,
-it must produce an output consisting of zero or more event occurrences, a new
-signal function with the same type (to enable reactivity and state), and, if
-it is responding to a time and sample update, a delta which represents updates
-to the sample of its output signals.
-
-Finally, there is a special case for a signal function at time 0. It must be
-provided with an initial input sample before it can respond to incremental time
-and sample updates or to event occurrences.
-
-Signal functions are represented as a datatype with two constructors. The first
-constructor wraps a function from an input sample to an output sample and an
-initialized signal function. The second wraps two functions. The first accepts
-a time delta and a signal delta, and produces a signal delta, a list of event
-occurrences, and a new signal function as the output. The second accepts an
-event occurrence and produces a list of event occurrences and a new signal
-function as the output. The signal function representation is shown in Figure~\ref{figure:signal_function_representation}.
-
-\begin{figure}[t]
-\begin{code}
-data Initialized
-data NonInitialized
-
-type :~> svIn svOut = SF NonInitialized svIn svOut
-
-data SF init svIn svOut where
-  SF :: (SVSample svIn -> (SVSample svOut, SF Initialized svIn svOut))
-        -> SF NonInitialized svIn svOut
-  SFInit :: (Double -> SVDelta svIn 
-             -> (SVDelta svOut, [SVOccurrence svOut],
-                 SF Initialized svIn svOut))
-            -> (SVOccurrence svIn -> ([SVOccurrence svOut],
-                                      SF Initialized svIn svOut))
-            -> SF Initialized svIn svOut
-\end{code}
-\hrule
-\caption{Signal function representation}
-\label{figure:signal_function_representation}
-\end{figure}
-
-This representation allows a signal function to respond to events and time
-updates separately, and does not enforce the representation of events during
-time updates, as previous signal function systems do.
-
-\subsection{Signal Function Implementations}
-\label{subsection:Signal_Function_Implementations}
-
-Now that we have established representations for signal functions and their
-inputs and outputs, we show several examples of how signal function combinators
-are implemented.
-
-The {\tt identity} signal function's initialization function takes a sample,
-and returns that sample as its output sample. The initialized function returned
-is the {\tt identityInit} function, which is not exported by the module.
-The time and sample function ignores the provided time delta, produces the
-input signal delta as the output signal delta, and returns {\tt identityInit} as
-the new signal function. The event function takes an event occurrence and
-returns a singleton list containing that occurrence, along with the
-{\tt identityInit} function as the replacement signal function. The
-implementation is shown in Figure~\ref{figure:identity_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-identity :: sv :~> sv
-identity = SF (\sample -> (sample, identityInit))
-
-identityInit :: SF Initialized sv sv
-identityInit = SFInit (\_ delta -> (delta, [], identityInit))
-                      (\occ -> ([occ], identityInit))
-\end{code}
-\hrule
-\caption{{\tt identity} signal function implementation.}
-\label{figure:identity_signal_function_implementation}
-\end{figure}
-
-The {\tt constant} signal function is initialized with a value, which it then
-produces as its output forever. Thus, the initialized version of {\tt constant}
-never produces any output other than itself as a replacement
-signal function, and an empty signal delta. The implementation of {\tt constant}
-is shown in Fig.~\ref{figure:constant_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-constant :: a -> SVEmpty :~> SVSignal a
-constant x = SF (\_ -> (constantInit, sample x))
-
-constantInit :: SF Initialized SVEmpty (SVSignal a)
-constantInit = SFInit (\_ _ -> (deltaNothing, [], constantInit))
-                      (\_ -> ([], constantInit))
-\end{code}
-\hrule
-\caption{{\tt constant} signal function implementation.}
-\label{figure:constant_signal_function_implementation}
-\end{figure}
-
-The {\tt asap} signal function implementation requires a parameter for the
-initialized version of the signal function, to specify the value of the event
-occurrence. It replaces itself with the initialized version of the {\tt never}
-signal function after the first time step. The implementation is shown in
-Fig.~\ref{figure:asap_signal_function_implementation}.
-
-\begin{figure}
-\begin{code}
-asap :: a -> SVEmpty :~> SVEvent a
-asap x = SF (\_ -> (sampleEvt, asapInit x))
-
-asapInit :: a -> SF Initialized SVEmpty (SVEvent a)
-asapInit x = SFInit (\_ _ -> (deltaNothing, [occurrence x], neverInit))
-                    (\_ -> (never))
-\end{code}
-\hrule
-\caption{{\tt asap} signal function implementation.}
-\label{figure:asap_signal_function_implementation}
-\end{figure}
-
-The {\tt filter} function accepts events, applies a {\tt Maybe} predicate to
-their values, and produces the value produced by the predicate as an event
-occurrence, or no occurrence if the value is {\tt Nothing}. It is implemented
-by having the event occurrence response function apply the {\tt maybe} function
-from the Haskell prelude to the value returned by the predicate, as shown in 
-Fig.~\ref{figure:filter_signal_function_implementation}
-
-\begin{figure}[t]
-\begin{code}
-filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-filter p = SF (\_ -> (sampleEvt, filterInit p))
-
-filterInit :: (a -> Maybe b) -> SF Initialized (SVEvent a) (SVEvent b)
-filterInit p = SFInit (\_ _ -> (deltaNothing, [], filterInit p))
-                      (\evtOcc -> (maybe [] ((:[]) . occurrence) $
-                                   fromOccurrence evtOcc,
-                                   filterInit p))
-\end{code}
-\hrule
-\caption{{\tt filter} signal function implementation.}
-\label{figure:filter_signal_function_implementation}
-\end{figure}
-
-The {\tt associate} signal function is a routing signal function. It transforms
-a signal vector which is left-associated at the top level to one that is
-right-associated at the top level. Its implementation is representative of all
-of the routing functions, and is shown in Fig.~\ref{figure:associate_signal_function_implementation}.
-
-\begin{figure}[t]
-\begin{code}
-associate :: SVAppend (SVAppend sv1 sv2) sv3
-             :~> SVAppend sv1 (SVAppend sv2 sv3)
-associate = 
-  SF (\sigSample -> let (sigSampleLeft, sigSampleRight) = 
-                          splitSample sigSample
-                        (sigSampleLeftLeft, sigSampleLeftRight) =
-                          splitSample sigSampleLeft
-                    in (combineSamples sigSampleLeftLeft 
-                        (combineSamples sigSampleLeftRight sigSampleRight),
-                        associateInit))
-
-associateInit :: SF Initialized (SVAppend (SVAppend sv1 sv2) sv3)
-                                (SVAppend sv1 (SVAppend sv2 sv3))
-associateInit = SFInit (\_ sigDelta -> let (sigDeltaLeft, sigDeltaRight) =
-                                             splitDelta sigDelta
-                                           (sigDeltaLeftLeft,
-                                            sigDeltaLeftRight) =
-                                             splitDelta sigDeltaLeft
-                                       in (combineDeltas sigDeltaLeftLeft
-                                           (combineDeltas sigDeltaLeftRight
-                                                          sigDeltaRight),
-                                           [], associateInit)) 
-                       (\evtOcc -> (case chooseOccurrence evtOcc of
-                                      Left leftOcc -> 
-                                        case chooseOccurrence leftOcc of
-                                          Left leftLeftOcc ->
-                                            [occLeft leftLeftOcc]
-                                          Right leftRightOcc ->
-                                            [occRight $ occLeft leftRightOcc]
-                                      Right rightOcc ->
-                                        [occRight $ occRight $ rightOcc],
-                                    associateInit))
-\end{code}
-\hrule
-\caption{{\tt associate} signal function implementation.}
-\label{figure:associate_signal_function_implementation}
-\end{figure}
-
-The {\tt switch} function is the basic combinator used to introduce reactivity.
-It is given a signal function whose output is the append of two signal vectors.
-The left side of the signal vector is passed on as the output of the reactive
-signal function, and the right side is an event carrying signal functions. When
-an event occurrence is present on the right-side event output, the signal
-function carried by this occurrence replaces the signal function constructed
-by switch. Due to space concerns, the implementation is elided, but a brief
-description will suffice. The signal function stores the input sample
-provided during initialization, and updates it with deltas. When an event
-occurrence carrying a signal function is produced by either the event or 
-time function of the wrapped signal function, the stored signal sample is used
-to initialize the new signal function. If this occurs while handling an event
-input, the sample output by the new signal function's initialization is stored
-and is combined with its first output delta to produce the output delta at the
-next time step.
-
-The {\tt loop} function allows a signal function to see a component of its own
-output as input. This is primarily useful when a signal function has components
-which mutually depend on each others outputs, such as in physics simulations
-or games. Care must be taken that the feedback output is not immediately
-dependent on the feedback input, or sampling the signal function will not
-terminate. Loop is implemented by generating a recursive list (which will
-be infinite if the feedback is not decoupled) of event inputs, and by splitting
-the output signal delta and using the right side output delta as the right
-side input delta within a recursive let-binding. (The let-binding in Haskell
-always admits recursive bindings.) If the system is not completely decoupled,
-this will result in non-termination during an evaluation step. Decoupling can
-be achieved using {\tt delay} primitive.
-
-\subsection{Evaluation Interface Implementation}
-\label{subsection:Evaluation_Interface_Implementation}
-
-The evaluation interface must maintain state including the current signal
-function, the current time (to produce time deltas), updates to the input
-sample which have yet to be sampled, and the output handlers.
-
-The output handlers are stored in a structure similar to that for a signal
-sample. The difference is that both signal and event leaves contain values, and
-these values are functions from the leaf type to another type. The handler
-datatype is shown in Fig.~\ref{figure:handler_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SVHandler out sv where
-  SVHandlerEmpty  :: SVHandler out SVEmpty
-  SVHandlerSignal :: (a -> out) -> SVHandler out (SVSignal a)
-  SVHandlerEvent  :: (a -> out) -> SVHandler out (SVEvent a)
-  SVHandlerBoth   :: SVHandler out svLeft -> SVHandler out svRight 
-                     -> SVHandler out (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Handler datatype.}
-\label{figure:handler_datatype}
-\end{figure}
-
-We now need a datatype to hold the various components of the evaluation state.
-This type is shown in Fig~\ref{figure:evaluation_state_datatype}.
-
-\begin{figure}[t]
-\begin{code}
-data SFEvalState m svIn svOut 
-  = SFEvalState { 
-                  esSF :: SF Initialized svIn svOut,
-                  esOutputHandlers :: SVHandler (m ()) svOut,
-                  esLastTime :: Double,
-                  esDelta :: SVDelta svIn
-                }
-
-\end{code}
-\hrule
-\caption{Evaluation state datatype.}
-\label{figure:evaluation_state_datatype}
-\end{figure}
-
-The evaluation interface itself is a monad transformer, which we implement
-as a Haskell newtype wrapping the {\tt StateT} monad transformer. We do not
-export the raw {\tt put} and {\tt get} actions of the state monad, but instead
-implement the push, update, and sampling operations for signal function
-evaluation using {\tt put} and {\tt get}. The evaluation monad transformer is
-shown in Fig.~\ref{figure:signal_function_evaluation_monad_transformer}.
-Instances of typeclasses including {\tt Monad} and {\tt MonadTrans} are derived
-using the ``GeneralizedNewtypeDeriving'' extension to the Glasgow Haskell
-Compiler.
-
-\begin{figure}[t]
-\begin{code}
-newtype SFEvalT svIn svOut m a = StateT (SFEvalState m svIn svOut) m a
-\end{code}
-\hrule
-\caption{Signal function evaluation monad transformer.}
-\label{figure:signal_function_evaluation_monad_transformer}
-\end{figure}
-
-\section{Discussion}
-\label{section:Discussion}
-
-The system presented here, TimeFlies, demonstrates how using signal vectors
-to type inputs and outputs enables push-based evaluation of events in a
-signal-function system. We take advantage of this representation in several ways.
-
-First, by separating components of inputs and outputs in the types, we are free
-to create distinct, and often partial, representations of the input or output
-of a signal function. This enables us to represent only the event occurrence
-being pushed at that time.
-
-Second, this separation also permits us to separate the process of gathering
-the input to a signal function, and the process of handling its output, into
-different points in a program. Using the evaluation interface described, an
-event occurrence may be pushed onto one input of a signal function from one
-point in a program (e.g. a mouse click handler), an input signal
-may be updated in another (e.g. a mouse movement handler), and finally the
-system may be sampled in a third place (e.g. an animation or audio timed
-callback).
-
-Finally, this approach enables further work on the implementation of the signal
-function system to be separated from changes in the interface. By enabling
-differing representations of the inputs and outputs of signal functions, we are
-free to change these representations without the need to further constrain the
-input and output types.
-
-\section{Ongoing and Further Work}
-\label{section:Ongoing_and_Further_Work}
-
-TimeFlies, the system described here, has been implemented, but not
-extensively tested. The immediate goal is to create a real-time application
-which will permit a performance and implementation comparison of TimeFlies with
-Yampa, the current state-of-the-art pull-based signal-function system.
-
-In the future, we hope to apply run-time optimizations, using the technique used
-for Yampa, to create a push-pull self-optimizing signal-function system. Further,
-we hope to use this system as a basis for exploring signal-function FRP as a
-basis for general-purpose application frameworks.
-
-\section{Related Work}
-\label{section:Related_Work}
-
-Signal Function FRP was introduced as a model for Graphical User Interfaces~\cite{Courtney2001-1}.
-The system was originally termed ``AFRP'' (Arrowized FRP). Yampa is a rewrite of
-AFRP where signal functions apply a number of ad-hoc optimizatons to themselves
-as they evolve. Yampa demonstrated a modest performance improvement
-over AFRP~\cite{Nilsson2005}.
-
-Reactive is a classic FRP system which implements push-based evaluation for events
-by transforming behaviors to ``reactive~normal~form,'' where a behavior
-is a non-reactive behavior running inside a switch, whose event stream carries
-behaviors in reactive normal form. The system is evaluated by forking a Haskell
-thread to repeatedly sample the non-reactive behavior, and then blocking on the
-evaluation of the first occurrence in the event stream. When this occurrence
-is yielded, the evaluation thread for the behavior is killed and a new
-thread forked to evaluate the new behavior~\cite{Elliott2009}.
-
-\section{Conclusion}
-\label{section:Conclusion}
-
-We have presented TimeFlies, a system for push-pull signal-function Functional
-Reactive Programming, and have shown how the use of a signal vectors as input
-and output types for signal functions, together with GADT-based representations
-of the inputs and outputs, permits the implementation of a push-pull system.
-
-We have also described a general and flexible monadic evaluation interface for
-TimeFlies, which permits us to interface the TimeFlies system with different
-styles of IO systems, including multiple IO systems in the same application.
-
-This opens up the exciting possibility that a signal-function FRP could become
-an efficient and general framework for writing interactive applications.
+\input{abstract.tex}
+\input{introduction.tex}
+\input{design.tex}
+\input{implementation.tex}
+\input{discussion.tex}
+\input{related.tex}
+\input{conclusion.tex}
 

### 2012-11-27 Docs/IFL2012/related.tex
Break IFL paper into multiple files.

\section{Related Work}
\label{section:Related_Work}

Signal Function FRP was introduced as a model for Graphical User Interfaces~\cite{Courtney2001-1}.
The system was originally termed ``AFRP'' (Arrowized FRP). Yampa is a rewrite of
AFRP where signal functions apply a number of ad-hoc optimizatons to themselves
as they evolve. Yampa demonstrated a modest performance improvement
over AFRP~\cite{Nilsson2005}.

Reactive is a classic FRP system which implements push-based evaluation for events
by transforming behaviors to ``reactive~normal~form,'' where a behavior
is a non-reactive behavior running inside a switch, whose event stream carries
behaviors in reactive normal form. The system is evaluated by forking a Haskell
thread to repeatedly sample the non-reactive behavior, and then blocking on the
evaluation of the first occurrence in the event stream. When this occurrence
is yielded, the evaluation thread for the behavior is killed and a new
thread forked to evaluate the new behavior~\cite{Elliott2009}.

An alternate push-based system, based on {\em signal segments}, was presented in
work on the Curry-Howard correspondence between a form of temporal logic (LTL)
and FRP~\cite{Jeffrey2012}. However, this system continues to represent events
as option-valued signals. It therefore cannot be push-based in its response to
individual events, and cannot be push-pull, evaluating signals by pull-based
evaluation and events by push-based evaluation. It is push-based not in the
sense of evaluating events only when they occur, but rather in that it provides
a means to push signal segments (samples of a signal over intervals of time) to
the system and wait on its response.

### 2012-11-28 Docs/IFL2012/design.tex
IFL draft updates

@@
 
-A rigorous and formal elucidation of the semantics of signal-function FRP remains
-unattempted, but there is a sufficient change to the practical semantics of
-signal-function FRP between previous signal-function systems and TimeFlies to warrant
-some description.
-
-In previous systems such as Yampa, events were understood (and implemented) as option-valued
-signals. This approach is undesirable for several reasons. The most pressing reason is
-that it prohibits push-based evaluation of events, because events are embedded in the
-samples of a signal and must be searched for when sampling.
+A rigorous and formal elucidation of the semantics of signal-function FRP with signal
+vectors remains unattempted, though a categorical semantics for signal-function
+FRP using {\em fan categories} has been explored~\cite{Jeffrey2012}. In both
+these semantics and their implementation, events are represented (following
+AFRP~\cite{Courtney2001-1} and Yampa~\cite{Nilsson2005}) as option-valued
+signals. A push-based implementation of these semantics is proposed, but the
+unit being pushed is {\em segments} of signals, and not event occurrences. This
+approach requires processing the whole signal function at each time step,
+whether an event occurs or not, and checking for its occurrence at each step.
 
@@
 
-This concern drives the next concern. Events are not instantaneous in this formulation.
-If a signal is option valued, the sampling instant must fall within the interval where
-there is an event occurrence present for that event to be observed. If events are
-instantaneous, the probability of observing an event occurrence is zero.
+This concern drives the next concern. Events are not instantaneous in this
+formulation. If events are signals, the sampling instant must fall within
+the interval where there is an event occurrence present for that event to be
+observed. If events are instantaneous, the probability of observing an event 
+occurrence is zero.
 
@@
 since no guarantee of their order is given.
-
-A formal semantic description of signal function FRP would clarify the consequences of this decision somewhat,
-but is outside the scope of this thesis.
 
@@
 
-\begin{figure}
-\begin{code}
--- Pass the input unmodified to the output
-identity :: sv :~> sv
-
--- Produce a signal which is at all times the supplied value
-constant :: a -> SVEmpty :~> SVSignal a
-
--- An event with no occurrences
-never    :: SVEmpty :~> SVEvent a
-
--- An event with one occurrence, as soon as possible after
--- the signal function is initialized
-asap     :: a -> SVEmpty :~> SVEvent a
-
--- An event with one occurrence
--- after the specified amount of time has elapsed.
-after    :: Double -> a -> SVEmpty :~> SVEvent a
-\end{code}
-\hrule
-\caption{Basic signal function combinators.}
-\label{figure:basic_signal_functions}
-\end{figure}
-
 Two combinators are provided to lift pure functions to signal functions
@@
 an input event.
-
-\begin{figure}
-\begin{code}
--- Apply the given function to a signal at all points in time
-pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal b
-
--- Apply the given function to each event occurrence
-pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent b
-\end{code}
-\hrule
-\caption{Lifting pure functions.}
-\label{figure:lifting_pure_functions}
-\end{figure}
 
@@
 
-\begin{figure}
-\begin{code}
--- Use the output of one signal function as the input for another
-(>>>) :: (svIn :~> svBetween) -> (svBetween :~> svOut) -> svIn :~> svOut
-
--- Pass through the right side of the input unchanged
-first :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
-
--- Swap the left and right sides
-swap :: (svLeft :^: svRight) :~> (svRight :^: svLeft)
-
--- Duplicate the input
-copy :: sv :~> (sv :^: sv)
-
--- Ignore the input
-ignore :: sv :~> svEmpty
-
--- Remove an empty vector on the left
-cancelLeft :: (SVEmpty :^: sv) :~> sv
-
--- Add an empty vector on the left
-uncancelLeft :: sv :~> (SVEmpty :^: sv)
-
--- Make right-associative
-associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))
-\end{code}
-\hrule
-\caption{Routing combinators.}
-\label{figure:routing_combinators}
-\end{figure}
-
 Reactivity is introduced by means of the {\tt switch} combinator
@@
 
-\begin{figure}
-\begin{code}
-switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
-          -> svIn :~> svOut
-\end{code}
-\hrule
-\caption{Reactivity combinator.}
-\label{figure:switch_combinator}
-\end{figure}
-
 It is often useful for a signal function to receive a portion of its
@@
 in Figure~\ref{figure:feedback_combinator}.
-
-\begin{figure}
-\begin{code}
-loop ::    ((svIn :^: svLoop) :~> (svOut :^: svLoop))
-        -> svIn :~> svOut
-\end{code}
-\hrule
-\caption{Feedback combinator.}
-\label{figure:feedback_combinator}
-\end{figure}
 
@@
 
-\begin{figure}
-\begin{code}
--- Apply the function to each input occurrence,
--- and produce an occurrence for each Just.
-filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b
-
--- Apply the function to the stored accumulator
--- and the event occurrence, replacing the accumulator
--- and possibly outputting an occurrence
-accumulate ::    (a -> b -> (Maybe c, a))
-              -> a
-              -> SVEvent b :~> SVEvent c
-\end{code}
-\hrule
-\caption{Event-specific combinators.}
-\label{figure:event_specific_combinators}
-\end{figure}
-
 The filter combinators are stateless, and thus apply the function to only the new
 input value. They are useful for suppressing events, as well as for extracting one
-of multiple cases of a datatype. For instance, a splitter for events carrying
-{\tt Either}-valued occurrences could be written as:
-
-\begin{code}
-getLeft :: Either a b -> Maybe a
-getLeft (Left x) = Just x
-getLeft _ = Nothing
-
-getRight :: Either a b -> Maybe b
-getRight (Right x) = Just x
-getRight _ = Nothing
-
-split :: SVEvent (Either a b) :~> (SVEvent a :^: SVEvent b)
-split = copy >>> first (filter getLeft) >>> second (filter getRight)
-\end{code}
+of multiple cases of a datatype.
 
@@
 
-\begin{figure}
-\begin{code}
-union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
-combineSignals :: (a -> b -> c) -> (SVSignal a :^: SVSignal b) :~> SVSignal c
-capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent (b, a)
-\end{code}
-\hrule
-\caption{Joining combinators.}
-\label{figure:joining_combinators}
-\end{figure}
-
 A set of combinators are provided for making use of time-dependence in a signal
@@
 are shown in Figure~\ref{figure:time_combinators}
-
-\begin{figure}
-\begin{code}
-time      :: SVEmpty :~> SVSignal Double
-delay     :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
-integrate :: TimeIntegrate i => SVSignal i :~> SVSignal i
-\end{code}
-\hrule
-\caption{Time-dependent combinators.}
-\label{figure:time_combinators}
-\end{figure}
 
@@
 
-\begin{figure}
-\begin{code}
--- A vector of handlers for outputs
-data SVHandler out sv
-
--- The state maintained when evaluating a signal function
-data SFEvalState m svIn svOut
-
--- Create the initial state for evaluating a signal function
-initSFEval ::    SVHandler (m ()) svOut   -- Output handlers
-
-              -> SVSample svIn            -- Initial input samples
-
-              -> Double                   -- Initial external time,
-                                          -- corresponding to time 0 for
-                                          -- the signal function
-
-              -> (svIn :~> svOut)         -- Signal function to evaluate
-              -> SFEvalState m svIn svOut
-\end{code}
-\hrule
-\caption{State maintained when evaluating a signal function.}
-\label{figure:evaluation_state}
-\end{figure}
-
 Sets which correspond to signal vectors are built with ``smart'' constructors.
@@
 
-\begin{figure}
-\begin{code}
--- Class to overload left and right functions
-class SVRoutable r where
-  svLeft          :: r svLeft -> r (svLeft :^: svRight)
-  svRight         :: r svRight -> r (svLeft :^: svRight)
-
--- An input event occurrence
-data SVEventInput sv
-instance SVRoutable SVEventInput sv
-
--- An updated sample for a signal
-data SVSignalUpdate sv
-instance SVRoutable SVSignalUpdate sv
-
--- Create an event occurrence
-svOcc           :: a -> SVEventInput (SVEvent a)
-
--- Create an updated sample
-svSig           :: a -> SVSignalUpdate (SVSignal a)
-\end{code}
-\hrule
-\caption{Data types for ongoing input.}
-\label{figure:ongoing_input}
-\end{figure}
-
 The {\tt SFEvalT} monad is actually a monad transformer, that is, it is
@@
 function. These combinators are shown in Figure~\ref{figure:evaluation_combinators}.
-
-\begin{figure}
-\begin{code}
--- The evaluation monad
-data SFEvalT svIn svOut m a
--- Appropriate instances for Monad, MonadTrans, MonadIO,
--- Functor, Applicative, etc
-
--- Obtain an action in the underlying monad
--- from an SFEvalT and a new state.
-runSFEvalT ::    SFEvalT svIn svOut m a
-              -> SFEvalState m svIn svOut
-              -> m (a, SFEvalState m svIn svOut)
-
--- Push an event occurrence.
-push :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
-
--- Update the value of an input signal sample
--- (not immediately observed)
-update :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
-
--- Step forward in time, observing the updated signal values
-step :: (Monad m) => Double -> SFEvalT svIn svOut m ()
-\end{code}
-\hrule
-\caption{Evaluation combinators}
-\label{figure:evaluation_combinators}
-\end{figure}

### 2012-11-28 Docs/IFL2012/implementation.tex
IFL draft updates

@@
 
-\begin{figure}
-\begin{code}
-data SVSample sv where
-  SVSample      ::    a
-                   -> SVSample (SVSignal a)
-  SVSampleEvent ::    SVSample (SVEvent a)
-  SVSampleEmpty ::    SVSample SVEmpty
-  SVSampleBoth  ::    SVSample svLeft
-                   -> SVSample svRight
-                   -> SVSample (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Datatype for signal samples.}
-\label{figure:signal_sample_datatype}
-\end{figure}
-
 What about the event components? We want to represent event occurrences,
@@
 
-\begin{figure}
-\begin{code}
-data SVOccurrence sv where
-  SVOccurrence ::    a
-                  -> SVOccurrence (SVEvent a)
-  SVOccLeft    ::    SVOccurrence svLeft
-                  -> SVOccurrence (SVAppend svLeft svRight)
-  SVOccRight   ::    SVOccurrence svRight 
-                  -> SVOccurrence (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Datatype for event occurrences.}
-\label{figure:event_occurrence_datatype}
-\end{figure}
-
 We add one more representation for signals, in order to avoid uneccessary
@@
 This datatype is shown in Figure~\ref{figure:signal_update_datatype}.
-
-\begin{figure}
-\begin{code}
-data SVDelta sv where
-  SVDeltaSignal  ::    a
-                    -> SVDelta (SVSignal a)
-  SVDeltaNothing ::    SVDelta sv
-  SVDeltaBoth    ::    SVDelta svLeft
-                    -> SVDelta svRight
-                    -> SVDelta (SVAppend svLeft svRight)
-\end{code}
-\hrule
-\caption{Datatype for signal updates.}
-\label{figure:signal_update_datatype}
-\end{figure}
 
@@
 is shown in Figure~\ref{figure:signal_function_datatype}.
-
-\begin{figure}
-\begin{code}
-data Initialized
-
-data NonInitialized
-
-data SF init svIn svOut where
-  SF     ::    (SVSample svIn 
-                  -> (SVSample svOut,
-                      SF Initialized svIn svOut)) 
-            -> SF NonInitialized svIn svOut
-  SFInit ::    (Double 
-                  -> SVDelta svIn
-                  -> (SVDelta svOut,
-                      [SVOccurrence svOut],
-                      SF Initialized svIn svOut)) 
-            -> (SVOccurrence svIn
-                  -> ([SVOccurrence svOut],
-                      SF Initialized svIn svOut))
-            -> SF Initialized svIn svOut
-\end{code}
-\hrule
-\caption{Datatype and empty types for signal functions.}
-\label{figure:signal_function_datatype}
-\end{figure}
 
@@
 update as the next time step. After this, it acts as the new signal function.
-This wrapping has some performance implications, which are discussed
+This wrapping has some performance implications, which are discussed in Section~\ref{section:Discussion}.
 
@@
 
-The {\tt first} and {\tt second} combinators are similar to serial composition,
-but they transform only one signal function. For signal changes, they must split
-the set of input changes into those which will be passed to the signal function
-and those which will be simply passed along to the output, and then recombine
-them on the other side. For event occurrences, the occurrence must be
-pattern-matched to determine whether to call the event continuation from the
-provided signal function or passed through, and output event occurrences must
-have the suitable routing constructor re-applied. In any case, when a
-continuation has been applied, the combinator must be recursively applied to the
-new signal function.
-
 The looping feedback combinator is particularly tricky. As it is currently
@@
 in Haskell, and the non-strict evaluation of Haskell, to implement feedback.
-
-The {\tt filter} combinators are simple to implement. Their sampling
-continuation is superfluous, and the event continuation merely applies the
-supplied function, and constructs an output list based on the result.
-
-The {\tt accumulate} combinators are implemented in terms of the {\tt filter}
-and {\tt switch} combinators, as shown in
-Figure~\ref{figure:accumulate_implementation}.
-
-The implementation of the joining combinators is simple. The {\tt union}
-combinator simply passes along every event occurrence it receives on either
-input, stripping off the left and right combinator. This is acceptable since we
-do not insist on a total ordering of events, or an event time resolution greater
-than the sampling rate. The {\tt combineSignals} combinator maintains the values
-of both signals, and applies the combination function whenever one is updated.
-The {\tt capture} combinator maintains the input signal value, and adds it to
-each event occurrence.
 
@@
 
-The {\tt SFEvalT} monad transformer is a newtype wrapper around the {\tt StateT}
-monad available in the Haskell {\tt transformers} package. The StateT monad
-wraps any other monad, providing a state which may be read using the {\tt get}
-action and replaced with the {\tt put} action. An action in the underlying
-monad is obtained by supplying a {\tt StateT} value and an initial state
-value to the {\tt runStateT} function.
-
 The {\tt SFEvalT} monad transformer does not make the {\tt get} and {\tt put}

### 2012-11-28 Docs/IFL2012/design.tex
Draft updates to design and implementation sections of IFL paper

@@
 
-The type constructor for signal functions is shown in
-Figure~\ref{figure:signal_function_types}. For the {\tt init} parameter, only
-one possible instantiation is shown. The usefulness of this type parameter,
-along with another instantation which is hidden from users of the library,
-is discussed in the section on implementation of signal functions
-(Section~\ref{section:Implementation}).
-
 Signal functions with signal vectors as input and output types form a
@@
 The representation of signal functions is discussed in
-Section~\ref{section:Implementation-Signal_Functions}. The type synonyms
-{\tt :\textasciitilde>} and {\tt :\textasciicircum:} are included for
-readability and are not crucial to the FRP system.
-
+Section~\ref{section:Implementation-Signal_Functions}.
 \begin{figure}
@@
 
-Monad transformers~\cite{Jones1995} provide a means to combine the functionality
-of multiple monads. A monad transformer is a monad with an extra type parameter.
-This type parameter is instantiated with the type constructor of the underlying
-monad, and an extra operation {\tt lift} is provided which converts values in
-the underlying monad to values in the monad transformer.
-
-The evaluation monad is a monad transformer. This permits it to be used in
-conjunction with Haskell's {\tt IO} monad (or any other monad) to describe how
-input is to be obtained for the signal function being evaluated, and how outputs
-are to be handled.
-
-The evaluation monad, in addition to the standard monad operators, provides a
-means of {\em initializing} a signal function, and a means of translating the
-monadic value describing evaluation to a value in the underlying monad. This
-means, for instance, that we can obtain an action in the {\tt IO} monad to
-evaluate a signal function.
-
-The type of the evaluation monad must track the input and output type of the
-signal function. The monad's context stores a mapping from outputs to handling
-actions. There are thus four type parameters to the monad's type
-constructor: the input signal vector, the output signal vector, the type of the
-underlying monad, and the monadic type parameter. The type is shown in
-Figure~\ref{figure:evaluation_monad_types}.
-
-\begin{figure}
-\begin{code}
--- A signal function's evaluations state
-data SFEvalState svIn svOut m
--- The evaluation monad
-data SFEvalT svIn svOut m a
-
-instance Monad m => Monad (SFEvalT svIn svOut m)
-\end{code}
-\hrule
-\caption{Evaluation monad types.}
-\label{figure:evaluation_monad_types}
-\end{figure}
+Our interface for evaluating signal functions takes the form of a monad
+transformer~\cite{Jones1995}. This formulation permits us to provide specific
+temporal actions to perform on a signal function in sequence, and in conjunction
+with actions taken in Haskell's {\tt IO} monad, or another monad. Actions are
+provided in this monad to push an event to the signal function's input, step
+the signal function's time, or modify a signal on the signal functions input to
+be observed at the next time step. Outputs are handled by a vector of actuation
+functions in the underlying monad, supplied by the user of the interface. This
+vector is structured by parameterizing it over the signal function's output
+signal vector.
 
@@
 
-With the exception of {\tt identity}, all of the basic signal functions have
-empty inputs. This allows these signal functions to be used to insert values
-into the network which are known when the signal function is created, without
-having to route those values from an input.
+Two combinators are provided to lift pure functions to signal functions.
+The {\tt pureSignalTransformer} combinator applies the pure function pointwise
+to a signal. The {\tt pureEventTransformer} combinator applies the function to
+the value carried by each occurrence of an input event.
 
-Two combinators are provided to lift pure functions to signal functions
-(Figure~\ref{figure:lifting_pure_functions}). The {\tt pureSignalTransformer}
-combinator applies the pure function to a signal at every sample point. The
-{\tt pureEventTransformer} combinator applies the function to each occurrence of
-an input event.
+The routing combinators are used to combine signal functions, and to re-arrange
+and discard components of signal vectors in order to connect signal functions.
 
-The routing combinators are used to combine signal functions, and
-to re-arrange signal vectors in order to connect signal functions.
-The routing combinators are shown in Figure~\ref{figure:routing_combinators}.
-
-Only those combinators which modify or combine signal functions
+Only those routing combinators which modify or combine signal functions
 ({\tt (>>>)}, {\tt first}, {\tt second}) are reactive (may replace themselves in
@@
 
-Several combinators are provided for manipulating, suppressing, and generating events.
-Each of the combinators has an option variant and a list variant. The option variant
-produces an output event occurrence whenever the application of the supplied function
-to the input event produces a value. The list version produces an event occurrence for
-each of the elements of the output list. The combinators are shown in Figure~\ref{figure:event_specific_combinators}.
+The {\tt filter} and {\tt filterList} combinators are stateless signal functions
+which apply a function to each event occurrence. The {\tt filter} combinator
+expects an option-typed output, while the {\tt list} combinator expects a list
+output. If the function yields a non-empty output, one or more output occurrences
+are produced with the elements of that output. If the output is empty, there
+are no output occurrences.
 
-The filter combinators are stateless, and thus apply the function to only the new
-input value. They are useful for suppressing events, as well as for extracting one
-of multiple cases of a datatype.
-
-The accumulate combinators are stateful, applying the supplied function
-to both the input value and an accumulator. This function has two results:
-the option or list of output event occurrence values, and the new value
-for the accumulator.
+The {\tt accumulate} and {\tt accumulateList} functions are similar, except that
+they maintain a state value which is replaced after each input event occurrence
+by an additional output from the function, and supplied to the function in
+addition to the value of each event occurrence.
 
@@
 
-The {\tt union} combinator is a non-deterministic merge of event
-streams. Any event which occurs on either input will occur
-on the output. For simultaneous event occurrences, the order of occurrence
-is not guaranteed, but the occurrence itself is. This construct
-is also guaranteed to respect the relationship of event occurrences to sampling
-intervals.
+The {\tt union} combinator is a non-deterministic merge of two event streams.
+Simultaneity of events is only observable up to the sampling rate, but events
+can occur in between sampling intervals. Thus, if two events occur
+simultaneously, one occurrence is arbitrarily selected to be first in the output
+stream.
 

### 2012-11-28 Docs/IFL2012/implementation.tex
Draft updates to design and implementation sections of IFL paper

@@
 
-We create concrete representations which are parameterized over
-these types. These concrete representations will be expressed as GADTs, allowing
-each data constructor of the representation to fill in a specific signal vector
-type for the parameter of the representation.
+Using the Glasgow Haskell Compiler's extension for Generalized Algebraic
+Datatypes~\cite{Cheney2003,Xi2003,PeytonJones2006}, we specify three concrete
+datatypes which are parameterized over signal vectors, and represent
+information about one temporal point in a signal function's evaluation.
 
-The first thing to represent is {\em samples}, which are sets of values for
-the signal components of a signal vector. Therefore, we create a representation
-which carries a value for every {\tt SVSignal} leaf of a signal vector. In order
-to do this, we restrict each of our constructors to a single signal vector type.
-So there are three leaf constructors: {\tt SVSample}, which carries a value; and
-{\tt SVSampleEvent} and {\tt SVSampleEmpty}, which do not. This ensures that the
-only way to represent a sample leaf is with the {\tt SVSample} constructor,
-which carries a value of the appropriate type. The datatype is shown in
-Figure~\ref{figure:signal_sample_datatype}.
+The first type carries the instantaneous values of all signals in a signal vector.
+There is precisely one constructor for each type constructor of a signal vector ({\tt SVEmpty},
+{\tt SVSignal}, {\tt SVEvent}, and {\tt SVAppend}). The constructor for {\tt SVSignal}
+carries a value typed with the type parameter of the {\tt SVSignal} constructor,
+the rest do not. Thus, in this representation, for any signal vector, there is
+a node in the representation for each node in the signal vector, and precisely
+the signal nodes have values.
 
-What about the event components? We want to represent event occurrences,
-each of which will correspond to at most one event in the vector. So a different
-representation is called for. In this case, there will be only three
-constructors. One constructor will represent an event leaf, and the other will
-represent a single value on the left or right side of the node ({\tt SVAppend}),
-ignoring all of the type structure on the other side. This representation
-describes a path from the root of the signal vector, terminating at an event
-leaf with a value.
+The next representation contains a constructor for {\tt SVEvent} nodes, carrying
+a value typed by the type parameter of {\tt SVEvent}, and two constructors, each
+of which carries a representation for one child of a {\tt SVAppend} node (the
+left or right), and is parametric in the other. Thus, an instance of this
+representation is a path from the root of a signal vector to an event node,
+carrying exactly one value for the event.
 
-By pattern matching on the path constructors, we can determine which subvector
-of a signal vector an event occurrence belongs to, repeatedly refining it until
-we determine which event in the vector the occurrence corresponds to. The
-datatype for occurrences is shown in Figure~\ref{figure:event_occurrence_datatype}.
+The last representation is used for efficient implementation, and has three
+constructors. One constructor represents a {\tt SVSignal} node and has a value
+typed by its type parameter. The second carries representations of both
+children of a {\tt SVAppend} node. The third is parametric in its signal vector,
+and thus represents the absence of information for an arbitrary signal vector.
+This representation contains values for all, some, or none of the {\tt SVSignal}
+nodes of a signal vector.
 
-We add one more representation for signals, in order to avoid uneccessary
-representations of the values of all signals when not all signals have changed
-their values. This representation allows us to represent the values of zero or
-more of the signals in a signal vector. To accomplish this, we replace the
-individual constructors for the {\tt SVEmpty} and {\tt SVEvent} leaves with %there is a better word, something about "useless but not really", for this
-a single, unconstrained constructor. This constructor can represent an arbitrary
-signal vector. We can use the constructor for signal vector nodes and the 
-constructor for sample leaves to represent the updated values, while filling
-in the unchanged portions of the signal vector with this general constructor.
-This datatype is shown in Figure~\ref{figure:signal_update_datatype}.
+Signal functions are implemented by combining two strategies for temporal
+updates. The first such strategy is sampling. A signal function has a 
+continuation which may be invoked at regular intervals, using the partial
+representation of signals described above as input and producing such a
+representation as output. This amounts to repeated sampling of the output
+signals of a signal function.
 
-We now have concrete datatypes for an implementation to operate on. Our next
-task is to represent transformers of temporal data, which themselves may change
-with time. The common approach to this task is sampling, in which a program
-repeatedly checks for updated information, evaluates it, updates some state,
-and produces an output. This is the essence of pull-based evaluation.
-
-Another approach is notification, in which the program exposes an interface
-which the source of updated information may invoke. This is a repeated entry
-point to the program, which causes the program to perform the same tasks
-listed above, namely, evaluate the updated information, update state, and
-produce output. The strategy of notification as opposed to repeated checking is
-the essence of push-based evaluation.
-
-Signal functions are declarative objects, and not running processes. They have
-no way to invoke sampling themselves. They can, however, expose separate
-interfaces for when sampling is invoked, and when they are notified of an event
-occurrence. This creates two control paths through a signal function. One of
-these control paths is intended to be invoked regularly and frequently with
-updates to the time and sample values, and the other is intended to be invoked
-only when an event occurs. The benefit of separating these control paths is that
-events are no longer defined in terms of sampling intervals, and need not even
-be considered in sampling, except when they are generated by a condition on a
-sample. On the other hand, events can be responded to even if the time has not
-yet come for another sample, and multiple events can be responded to in a single
-sampling interval.
+The other approach is notification. Signal functions have a second continuation
+which is only invoked when there is an event occurence on their input. The input
+to this continuation is, of course, the representation of event occurrences. If
+events were signals, they would be sampled as above, and the sampling interval
+would not match the event occurrence interval. It would thus be necessary to
+represent these non-occurrences. By implementing event handling separately from
+signal updates, we eliminate the need for a representation of event
+non-occurences and the need for invoking event-handling code at every time step.
 
@@
 and is specialized to {\tt Initialized} or {\tt NonInitialized} depending on the
-constructor. The other two type parameters are the input and output signal
-vectors, respectively. The signal functions that a user will compose are\
-non-initialized signal functions. They must be provided with an initial set of
-input signals (corresponding to time zero). When provided with this input, they
-produce their time-zero output, and an initialized signal function. The datatype
-is shown in Figure~\ref{figure:signal_function_datatype}.
+constructor. It would of course be possible to represent these as two distinct
+datatypes, but this representation communicates the intuition that these are
+two states of the same object, rather than separate objects. The other two type
+parameters are the input and output signal vectors, respectively. The signal
+functions that a user will compose are non-initialized signal functions.
+They must be provided with an initial set of input signal values, which are considered
+the sample for time zero, and represented by the full representation of signals
+described above. When provided with this input, they produce their time-zero
+output and an initialized signal function.
 
-Initialized signal functions carry two continuations. The first continuation
-takes a time differential and a set of signal updates, and returns a set of
-signal updates, a collection of event occurrences, and a new initialized signal
-function of the same type. This is the continuation called when sampling.
+Initialized signal functions carry the two continuations described above.
+The first continuation takes a time differential and a set of signal updates
+(the partial representation of signals) and returns a set of signal updates, a
+collection of event occurrences, and a new initialized signal function of the
+same type. This is the continuation called when sampling.
 
@@
 
-Note that each of these continuations uses one or more of the concrete
-representations of signal vectors, and applies the type constructor for the
-representation to the input or output signal vector for the signal function.
-
-Having specified a type constructor for signal functions, we must now provide
-combinators which produce signal functions typed with this type constructor.
-Space precludes a complete description of the combinators in the TimeFlies
-system, but we will present
+This is the type and framework for our signal function implementation.
+Combinators are implemented as functions which return these types with specific
+implementations of the continuations. Space precludes a full discussion of the
+implemenation of each combinator, but some of the examples will be discussed.
 

### 2012-11-28 Docs/IFL2012/design.tex
Thesis draft updates

@@
 The representation of signal functions is discussed in
-Section~\ref{section:Implementation-Signal_Functions}.
+Section~\ref{subsection:Implementation-Signal_Functions}.
 \begin{figure}
@@
 
-The basic signal functions (Figure~\ref{figure:basic_signal_functions})
+The basic signal functions
 provide very simple operations. The {\tt identity} signal function, as expected,
@@
 
-Reactivity is introduced by means of the {\tt switch} combinator
-(Figure~\ref{figure:switch_combinator}). The design of this combinator
+Reactivity is introduced by means of the {\tt switch} combinator.
+The design of this combinator
 allows modular packaging of reactivity. A signal function can determine
@@
 back around to the first. Many signal-processing algorithms also depend
-on feedback. The combinator which provides this ability is shown
-in Figure~\ref{figure:feedback_combinator}.
+on feedback. This ability is provided by the {\tt loop} operator.
 
@@
 
-The accumulator is useful when responses to multiple event occurrences
-(from one or more sources) must be coordinated. For instance, in the
-benchmark application (see Chapter~\ref{chapter:Evaluation_and_Comparisons}),
-a table is maintained that allows knowledge from previous event occurrences
-(packets from a network switch) to be used in deciding where the present
-packet ought to go.
-
 The joining combinators provide the ability to combine two event
-streams, two signals, or a signal and an event stream. These
-combinators are shown in Figure~\ref{figure:joining_combinators}
+streams, two signals, or a signal and an event stream.
 
@@
 function. These combinators allow the modification of signals and events with
-respect to time, and the observation of the current time value. The combinators
-are shown in Figure~\ref{figure:time_combinators}
+respect to time, and the observation of the current time value.
 
@@
 The evaluation interface provides a modified state monad which holds a signal
-function, together with some additional information, as its state (shown in Figure~\ref{figure:evaluation_state}.
+function, together with some additional information, as its state.
 Rather than monadic instructions to put and get the state, the monad provides instructions
@@
 In order to initialize the state, the user must supply a set of handlers, the
-signal function to evaluate, and initial values for all of the signal inputs
-(Figure~\ref{figure:initial_input}).
+signal function to evaluate, and initial values for all of the signal inputs.
 
@@
 functions to construct sample updates and event occurrences, and to specify
-their place in the vector (Figure~\ref{figure:ongoing_input}).
+their place in the vector.
 
@@
 which produces the result and a new state is obtained with the {\tt runSFEvalT}
-function. These combinators are shown in Figure~\ref{figure:evaluation_combinators}.
+function.

### 2012-11-28 Docs/IFL2012/discussion.tex
Thesis draft updates

@@
 input and output types.
-
-\section{Ongoing and Further Work}
-\label{section:Ongoing_and_Further_Work}
-
-TimeFlies, the system described here, has been implemented, but not
-extensively tested. The immediate goal is to create a real-time application
-which will permit a performance and implementation comparison of TimeFlies with
-Yampa, the current state-of-the-art pull-based signal-function system.
-
-In the future, we hope to apply run-time optimizations, using the technique used
-for Yampa, to create a push-pull self-optimizing signal-function system. Further,
-we hope to use this system as a basis for exploring signal-function FRP as a
-basis for general-purpose application frameworks.

### 2012-11-28 Docs/IFL2012/implementation.tex

@@
 
-The simplest example of the implementation of is the {\tt identity} combinator,
-shown in Figure~\ref{figure:identity_implementation}. This signal function
+The simplest example of the implementation of is the {\tt identity} combinator.
+This signal function
 simply passes all of its inputs along as outputs. The initialization function
@@
 Combinators which take one or more signal functions as input must recursively
-apply themselves, as is shown in the implementation of serial composition
-(Figure~\ref{figure:serial_composition_implementation}). They must also
+apply themselves, as is shown in the implementation of serial composition.
+They must also
 handle initialization, retaining the initialized signal functions and passing
@@
 of changes to the signal. This has some performance implications, which are
-discussed in Chapter~\ref{chapter:Evaluation_and_Comparisons}.
+discussed in Section~\ref{section:Discussion}.
 
@@
 Since these signal functions are stateless and primitive, they can simply
-return themselves as their replacements. The {\tt swap} combinator is shown as
-an example in Figure~\ref{figure:swap_implementation}
+return themselves as their replacements.
 
@@
 Nevertheless, this implementation performs quite well as it currently exists, as
-we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparisons}.
+we demonstrate in Section~\ref{section:Discussion}.
 
@@
 
-The evaluation interface is exported as shown in
-Section~\ref{section:System_Design_and_Interface-Evaluator}.
-
 The {\tt SFEvalState} type constructor parameterizes over input types for signal
@@
 
-The {\tt update} action updates the current input signal delta, which will be
+The {\tt update} action modifies the current input signal update, which will be
 sampled at the next {\tt step} action. It has no immediately observable effect.
-It simply updates the value of one signal in the input signal vector, a delta
-of which is stored in the {\tt SFEvalState} record.
+It simply updates the value of one signal in the input signal vector.
 
 The {\tt step} action takes a time delta, and calls the signal continuation of
-the stored signal function. It actuates the outputs using the stored handlers,
-replaces the stored delta with an empty delta, and stores the resulting new
-signal function in the state.
+the stored signal function with the time delta and the stored signal update. It
+actuates the outputs using the stored handlers, replaces the stored delta with
+an empty delta, and stores the resulting new signal function in the state.

### 2012-11-28 Docs/IFL2012/abstract.tex
Tentative edits to abstract - come back to this later

@@
 \begin{abstract}
-Functional Reactive Programming is a promising class of systems for writing
-interactive and time-dependent programs. Signal-function FRP is a subclass of
-these systems which provides advantages of modularity and correctness, but
-has proven difficult to efficiently implement.
+Functional Reactive Programming is a promising \new{approach} 
+\rn{systems -- seems weird to identify it with the implementations... model, approach?} 
+for writing interactive and time-dependent programs. Signal-function FRP 
+is variant of FRP 
+% a subclass of these systems 
+which provides advantages in modularity and correctness, but
+has proven difficult to implement efficiently.
 
-The abstraction of signal vectors provides the necessary type apparatus to
+\textred{The abstraction of signal vectors provides the necessary type apparatus to
 distinguish components of the input and output of signal functions which benefit
 from a push-based implementation from those which benefit from a pull-based
-implementation, and to combine both implementation strategies in a single system.
+implementation, and to combine both implementation strategies in a single system.}
 
-We describe a signal-function FRP system which provides push-based evaluation
+\rn{Without understanding it a priori, the above para is hard to
+  parse.  How about the following?}
+
+\new{One important implementation trade-off is whether evaluation of a composition 
+of signal functions is {\em push-based} or {\em pull-based}.
+The {\em signal vector} technique provides a type-level
+  description of the inputs and outputs of signal functions, while
+  distinguishing} components that benefit from a push-based
+implementation from those that benefit from a pull-based
+implementation.  This makes it possible to combine both implementation
+strategies in a single system.
+
+We describe a novel signal-function FRP system which provides push-based evaluation
 for events, pull-based evaluation for signals, and a simple monadic evaluation
-interface which permits the system to be easily integrated with one or more
+interface that permits the system to easily integrate with one or more
 IO systems.

### 2012-11-29 Docs/IFL2012/abstract.tex
Edits to intro.

@@
 
-\new{One important implementation trade-off is whether evaluation of a composition 
+\new{One important implementation trade-off is whether evaluation 
+% of a composition 
 of signal functions is {\em push-based} or {\em pull-based}.
-The {\em signal vector} technique provides a type-level
-  description of the inputs and outputs of signal functions, while
+The {\em signal-vector} technique addresses this by providing a type-level
+  description of the inputs and outputs of signal functions, 
   distinguishing} components that benefit from a push-based
@@
 
-We describe a novel signal-function FRP system which provides push-based evaluation
+We describe a \new{novel} \rn{is it novel? how? need clear contributions}
+ signal-function FRP system which provides push-based evaluation
 for events, pull-based evaluation for signals, and a simple monadic evaluation

### 2012-11-29 Docs/IFL2012/introduction.tex
Edits to intro.

@@
 Functional Reactive Programming (FRP) is a class of systems for describing
-reactive programs. Reactive programs are programs which, rather than taking
-a single input and producing a single output, must accept multiple inputs and 
-alter temporal behavior, including the production of multiple outputs, based
-on these inputs.
+reactive programs. Reactive programs are programs which, rather than 
+\new{batch-processing inputs to produce an output,}
+% taking a single input and producing a single output, 
+\new{instead map time-varying inputs onto outputs.}
+%% must accept multiple inputs and 
+%% alter temporal behavior, including the production of multiple outputs, based
+%% on these inputs.
 
-An FRP system will provide a means of manipulating {\em behaviors} and
-{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
-but the definition is the same. A behavior or signal is, semantically, a
-function from time to a value. An event is a discrete, possibly infinite, and 
-time-ordered sequence of occurrences, which are times paired with values.
+An FRP system provides a means of manipulating {\em behaviors} and
+{\em events}. 
+\rn{Switched to present tense.  Don't need to add additional future or past tense...}
+Behaviors are often referred to as {\em signals} in FRP literature,
+but the definition is the same.  Semantically, a behavior or signal is a
+function from time to a value.  
+\new{Whereas an event is a ordered sequence of discrete occurrences,
+  carrying both values and timestamps.}
+%% An event is a discrete, possibly infinite, and 
+%% time-ordered sequence of occurrences, which are times paired with values.
+
+\rn{Looks like you should standardize early on {\bf signal} and leave
+  extra occurrences of {\bf behavior} out of this intro.}
 
 FRP systems can generally be categorized as ``classic FRP,'' 
-which corresponds to the originally described FRP system in that behaviors
-and events are manipulated directly and are first-class values in the FRP
-system, or ``signal-function FRP,'' in which behaviors (generally termed
-signals in this approach) and events are not first-class values, but signal
-functions are first class values. Signal functions are time-dependent and
+% which corresponds to the originally described FRP system 
+in which behaviors
+and events are manipulated directly as first-class values,
+% in the FRP system,
+or {\em signal-function FRP}. 
+% in which behaviors  (generally termed signals in this approach) 
+\new{In signal-function FRP, the programmer cannot directly manipulate
+  signals, rather, functions from input-signal to output-signal are
+  the central abstraction.  Thus}
+%% and events are not first-class values, but signal
+%% functions are first class values. 
+signal functions are time-dependent,
 reactive transformers of signals, events, or combinations of signals and events.
-
-FRP combines behaviors and events through the use of {\em switching}, in which
+%
+FRP {\em combines} behaviors and events through the use of {\em switching}, in which
 a behavior (in classic FRP) or a signal function (in signal-function FRP) is
@@
 
-Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
-Recent work on classic FRP has focused on efficient implementation. One approach to
+\todo{TODO: it would be good to give SOME sense for how an FRP
+  evaluation proceeds (e.g. a strawman) before getting into the next para.}
+
+Classic FRP was first described as a system for interactive animations~\cite{Elliott1997}.
+Recent work on classic FRP has focused on efficient implementation \rn{CITATION?}.  One approach to
 efficiency is to separate the evaluation of behaviors and events, since suitable 
@@
 The initial implementations of FRP made use of pull-based evaluation for both
-behaviors and events. Reactive~\cite{Elliott2009}, as well as more recent systems such as
+behaviors and events. ``Reactive''~\cite{Elliott2009}, as well as more recent systems such as
 ``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
@@
 have used pull-based evaluation for both signals and events. This is due to
-the ease of implementation of pull-based evaluation, and the types used for
-signal functions which do not permit distinguishing signals and events, or
-constructing only part of the input (for instance, one event occurrence.)
+the ease of implementation of pull-based evaluation, and because the types used for
+signal functions do not permit distinguishing signals and events, or
+constructing only part of the input (for instance, one event occurrence).
+
+\rn{What does it mean to construct one event occurrence?  As opposed to?}
 
@@
 may contain multiple values, including option values for events. Signal vectors
-are uninhabited types, which can be used to type partial or full representations
+are uninhabited types, which can be used to describe partial or full representations
 of the signal function inputs and outputs.
@@
 
+\rn{Need a concrete contributions.   ``The first system to apply
+  push-pull implementation strategies to signal-function FRP'', etc.
+  It doesn't hurt for it to be bulletted.}
+
 Section~\ref{section:System_Design_and_Interface} describes design choices for the system,
@@
 between events and signals is achieved. Section~\ref{section:Discussion} is a
-discussion of the usefulness of our implementation. 
+discussion of the \textred{usefulness} \rn{more specific?} of our implementation. 
 Section~\ref{section:Ongoing_and_Further_Work} describes the current and future

### 2012-11-29 Docs/IFL2012/design.tex
edits to 2.0

@@
 
-The goal of FRP is to provide an efficient, declarative abstraction for creating
-reactive programs. Towards this overall goal, there are three goals which this
-system is intended to meet.
+The purpose of FRP is to provide an efficient, declarative abstraction for creating
+reactive programs. To this end, there are three goals which the TimeFlies
+system is intended to meet. \rn{What are the three!?  Enumerate.  Push
+based, composable, IO-system agnostic?}
 
@@
 producing output at a consistently high sample rate.
+
+\todo{At this point we still need a clear notion of what a pull-based
+  evaluation of events would actually do.}
 
@@
 
-It is fine for a system to be composable with regards to itself, but an FRP
-system must interact with the outside world. Since we cannot anticipate every
+\rn{Can the above be made into a slightly more technical claim?
+  Composable abstractions enable composing $O(N)$ components while
+  reasoning about the $O(N)$ local interactions, rather than the
+  $O(N^2)$ global interactions between all components?  Second, I
+  think not needing to know the implementation is separate from
+  ``composability''; surely there is some name for that: ADTs,
+  ``encapsulation'', information hiding, etc.}
+
+% It is fine for a system to be composable with regards to itself, but
+% an FRP system must
+\new{Further, an FRP program is not a closed system; it must}
+interact with the outside world. Since we cannot anticipate every
 possible form of input and output that the system will be asked to interact
-with, we must interface with Haskell's IO system. In particular, most libraries
+with, we must interface with Haskell's IO system. 
+\rn{First mention of Haskell... need some transition?}
+In particular, most libraries
 for user interaction (e.g. GUI and graphics libraries such as GTK+ and GLUT) and
 most libraries for time-dependent IO (e.g. audio and video systems) make use of
-the event loop abstraction. In this abstraction, event handlers are registered
-with the system, and then a command is issued to run a loop which detects events
-and runs the handlers, and uses the results of the handlers to render the
-appropriate output.
+the event-loop abstraction. In this abstraction, event handlers are registered
+\new{centrally}; 
+\new{once initiated, the running event-loop detects events and invokes handlers; and
+  handlers are responsible for rendering output.}
+%% and then a command is issued to run a loop which detects events
+%% and runs the handlers, and uses the results of the handlers to render the
+%% appropriate output.
+
+\rn{Is the event loop really an {\em abstraction}?  (What is
+  abstracted?)  Or is it just a ``model'' or an API?}
 
@@
 such as simple imperative systems, threaded systems, or network servers.
+\rn{vague... not sure what these mean.  Threaded = multiple event
+  loop?  Alternative to an event loop is what... blocking IO operations?}
 

### 2012-11-29 Docs/IFL2012/design.tex
pass over section 2.1

@@
 The purpose of FRP is to provide an efficient, declarative abstraction for creating
-reactive programs. To this end, there are three goals which the TimeFlies
+reactive programs. To this end, there are three goals that the TimeFlies
 system is intended to meet. \rn{What are the three!?  Enumerate.  Push
@@
 % an FRP system must
-\new{Further, an FRP program is not a closed system; it must}
-interact with the outside world. Since we cannot anticipate every
+\new{Further, an FRP program is not a closed system; it}
+interacts with the outside world. Since we cannot anticipate every
 possible form of input and output that the system will be asked to interact
@@
 
-\rn{Is the event loop really an {\em abstraction}?  (What is
-  abstracted?)  Or is it just a ``model'' or an API?}
+\rn{Is the event loop really an {\em abstraction}?  What is
+  abstracted?  Or is it just a ``model'' or an API?}
 
@@
 \rn{vague... not sure what these mean.  Threaded = multiple event
-  loop?  Alternative to an event loop is what... blocking IO operations?}
+  loop?  Alternative to an event loop is... what?... blocking IO operations?}
 
@@
 
-A rigorous and formal elucidation of the semantics of signal-function FRP with signal
+A rigorous 
+%and formal elucidation of the 
+semantics of signal-function FRP with signal
 vectors remains unattempted, though a categorical semantics for signal-function
@@
 signals. A push-based implementation of these semantics is proposed, but the
-unit being pushed is {\em segments} of signals, and not event occurrences. This
+unit being pushed is a {\em segment} of signals, and not event occurrences. This
 approach requires processing the whole signal function at each time step,
@@
 
+\rn{Might want to make it clear -- the ``whole'' signal function means
+  ``all signal functions in the program'' (i.e. which are composed
+  to form the global signal function) right?}
+
 Another concern is that this approach limits the event rate to the sampling rate.
 The rate of sampling should, at some level, not matter to the FRP system. Events
-which occur between sampling intervals are never observed by the system. Since
-the signal function is never evaluated except at the sampling intervals, no
+which occur between sampling intervals are never observed by the
+system. \rn{An FRP system has sampling intervals?  Where were we told
+  this!?}  
+Since the signal function is never evaluated except at the sampling intervals, no
 events can otherwise be observed.
 
-This concern drives the next concern. Events are not instantaneous in this
+This \new{raises another} concern. Events are not instantaneous in this
 formulation. If events are signals, the sampling instant must fall within
@@
 observed. If events are instantaneous, the probability of observing an event 
-occurrence is zero.
+occurrence is zero. \rn{So.... how do existing systems/semantics
+  compensate for this?  They allow events to have temporal extent?}
 
@@
 representation and evaluation of signals.
-
+%
 This freedom yields the additional ability to make events independent of the sampling
 interval altogether. The semantics of event handling in TimeFlies is that an event occurrence
-is responded to immediately, and does not wait for the next sampling instant. This allows events
-to be instantaneous, and further, allows multiple events to occur within a single sampling interval.
+is responded to immediately, and does not wait for the next sampling
+instant. 
+\rn{Could you flesh out how this is a {\em semantics} issue and not
+  merely a more efficient or accurate implementation?  Ah, that's
+  answered in the next para, but might be good to use the word
+  semantics somewhere and connect this explicitly.} 
+This allows events to be instantaneous, and further, allows multiple
+events to occur within a single sampling interval.
+\rn{Clarification---even in traditional approaches multiple {\bf
+    different} events could occur in one sampling round right?  It's
+  just that {\em each} event [stream] could only have one occurrence per
+  sampling interval?}
 
@@
 same sampling interval. Further, the precise time of an event occurrence cannot be observed, only the 
-time of the last sample prior to the occurrence.
+time of the last \rn{[signal?]} sample prior to the occurrence.
+\rn{That last sentence had an ambiguous parse wherein the first time I read it I thought it was saying that you
+couldn't tell the timestamp of event N only of the preceding event $N-1$, which of course makes no sense,
+but it might be good to clarify what {\em sample} you're referring to.}
 
@@
 since no guarantee of their order is given.
+\rn{Nice.}
 

### 2012-11-29 Docs/IFL2012/design.tex
1st pass over section 2.2

@@
 In a strongly and statically typed functional language, types are a key part of
-an interface. Types provide a mechanism for describing and ensuring properties
+any interface. Types provide a mechanism for describing and ensuring properties
 of the interface's components and about the systems created with these
@@
 output. In most signal function systems, a signal function takes exactly one
-input and produces exactly one output. Multiple inputs or outputs are handled
-by making the input or output a tuple, and combinators which combine or split
+input signal and produces exactly one output signal. Multiple inputs or outputs are handled
+by making the input or output a signal of tuples, and combinators which combine or split
 the inputs or outputs of a signal assume this.  Events are represented at the
-type level as a particular type of signal, and at the value level as an option,
+type level as a particular type of signal, and at the value level as an option:
 either an event occurrence or not. 
+\rn{That's a bit confusing, because of the issue of whether we're talking about {\em event occurrences} or
+  event streams.  At runtime there is only a representation for the former, which is a simple {\tt Maybe},
+  right?}
 
@@
 {\tt SVEmpty} type constructor forms the unit type.
-
+%
 The representation of signal functions is discussed in
@@
 the signal function's time, or modify a signal on the signal functions input to
-be observed at the next time step. Outputs are handled by a vector of actuation
+be observed at the next time step. Outputs are handled by a vector \rn{vector? in the same sense as a
+  signal vector?} of actuation
 functions in the underlying monad, supplied by the user of the interface. This
 vector is structured by parameterizing it over the signal function's output
-signal vector.
+signal vector. \rn{huh? more concrete way to say it?}
 

### 2012-11-29 Docs/IFL2012/design.tex
1st pass of section 2.4

@@
+% (set-fill-column 100)
+
 \section{System Design and Interface}
@@
 
-Signal functions are constructed from combinators, which are primitive signal
-functions and operations to combine these primitives. These combinators are
-grouped as basic signal functions, lifting operations for pure functions,
+%% Signal functions are constructed from combinators, which are primitive signal
+%% functions and operations to combine these primitives. 
+\new{TimeFlies includes basic signal functions and a library of signal function combinators, falling
+  into these categories:}
+% These combinators are grouped as basic signal functions, 
+lifting operations from pure functions to signal functions,
 routing, reactivity, feedback, event processing, joining, and time dependence.
+%
+\rn{Eek, I get thrown off by a definition of combinators which is so broad as to include everything.  I
+  like the definition here 
+   \url{http://stackoverflow.com/tags/combinators/info} : 
+   {\em A combinator is a higher-order function that uses only function application and earlier defined
+    combinators to define a result from its arguments.}  Does everything in the above list meet that
+   definition?}
 
@@
 and discard components of signal vectors in order to connect signal functions.
-
+%
 Only those routing combinators which modify or combine signal functions
-({\tt (>>>)}, {\tt first}, {\tt second}) are reactive (may replace themselves in
+({\tt (>>>)}, {\tt first}, {\tt second}) are reactive (\ie{} may replace themselves in
 response to events, and then only if they inherit their reactivity from the
@@
 does not generate an infinite sequence of events in a single time-step.
+\rn{Presumably that is different from previous formulations of FRP?  Because ... there was no
+  notion of multiple events in a sampling interval?}
 
@@
 expects an option-typed output, while the {\tt list} combinator expects a list
-output. If the function yields a non-empty output, one or more output occurrences
+output. \rn{and input right?} If the function yields a non-empty output, one or more output occurrences
 are produced with the elements of that output. If the output is empty, there
@@
 by an additional output from the function, and supplied to the function in
-addition to the value of each event occurrence.
+addition to the value of each event occurrence. \rn{Murky prose definition...}
 
@@
 streams, two signals, or a signal and an event stream. 
+\rn{How does that work temporally?  For event/event is this a zip
+  that matches up the left events w/ the right events?}
 
@@
 time step, and must produce exactly one output sample.
+\rn{Wait that sounded like it was covered by join?}
 
@@
 of an event occurrence to that event occurrence.
+\rn{Again, what is a signal-event {\bf join} if not that?}
 
-These three combinators together provide the ability to combine elements of
+These three combinators \rn{Which?} together provide the ability to combine elements of
 signal vectors. By combining these combinators, along with the {\tt cancelLeft}
@@
 respect to time, and the observation of the current time value.
-
+%
 The simplest time-dependent combinator is {\tt time}, which simply outputs
@@
 introduced through {\tt switch}.
+
+\rn{If that's the simplest what are the more complex ones?  If not worth mentioning it seems easier to just
+    describe {\bf time}.}
 
@@
 integral of its input signal with respect to time.
+
+\todo{What's novel here?  After this section is de-listified need to emphasize that.}
 

### 2012-11-29 Docs/IFL2012/design.tex
more on 2.4

@@
 
+\rn{Evaluation interface: Very cool.  It sounds a bit your FRP implementation itself is kind of
+  ``scriptable'' rather than black box...}
+
 \subsection{Combinators}
@@
 
-\todo{What's novel here?  After this section is de-listified need to emphasize that.}
+\todo{What's novel here?  After this section is de-listified, need to emphasize that.}
 
@@
 \label{subsection:System_Design_and_Interface-Evaluation_Interface}
+
+\rn{Is the thing that really sets TimeFlies apart?  (Apart from being the first push/pull for SFFRP.)}
 
@@
 Sets which correspond to signal vectors are built with ``smart'' constructors.
-For instance, to construct a set of handlers, individual handling functions are
+For instance, to construct a set of handlers, 
+\rn{is this ``set'' the same as the earlier ``vector of actuation functions''?}
+individual handling functions are
 lifted to handler sets with the {\tt signalHandler} and {\tt eventHandler}
@@
 using the {\tt combineHandlers} function.
+%
+\rn{Examples, examples, examples.  There's a lot of prose here that gives the reader some idea, but doesn't
+  show them exactly what it would look like.  Too much guesswork required to fill in the gaps.}
 
@@
 update inputs, and step time, as well as actions lifted from the underlying
-monad (used to obtain these inputs). An action in the underlying monad
+monad (used to obtain these inputs). 
+\rn{That sounds introductory... I thought that was supposed to be what we were learning about throughout
+  this subsection?}
+An action in the underlying monad
 which produces the result and a new state is obtained with the {\tt runSFEvalT}

### 2012-11-29 Docs/IFL2012/implementation.tex
implementation section

@@
 respect to the temporal semantics of their inputs and outputs.
+\rn{What does that mean?  ``implement with respect to the temporal
+  semantics'' seems like an underspecified relation.  I'm also not
+  sure about the use of ``purest'' -- isn't this a semantics
+  vs. runtime/implementation issue?}
 
@@
 the library. Rather, they will specify how data is represented during the
-temporal evaluation of signal functions. We then implement signal functions
-as records of functions from these concrete types to these concrete types paired
+temporal evaluation of signal functions. 
+\rn{Maybe ``temporal evaluation'' could be more concrete... this means
+  repeated running sampling computation, right?  Maybe there should be a term
+  for each sampling interval, epoch, whatever?}
+We then implement signal functions
+as records of functions from these concrete types \textred{to these concrete types} paired
 with new signal functions. The evaluation system for signal functions then
@@
 extension for this was far from stable at the time this system was created.
+\rn{This should probably be a footnote... doesn't need to be highlighted.}
 
@@
 datatypes which are parameterized over signal vectors, and represent
-information about one temporal point in a signal function's evaluation.
+information about \textred{one} temporal point \rn{and not others?
+  rephrase?} in a signal function's evaluation.
 
@@
 of which carries a representation for one child of a {\tt SVAppend} node (the
-left or right), and is parametric in the other. Thus, an instance of this
+left or right), and is parametric in the other. \rn{?? Example?} Thus, an instance of this
 representation is a path from the root of a signal vector to an event node,
@@
 
-This is the type and framework for our signal function implementation.
+This is the \textred{type and} framework for our signal function implementation.
 Combinators are implemented as functions which return these types with specific
@@
 them as output when time advances far enough. The integrate combinator performs
-rectangle-rule integration on signal samples with respect to time.
+rectangle-rule integration on signal samples with respect to
+time. \rn{repetitive with previous section...}
 

### 2012-11-29 Docs/IFL2012/abstract.tex
edits to ifl abstract and intro

--- 2012-11-27 Docs/IFL2012/abstract.tex
@@
 \begin{abstract}
-Functional Reactive Programming is a promising class of systems for writing
-interactive and time-dependent programs. Signal-function FRP is a subclass of
-these systems which provides advantages of modularity and correctness, but
-has proven difficult to efficiently implement.
+Functional Reactive Programming is a promising model for writing interactive and
+time-dependent programs. Signal-function FRP is a variant of FRP which provides
+advantages in modularity and correctness, but has proven difficult to implement
+efficiently.
 
-The abstraction of signal vectors provides the necessary type apparatus to
-distinguish components of the input and output of signal functions which benefit
-from a push-based implementation from those which benefit from a pull-based
-implementation, and to combine both implementation strategies in a single system.
+One important implementation trade-off is whether evaluation of an FRP program
+is push-based or pull-based. Signal vectors are an abstraction which addresses
+this by providing a type-level description of the inputs and outputs of signal
+functions, distinguishing components that benefit from a push-based
+implementation from those that benefit from a pull-based implementation. This
+makes it possible to combine both implementation strategies in a single system.
 
@@
 for events, pull-based evaluation for signals, and a simple monadic evaluation
-interface which permits the system to be easily integrated with one or more
-IO systems.
+interface that permits the system to easily integrate with one or more IO
+systems.
 \end{abstract}

### 2012-11-29 Docs/IFL2012/introduction.tex
edits to 2.0
edits to ifl abstract and intro

@@
 Functional Reactive Programming (FRP) is a class of systems for describing
-reactive programs. Reactive programs are programs which, rather than 
-\new{batch-processing inputs to produce an output,}
-% taking a single input and producing a single output, 
-\new{instead map time-varying inputs onto outputs.}
-%% must accept multiple inputs and 
-%% alter temporal behavior, including the production of multiple outputs, based
-%% on these inputs.
+interactive programs. These are programs which, rather than batch-processing
+inputs to produce an output, instead map time-varying inputs onto outputs.
 
 An FRP system provides a means of manipulating {\em behaviors} and
-{\em events}. 
-\rn{Switched to present tense.  Don't need to add additional future or past tense...}
-Behaviors are often referred to as {\em signals} in FRP literature,
+{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
 but the definition is the same.  Semantically, a behavior or signal is a
-function from time to a value.  
-\new{Whereas an event is a ordered sequence of discrete occurrences,
-  carrying both values and timestamps.}
-%% An event is a discrete, possibly infinite, and 
-%% time-ordered sequence of occurrences, which are times paired with values.
+function from time to a value. An event is a ordered sequence of discrete
+occurrences, carrying both values and timestamps.
 
-\rn{Looks like you should standardize early on {\bf signal} and leave
-  extra occurrences of {\bf behavior} out of this intro.}
+FRP systems can generally be categorized as ``classic FRP,'' in which behaviors
+and events are manipulated directly as first-class values,
+or {\em signal-function FRP}. In signal-function FRP, the programmer cannot
+directly manipulate signals, rather, the programmer manipulates time-dependent,
+reactive transformers of signals, events, or combinations of signals and events.
 
-FRP systems can generally be categorized as ``classic FRP,'' 
-% which corresponds to the originally described FRP system 
-in which behaviors
-and events are manipulated directly as first-class values,
-% in the FRP system,
-or {\em signal-function FRP}. 
-% in which behaviors  (generally termed signals in this approach) 
-\new{In signal-function FRP, the programmer cannot directly manipulate
-  signals, rather, functions from input-signal to output-signal are
-  the central abstraction.  Thus}
-%% and events are not first-class values, but signal
-%% functions are first class values. 
-signal functions are time-dependent,
-reactive transformers of signals, events, or combinations of signals and events.
-%
 FRP {\em combines} behaviors and events through the use of {\em switching}, in which
@@
 Classic FRP was first described as a system for interactive animations~\cite{Elliott1997}.
-Recent work on classic FRP has focused on efficient implementation \rn{CITATION?}.  One approach to
-efficiency is to separate the evaluation of behaviors and events, since suitable 
-strategies give best performance in each case. Push-based evaluation evaluates a
-system only when input is available, and is thus suitable for discrete inputs
-such as events. Pull-based evaluation evaluates the system as quickly as
-possible, polling for input, and is preferable for behaviors and signals.
-The initial implementations of FRP made use of pull-based evaluation for both
-behaviors and events. ``Reactive''~\cite{Elliott2009}, as well as more recent systems such as
-``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
-events and pull-based evaluation for behaviors. This is known as ``push-pull''
-evaluation.
+Recent work on classic FRP has focused on efficient implementation. One approach
+to efficiency is to separate the evaluation of behaviors and events, since
+suitable strategies give best performance in each case. Push-based evaluation
+evaluates a system only when input is available, and is thus suitable for
+discrete inputs such as events. Pull-based evaluation evaluates the system as
+quickly as possible, polling for input, and is preferable for behaviors and
+signals. The initial implementations of FRP made use of pull-based evaluation
+for both behaviors and events. ``Reactive''~\cite{Elliott2009}, as well as more
+recent systems such as ``reactive-banana''~\cite{Apfelmus}, make use of
+push-based evaluation for events and pull-based evaluation for behaviors. This
+is known as ``push-pull'' evaluation.
 
-All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
-have used pull-based evaluation for both signals and events. This is due to
-the ease of implementation of pull-based evaluation, and because the types used for
+Almost all implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
+have used pull-based evaluation for both signals and events. This is due to the
+ease of implementation of pull-based evaluation, and because the types used for
 signal functions do not permit distinguishing signals and events, or
-constructing only part of the input (for instance, one event occurrence).
-
-\rn{What does it mean to construct one event occurrence?  As opposed to?}
+constructing only part of the input (for instance, an occurrence on one of the
+event inputs). One signal function system~\cite{Jeffery2012} does permit the
+user to push {\em segments} of signals, but, like the others, does not
+distinguish events in a way that permits them to be separately pushed.
 
@@
 
-We present TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
+TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
 favorite quotation of one of the author's philosophy instructors, used to
 demonstrate the ambiguity of language. The origin of the quotation is unknown.}
-a push-pull signal-function FRP system. We hope to demonstrate the feasibility
-of such an approach to FRP, and provide a basis for further research into
-efficient implementation of signal-function FRP. We also describe a powerful
-evaluation interface for TimeFlies, which permits us to use TimeFlies to
-describe applications which make use of multiple and differing IO libraries.
-
-\rn{Need a concrete contributions.   ``The first system to apply
-  push-pull implementation strategies to signal-function FRP'', etc.
-  It doesn't hurt for it to be bulletted.}
+is a push-pull signal-function FRP system, implemented in the Haskell
+programming language. We show that typing signal-function FRP with signal
+vectors allows push-pull evaluation of signal functions. We demonstrate an
+evaluation interface that allows the user of TimeFlies to take advantage of this
+behavior in interfacing TimeFlies to a wide range of IO systems. Finally, we
+give benchmark results which demonstrate that the push-pull evaluation strategy
+makes TimeFlies more efficient than the current pull-based implementation of
+signal-function FRP.
 

### 2012-11-30 Docs/IFL2012/design.tex
Performance results in discussion section - graphic not rendering properly

--- 2012-11-29 Docs/IFL2012/design.tex
@@
 reactive programs. To this end, there are three goals that the TimeFlies
-system is intended to meet. \rn{What are the three!?  Enumerate.  Push
-based, composable, IO-system agnostic?}
+system is intended to meet:
 
-Efficient evaluation is the motivation for push-based evaluation of events.
-Since FRP programs are expected to  interact with an outside world in real time,
-efficiency cannot simply be measured by the runtime of a program. Thus, when speaking of efficiency,
-we are expressing a desire that the system utilize as few system resources as possible
-for the task at hand, while responding as quickly as possible to external inputs and
-producing output at a consistently high sample rate.
-
-\todo{At this point we still need a clear notion of what a pull-based
-  evaluation of events would actually do.}
+An FRP system should permit efficient evaluation. Since FRP programs are
+expected to  interact with an outside world in real time, efficiency cannot
+simply be measured by the runtime of a program. Thus, when speaking of
+efficiency, we are expressing a desire that the system utilize as few system
+resources as possible for the task at hand, while responding as quickly as
+possible to external inputs and producing output at a consistently high sample
+rate.
 
@@
 
-\rn{Can the above be made into a slightly more technical claim?
-  Composable abstractions enable composing $O(N)$ components while
-  reasoning about the $O(N)$ local interactions, rather than the
-  $O(N^2)$ global interactions between all components?  Second, I
-  think not needing to know the implementation is separate from
-  ``composability''; surely there is some name for that: ADTs,
-  ``encapsulation'', information hiding, etc.}
-
-% It is fine for a system to be composable with regards to itself, but
-% an FRP system must
-\new{Further, an FRP program is not a closed system; it}
-interacts with the outside world. Since we cannot anticipate every
-possible form of input and output that the system will be asked to interact
-with, we must interface with Haskell's IO system. 
-\rn{First mention of Haskell... need some transition?}
-In particular, most libraries
-for user interaction (e.g. GUI and graphics libraries such as GTK+ and GLUT) and
-most libraries for time-dependent IO (e.g. audio and video systems) make use of
-the event-loop abstraction. In this abstraction, event handlers are registered
-\new{centrally}; 
-\new{once initiated, the running event-loop detects events and invokes handlers; and
-  handlers are responsible for rendering output.}
-%% and then a command is issued to run a loop which detects events
-%% and runs the handlers, and uses the results of the handlers to render the
-%% appropriate output.
-
-\rn{Is the event loop really an {\em abstraction}?  What is
-  abstracted?  Or is it just a ``model'' or an API?}
+Further, an FRP program is not a closed system; it interacts with the outside
+world. Since we cannot anticipate every possible form of input and output that
+the system will be asked to interact with, we must interface with IO systems.
+In particular, most libraries for user interaction (e.g. GUI and graphics
+libraries such as GTK+ and GLUT) and most libraries for time-dependent IO
+(e.g. audio and video systems) make use of the event-loop model. In this
+model, event handlers are registered for various forms of input, and then a
+loop is initiated which checks for, or blocks waiting for, input and then
+actuates the appropriate handlers to compute on the input and actuate outputs.
 
@@
 while being flexible enough to enable its use with other forms of IO systems,
-such as simple imperative systems, threaded systems, or network servers.
-\rn{vague... not sure what these mean.  Threaded = multiple event
-  loop?  Alternative to an event loop is... what?... blocking IO operations?}
+such as systems which use blocking IO and threads, rather than event loops.
 
@@
 
-A rigorous 
-%and formal elucidation of the 
-semantics of signal-function FRP with signal
-vectors remains unattempted, though a categorical semantics for signal-function
-FRP using {\em fan categories} has been explored~\cite{Jeffrey2012}. In both
-these semantics and their implementation, events are represented (following
+A rigorous semantics of signal-function FRP with signal vectors remains
+unattempted, though a categorical semantics for signal-function FRP using
+{\em fan categories} has been explored~\cite{Jeffrey2012}. In both these
+semantics and their implementation, events are represented (following
 AFRP~\cite{Courtney2001-1} and Yampa~\cite{Nilsson2005}) as option-valued
@@
 
-\rn{Might want to make it clear -- the ``whole'' signal function means
-  ``all signal functions in the program'' (i.e. which are composed
-  to form the global signal function) right?}
-
 Another concern is that this approach limits the event rate to the sampling rate.
@@
 which occur between sampling intervals are never observed by the
-system. \rn{An FRP system has sampling intervals?  Where were we told
-  this!?}  
-Since the signal function is never evaluated except at the sampling intervals, no
-events can otherwise be observed.
+system. Since the signal function is never evaluated except at the sampling
+intervals, no events can otherwise be observed.
 
-This \new{raises another} concern. Events are not instantaneous in this
+This raises another concern. Events are not instantaneous in this
 formulation. If events are signals, the sampling instant must fall within
@@
 observed. If events are instantaneous, the probability of observing an event 
-occurrence is zero. \rn{So.... how do existing systems/semantics
-  compensate for this?  They allow events to have temporal extent?}
+occurrence is zero. So in current signal-function semantics, events occurrences
+implicitly have some duration which contains the point of sampling.
 
@@
 representation and evaluation of signals.
-%
-This freedom yields the additional ability to make events independent of the sampling
-interval altogether. The semantics of event handling in TimeFlies is that an event occurrence
-is responded to immediately, and does not wait for the next sampling
-instant. 
-\rn{Could you flesh out how this is a {\em semantics} issue and not
-  merely a more efficient or accurate implementation?  Ah, that's
-  answered in the next para, but might be good to use the word
-  semantics somewhere and connect this explicitly.} 
-This allows events to be instantaneous, and further, allows multiple
-events to occur within a single sampling interval.
-\rn{Clarification---even in traditional approaches multiple {\bf
-    different} events could occur in one sampling round right?  It's
-  just that {\em each} event [stream] could only have one occurrence per
-  sampling interval?}
 
-There are two tradeoffs intrinsic to this approach. The first is that events are only partially ordered
-temporally. There is no way to guarantee the order of observation of event occurrences occurring in the
-same sampling interval. Further, the precise time of an event occurrence cannot be observed, only the 
-time of the last \rn{[signal?]} sample prior to the occurrence.
-\rn{That last sentence had an ambiguous parse wherein the first time I read it I thought it was saying that you
-couldn't tell the timestamp of event N only of the preceding event $N-1$, which of course makes no sense,
-but it might be good to clarify what {\em sample} you're referring to.}
+This freedom yields the additional ability to make events independent of the
+sampling interval altogether. The semantics of event handling in TimeFlies is
+that an event occurrence is responded to immediately, and does not wait for the
+next sampling instant. This allows events to be instantaneous, and further,
+allows multiple occurrences for an event within a single sampling interval.
 
-In return for giving up total ordering and precise observation of the timing of events, we obtain the
-ability to employ push-based evaluation for event occurrences, and the ability to non-deterministically
-merge event occurrences. When events being input to a non-deterministic merge have simultaneous occurrences,
-we arbitrarily select one to occur first. This does not violate any guarantee about time values, since
-they will both have the same time value in either case, and does not violate any guarantee about ordering,
-since no guarantee of their order is given.
-\rn{Nice.}
+There are two tradeoffs intrinsic to this approach. The first is that events are
+only partially ordered temporally. There is no way to guarantee the order of
+observation of event occurrences occurring in the same sampling interval.
+Further, the precise time of an event occurrence cannot be observed, only the 
+time that the signal function was last sampled prior to the occurrence.
+
+In return for giving up total ordering and precise observation of the timing of 
+events, we obtain the ability to employ push-based evaluation for event
+occurrences, and the ability to non-deterministically merge event occurrences.
+When events being input to a non-deterministic merge have simultaneous
+occurrences, we arbitrarily select one to occur first. This does not violate any
+guarantee about time values, since they will both have the same time value in
+either case, and does not violate any guarantee about ordering, since no
+guarantee of their order is given.
 
@@
 output. In most signal function systems, a signal function takes exactly one
-input signal and produces exactly one output signal. Multiple inputs or outputs are handled
-by making the input or output a signal of tuples, and combinators which combine or split
-the inputs or outputs of a signal assume this.  Events are represented at the
-type level as a particular type of signal, and at the value level as an option:
-either an event occurrence or not. 
-\rn{That's a bit confusing, because of the issue of whether we're talking about {\em event occurrences} or
-  event streams.  At runtime there is only a representation for the former, which is a simple {\tt Maybe},
-  right?}
+input signal and produces exactly one output signal. Multiple inputs or outputs
+are handled by making the input or output a signal of tuples, and combinators
+which combine or split the inputs or outputs of a signal assume this. Events are
+represented at the type level as a particular type of signal, and at the value
+level as an option: either an event occurrence or not.
 
@@
 the signal function's time, or modify a signal on the signal functions input to
-be observed at the next time step. Outputs are handled by a vector \rn{vector? in the same sense as a
-  signal vector?} of actuation
+be observed at the next time step. Outputs are handled by a set of actuation
 functions in the underlying monad, supplied by the user of the interface. This
-vector is structured by parameterizing it over the signal function's output
-signal vector. \rn{huh? more concrete way to say it?}
+set's type is parameterized over the signal function's output signal vector.
 
@@
 
-Signal functions are constructed from combinators, which are primitive signal
-functions and operations to combine these primitives. These combinators are
-grouped as basic signal functions, lifting operations for pure functions,
-routing, reactivity, feedback, event processing, joining, and time dependence.
+TimeFlies includes basic signal functions and a library of signal function
+combinators, grouped into lifting operations from pure functions to signal
+functions, routing, reactivity, feedback, event processing, joining, and time
+dependence.
 
-The basic signal functions
-provide very simple operations. The {\tt identity} signal function, as expected,
-simply copies its input to its output. The {\tt constant} signal function
-produces the provided value as a signal at all times. The {\tt never} signal
-function has an event output which never produces occurrences. The {\tt asap}
-function produces an event occurrence with the given value at the first time
-step after it is switched into the network. The {\tt after} function waits for
-the specified amount of time before producing an event occurrence.
+These signal functions can be combined in serial (the output of one becomes
+the input of the other) or in parallel (their input and output vectors are
+combined). Signal functions are independently reactive or non-reactive, which
+is the basis of their compositionality. Reactivity is introduced using the
+{\tt switch} combinator, which takes an initial signal function, watches
+one of its outputs for an event occurrence carrying a new signal function,
+and replaces itself with this new signal function. Its type is:
 
-Two combinators are provided to lift pure functions to signal functions.
-The {\tt pureSignalTransformer} combinator applies the pure function pointwise
-to a signal. The {\tt pureEventTransformer} combinator applies the function to
-the value carried by each occurrence of an input event.
+\begin{code}
+switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
+          -> svIn :~> svOut
+\end{code}
 
-The routing combinators are used to combine signal functions, and to re-arrange
-and discard components of signal vectors in order to connect signal functions.
-
-Only those routing combinators which modify or combine signal functions
-({\tt (>>>)}, {\tt first}, {\tt second}) are reactive (may replace themselves in
-response to events, and then only if they inherit their reactivity from the
-signal function(s) which are inputs to combinator. The rest do not
-react to or modify the input in any way, except to re-arrange it, copy it, or
-discard it altogether.
-
-Reactivity is introduced by means of the {\tt switch} combinator.
-The design of this combinator
-allows modular packaging of reactivity. A signal function can determine
-autonomously when to replace itself, based only on its input and state,
-by emitting an event occurrence carrying its replacement. The combinator
-consumes and hides the event carrying the replacement signal function, 
-so the reactivity is not exposed by the resulting reactive signal function.
-
-It is often useful for a signal function to receive a portion of its
-own output as input. This is especially useful when we have two
-signal functions which we would like to mutually interact. We cannot
-just serially compose them, we must also bring the output of the second
-back around to the first. Many signal-processing algorithms also depend
-on feedback. This ability is provided by the {\tt loop} operator.
-
-This combinator provides decoupling for signals
-(the input signal is the output signal at the previous time-step)
-but not events (event occurrences are supplied to the combinator immediately).
-This means that the programmer has the responsibility to ensure that feedback
-does not generate an infinite sequence of events in a single time-step.
-
-The {\tt filter} and {\tt filterList} combinators are stateless signal functions
-which apply a function to each event occurrence. The {\tt filter} combinator
-expects an option-typed output, while the {\tt list} combinator expects a list
-output. If the function yields a non-empty output, one or more output occurrences
-are produced with the elements of that output. If the output is empty, there
-are no output occurrences.
-
-The {\tt accumulate} and {\tt accumulateList} functions are similar, except that
-they maintain a state value which is replaced after each input event occurrence
-by an additional output from the function, and supplied to the function in
-addition to the value of each event occurrence.
-
-The joining combinators provide the ability to combine two event
-streams, two signals, or a signal and an event stream.
-
-The {\tt union} combinator is a non-deterministic merge of two event streams.
-Simultaneity of events is only observable up to the sampling rate, but events
-can occur in between sampling intervals. Thus, if two events occur
-simultaneously, one occurrence is arbitrarily selected to be first in the output
-stream.
-
-The {\tt combineSignals} combinator applies a binary function pointwise to two
-signals, and produces the result of this application as a third signal. The
-combining function is necessary because we will have two input samples at each
-time step, and must produce exactly one output sample.
-
-The {\tt capture} combinator adds the last-sampled value of a signal at the time
-of an event occurrence to that event occurrence.
-
-These three combinators together provide the ability to combine elements of
-signal vectors. By combining these combinators, along with the {\tt cancelLeft}
-and {\tt cancelRight} routing combinators, arbitrary signal vectors can be
-reduced.
-
-A set of combinators are provided for making use of time-dependence in a signal
-function. These combinators allow the modification of signals and events with
-respect to time, and the observation of the current time value.
-
-The simplest time-dependent combinator is {\tt time}, which simply outputs
-the time since it began to be evaluated. This does not necessarily correspond to
-the time since the global signal function began to be evaluated, since the
-signal function in which the {\tt time} combinator is used may have been
-introduced through {\tt switch}.
-
-The {\tt delay} signal function allows events to be delayed. An initial delay
-time is given, and event occurrences on the right input can carry a new delay
-time. Event occurrences on the left input are stored and output when their delay
-time has passed. Changing the delay time does not affect occurrences already
-waiting.
-
-The {\tt integrate} combinator outputs the rectangle-rule approximation of the
-integral of its input signal with respect to time.
+The full interface is described in the API documentation~\cite{TimefliesAPI}.
 
@@
 update inputs, and step time, as well as actions lifted from the underlying
-monad (used to obtain these inputs). An action in the underlying monad
-which produces the result and a new state is obtained with the {\tt runSFEvalT}
-function.
+monad (used to obtain these inputs).
+
+{\tt SFEvalT} actions are run using a supplied function which takes a signal
+function's evaluation state and an {\tt SFEvalT} action and provides an action
+in its underlying monad. This action includes the actions to actuate the outputs,
+and returns the new evaluation state, along with the return value of the
+{\tt SFEvalT} here.
+
+Using this interface, the state can be stored in a mutable reference, or in a
+closure. Handlers can be registered with the IO interface which push events, and
+the actuation function can be constructed from the primitives given by the IO
+library for actuation.

### 2012-11-30 Docs/IFL2012/discussion.tex
Performance results in discussion section - graphic not rendering properly

@@
 input and output types.
+
+We have described the implementation of a push-pull signal-function FRP system.
+Push-based event evaluation should have advantages in event latency and overall
+performance over pull-based event evaluation. We would also like to compare
+the evaluation interface in the current widely-used system with the one proposed
+here, again in order to verify that the theoretical benefits do appear in
+practice.
+
+One way to evaluate this is to write an application in two systems, and benchmark
+these systems for comparison. In order to compare TimeFlies with Yampa, the most
+recently released pull-based signal-function FRP system, we decided to implement
+a suitable application in both.
+
+\subsection{Evaluation Application}
+\label{subsection:Evaluation_and_Comparisons-Evaluation_Application}
+
+The application of choice is a learning switch controller for OpenFlow~\cite{OpenflowSpec}
+switches. This application must handle input events triggered by switches
+signalling that they have received a packet for which they have no handling
+rules, and produce output events which install handling rules for dealing with
+these packets.
+
+The reactive algorithm is simple. A table from machine addresses to switch ports
+is maintained. This table is updated in two ways. When a packet is sent to the
+controller, meta-data for the packet indicates which port it arrived on. The
+source address of the packet can then be mapped to that port in the table.
+Once this occurs, rules are installed for each packet with a destination address
+matching the source, along with reciprocal rules.
+
+Each mapping in the table has a timestamp. After a certain amount of time, the
+table entry is deleted. The table entry is also deleted if a packet is received
+with the same source address and a different port. When an entry is deleted,
+the switch is signaled to remove all corresponding rules.
+
+In both Yampa and TimeFlies, this algorithm is implemented by using an accumulator,
+which applies functions carried by event occurrences to a state value. It stores
+the new state and outputs new event occurrences based on the output of the function.
+
+The state stored is the table (implemented using the {\tt Data.Map} module in the
+Haskell {\tt containers} library) and the output event occurrences carry messages
+to be sent to the switch. The events are sourced internally to the learning
+switch signal function (using the {\tt after} combinator in TimeFlies and the {\tt repeatedly}
+function in Yampa) to delete expired rules, and externally, by transforming messages
+from the switch to the controller into accumulator functions.
+
+\subsection{Methodology and Results}
+\label{subsection:Evaluation_and_Comparisons-Methodology_and_Results}
+
+The controllers were benchmarked using the {\tt cbench} program~\cite{cbench}. The
+amount to time to pause between time and signal updates (the inverse of the sampling rate for signals)
+was passed as a parameter to both programs. The {\tt cbench} benchmark operates by sending a simulated packet in
+event, waiting for a rule response, and then repeating. The output of the benchmark
+is the number of times this occurs in a second. This is thus a very good test of
+event latency in a reactive system.
+
+Figure \ref{fig:timeflies-yampa-comparison} shows the performance comparison.
+Toward the left of the graph is lower pause times (more frequent sampling). The
+y-axis is a higher rate of responses (higher numbers are preferred).
+
+\begin{figure}
+\epsfig{file=graph.eps,angle=-90,width=\linewidth}
+\hrule
+\caption{Comparison of timeflies vs. Yampa implementing an OpenFlow learning switch controller.}
+\label{fig:timeflies-yampa-comparison}
+\end{figure}
+
+There is some correlation between the sampling rate and the response rate
+for TimeFlies, which is due to the continued wrapping of a replacement signal
+function by {\tt switch} until the next time step (see Section~\ref{subsection:Implementation-Signal_Functions-Implementation_of_Signal_Function_Combinators}).
+Nevertheless, TimeFlies outperforms Yampa in event response by several orders of magnitude,
+even at a very high sampling rate. This is due to two factors. First, TimeFlies can react
+to individual events without evaluating the whole signal function. Second, the
+interface of time-flies permits us to connect its evaluation directly to GHC's event
+manager, while the design of Yampa requires us to use threads and synchronized
+communication\footnote{Yampa does include a step-wise evaluation function, but
+it still couples event evaluation firmly to time steps, and is not well
+documented.}.

### 2012-11-30 Docs/IFL2012/implementation.tex
ifl draft edits

--- 2012-11-28 Docs/IFL2012/implementation.tex
@@
 The design of signal functions specifies a family of types for the inputs and
-outputs of signal functions. Signal functions are not functions in the purest
-sense, however. They are not mappings from a single instance of their input
-type to a single instance of their output type. They must be implemented with
-respect to the temporal semantics of their inputs and outputs.
+outputs of signal functions. ``Signal functions'' are not simple functions,
+rather, they are transformers of signals and events. They are reactive, that is,
+they can alter themselves in response to their inputs.
 
-We therefore start by creating a set of concrete datatypes for the inputs and
-outputs of signal functions. These datatypes will be parameterized by the input
-and output types of the signal function, and will not be exposed to the user of
+We start by creating a set of concrete datatypes for the inputs and outputs of
+signal functions. These datatypes will be parameterized by the input and output
+signal vectors of the signal function, and will not be exposed to the user of
 the library. Rather, they will specify how data is represented during the
-temporal evaluation of signal functions. We then implement signal functions
+evaluation of signal functions. We then implement signal functions
 as records of functions from these concrete types to these concrete types paired
@@
 In Section~\ref{section:System_Design_and_Interface} we presented signal vectors
-as a set of types. In order to be completely principled, we should isolate these
-types into their own {\em kind} (a sort of type of types); however, the Haskell
-extension for this was far from stable at the time this system was created.
+as a set of types. \footnote{In order to be completely principled, we should
+isolate these types into their own {\em kind} (a sort of type of types);
+however, the Haskell extension for this was far from stable at the time this
+system was created.} Using the Glasgow Haskell Compiler's extension for
+Generalized Algebraic Datatypes~\cite{Cheney2003,Xi2003,PeytonJones2006}, we
+specify three concrete datatypes which are parameterized over signal vectors,
+and represent information about a temporal point during evaluation.
 
-Using the Glasgow Haskell Compiler's extension for Generalized Algebraic
-Datatypes~\cite{Cheney2003,Xi2003,PeytonJones2006}, we specify three concrete
-datatypes which are parameterized over signal vectors, and represent
-information about one temporal point in a signal function's evaluation.
-
-The first type carries the instantaneous values of all signals in a signal vector.
-There is precisely one constructor for each type constructor of a signal vector ({\tt SVEmpty},
-{\tt SVSignal}, {\tt SVEvent}, and {\tt SVAppend}). The constructor for {\tt SVSignal}
-carries a value typed with the type parameter of the {\tt SVSignal} constructor,
-the rest do not. Thus, in this representation, for any signal vector, there is
-a node in the representation for each node in the signal vector, and precisely
-the signal nodes have values.
+The first type carries the instantaneous values of all signals in a signal
+vector. There is precisely one constructor for each type constructor of a signal
+vector ({\tt SVEmpty}, {\tt SVSignal}, {\tt SVEvent}, and {\tt SVAppend}). The
+constructor for {\tt SVSignal} carries a value typed with the type parameter of
+the {\tt SVSignal} constructor, the rest do not. Thus, in this representation,
+for any signal vector, there is a node in the representation for each node in
+the signal vector, and precisely the signal nodes have values.
 
@@
 representation is a path from the root of a signal vector to an event node,
-carrying exactly one value for the event.
+carrying exactly one value for the event. For instance, for an input occurrence
+to the signal function:
+
+\begin{code}
+example :: (SVEvent Int :^: (SVEvent String :^: SVSignal Float))
+\end{code}
+
+might be:
+
+\begin{code}
+SVEventRight (SVEventLeft (SVEventOcc "Example"))
+\end{code}
 

### 2012-11-30 Docs/IFL2012/introduction.tex
ifl draft edits

@@
 constructing only part of the input (for instance, an occurrence on one of the
-event inputs). One signal function system~\cite{Jeffery2012} does permit the
+event inputs). One signal function system~\cite{Jeffrey2012} does permit the
 user to push {\em segments} of signals, but, like the others, does not
@@
 between events and signals is achieved. Section~\ref{section:Discussion} is a
-discussion of the \textred{usefulness} \rn{more specific?} of our implementation. 
-Section~\ref{section:Ongoing_and_Further_Work} describes the current and future
-work on this system. Section~\ref{section:Related_Work} gives an overview of
+discussion of the performance of our implementation in the implementation of a
+network application. Section~\ref{section:Related_Work} gives an overview of
 related efforts. Section~\ref{section:Conclusion} concludes.

### 2012-11-30 Docs/IFL2012/introduction.tex
IFL draft update

@@
 
-\todo{TODO: it would be good to give SOME sense for how an FRP
-  evaluation proceeds (e.g. a strawman) before getting into the next para.}
+An FRP program is generally evaluated by supplying a behavior or signal function
+to an evaluation function, which supplies input (in the case of a signal
+function) and actuates its outputs as they are produced. For instance, if a
+video game is implemented in a signal-function FRP system, the evaluation
+function is responsible for obtaining keyboard and mouse inputs and supplying
+them to the signal function, and actuating the outputs of the signal function
+as graphics on the monitor and sound through the speakers.
 

### 2012-11-30 Docs/IFL2012/discussion.tex
IFL draft update
IFL paper draft update

@@
 
-The reactive algorithm is simple. A table from machine addresses to switch ports
-is maintained. This table is updated in two ways. When a packet is sent to the
-controller, meta-data for the packet indicates which port it arrived on. The
-source address of the packet can then be mapped to that port in the table.
-Once this occurs, rules are installed for each packet with a destination address
-matching the source, along with reciprocal rules.
+The reactive algorithm maintains a table of machine addresses to switch ports.
+This table is updated in two ways. When a packet is sent to the controller,
+meta-data for the packet indicates which port it arrived on. The source address
+of the packet can then be mapped to that port in the table. Once this occurs,
+rules are installed for each packet with a destination address matching the
+source, along with reciprocal rules.
 
@@
 \begin{figure}
-\epsfig{file=graph.eps,angle=-90,width=\linewidth}
+\includegraphics[width=0.7\linewidth,angle=-90]{graph}
 \hrule
-\caption{Comparison of timeflies vs. Yampa implementing an OpenFlow learning switch controller.}
+\caption{Comparison of TimeFlies vs. Yampa implementing an OpenFlow learning switch controller.}
 \label{fig:timeflies-yampa-comparison}
@@
 for TimeFlies, which is due to the continued wrapping of a replacement signal
-function by {\tt switch} until the next time step (see Section~\ref{subsection:Implementation-Signal_Functions-Implementation_of_Signal_Function_Combinators}).
+function by {\tt switch} until the next time step (see Section~\ref{subsection:Implementation-Signal_Functions}).
 Nevertheless, TimeFlies outperforms Yampa in event response by several orders of magnitude,

### 2012-11-30 Docs/IFL2012/introduction.tex
IFL paper draft update

@@
 
-FRP {\em combines} behaviors and events through the use of {\em switching}, in which
+FRP combines behaviors and events through the use of {\em switching}, in which
 a behavior (in classic FRP) or a signal function (in signal-function FRP) is

### 2012-11-30 Docs/IFL2012/introduction.tex
Add weather example to intro

@@
 
-An FRP system provides a means of manipulating {\em behaviors} and
-{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
-but the definition is the same.  Semantically, a behavior or signal is a
+For example, consider an application for monitoring thunderstorms. Such an
+application might be employed by a sports venue, or a manufacturing center with
+sensitive equipment. The application receives input from several sources,
+including barometers (pressure), anemometers (wind speed and direction), rain
+gauges, and lightning detectors. The output of the application is a pictoral map
+with information from the sensors displayed, as well as an overlay showing an
+estimate of the sizes and motions of storms.
+
+An FRP system provides a means of manipulating {\em signals} and
+{\em events}. Signals are often referred to as {\em behaviors} in FRP literature,
+but the definition is the same.  Semantically, a signal or behavior is a
 function from time to a value. An event is a ordered sequence of discrete
-occurrences, carrying both values and timestamps.
+occurrences, carrying both values and timestamps. In our example, inputs from
+the barometers, anemometers, and rain gauges would be abstracted as signals. The
+information carried by the signal would be the measurement made by the
+instrument, perhaps tagged with the instrument's location. The input from the
+lighting detector would be an event, whose occurrences would each carry the
+location and intensity of a lightning strike.
+
+One key distinction in FRP systems concerns how signals and events are
+evaluated. There are two possibilities. Pull-based evaluation uses repeated and
+frequent updates in a loop to check for updated values (similar to busy waiting
+in concurrency, except that for signals, there should be no waiting). Push-based
+evaluation blocks or waits to be notified about an updated value, similar to
+notification in concurrency. Pull-based evaluation is appropriate for signals,
+which approximate continuous values and should have a value available whenever
+they are sampled. Push-based evaluation is appropriate for events, which must be
+reacted to quickly without consuming unnecessary resources to wait. Turning
+again to our example: it is appropriate for our system to repeatedly query the
+pressure, wind speed, and rain levels. The system should respond quickly to
+lightning strikes, but should not use resources in repeatedly sampling for the
+occurrence, or not, of a lightning strike. The system should instead be notified
+when a strike has occurred.
 
@@
 to an evaluation function, which supplies input (in the case of a signal
-function) and actuates its outputs as they are produced. For instance, if a
-video game is implemented in a signal-function FRP system, the evaluation
-function is responsible for obtaining keyboard and mouse inputs and supplying
-them to the signal function, and actuating the outputs of the signal function
-as graphics on the monitor and sound through the speakers.
+function) and actuates its outputs as they are produced. For our weather system
+example, the evaluation function is responsible for sampling the sensors,
+receiving notification from the lightning strike detectors and pushing the
+occurrences to the signal function, and drawing the picture which is the output
+signal of the signal function.
 
@@
 to efficiency is to separate the evaluation of behaviors and events, since
-suitable strategies give best performance in each case. Push-based evaluation
-evaluates a system only when input is available, and is thus suitable for
-discrete inputs such as events. Pull-based evaluation evaluates the system as
-quickly as possible, polling for input, and is preferable for behaviors and
-signals. The initial implementations of FRP made use of pull-based evaluation
-for both behaviors and events. ``Reactive''~\cite{Elliott2009}, as well as more
-recent systems such as ``reactive-banana''~\cite{Apfelmus}, make use of
-push-based evaluation for events and pull-based evaluation for behaviors. This
-is known as ``push-pull'' evaluation.
+as was discussed above, the suitable strategy for evaluating signals is not the
+same as that for events. The initial implementations of FRP made use of
+pull-based evaluation for both behaviors and events.
+``Reactive''~\cite{Elliott2009}, as well as more recent systems such as
+``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
+events and pull-based evaluation for behaviors. This is known as ``push-pull''
+evaluation.
 
-Almost all implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
-have used pull-based evaluation for both signals and events. This is due to the
-ease of implementation of pull-based evaluation, and because the types used for
-signal functions do not permit distinguishing signals and events, or
-constructing only part of the input (for instance, an occurrence on one of the
-event inputs). One signal function system~\cite{Jeffrey2012} does permit the
-user to push {\em segments} of signals, but, like the others, does not
-distinguish events in a way that permits them to be separately pushed.
+Almost all implementations of signal-function FRP to
+date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011} have used
+pull-based evaluation for both signals and events. This is due to the ease of
+implementation of pull-based evaluation, and because the types used for signal
+functions do not permit distinguishing signals and events, or constructing only
+part of the input (for instance, an occurrence on one of the event inputs). One
+signal function system~\cite{Jeffrey2012} does permit the user to push
+{\em segments} of signals, but, like the others, does not distinguish events
+from signals in a way that permits events to be separately pushed.
 
@@
 may contain multiple values, including option values for events. Signal vectors
-are uninhabited types, which can be used to describe partial or full representations
-of the signal function inputs and outputs.
+are uninhabited types, which can be used to describe partial or full
+representationsof the signal function inputs and outputs.
 
@@
 is a push-pull signal-function FRP system, implemented in the Haskell
-programming language. We show that typing signal-function FRP with signal
-vectors allows push-pull evaluation of signal functions. We demonstrate an
-evaluation interface that allows the user of TimeFlies to take advantage of this
-behavior in interfacing TimeFlies to a wide range of IO systems. Finally, we
-give benchmark results which demonstrate that the push-pull evaluation strategy
-makes TimeFlies more efficient than the current pull-based implementation of
-signal-function FRP.
+programming language. Our contributions with this system are as follows:
 
-Section~\ref{section:System_Design_and_Interface} describes design choices for the system,
-and provides an overview of the interface. Section~\ref{section:Implementation}
-describes how the system is implemented, and how the separation of evaluation
-between events and signals is achieved. Section~\ref{section:Discussion} is a
-discussion of the performance of our implementation in the implementation of a
-network application. Section~\ref{section:Related_Work} gives an overview of
+\begin{itemize}
+  \item We show that typing signal-function FRP with signal vectors allows
+        push-pull evaluation of signal functions.
+  \item We demonstrate a programmable evaluation interface that allows the user
+        of TimeFlies to take advantage of this behavior while interfacing
+        TimeFlies to a wide range of IO systems.
+  \item We give benchmark results which demonstrate that the push-pull
+        evaluation strategy makes TimeFlies more efficient than the current
+        pull-based implementation of signal-function FRP.
+\end{itemize}
+
+Section~\ref{section:System_Design_and_Interface} describes design choices for
+the system, and provides an overview of the interface.
+Section~\ref{section:Implementation} describes how the system is implemented,
+and how the separation of evaluation between events and signals is achieved.
+Section~\ref{section:Discussion} is a discussion of the performance of our
+implementation in the implementation of a network application. Section~\ref{section:Related_Work} gives an overview of
 related efforts. Section~\ref{section:Conclusion} concludes.

### 2012-11-30 Docs/IFL2012/introduction.tex
Partial pass on intro

@@
 An FRP system provides a means of manipulating {\em signals} and
-{\em events}. Signals are often referred to as {\em behaviors} in FRP literature,
+{\em events}. Signals are often referred to as {\em behaviors} in earlier FRP literature,
 but the definition is the same.  Semantically, a signal or behavior is a
 function from time to a value. An event is a ordered sequence of discrete
-occurrences, carrying both values and timestamps. In our example, inputs from
-the barometers, anemometers, and rain gauges would be abstracted as signals. The
+occurrences, \textred{carrying both values and timestamps}. In our example, inputs from
+the barometers, anemometers, and rain gauges would be signals. The
 information carried by the signal would be the measurement made by the
@@
 evaluated. There are two possibilities. Pull-based evaluation uses repeated and
-frequent updates in a loop to check for updated values (similar to busy waiting
-in concurrency, except that for signals, there should be no waiting). Push-based
-evaluation blocks or waits to be notified about an updated value, similar to
-notification in concurrency. Pull-based evaluation is appropriate for signals,
+frequent updates in a loop to check for updated values (polling).
+% (similar to busy waiting in concurrency, except that for signals, there should be no waiting). 
+\rn{``In concurrency'' seems awkward.  Maybe ``in concurrent
+  programming'', or just nix.  What do you think of this alternative?}
+Push-based
+evaluation blocks or waits to be notified about an updated value (blocking).
+% , similar to notification in concurrency. 
+Pull-based evaluation is appropriate for signals,
 which approximate continuous values and should have a value available whenever
@@
 again to our example: it is appropriate for our system to repeatedly query the
-pressure, wind speed, and rain levels. The system should respond quickly to
+pressure, wind speed, and rain levels \new{to update the real-time model}. The system should respond quickly to
 lightning strikes, but should not use resources in repeatedly sampling for the
-occurrence, or not, of a lightning strike. The system should instead be notified
-when a strike has occurred.
+occurrence (or not) of a lightning strike.  The system should instead be notified
+when a strike has occurred.  \new{Further, this notification should {\em not} result in 
+recomputeing all signals upon each strike event.}
 

### 2012-11-30 Docs/IFL2012/introduction.tex
minor.. merge paras, bit of new text

@@
 occurrence (or not) of a lightning strike.  The system should instead be notified
-when a strike has occurred.  \new{Further, this notification should {\em not} result in 
-recomputeing all signals upon each strike event.}
+when a strike has occurred, \new{and that notification should {\em not} result in 
+recomputing all signals upon each strike event.}
+
+\rn{Shall we rename {\em behavior} to {\em signal} below this point?}
 
@@
 reactive transformers of signals, events, or combinations of signals and events.
-
+%
 FRP combines behaviors and events through the use of {\em switching}, in which
@@
 are uninhabited types, which can be used to describe partial or full
-representationsof the signal function inputs and outputs.
+representations of the signal function inputs and outputs.
 

### 2012-11-30 Docs/IFL2012/introduction.tex
more intro

@@
 An FRP program is generally evaluated by supplying a behavior or signal function
-to an evaluation function, which supplies input (in the case of a signal
+to an evaluation function, which provides input (in the case of a signal
 function) and actuates its outputs as they are produced. For our weather system
@@
 Classic FRP was first described as a system for interactive animations~\cite{Elliott1997}.
-Recent work on classic FRP has focused on efficient implementation. One approach
-to efficiency is to separate the evaluation of behaviors and events, since
-as was discussed above, the suitable strategy for evaluating signals is not the
-same as that for events. The initial implementations of FRP made use of
+%% Recent work on classic FRP has focused on efficient implementation. One approach
+%% to efficiency is to separate the evaluation of behaviors and events, since
+%% as was discussed above, the suitable strategy for evaluating signals is not the
+%% same as that for events. 
+\rn{Let's not be non-specific: ``focus on efficient implementation.
+  One approach...'', if there aren't actually other examples of
+  ``efficient implementation'' and we're really just talking about push-pull.}
+%%
+The initial implementations of FRP made use of
 pull-based evaluation for both behaviors and events.
@@
 events and pull-based evaluation for behaviors. This is known as ``push-pull''
-evaluation.
+evaluation \new{and is asymptotically more efficient}.
 
@@
 the push-based evaluation of events in a signal-function FRP system. The notion
-of signal vectors allows the representation of signal function inputs and
-outputs as combinations of signals and events, rather than a single signal which
-may contain multiple values, including option values for events. Signal vectors
+of {\em signal vectors} allows the \new{(type-level)} representation of signal-function inputs and
+outputs as combinations of signals and events, rather than a single
+signal \new{of composite} values, including option values for events. Signal vectors
 are uninhabited types, which can be used to describe partial or full
 representations of the signal function inputs and outputs.
+\rn{But what's the limitation of N-Ary FRP?! It doesn't allow full
+  push-pull right?  Or wait, the SV technique ``enables'' push-pull,
+  but the previous sculthorpe implementation just didn't do it?  Clarify.}
 
+\new{In this paper we introduce}
 TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
 favorite quotation of one of the author's philosophy instructors, used to
-demonstrate the ambiguity of language. The origin of the quotation is unknown.}
-is a push-pull signal-function FRP system, implemented in the Haskell
-programming language. Our contributions with this system are as follows:
+demonstrate the ambiguity of language. The origin of the quotation is unknown.},
+a push-pull signal-function FRP system, implemented in the Haskell
+programming language. Our contributions are as follows:
+% \rn{Should be the paper contributions, not the system!}
 
@@
   \item We show that typing signal-function FRP with signal vectors allows
-        push-pull evaluation of signal functions.
+        push-pull evaluation of signal functions. \rn{FIRST?}
   \item We demonstrate a programmable evaluation interface that allows the user

### 2013-01-11 Docs/Thesis/example.tex
Add example chapter

\chapter{Example Application}
\label{chapter:Example_Application}

TimeFlies is a library for Functional Reactive Programming, which is a paradigm
for creating interactive and time-dependent applications. This chapter presents
the design of one such application, and its implementation using TimeFlies. The
application is an OpenFlow controller which implements a learning switch. In
short, it is a re-implementation of the standard kind of switch used in local
area networks, using ``software-defined networking.'' This
is the application which is benchmarked for performance comparisons in
Chapter~\ref{chapter:Evaluation_and_Comparisons}.

\section{OpenFlow}
\label{section:Example_Application-OpenFlow}

The OpenFlow protocol~\cite{OpenflowSpec} is a protocol for software-defined
networking applications. In particular, it defines the communication between
switches (devices which quickly route packets from input ports to output ports
based on learned rules) and {\em controllers}, which are generally devices with
large computational resources, such as servers. OpenFlow allows switches to
report packets for which no rule exists to a controller, and provides a means
for the controller to install new rules on a switch either preemptively, or in
response to a reported packet.

One of the simplests tasks which may be implemented as a OpenFlow controller is
a ``learning switch''. Such a switch uses the source and destination hardware
addresses in network packets to make routing decisions. A table is kept which
records the ports where source addresses are observed on incoming packets. This
table thus contains knowledge of which hardware address(es) can be reached on
which port. Using this table, rules are constructed to route packets with
particular source-destination pairs to the correct port. When a packet is seen
for which no rule exists, it is reported to the controller, which updates the
table and rules based on the new knowledge, and broadcasts the packet so that
it can reach its intended destination and receive a response.

Our example application is a controller for a ``learning switch.'' We describe
its implementation as a TimeFlies signal function, and two approaches to running
this signal function using the TimeFlies evaluation interface.

\section{Implementation}
\label{section:Example_Application-Implementation}

The first component for our learning switch is the table which maps addresses
to ports. This is a stateful data structure which will be updated by input
events and possibly produce output events. We employ the {\tt accumulateList}
combinator to produce the signal function:

\begin{figure}
\begin{code}
-- | Function type to modify a table and produce messages to the switch
type TableAccumulator = 
     SwitchTable 
  -> ([(SwitchHandle EthernetFrame,
        TransactionID,
        CSMessage)],
      SwitchTable)

-- | Empty map
M.empty :: M.Map k v

-- | Reverse application
rapp :: a -> (a -> b) -> b
rapp x f = f x

-- | Accumulate a switch table, producing output messages
--   based on the incoming functions
switchTable ::    SVEvent TableAccumulator
              :~> SVEvent (SwitchHandle EthernetFrame,
                           TransactionID,
                           CSMessage)
switchTable = accumulateList rapp M.empty
\end{code}
\hrule
\caption{Signal function for switch table.}
\label{figure:switch_table_sf}
\end{figure}

Note that the input events are closures which expect the table as input and
produce an updated table as output. This enables us to write several different
event sources whose final events require state (the table) and produce
messages to the switch.

The first source of such events is packets forwarded to the controller by
switches. The closures carried by these events implement the response to
packet inputs. The signal function from input switch message events to events
from the accumulator is shown in Figure~\ref{figure:packet_in_sf}, but the
details of the switch routing algorithm are elided.

\begin{figure}
\begin{code}
getPacketIn ::     SVEvent (SwitchHandle EthernetFrame,
                            TransactionID,
                            SCMessage EthernetFrame)
               :~> SVEvent (SwitchHandle EthernetFrame,
                            TransactionID,
                            PacketInfo EthernetFrame)

-- | Handle a packet in event
handlePacketIn ::    ((SwitchHandle EthernetFrame,
                       TransactionID,
                       PacketInfo EthernetFrame), Double)
                  -> TableAccumulator

packetIn ::     SVEvent ((SwitchHandle EthernetFrame,
                          TransactionID,
                          PacketInfo EthernetFrame),
                         Double)
            :~> SVEvent TableAccumulator
packetIn = pureEventTransformer handlePacketIn

-- | Capture the occurrence time of an event.
captureTime :: SVEvent a :~> SVEvent (a, Double)
captureTime = uncancelLeft >>>
              first time >>>
              capture

-- | Build TableAccumulator events from input packets.
handleSCMessage ::     SVEvent (SwitchHandle EthernetFrame,
                                TransactionID,
                                SCMessage EthernetFrame)
                   :~> SVEvent TableAccumulator
handleSCMessage = getPacketIn >>> captureTime >>> packetIn

\end{code}
\hrule
\caption{Signal function handling incoming packets.}
\label{figure:packet_in_sf}
\end{figure}

The {\tt captureTime} signal function attaches the time to incoming packets. The
{\tt getPacketIn} signal function extracts packet-in messages from the variety
of message events that a switch may send to the controller. These two signal
functions feed into the {\tt packetIn} signal function, which builds the
closures for the accumulator using the {\tt handlePacketIn} pure function.

The other task is to periodically eliminate all rules which are older than a
specified threshold. For this purpose, we construct another signal function
which produces {\tt TableAccumulator} events. This signal function runs a timer
(the {\tt every} combinator) and produces events carrying closures which scan
the table for old rules, delete them, and produces messages to inform the
switches of their deletion. This signal function is shown in Figure~\ref{figure:clean_sf},
again eliding the implementation of the closures.

\begin{figure}
\begin{code}
-- | Take a function-value pair and return the application.
pairApp :: (a -> b, a) -> b
pairApp (f, x) = f x

-- | Alter a table and generate messages to remove expired entries and rules
cleanTable :: Double -> Double -> TableAccumulator

-- | Produce an event with the rule-cleaning accumulator.
cleanRules :: Double -> SVEmpty :~> SVEvent TableAccumulator
cleanRules t = copy >>>
               second (every t cleanRulesAcc) >>>
               first time >>>
               capture >>>
               pureEventTransformer pairApp
  where 
    cleanRulesAcc :: Double -> TableAccumulator
    cleanRulesAcc nowT table = cleanTable t nowT table
\end{code}
\hrule
\caption{Table-cleaning signal function.}
\label{figure:clean_sf}
\end{figure}

We now have all of the pieces for a signal function implementing a learning
switch, as shown in Figure~\ref{figure:learn_sf}.

\begin{figure}
\begin{code}
-- | A constant, for how long rules may survive
tableTime :: Double

-- | Make IO actions to send each packet.
sendCSMessage ::     SVEvent (SwitchHandle EthernetFrame,
                              TransactionID,
                              CSMessage)
                 :~> SVEvent (IO ())

-- | The learning switch
learn ::     SVEvent (SwitchHandle EthernetFrame,
                      TransactionID,
                      SCMessage EthernetFrame)
         :~> SVEvent (IO ())
learn = uncancelLeft >>> 
        first (cleanRules tableTime) >>>
        second handleSCMessage >>>
        union >>>
        switchTable >>>
        sendCSMessage
\end{code}
\hrule
\caption{The signal function for a learning switch.}
\label{figure:learn_sf}
\end{figure}

The learning switch composes the {\tt cleanRules} and {\tt handleSCMessage}
signal functions in parallel, but routes input only to {\tt handleSCMessage},
by using the {\tt uncancelLeft} routing combinator. The event streams are
merged and fed to the {\tt switchTable} signal function, and the resulting
packets are fed to the {\tt sendCSMessage} signal function which constructs
{\tt IO ()} events.

Note that this approach enables a modular design where each task is implemented
in its own signal function. Additional behaviors could be coded as additional
signal functions, whose events were also routed to the table accumulator.

Finally, we show how the signal function is interfaced with IO code using the
evaluation interface. First, we need a few extra definitions. For starters,
a controller must also handle switches when they first connect. We create a
separate signal function for this purpose, and compose it in parallel with our
learning switch signal function. We also need a definition of which port to
listen on, and an IO action for reading the time. These are shown in
Figure~\ref{figure:example_io_prelim}.

\begin{figure}
\begin{code}

-- | The port for the server to listen on.
port :: Word16

-- | Get the current time from the monotonic timer
getMonoTimeAsDouble :: IO Double

-- | Handle the connection of a new switch          
switchConnect ::     SVEvent (SwitchHandle EthernetFrame)
                 :~> SVEvent (IO ())
switchConnect = pureEventTransformer (void . handshake)

-- | Run switch connection and learning concurrently
learningSwitch ::     (SVEvent (SwitchHandle EthernetFrame) :^:
                       SVEvent (SwitchHandle EthernetFrame,
                                TransactionID,
                                SCMessage EthernetFrame))
                  :~> SVEvent (IO ())
learningSwitch = first switchConnect >>> second learn >>> union
\end{code}
\hrule
\caption{Evaluation preliminaries.}
\label{figure:example_io_prelim}
\end{figure}

Now we can create the main procedure for our learning switch. Note that
the code for placing an event on the input of a signal function is a simple
monadic sequence, sandwich between {\tt IORef} reads and writes of an opaque
value (Figure~\ref{figure:example_io}).

\begin{figure}
\begin{code}
-- | Entry point
main :: IO ()
main = do (argS:_) <- getArgs
          let sampleTime = read argS
          time <- getMonoTimeAsDouble
          reactiveRef <- newIORef $ initSFEval 
                                      (eventHandler id)
                                      (combineSamples sampleEvt sampleEvt)
                                      time
                                      learningSwitch
          ofEventManager <- 
            openFlowEventManager
              Nothing
              port 
              -- Switch addition inputs
              (\handle -> do rState <- readIORef reactiveRef
                             ((), rState') <- runSFEvalT 
                                                (push $ svLeft $ svOcc handle)
                                                rState
                             writeIORef reactiveRef rState')
              -- Switch message inputs:
              (\((tid, msg), handle) -> 
                 do rState <- readIORef reactiveRef
                              ((), rState') <- runSFEvalT 
                                                 (push $ svRight $
                                                  svOcc (handle, tid, msg))
                                                 rState
                              writeIORef reactiveRef rState')
          let evtMgr = getEventManager ofEventManager
              -- Sampling:
              sample = void $ registerTimeout evtMgr sampleTime $ 
                         do rState <- readIORef reactiveRef
                            time <- getMonoTimeAsDouble
                            ((), rState') <- runSFEvalT (step time) rState
                            writeIORef reactiveRef rState'
                            sample
          sample 
          loop $ evtMgr      
\end{code}
\hrule
\caption{Using the evaluation interface for IO.}
\label{figure:example_io}
\end{figure}

### 2013-01-11 Docs/Thesis/background.tex
Take care of revisions from mtf

@@
 switcher b [] t = b t
-switcher b [(to, bo)] t = if t < to
+switcher b ((to, bo):_) t = if t < to
                             then b t
@@
 Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs.
-This library utilized the abstraction of Arrows~\cite{Hughes2000} to structure
-{\em signal functions}. Arrows are abstract type constructors with input and
+This library utilized the concept of Arrows~\cite{Hughes2000} as an abstraction
+for {\em signal functions}. Arrows are abstract type constructors with input and
 output type parameters, together with a set of routing combinators (see Figure~\ref{figure:arrow_combinators}).
-Signal functions form an arrow. Signal functions are the first-class abstraction
-in this approach. They represent time-dependent and reactive transformers of
-events and signals, which are themselves not first class, since such values
+Signal functions form an arrow. Signal functions represent time-dependent and
+reactive transformers of events and signals. Signals and events
 cannot be directly manipulated by the programmer. This approach has two
@@
 consisted in using the dependent type system to prove the correctness of the FRP
-system discussed. This work introduced a signal vectors, a typing construct
+system discussed. This work introduced signal vectors, a typing construct
 which permits the distinction of signal and event types at the level of the FRP

### 2013-01-11 Docs/Thesis/comparison.tex
Thesis pdf update
Take care of revisions from mtf

@@
 the new state and outputs new event occurrences based on the output of the function.
+The accumulator has a simple implementation using {\tt switch}.
 
-The state stored is the table (implemented using the {\tt Data.Map} module in the
-Haskell {\tt containers} library) and the output event occurrences carry messages
-to be sent to the switch. The events are sourced internally to the learning
-switch signal function (using the {\tt after} combinator in TimeFlies and the {\tt repeatedly}
-function in Yampa) to delete expired rules, and externally, by transforming messages
-from the switch to the controller into accumulator functions.
+The state stored is the table (implemented using the {\tt Data.Map} module in
+the Haskell {\tt containers} library) and the output event occurrences carry
+messages to be sent to the switch. The events are sourced internally to the
+learning switch signal function (using the {\tt after} combinator in TimeFlies
+and the {\tt repeatedly} function in Yampa) to delete expired rules, and
+externally, by transforming messages from the switch to the controller into
+accumulator functions. This implementation is shown in Figure~\ref{figure:openflow_implementation}.
+
+\begin{figure}
+\begin{code}
+learn ::     SVEvent (SwitchHandle EthernetFrame,
+                      TransactionID,
+                      SCMessage EthernetFrame)
+         :~> SVEvent (IO ())
+learn = uncancelLeft >>>
+        -- Timer to clean the table
+        first (cleanRules tableTime) >>>
+        -- Builds accumulator actions from input messages 
+        second handleSCMessage >>>
+        -- Bring together accumulator inputs     
+        union >>> 
+        -- The accumulator                       
+        switchTable >>>
+        -- Produce IO actions as event outputs                 
+        sendCSMessage
+
+cleanRules ::     Double
+               -> SVEmpty
+              :~> SVEvent TableAccumulator
+cleanRules t = copy >>> -- two empty inputs
+               -- Fire every t seconds
+               second (every t cleanRulesAcc) >>> 
+               -- capture the time each fires
+               first time >>>
+               capture >>>                        
+               pureEventTransformer (uncurry ($))
+  where 
+    cleanRulesAcc :: Double -> TableAccumulator
+    cleanRulesAcc nowT table = cleanTable t nowT table
+
+-- | Produce an event with the given value at the given time interval
+every :: Double -> a -> SVEmpty :~> SVEvent a
+every t x = let e = switch (after t x >>>
+                            copy >>>
+                            second (pureEventTransformer $ const e))
+            in e
+\end{code}
+\hrule
+\caption{Partial implementation of the OpenFlow switch using TimeFlies}.
+\label{figure:openflow_implementation}
+\end{figure}
 
@@
 \begin{figure}
-\epsfig{file=graph.eps,angle=-90,width=\linewidth}
+\includegraphics{graph}
 \hrule

### 2013-01-11 Docs/Thesis/introduction.tex
Take care of revisions from mtf

@@
 Functional reactive programming (FRP) is a promising class of abstractions for
-writing reactive programs in functional languages. The FRP abstractions are
+writing interactive programs in functional languages. The FRP abstractions are
 behaviors or signals, which are functions of continuous time, and events, which
@@
 literature on signal functions, but we provide a documented and improved
-interface for evaluating signal functions. A further advantage of signal
-function FRP programs is that they are more readily composed, since additional
-signal functions may be composed on the input or output side of a signal
-function, rather than simply the output side as in classic (non-signal-function)
-FRP systems.
+interface for evaluating signal functions. A further advantage of
+signal-function FRP programs is that they are more readily composed,
+since additional signal functions may be composed on the input or output side of
+a signal function, rather than simply the output side as in classic
+(non-signal-function) FRP systems.
 

### 2013-01-11 Docs/Thesis/system_design_and_interface.tex
Take care of revisions from mtf

@@
 
-In previous systems such as Yampa, events were understood (and implemented) as option-valued
-signals. This approach is undesirable for several reasons. The most pressing reason is
-that it prohibits push-based evaluation of events, because events are embedded in the
+In previous systems such as Yampa, events were understood (and implemented) as
+option-valued signals.\footnote{{\tt type Event a = Signal (Maybe a)}} This
+approach is undesirable for several reasons. The most pressing reason is that it
+prohibits push-based evaluation of events, because events are embedded in the
 samples of a signal and must be searched for when sampling.
@@
 ({\tt (>>>)}, {\tt first}, {\tt second}) are reactive (may replace themselves in
-response to events, and then only if they inherit their reactivity from the
+response to events), and then only if they inherit their reactivity from the
 signal function(s) which are inputs to combinator. The rest do not
@@
 There are other formulations of a reactive combinator found in FRP literature
-and libraries, which may be implemented using the one supplied. These are shown
-in Figure~\ref{figure:alternate_switching_combinators} and may be provided in a
-future version of the TimeFlies library.
+and libraries, which may be implemented using the one supplied. Some of these
+are shown in Figure~\ref{figure:alternate_switching_combinators} and may be
+provided in a future version of the TimeFlies library.
 
@@
   switch (sf >>> second (pureEventTransformer f))
-
--- Supplied version in terms of alternate version
-switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
-          -> svIn :~> svOut
-switch sf = switch_gen sf id
 
@@
 
+The goal of the evaluation interface is to provide a means to evaluate
+declaratively specified signal functions in the context of a monadic
+input/output system, such as Haskell's IO monad. The means providing a simple
+monadic interface for taking actions on a signal function in response to
+inputs, and for actuating the outputs of a signal function.
+
 The evaluation interface provides a modified state monad which holds a signal
-function, together with some additional information, as its state (shown in Figure~\ref{figure:evaluation_state}.
+function, together with some additional information, as its state (shown in
+Figure~\ref{figure:evaluation_state}).
 Rather than monadic instructions to put and get the state, the monad provides instructions

### 2013-01-29 Docs/Thesis/background.tex
Revisions for MTF

@@
 signal-function literature. This is unfortunate, since a {\em signal function}
-may operate on events, signals, or some combination of the two.}, which are
+may operate on events, behaviors, or some combination of the two.}, which are
 time-dependent values defined at every point in continuous time, and events,
 which are values defined at countably many points in time. An FRP system will
-provide combinators to manipulate events and signals and to react to events by
+provide combinators to manipulate events and behaviors and to react to events by
 replacing a portion of the running program in response to an event. Behaviors
@@
 
-The two general approaches to FRP are ``classic'' FRP, where behaviors and signals are first-class and reactive objects, and ``signal-function'' FRP,
-where transformers of signals and events are first-class and reactive objects.
+The two general approaches to FRP are ``classic'' FRP, where behaviors and
+events are first-class and reactive objects, and ``signal-function'' FRP, where
+transformers of signals and events are first-class and reactive objects.
 

### 2013-01-29 Docs/Thesis/comparison.tex
Revisions for MTF

--- 2012-10-30 Docs/Thesis/comparison.tex
@@
 
-\section{Evaluation Application}
-\label{section:Evaluation_and_Comparisons-Evaluation_Application}
-
-The application of choice is a learning switch controller for OpenFlow~\cite{OpenflowSpec}
-switches. this application must handle input events triggered by switches
-signalling that they have received a packet for which they have no handling
-rules, and produce output events which install handling rules for dealing with
-these packets.
-
-The reactive algorithm is simple. A table from machine addresses to switch ports
-is maintained. This table is updated in two ways. When a packet is sent to the
-controller, meta-data for the packet indicates which port it arrived on. The
-source address of the packet can then be mapped to that port in the table.
-Once this occurs, rules are installed for each packet with a destination address
-matching the source, along with reciprocal rules.
-
-Each mapping in the table has a timestamp. After a certain amount of time, the
-table entry is deleted. The table entry is also deleted if a packet is received
-with the same source address and a different port. When an entry is deleted,
-the switch is signaled to remove all corresponding rules.
-
-In both Yampa and TimeFlies, this algorithm is implemented by using an accumulator,
-which applies functions carried by event occurrences to a state value. It stores
-the new state and outputs new event occurrences based on the output of the function.
-
-The state stored is the table (implemented using the {\tt Data.Map} module in the
-Haskell {\tt containers} library) and the output event occurrences carry messages
-to be sent to the switch. The events are sourced internally to the learning
-switch signal function (using the {\tt after} combinator in TimeFlies and the {\tt repeatedly}
-function in Yampa) to delete expired rules, and externally, by transforming messages
-from the switch to the controller into accumulator functions.
-
 \section{Methodology and Results}
@@
 
-The controllers were benchmarked using the {\tt cbench} program~\cite{cbench}. The
-amount to time to pause between time and signal updates was passed as a parameter to
-both programs. The {\tt cbench} benchmark operates by sending a simulated packet in
-event, waiting for a rule response, and then repeating. The output of the benchmark
-is the number of times this occurs in a second. This is thus a very good test of
-event latency in a reactive system.
+The example application from Chapter~\ref{chapter:Example_Application}, an
+OpenFlow controller, was also implemented in Yampa. A slightly-modified
+TimeFlies implementation, which used a polling loop rather than the GHC event
+system to obtain events, but still "pushed" events to TimeFlies once they were
+acquired, was also implemented.
+
+These threee controllers were benchmarked using the {\tt cbench}
+program~\cite{cbench}. The amount to time to pause between time and signal
+updates (the inverse of the sampling rate for signals) was passed as a parameter
+to each program. The {\tt cbench} benchmark operates by sending a simulated
+packet-in event, waiting for a rule response, and then repeating. The output of
+the benchmark is the number of times this occurs in a second. This is thus a
+test of event latency in a reactive system.
 
@@
 even at a very high sampling rate. This is due to two factors. First, TimeFlies can react
-to individual events without evaluating the whole signal function. Second, the
-interface of time-flies permits us to connect its evaluation directly to GHC's event
-manager, while the design of Yampa requires us to use threads and synchronized
-communication\footnote{Yampa does include a step-wise evaluation function, but
-it still couples event evaluation firmly to time steps, and is not well
-documented.}.
+to individual events without evaluating the whole signal function. Second, in
+the implementation described in Chapter~\ref{chapter:Example_Application}, the
+interface of TimeFlies permits us to connect its evaluation directly to GHC's
+event manager, while the design of Yampa requires us to use threads and
+synchronized communication\footnote{Yampa does include a step-wise evaluation
+function, but it still couples event evaluation firmly to time steps, and is not well
+documented.}. Even the TimeFlies implementation using polling IO outperforms the
+Yampa version, due to the ability to evaluate multiple events per time step.

### 2013-01-29 Docs/Thesis/conclusions-and-further-work.tex
Revisions for MTF

@@
 \label{section:Conclusions_and_Further_Work-Conclusions}
-The TimeFlies system is a fully-implemented and -documented FRP library which
+The TimeFlies system is a fully-implemented and fully-documented FRP library which
 may be extended with utility functions and further optimized. Its performance in
@@
 modeled as a signal function, and these signal functions can be added to and removed from the
-collection.
+collection. An attempt was made to add such a function to TimeFlies. Currently,
+it is not clear what the type of such a combinator would be in a system
+employing signal vectors, or how the "routing" of inputs to individual signal
+functions in the collection would be specified.
 

### 2013-01-29 Docs/Thesis/introduction.tex
Revisions for MTF

@@
 writing interactive programs in functional languages. The FRP abstractions are
-behaviors or signals, which are functions of continuous time, and events, which
-are sequences of time-value pairs. These abstractions translate desirable
-properties of the underlying functional languages, such as higher-order
-functions and referential transparency, to reactive constructs, generally
-without modifications to the underlying language. This permits the compositional
-construction of reactive programs in purely functional languages.
+{\em behaviors} (also termed {\em signals}), which are functions of continuous
+time, and {\em events}, which are sequences of time-value pairs. These
+abstractions translate desirable properties of the underlying functional
+languages, such as higher-order functions and referential transparency, to
+reactive constructs, generally without modifications to the underlying language.
+This permits the compositional construction of reactive programs in purely
+functional languages.
 
-Functional reactive programming was introduced, though not quite by name, with Fran~\cite{Elliott1997}, a compositional system for writing
-reactive animations. From the start, two key challenges emerged: the enforcement of {\em causality}\footnote{Causality is the property 
-that a value in an FRP program depends only on present or past values. Obviously, a program which violates this property is impossible to
-evaluate, but in some FRP systems such programs are unfortunately quite easy to write down.} in FRP programs, and the efficient
-evaluation of FRP programs.
+Functional reactive programming was introduced, though not quite by name, with
+Fran~\cite{Elliott1997}, a compositional system for writingm reactive
+animations. From the start, two key challenges emerged: the enforcement of
+{\em causality}\footnote{Causality is the property that a value in an FRP
+program depends only on present and past values. Obviously, a program which
+violates this property is impossible to evaluate, but in some FRP systems such
+programs are unfortunately quite easy to write down.} in FRP programs, and the
+efficient evaluation of FRP programs.
 
 The first challenge was addressed by representing FRP programs, not as
-compositions of signals and events, but as compositions of transformers of
+compositions of behaviors and events, but as compositions of transformers of
 behaviors and events called {\em signal functions}. By not permitting direct
@@
 To address both challenges in a FRP implementation, we combine the approaches
-and demonstrate a push-based signal-function FRP system. The current
+and demonstrate a push-pull signal-function FRP system. The current
 implementation approach for signal-function FRP is inherently
@@
 which suggests a method of push-based evaluation. In previous implementations of
-FRP, events were simply option-valued signals. This approach has the advantage
-of being simple to implement. However, it does not have an obvious semantics for
-event merging. Further, it necessitates pull-based evaluation, since there is no
-way to separate the events which may be pushed from the truly continuous values
-which must be repeatedly sampled.
+signal-function FRP, events were simply option-valued signals. This approach has
+the advantage of being simple to implement. However, it does not have an obvious
+semantics for event merging. Further, it necessitates pull-based evaluation,
+since there is no way to separate the events which may be pushed from the truly
+continuous values which must be repeatedly sampled.
 

### 2013-01-29 Docs/Thesis/system_design_and_interface.tex
Revisions for MTF

@@
 
-The evaluation monad is a monad transformer. This permits it to be used in
-conjunction with the {\tt IO} monad (or any other monad) to describe how input
-is to be obtained for the signal function being evaluated, and how outputs are
-to be handled.
+In TimeFlies, the evaluation monad is a monad transformer. This permits it to be
+used in conjunction with the {\tt IO} monad (or any other monad) to describe how
+input is to be obtained for the signal function being evaluated, and how outputs
+are to be handled.
 
@@
 The type of the evaluation monad must track the input and output type of the
-signal function. The monad's context stores a mapping from outputs to handling
-actions. There are thus four type parameters to the monad's type
-constructor: the input signal vector, the output signal vector, the type of the
-underlying monad, and the monadic type parameter. The type is shown in
-Figure~\ref{figure:evaluation_monad_types}.
+signal function. The monad is a state monad, and its state is a record with
+fields for a mapping from outputs to handling actions, the current state of the
+signal function, the last time the signal function was sampled, and the
+currently-unsampled input. There are thus four type parameters to the monad's
+type constructor: the input signal vector (to type the signal function and
+unsampled input), the output signal vector (to type the signal function and
+output handlers), the type of the underlying monad, and the monadic type
+parameter. The type is shown in Figure~\ref{figure:evaluation_monad_types}.
 

### 2013-01-29 Docs/Thesis/background.tex
Clarify definition of signal functions in background.

@@
 for {\em signal functions}. Arrows are abstract type constructors with input and
-output type parameters, together with a set of routing combinators (see Figure~\ref{figure:arrow_combinators}).
-Signal functions form an arrow. Signal functions represent time-dependent and
-reactive transformers of events and signals. Signals and events
-cannot be directly manipulated by the programmer. This approach has two
-motivations: it increases modularity since both the input and output of signal
-functions may be transformed (as opposed to signals or events which may only be
-transformed in their output) and it avoids a large class of time and space leaks
-which have emerged when implementing FRP with first-class signals and events.
+output type parameters, together with a set of routing combinators (see
+Figure~\ref{figure:arrow_combinators}). The concept of an Arrow in Haskell,
+including the axioms the combinators of an arrow must satisfy, are derived from
+the concept of arrows from category theory.
+
+Signal functions are time-dependent and reactive transformers of events and
+signals. Signals and events cannot be directly manipulated by the programmer.
+This approach has two motivations: it increases modularity since both the input
+and output of signal functions may be transformed (as opposed to signals or
+events which may only be transformed in their output) and it avoids a large
+class of time and space leaks which have emerged when implementing FRP with
+first-class signals and events.
 

### 2013-01-30 Docs/Thesis/appendix-haskell.tex
ANH thesis revisions

\chapter{Haskell Concepts}
\label{chapter:Haskell_Concepts}

One of the primary attractions of the Haskell language, and the reason for its
use throughout this work, is its advanced yet practical and usable type system.
This type system enables the use of compositional software design that would
be rendered infeasible without a type system to both inform and verify
composition and implementation. This appendix gives an overview of Haskell
concepts, design patterns, and idioms which are used in this thesis.

\section{Datatypes and Pattern Matching}
\label{section:Haskell_Concepts-Datatypes_and_Pattern_Matching}

In Haskell, new type constructors are introduced by defining Algebraic Datatypes.
An ADT declaration can take one of two forms. The first is a {\tt data}
declaration, e.g.:

\begin{code}
data Bool = True | False
data Maybe a = Just a | Nothing
\end{code}

In this form the identifier(s) preceding the {\tt =} character are a type
constructor followed by zero or more type variables. Following the {\tt =} 
character, and separated by {\tt |} characters, are data constructors. Each data
constructor is an identifier followed by zero or more types, which are the types
of its arguments. Data constructors can be used in expressions to construct a
value of this newly-declared type constructor, and in pattern matching to
``take apart'' the value and observe its components (the arguments to the data
constructor.

The second form is the {\tt newtype} declaration. This form is more restricted.
It is limited to one data constructor with exactly one argument. It introduces
a new type without introducing a new runtime restriction, though the Haskell
code must still use the data constructor in pattern matches and expressions
to explicitly coerce between the new type and the type of its data constructor's
parameter. This behavior is most often used to hide implementation types without
introducing the runtime overhead of value construction and pattern-matching, as
these are erased for constructors declared using {\tt newtype} once type-checking
is complete.

Once a type constructor has been introduced, it can be used in a type, with
its arguments replaced by any valid Haskell type. For instance:

\begin{code}
not      :: Bool -> Bool
cbool    :: Bool -> Int
maybeInt :: Int -> Maybe Int -> Int
isJust   :: Maybe a -> Bool
mapMaybe :: (a -> b) -> Maybe a -> Maybe b
\end{code}

Its data constructors can be used in the patterns of functions, and in their
expressions. For instance:

\begin{code}
not :: Bool -> Bool
not True  = False
not False = True

isJust :: Maybe a -> Bool
isJust (Just _) = True
isJust Nothing  = False

mapMaybe :: (a -> b) -> Maybe a -> Maybe b
mapMaybe f (Just x) = Just (f x)
mapMaybe _ Nothing  = Nothing
\end{code}

\subsection{Generalized Algebraic Datatypes}
\label{subsection:Haskell_Concepts-Datatypes_and_Pattern_Matching-Generalized_Algebraic_Datatypes}

GADTs~\cite{Cheney2003,Xi2003} permit us to specify what types fill in the type
parameters for specific constructors. For instance, if we wish to build a tiny
expression language, we can use a standard ADT:

\begin{code}
data Exp a = Const a 
           | Plus (Exp a) (Exp a)
           | LessThan (Exp a) (Exp a)
           | If (Exp a) (Exp a) (Exp a)
\end{code}

Let us assume for the moment that Haskell exports functions\footnote{The types
for addition and comparison actually involve a typeclass constraint, but the
point is that the functions' types are not parametric.}:

\begin{code}
(+)  :: Int -> Int -> Int
(<)  :: Int -> Int -> Bool
\end{code}

An attempt to write an evaluation function for our expression type is:

\begin{code}
eval :: Exp a -> a
eval (Const x)  = x
eval (Plus x y) = eval x + eval y
eval (LessThan x y) = eval x < eval y
eval (If p c a) = if (eval p)
                  then eval c
                  else eval a
\end{code}

But this function will not typecheck. In the {\tt Plus} case, we cannot
assume that the argument to the type constructor {\tt Exp} typing {\tt x} or {\tt y}
is {\tt Int}, and similarly for the {\tt LessThan} case. Again in the {\tt If}
case, we cannot assume that the predicate is of type {\tt Exp Bool}, and if we
could, that would force our results to be of type {\tt Exp Bool} as well.

Let's try a slightly modified ADT:

\begin{code}
data Exp a = Const a
           | Plus (Exp Int) (Exp Int)
           | LessThan (Exp Int) (Exp Int)
           | IfThenElse (Exp Bool) (Exp a) (Exp a)
\end{code}

Here the motivation for a type parameter to our type constructor becomes clear.
We can introduce both {\tt Bool} and {\tt Int} constants (as well as others, but
we cannot do anything with them unless we extend the language). Further, we
can constrain the types of the input expressions to each of our constructors to
be of the appropriate type.

The code for our our evaluator is the same. But now note that the in the
{\tt Plus} and {\tt LessThan} cases, even though the input types are compatible
with the functions used, the output type expected from our function is not.
{\tt Plus x y} has type {\tt Exp a} in the pattern match, so our output is
expected to be of type {\tt a} for any {\tt a} argument to type of an 
input {\tt Exp}.

Here is our expression type as a GADT:
\begin{code}
data Exp a where
  Const    :: a                               -> Exp a
  Plus     :: Exp Int  -> Exp Int             -> Exp Int
  LessThan :: Exp Int  -> Exp Int             -> Exp Bool
  If       :: Exp Bool -> Exp a   -> Exp a    -> Exp a
\end{code}

Now our evaluation function can typecheck. Each constructor is able to constrain
the type parameter for output, not just its arguments. So when pattern matching
on the {\tt Plus} case, we know that each of our inputs will be of type
{\tt Exp Int}, and that the type of the expression we are pattern matching has
type {\tt Exp Int}, so the output from our function can be constrained to type
{\tt Int}, similarly for {\tt LessThan}. The type argument to {\tt Exp} in the
type of the {\tt If} is constrainted to be the same as that as the argument
to the types of the consequent and alternate to the conditional. This permits
our {\tt If} statement to be parametric while still allowing our evaluation to
typecheck.

This capacity to constrain the output types of data constructors, and thus, to
constrain the types of expressions in the scope of pattern matches of these data
constructors, is called {\em type refinement}. We will make use of this ability
to parameterize concrete datatypes over abstract type structures, rather than to
permit typechecking in specific cases, but the principle is the same.

\section{Typeclasses}
\label{section:Haskell_Concepts-Typeclasses}

Typeclasses in Haskell provide a means to implement functions that are openly
polymorphic while not being parametric. A typeclass is declared as follows:

\begin{code}
class Show t where
  show :: t -> String
\end{code}

A typeclass has an identifier and a single type parameter. This type parameter
is used in the type of one or more functions which are members of the class.

The class can then be instantiated:

\begin{code}
instance Show Bool where
  show True  = "True"
  show False = "False"

instance (Show a) => Show (Maybe a) where
  show (Just x) = "Just " ++ show x
  show Nothing  = Nothing
\end{code}

Functions can now be written polymorphically over the types instantiating the
typeclass, by including the typeclass as a constraint:

\begin{code}
repL :: (Show t) => t -> Int
repL x = length (show x)
\end{code}

Typeclasses are used in Haskell to provide common interfaces or functionality
across types. The {\tt Show} class used as an example is exported, along with
instances for most of the basic Haskell types, from Haskell's Prelude (the
standard module imported into every Haskell module).

\section{Monads and Monad Transformers}
\label{section:Haskell_Concepts-Monads_and_Monad_Transformers}

One of the primary concepts employed in Haskell programs is that of the
monad~\cite{PeytonJones1993,PeytonJones2001}. The concept of the monad is
borrowed from category theory, but it is quite simple when understood within
Haskell. A monad is a type constructor with a single parameter, and two
associated functions. In Haskell's {\tt Monad} typeclass, these functions are
denoted {\tt return} and {tt (>>=)}.

\begin{code}
class Monad m where
  return :: a -> m a
  (>>=)  :: m a -> (a -> m b) -> m b
\end{code}

A monad must obey the following axioms:
\begin{itemize}
\item Left identity: {\tt return x >>= f = f x}
\item Right identity: {\tt m >>= return = m}
\item Associativity: {\tt (a >>= b) >>= c) = (a >>= (\\x -> b x >>= c))}
\end{itemize}

Several standard Haskell type constructor are monads in interesting ways, but
the most well-known is Haskell's {\tt IO} type constructor. This is the basis of
Haskell's input/output system. The entry point to a Haskell program is the
{\tt main} function, of type {\tt IO ()}. This function can be constructed by
using the monadic functions to sequence an arbitrary number of other functions
whose output type is {\tt IO a}. Since the sequencing operator takes an
arbitrary function, this allows the full power of Haskell functions, including
first-class and higher-order functions, to be employed in defining a program's
input and output. A convenience function is commonly used when the result is not
necessary as part of the sequencing:

\begin{code}
(>>) :: Monad m => m a -> m b -> m b
(>>) m1 m2 = m1 >>= (const m2)

\subsection{Do-notation}
\label{subsection:Haskell_Concepts-Monads_and_Monad_Transformers-Do_notation}

Because monads are such a pervasive concept in Haskell, the language includes
special syntax for writing monadic expressions. Do-notation is expression syntax
which begins with the keyword {\tt do} and is followed by lines of two forms:

\begin{code}
do
  x <- m1
  m2
\end{code}

The first form is a binding expression: it binds the variable {\tt x} to the
output of the monadic value {\tt m2}. The second form simply sequences the monad
value {\tt m2} into the monadic value being built. Do-notation has a syntax-driven
translation to desugared Haskell expression:

\begin{code}
desugar {
do x <- m1
   ...} = 
m1 >>= \x -> desugar {do ...}

desugar {
do m1
   ...} =
m1 >> desugar {do ...}
\end{code}

\subsection{Monad Transformers}
\label{subsection:Haskell_Concepts-Monads_and_Monad_Transformers-Monad_Transformers}
A monad transformer is a type constructor with two parameters. The first
parameter parameterizes over a one-parameter type constructor, rather than a
type. The second is the monadic type parameter. A type constructor {T} is a monad
transformer if it has the following instance of the Monad typeclass:

\begin{code}
instance Monad m => Monad T m
\end{code}

and is also an instance of the class

\begin{code}
class MonadTrans t where
  lift :: (Monad m) => m a -> t m a
\end{code}

The axioms for the lift function are:
\begin{itemize}
\item {\tt lift . return = return}
\item {\tt lift (m >>= f) = lift m >>= (lift . f)}
\end{itemize}

Restated, {\tt lift} does not modify return and distributes over monadic
sequencing.

As an example of a monad transformer, we can consider the {\tt StateT} type,
which is employed in the implementation of this thesis.

The type is declared:
\begin{code}
newtype StateT s m a = S { runStateT :: s -> m (s, a) }
\end{code}

Its monad instance, for any {\tt s}, is

\begin{code}
instance Monad m => Monad (StateT s m) where
  return x  = S (\s -> return (x, s))
  (>>=) (S f) mf = S (\s -> f s >>= (\ (x, s') -> let (S f') = mf x in f' s ))
\end{code}

The return and sequencing functions carry the state through the underlying monad.

The {\tt MonadTrans} instance is:
\begin{code}
instance MonadTrans (StateT s) where
  lift m = S (\s -> m >>= (\x -> return (x, s))
\end{code}

Finally, there are two functions provided to access and set the state:

\begin{code}
put :: s -> StateT s m ()
put = S (\_ -> return ((), s'))

get :: StateT s m s
get = S (\s -> return (s, s))
\end{code}

If we use {\tt StateT} as a wrapper around the IO monad, we might employ it as
a way to generate a unique line number for each "putStrLn" we call.

\begin{code}
putStrLnN :: String -> StateT Int IO ()
putStrLnN s = do i <- get
                 put (i + 1)
                 lift (putStrLn (show i ++ " " ++ s))

main = runStateT mainSt 1
  where mainSt = do g <- lift getLine
                    putStrLnN g
                    mainSt
\end{code}

\section{Glossary of Type Terminology}
\label{section:Haskell_Concepts-Glossary_of_Type_Terminology}
\begin{description}
\item[ADT] See {\em Algebraic Datatype}.

\item[Algebraic Datatype] An Algebraic Datatype or ADT is a type whose terms
are {\em data constructors}. An ADT is defined by naming a
{\em type constructor} and its parameters (zero or more)
(as {\em type variables}), together with one or more data constructors and the
types of their members. Each data constructor takes a fixed number
(zero or more) data members, whose types are given following the constructor
name. These types are defined in terms of the type variables named as parameters
of the type constructor and any type constructors (including the type
constructor associated with the ADT) in scope in the module.

\item[Data Constructor] A Data Constructor is a component of an
Algebraic~Datatype which, when applied to values of the appropriate type, 
creates a value typed with the ADT. Data constructors are the primary element
which may be pattern matched in languages such as Haskell.

\item[GADT] See {\em Generalized Algebraic Datatype}.

\item[Generalized Algebraic Datatype] Similar to an {\em Algebraic Datatype},
but {\em type variables} in the {\em type constructor} declaration serve merely
to denote the number of type parameters (and thus may be replaced by a
{\em kind signature}) and types are given for each {\em data constructor}. These
types must have the type constructor as their top-level term, but may fill in
the parameters of the type constructor with arbitrary types. Variables which
appear in the data member types but on in the data constructor type are
{\em existentially quantified}, and types appearing in the data constructor
type but not the data member types may be instantiated arbitrarily.

\item[Kind] A ``type of types.'' Kinds are used to verify that types are
consistent during typechecking. The kind of types which contain values is
{\tt *}, and the kind of single-parameter type constructors which take a
type is {\tt * -> *}. Other kinds may also be introduced. For instance,
signal vectors should be their own separate kind, but the Haskell type mechanism
was not mature enough to support this at the time of this writing..

\item[Kind Signature] A means of specifying the number and kind of types which
may instantiate type variables. Type variables in Haskell are not restricted to
types, but may be instantiated by type constructors as well. The kind of a
variable restricts what it may be instantiated with. A kind signature gives
kinds to a type constructor, and thus to its parameters. Specifying the kind
of a type constructor perfectly constrains the number of parameters.

\item[Type Constructor] A type constructor is a type level term which, when
applied to the proper number of types, produces a type. Type constructors,
together with {\em type variables}, form the basis of polymorphism in Haskell
and similar languages.

\item[Type Variable] A type variable is a type-level term which may be 
instantiated (by the typechecker via inference, or by the user via annotation)
with any type at the point where the value so typed is used. 
Together with {\em type constructors}, type variables form the basis of
polymorphism in Haskell and similar languages.
\end{description}

### 2013-01-30 Docs/Thesis/implementation-evalinterface.tex
ANH thesis revisions

@@
 system for external IO.
-
-The evaluation interface translates between signal functions and the standard
-Haskell construct for sequencing effects and external inputs, namely,
-{\em monads}~\cite{PeytonJones1993}. The inspiration for monads is drawn from
-the rather esoteric domain of category theory, but the concept as applied to
-programming languages is actually rather simple.
-
-In a programming language with type constructors, a monad is simply a 1-arity
-type constructor together with two operations. We shall call this type
-constructor $m$. The first operation takes a value of any type
-(call the type $a$) and produces a value of type $m \: a$ ($m$ applied to $a$).
-The second operation takes a value of type $m \: a$ and a function from $a$ to
-$m \: b$, and produces a $m \: b$. This means that this second operation,
-monadic application, applies the function to the value in the context of the
-monad.
-
-For our purposes, there are two interesting properties of monads. One is that
-for a specific type constructor, opaque primitives may be defined which
-can be used, along with return, to build functions for the second argument
-of the monadic application operation. In our case, we would like operations
-to push an event to a signal function, to update the signal components of
-the signal function, and to increment the time and sample the signal function.
-
-This property is what enables monads to be used for input and output. A Haskell
-{\tt main} program is an IO action. An IO action is a value of the monadic type
-IO. This action is comprised of a primitive action (either a {\tt return} or a 
-true IO primitive), possibly together with a function from the output of that
-action to another IO action. Note that values with the IO type may appear
-anywhere in a Haskell program, but they are only execute when they are returned
-as a part of the {\tt main} IO program.
-
-The Haskell typeclass for monads is defined:
-
-\begin{code}
-class Monad m where
-  return :: a -> m a
-  -- | Infix operator for monadic application
-  (>>=)  :: m a -> (a -> m b) -> m b 
-\end{code}
-
-The other interesting property of monads is the existence of a class of monads
-called monad transformers. These are arity-2 type constructors which take as
-their first parameter a monad type constructor. The partial application of
-such a type constructor to any monad type constructor then results in a monad.
-A construct is then provided to lift values of the underlying monad to values
-in the transforming monad, which may then be sequenced as normal. This leads
-to the powerful concept of monad transformer stacks, where monads with many
-different capabilities are combined without any more plumbing than explicit
-{\tt lift} operations.
-
-This property will be useful because we want our evaluation interface to be
-part of a monad stack, the base of which is Haskell's IO monad. By formulating
-the evaluation interface as a monad transformer, we need only define the
-operations relevant to the evaluation of signal functions, and we can depend
-on the constructs of the IO monad to interact with whatever inputs and outputs
-are necessary. In some cases, we may not wish to use the IO monad at all (e.g
-for testing or simulation). In this case, we can parameterize over another
-monad, such as the Identity monad (which has no special operations and whose
-context is just the value), or the State monad (which maintains an implicit 
-state accessible by {\tt get} and {\tt put} operations).
-
-The monad tranformer typeclass is defined as:
-
-\begin{code}
-class MonadTrans t where
-  lift :: Monad m => m a -> t m a
-\end{code}
-
-Haskell also supports {\em do}-notation for monads. Do notation simply
-makes monadic application implicit between lines and reverses the syntatic
-order of lambda bindings, providing a way to use monads in a manner similar
-to imperative programming. (See Figure~\ref{figure:monad-example} for an
-example.)
-
-\begin{figure}
-\begin{code}
--- Do notation:
-do putStr "Hello: "
-   string <- getStr
-   putStrLn $ "Nice to meet you " ++ string
-   time <- getCurrentTime
-   print time
-   putStrLn "Look at the time! Bye."
-
--- Translates to:
-putStr "Hello: " >> getStr >>= 
-(\string -> (putStrLn $ "Nice to meet you " ++ string) >>
- getCurrentTime >>=
- (\time -> print time >> putStrLn "Look at the time! Bye."))
-\end{code}
-\hrule
-\caption{An example of monadic $do$-notation and its equivalent monadic
-expression.}
-\label{figure:monad-example}
-\end{figure}
 

### 2013-01-30 Docs/Thesis/implementation-signalfunctions.tex
Take care of revisions from mtf
ANH thesis revisions

@@
 types, along with higher-order functions and continuations.
-
-\subsection{Generalized Algebraic Datatypes}
-\label{subsection:Implementation-Signal_Functions-Generalized_Algebraic_Datatypes}
-The first consideration is how to make use of the information encoded in the
-signal vectors used as the input and output types of signal functions. In order
-to parameterize over these types, we turn to an extension of the Haskell type
-system known as Generalized Algebraic Datatypes (GADTs).
-
-Algebraic Datatypes (ADTs) are the means by which new primitive types are
-introduced into Haskell programs. For instance, the following datatypes are
-declared in the Haskell Prelude (the Haskell module which is in scope for every
-Haskell program):
-
-\begin{code}
-data Bool    = True   | False
-data Maybe a = Just a | Nothing
-\end{code}
-
-The first example is simply a set of alternatives, representing the values in
-Boolean algebras. These alternatives may be pattern-matched in a Haskell
-function either at the top level:
-
-\begin{code}
-boolToString :: Bool -> String
-boolToString True  = "Boolean True"
-boolToString False = "Boolean False"
-\end{code}
-
-Or in a {\tt case} expression:
-
-\begin{code}
-boolToString :: Bool -> String
-boolToString b = "Boolean " ++ case b of
-                                 True  -> "True"
-                                 False -> "False"
-\end{code}
-
-The second example is a parameterized ADT, or type constructor. The actual type
-is constructed by filling in a concrete type for the variable {\tt a}.
-Polymorphic functions can fill in quantified variables for this type or a
-component of it. However, there are no restrictions on what types may be used.
-In the {\tt Just} data constructor, the use of the type variable indicates a
-parameter to the data constructor which must take the type filled in for {\tt a}.
-
-This allows the {\tt Maybe} type to generalize the pattern of an optional value.
-The type {\tt Maybe Int} represents 0 or 1 integer values, and the type
-{\tt Maybe String} represents 0 or 1 strings. In order to consume values of these
-optional types, we can pattern match\footnote{The underscore character {\tt\_} 
-represents an input to the pattern which we disregard. It is called a wildcard.}:
-
-\begin{code}
-stringMaybe :: String -> Maybe String -> String
-stringMaybe s Nothing = s
-stringMaybe _ (Just s)  = s
-\end{code}
-
-But since we can fill in a type variable for the parameter of the Maybe type
-constructor, we can generalize this function:
-
-\begin{code}
-maybe :: a -> Maybe a -> a
-maybe x Nothing  = x
-maybe _ (Just x) = x
-\end{code}
-
-GADTs permit us to specify what types fill in the type parameters for specific
-constructors. For instance, if we wish to build a tiny expression language,
-we can use a standard ADT:
-
-\begin{code}
-data Exp a = Const a | Plus (Exp a) (Exp a)
-\end{code}
-
-Let us assume for the moment that the Haskell addition function is typed:
-
-\begin{code}
-(+) :: Int -> Int -> Int
-\end{code}
-
-The type actually involves a typeclass constraint, but the point is that the
-function's type is not parametric.
-
-An attempt to write an evaluation function for our expression type is:
-
-\begin{code}
-eval :: Exp a -> a
-eval (Const x) = x
-eval (Plus x y) = eval x + eval y
-\end{code}
-
-But this will not typecheck, since we cannot assume that {\tt a} is {\tt Int},
-so the output type of {\tt (+)} does not match the output type of our function,
-and the input types to {\tt (+)} do not match the input types to our function.
-
-Let's try a slightly modified ADT:
-
-\begin{code}
-data Exp a = Const a | Plus (Exp Int) (Exp Int)
-\end{code}
-
-The code for our our evaluator is the same. Now the input types to {\tt (+)}
-match, but the output type still does not.
-
-Here is our expression type as a GADT:
-\begin{code}
-data Exp a where
-  Const  ::  a -> Exp a
-  Plus   :: Exp Int -> Exp Int -> Exp Int
-\end{code}
-
-Now our evaluation function will typecheck. But why? The type of the {\tt Plus}
-constructor has had its {\em output} type restricted to {\tt Int}, and this
-allows us to assume, in the case where {\tt Plus} is matched, that the type
-{\tt a} has been restricted to {\tt Int}. Put another way, if a value of an
-{\tt Exp} type is not an {\tt Exp Int}, then the {\tt Plus} case will not occur.
-
-This capacity to constrain the output types of data constructors, and thus, to
-constrain the types of expressions in the scope of pattern matches of these data
-constructors, is called {\em type refinement}. We will make use of this ability
-to parameterize concrete datatypes over abstract type structures, rather than to
-permit typechecking in specific cases, but the principle is the same.
 

### 2013-01-30 Docs/Thesis/system_design_and_interface.tex
ANH thesis revisions

@@
 
-\subsection{Evaluation Monad}
-\label{section:System_Design_and_Interface-Types-Evaluation_Monad}
+\subsection{Evaluation Monad Transformer}
+\label{section:System_Design_and_Interface-Types-Evaluation_Monad_Transformer}
 
-A {\em monad} is a standard, composable abstraction for writing functions with
-a context, used in Haskell for IO~\cite{PeytonJones1993,PeytonJones2001} and
-other tasks. A monad is simply a 1-arity type constructor together with two
-functions. The first function, {\tt return}, takes a value of type {\tt a} and
-produces a value of type {\tt m a}. The second, called {\tt bind} and stylized
-in the Haskell standard library as the infix operator {\tt (>>=)}, takes a value
-of type {\tt m a} and a function of type {\tt a -> m b} and produces a value of
-type {\tt m b}. This allows a value to be operated on out of the context and a
-new context to be assigned.
-
-A monad can have other primitives which manipulate the context in some way. For 
-instance, the primtives in Haskell's {\tt IO} monad produce actions which, when
-interpreted as part of the {\tt main} action, produce some side-effect. The
-{\tt State} monad provides {\tt get} and {\tt put} operations to work with a 
-state value stored in the context.
-
-Monad transformers~\cite{Jones1995} provide a means to combine the functionality
-of multiple monads. A monad transformer is a monad with an extra type parameter.
-This type parameter is instantiated with the type constructor of the underlying
-monad, and an extra operation {\tt lift} is provided which converts values in
-the underlying monad to values in the monad transformer.
-
-In TimeFlies, the evaluation monad is a monad transformer. This permits it to be
-used in conjunction with the {\tt IO} monad (or any other monad) to describe how
-input is to be obtained for the signal function being evaluated, and how outputs
-are to be handled.
+TimeFlies provides a monad transformer for evaluating signal functions. Thi
+monad transformer may be used in conjunction with the {\tt IO} monad (or any
+other monad) to describe how input is to be obtained for the signal function
+being evaluated, and how outputs are to be handled.
 
@@
 The type of the evaluation monad must track the input and output type of the
-signal function. The monad is a state monad, and its state is a record with
-fields for a mapping from outputs to handling actions, the current state of the
-signal function, the last time the signal function was sampled, and the
-currently-unsampled input. There are thus four type parameters to the monad's
-type constructor: the input signal vector (to type the signal function and
-unsampled input), the output signal vector (to type the signal function and
-output handlers), the type of the underlying monad, and the monadic type
-parameter. The type is shown in Figure~\ref{figure:evaluation_monad_types}.
+signal function. The monad is a state monad, and its state is a record (the
+{\tt SFEvalState} type) with fields for a mapping from outputs to handling
+actions, the current state of the signal function, the last time the signal
+function was sampled, and the currently-unsampled input. There are thus four
+type parameters to the monad's type constructor: the input signal vector (to
+type the signal function and unsampled input), the output signal vector (to type
+the signal function and output handlers), the type of the underlying monad, and
+the monadic type parameter. The type is shown in
+Figure~\ref{figure:evaluation_monad_types}.
 

### 2013-07-17 Docs/Thesis/conclusions-and-further-work.tex
thesis updates

@@
 modeled as a signal function, and these signal functions can be added to and removed from the
-collection. An attempt was made to add such a function to TimeFlies. Currently,
-it is not clear what the type of such a combinator would be in a system
-employing signal vectors, or how the "routing" of inputs to individual signal
-functions in the collection would be specified.
+collection. An attempt was made to add such a function to TimeFlies, but it is not clear what 
+the type of such a combinator would be in a system employing signal vectors, or how the
+"routing" of inputs to individual signal functions in the collection would be specified.
 

### 2013-07-17 Docs/Thesis/implementation-signalfunctions.tex
thesis updates

@@
 data SVSample sv where
+  -- Store a value at an SVSignal leaf
   SVSample      ::    a
                    -> SVSample (SVSignal a)
+  -- Empty constructor for event leaves
   SVSampleEvent ::    SVSample (SVEvent a)
+  -- Empty constructor for empty leaves
   SVSampleEmpty ::    SVSample SVEmpty
+  -- Construct a sample from two samples, typed
+  --  with the SVAppend of their signal vectors
   SVSampleBoth  ::    SVSample svLeft
@@
 data SVOccurrence sv where
+  -- Leaf, carrying a value for the occurrence
   SVOccurrence ::    a
                   -> SVOccurrence (SVEvent a)
+  -- Lift an occurrence to be on the left side of a signal vector
   SVOccLeft    ::    SVOccurrence svLeft
                   -> SVOccurrence (SVAppend svLeft svRight)
+  -- Lift an occurrence to be on the right side of a signal vector
   SVOccRight   ::    SVOccurrence svRight 
@@
 data SVDelta sv where
+  -- A replacement value for a sample of a single signal
   SVDeltaSignal  ::    a
                     -> SVDelta (SVSignal a)
+  -- No replacement value for a sample of an arbitrary
+  --  signal vector
   SVDeltaNothing ::    SVDelta sv
+  -- Combine replacements for component signal vectors into 
+  --   replacements for the SVAppend of the vectors.
   SVDeltaBoth    ::    SVDelta svLeft
@@
 \begin{code}
+-- Type tag for signal functions which don't require
+--  an initial input
 data Initialized
 
+-- Type tag for signal functions which require
+--  an initial input
 data NonInitialized
 
+-- Type of signal functions
 data SF init svIn svOut where
+  -- Non-initialized signal functions consist of a function
+  --  from an input sample to a pair of an output sample
+  --  and an initialized continuation of the signal function
   SF     ::    (SVSample svIn 
@@
             -> SF NonInitialized svIn svOut
+
+  -- Initialized signal functions consist of two functions.
+  -- The first function's inputs are the time delta, and updates
+  --   to the input signals. Its outputs are updates to the output
+  --   signals, a set of output events, and a continuation of the
+  --   signal function.
   SFInit ::    (Double 
@@
                       SF Initialized svIn svOut))
+
+  -- The second function's input is an event occurrence for an
+  --   event in its input signal vector. Its outputs are a set 
+  --   of output occurrences for events in its output signal
+  --   vector, and a continuation of the signal function.
             -> (SVOccurrence svIn
@@
     (\evtOcc -> 
-      let (sf1Outputs, newSf1) = inputCont1 evtOcc
-          (sf2FoldOutputs, newSf2) = applySF sf2 sf1Outputs
-      in (sf2FoldOutputs, composeInit newSf1 newSf2)   
+      let (sf1Outputs, sf1New) = inputCont1 evtOcc
+          (sf2FoldOutputs, sf2New) = applySF sf2 sf1Outputs
+      in (sf2FoldOutputs, composeInit sf1New sf2New)   
     )
 
+-- The programmer should not rely on any 
 applySF ::    SF Initialized svIn svOut
@@
   foldr (\evtOcc (changes, SFInit _ changeCont) ->
-           let (newChanges, nextSF) = changeCont evtOcc
-               in (newChanges ++ changes, nextSF))
+           let (newChanges, sfNew) = changeCont evtOcc
+               in (newChanges ++ changes, sfNew))
         ([], sf)
@@
 
-The switch combinator is the means of introducing reactivity into a signal
+The {\tt switch} combinator is the means of introducing reactivity into a signal
 function. This combinator allows a signal function to replace itself by
@@
 
-The switch combinator stores the input sample provided during initialization,
+The {\tt switch} combinator stores the input sample provided during initialization,
 and updates it with the input signal updates. When the wrapped signal function
@@
 -- T.swap is imported from Data.Tuple
+-- chooseOccurrence, splitSample, splitDelta,
+--  combineSamples, combineDeltas, occRight, and
+--  occLeft are convenience functions for manipulating
+--  signal sample
 swap :: (sv1 :^: sv2) :~> (sv2 :^: sv1)
 swap =
-  SF ((, swapInit) . 
-      uncurry combineSamples .
-      T.swap . splitSample)
+  SF (\sigSample ->  
+       (uncurry combineSamples $ T.swap $ splitSample sigSample, 
+        swapInit)
 
@@
 swapInit =
-  SFInit (flip (const .
-                (, [], swapInit) .
-                uncurry combineDeltas .
-                T.swap . splitDelta))
+  SFInit (\_ sigDelta ->
+           (uncurry combineDeltas $ T.swap $ splitDelta sigDelta,
+            swapInit) 
          (\evtOcc ->
@@
 
-The implementation of the joining combinators is simple. The {\tt union}
-combinator simply passes along every event occurrence it receives on either
-input, stripping off the left and right combinator. This is acceptable since we
-do not insist on a total ordering of events, or an event time resolution greater
-than the sampling rate. The {\tt combineSignals} combinator maintains the values
-of both signals, and applies the combination function whenever one is updated.
-The {\tt capture} combinator maintains the input signal value, and adds it to
-each event occurrence.
+The {\tt union} combinator's implementation passes along every event occurrence 
+it receives on either input, stripping off the left and right combinator. This is 
+acceptable since we do not insist on a total ordering of events, or an event time
+resolution greater than the sampling rate. The {\tt combineSignals} combinator 
+maintains the values of both signals, and applies the combination function whenever
+one is updated, producing an update to the output signal's value.
+The {\tt capture} combinator similarly maintains the input signal value, and adds it
+to each event occurrence.
 
@@
 portions of a signal function to forgo evaluation during time steps unless they
-had signal updates. Optimizations in the style of Yampa, observed by keeping
-an updated AST for the signal function and pattern-matching on it when switching,
+had signal updates. Optimizations in the style of Yampa, where an AST for the
+current signal function is maintained alongside the continations and checked for
+opportunities to eliminate redundant components of the signal function,
 might further improve performance. In particular, collapsing nested or
-serially-composed versions of the {\tt switchWait} step when switching would
+serially-composed switches which have occurred since the last time step would
 remove at least some of the observed dependence of performance on sampling rate.

### 2013-07-17 Docs/Thesis/implementation.tex
thesis updates

--- 2012-10-30 Docs/Thesis/implementation.tex
@@
 Having established the design and semantics of the reactive system, I present a purely 
-functional implementation. For this implementation, I will make significant use
-of an extension to the Haskell type system. This extension, called
-GADTs~\cite{Cheney2003,Xi2003}, provides a mechanism for expressing abstractions
-which are not easily expressible in a standard Hindley-Milner type system, while
-maintaining the type soundness of Haskell. In particular, GADTs provide the
-power we require to take advantage of the signal vector representation of signal
-function inputs and outputs.
+functional implementation. For this implementation, I will make significant use of
+Generalized Algebraic Datatypes (GADTs)~\cite{Cheney2003,Xi2003}. GADTs provide a mechanism
+for expressing abstractions which are not easily expressible in a standard
+Hindley-Milner type system, while maintaining the type soundness of Haskell.
+In particular, GADTs provide the power we require to take advantage of the
+signal vector representation of signal function inputs and outputs.
 

### 2013-07-17 Docs/Thesis/system_design_and_interface.tex
thesis updates

@@
 -- Make right-associative
-associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))
+associateRight :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))
 
 -- Make left-associative
-unassociate :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
+associateLeft :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
 \end{code}
@@
 combining function is necessary because we will have two input samples at each
-time step, and must produce exactly one output sample.
+time step, and must produce exactly one output sample. Even if we restrict the types
+to be the same, parametricity forbids assuming that we can combine two values of
+an arbitrary type.\footnote{The {\tt union} combinator does not encounter this difficulty
+since event occurrences, being temporally discrete, can be temporally interleaved without
+ combining their individual values}
 

### 2013-08-19 Code/gloss-demo/src/Main.hs
Defense demo app

@@ -1,0 +1,21 @@
+{-# LANGUAGE TypeOperators #-}
+module Main where
+
+import FRP.TimeFlies.SignalFunctions
+import Graphics.Gloss.Data.Picture
+import Graphics.Gloss.Interface.IO.Game
+import Prelude hiding (filter)
+import Stopwatch (timeTracker)
+import World (WorldSVIn, WorldSVOut, GlossEvent (..), playSF)
+
+input :: WorldSVIn :~> SVEvent ()
+input = first ignore >>> cancelLeft >>> 
+        filter (\e -> case e of
+                        GlossEventKey (MouseButton _) Down _ -> Just ()
+                        _ -> Nothing)
+
+output :: SVSignal Double :~> WorldSVOut
+output = pureSignalTransformer (color black . text . show)
+
+main :: IO ()
+main = playSF (input >>> timeTracker 0 >>> output)

### 2013-08-19 Code/gloss-demo/src/Stopwatch.hs

@@ -1,0 +1,26 @@
+{-# LANGUAGE TypeOperators #-}
+module Stopwatch where
+
+import FRP.TimeFlies.SignalFunctions
+
+import World (GlossEvent(..))
+
+timeTracker :: Double -> (SVEvent a :~> SVSignal Double)
+timeTracker initialInterval = switch $ uncancelLeft >>> 
+                                       (second $ pureEventTransformer startTimeTracker) >>>
+                                       (first $ constant initialInterval)
+  where
+    startTimeTracker :: a -> (SVEvent a :~> SVSignal Double)
+    startTimeTracker _ = runningTimeTracker initialInterval
+
+runningTimeTracker :: Double -> (SVEvent a :~> SVSignal Double)
+runningTimeTracker initialTimeInterval = switch $ uncancelLeft >>> 
+                                                  first (time >>> 
+                                                         pureSignalTransformer (+ initialTimeInterval) >>> 
+                                                         copy) >>>
+                                                  associate >>>
+                                                  second (capture >>> pureEventTransformer stopTimeTracker)
+  where
+    stopTimeTracker :: (a, Double) -> (SVEvent a :~> SVSignal Double)
+    stopTimeTracker (_, t) = timeTracker t
+                                     

### 2013-08-19 Code/gloss-demo/src/World.hs

@@ -1,0 +1,52 @@
+{-# LANGUAGE TypeOperators #-}
+module World where
+
+import Control.Monad.Writer
+import Data.Monoid
+import FRP.TimeFlies.SignalFunctions
+import Graphics.Gloss
+import Graphics.Gloss.Interface.Pure.Game
+
+type WorldSVIn  = SVSignal (Float, Float) :^: SVEvent GlossEvent
+type WorldSVOut = SVSignal Picture
+
+type WorldSF = SF WorldSVIn WorldSVOut
+
+data GlossEvent = GlossEventKey Key KeyState Modifiers
+
+type World = (SFEvalState (Writer (Last Picture)) WorldSVIn WorldSVOut, Picture, Double)
+
+toWorldTransformer :: SFEvalT WorldSVIn WorldSVOut (Writer (Last Picture)) () -> World -> World
+toWorldTransformer m (st, p, t) =
+  let ((_, st'), l) = runWriter $ runSFEvalT m st
+      p'            = case l of
+                        Last Nothing  -> p
+                        Last (Just p') -> p'
+  in (st', p', t)
+
+handleEvent :: Event -> World -> World
+handleEvent e = toWorldTransformer $ case e of
+  EventKey k ks m pos -> do
+                           push $ svRight $ svOcc $ GlossEventKey k ks m
+                           update $ svLeft $ svSig pos
+  EventMotion pos     -> update $ svLeft $ svSig pos
+
+handleStep :: Float -> World -> World
+handleStep td (st, p, t) =
+  let t' = (realToFrac td) + t
+  in toWorldTransformer (step t') (st, p, t')
+
+playSF :: (WorldSVIn :~> WorldSVOut) -> IO ()
+playSF sf = play (InWindow "TimeFlies" (500, 500) (0, 0)) 
+                 white 
+                 20 
+                 (initSFEval (signalHandler $ tell . Last . Just) 
+                             (combineSamples (sample (0, 0)) sampleEvt)
+                             0
+                             sf,
+                  blank,
+                  0)
+                 (\(_, p, d) -> d `seq` p)
+                 handleEvent
+                 handleStep
+ 

### 2013-08-19 Code/gloss-demo/src/Stopwatch.hs
Remove extraneous import line from stopwatch demo source

@@ -3,8 +3,6 @@ module Stopwatch where
 
 import FRP.TimeFlies.SignalFunctions
 
-import World (GlossEvent(..))
-
 timeTracker :: Double -> (SVEvent a :~> SVSignal Double)
 timeTracker initialInterval = switch $ uncancelLeft >>> 
                                        (second $ pureEventTransformer startTimeTracker) >>>

### 2013-08-19 Docs/Thesis/background.tex
Final thesis revisions

@@
 The goal of FRP is to provide compositional and high-level abstractions for
-creating reactive programs. The key abstractions of FRP are behaviors or
-signals\footnote{Behaviors are generally referred to as {\em signals} in
+creating reactive programs. The key abstractions of FRP are behaviors
+\footnote{Behaviors are generally referred to as {\em signals} in
 signal-function literature. This is unfortunate, since a {\em signal function}
@@
 events are first-class and reactive objects, and ``signal-function'' FRP, where
-transformers of signals and events are first-class and reactive objects.
+transformers of behaviors and events are first-class and reactive objects.
 
@@
 
-The first implementation of FRP outside the Haskell language was Frapp\'{e}, 
+The first implementation of FRP outside the Haskell language was Frapp\'{e}~\cite{Courtney2001-2},
 implemented in the Java Beans framework. Frapp\'{e} built on the notion
@@
 sinks, and the implementation of Bean events and bound properties to propagate
-changes to the network~\cite{Courtney2001-2}.
+changes to the network.
 
@@
 
-The FrTime\footnote{FrTime is available in the current version of Dr Racket (5.2.1).}~\cite{Cooper2006}
+The FrTime\footnote{FrTime is available in all recent versions of Dr. Racket}~cite{Cooper2006}
 language extends the Scheme evaluator with a mutable dependency graph, which is
@@
 FrTime does not provide a distinct concept of events, and chooses branches of
-the dependency graph by conditional evaluation of signals, rather than by the
-substitution of signals used by FRP systems.
+the dependency graph by conditional evaluation of behaviors, rather than by the
+substitution of behaviors used by FRP systems.
 
@@
 Signal functions are time-dependent and reactive transformers of events and
-signals. Signals and events cannot be directly manipulated by the programmer.
+behaviors. Behaviors and events cannot be directly manipulated by the programmer.
 This approach has two motivations: it increases modularity since both the input
-and output of signal functions may be transformed (as opposed to signals or
+and output of signal functions may be transformed (as opposed to behaviors or
 events which may only be transformed in their output) and it avoids a large
 class of time and space leaks which have emerged when implementing FRP with
-first-class signals and events.
+first-class behaviors and events.
 
@@
 system discussed. This work introduced signal vectors, a typing construct
-which permits the distinction of signal and event types at the level of the FRP
-system, instead of making events merely a special type of signal.
+which permits the distinction of behavior and event types at the level of the FRP
+system, instead of making events merely a special type of behavior.
 

### 2013-08-19 Docs/Thesis/comparison.tex
Final thesis revisions

@@
 TimeFlies implementation, which used a polling loop rather than the GHC event
-system to obtain events, but still "pushed" events to TimeFlies once they were
+system to obtain events, but still ``pushed'' events to TimeFlies once they were
 acquired, was also implemented.
 
-These threee controllers were benchmarked using the {\tt cbench}
+These three controllers were benchmarked using the {\tt cbench}
 program~\cite{cbench}. The amount to time to pause between time and signal

### 2013-08-19 Docs/Thesis/conclusions-and-further-work.tex
Final thesis revisions

@@
 the type of such a combinator would be in a system employing signal vectors, or how the
-"routing" of inputs to individual signal functions in the collection would be specified.
+``routing'' of inputs to individual signal functions in the collection would be specified.
 

### 2013-08-19 Docs/Thesis/example.tex
Final thesis revisions

@@
 events and possibly produce output events. We employ the {\tt accumulateList}
-combinator to produce the signal function:
+combinator to produce the signal function shown in Figure~\ref{figure:switch_table_sf}.
 
@@
 which produces {\tt TableAccumulator} events. This signal function runs a timer
-(the {\tt every} combinator) and produces events carrying closures which scan
+(the {\tt every} combinator, which is implemented using {\tt switch} and {\tt after}
+and produces an event repeatedly on a given interval) and produces events carrying closures which scan
 the table for old rules, delete them, and produces messages to inform the
@@
           sample 
-          loop $ evtMgr      
+          loop evtMgr      
 \end{code}

### 2013-08-19 Docs/Thesis/implementation-signalfunctions.tex
Final thesis revisions

@@
 types is that they have {\em no data constructors}. There are no values which
-take these types.
+have these types.
 
@@
 produces an occurrence carrying a new signal function, that signal function is
-initialized with the stored input sample. It is then "wrapped" by another
+initialized with the stored input sample. It is then ``wrapped'' by another
 function which closes over its output sample, and outputs the sample as a signal
 update as the next time step. After this, it acts as the new signal function.
-This wrapping has some performance implications, which are discussed
+This wrapping has some performance implications, which are discussed in Chapter~\ref{chapter:Evaluation_and_Comparisons}.
 

### 2013-08-19 Docs/Thesis/system_design_and_interface.tex
Final thesis revisions

@@
 most libraries for time-dependent IO (e.g. audio and video systems) make use of
-the event loop abstraction. In this abstraction, event handlers are registered
+the event-loop abstraction. In this abstraction, event handlers are registered
 with the system, and then a command is issued to run a loop which detects events
@@
 
-TimeFlies provides a monad transformer for evaluating signal functions. Thi
+TimeFlies provides a monad transformer for evaluating signal functions. This
 monad transformer may be used in conjunction with the {\tt IO} monad (or any
@@
 \begin{code}
--- A signal function's evaluations state
+-- A signal function's evaluation state
 data SFEvalState svIn svOut m
@@
 into the network which are known when the signal function is created, without
-having to route those values from an input.
+having to route those values from an external input.
 
@@
 -- (not immediately observed)
-update :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
+update :: (Monad m) => SVSignalUpdate svIn -> SFEvalT svIn svOut m ()
 
